---
title: "Comparative Statistical Power of N-of-1 Trials versus
  Randomized Controlled Trials: A Simulation Study of Prazosin
  for PTSD Nightmares"
author:
  - pmsimstats team
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
abstract: |
  **Background:** N-of-1 trials offer a promising approach for personalised medicine by evaluating treatment effects within individual patients through repeated crossover designs. Their statistical properties relative to traditional randomised controlled trials remain incompletely characterised across clinically realistic effect-size and carryover regimes. **Methods:** We conducted an ADEMP-compliant Monte Carlo simulation study comparing the statistical power of aggregated N-of-1 hybrid trials versus parallel-group RCTs for detecting treatment effects of prazosin in reducing PTSD nightmares. The N-of-1 design enrolled 35 individuals each completing 8-period crossover trials analysed by a linear mixed-effects model with continuous-time AR(1) residuals; the RCT arm used ANCOVA at matched sample size $N = 35$. Carryover-contaminated placebo periods were excluded from the N-of-1 analysis before fitting. We ran $n_{\text{sim}} = 2{,}000$ replicates per cell, uniform across all primary and sensitivity analyses. **Results:** The aggregated N-of-1 hybrid dominated the parallel-group RCT across the full effect-size range. At a true effect of $-2.0$ nightmares per week, N-of-1 power was 1.000 versus 0.675 (MCSE 0.010) for the RCT; at $-1.0$, N-of-1 power was 0.830 (MCSE 0.008) versus 0.230 (MCSE 0.009). Bias was negligible in both designs. Coverage of nominal-95% CIs was 0.940--0.949 for the N-of-1 arm and 0.939 for the RCT arm across all effect sizes. **Conclusions:** Aggregated N-of-1 hybrid trials achieve substantially higher statistical power than parallel-group RCTs at matched sample size when within-subject carryover is managed by period exclusion. The efficiency advantage is largest in the small-to-moderate effect-size regime where parallel-group RCTs are essentially underpowered. **Keywords:** N-of-1 trials; randomised controlled trials; statistical power; crossover design; carryover effects; PTSD; prazosin; personalised medicine.
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: true
    number_sections: true
    keep_tex: true
    highlight: tango
fontsize: 11pt
geometry: "left=3cm,right=5cm,top=2cm,bottom=2cm"
bibliography: references.bib
csl: statistics-in-medicine.csl
link-citations: true
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{lineno}
  - \linenumbers
  - \usepackage{setspace}
  - \setstretch{1.1}
  - \providecommand{\rmdsource}{}
  - \providecommand{\rendertimestamp}{}
knit: (function(input, encoding) source(normalizePath(file.path(dirname(input), '..', '..', '..', 'tools', 'stage-render.R')))$value(input, encoding))
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 6,
  dpi = 300,
  fig.pos = "H",
  eval = TRUE
)

library(knitr)
library(kableExtra)

# Fallback stubs: populated by the load-simulation-results chunk.
# These ensure the document knits (with placeholder values) even
# before production-run.R has been executed.
n_of_1_power <- n_of_1_power_mcse <- rct_power <-
  rct_power_mcse <- n_of_1_bias <- rct_bias <- NA_real_
n_of_1_effects <- rct_effects <- numeric(0)
i_squared_mean <- tau_squared_mean <- NA_real_
prazosin_periods <- pure_placebo_periods <- numeric(0)
n_of_1_type1 <- n_of_1_type1_mcse <- NA_real_
rct_type1 <- rct_type1_mcse <- NA_real_
n_of_1_coverage <- n_of_1_coverage_mcse <- NA_real_
rct_coverage <- rct_coverage_mcse <- NA_real_
n_of_1_rel_err <- rct_rel_err <- NA_real_
exclusion_power <- explicit_power <- NA_real_
n_strategy_sims  <- NA_real_
sensitivity_n_sims <- NA_real_
power_diff <- NA_real_
decay_sensitivity_results <- list(power_range = NA_real_, robust = NA)
sensitivity_summary <- data.frame(
  Model = character(0), Power_Pct = numeric(0),
  Power_MCSE = numeric(0)
)
```

```{r footer-setup, echo = FALSE, results = 'asis'}
if (knitr::is_latex_output()) {
  src <- tryCatch(normalizePath(knitr::current_input()),
                  error = function(e) 'unknown')
  src_disp <- gsub('^/Users/zenn/Dropbox', '~/Dropbox', src)
  src_disp <- gsub('^/Users/zenn', '~', src_disp)
  src_tex <- gsub('([_~^&%#$])', '\\\\\\1', src_disp)
  ts <- format(Sys.time(), '%Y-%m-%d %H:%M %Z')
  cat(sprintf('\\renewcommand{\\rmdsource}{Source: \\texttt{%s}}\n',
              src_tex))
  cat(sprintf('\\renewcommand{\\rendertimestamp}{Rendered: %s}\n', ts))
}
```

# Abstract

**Background.** Aggregated N-of-1 trial designs and parallel-
group randomised controlled trials are the two principal
methodological alternatives for evaluating treatment effects
under substantial individual heterogeneity. The published
literature characterises the variance-component foundations of
each design separately, but has not delivered a comprehensive
sample-size and power comparison across clinically realistic
effect-size and carryover regimes that admits Monte Carlo
standard-error reporting in the @morris2019using ADEMP framework.

**Methods.** We conduct an ADEMP-compliant Monte Carlo
simulation study comparing the aggregated N-of-1 hybrid design
to a parallel-group RCT at matched sample size $N = 35$, across
a grid of true-effect sizes $\{0, -0.5, -1.0, -2.0\}$
nightmares per week, calibrated to the prazosin/PTSD reference
application. The hybrid N-of-1 fits an `nlme::lme` model with
continuous-time AR(1) residual correlation; the parallel-group
RCT fits a baseline-adjusted linear model (ANCOVA).
Carryover-contaminated placebo periods are excluded before
fitting the N-of-1 model. Each cell is run at
$n_{\text{sim}} = 2{,}000$ replicates; null cells (true effect
zero) provide explicit Type I error checks. Estimands include
power, empirical effect-size bias, empirical SE, model SE,
and nominal-95% CI coverage; all performance measures are
reported with Monte Carlo standard errors. Sensitivity sweep
S1 examines carryover half-lives in $\{0.5, 1, 3, 5, 7\}$ days.

**Results.** The aggregated N-of-1 hybrid dominated the
parallel-group RCT across the entire effect-size range.
At true effect $-2.0$, N-of-1 power was $1.000$ versus
parallel-group $0.675 \pm 0.010$; at $-1.0$, $0.830 \pm 0.008$
versus $0.230 \pm 0.009$; at $-0.5$, $0.311 \pm 0.010$ versus
$0.104 \pm 0.007$. The two power curves did not cross within
the simulated range; gaps exceeded five combined MCSEs at
every non-null effect size. The relative N-of-1-to-RCT power
ratio was approximately $3.0\times$ at the moderate-effect
$-0.5$ cell. Bias was negligible across all cells for both
designs. Coverage of nominal-95% CIs was 0.940--0.949 for
the N-of-1 arm and 0.939 for the RCT arm, each within two MCSEs
of the nominal level. The N-of-1 analysis was insensitive to the
true carryover half-life, confirming robustness of the
period-exclusion strategy. Type I error was $0.051 \pm 0.005$
for the N-of-1 hybrid and $0.061 \pm 0.005$ for the RCT; the
RCT rate is marginally above nominal, consistent with
small-sample effects in $\text{lm}$ at $N = 35$.

**Conclusions.** Aggregated N-of-1 hybrid trials achieve
substantially higher statistical power than parallel-group RCTs
at matched sample size when individual variability is
substantial and within-subject carryover is manageable. The
efficiency advantage is largest in the small-to-moderate
effect-size regime where parallel-group RCTs are essentially
powerless. Trial designers in personalised-medicine settings
with high inter-individual variability should consider the
aggregated N-of-1 design as the preferred default.

# Introduction

Clinical research has moved steadily toward personalized medicine
approaches that recognize substantial inter-individual variation
in treatment response [@lillieNof1ClinicalTrial2011;
@nam2021expanding]. Traditional randomized controlled trials
(RCTs), while remaining the standard for establishing
population-level treatment effects, may not capture the
individual-level treatment heterogeneity that is central to
personalized therapeutic decision-making
[@kravitzSinglepatientNof1Trials2014;
@marshallScopingReviewRandomized2022].

N-of-1 trials, also known as single-patient crossover trials,
evaluate treatment efficacy within individual patients through
repeated crossover periods [@Miller1999n;
@vohraCONSORTExtensionReporting2015]. In these designs each
patient serves as their own control, which reduces the
confounding effect of between-individual variability while
providing personalized treatment-effect estimates
[@zuckerIndividualNof1Trials2010; @schmidBayesianModelsNof12021].
When appropriately aggregated across patients, N-of-1 trials can
support population-level inference while simultaneously informing
individual treatment decisions [@chenComparisonAggregatedNof12020;
@araujoComparisonFourMethods2014].

## Methodological Considerations in N-of-1 Trial Design

The statistical analysis of N-of-1 trials presents unique methodological challenges that differ substantially from traditional parallel-group designs [@schmidStatisticalDesignAnalytic2014; @sennAnalysisContinuousData2024]. Key considerations include the appropriate handling of carryover effects, period effects, and temporal trends within individual patients [@tangTtestsNof1Trials2020; @yapBehavioralCarryoverEffect2024]. Carryover effects, representing the residual influence of treatment from previous periods, are particularly critical in crossover designs and can substantially impact both the validity and power of statistical analyses [@liuConsiderationsCrossoverDesign2021; @patsopoulosUnderstandingControlledTrials2011].

The aggregation of individual N-of-1 trials to provide population-level estimates requires sophisticated meta-analytic approaches that account for both within-patient and between-patient variability [@zuckerIndividualNof1Trials2010]. Random-effects meta-analysis methods, such as the DerSimonian-Laird approach, can appropriately combine individual treatment effect estimates while incorporating heterogeneity between patients [@chenMethodologicalReviewRandomised2024].

## Sample Size and Power Considerations

Sample size determination for N-of-1 trials involves complex considerations of the number of patients, number of crossover periods per patient, and the expected magnitude of within-patient versus between-patient variability [@arendsSampleSizeConsiderations2019]. Simulation studies have suggested that N-of-1 designs may achieve adequate statistical power with smaller total sample sizes compared to parallel-group RCTs, particularly when treatment effects are moderately large and individual variability is substantial [@blackstonComparisonAggregatedNof12019].

However, the relative statistical efficiency of N-of-1 trials compared
to RCTs remains context-dependent and has not been comprehensively
evaluated across varying effect sizes and carryover scenarios
[@chenComparisonAggregatedNof12020]. Understanding these relationships
is crucial for optimal study design selection in clinical research
applications.

### Analytic Frameworks for Power in N-of-1 Designs

The statistical power of an N-of-1 trial depends on a fundamentally
different set of variance components than a parallel-group RCT. Because
each patient serves as their own control, the between-patient variance
$\sigma^2_{\text{between}}$ is eliminated from the treatment effect
estimator, and the relevant sources of uncertainty are the
within-patient, between-period variance $\sigma^2$ and the residual
measurement error [@sennSampleSizeConsiderations2019;
@schmidStatisticalDesignAnalytic2014].

#### Variance decomposition and the paired-cycles model

Senn [-@sennSampleSizeConsiderations2019] formalized the sample size
problem for sets of N-of-1 trials under a Normal--Normal mixture model.
Suppose $n$ patients are each randomized in $k$ cycles of paired
treatment periods (each cycle containing one period of treatment A
and one of treatment B, in random order). Let $d_{ij}$ denote the
within-cycle treatment difference for patient $i$ in cycle $j$.
The model assumes

$$d_{ij} = \delta_i + \varepsilon_{ij}, \quad
  \delta_i \sim N(\Delta, \tau^2), \quad
  \varepsilon_{ij} \sim N(0, 2\sigma^2)$$

where $\Delta$ is the population mean treatment effect, $\tau^2$ is
the between-patient variance of individual treatment effects
(treatment-by-patient interaction), and $2\sigma^2$ is the
within-patient, within-cycle variance of the paired differences
[@sennSampleSizeConsiderations2019]. The factor of 2 arises because
$d_{ij}$ is the difference of two independent within-period
observations, each with variance $\sigma^2$.

Under this model, the patient-level mean difference
$\bar{d}_{i\cdot} = k^{-1}\sum_j d_{ij}$ has variance
$\tau^2 + 2\sigma^2/k$, and the grand mean
$\hat{\Delta} = n^{-1}\sum_i \bar{d}_{i\cdot}$ has variance

$$\text{Var}(\hat{\Delta}) =
  \frac{\tau^2 + 2\sigma^2/k}{n}$$

This expression is central to the trade-off between the number of
patients $n$ and the number of cycles per patient $k$: increasing
$k$ reduces only the $2\sigma^2/k$ component, whereas increasing
$n$ reduces the entire expression proportionally
[@zuckerIndividualNof1Trials2010;
@sennSampleSizeConsiderations2019].

#### Closed-form power for fixed- and random-effects analyses

Senn [-@sennSampleSizeConsiderations2019] distinguished two inferential
goals with distinct power formulas. For a *random-effects* analysis
targeting the population treatment effect $\Delta$, the patient-level
summary measures approach yields a one-sample $t$-test on the $n$
patient means $\bar{d}_{i\cdot}$ with $n - 1$ degrees of freedom.
The variance at the patient level is $\tau^2 + 2\sigma^2/k$,
estimated directly from the $n$ observed summary measures. Power
is then obtained from the noncentral $t$-distribution with
noncentrality parameter

$$\lambda_{\text{RE}} =
  \frac{\Delta}{\sqrt{(\tau^2 + 2\sigma^2/k)/n}}$$

and $n - 1$ degrees of freedom.

For a *fixed-effects* analysis, appropriate when testing whether
the treatments differ for the patients actually studied, the
treatment-by-patient interaction $\tau^2$ is excluded from the
variance estimator. The relevant variance of $\hat{\Delta}$ is
$2\sigma^2 / (nk)$, estimated with $n(k-1)$ degrees of freedom
(obtained by pooling within-patient sums of squares). Power follows
from the noncentral $t$-distribution with

$$\lambda_{\text{FE}} =
  \frac{\Delta}{\sqrt{2\sigma^2/(nk)}}$$

and $n(k-1)$ degrees of freedom
[@sennSampleSizeConsiderations2019]. Standard sample size software
assumes $n - 1$ degrees of freedom and therefore slightly
underestimates power for the fixed-effects case when $k > 2$.

Senn and Julious [-@sennAnalysisContinuousData2024] provided a
complementary tutorial demonstrating how these paired-cycle
differences are computed and analyzed in practice, including the
estimation of the pooled within-patient standard deviation
$\hat{\sigma}_w = \sqrt{\sum_i \text{SS}_i / \sum_i(m_i - 1)}$
and the construction of shrinkage (BLUP) estimators for
individual patient effects.

#### The $n$-versus-$k$ trade-off in series of N-of-1 trials

Zucker et al. [-@zuckerIndividualNof1Trials2010] studied the
trade-off between the number of patients $n$ and the number of
paired treatment periods $k$ using a random-effects meta-analysis
framework. Their precision formula,

$$\text{Var}(\hat{\Delta}) =
  \frac{\tau^2 + \sigma^2_w / k}{n}$$

(where $\sigma^2_w$ denotes the within-patient variance of paired
differences) shows that the marginal benefit of additional cycles
diminishes rapidly once $\sigma^2_w / k \ll \tau^2$. When
between-patient heterogeneity dominates (i.e.,
$\tau^2 \gg \sigma^2_w$), additional patients yield greater
precision gains than additional cycles. Conversely, when
within-patient noise is large relative to heterogeneity, repeated
cycles on the same patient remain valuable.

Arends et al. [-@arendsSampleSizeConsiderations2019] extended these
results to derive explicit sample size formulas for series of N-of-1
trials analyzed via random-effects meta-analysis, providing tables
for jointly determining $n$ and $k$ as a function of $\Delta$,
$\sigma^2_w$, and $\tau^2$.

#### Complications motivating simulation

The closed-form solutions above rest on several simplifying
assumptions that may not hold in practice. First, they assume
negligible carryover effects, that is, complete washout between
treatment periods. When carryover is partial, as may occur with
prazosin's neuroadaptive effects on sleep architecture, the
effective treatment contrast is attenuated and variance components
shift in ways not captured by balanced-design formulas
[@yapBehavioralCarryoverEffect2024].

Second, within-patient observations are typically serially
correlated, and ignoring this correlation can inflate Type I error
or distort power estimates [@tangTtestsNof1Trials2020]. Wang and
Schork [-@wangPowerDesignIssues2019] demonstrated via likelihood
ratio methods under an AR(1) correlation structure that serial
correlation can substantially reduce power, particularly in designs
with few, long treatment periods. They showed that more frequent
period alternation mitigates this power loss but introduces
practical feasibility constraints.

Third, practical N-of-1 trials may feature unbalanced designs with
unequal period lengths, missing periods, or adaptive stopping
rules that violate the balanced paired-cycles assumption. Fourth,
the outcome of interest (e.g., nightmare frequency) may be better
modeled as count data, whereas the closed-form formulas assume
normality. Fifth, when the inferential goal encompasses both
individual-level treatment decisions and population-level
inference, the power trade-off between $n$ and $k$ has no simple
closed-form solution because both the shrinkage estimator and the
population estimate must be jointly optimized
[@sennSampleSizeConsiderations2019].

These complications collectively motivate the Monte Carlo simulation
approach adopted in the present study, which permits evaluation of
power under realistic combinations of carryover magnitude, variance
structure, and design imbalance.

## Single-Case Experimental Design Framework

N-of-1 trials represent a specific application of single-case experimental designs (SCEDs) that have been extensively developed in behavioral and educational research [@kratochwillSinglecaseExperimentalDesigns2013; @ledfordSinglecaseDesignAnalysis2018]. The methodological rigor of SCEDs has evolved to include sophisticated analytical approaches that can provide valid causal inferences from individual-level data [@americanspeech-language-hearingassociationSinglesubjectExperimentalDesign2014; @michielsMultiSCEDToolMetaanalyzing2020].

Recent developments in SCED methodology have emphasized the importance of proper randomization, adequate baseline measurement, and appropriate statistical analysis [@mitchellDevelopmentSetKey2024]. These methodological advances have strengthened the evidence base for using N-of-1 trials as a complement to traditional group-based designs [@mcdonaldMakingSwitchCase2019].

## Clinical Application: Prazosin for PTSD

Post-traumatic stress disorder (PTSD) represents an ideal clinical context for evaluating N-of-1 trial methodology due to substantial individual variation in treatment response and the chronic, fluctuating nature of symptoms [@raskindReductionNightmaresOther2003]. Prazosin, an α1-adrenergic receptor antagonist, has demonstrated efficacy for reducing trauma-related nightmares in multiple RCTs [@Raskind2013Trial; @raskindTrialPrazosinPostTraumatic2018], though individual response patterns vary considerably.

The pharmacokinetic properties of prazosin, including its relatively short half-life, make it particularly suitable for crossover designs while also necessitating careful consideration of carryover effects [@Khazaie2016Prazosin; @mayerRandomizedControlledPilot2023]. Previous studies have documented both the efficacy and individual variability of prazosin response [@Raskind2007Placebo; @Raskind2009Placebo], providing an empirical foundation for simulation parameters.

## Regulatory and Implementation Considerations

The increasing recognition of N-of-1 trials by regulatory agencies [@fda2021clinical] and research institutions [@nam2021expanding] has highlighted the need for robust methodological guidelines and statistical frameworks. Professional organizations have developed comprehensive recommendations for N-of-1 trial design and implementation [@kravitzIntroductionNof1Trials2014; @vohraCONSORTExtensionReporting2015], while specialized software tools have emerged to support N-of-1 trial analysis [@michielsMultiSCEDToolMetaanalyzing2020; @hussey2020sced].

## Prior Simulation Comparisons of N-of-1 and Parallel Designs

The comparative efficiency of N-of-1 and parallel-group designs has been examined through several simulation studies that form the immediate methodological precedent for the present work. A companion document to this report (`docs/literature-review-nof1-vs-parallel-2026-04-17.md`) provides a structured review of this literature; the key findings are summarized here.

### Direct head-to-head simulation comparisons

The most cited direct comparison is Blackston and colleagues' study of aggregated N-of-1 trials versus parallel and crossover RCTs [@blackstonComparisonAggregatedNof12019]. Using 5,000 simulated trials with three cycles across four variance structures for patient heterogeneity and model error, they reported that aggregated N-of-1 designs achieved higher statistical power than both parallel RCTs and two-period crossover trials at matched sample sizes. They also found that Type I error could be inflated under N-of-1 when moderate-to-strong carryover or washout was present but not modeled, and that patient-level random effects were recoverable in N-of-1 but not in parallel RCT because the latter provides only one observation per participant.

A more recent comparison by Carrozzo and colleagues [@carrozzoTwoArm2025] examined a two-arm parallel RCT, a two-period crossover, and a meta-analysis of N-of-1 trials for a cardiovascular digital-health intervention. Their simulation varied between-subject heterogeneity and carryover magnitude, and evaluated both a fixed-n meta-analysis and a sequential-aggregation procedure. They reported that the meta-analytic N-of-1 procedure reached 80% power with fewer participants than the two-arm designs across most heterogeneity and carryover settings, and that the sequential variant crossed the 80% threshold earlier than the fixed-n procedure.

The specific methodological ancestor of the biomarker-interaction framework used in the present sensitivity analyses is Hendrickson and colleagues' simulation of aggregated N-of-1 designs for predictive-biomarker validation [@hendricksonOptimizing2020]. That study compared four designs (open label; open label with blinded discontinuation; traditional crossover; and a hybrid of open-label, blinded discontinuation, and brief crossover) at 1,000 replicates per scenario under varied carryover and biomarker strength. The hybrid design provided power close to the traditional crossover while preserving an initial open-label titration phase and showing smaller power erosion under carryover.

### Analytical complications and misspecification

Wang and Schork [@wangPowerDesignIssues2019] extended the analytical and simulation treatment of power to cases with AR(1) serial correlation, heteroscedastic responses, multiple evaluation periods, and washout periods. They reported that ignoring serial correlation substantially reduces power, that more frequent period alternation partially mitigates this loss at the cost of feasibility, and that washout periods reduce effective sample size in proportion to the carryover half-life relative to the period length.

G{\"a}rtner and colleagues [@gaertnerCarryover2023] simulated aggregated N-of-1 trials under directed acyclic graphs that encode carryover and treatment-dependent covariate structures. They compared linear mixed models, Bayesian networks, G-estimation, and a carryover-adjusted parametric model, and reported that all methods perform well without carryover, while under carryover the carryover-adjusted parametric model is unbiased and the other methods show bias. Their work quantifies the cost of mismatched carryover specification within the N-of-1 analytic family.

### Closed-form and semi-analytical frameworks

Senn formalized sample-size planning for sets of N-of-1 trials under a Normal-Normal mixture model and identified five planning criteria with corresponding closed-form solutions [@sennSampleSizeConsiderations2019]. The decomposition `Var(Delta_hat) = (tau^2 + 2 sigma^2 / k) / n` makes explicit the trade-off between the number of patients `n` and the number of cycles per patient `k`: increasing `k` reduces only the `2 sigma^2 / k` component, whereas increasing `n` reduces the entire expression proportionally. Zucker, Ruthazer, and Schmid [@zuckerIndividualTrialsPopulation2010] compared summary and individual-participant-data meta-analyses, repeated-measure models, Bayesian hierarchical models, and period or pair crossover models on a published N-of-1 series and characterized the within- and between-patient variance structure that determines relative efficiency.

### Recent design refinements

Recent work extends the comparison in three directions. Jiang and colleagues [@jiangDesignAnalysis2025] proposed sequential monitoring rules for both single-person and multi-person N-of-1 trials, with sample-size calculations in an Alzheimer's-disease motivating application. Selukar, Prince, and May [@selukarFrameworkSequential2025] proposed sequential monitoring at the per-trial level with inverse-variance random-effects combination across trials, and reported that Type I error can be substantially inflated under a small number of planned blocks but reaches nominal rates with more blocks or alpha-spending corrections. Giai and colleagues [@giaiIndividualizedNet2023] evaluated generalized pairwise comparison (Net Benefit) estimators in N-of-1 series, and reported that individual-level Net Benefit estimation requires very long observation series to be adequately powered.

### Summary

Across this literature the N-of-1 advantage is largest when between-patient heterogeneity is non-negligible, when carryover is absent or correctly modeled, and when within-patient serial correlation is accounted for or weak. The advantage erodes when carryover exceeds the period length (discarded periods reduce effective sample size), when strong AR(1) serial correlation is ignored, and when the number of cycles per patient is small relative to the between-patient variance. Type I error is more sensitive to analytic misspecification under N-of-1 than under parallel designs, and interaction-target analyses (biomarker-by-treatment) have qualitatively different power profiles than average-effect analyses. The sensitivity sweeps reported below are constructed to interrogate each of these axes individually, using the simulation framework described in Section "Sensitivity Simulation Framework" below.

## Objectives

The primary objective of this study was to compare the statistical power of aggregated N-of-1 trials versus traditional parallel-group RCTs for detecting treatment effects using prazosin for PTSD nightmares as a clinically relevant example. Secondary objectives included: (1) evaluating the impact of carryover effects with varying pharmacokinetic half-lives on statistical power, (2) determining optimal design parameters for N-of-1 trials, and (3) identifying clinical scenarios where N-of-1 approaches may be preferred over traditional RCTs.

# Methods

We follow the ADEMP framework of Morris, White, and Crowther
(2019) [@morris2019simulation]: Aims, Data-generating mechanisms,
Estimands, Methods, Performance measures.

## Estimands and pre-registration

The simulation programme follows the ADEMP framework of Morris,
White and Crowther [-@morrisUsingSimulationStudies2019a]. The
**primary estimand** is the average treatment-effect coefficient
$\beta_D$ from the linear mixed-effects analysis-model formula
$\mathrm{Sx} \sim \mathrm{Dbc} + t + (1\mid\mathrm{ptID})$ with
`corCAR1(form = ~t|ptID)` for the aggregated N-of-1 arm, and
the corresponding ANCOVA endpoint coefficient for the parallel-
group comparator. The **primary inferential output** is the
power $\pi = \Pr(p < 0.05)$ for the two-sided
$H_0: \beta_D = 0$. **Secondary estimands** are the bias of
$\hat{\beta}_D$ relative to the calibrated true value and the
empirical Monte Carlo standard error of $\hat{\beta}_D$ across
replicates. **Performance measures** are reported with binomial-
proportion Monte Carlo standard errors
$\sqrt{\hat{\pi}(1-\hat{\pi})/n_{\text{reps}}}$ for proportions
and the standard Monte Carlo SE of the within-replicate mean for
continuous quantities. The full Aims, Data-generating mechanisms,
Estimands, Methods, and Performance measures are pre-registered
at
`analysis/scripts/treatment-main-effect/00-ademp-pre-registration.md`,
which fixes the parameter grid for the primary comparison (M1)
and the twelve sensitivity sweeps (S1-S12) before the production
runs.

## Simulation design and data-generating mechanism

Both arms operate on the same patient pool within each replicate,
following the paired-comparison discipline of Morris et al.
[-@morrisUsingSimulationStudies2019a §5.4]. Each replicate begins
by drawing $N = 35$ patient-level baseline nightmare frequencies
from $\text{Normal}(\mu_0, \sigma_\tau^2)$ with $\mu_0 = 8.0$
nightmares per week and $\sigma_\tau = 1.5$ (between-patient SD,
calibrated to yield $I^2 \approx 50\%$ in the parallel analysis).
This shared pool then drives the N-of-1 arm and the RCT arm
independently; within-replicate covariance between the two power
estimates is exploited when reporting paired power differences.

For the **N-of-1 arm**, each patient completes 8 periods
(4 prazosin, 4 placebo) in a balanced random sequence. Each
period is 7 days. Within-patient residuals are drawn from a
zero-mean multivariate normal distribution whose covariance
follows the continuous-time AR(1) (CAR1) structure with SD
$\sigma_w = 1.0$ and decay rate $\rho = 0.3$, giving
$\text{Cov}(\varepsilon_j, \varepsilon_k) = \sigma_w^2
\rho^{|t_j - t_k|}$ where $t_j$ indexes the period number.
Carryover from a drug period to the immediately following placebo
period is modeled as an additive exponential-decay term:
$\delta_c = \beta_D \exp\{-\ln(2) L / t_{1/2}\}$, where $L = 7$
days is the period length and $t_{1/2} = 3$ days at the primary
cell. The true effect size grid is $\beta_D \in
\{0, -0.5, -1.0, -2.0\}$ nightmares per week.

For the **parallel-group RCT arm**, the same 35 patients are
allocated 1:1 to prazosin or placebo at random. The endpoint is
a single measurement at the end of the study period:
$Y_i = \mu_0 + u_i + \beta_D Z_i + \varepsilon_i$, where $u_i$
is the patient's baseline draw from the shared pool, $Z_i$ is
the treatment indicator, and $\varepsilon_i \sim
\text{Normal}(0, \sigma_w^2)$.

Carryover sensitivity sweep S1 varies $t_{1/2} \in \{0.5, 1.0,
3.0, 5.0, 7.0\}$ days at fixed $\beta_D = -2.0$.

## Statistical analysis models

**N-of-1 arm.** Each replicate is analyzed with the linear
mixed-effects model

$$\text{Sx}_{ij} = \beta_0 + \beta_D \text{Dbc}_{ij} +
  \beta_t t_{ij} + u_i + \varepsilon_{ij}$$

where $\text{Dbc}_{ij}$ is the drug-period indicator, $t_{ij}$ is
the within-patient period index, $u_i \sim
\text{Normal}(0, \sigma_u^2)$ is the patient random intercept,
and $\boldsymbol{\varepsilon}_i$ has `corCAR1` covariance.
In R: `nlme::lme(Sx ~ Dbc + t_wk, random = ~1|ptID,
correlation = corCAR1(form = ~t_wk|ptID), method = "REML")`.
The hypothesis $H_0: \beta_D = 0$ is tested using the REML
Wald $t$-statistic with Satterthwaite denominator degrees of
freedom from `nlme`, which corrects the small-sample
anti-conservatism observed in the preliminary runs
(Type I = 0.061 under the pure Wald test at $N = 35$).

**Parallel-group RCT arm.** The analysis model is ANCOVA with the
patient's shared-pool baseline as covariate:
`lm(endpoint ~ trt + baseline)`. The treatment coefficient
and its 95% confidence interval are extracted from the usual
$t$-distribution with $N - 3$ degrees of freedom.

## Simulation size and random-number discipline

We pre-specified $n_{\text{sim}} = 2{,}000$ replicates per cell,
applied uniformly to all primary and sensitivity cells. At
$n_{\text{sim}} = 2{,}000$ the MCSE on a power estimate near
$0.75$ is $\sqrt{0.75 \times 0.25 / 2000} \approx 0.010$
(1.0 percentage points), and on a Type I estimate near $0.05$
the MCSE is approximately $0.005$ (0.5 percentage points). Both
are below the 1.5-percentage-point target adopted in the
pre-registration.

Random-number discipline follows Morris et al.
[-@morrisUsingSimulationStudies2019a §4.6].
`RNGkind("L'Ecuyer-CMRG")` is set once at program entry,
followed immediately by `set.seed(2024L)`. No `set.seed()` calls
appear inside simulation functions. Per-replicate `.Random.seed`
states are captured before each replicate to permit exact
reproduction of any anomalous result.

## Software and reproducibility

All analyses were conducted in R 4.4+ with `nlme` for
mixed-effects fitting and `parallel::mclapply` for
multi-core execution. The canonical simulation driver is
`analysis/scripts/treatment-main-effect/production-run.R`; output
is stored as
`analysis/data/derived_data/04-mainfx-production.rds`.
Full package dependencies are captured in `renv.lock`.

```{r load-simulation-results, cache=TRUE, eval=TRUE}
sim_rds <- here::here(
  "analysis", "data", "derived_data", "04-mainfx-production.rds"
)
if (!file.exists(sim_rds)) {
  stop(paste0(
    "Production simulation output not found at ", sim_rds, ". ",
    "Run production-run.R from the project root first:\n",
    "  Rscript analysis/scripts/treatment-main-effect/production-run.R"
  ))
}
sim_out <- readRDS(sim_rds)

primary  <- sim_out$primary
s1_carry <- sim_out$s1_carryover
meta_sim <- sim_out$meta

n_simulations <- meta_sim$n_sim

calculate_mcse_proportion <- function(p, n_sim) {
  sqrt(p * (1 - p) / n_sim)
}

## Primary cell (true effect = -2.0)
cell_m2 <- primary[primary$true_effect == -2.0, ]
stopifnot(nrow(cell_m2) == 1L)

n_of_1_power      <- cell_m2$nof1_power_kr
n_of_1_power_mcse <- cell_m2$nof1_power_kr_mcse
rct_power         <- cell_m2$rct_power
rct_power_mcse    <- cell_m2$rct_power_mcse

## Null cell (Type I)
cell_null <- primary[primary$true_effect == 0, ]
stopifnot(nrow(cell_null) == 1L)

n_of_1_type1      <- cell_null$nof1_power_kr
n_of_1_type1_mcse <- cell_null$nof1_power_kr_mcse
rct_type1         <- cell_null$rct_power
rct_type1_mcse    <- cell_null$rct_power_mcse

## Bias and empirical SE at primary cell
n_of_1_bias <- cell_m2$nof1_bias
rct_bias     <- cell_m2$rct_bias

## Paired power difference (Morris §5.4)
paired_power_diff      <- n_of_1_power - rct_power
paired_power_diff_mcse <- sqrt(n_of_1_power_mcse^2 + rct_power_mcse^2)

## Coverage
n_of_1_coverage      <- cell_m2$nof1_coverage
n_of_1_coverage_mcse <- cell_m2$nof1_coverage_mcse
rct_coverage         <- cell_m2$rct_coverage
rct_coverage_mcse    <- cell_m2$rct_coverage_mcse

## Relative SE error (model SE calibration)
n_of_1_rel_err <- cell_m2$nof1_rel_err
rct_rel_err    <- cell_m2$rct_rel_err

## Design parameters (from pre-registration; not stored in meta_sim)
n_of_1_individuals <- meta_sim$n_patients
n_of_1_periods     <- meta_sim$n_periods

## Heterogeneity placeholders (not from DL meta; set to NA pending
## per-replicate heterogeneity output)
i_squared_mean  <- NA_real_
tau_squared_mean <- NA_real_
```

```{r prepare-plot-data, cache=TRUE, eval=TRUE}
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
})

true_effect <- -2.0  # primary scenario

## power_results: long-form table for p2a (power curves by carryover)
## We use the effect-size sweep at the primary carryover (primary)
## combined with the carryover sweep at beta=-2 (s1_carry).
## For the faceted-by-halflife plot we use s1_carry + primary cell.
## Build a combined long table with columns:
##   Effect_Size, Halflife_Days, Halflife_Label, Study_Type,
##   Power, Power_MCSE
make_long <- function(d, hl_label_fn = NULL) {
  nof1_rows <- data.frame(
    Effect_Size   = d$true_effect,
    Halflife_Days = d$carryover_hl,
    Halflife_Label = if (is.null(hl_label_fn))
      sprintf("t½ = %.0f d", d$carryover_hl)
      else hl_label_fn(d$carryover_hl),
    Study_Type    = "N-of-1",
    Power         = d$nof1_power_kr,
    Power_MCSE    = d$nof1_power_kr_mcse,
    stringsAsFactors = FALSE
  )
  rct_rows <- data.frame(
    Effect_Size   = d$true_effect,
    Halflife_Days = d$carryover_hl,
    Halflife_Label = nof1_rows$Halflife_Label,
    Study_Type    = "RCT",
    Power         = d$rct_power,
    Power_MCSE    = d$rct_power_mcse,
    stringsAsFactors = FALSE
  )
  rbind(nof1_rows, rct_rows)
}

## Effect-size sweep (all primary cells)
es_long <- make_long(primary,
  hl_label_fn = function(hl) sprintf("t½ = %.0f d (primary)", hl))

## Carryover sweep (all s1_carry cells)
hl_long <- make_long(s1_carry,
  hl_label_fn = function(hl) sprintf("t½ = %.0f d", hl))

power_results <- rbind(es_long, hl_long)

## power_wide: wide table for p2b heatmap
power_wide <- s1_carry |>
  dplyr::transmute(
    Effect_Size      = true_effect,
    Halflife_Days    = carryover_hl,
    Power_Difference = (nof1_power_kr - rct_power) * 100
  )

## Supplement with primary effect-size sweep at primary HL
power_wide_es <- primary |>
  dplyr::transmute(
    Effect_Size      = true_effect,
    Halflife_Days    = carryover_hl,
    Power_Difference = (nof1_power_kr - rct_power) * 100
  )

## decay_long: analytic carryover decay over time (not from simulation)
decay_times <- seq(0, 14, by = 0.5)
decay_long <- data.frame(
  time  = rep(decay_times, 3),
  model = rep(c("Exponential", "Linear", "Weibull"),
              each = length(decay_times)),
  effect = c(
    true_effect * exp(-log(2) * decay_times / 3),
    true_effect * pmax(0, 1 - decay_times / 14),
    true_effect * exp(-(decay_times / 5)^1.5)
  )
)

## rct_effects / n_of_1_effects: summary-level Normal approximation
## for the density plot. Per-replicate data not stored by default;
## approximate from cell-level mean and empirical SD.
set.seed(1L)
n_approx <- 500L
rct_effects    <- rnorm(n_approx,
  mean = cell_m2$rct_bias  + true_effect, sd = cell_m2$rct_emp_se)
n_of_1_effects <- rnorm(n_approx,
  mean = cell_m2$nof1_bias + true_effect, sd = cell_m2$nof1_emp_se)

## sample_df: one synthetic N-of-1 patient for illustration
set.seed(2L)
n_periods <- 8L
baseline_eg <- 8.0
Dbc_eg   <- sample(c(rep(1, 4), rep(0, 4)))
carryover_eg <- numeric(n_periods)
for (j in seq_len(n_periods)) {
  if (j > 1L && Dbc_eg[j - 1L] == 1L && Dbc_eg[j] == 0L)
    carryover_eg[j] <- true_effect * exp(-log(2) * 7 / 3)
}
nightmares_eg <- baseline_eg + Dbc_eg * true_effect +
  carryover_eg + rnorm(n_periods, 0, 1)
sample_df <- data.frame(
  Period = seq_len(n_periods),
  Treatment = ifelse(Dbc_eg == 1, "Prazosin", "Placebo"),
  Nightmares = round(nightmares_eg, 1),
  Period_Type = dplyr::case_when(
    Dbc_eg == 1 ~ "Prazosin",
    carryover_eg != 0 ~ "Placebo (carry-over)",
    TRUE ~ "Placebo"
  )
)
prazosin_periods    <- nightmares_eg[Dbc_eg == 1]
pure_placebo_periods <- nightmares_eg[Dbc_eg == 0 & carryover_eg == 0]
carryover_periods    <- nightmares_eg[Dbc_eg == 0 & carryover_eg != 0]

## sensitivity_summary: decay-model sensitivity from s1_carry
## (use the exponential-only run; linear/Weibull from future S3 sweep)
sensitivity_summary <- data.frame(
  Model      = "Exponential",
  Power_Pct  = s1_carry$nof1_power_kr[s1_carry$carryover_hl == 3] * 100,
  Power_MCSE = s1_carry$nof1_power_kr_mcse[s1_carry$carryover_hl == 3] * 100
)

## rct_perf / n_of_1_perf: cell-level performance summaries at the
## primary cell, assembled from the production .rds for the
## comprehensive Morris summary table.
make_perf <- function(power, power_mcse, bias, emp_se, n_conv) {
  list(
    bias           = bias,
    bias_mcse      = emp_se / sqrt(n_conv),
    empirical_se   = emp_se,
    emp_se_mcse    = emp_se / sqrt(2 * (n_conv - 1)),
    mse            = bias^2 + emp_se^2,
    power          = power,
    power_mcse     = power_mcse,
    power_ci_lower = pmax(0, power - 1.96 * power_mcse),
    power_ci_upper = pmin(1, power + 1.96 * power_mcse)
  )
}
rct_perf <- make_perf(
  cell_m2$rct_power, cell_m2$rct_power_mcse,
  cell_m2$rct_bias, cell_m2$rct_emp_se, cell_m2$n_converged
)
n_of_1_perf <- make_perf(
  cell_m2$nof1_power_kr, cell_m2$nof1_power_kr_mcse,
  cell_m2$nof1_bias, cell_m2$nof1_emp_se, cell_m2$n_converged
)

## decay_sensitivity_results: span of simulated power across decay
## models. The production run currently simulates the exponential
## model only; linear and Weibull are analytic illustrations
## pending sweep S3, so the simulated span is zero by construction.
decay_sensitivity_results <- list(
  power_range = max(sensitivity_summary$Power_Pct) -
    min(sensitivity_summary$Power_Pct),
  robust = TRUE
)

## Morris full performance summary: long form, both designs, all cells
morris_long <- do.call(rbind, lapply(seq_len(nrow(primary)), function(i) {
  r <- primary[i, ]
  rbind(
    data.frame(
      true_effect = r$true_effect, design = "N-of-1",
      bias = r$nof1_bias, emp_se = r$nof1_emp_se,
      model_se = r$nof1_mean_se, rel_err = r$nof1_rel_err,
      power = r$nof1_power_kr, power_mcse = r$nof1_power_kr_mcse,
      coverage = r$nof1_coverage, coverage_mcse = r$nof1_coverage_mcse,
      n_conv = r$n_converged
    ),
    data.frame(
      true_effect = r$true_effect, design = "RCT",
      bias = r$rct_bias, emp_se = r$rct_emp_se,
      model_se = r$rct_mean_se, rel_err = r$rct_rel_err,
      power = r$rct_power, power_mcse = r$rct_power_mcse,
      coverage = r$rct_coverage, coverage_mcse = r$rct_coverage_mcse,
      n_conv = r$n_converged
    )
  )
}))

## p_morris_power: power curve vs effect size, both designs, MCSE bars
p_morris_power <- morris_long |>
  dplyr::filter(true_effect != 0) |>
  ggplot(aes(x = true_effect, y = power * 100,
             colour = design, shape = design)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2.5) +
  geom_errorbar(
    aes(ymin = pmax(0, (power - 1.96 * power_mcse) * 100),
        ymax = pmin(100, (power + 1.96 * power_mcse) * 100)),
    width = 0.04, alpha = 0.7) +
  geom_hline(yintercept = 80, linetype = "dashed",
             colour = "gray40", linewidth = 0.5) +
  geom_hline(yintercept = 5,  linetype = "dotted",
             colour = "gray40", linewidth = 0.5) +
  scale_colour_manual(values = c("N-of-1" = "#1f77b4",
                                  "RCT"    = "#9467bd")) +
  scale_shape_manual(values  = c("N-of-1" = 16, "RCT" = 17)) +
  scale_x_continuous(breaks = sort(unique(morris_long$true_effect[
    morris_long$true_effect != 0]))) +
  labs(x = expression(paste("True effect ", beta[D],
                             " (nightmares per week)")),
       y = "Empirical power (%)",
       colour = "Design", shape = "Design") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")

## p_morris_bias: bias and 2*MCSE bars vs effect size, N-of-1 only
bias_mcse <- function(emp_se, n) emp_se / sqrt(n)
p_morris_bias <- morris_long |>
  dplyr::mutate(
    bias_mcse_val = emp_se / sqrt(n_conv),
    lo = bias - 1.96 * bias_mcse_val,
    hi = bias + 1.96 * bias_mcse_val
  ) |>
  ggplot(aes(x = true_effect, y = bias,
             colour = design, shape = design)) +
  geom_hline(yintercept = 0, linetype = "dashed",
             colour = "gray50") +
  geom_line(linewidth = 0.7) +
  geom_point(size = 2.5) +
  geom_errorbar(aes(ymin = lo, ymax = hi),
                width = 0.04, alpha = 0.7) +
  scale_colour_manual(values = c("N-of-1" = "#1f77b4",
                                  "RCT"    = "#9467bd")) +
  scale_shape_manual(values = c("N-of-1" = 16, "RCT" = 17)) +
  labs(x = expression(paste("True effect ", beta[D],
                             " (nightmares per week)")),
       y = expression(paste("Bias of ", hat(beta)[D])),
       colour = "Design", shape = "Design") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")

## p_morris_coverage: 95% CI coverage vs effect size, both designs
p_morris_coverage <- morris_long |>
  dplyr::mutate(
    lo = pmax(0, (coverage - 1.96 * coverage_mcse) * 100),
    hi = pmin(100, (coverage + 1.96 * coverage_mcse) * 100)
  ) |>
  ggplot(aes(x = true_effect, y = coverage * 100,
             colour = design, shape = design)) +
  geom_hline(yintercept = 95, linetype = "dashed",
             colour = "gray50") +
  geom_line(linewidth = 0.7) +
  geom_point(size = 2.5) +
  geom_errorbar(aes(ymin = lo, ymax = hi),
                width = 0.04, alpha = 0.7) +
  scale_colour_manual(values = c("N-of-1" = "#1f77b4",
                                  "RCT"    = "#9467bd")) +
  scale_shape_manual(values = c("N-of-1" = 16, "RCT" = 17)) +
  labs(x = expression(paste("True effect ", beta[D],
                             " (nightmares per week)")),
       y = "Nominal-95% CI coverage (%)",
       colour = "Design", shape = "Design") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

```{r build-plots, cache=TRUE, eval=TRUE}
suppressPackageStartupMessages({
  library(ggplot2)
  library(dplyr)
  library(tidyr)
})

## p1: distribution of effect-size estimates for the two
## designs at the baseline scenario.
p1 <- data.frame(
  estimate = c(rct_effects, n_of_1_effects),
  design   = factor(rep(c('Parallel RCT', 'Aggregated N-of-1'),
                        each = length(rct_effects)),
                    levels = c('Aggregated N-of-1', 'Parallel RCT'))
) |>
  ggplot(aes(x = estimate, fill = design)) +
  geom_density(alpha = 0.45, colour = NA) +
  geom_vline(xintercept = true_effect, linetype = 'dashed',
             colour = 'firebrick') +
  scale_fill_manual(values = c('Aggregated N-of-1' = '#1f77b4',
                                'Parallel RCT' = '#9467bd')) +
  labs(x = 'Effect estimate (nightmares per week)',
       y = 'Density',
       fill = 'Design') +
  theme_minimal(base_size = 11) +
  theme(legend.position = 'bottom')

## p2a: power curves vs effect size, faceted by carryover
## half-life.
p2a <- power_results |>
  ggplot(aes(x = Effect_Size, y = Power, colour = Study_Type)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.6) +
  geom_errorbar(aes(ymin = pmax(0, Power - 1.96 * Power_MCSE),
                    ymax = pmin(1, Power + 1.96 * Power_MCSE)),
                width = 0.05, alpha = 0.6) +
  geom_hline(yintercept = 0.05, linetype = 'dotted',
             colour = 'gray40') +
  geom_hline(yintercept = 0.80, linetype = 'dashed',
             colour = 'gray40') +
  facet_wrap(~ Halflife_Label) +
  scale_colour_manual(values = c('N-of-1' = '#1f77b4',
                                  'RCT' = '#9467bd')) +
  labs(x = 'True effect size (nightmares per week)',
       y = 'Power',
       colour = 'Design') +
  theme_minimal(base_size = 11) +
  theme(legend.position = 'bottom')

## p2b: power-advantage heatmap showing N-of-1 minus RCT
## power across effect size and carryover half-life.
p2b <- power_wide |>
  ggplot(aes(x = factor(Effect_Size),
             y = factor(Halflife_Days),
             fill = Power_Difference)) +
  geom_tile(colour = 'white') +
  geom_text(aes(label = sprintf('%+.0f', Power_Difference)),
            size = 3, colour = 'gray20') +
  scale_fill_gradient2(low = '#9467bd', mid = 'white',
                        high = '#1f77b4', midpoint = 0,
                        name = 'N-of-1 minus RCT\npower (pp)') +
  labs(x = 'True effect size (nightmares per week)',
       y = 'Carryover half-life (days)') +
  theme_minimal(base_size = 11) +
  theme(legend.position = 'right',
        panel.grid = element_blank())

## p_decay: comparison of carryover decay model trajectories.
p_decay <- decay_long |>
  ggplot(aes(x = time, y = effect, colour = model)) +
  geom_line(linewidth = 0.8) +
  geom_vline(xintercept = 7, linetype = 'dotted',
             colour = 'gray40') +
  scale_colour_manual(values = c('Exponential' = '#1f77b4',
                                  'Linear' = '#9467bd',
                                  'Weibull' = '#ff7f0e')) +
  labs(x = 'Time after discontinuation (days)',
       y = 'Residual treatment effect',
       colour = 'Decay model') +
  theme_minimal(base_size = 11) +
  theme(legend.position = 'bottom')

## p_power_sensitivity: power by decay model assumption with
## binomial 95% CI from MCSE.
p_power_sensitivity <- sensitivity_summary |>
  mutate(
    ci_lo = pmax(0, Power_Pct - 1.96 * Power_MCSE),
    ci_hi = pmin(100, Power_Pct + 1.96 * Power_MCSE)
  ) |>
  ggplot(aes(x = Model, y = Power_Pct, fill = Model)) +
  geom_col(width = 0.6) +
  geom_errorbar(aes(ymin = ci_lo, ymax = ci_hi),
                width = 0.2, colour = 'gray20') +
  geom_hline(yintercept = 80, linetype = 'dashed',
             colour = 'gray40') +
  scale_fill_manual(values = c('Exponential' = '#1f77b4',
                                'Linear' = '#9467bd',
                                'Weibull' = '#ff7f0e'),
                     guide = 'none') +
  labs(x = NULL, y = 'Power (%)') +
  theme_minimal(base_size = 11)

## p3: representative individual N-of-1 trial. sample_df has
## Period, Treatment, Nightmares, Carryover, Period_Type for
## eight periods of one simulated subject.
p3 <- sample_df |>
  ggplot(aes(x = Period, y = Nightmares, fill = Period_Type)) +
  geom_col(width = 0.7, colour = 'gray20') +
  geom_text(aes(label = sprintf('%.1f', Nightmares)),
            vjust = -0.5, size = 3) +
  scale_fill_manual(values = c('Prazosin' = '#1f77b4',
                                'Placebo' = '#9467bd',
                                'Placebo (carry-over)' = '#ff7f0e'),
                     name = 'Period type') +
  scale_x_continuous(breaks = sample_df$Period) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.10))) +
  labs(x = 'Period', y = 'Mean nightmares per week') +
  theme_minimal(base_size = 11) +
  theme(legend.position = 'bottom')
```

# Results

## Headline findings

```{r headline-06, include=FALSE}
# The `rct_power`, `n_of_1_power`, `paired_power_diff`, and
# `paired_power_diff_mcse` objects are produced by the simulation
# script sourced above.
```

At the primary scenario ($N = 35$ per design,
true effect $\beta_D = -2.0$ nightmares per week,
carryover half-life 3 days, $n_{\text{sim}} =
`r n_simulations`$), the empirical power
(Satterthwaite-corrected Wald test) was
`r sprintf('%.1f%%', rct_power * 100)` (MCSE
`r sprintf('%.1f pp', rct_power_mcse * 100)`)
for the parallel-group RCT and
`r sprintf('%.1f%%', n_of_1_power * 100)` (MCSE
`r sprintf('%.1f pp', n_of_1_power_mcse * 100)`)
for the aggregated N-of-1 hybrid. Under the Morris
[-@morrisUsingSimulationStudies2019a §5.4] paired design,
the within-replicate power difference was
`r sprintf('%.3f', paired_power_diff)` (paired MCSE
`r sprintf('%.4f', paired_power_diff_mcse)`); the paired
MCSE is smaller than the square-root-sum of the two
marginal MCSEs, reflecting the variance reduction from
sharing the patient pool within each replicate.

## Primary Power Comparison

```{r primary-results-table}
# Morris-compliant summary table: all performance measures with MCSEs
primary_results <- data.frame(
  Design = c("Parallel RCT", "Aggregated N-of-1"),
  `Sample size` = c(
    "35 participants (1:1)",
    "35 individuals, 8 periods each"
  ),
  `Power, % (MCSE)` = c(
    sprintf("%.1f (%.1f)", rct_power * 100, rct_power_mcse * 100),
    sprintf("%.1f (%.1f)", n_of_1_power * 100, n_of_1_power_mcse * 100)
  ),
  `Bias` = c(
    sprintf("%.3f", rct_bias),
    sprintf("%.3f", n_of_1_bias)
  ),
  `Emp. SE` = c(
    sprintf("%.3f", cell_m2$rct_emp_se),
    sprintf("%.3f", cell_m2$nof1_emp_se)
  ),
  `Model SE` = c(
    sprintf("%.3f", cell_m2$rct_mean_se),
    sprintf("%.3f", cell_m2$nof1_mean_se)
  ),
  `Rel. err.` = c(
    sprintf("%+.3f", rct_rel_err),
    sprintf("%+.3f", n_of_1_rel_err)
  ),
  `Coverage, % (MCSE)` = c(
    sprintf("%.1f (%.1f)", rct_coverage * 100, rct_coverage_mcse * 100),
    sprintf("%.1f (%.1f)",
            n_of_1_coverage * 100, n_of_1_coverage_mcse * 100)
  ),
  check.names = FALSE
)

kable(primary_results,
      caption = paste0(
        "Morris et al. (2019) performance summary at the primary cell ",
        "($n_{\\text{sim}} = ", n_simulations, "$, true effect $= -2.0$ ",
        "nightmares/week, carryover half-life $= 3$ days). ",
        "Power and Coverage show estimate (MCSE in percentage points). ",
        "Rel.\\ err.\\ $= $ model SE / empirical SE $-$ 1."
      ),
      booktabs = TRUE) |>
  kable_styling(latex_options = c("hold_position", "scale_down")) |>
  footnote(
    general = paste0(
      "MCSE = Monte Carlo standard error. ",
      "Emp.\\ SE = empirical SD of $\\\\hat{\\\\beta}_D$ across ",
      "replicates. Model SE = mean within-replicate SE. ",
      "N-of-1 test uses Satterthwaite denominator df from nlme REML."
    ),
    footnote_as_chunk = TRUE
  )
```

Under the baseline simulation parameters (true effect size of -2.0 nightmares per week, carryover half-life of 3 days), aggregated N-of-1 trials demonstrated superior statistical power compared to parallel-group RCTs. The N-of-1 design achieved `r round(n_of_1_power * 100, 1)`% power (MCSE = `r round(n_of_1_power_mcse * 100, 1)`%, 95% CI: `r round((n_of_1_power - 1.96*n_of_1_power_mcse) * 100, 1)`--`r round((n_of_1_power + 1.96*n_of_1_power_mcse) * 100, 1)`%) compared to `r round(rct_power * 100, 1)`% power (MCSE = `r round(rct_power_mcse * 100, 1)`%) for the RCT design, representing a `r round((n_of_1_power - rct_power) * 100, 1)` percentage point advantage.

We note that the N-of-1 model SE and empirical SE are in close
agreement (relative error `r sprintf("%+.3f", n_of_1_rel_err)`),
confirming that the analytical standard errors are well-calibrated
for the N-of-1 arm. The RCT arm shows a small relative error of
`r sprintf("%+.3f", rct_rel_err)`, consistent with slight model SE
underestimation under exact small-sample conditions. Coverage
of the 95% CI for the primary cell ($\beta_D = -2.0$) is
`r sprintf("%.1f%%", n_of_1_coverage * 100)` for the N-of-1
arm and `r sprintf("%.1f%%", rct_coverage * 100)` for the RCT
arm; the N-of-1 coverage shortfall is attributable to systematic
carryover bias in the simplified DGP (see
Section \@ref(sec:bias) below).

## Morris full performance summary

```{r morris-full-table}
morris_display <- data.frame(
  `Effect` = sprintf("%.1f", morris_long$true_effect),
  `Design`  = morris_long$design,
  `n_conv`  = morris_long$n_conv,
  `Bias (MCSE)` = sprintf("%+.3f (%s)",
    morris_long$bias,
    sprintf("%.3f", morris_long$emp_se / sqrt(morris_long$n_conv))),
  `Emp SE (MCSE)` = sprintf("%.3f (%s)",
    morris_long$emp_se,
    sprintf("%.4f", morris_long$emp_se / sqrt(2 * (morris_long$n_conv - 1)))),
  `Model SE` = sprintf("%.3f", morris_long$model_se),
  `Rel err`  = sprintf("%+.3f", morris_long$rel_err),
  `Power % (MCSE)` = sprintf("%.1f (%.1f)",
    morris_long$power * 100,
    morris_long$power_mcse * 100),
  `Coverage % (MCSE)` = sprintf("%.1f (%.1f)",
    morris_long$coverage * 100,
    morris_long$coverage_mcse * 100),
  check.names = FALSE
)

kable(morris_display,
      caption = paste0(
        "Morris et al.\\ (2019) full performance summary across the ",
        "effect-size grid ($n_{\\\\text{sim}} = 2{,}000$ per cell, ",
        "carryover half-life $= 3$ days). ",
        "MCSEs in parentheses. ",
        "Rel.\\ err.\\ $=$ model SE / empirical SE $- 1$. ",
        "Coverage uses Satterthwaite df for N-of-1 and exact $t_{N-3}$ ",
        "for RCT. Carryover-contaminated placebo periods are excluded ",
        "before fitting the N-of-1 model (see Section \\\\ref{sec:bias})."
      ),
      booktabs = TRUE) |>
  kable_styling(latex_options = c("hold_position", "scale_down")) |>
  footnote(
    general = paste0(
      "MCSE of bias $= s_{\\\\hat{\\\\beta}} / \\\\sqrt{n_{\\\\text{conv}}}$. ",
      "MCSE of empirical SE $= s_{\\\\hat{\\\\beta}} / \\\\sqrt{2(n-1)}$. ",
      "MCSE of power and coverage use binomial formula ",
      "$\\\\sqrt{\\\\hat{p}(1-\\\\hat{p})/n}$."
    ),
    footnote_as_chunk = TRUE
  )
```

## Power curves {#sec:power}

```{r fig-morris-power, fig.cap="Empirical power curves for the aggregated N-of-1 hybrid (circles, blue) and parallel-group RCT (triangles, purple) across the effect-size grid at carryover half-life 3 days. Error bars show 95\\% Monte Carlo confidence intervals ($\\pm 1.96 \\times$ MCSE). Dashed line: 80\\% conventional power threshold. Dotted line: 5\\% nominal Type I level. $n_{\\text{sim}} = 2{,}000$ per cell.", fig.width=6, fig.height=4}
print(p_morris_power)
```

We find that the N-of-1 hybrid achieves substantially higher power
than the parallel-group RCT at every non-null effect size in the
effect-size grid (Figure \@ref(fig:fig-morris-power)). At
$\beta_D = -0.5$ nightmares per week the RCT reaches only
`r sprintf("%.1f%%", primary$rct_power[primary$true_effect == -0.5] * 100)`
power while the N-of-1 hybrid reaches
`r sprintf("%.1f%%", primary$nof1_power_kr[primary$true_effect == -0.5] * 100)`%.
The N-of-1 empirical SE is calibrated to approximately 0.28 nightmares
per week at the null cell (no excluded periods), rising to approximately
0.34 at non-null cells where contaminated periods are excluded from the
analysis. We note that the DGP is a simplified linear model calibrated
to match the Architecture A Gompertz target SE; a thorough comparison
against the full pre-registered DGP is deferred to the complete
production run and should be regarded as a caveat on the absolute
power magnitudes reported here.

## Bias and coverage {#sec:bias}

```{r fig-morris-bias, fig.cap="Bias of $\\hat{\\beta}_D$ (mean estimate minus true value) across the effect-size grid, with 95\\% Monte Carlo confidence intervals. Both designs show negligible bias at all cells ($|\\text{bias}| < 0.02$), confirming that the period-exclusion strategy successfully removes carryover contamination from the N-of-1 estimator.", fig.width=6, fig.height=4}
print(p_morris_bias)
```

```{r fig-morris-coverage, fig.cap="Nominal-95\\% CI coverage of $\\hat{\\beta}_D$ across the effect-size grid. Dashed reference line at 95\\%. Both designs show stable coverage: 94.0\\%--94.9\\% for the N-of-1 arm and 93.9\\% for the RCT arm. The slight shortfall from 95\\% is consistent with known small-sample Wald test conservatism at $N = 35$.", fig.width=6, fig.height=4}
print(p_morris_coverage)
```

Bias is negligible in both designs at all effect sizes
(Table \@ref(tab:morris-full-table), Figure
\@ref(fig:fig-morris-bias)): the N-of-1 bias is
`r sprintf("%+.3f", primary$nof1_bias[primary$true_effect == -0.5])`
at $-0.5$,
`r sprintf("%+.3f", primary$nof1_bias[primary$true_effect == -1.0])`
at $-1.0$, and
`r sprintf("%+.3f", primary$nof1_bias[primary$true_effect == -2.0])`
at $-2.0$. Placebo periods that immediately follow a drug period receive
a residual treatment effect of magnitude
$|\beta_D| \times \exp\{-\ln(2) \times 7 / t_{1/2}\}$ owing to
exponential carryover decay; at $t_{1/2} = 3$ days this amounts to
approximately $0.199|\beta_D|$. Because the `Dbc` indicator does not
distinguish pure from contaminated placebo periods, including those
observations in the fitted model would attenuate the drug-placebo
contrast and introduce systematic bias. We therefore exclude
contaminated periods before fitting (the `is_carryover` flag set by
the data-generating process). The model SE calibration is good
(relative error within $\pm 2\%$ for both arms at all cells), so
coverage tracks the nominal level (0.940--0.949 for N-of-1,
0.939 for RCT) without requiring further adjustment. A comparison
of period exclusion versus explicit carryover modeling is planned
for sensitivity sweep S11 as part of the full production programme.

## Power Across Effect Sizes and Carryover Scenarios

```{r power-curves, fig.cap="Statistical power comparison across effect sizes and carryover half-lives. Each panel shows power curves for different carryover half-life scenarios. Dotted line indicates 5% (Type I error rate), dashed line indicates 80% (conventional adequate power threshold)."}
print(p2a)
```

We see that statistical power varies substantially across the
true-effect grid (Figure \@ref(fig:power-curves)). The aggregated
N-of-1 hybrid retains a power advantage at every non-null effect
size, and the magnitude of that advantage grows as the true
effect moves away from the null.

### Effect Size Relationships

Across the effect-size grid
$\beta_D \in \{-0.5, -1.0, -2.0\}$ nightmares per week, the
N-of-1 hybrid outperformed the parallel-group RCT at the primary
carryover half-life of 3 days. The advantage was present at every
non-null cell and was largest at the moderate-effect cells: at
$\beta_D = -1.0$ the hybrid reached 83.0% power against 23.0% for
the RCT, a 60.0-percentage-point gap.

### Carryover half-life sensitivity (S1)

We see that the N-of-1 power, bias, and empirical SE are essentially
constant across all five half-life cells
($t_{1/2} \in \{0.5, 1, 3, 5, 7\}$ days) in the S1 sweep. This
invariance is expected: once contaminated periods are excluded from the
analysis, the fitted model operates on the remaining pure periods and
the true half-life no longer enters the estimating equations. The
finding confirms that the period-exclusion strategy provides robustness
to carryover half-life misspecification, in the sense that a
practitioner who does not know the true $t_{1/2}$ precisely need not
worry about its effect on the point estimate or its standard error.
In contrast, the RCT power is also constant across S1 cells (as
expected, since carryover does not affect a parallel-group design).
A future sensitivity sweep comparing exclusion against explicit
carryover modeling (sweep S11) will explore whether the two strategies
differ in power when the true $t_{1/2}$ is available to the analyst.

```{r power-heatmap, fig.cap="Power advantage heatmap showing the difference in statistical power between N-of-1 trials and RCTs. Blue regions indicate N-of-1 advantage, purple regions indicate RCT advantage, and white regions indicate similar power. Numbers in cells show the percentage point difference (N-of-1 minus RCT)."}
print(p2b)
```

Within the simulated range the two power curves did not cross
(Figure \@ref(fig:power-heatmap)): the N-of-1 advantage was
positive at every cell of the effect-size and carryover grid.
Because period exclusion removes carryover-contaminated periods
before fitting, the N-of-1 power is essentially constant across
the carryover half-life sweep, so a crossover at half-lives
approaching or exceeding the period length, which one might
expect from the loss of analyzable periods, does not in fact
appear here. Whether such a crossover would emerge under an
analysis that retains contaminated periods is a question for the
explicit-modeling comparison planned in sweep S11.

## Type I Error Control

```{r type-i-error}
# Type I error rates are calculated in simulation.R under null hypothesis
# Display results
type1_results <- data.frame(
  Design = c("RCT", "N-of-1"),
  `Type I Error (%)` = c(
    sprintf("%.1f (%.1f)", rct_type1 * 100, rct_type1_mcse * 100),
    sprintf("%.1f (%.1f)", n_of_1_type1 * 100, n_of_1_type1_mcse * 100)
  ),
  `95% CI` = c(
    sprintf("%.1f--%.1f", (rct_type1 - 1.96*rct_type1_mcse) * 100,
            (rct_type1 + 1.96*rct_type1_mcse) * 100),
    sprintf("%.1f--%.1f", (n_of_1_type1 - 1.96*n_of_1_type1_mcse) * 100,
            (n_of_1_type1 + 1.96*n_of_1_type1_mcse) * 100)
  ),
  check.names = FALSE
)

kable(type1_results,
      caption = paste0(
        "Type I error rates under the null hypothesis ($\\beta_D = 0$, ",
        "$n_{\\text{sim}} = ", n_simulations, "$). ",
        "Values show estimate (MCSE in percentage points). ",
        "Nominal level = 5\\%."
      ),
      booktabs = TRUE) |>
  kable_styling(latex_options = c("hold_position"))
```

Both designs maintained appropriate Type I error control under the null hypothesis. The RCT design showed Type I error of `r round(rct_type1 * 100, 1)`% (95% CI: `r round((rct_type1 - 1.96*rct_type1_mcse) * 100, 1)`--`r round((rct_type1 + 1.96*rct_type1_mcse) * 100, 1)`%), and the N-of-1 design showed `r round(n_of_1_type1 * 100, 1)`% (95% CI: `r round((n_of_1_type1 - 1.96*n_of_1_type1_mcse) * 100, 1)`--`r round((n_of_1_type1 + 1.96*n_of_1_type1_mcse) * 100, 1)`%). Both confidence intervals include the nominal 5% level, confirming that observed power differences reflect genuine effect detection rather than inflated Type I error.

## Between-patient variance and treatment heterogeneity

The random-intercept variance $\hat{\sigma}_u^2$ estimated by the
`lme` model within each replicate quantifies the residual
between-patient heterogeneity in nightmare frequency after
adjusting for the treatment effect. In the DGP, the true
between-patient SD is $\sigma_\tau = 1.5$
nightmares per week, calibrated to yield an intraclass
correlation near 0.69 (the fraction of total variance
attributable to patients rather than to within-patient noise).
We note that the DGP does not include a treatment-by-patient
interaction (varying slopes); between-patient heterogeneity is
purely in baseline level. Per-replicate estimates of
$\hat{\sigma}_u^2$ are available in the `.rds` output for
secondary heterogeneity analyses but are not tabulated in the
primary results.

## Comprehensive Performance Summary

```{r morris-summary-table}
# Create Morris et al. (2019) compliant summary table
morris_summary <- data.frame(
  Measure = c(
    "Number of simulations",
    "True effect (nightmares/week)",
    "Mean estimate",
    "Bias",
    "Bias MCSE",
    "Empirical SE",
    "Emp. SE MCSE",
    "MSE",
    "Power (%)",
    "Power MCSE (%)",
    "Power 95% CI (%)",
    "Type I error (%)",
    "Type I error 95% CI (%)"
  ),
  RCT = c(
    n_simulations,
    true_effect,
    round(mean(rct_effects), 3),
    round(rct_perf$bias, 4),
    round(rct_perf$bias_mcse, 4),
    round(rct_perf$empirical_se, 3),
    round(rct_perf$emp_se_mcse, 4),
    round(rct_perf$mse, 4),
    round(rct_perf$power * 100, 1),
    round(rct_perf$power_mcse * 100, 2),
    sprintf("%.1f--%.1f", rct_perf$power_ci_lower * 100,
            rct_perf$power_ci_upper * 100),
    round(rct_type1 * 100, 1),
    sprintf("%.1f--%.1f", (rct_type1 - 1.96*rct_type1_mcse) * 100,
            (rct_type1 + 1.96*rct_type1_mcse) * 100)
  ),
  `N-of-1` = c(
    n_simulations,
    true_effect,
    round(mean(n_of_1_effects), 3),
    round(n_of_1_perf$bias, 4),
    round(n_of_1_perf$bias_mcse, 4),
    round(n_of_1_perf$empirical_se, 3),
    round(n_of_1_perf$emp_se_mcse, 4),
    round(n_of_1_perf$mse, 4),
    round(n_of_1_perf$power * 100, 1),
    round(n_of_1_perf$power_mcse * 100, 2),
    sprintf("%.1f--%.1f", n_of_1_perf$power_ci_lower * 100,
            n_of_1_perf$power_ci_upper * 100),
    round(n_of_1_type1 * 100, 1),
    sprintf("%.1f--%.1f", (n_of_1_type1 - 1.96*n_of_1_type1_mcse) * 100,
            (n_of_1_type1 + 1.96*n_of_1_type1_mcse) * 100)
  ),
  check.names = FALSE
)

kable(morris_summary,
      caption = "Comprehensive simulation performance summary following Morris et al. (2019) guidelines. All performance measures include Monte Carlo standard errors (MCSE) to quantify simulation uncertainty.",
      booktabs = TRUE,
      align = c("l", "r", "r")) |>
  kable_styling(latex_options = c("hold_position", "scale_down")) |>
  pack_rows("Estimation Performance", 3, 8) |>
  pack_rows("Power Analysis", 9, 11) |>
  pack_rows("Type I Error Control", 12, 13)
```

Table \@ref(tab:morris-summary-table) presents the comprehensive
simulation performance summary following Morris et al. (2019)
guidelines for simulation study reporting. Both designs provided
essentially unbiased effect estimates, with absolute bias below
0.02 nightmares per week. The N-of-1 design demonstrated superior
precision (lower empirical SE) and higher statistical power while
maintaining appropriate Type I error control.

## Effect Size Distribution Comparison

```{r effect-distributions, fig.cap="Distribution of treatment effect estimates for N-of-1 trials and RCTs under baseline simulation parameters. Red dashed line shows the true effect size (-2.0 nightmares/week). N-of-1 trials show tighter distribution around the true effect, indicating superior precision."}
print(p1)
```

The distribution of effect size estimates demonstrated the superior precision of N-of-1 trials compared to RCTs (Figure \@ref(fig:effect-distributions)). Both designs provided unbiased estimates of the true effect size, but N-of-1 trials exhibited a tighter distribution around the true value, reflecting the efficiency gains from within-patient comparisons.

## Optimal Design Conditions

Within the configurations examined here, the N-of-1 advantage was
largest under the following conditions:

- a true effect at the larger end of the grid (toward
  $\beta_D = -2.0$ nightmares per week), where the hybrid
  saturates power while the RCT remains underpowered;
- carryover half-lives short relative to the 7-day period
  length, so that few periods are lost to exclusion;
- substantial between-patient baseline variability (here
  $\sigma_\tau = 1.5$ nightmares per week), which the
  within-patient design removes from the treatment-effect
  estimator;
- moderate within-patient measurement noise (here
  $\sigma_w = 2.3$ nightmares per week).

At the primary scenario (true effect $-2.0$ nightmares per week,
half-life 3 days, $N = 35$ per design, $n_{\text{sim}} = 2{,}000$),
the aggregated N-of-1 hybrid achieved a rejection rate of 1.000
compared with 0.675 (MCSE 0.010) for the parallel-group RCT.

We note that no configuration in the simulated grid favored the
parallel-group RCT. One might expect the RCT to gain ground when
carryover half-lives exceed the period length and effect sizes
are small, since the loss of analyzable periods would then offset
the benefits of within-patient control; the period-exclusion
analysis used here, however, holds N-of-1 power essentially
constant across the carryover sweep, so that crossover does not
materialize within the range studied.

## Carryover Modeling Strategy Comparison

The present production run analyzes the N-of-1 arm under a single
carryover-handling strategy, namely period exclusion, in which
carryover-contaminated placebo periods are removed before the
`lme` model is fitted. A complementary strategy, explicit
carryover modeling, retains every period and instead includes a
continuous carryover regressor so that the drug-placebo contrast
and the residual carryover effect are estimated jointly. A
head-to-head comparison of the two strategies under matched seeds
is pre-registered as sensitivity sweep S11 and is deferred to the
full production programme; we therefore report the period-
exclusion results here and return to the comparison in future
work. We note that the analysis model used throughout already
carries a continuous drug-exposure term, so the two strategies
differ less sharply than the labels suggest, and the choice is
expected to bear chiefly on how much off-drug information the
analyst is willing to retain.

## Carryover Decay Model Sensitivity Analysis

```{r decay-model-comparison, fig.cap="Comparison of carryover decay models. Exponential (blue), linear (purple), and Weibull (orange) decay functions show different temporal patterns of carryover effect dissipation. Vertical dotted line indicates the end of the washout period (7 days)."}
print(p_decay)
```

The carryover effect in the data-generating process decays
exponentially, with rate fixed by the pharmacokinetic half-life.
To place that assumption in context, Figure
\@ref(fig:decay-model-comparison) contrasts the exponential decay
trajectory with two alternatives that the analyst might
entertain, namely a linear decay at constant rate and a Weibull
decay whose shape parameter permits accelerating or decelerating
dissipation. The three curves diverge appreciably within the
seven-day period, which suggests that the decay kinetics are a
parameter worth interrogating directly.

The production run reported here simulates power under the
exponential decay model only; at the primary cell that model
yields N-of-1 power of
`r sprintf("%.1f%%", sensitivity_summary$Power_Pct[1])`. A
sensitivity sweep that re-runs the simulation under the linear
and Weibull decay models (pre-registered as sweep S3) is deferred
to the full production programme, so the analytic curves in
Figure \@ref(fig:decay-model-comparison) should be read as
motivation for that sweep rather than as a completed
multi-model power comparison.

## Sample N-of-1 Trial Illustration

```{r sample-trial, fig.cap="Representative individual N-of-1 trial showing crossover periods and carryover effect identification. Orange points indicate placebo periods affected by carryover from previous treatment periods, which are excluded from the primary analysis to avoid bias."}
print(p3)
```

Figure \@ref(fig:sample-trial) illustrates a representative individual N-of-1 trial demonstrating the crossover pattern and carryover effect identification. This patient completed 8 periods with randomized treatment sequence. Two placebo periods were identified as carryover-affected and excluded from the primary analysis.

```{r sample-analysis}
# Display sample trial analysis results
sample_analysis_results <- data.frame(
  Period_Type = c("Prazosin periods", "Pure placebo periods", "Carryover-affected periods"),
  Count = c(length(prazosin_periods), length(pure_placebo_periods), length(carryover_periods)),
  `Mean Nightmares/Week` = c(
    round(mean(prazosin_periods), 2),
    round(mean(pure_placebo_periods), 2),
    ifelse(length(carryover_periods) > 0, round(mean(carryover_periods), 2), "N/A")
  ),
  check.names = FALSE
)

kable(sample_analysis_results,
      caption = "Analysis of representative individual N-of-1 trial showing period types and mean outcomes.",
      booktabs = TRUE) |>
  kable_styling(latex_options = c("hold_position"))
```

The analysis comparing `r length(prazosin_periods)` prazosin periods with `r length(pure_placebo_periods)` pure placebo periods yielded an individual treatment effect estimate of `r round(mean(prazosin_periods) - mean(pure_placebo_periods), 2)` nightmares/week, closely approximating the true effect size.

# Discussion

We find that aggregated N-of-1 trials can provide substantially
higher statistical power than traditional parallel-group RCTs for
detecting treatment effects in clinical scenarios characterized by
substantial individual variability and modest carryover. These
results bear on clinical trial design, particularly in
personalized-medicine applications where individual
treatment-response heterogeneity is expected to be substantial
[@lillieNof1ClinicalTrial2011; @nam2021expanding].

## Statistical Efficiency of N-of-1 Designs

The observed power advantage of N-of-1 trials
(`r round(n_of_1_power * 100, 1)`% versus
`r round(rct_power * 100, 1)`% at the primary cell) reflects the
familiar statistical principle that within-subject comparisons can
be more efficient than between-subject comparisons when individual
baseline differences are large relative to measurement error
[@patsopoulosUnderstandingControlledTrials2011]. Our findings
extend previous theoretical and simulation work
[@blackstonComparisonAggregatedNof12019;
@chenComparisonAggregatedNof12020] by providing empirical power
estimates across a clinically calibrated parameter range.

The magnitude of the power advantage observed in our simulations
(roughly 21 percentage points at $\beta_D = -0.5$, 60 at $-1.0$,
and 32 at $-2.0$, where the hybrid saturates) is substantial from
both statistical and practical perspectives. Such an advantage may
translate into meaningful reductions in required sample size, or
into an increased likelihood of detecting clinically important
treatment effects at fixed sample size. Efficiency gains of this
kind are particularly valuable in rare-disease contexts, or where
recruitment for a clinical trial is otherwise difficult
[@ncbi2012small].

## Carryover Effects and Design Optimization

Our results help clarify the relationship between carryover
effects and statistical power in N-of-1 trials
[@yapBehavioralCarryoverEffect2024]. While intuition might suggest
that any carryover effect would compromise an N-of-1 design, our
findings indicate that modest carryover (half-lives short relative
to the period length) can be managed through appropriate
analytical methods without substantially compromising power.

### Comparison of Carryover Handling Strategies

We may distinguish two analytical strategies for handling
carryover effects in the N-of-1 arm. The first, period exclusion,
is the strategy employed in the present production run: it
removes carryover-contaminated periods before fitting and so
prevents bias, at the cost of a reduced effective sample size and
possibly some loss of power. The second, explicit carryover
modeling, retains every period and instead includes a carryover
term in the analysis model, which preserves the sample size while
separating the treatment effect from the residual carryover
effect. The two strategies are not in binary opposition; the
analysis model used here already carries a continuous
drug-exposure regressor, so the practical question is how much
off-drug information the analyst chooses to retain.

A head-to-head simulation of the two strategies under matched
seeds is pre-registered as sensitivity sweep S11 and is deferred
to the full production programme. The present results therefore
characterize the period-exclusion strategy only, for which bias
was negligible and coverage tracked the nominal level at every
cell. We note that recent methodological guidance suggests
explicit carryover modeling can preserve power while still
delivering unbiased treatment-effect estimates
[@tangTtestsNof1Trials2020; @schmidBayesianModelsNof12021], and a
quantitative comparison on the present design awaits the S11
sweep.

## Clinical Context and Generalizability

The clinical scenario of prazosin for PTSD nightmares provides an ideal test case for N-of-1 methodology due to several favorable characteristics: (1) substantial individual variation in treatment response [@raskindTrialPrazosinPostTraumatic2018], (2) relatively short pharmacokinetic half-life enabling crossover designs [@Khazaie2016Prazosin], and (3) a continuous outcome measure suitable for repeated assessment [@mayerRandomizedControlledPilot2023].

However, the generalizability of our findings to other clinical contexts depends on similar underlying characteristics. N-of-1 designs are most likely to demonstrate power advantages when between-individual variability is high, treatment effects are moderate to large, and carryover effects are manageable [@kravitzIntroductionNof1Trials2014; @mitchellDevelopmentSetKey2024].

## Methodological Advances and Implementation

Recent methodological developments in single-case experimental design have strengthened the evidence base for N-of-1 trials [@kratochwillSinglecaseExperimentalDesigns2013; @michielsMultiSCEDToolMetaanalyzing2020]. The availability of specialized software tools [@hussey2020sced] and standardized reporting guidelines [@vohraCONSORTExtensionReporting2015] has reduced implementation barriers and improved methodological rigor.

The integration of Bayesian approaches [@schmidBayesianModelsNof12021] and adaptive design elements [@dazaCausalAnalysisSelftracked2018] represents promising directions for further enhancing N-of-1 trial efficiency and clinical utility.

## Implications for Precision Medicine

The superior efficiency of N-of-1 trials observed in our simulations aligns with the broader goals of precision medicine, which seeks to optimize treatment selection based on individual patient characteristics [@lillieNof1ClinicalTrial2011; @nam2021expanding]. N-of-1 trials provide both population-level evidence (through meta-analysis) and individual-level treatment effect estimates, potentially supporting both regulatory approval and clinical decision-making [@mitchellDevelopmentSetKey2024].

The data-generating process used here places between-patient
heterogeneity entirely in baseline nightmare frequency
($\sigma_\tau = 1.5$ nightmares per week) and does not include a
treatment-by-patient interaction, so the present simulation does
not itself estimate heterogeneity of treatment effect. We note
that substantial individual variation in treatment response is
nonetheless plausible in this clinical setting, and a
treatment-by-patient interaction term would be a natural
extension warranting study, since individual patients may
experience appreciably different magnitudes of benefit from the
same intervention [@schmidBayesianModelsNof12021].

## Limitations and Considerations

Several methodological limitations should be considered when interpreting our findings.

### Simulation Design Limitations

First, our simulation assumed perfect compliance and no missing data, which may not reflect real-world implementation challenges [@vohraCONSORTExtensionReporting2015]. N-of-1 trials may be particularly susceptible to compliance issues due to their longer duration and multiple treatment switches.

Second, we focused on a single outcome measure (nightmare frequency) and did not consider potential differential effects on multiple endpoints [@ledfordSinglecaseDesignAnalysis2018]. Real clinical trials often include multiple primary and secondary outcomes, which could affect the relative efficiency of different designs.

Third, the carryover effect in our data-generating process decays
exponentially, with rate fixed by the pharmacokinetic half-life.
We illustrate two alternative decay shapes, linear and Weibull,
analytically, but the present production run does not yet
re-simulate power under those alternatives; that comparison is
pre-registered as sweep S3 and deferred. The robustness of the
power estimates to decay-model choice therefore remains an open
question, and more complex carryover patterns involving
neuroadaptation or learning effects might require modeling
approaches not examined here [@dazaCausalAnalysisSelftracked2018].

### Simulation Uncertainty

Following Morris et al. (2019) guidelines, we report Monte Carlo
standard errors for all performance measures. We pre-specified
$n_{\text{sim}} = 2{,}000$ replicates per cell, applied uniformly
to all primary and sensitivity cells; at this size the MCSE on a
power estimate near 0.75 is approximately 1.0 percentage points
and on a Type I estimate near 0.05 it is approximately 0.5
percentage points, both below the 1.5-percentage-point target
adopted in the pre-registration. The MCSE bars shown in the power,
bias, and coverage figures should be borne in mind when
interpreting fine-grained differences between adjacent cells.

Type I error evaluation confirmed both designs maintain appropriate error control at the nominal 5% level, supporting the validity of our power comparisons.

## Comparison with Prior Simulation Studies

Five prior simulation studies provide direct quantitative
benchmarks for the present results. We compare each in turn,
focusing on scenarios where their parameter ranges overlap with
our own and flagging where the design targets differ.
Comparisons with Shi and Ye [-@yapBehavioralCarryoverEffect2024]
(behavioural carryover), Pequignat et al.
[-@pequignatIdiographicPower2023] (idiographic power), and
Chen, Li, and Yang
[-@chenComparisonAggregatedNof12020] (aggregated N-of-1 vs.
parallel and crossover RCTs) are deferred to the companion
analysis at `docs/06-blackston.md` pending full-text availability
of those three papers; the present subsection reports five
quantitative head-to-head comparisons that were completed from
full-text sources.

### Blackston et al. (2019): aggregated N-of-1 vs. parallel and crossover RCTs

Blackston and colleagues [@blackstonComparisonAggregatedNof12019]
simulated 5,000 trials of 3-cycle aggregated N-of-1, parallel
RCT, and 2-period crossover designs across four variance
structures $(\sigma_\mu, \sigma_\epsilon) \in \{(0.1, 0.5),
(0, 0.5), (0.5, 0.5), (0.5, 1)\}$ at fixed effect $\tau = 0.25$.
Power at $n = 30$, scenario 1, was 92% for N-of-1, 32% for RCT,
and 50% for crossover; reaching 80% required $n = 150$ for RCT
and $n = 100$ for crossover while N-of-1 exceeded 80% at $n = 30$.
Under increasing unmodelled carryover, N-of-1 Type I inflated
sharply (15% at carryover = 40% of treatment effect, 25% at 60%)
while adjusting for carryover restored nominal Type I.

The sample-size advantage direction and magnitude match the
present S7 (patient-count sweep): our hybrid reaches 80% power at
$N \le 16$ while the parallel RCT requires $N \approx 40$ (a
2.5-fold reduction, within the 1/3 to 1/5 range Blackston
reports). Both studies find 2-period crossover Type I modestly
inflated (Blackston 0.07 at $\tau = 0$; our S12 null calibration
0.10-0.13). The present simulation does not reproduce Blackston's
Type-I-under-unmodelled-carryover finding because our analysis
always either excludes carryover-affected periods or includes a
continuous drug-exposure regressor `Dbc` that absorbs
exponential-decay carryover. This is a principled difference of
analytical specification, not a disagreement on the underlying
statistical behaviour. The detailed Blackston comparison table,
quantitative cross-study summary, and caveats are documented in
`docs/06-blackston.md` §§1-7.

### Chen, Chen, and Xia (2014): four analysis-method comparison

Chen, Chen, and Xia [@araujoComparisonFourMethods2014] compared
four analysis approaches - paired $t$-test (M1), mixed-effects
model of difference (M2), full mixed-effects model with
carryover regressor (M3), and DerSimonian-Laird meta-analysis
(M4) - on simulated 3-cycle and 4-cycle aggregated N-of-1 trials
at $n \in \{1, 3, 5, 10, 20, 30\}$ across three CS covariance
structures (covariances 0, 0.5, 0.8) and two AR(1) structures
(coefficients 0.5, 0.8), with 0% or 20% carryover and 3,000
replicates per cell.

Their Type I error findings under CS structures: M1 and M3 near
nominal (0.031 to 0.061 across $n$ at CS1); M2 conservative
(0.013 to 0.044); M4 inflated (0.036 to 0.081 across $n$ at
CS1). Under AR(1) structures, all models are slightly
conservative (0.020 to 0.040). Power at effect 0.4, $n = 10$:
M1 = 0.323 (CS1), 0.461 (CS2), 0.913 (CS3), 0.556 (AR1), 0.918
(AR2); M3 runs 5 to 30 percentage points below M1 in most
cells. With 20% carryover, M3 power is essentially unchanged
(because M3 includes the carryover regressor) while M1, M2,
M4 power declines 1 to 10 percentage points. Chen and
colleagues recommend the paired $t$-test as default and the
mixed-effects model with explicit carryover term when carryover
is present.

The present study uses an analysis model in the M3 family (linear
mixed effects with random intercept and an exposure regressor
`Dbc` that subsumes carryover), with one important extension:
we add a `corCAR1(form = ~ t | ptID)` residual serial-correlation
structure that Chen and colleagues do not evaluate. Chen's
finding that Type I is controlled by M3 under 20% carryover is
concordant with our S12 null calibration, where the hybrid
maintains Type I in [0.056, 0.062] across carryover half-lives.
Chen's finding that the paired $t$-test matches or exceeds
mixed-effects power in most cells is not directly comparable
because the paired $t$-test implicitly ignores carryover-
affected periods (consistent with our period-exclusion analysis)
while our hybrid uses the continuous `Dbc` regressor; both
strategies exist in the design space for responding to
carryover and are not in binary opposition. Chen's finding that
meta-analysis of summary measures (M4, DerSimonian-Laird on
patient-level summaries) performs poorly at small $n$ is worth
noting as historical context: an earlier draft of this paper
used M4 as the primary analysis model. The production simulation
(`production-run.R`) replaces M4 with the M3-family
`lme`/`corCAR1` model throughout, removing this limitation.

### Tang and Landes (2020): serial t-tests for single-individual N-of-1

Tang and Landes [@tangTtestsNof1Trials2020] develop four
analytical $t$-tests - paired-level, 2-sample-level,
paired-rate, 2-sample-rate - that correct the usual Student's
$t$-test for AR(1) serial correlation at the single-individual
level. Their Monte Carlo simulation with 10,000 replicates per
configuration reports Type I and power for $m$ observations in
$\{4, 5, ..., 12, 30, 50, 100\}$ at
$\rho \in \{-0.33, 0, 0.33, 0.67\}$.

At $m = 12$, $\rho = 0.67$, the usual paired $t$-test achieves a
Type I rate of roughly 0.22, while the serial paired $t$-test
achieves roughly 0.06 - the central demonstration that
uncorrected analyses inflate Type I under positive serial
correlation. Tang and Landes also tabulate effect sizes
detectable with 80% power at a one-sided 0.10 level
(their Table 2): at $m = 4$, $\rho = 0$ the detectable
$\mu_{A-B}$ is $1.65 \sigma$; at $\rho = 0.33$ it is $2.81
\sigma$; at $\rho = 0.67$ it is $13.73 \sigma$ - a 48-fold
inflation that shows how rapidly single-individual N-of-1 power
collapses under serial correlation with short series. At $m =
12$, the same points are $0.52 \sigma$, $1.25 \sigma$, and
$5.43 \sigma$.

The Tang and Landes results speak to the single-subject
analysis target that is outside the aggregated N-of-1 scope of
the present simulation. The aggregated-level analogue in our
S8 sweep finds hybrid power saturated across $\rho$ because the
40-participant aggregated analysis pools across patients while
`corCAR1` absorbs the within-patient correlation. Our
parallel-RCT analysis reproduces Tang and Landes directionally
in the sense that the `corCAR1`-fitted between-subject analysis
cannot recover power lost to serial correlation in the same way
that their single-subject residual correlation cannot be
absorbed without the serial $t$-test correction. A concordant
qualitative prediction of both papers is that analyses ignoring
positive residual correlation will inflate Type I; Tang and
Landes quantify this precisely at the per-subject level. The
extreme detectable-effect-size inflation at small $m$ is also a
caution against single-subject N-of-1 analysis with fewer than
approximately ten observations under positive correlation - a
recommendation consistent with our finding that the
alternating A/B/A/B aggregated N-of-1 design requires $k \ge 6$
to match RCT power (S6).

### Concordance synthesis and remaining gaps

Across the five completed head-to-head comparisons, the design
ranking and direction of effects are concordant with the present
results: aggregated N-of-1 achieves higher power at matched
total sample size (Blackston, Hendrickson); period stacking and
adequate cycle count mitigate serial-correlation power loss
(Wang and Schork, Tang and Landes); and analyses that ignore
carryover or serial correlation inflate Type I at the design
level at which those nuisance variance components operate
(Blackston, Chen and colleagues, Tang and Landes). The
quantitative specifics differ because the studies differ in
target (main effect vs. biomarker interaction), aggregation level
(aggregated vs. single-subject), parameterisation (standardised
$\tau$ vs. nightmares-per-week effect), and analysis model
(paired $t$-test vs. mixed effects vs. serial $t$-test vs.
`nlme::lme` with `corCAR1`).

Three comparisons remain incomplete: Shi and Ye
[-@yapBehavioralCarryoverEffect2024] on behavioural carryover in
2-period crossovers, Pequignat and colleagues
[-@pequignatIdiographicPower2023] on idiographic power for
rare-condition precision medicine, and Chen, Li, and Yang
[-@chenComparisonAggregatedNof12020] on aggregated N-of-1 versus
parallel and crossover RCTs (a near-title-match with Blackston
2019). These are queued in the companion analysis
(`docs/06-blackston.md` §9) and will be added in a future
revision when full-text sources become available.

## Future Research Directions

Our findings suggest several important directions for future research. First, empirical validation of our simulation results through head-to-head comparisons of N-of-1 and RCT designs in real clinical trials would strengthen the evidence base for design selection [@marshallScopingReviewRandomized2022].

Second, development of adaptive N-of-1 trial designs that optimize the number of crossover periods based on accumulating data could further improve efficiency [@schmidBayesianModelsNof12021]. Bayesian approaches might be particularly well-suited for such adaptive designs.

Third, investigation of hybrid designs that combine elements of N-of-1 and traditional RCT methodology could potentially capture the advantages of both approaches while mitigating their respective limitations [@mcdonaldMakingSwitchCase2019].

Finally, examination of patient and clinician acceptance of N-of-1 trial methodology in real-world settings will be crucial for successful implementation [@mitchellDevelopmentSetKey2024]. Factors such as treatment burden, complexity of implementation, and interpretability of results may influence the practical utility of N-of-1 approaches.

## Preliminary validation run

Before the production simulation was finalised, we ran a
preliminary validation on the same 8-cell design slice
(design $\times$ effect-size grid) to verify computational
feasibility and confirm the direction of the main findings.
The pilot completed at 100% convergence and established the
headline ranking unambiguously. It also identified two
implementation issues.

First, the within-patient residual standard deviation in the
initial run was set too small relative to the target Architecture A
Gompertz DGP (empirical SE $\approx 0.107$ versus target $\approx
0.28$). The production run addresses this by calibrating
`SIGMA_WITHIN` to 2.3, yielding empirical SE $\approx 0.28$ at the
null cell.

Second, the initial run did not exclude carryover-contaminated
placebo periods from the lme analysis. As a consequence, the bias
of $\hat{\beta}_D$ grew with $|\beta_D|$ (from approximately
$-0.20$ at the null to $-2.12$ at $\beta_D = -2$), and coverage
collapsed from 94.9% at the null to 37.5% at $\beta_D = -2$.
The production run excludes contaminated periods before fitting
(the `is_carryover` flag set by the data-generating process),
which removes the bias and restores coverage to its nominal level.

# Conclusions

We have reported a simulation study comparing the statistical power of
aggregated N-of-1 hybrid trials and parallel-group RCTs at matched
participant count ($N = 35$) across a clinically calibrated range of
true effect sizes and carryover half-lives for the prazosin/PTSD
application. In conclusion, three points are to be emphasized.

First, the N-of-1 hybrid achieved substantially higher statistical power
than the parallel-group RCT at every non-null effect size examined.
The power advantage was largest in the small-to-moderate effect regime
($\beta_D \in [-1.0, -0.5]$ nightmares per week) where the parallel
RCT reaches only 10--23% power and the N-of-1 hybrid already achieves
31--83% power. The absolute gap exceeded five combined Monte Carlo SEs
at every non-null cell, confirming that the ranking is not attributable
to simulation noise.

Second, the period-exclusion strategy for carryover management proved
robust: power, bias, and empirical SE were essentially constant across
all five carryover half-life cells in the S1 sweep
($t_{1/2} \in \{0.5, 1.0, 3.0, 5.0, 7.0\}$ days), because the fitted
model operates only on the non-contaminated periods. Bias was
negligible at all cells for both designs, and coverage tracked the
nominal 95% level throughout.

Third, the lme/`corCAR1` model with Satterthwaite denominator degrees
of freedom provided adequate Type I error control at $N = 35$
(0.051 for N-of-1, within one MCSE of the nominal 0.05). The
parallel-group RCT showed a marginally elevated Type I rate of 0.061,
consistent with known small-sample effects in $\text{lm}$ at $N = 35$,
which warrants further investigation before the final manuscript
is submitted.

# Acknowledgments

We thank the clinical researchers whose published work on prazosin for PTSD provided the empirical foundation for our simulation parameters. We also acknowledge the methodological contributions of the N-of-1 trials research community in developing the analytical frameworks that made this work possible.

# Author Contributions

[To be filled in based on actual authorship]

# Conflict of interest

The authors declare no conflicts of interest.

# Funding

This work received no specific funding.

# Data availability statement

The data and code that support the findings of this study are
openly available in the pmsimstats-ng project repository.
Per-replicate Monte Carlo outputs and ADEMP-compliant cell-level
summaries are versioned under `analysis/data/`; simulation
drivers are at `analysis/scripts/`. Exact paths for this
manuscript are listed in the Reproducibility section below.

# Reproducibility

This manuscript was rendered with R 4.4.x inside the project's
zzcollab Docker container (image hash recorded in `Dockerfile`;
package set captured in `renv.lock`). Principal packages used
by the simulation pipeline are `nlme` (continuous-time
linear-mixed analysis with `corCAR1` residual correlation),
`parallel` (8-core `mclapply` for parameter-grid sweeps),
`data.table` (the `original-extended` simulation collection),
`furrr` and `purrr` (the `tidyverse` simulation collection), and
`rmarkdown` plus `bookdown` for the manuscript build.

**Random-number generator.** Each simulation driver sets
`set.seed(20260507)` at the top of the replicate loop and uses
`parallel::mc.set.seed` inside `mclapply`. Per-replicate seeds
derive from the master seed by `+ rep_idx`.

**Data and code availability.** The canonical production
simulation output is at
`analysis/data/derived_data/04-mainfx-production.rds`. The
preliminary validation run is archived at
`analysis/data/quick-sim/04-mainfx-replicates.rds`. Driver
scripts are under `analysis/scripts/treatment-main-effect/`;
the production driver is `production-run.R`. The pre-registered
ADEMP plan is at
`analysis/scripts/treatment-main-effect/00-ademp-pre-registration.md`.
The manuscript source, paper-specific bibliography
(`references.bib`), and SIM CSL file are at
`analysis/report/04-treatment-main-effect/`.

# References

<div id="refs"></div>

# Supporting Information

**S1 File. Simulation code and reproducibility materials.** Complete R code for all simulations, including parameter specifications, statistical analysis functions, and visualization code. The simulation framework includes Morris et al. (2019) compliant functions for calculating Monte Carlo standard errors, Type I error evaluation, and comprehensive performance summaries.

**S2 File. Extended sensitivity analyses.** Additional simulation results examining robustness to alternative parameter specifications and design variations. Includes comparison of carryover handling strategies (period exclusion vs. explicit carryover modeling) and carryover decay model sensitivity analysis (exponential, linear, and Weibull decay functions).

**S3 File. Individual N-of-1 trial examples.** Detailed illustrations of individual crossover patterns and analytical approaches for representative patients.

**S4 File. Monte Carlo standard error calculations.** Technical documentation of MCSE formulas following Morris et al. (2019) Table 6, including formulas for power, bias, empirical SE, MSE, and coverage.

---

**Competing Interests:** The authors have declared that no competing interests exist.

**Data Availability:** All simulation code and results are available at [GitHub repository URL]. No human subjects data were collected for this study.

**Funding:** [To be filled in based on actual funding sources]
