---
title: "Carryover-sensitivity study: development-run review"
subtitle: "96-cell grid, five analysis methods, 150 replicates per cell"
author:
  - pmsimstats team
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
knit: (function(input, ...) { d <- dirname(input); while (!file.exists(file.path(d, 'tools', 'stamp-render.R')) && d != dirname(d)) d <- dirname(d); source(file.path(d, 'tools', 'stamp-render.R'))$value(input) })
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: true
    number_sections: true
    keep_tex: false
    highlight: tango
fontsize: 11pt
geometry: "left=2.5cm,right=2.5cm,top=2cm,bottom=2cm"
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
---

<!--
  DEV REVIEW DOCUMENT
  ===================
  This Rmd loads the development-run simulation outputs directly and
  renders all figures inline. It does NOT depend on the pre-rendered
  PDFs in analysis/figures/ or the summarised RDS files in
  analysis/data/. Run it with:

    Rscript -e "rmarkdown::render(
      'analysis/report/02-carryover-sensitivity/report-devresults.Rmd')"

  The dev run was executed at 150 reps/cell with the --robust flag,
  which adds the lme+CR2 and GEE+MD analyses to M1/M2/M3:
    Rscript analysis/scripts/carryover-sensitivity/01-run-factorial.R \
      --dev --robust --reps 150
    Rscript analysis/scripts/carryover-sensitivity/04-run-sensitivity-blocks.R \
      --dev --robust --reps 150

  All figures and tables are computed inline from:
    analysis/scripts/carryover-sensitivity/output/01-dev.rds
    analysis/scripts/carryover-sensitivity/output/04-sensitivity-dev.rds

  MC SEs at 150 reps are approximately 0.041 at p = 0.5 (vs 0.022 at
  500 reps). Use this document for ranking and direction checks, not
  for precise power estimates.
-->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  warning = FALSE,
  message = FALSE
)

suppressPackageStartupMessages({
  library(dplyr)
  library(tibble)
  library(tidyr)
  library(ggplot2)
  library(knitr)

## The archived simulation output labels the three analysis
## specifications A1/A2/A3. The manuscript reports them as M1/M2/M3 so
## that the letters A/B/C stay reserved for DGP architectures. Data
## values are never rewritten; they are mapped at display time only.
spec_display <- function(x) {
  unname(c(A1 = 'M1', A2 = 'M2', A3 = 'M3')[as.character(x)])
}
})

alpha <- 0.05

## ------------------------------------------------------------------
## summarise_cells(): inline copy of 02-summarise-grid.R logic.
## Groups by spec, so it handles the five analysis methods without
## modification.
## ------------------------------------------------------------------
summarise_cells <- function(results, grid_names) {
  if (!('std_error' %in% colnames(results))) {
    results <- results |> mutate(std_error = NA_real_)
  }
  results |>
    group_by(across(all_of(c(grid_names, 'spec')))) |>
    summarise(
      n_reps             = n(),
      n_converged        = sum(converged, na.rm = TRUE),
      power              = mean(p_value < alpha, na.rm = TRUE),
      mc_se_power        = sqrt(power * (1 - power) / n_reps),
      sigma_BR           = 8,
      true_value         = -first(c_bm) * sigma_BR,
      mean_estimate      = mean(estimate, na.rm = TRUE),
      bias               = mean_estimate - true_value,
      empirical_se       = sd(estimate, na.rm = TRUE),
      mc_se_bias         = empirical_se / sqrt(n_reps),
      mse                = bias^2 + empirical_se^2,
      mc_se_mse          = sd((estimate - true_value)^2,
                              na.rm = TRUE) / sqrt(n_reps),
      coverage           = mean(
        abs(estimate - true_value) < qnorm(0.975) * std_error,
        na.rm = TRUE),
      mc_se_coverage     = sqrt(coverage * (1 - coverage) / n_reps),
      converged_frac     = mean(converged, na.rm = TRUE),
      non_convergence_rate = 1 - converged_frac,
      mc_se_non_convergence = sqrt(
        non_convergence_rate * (1 - non_convergence_rate) / n_reps),
      .groups = 'drop'
    )
}

## ------------------------------------------------------------------
## Load and summarise the dev outputs
## ------------------------------------------------------------------
repo_root  <- here::here()
raw1_path  <- file.path(repo_root,
  'analysis/scripts/carryover-sensitivity/output/01-dev.rds')
raw1       <- readRDS(raw1_path)
grid1      <- summarise_cells(raw1$results, names(raw1$grid))

raw2_path  <- file.path(repo_root,
  'analysis/scripts/carryover-sensitivity/output/04-sensitivity-dev.rds')
raw2       <- readRDS(raw2_path)
grid2      <- summarise_cells(raw2$results, names(raw2$grid))

## ------------------------------------------------------------------
## Five analysis methods: three model-based specifications plus two
## robust-inference analyses (both on the M2 / Dbc predictor).
## ------------------------------------------------------------------
method_levels <- c('M1', 'M2', 'M3', 'lme+CR2', 'GEE+MD')
method_labels <- c('M1' = 'M1 binary',
                    'M2' = 'M2 Dbc (matched)',
                    'M3' = 'M3 lagged',
                    'lme+CR2' = 'lme+CR2 (robust)',
                    'GEE+MD'  = 'GEE+MD (robust)')

theme_paper <- theme_bw(base_size = 10) +
  theme(
    panel.grid.minor   = element_blank(),
    strip.background   = element_rect(fill = 'grey92', colour = NA),
    legend.position    = 'top'
  )

spec_scale <- scale_colour_manual(
  values = c('M1' = '#1f78b4', 'M2' = '#33a02c', 'M3' = '#e31a1c',
             'lme+CR2' = '#ff7f00', 'GEE+MD' = '#6a3d9a'),
  labels = method_labels,
  limits = method_levels)

as_method <- function(x) factor(x, levels = method_levels)
```

# Overview

This document presents pipeline-verification results for the
carryover-sensitivity manuscript from the development simulation
run completed 2026-05-20. The run uses a 96-cell reduced Tier 1
grid (150 replicates per cell, MC SE $\approx 0.041$ at $p = 0.5$)
and a 31-cell Tier 2 sensitivity grid. All figures and tables are
computed inline from the development RDS files; nothing is loaded
from pre-rendered figures or the shared `analysis/data/` summaries.

**Five analysis methods.** Beyond the three analysis-model
specifications of the parent study, M1 (binary $D_b$), M2
(exposure-weighted $D_{bc}$), and M3 (lagged-treatment), this dev
run carries two robust-inference analyses, both applied to the M2
($D_{bc}$) predictor. `lme+CR2` is the conditional `lme` fit with
its model-based standard error replaced by a CR2 bias-reduced
cluster-robust sandwich. `GEE+MD` is the marginal
generalized-estimating-equations fit with a Mancl-DeRouen
bias-corrected sandwich. The two robust analyses are produced by
the `--robust` flag of the simulation drivers, in anticipation of
carrying them into the production pipeline; their rationale and
validation are documented in manuscript 10
(`analysis/report/10-interaction-test-calibration/`).

The purpose of this document is to verify that:

a) the M2 $>$ M1 $\approx$ M3 power ranking is visible at 150 reps;
b) the three model-based specifications are conservative under the
   null, while the robust analyses move the type-I error toward
   the nominal 0.05;
c) lme+CR2 recovers the power lost to conservatism, while GEE+MD
   carries a modest residual efficiency cost as a marginal-model
   estimator;
d) the Tier 2 sensitivity-block trends are plausible.

Precise point estimates should be drawn from the production run
(500 reps, 540 cells) rather than from this document.

**Dev grid coverage.** The Tier 1 dev grid retains both DGP
architectures and all three DGP decay forms (the primary axes),
trimming secondary axes to two representative values each.
`t1half` is reduced to $\{0.0, 1.0\}$ weeks; design to
$\{\text{CO}, \text{Hybrid}\}$; $N$ to $\{35, 70\}$; $c_{bm}$ to
$\{0.0, 0.45\}$; Weibull uses only $k = 0.7$. The OLBDC design and
$k = 1.5$ Weibull shape are absent from the dev grid; any figure
filtered to those levels will appear empty.

```{r meta-table, echo = FALSE}
meta_df <- tibble::tibble(
  Item  = c('Tier 1 cells', 'Tier 1 reps/cell',
            'Tier 2 cells', 'Tier 2 reps/cell',
            'Analysis methods per cell',
            'Total Tier 1 rows', 'Total Tier 2 rows',
            'Run date'),
  Value = c(
    nrow(raw1$grid),
    raw1$meta$n_reps,
    nrow(raw2$grid),
    raw2$meta$n_reps,
    length(unique(raw1$results$spec)),
    nrow(raw1$results),
    nrow(raw2$results),
    format(raw1$meta$date, '%Y-%m-%d')
  )
)
knitr::kable(meta_df, booktabs = TRUE,
  caption = 'Development-run metadata read from the output RDS files.')
```

# Tier 1 results

## Headline performance at the reference cell

Table \@ref(tab:headline-dev) shows the Morris-style performance
measures at the reference cell (Architecture B, exponential DGP,
Hybrid design, $N = 70$, $c_{bm} = 0.45$), for all five analysis
methods at both carryover half-lives present in the dev grid.

```{r headline-dev}
hd <- grid1 |>
  filter(carryover_form == 'exponential', dgp_arch == 'mvn',
         design == 'Hybrid', N == 70, c_bm == 0.45) |>
  mutate(
    `t1/2`             = t1half,
    `Power (MC SE)`    = sprintf('%.3f (%.3f)',
                                  power, mc_se_power),
    `Bias (MC SE)`     = sprintf('%+.3f (%.3f)',
                                  bias, mc_se_bias),
    `Emp SE`           = sprintf('%.3f', empirical_se),
    `Coverage`         = sprintf('%.3f', coverage),
    `Non-conv`         = sprintf('%.3f', non_convergence_rate)
  ) |>
  arrange(`t1/2`, match(spec, method_levels)) |>
  mutate(spec = spec_display(spec)) |>
  select(`t1/2`, spec, `Power (MC SE)`, `Bias (MC SE)`,
         `Emp SE`, `Coverage`, `Non-conv`)

knitr::kable(
  hd, booktabs = TRUE,
  caption = 'Reference-cell performance (Architecture B, exponential
    DGP, Hybrid, $N = 70$, $c_{bm} = 0.45$, $n_{sim} = 150$). The
    lme+CR2 row shares the M2 interaction estimate (its bias and
    empirical SE track M2) and differs only in the standard error;
    GEE+MD is a separate marginal fit with its own estimate,
    modestly less efficient than the conditional lme. MC SEs are
    about twice those of the 500-rep production run.'
)
```

## Power by analysis method

```{r fig-power-by-spec, fig.cap = "Power versus carryover half-life under the exponential DGP, stratified by architecture and design, for all five analysis methods. Dev grid: design in {CO, Hybrid} only; $N = 70$, $c_{bm} = 0.45$, 150 reps per cell. lme+CR2 lifts power above M2 by recalibrating the conservative standard error; GEE+MD tracks just below M2, a modest residual efficiency cost of the marginal estimator relative to the conditional lme.", fig.width = 7.5, fig.height = 5}
d_a <- grid1 |>
  filter(carryover_form == 'exponential', c_bm == 0.45, N == 70) |>
  mutate(
    dgp_arch = factor(dgp_arch,
      levels  = c('mean_moderation', 'mvn'),
      labels  = c('Arch A (mean mod)', 'Arch B (MVN)')),
    spec   = as_method(spec),
    design = factor(design, levels = c('CO', 'Hybrid'))
  )

ggplot(d_a, aes(t1half, power, colour = spec, group = spec)) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 1.5) +
  facet_grid(dgp_arch ~ design) +
  scale_y_continuous(limits = c(0, 1),
    breaks = c(0, 0.25, 0.5, 0.75, 1)) +
  labs(
    x      = expression('Carryover half-life'~t['1/2']~'(weeks)'),
    y      = 'Power',
    colour = 'Analysis method',
    title  = 'Power vs carryover under matched exponential DGP',
    subtitle = expression(
      N==70*','~c[bm]==0.45*','~'150 reps/cell (dev)')
  ) +
  theme_paper +
  spec_scale
```

## Decay-form by analysis-method heatmap

```{r fig-heatmap, fig.cap = "Power at the reference carryover cell ($t_{1/2} = 1.0$ week, Architecture B, Hybrid design, $N = 70$, $c_{bm} = 0.45$) across DGP decay forms and the five analysis methods. The dev grid contains three DGP forms: Linear, Exponential, and Weibull $k = 0.7$.", fig.width = 6.5, fig.height = 3.4}
d_b <- grid1 |>
  filter(
    dgp_arch == 'mvn', design == 'Hybrid',
    c_bm == 0.45, N == 70, t1half == 1.0
  ) |>
  mutate(
    dgp_label = case_when(
      carryover_form == 'linear'      ~ 'Linear',
      carryover_form == 'exponential' ~ 'Exponential',
      carryover_form == 'weibull' & weibull_shape == 0.7 ~
        'Weibull (k=0.7)',
      TRUE ~ paste0(carryover_form, ' k=', weibull_shape)
    ),
    dgp_label = factor(dgp_label,
      levels = c('Linear', 'Exponential', 'Weibull (k=0.7)')),
    spec = as_method(spec)
  )

ggplot(d_b, aes(spec, dgp_label, fill = power)) +
  geom_tile(colour = 'white') +
  geom_text(aes(label = sprintf('%.2f', power)), size = 3.2) +
  scale_fill_gradient2(
    low = 'white', mid = '#fdd', high = '#08519c',
    midpoint = 0.4, limits = c(0, 1)) +
  labs(
    x        = 'Analysis method',
    y        = 'DGP decay form',
    fill     = 'Power',
    title    = 'Decay-form x analysis-method sensitivity',
    subtitle = expression(
      'Arch B, Hybrid,'~N==70*','~
      c[bm]==0.45*','~t['1/2']==1.0*' (dev, 150 reps)')
  ) +
  theme_paper
```

## Type-I error control

```{r fig-type1, fig.cap = "Empirical type-I error rates under the null ($c_{bm} = 0$), pooled across DGP decay forms and carryover half-lives, stratified by design and analysis method. $N = 70$; dashed line at nominal 0.05. The three model-based specifications sit below nominal; lme+CR2 sits near nominal; GEE+MD is near nominal under the Hybrid design but anti-conservative under CO.", fig.width = 7, fig.height = 3.6}
d_c <- grid1 |>
  filter(c_bm == 0, N == 70) |>
  mutate(
    spec = as_method(spec),
    design = factor(design, levels = c('CO', 'Hybrid')),
    dgp_arch = factor(dgp_arch,
      levels = c('mean_moderation', 'mvn'),
      labels = c('Arch A', 'Arch B'))
  )

ggplot(d_c, aes(spec, power, fill = dgp_arch)) +
  geom_boxplot(outlier.size = 0.8) +
  geom_hline(yintercept = 0.05, linetype = 'dashed',
             colour = 'grey40') +
  facet_wrap(~ design) +
  scale_y_continuous(limits = c(0, NA)) +
  labs(
    x        = 'Analysis method',
    y        = 'Empirical type-I error rate',
    fill     = NULL,
    title    = 'Type-I error control under null',
    subtitle = expression(
      c[bm]==0*','~N==70*','~
      'pooled across DGP forms and carryover levels (dev)')
  ) +
  theme_paper +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

The three model-based specifications M1, M2, and M3 sit visibly
below the nominal 0.05 line in Figure \@ref(fig:fig-type1), the
conservatism diagnosed in manuscript 10
(`analysis/report/10-interaction-test-calibration/`); lme+CR2 and
GEE+MD sit on or near it, confirming that the robust analyses
recalibrate the interaction test. Figure
\@ref(fig:fig-type1-arch) shows the same null cells split by DGP
architecture rather than by design. Because the two architectures
reduce to a bit-identical data-generating process when
$c_{bm} = 0$, the Arch A and Arch B panels are two samples from
one distribution and should coincide up to Monte Carlo noise; that
they do is a consistency check on the simulation.

```{r fig-type1-arch, fig.cap = "Empirical type-I error rates under the null ($c_{bm} = 0$, $N = 70$), the same null cells as the preceding figure but split by DGP architecture, with design as the fill. At the null the two architectures are a bit-identical data-generating process, so the two panels coincide up to Monte Carlo noise. Dashed line at nominal 0.05.", fig.width = 7, fig.height = 3.6}
ggplot(d_c, aes(spec, power, fill = design)) +
  geom_boxplot(outlier.size = 0.8) +
  geom_hline(yintercept = 0.05, linetype = 'dashed',
             colour = 'grey40') +
  facet_wrap(~ dgp_arch) +
  scale_y_continuous(limits = c(0, NA)) +
  scale_fill_manual(values = c(CO = '#fdbf6f',
                               Hybrid = '#a6cee3')) +
  labs(
    x        = 'Analysis method',
    y        = 'Empirical type-I error rate',
    fill     = NULL,
    title    = 'Type-I error under null, split by architecture',
    subtitle = expression(
      c[bm]==0*','~N==70*','~
      'pooled across DGP forms and carryover levels (dev)')
  ) +
  theme_paper +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

## Summary table: all cells at $N = 70$, $c_{bm} = 0.45$, Hybrid

```{r all-cells-table}
tab_all <- grid1 |>
  filter(N == 70, c_bm == 0.45, design == 'Hybrid') |>
  mutate(
    DGP   = paste0(carryover_form,
      ifelse(carryover_form == 'weibull',
             paste0(' k=', weibull_shape), '')),
    Arch  = ifelse(dgp_arch == 'mvn', 'B', 'A'),
    `t1/2` = t1half,
    `Power (MC SE)` = sprintf('%.3f (%.3f)',
                               power, mc_se_power),
    `Non-conv` = sprintf('%.3f', non_convergence_rate)
  ) |>
  arrange(Arch, DGP, `t1/2`, match(spec, method_levels)) |>
  select(Arch, DGP, `t1/2`, spec,
         `Power (MC SE)`, `Non-conv`)

knitr::kable(
  tab_all, booktabs = TRUE, longtable = TRUE,
  caption = 'Power and non-convergence at $N = 70$, $c_{bm} = 0.45$,
    Hybrid design, dev grid (150 reps per cell), for all five
    analysis methods across the DGP-architecture combinations
    present in the dev grid.'
)
```

# Tier 2 sensitivity results

## S1: within-factor autocorrelation

```{r fig-s1, fig.cap = "Power by analysis method across AR(1) autocorrelation $\\rho \\in \\{0.5, 0.7, 0.8, 0.9\\}$. Reference: Arch B, Hybrid, $N = 70$, $c_{bm} = 0.45$, exponential DGP, $t_{1/2} = 1.0$ weeks.", fig.width = 6, fig.height = 3.9}
d1 <- grid2 |>
  filter(block == 'S1') |>
  mutate(spec = as_method(spec))

ggplot(d1, aes(rho, power, colour = spec, group = spec)) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 1.5) +
  scale_y_continuous(limits = c(0, 1)) +
  spec_scale +
  labs(
    x        = expression('AR(1) autocorrelation'~rho),
    y        = 'Power',
    colour   = 'Analysis method',
    title    = 'S1: Sensitivity to within-factor autocorrelation',
    subtitle = 'Hybrid, Arch B, N=70, c_bm=0.45, exp DGP (dev, 150 reps)'
  ) +
  theme_paper
```

## S2: analyst-truth half-life mismatch

```{r fig-s2, fig.cap = "Power as the analyst's assumed half-life departs from the true DGP half-life. M1 and M3 do not consume the assumed half-life and are flat; M2 and the two robust analyses are all built on the $D_{bc}$ predictor and vary with it.", fig.width = 7, fig.height = 3.9}
d2 <- grid2 |>
  filter(block == 'S2') |>
  mutate(
    spec = as_method(spec),
    true = factor(paste0('true t_{1/2} = ', t1half, ' wk'))
  )

ggplot(d2, aes(analysis_t1half, power,
               colour = spec, group = spec)) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 1.5) +
  facet_wrap(~ true) +
  scale_x_log10(breaks = c(0.25, 0.5, 1, 2)) +
  scale_y_continuous(limits = c(0, 1)) +
  spec_scale +
  labs(
    x        = 'Analyst-assumed half-life (log scale, weeks)',
    y        = 'Power',
    colour   = 'Analysis method',
    title    = 'S2: Cost of analyst-truth half-life mismatch (dev)',
    subtitle = 'M1 and M3 are invariant; M2, lme+CR2, GEE+MD are Dbc-based'
  ) +
  theme_paper
```

## S3: dropout

```{r fig-s3, fig.cap = "Power at dropout rates $\\in \\{0, 0.1, 0.2, 0.3\\}$ under MCAR and MAR mechanisms, for all five analysis methods. With its exchangeable working correlation GEE+MD runs cleanly across the dropout range, power declining monotonically as expected.", fig.width = 7, fig.height = 3.9}
d3 <- grid2 |>
  filter(block == 'S3') |>
  mutate(spec = as_method(spec))

ggplot(d3, aes(dropout_rate, power,
               colour = spec, group = spec)) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 1.5) +
  facet_wrap(~ dropout_mech) +
  scale_y_continuous(limits = c(0, 1)) +
  spec_scale +
  labs(
    x        = 'Dropout rate',
    y        = 'Power',
    colour   = 'Analysis method',
    title    = 'S3: Sensitivity to dropout (dev)',
    subtitle = 'MCAR and MAR (severity-biased); reference cell as above'
  ) +
  theme_paper
```

## S4: biomarker-moderation effect-size curve

```{r fig-s4, fig.cap = "Power across $c_{bm} \\in \\{0, 0.10, 0.20, 0.30, 0.45, 0.60\\}$, for all five analysis methods. The $c_{bm} = 0$ point is the null; dashed line at nominal $\\alpha = 0.05$.", fig.width = 6, fig.height = 3.9}
d4 <- grid2 |>
  filter(block == 'S4') |>
  mutate(spec = as_method(spec))

ggplot(d4, aes(c_bm, power, colour = spec, group = spec)) +
  geom_hline(yintercept = 0.05, linetype = 'dashed',
             colour = 'grey50') +
  geom_line(linewidth = 0.6) +
  geom_point(size = 1.5) +
  scale_y_continuous(limits = c(0, 1)) +
  spec_scale +
  labs(
    x        = expression(c[bm]),
    y        = 'Power / type-I error',
    colour   = 'Analysis method',
    title    = 'S4: Biomarker-moderation effect-size curve (dev)',
    subtitle = 'Dashed line at nominal alpha = 0.05'
  ) +
  theme_paper
```

## S5: autocorrelation by carryover (exploratory)

```{r fig-s5, fig.cap = "Power across carryover half-lives $\\{0, 0.5, 1.0\\}$ weeks at $\\rho \\in \\{0.5, 0.8\\}$, for all five analysis methods. Facets by $\\rho$.", fig.width = 7, fig.height = 3.9}
d5 <- grid2 |>
  filter(block == 'S5') |>
  mutate(spec = as_method(spec))

ggplot(d5, aes(t1half, power, colour = spec, group = spec)) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 1.5) +
  facet_wrap(~ rho, labeller = label_bquote(rho == .(rho))) +
  scale_y_continuous(limits = c(0, 1)) +
  spec_scale +
  labs(
    x        = expression('Carryover half-life'~t['1/2']~'(weeks)'),
    y        = 'Power',
    colour   = 'Analysis method',
    title    = 'S5: Autocorrelation x carryover interaction (dev)',
    subtitle = 'Is the method ranking rho-sensitive?'
  ) +
  theme_paper
```

# Dev-run review checklist

```{r checklist}
ref <- grid1 |>
  filter(carryover_form == 'exponential', dgp_arch == 'mvn',
         design == 'Hybrid', N == 70, c_bm == 0.45,
         t1half == 1.0)
pw <- function(s) {
  v <- ref |> filter(spec == s) |> pull(power)
  if (length(v) == 0) NA_real_ else v[1]
}

null_t1 <- grid1 |> filter(c_bm == 0, N == 70)
cr2_t1 <- null_t1 |> filter(spec == 'lme+CR2') |>
  pull(power) |> mean(na.rm = TRUE)

nonconv_max <- grid1 |>
  pull(non_convergence_rate) |> max(na.rm = TRUE)

checks <- tibble::tibble(
  Check = c(
    'M2 power > M1 power at the reference cell',
    'lme+CR2 power >= M2 power (robust SE recovers power)',
    'lme+CR2 null type-I within 0.015 of nominal 0.05',
    'Max non-convergence rate below 0.20'
  ),
  Result = c(
    sprintf('M2=%.3f, M1=%.3f -> %s', pw('M2'), pw('M1'),
            ifelse(pw('M2') > pw('M1'), 'PASS', 'FAIL')),
    sprintf('lme+CR2=%.3f, M2=%.3f -> %s',
            pw('lme+CR2'), pw('M2'),
            ifelse(pw('lme+CR2') >= pw('M2'), 'PASS', 'FAIL')),
    sprintf('lme+CR2 mean type-I=%.3f -> %s', cr2_t1,
            ifelse(abs(cr2_t1 - 0.05) < 0.015, 'PASS', 'FAIL')),
    sprintf('Max=%.3f -> %s', nonconv_max,
            ifelse(nonconv_max < 0.20, 'PASS', 'FAIL'))
  )
)

knitr::kable(checks, booktabs = TRUE,
  caption = 'Automated sanity checks against the dev-run results.
    Thresholds are deliberately loose (MC SE ~0.041 at 150 reps).')
```

---

*Dev review document rendered from
`analysis/report/02-carryover-sensitivity/report-devresults.Rmd`
on `r format(Sys.time(), '%Y-%m-%d at %H:%M %Z')`.*
*Source data: `output/01-dev.rds` (`r nrow(raw1$results)` rows),
`output/04-sensitivity-dev.rds` (`r nrow(raw2$results)` rows).*
