---
title: "Unadjusted, lag-adjusted, and exposure-weighted carryover
  specifications under differential-correlation biomarker
  interactions: a factorial simulation study"
author: 'pmsimstats team'
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: true
    toc_depth: 2
    number_sections: true
    keep_tex: true
    highlight: tango
    includes:
      in_header: ../sim-preamble.tex
bibliography: references.bib
csl: statistics-in-medicine.csl
link-citations: true
linestretch: 1.2
fontsize: 12pt
geometry: "left=2.5cm,right=6cm,top=2.5cm,bottom=2.5cm"
header-includes:
  - \usepackage{amssymb}
knit: (function(input, ...) { d <- dirname(input); while (!file.exists(file.path(d, 'tools', 'stamp-render.R')) && d != dirname(d)) d <- dirname(d); source(file.path(d, 'tools', 'stamp-render.R'))$value(input) })
---

<!--
  Build pipeline for this manuscript
  ==================================

  This is the single master source for this manuscript. Its scope is
  the covariance-moderation architecture (MVN differential correlation), the exponential and
  Weibull decay forms, and all three analysis specifications
  (stored E1/E3/E2; displayed Unadjusted, Lag-adjusted,
  Exposure-weighted via spec-labels.R).
  Earlier, wider-scope drafts are retained under archive/ and are no
  longer maintained; see README.md in this directory for the lineage
  and for what each archived draft uniquely contains.

  This Rmd does NOT run any simulation code. It loads pre-computed
  summaries and figures produced by the scripts below. Re-run the
  relevant steps when the DGP, analysis specs, or parameter grid
  change. All paths are relative to the repo root.

  1. Tier 1 principal factorial (heaviest step, ~85 min on 8 cores):
       Rscript analysis/scripts/carryover-sensitivity/01-run-factorial.R
     Writes: analysis/scripts/carryover-sensitivity/output/01-factorial.rds

  2. Tier 2 sensitivity blocks S1..S5 (~4 min on 8 cores; requires
     Tier 1 not to be running concurrently - both request all cores):
       Rscript analysis/scripts/carryover-sensitivity/04-run-sensitivity-blocks.R
     Writes: analysis/scripts/carryover-sensitivity/output/04-sensitivity.rds

  3. Summarize both tiers into cell-level power/bias/convergence
     (~seconds; re-run after either step 1 or step 2 changes):
       Rscript analysis/scripts/carryover-sensitivity/02-summarise-grid.R
     Writes:
       analysis/data/02-grid-summary.rds        (from Tier 1)
       analysis/data/02-sensitivity-summary.rds (from Tier 2)

  4. Filtered Tier 1 figures (the covariance-moderation architecture, no linear decay, all
     three specifications; writes analysis/figures/02xs-*.pdf):
       Rscript analysis/scripts/carryover-sensitivity/03-render-figures-extra-slim.R

  5. Filtered Tier 2 sensitivity figures (writes
     analysis/figures/02xs-sens-S1..S5.pdf):
       Rscript analysis/scripts/carryover-sensitivity/04-sensitivity-figures-extra-slim.R

  5b. S6 cluster-robust recalibration data
      (analysis/scripts/carryover-sensitivity/output/diag-s6-cr2.rds)
      is produced by 06-cr2-recalibration.R; no re-run needed for a
      plain re-knit.

  5c. Filtered S6 Type I figure (writes
      analysis/figures/02xs-type1-cr2.pdf):
       Rscript analysis/scripts/carryover-sensitivity/07-type1-cr2-figure-extra-slim.R

  The unsuffixed figure scripts (03-render-figures.R,
  04-sensitivity-figures.R, 07-type1-cr2-figure.R) write the
  full-scope 02-*.pdf figures consumed only by the archived drafts.
  They are retained for that reason and are not part of this
  manuscript's build.

  6. Knit this Rmd (this step; ~1 min; no simulation code runs):
       Rscript -e "rmarkdown::render('analysis/report/02-carryover-sensitivity/report.Rmd')"

  Grid design and rationale (full grid, of which this manuscript is
  a slice):
    analysis/scripts/carryover-sensitivity/simulation-grid-plan.md
-->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.path = "../../figures/02-cache-",
  cache.path = "_cache/"
)
library(tibble)
library(dplyr)
library(knitr)

## Display names for the three analysis specifications. Stored values
## stay E1/E2/E3 everywhere; only the display layer maps them, so the
## archived drafts and every .rds under analysis/data/ remain readable
## without regeneration. Display order is Unadjusted, Lag-adjusted,
## Exposure-weighted: the first two share the estimand bm:Db and are
## comparable on every measure, while the third targets bm:Dbc.
source(here::here(
  'analysis/scripts/carryover-sensitivity/spec-labels.R'))
```

# Abstract {-}

**Background.** Aggregated N-of-1 trials pool repeated within-patient
on/off contrasts to validate predictive biomarkers. Under the
multivariate-normal (MVN) differential-correlation architecture
(the covariance-moderation architecture), in which the biomarker-treatment interaction is
encoded in a treatment-state-dependent correlation, a companion study
found that carryover, that is, residual drug effect that decays after
discontinuation, costs up to roughly 31% of power in relative terms.
The analyst must still decide how to represent that residual exposure
in the linear predictor. Unfortunately, it is not established which
representation is most robust to the ordinarily unknown
pharmacokinetic decay shape.

**Methods.** We compare three analysis-model specifications for the
drug-exposure regressor under the covariance-moderation architecture:
the binary on-drug indicator, which ignores carryover altogether
(Unadjusted); that indicator augmented by a lagged just-off-drug term
that estimates carryover non-parametrically (Lag-adjusted, the
classical crossover device); and an exposure-weighted continuous
indicator that commits to a parametric decay and half-life
(Exposure-weighted, the current default). We embed them in an
ADEMP-structured factorial simulation [@morris2019simulation]
crossing two data-generating process (DGP) decay forms (exponential,
Weibull) with the three specifications, under three trial designs,
two sample sizes, and three interaction strengths (216 cells, 500
replicates each), together with four single-axis sensitivity blocks
and a cluster-robust recalibration. All three specifications are
fitted to the same simulated datasets, so every contrast between them
is paired. We report each performance measure with its Monte Carlo
standard error (MCSE).

**Results.** Two findings stand out. First, the lagged-treatment term
buys nothing: Lag-adjusted and Unadjusted are statistically
indistinguishable across all 216 cells, differing by a mean of 0.003
in power against an MCSE of 0.018, and agreeing to three decimal
places on bias, mean squared error, and coverage. Because the two
share an estimand and differ by exactly one nuisance term, this is a
clean null. Second, Exposure-weighted attains the highest power only
in the two designs with a blinded-discontinuation phase (Hybrid and
OL+BDC), where its advantage is about 10 percentage points (0.860 vs
0.766 for Unadjusted at the reference cell); under the classical
crossover (CO) design it is markedly inferior (0.488 vs 0.830).
Where it leads, the lead is robust to decay-shape and half-life
mis-specification and to autocorrelation strength. The model-based
interaction test is mildly conservative for all three (mean null
rejection 0.036 to 0.039), an artifact of the working covariance that
a CR2 cluster-robust standard error removes without disturbing the
ranking.

**Conclusions.** Three points warrant emphasis. First, under the
covariance-moderation architecture we recommend the exposure-weighted
predictor, reported with a cluster-robust standard error, for the
Hybrid and OL+BDC designs; under a crossover design it should not be
used, and the unadjusted binary indicator is materially better.
Second, the classical lagged-treatment remedy cannot be recommended
in any design examined here, since it never improved on ignoring
carryover altogether; we trace this to its entering the model as a
nuisance main effect, which leaves the attenuation of the interaction
untouched. Third, across all conditions anticipated dropout governs
power far more than the choice among specifications, and should
dominate design effort accordingly. This manuscript restricts the
grid to the covariance-moderation architecture and to the exponential
and Weibull decay forms; the mean-moderation architecture and linear
decay are evaluated in the archived companion drafts
`archive/report.Rmd` and `archive/report-slim.Rmd`.

# Introduction

In a single N-of-1 trial, one patient is crossed repeatedly between
active treatment and a comparator over a sequence of measurement
occasions, and the within-patient contrast between the on-drug and
off-drug periods estimates that patient's individual treatment
effect. An aggregated N-of-1 trial pools many such single-patient
series into a mixed model, trading some of the resolution available
within a single patient for population-level inference and for
sample sizes considerably smaller than a parallel-group trial would
require [@hendrickson2020]. The estimand of interest is the
biomarker-treatment interaction, that is, whether a baseline
covariate $B_i$ modifies the treatment effect. We examine three
trial designs that differ in how the on-drug and off-drug occasions
are arranged, a classical crossover (CO), an open-label design
followed by blinded discontinuation (OL+BDC), and a Hybrid N-of-1
design; each is defined in Section 2.

The inferential complication specific to this setting is carryover.
When a drug is discontinued, its pharmacological effect does not
switch off abruptly but decays over days or weeks, so that
observations recorded during the nominally off-drug period still
carry residual exposure. Treating those observations as though they
were cleanly off-drug mis-specifies the design matrix, and the
consequences for the interaction test are worth setting out
explicitly.

The exposure regressor the analyst writes down is, in that case, a
mis-measured version of the exposure the patient actually received,
and the damage is of two kinds. First, the contrast that the
interaction term is meant to estimate is diluted. Occasions that
still carry a substantial fraction of the drug effect are entered
with the same value as occasions that carry none, so the coefficient
on the biomarker-by-exposure product is biased toward zero, in the
familiar manner of a covariate measured with error. Second, those
same occasions are heterogeneous in a way the model does not
represent, and the unexplained variation is absorbed into the
residual variance, inflating the standard error of the very
coefficient that has just been attenuated. The test statistic loses
on both counts, its numerator shrinking while its denominator grows,
and power falls accordingly.

We note that this is a loss of power and not a loss of validity.
Under the null, $c_{bm} = 0$, the covariance-moderation architecture generates no
biomarker-response association at any level of residual exposure, so
a mis-coded exposure regressor has no interaction to distort, and the
size of the test is unaffected by the choice of exposure coding. The
mild conservatism we do observe in Section 3.5 has a separate origin
in the working covariance and applies equally to both specifications.
Carryover mis-specification is therefore a question of efficiency
rather than a source of spurious findings, and it is on that basis
that we frame what follows in terms of recovered power.

But how should residual exposure be represented in the linear
predictor? Three answers are available, and the question we take up
here is which of them recovers more of the power that carryover would
otherwise cost. They correspond to the three positions an analyst can
take toward the unknown decay: deny it, and code exposure as strictly
on or off; estimate it, by adding a nuisance term whose magnitude the
data determine; or assume it, by committing to a decay curve and a
half-life.

A companion study [@pmsimstats_dgp_architectures2026], building on
Hendrickson [-@hendrickson2020], shows that under what we call
the covariance-moderation architecture, a multivariate-normal differential-correlation
formulation in which the interaction lives in the
treatment-state-dependent covariance
$\text{Cor}(B_i, BR_{it}) = c_{bm} \cdot \phi(t_{sd})$, carryover
erodes power directly, and the loss is strongly design-dependent: at
a half-life of one week, the relative loss is roughly 31% under
OL+BDC, but only 3 to 5% under the crossover and Hybrid designs.
Under an alternative architecture, in which the biomarker instead
moderates the mean response directly, the corresponding loss is only
1 to 3% in every design; that architecture, together with a third
combined architecture, falls outside the scope of the present
manuscript,
which restricts attention to the covariance-moderation architecture, where the carryover
problem is largest and where Hendrickson's original results were
obtained. Unfortunately, that earlier work fixed the decay shape at
exponential and evaluated only a single analysis specification,
leaving open the question of which representation of residual
exposure performs best when the true decay shape is not known with
any precision, as is ordinarily the case in practice.

We compare three ways of representing drug exposure in the linear
predictor. The first is the binary on-drug indicator, which ignores
carryover entirely and is what an analyst obtains by default from a
standard mixed model; we call it Unadjusted, and it serves as the
benchmark any remedy must beat. The second augments that indicator
with a lagged just-off-drug term that estimates the magnitude of
carryover without assuming any particular functional form, which we
call Lag-adjusted, following the classical crossover-trial device
described by Jones and Kenward [-@jones2014crossover]. The third is
an exposure-weighted continuous indicator that commits to a
parametric decay form and half-life, which we call Exposure-weighted
and which is the current default in the pmsimstats-ng package.

Including the unadjusted benchmark is what makes the comparison
interpretable rather than merely competitive. Unadjusted and
Lag-adjusted estimate the same coefficient and are fitted to the same
simulated datasets, differing by exactly one nuisance term, so the
contrast between them isolates the contribution of that term alone.
Exposure-weighted differs on a second axis as well, since it changes
the regressor that carries the interaction. We cross the three
specifications against exponential and Weibull data-generating decay
forms and ask whether one can be recommended outright, or whether the
choice depends on the trial design. The study follows the ADEMP reporting
framework of Morris, White, and Crowther
[-@morris2019simulation], and every performance measure we report
carries its Monte Carlo standard error. Section 2 describes the
simulation design in ADEMP order, Section 3 gives the results, and
Section 4 takes up the practical implications for trialists.

# Simulation design

## Aims

The aim of this study is to quantify, under the covariance-moderation architecture, the cost
of mis-specifying the carryover representation in the analysis
model, and to determine which of the three candidate strategies,
Unadjusted, Lag-adjusted or Exposure-weighted, performs better across
the grid of data-generating decay forms
and trial designs. A second and related aim, addressed in the Tier 2
sensitivity blocks of Section 3.6, is to check whether the resulting
ranking survives perturbation of the within-subject autocorrelation,
the dropout rate, the accuracy of the assumed half-life, and the
size of the biomarker-treatment interaction itself.

## Data-generating mechanism

Under the covariance-moderation architecture, the biomarker and the biological response are
drawn jointly from a multivariate normal distribution with a
treatment-state-dependent correlation,
$\text{Cor}(B_i, BR_{it}) = c_{bm} \cdot \phi(t_{sd})$, where $\phi$
denotes the decay form and $t_{sd}$ is the time elapsed since drug
discontinuation. The mean-moderation architecture, in which the biomarker instead
moderates the mean response directly, and a combined the dual-channel architecture,
fall outside the scope of the present manuscript; the full
comparison across architectures is reported in the archived
full-scope drafts (`archive/report.Rmd`,
`archive/report-slim.Rmd`). In both architectures,
the decay function $\phi : \mathbb{R}_+ \to [0, 1]$ maps $t_{sd}$ to
a residual-effect multiplier. We evaluate two such forms here, each
parameterized to match a common half-life $t_{1/2}$. A third form,
linear decay, appears in the full-scope manuscripts but is not
considered further here.

### Exponential decay

$$
\phi_{\text{exp}}(t_{sd}) =
  \exp\left(-\lambda t_{sd}\right), \qquad
  \lambda = \frac{\ln 2}{t_{1/2}}
$$

This is the decay form used throughout the companion manuscript, and
it is the one currently implemented in `functions.R` under
`implementations/tidyverse/`. It corresponds to
classical first-order elimination kinetics and is the default
assumption in pharmacokinetics; we retain it here as the reference
against which the Weibull form is compared.

### Weibull decay

$$
\phi_{\text{weib}}(t_{sd}) =
  \exp\left(-\left(\lambda_w t_{sd}\right)^{k}\right), \qquad
  \lambda_w = \frac{(\ln 2)^{1/k}}{t_{1/2}}
$$

A shape parameter $k$ governs whether the elimination tail is
heavier than exponential ($k < 1$, slower elimination) or lighter
($k > 1$, faster off-drug clearance); the exponential form is
recovered exactly at $k = 1$. We report results at $k = 0.7$,
representing a drug with a prolonged low-concentration tail, and at
$k = 1.5$, representing a drug cleared more rapidly once
concentration falls below some saturation threshold, perhaps through
active transport. Together with $k = 1.0$, which is numerically
indistinguishable from the exponential form, this gives four
decay-shape levels in the grid.

## Estimand and analysis specifications

Under the covariance-moderation architecture, the interaction estimand is the on-drug
biomarker-response correlation $c_{bm}$, a dimensionless quantity.
Bias and empirical standard error are reported on a calibrated
effect-size scale, namely the expected difference in on-drug
response per standard deviation of the biomarker at $t_{1/2} = 0$,
as defined in `docs/19-mean-moderation-implementation-notes.tex`;
power and Type I error are reported on the native test-statistic
scale. The null estimand used for Type I error verification is
$c_{bm} = 0$, realized by the corresponding row of the parameter
grid.

The analysis model is `nlme::lme(Sx ~ bm * X + t, random = ~1
| ptID, correlation = corCAR1(form = ~ t | ptID))` for all three
specifications considered here; they differ only in how the
drug-exposure predictor $X$ is constructed, and in whether a nuisance
term is added.

One distinction governs how the results below must be read.
Unadjusted and Lag-adjusted both test the coefficient on
$\text{bm:}D_{it}$, so they estimate the same quantity and every
performance measure is comparable between them. Exposure-weighted
tests the coefficient on $\text{bm:}D_{bc,it}$, a different quantity.
Power and Type I error are properties of the test rather than of the
coefficient tested and so remain comparable across all three, but
bias, mean squared error and coverage are not comparable across that
divide, for the reason given in Section 3.1.

### Unadjusted: binary on-drug indicator

Under Unadjusted, $X = D_{it}$, the binary on-drug indicator, which
takes the value one at on-drug timepoints and zero otherwise.
Carryover is not represented at all: an off-drug occasion still
carrying most of the pharmacological effect is coded identically to
one carrying none. This is the specification an analyst obtains
without considering carryover, and it enters the comparison as the
benchmark against which the other two must justify their additional
structure. It consumes no half-life and is therefore invariant to the
analyst's pharmacokinetic assumptions.

### Lag-adjusted: binary indicator plus estimated carryover term

Under Lag-adjusted, $X = D_{it}$ is retained and the model is
augmented with a lagged-treatment indicator
$L_{it} = D_{i,t-1}(1 - D_{it})$, which marks the off-drug timepoints
immediately following an on-drug timepoint. The magnitude of
carryover is estimated from the data rather than assumed in advance,
so this specification does not require the analyst to posit any
particular decay form, and like Unadjusted it is invariant to the
assumed half-life. The device follows the classical crossover-trial
tradition surveyed by Jones and Kenward [-@jones2014crossover].

One structural feature bears directly on how its results should be
read: the lagged term $L_{it}$ enters only as a nuisance main effect,
with no $\text{bm:}L_{it}$ product, and the interaction actually
tested remains $\text{bm:}D_{it}$. Lag-adjusted can therefore absorb
mean-level carryover, but it cannot repair the attenuation of the
biomarker interaction described in Section 1. It is best understood
as Unadjusted with a carryover nuisance control added, rather than as
a fundamentally different model of the interaction, and Section 4.1
takes up what follows from that.

### Exposure-weighted: continuous decayed predictor

Under Exposure-weighted, $X = \text{Dbc}_{it}$, a continuous exposure-decayed
predictor constructed to match the analyst's assumed decay form and
half-life; this is the current pmsimstats-ng default. At on-drug
timepoints $D_{bc,it} = 1$ regardless of the assumed half-life. At
off-drug timepoints the predictor decays as
$D_{bc,it} = \exp(-(\ln 2/\hat{t}_{1/2})\,t_{sd,it}) =
(1/2)^{t_{sd,it}/\hat{t}_{1/2}}$, losing exactly half its value for
every $\hat{t}_{1/2}$ weeks that elapse after discontinuation. When
the analyst's assumed decay form matches the true data-generating
form, the predictor is well matched; when it does not, it is
mis-specified in a way we probe directly in Tier 2 block S2. We note
that the half-life assumption enters only through the predictor
values handed to `lme`, not through the model formula, which is
identical regardless of $\hat{t}_{1/2}$. That is, Exposure-weighted
is in effect a different model for every value of the assumed
half-life, even though it is written the same way each time. It is
the only one of the three that consumes a half-life at all, and that
asymmetry, one specification indexed continuously by
$\hat{t}_{1/2}$ and two that are not members of that family, is what
the Tier 2 block S2 sensitivity contrast in Section 3.6 is designed
to probe.

A limiting-case relationship connects the first and third
specifications and is worth stating, because it makes the set a
continuum rather than three unrelated choices. As
$\hat{t}_{1/2} \rightarrow 0$ the decay rate diverges and
$D_{bc,it} \rightarrow D_{it}$, so Unadjusted is exactly the boundary
member of the Exposure-weighted family at an assumed half-life of
zero. The S2 sweep in Section 3.6 is therefore a traverse along a
path anchored at Unadjusted, and not an independent robustness check.

## Factorial grid and parameter settings

The principal comparison is a factorial of DGP decay form (four
levels: exponential; Weibull at $k \in \{0.7, 1.0, 1.5\}$) against
analysis-model specification (three levels), under
the covariance-moderation architecture only:

```{r grid-table}
grid <- tibble::tribble(
  ~`DGP decay`,        ~Unadjusted, ~`Lag-adjusted`, ~`Exposure-weighted`,
  "Exponential",       "cell 1",    "cell 2",        "cell 3 (matched)",
  "Weibull (k = 0.7)", "cell 4",    "cell 5",        "cell 6",
  "Weibull (k = 1.0)", "cell 7",    "cell 8",        "cell 9",
  "Weibull (k = 1.5)", "cell 10",   "cell 11",       "cell 12"
)
knitr::kable(
  grid,
  caption = "Principal simulation grid (the covariance-moderation architecture only). Cell 3
    is the matched reference: exponential DGP analyzed under the
    current pmsimstats-ng default. Cells in the Exposure-weighted
    column below the first row quantify the cost of decay-form
    mis-specification; the Unadjusted and Lag-adjusted columns are
    invariant to the assumed decay form by construction."
)
```

| Parameter | Values |
|---|---|
| Biomarker moderation ($c_{bm}$) | 0.0, 0.30, 0.45 |
| Carryover half-life ($t_{1/2}$) | 0.0, 0.5, 1.0 weeks |
| Weibull shape ($k$) | 0.7, 1.0, 1.5 (Weibull form only) |
| Sample size ($N$) | 35, 70 |
| Trial design | CO, N-of-1 (Hybrid), OL+BDC |
| DGP architecture | B (MVN differential correlation) only |
| Replicates per cell | 500 (target); 50 for development |

The 216-cell count follows the same nesting logic used in the
full-scope manuscript, with the architecture dimension fixed at B
and the linear decay form removed. The decay-shape combinations
number four (exponential, and Weibull at $k \in \{0.7, 1.0, 1.5\}$),
so the grid comes to $4 \times 3 \times 3 \times 2 \times 3 = 216$
cells, the factors being decay shape, carryover half-life, trial
design, sample size, and biomarker level. The $c_{bm} = 0$ row
serves as the Type I error check, targeting the nominal 5% level,
and the $t_{1/2} = 0$ row, in which carryover is absent altogether,
gives a best-case power reference at each combination of sample
size, design, and $c_{bm}$, against which the power cost of nonzero
carryover can be read off directly.

## Performance measures and computing environment

For each cell we report Power, Type I error (at $c_{bm} = 0$), Bias
and Empirical SE of the interaction estimate (on the calibrated
scale), Coverage of the nominal 95% Wald CI, MSE, the
non-convergence rate, and the MCSE of each measure, following Morris
et al. [-@morris2019simulation, Tables 5-6].

We chose $n_{sim} = 500$ so that the Monte Carlo standard error on
power at $p = 0.5$ would not exceed 0.022
($\sqrt{0.5 \cdot 0.5 / 500} \approx 0.022$). No simulation is run
for this manuscript; we instead filter the same production-scale
grid summary used in the archived full-scope drafts (540 cells,
`analysis/data/02-grid-summary.rds`,
computed with `implementations/tidyverse/` as the primary
implementation under R 4.5.3, 128 minutes on eight cores, commit SHA
`7018b59`, dated 2026-04-15) down to the 216-cell slice defined by
the covariance-moderation architecture and the two non-linear decay
forms. Within each cell, all three specifications were
fit to the same simulated dataset (common random numbers), so every
pairwise contrast among them is estimated with reduced Monte Carlo
variance relative to what independent draws would give. This matters
most for the Lag-adjusted minus Unadjusted contrast, which differs by
a single nuisance term. The Tier 2
sensitivity blocks and the S6 cluster-robust recalibration are
filtered from the same shared outputs,
`02-sensitivity-summary.rds` under `analysis/data/` and
`diag-s6-cr2.rds` under the driver output directory; their designs
were already anchored at the covariance-moderation architecture. The filtered figures used
here are produced by `03-render-figures-extra-slim.R`,
`04-sensitivity-figures-` `extra-slim.R`, and
`07-type1-cr2-figure-extra-slim.R`. The Reproducibility section
gives the full paths.

# Results

All results reported below derive from the same 500-replicate
production run used in the full-scope manuscripts, filtered to
the covariance-moderation architecture and the exponential and
Weibull decay forms. The Monte Carlo standard error on power does
not exceed 0.022 at $p = 0.5$, and differences smaller than this
should not be over-interpreted. Coverage of the nominal 95% Wald
confidence interval is reported for the reference cell, and
non-convergence rates are reported per cell in the tables that
follow.

## Headline performance table {#sec-headline}

Table \@ref(tab:headline) reports performance at the reference cell
(the covariance-moderation architecture, exponential DGP, Hybrid design, $N = 70$,
$c_{bm} = 0.45$), where non-convergence was zero throughout. A
caveat applies to the bias, mean squared error, and coverage
columns: these are computed against Exposure-weighted's calibrated
target, $\theta_{\text{true}} = -c_{bm}\sigma_{BR} = -3.6$, while
Unadjusted and Lag-adjusted estimate a different coefficient,
$\text{bm:}D_{it}$. Their apparent bias and under-coverage in this
table reflect that mismatch in estimand rather than any defect in the
estimators. Those three columns are therefore comparable *within* the
Unadjusted and Lag-adjusted pair, which share a target, but not
across to Exposure-weighted; for the three-way comparison we rely on
power and Type I error, which are properties of the test rather than
of the coefficient tested.

```{r headline, echo = FALSE}
suppressPackageStartupMessages({library(dplyr)})
hd <- readRDS(here::here(
  'analysis/data/02-grid-summary.rds'))$summary |>
  filter(carryover_form == 'exponential', dgp_arch == 'mvn',
         design == 'Hybrid', N == 70, c_bm == 0.45,
         spec %in% spec_order) |>
  mutate(
    Specification = spec_factor(spec),
    `t1/2` = t1half,
    `Power (MCSE)` = sprintf('%.3f (%.3f)', power, mc_se_power),
    `Bias (MCSE)` = sprintf('%+.3f (%.3f)', bias, mc_se_bias),
    `Empirical SE` = sprintf('%.3f', empirical_se),
    `Coverage (MCSE)` = sprintf('%.3f (%.3f)',
                                  coverage, mc_se_coverage),
    `MSE (MCSE)` = sprintf('%.3f (%.3f)', mse, mc_se_mse),
    `Non-conv` = sprintf('%.3f', non_convergence_rate)
  ) |>
  arrange(`t1/2`, Specification) |>
  select(`t1/2`, Specification, `Power (MCSE)`, `Bias (MCSE)`,
         `Empirical SE`, `Coverage (MCSE)`, `MSE (MCSE)`,
         `Non-conv`)

knitr::kable(
  hd, booktabs = TRUE,
  caption = 'Headline performance at the reference cell
    (the covariance-moderation architecture, exponential DGP, Hybrid design, $N = 70$,
    $c_{bm} = 0.45$, $n_{sim} = 500$). Power is the rejection rate of
    $H_0: \\beta_{bm:X} = 0$ at $\\alpha = 0.05$. Bias is on the
    calibrated effect-size scale,
    $\\theta_{\\text{true}} = -c_{bm} \\cdot \\sigma_{BR} = -3.6$
    (per docs/19 calibration). MCSEs are reported per Morris et al.
    (2019, Table 6).'
) |>
  kableExtra::kable_styling(latex_options = "scale_down")
```

## Matched-decay baseline

At the matched cell (exponential DGP, exposure-weighted analysis,
the covariance-moderation architecture, $N = 70$, $c_{bm} = 0.45$, Hybrid design), power
declines from $0.988 \pm 0.005$ at $t_{1/2} = 0$ to
$0.948 \pm 0.010$ at $t_{1/2} = 0.5$ to $0.860 \pm 0.016$ at
$t_{1/2} = 1.0$ weeks (point estimate plus or minus MCSE,
$n_{sim} = 500$). All 500 replicates converged at every half-life in
this cell.

We had previously read the value at $t_{1/2} = 1.0$ as agreeing with
the power of 0.82 reported by Hendrickson et al. for a comparable
cell [@hendrickson2020], and taken that as evidence the pipeline
reproduces the original result. We withdraw that comparison. On
inspection of the reference implementation, the analysis-side and
data-generating carryover parameters are set independently, and the
drivers that reproduce the original figures supply a carryover
half-life to the data-generating step only. The analysis options they
pass carry no half-life, so the fitter falls back to its default of
zero, at which the exposure-decayed predictor collapses exactly onto
the binary on-drug indicator. The reference analysis is therefore the
unadjusted specification, and the appropriate comparator for a figure
produced under it is our unadjusted power at the same cell, $0.766$,
rather than the exposure-weighted $0.860$ quoted above. The grids
also differ on the interaction strengths and half-lives examined, so
"the same cell" was in any case approximate.

We have not resolved what the published figure was computed under,
which would require the paper rather than the vendored code, and we
therefore make no claim of agreement or disagreement here. Nothing
elsewhere in this manuscript depends on that comparison: the
specifications are compared against one another within a single
pipeline under common random numbers, so the ranking does not rest on
external calibration.

## Power by analysis specification

```{r fig-power-by-spec, echo = FALSE, fig.cap = "Power versus carryover half-life under the exponential DGP, the covariance-moderation architecture, stratified by analysis specification. Unadjusted and Lag-adjusted are indistinguishable throughout, their curves overlapping in all three panels. Exposure-weighted retains a modest advantage in the Hybrid and OL+BDC designs at $t_{1/2} = 1.0$ weeks, but is markedly inferior under the crossover (CO) design.", out.width = "95%"}
knitr::include_graphics("../../figures/02xs-power-by-spec.pdf")
```

Two features of Figure \@ref(fig:fig-power-by-spec) carry the
substance of this study.

The first is that the Unadjusted and Lag-adjusted curves lie on top of
one another in every panel and at every half-life. At the reference
cell they give $0.766$ and $0.770$ at $t_{1/2} = 1.0$ weeks, a gap of
0.004 against an MCSE of 0.018. Across all 216 cells in scope the mean
difference is 0.003, only three cells differ by more than 0.02, and
mean bias (0.372 versus 0.373), mean squared error (1.876 versus
1.885) and coverage (0.938 versus 0.936) agree to three decimal
places. Since the two share an estimand, are fitted to the same
simulated datasets, and differ by exactly one nuisance term, this is
as clean a null as the design can produce: the lagged-treatment term
contributes nothing on any measure. Section 4.1 takes up why.

The second is that the three specifications give nearly identical
power at $t_{1/2} = 0$, where there is no carryover to model, and
diverge as the half-life increases, with Exposure-weighted separating
from the other two. It retains a modest but consistent advantage in
the Hybrid and OL+BDC designs, first appearing at $t_{1/2} = 0.5$
weeks and growing through $t_{1/2} = 1.0$ weeks. That advantage does
not appear under the crossover design, where Exposure-weighted is in
fact markedly inferior to both alternatives, $0.488$ against $0.830$
for Unadjusted at the matched cell. The mechanism appears
straightforward: in a crossover trial the within-subject correlation
already carries most of the signal that the decaying predictor was
designed to recover, so it has little left to gain, and its
down-weighting of the off-drug observations only works against it. We
return to this in Section 4.

## Decay-form mis-specification

```{r fig-heatmap-mismatch, echo = FALSE, fig.cap = "Power at the critical cell ($t_{1/2} = 1.0$ weeks, the covariance-moderation architecture, Hybrid design, $N = 70$, $c_{bm} = 0.45$) across DGP decay forms (rows) and analysis specifications (columns). Under the heavy-tailed Weibull DGP ($k = 0.7$), the Exposure-weighted advantage narrows because its assumed exponential decay no longer matches truth. The Unadjusted and Lag-adjusted columns are near-identical throughout. The fill scale is stretched to the observed power range, not the full [0,1] range, so that cell-to-cell differences are visible; the absolute shading should not be compared to other figures in this manuscript.", out.width = "78%"}
knitr::include_graphics("../../figures/02xs-heatmap-matched-vs-mismatched.pdf")
```

The heatmap cell at the intersection of the exponential DGP row and
the Exposure-weighted column is the matched reference, with power
$0.860 \pm 0.016$. The Unadjusted and Lag-adjusted columns are
invariant to the assumed decay form by construction, since neither
consumes one, so variation down those columns reflects only the
data-generating decay and Monte Carlo noise; the two columns are
near-identical to each other throughout, as in Figure
\@ref(fig:fig-power-by-spec). The remaining rows of the
Exposure-weighted column give the cost of assuming exponential decay
in the analysis when the true decay form is Weibull. Even under the
heaviest-tailed data-generating process we examined ($k = 0.7$), the
Exposure-weighted advantage narrows but does not reverse. We take up
the practical interpretation of this in Section 4.

## Type I error control and cluster-robust recalibration

```{r fig-type1, echo = FALSE, fig.cap = "Empirical Type I error rates under the null ($c_{bm} = 0$), the covariance-moderation architecture, pooled across DGP decay forms and carryover half-lives, stratified by design and analysis specification. Boxplots cover the distribution across cells at $N = 70$; the dashed reference line marks the nominal 0.05 level.", out.width = "80%"}
knitr::include_graphics("../../figures/02xs-type1-boxplots.pdf")
```

Across the null cells, empirical Type I error sits uniformly below
the nominal 0.05 level for all three specifications and in all three
designs, with design-level means of 0.032 to 0.041. This shortfall is
not sampling noise but a genuine
conservatism of the model-based interaction test, which the
companion calibration study [@pmsimstats_calibration2026] traces to
the corCAR1 working covariance: the estimated standard error runs
six to ten percent too large, and the realized rejection rate falls
correspondingly. Because the conservatism is uniform across
specifications, it does not disturb the comparison among them, and
the cluster-robust recalibration reported below (block S6) confirms
that a CR2 cluster-robust standard error restores the rejection rate
to nominal without changing which specification leads. This also
establishes that the null result of Section 3.3 is not an artifact of
differential test size: Unadjusted and Lag-adjusted are conservative
to the same degree, so their equal power reflects equal information
rather than an offsetting calibration difference. Table
\@ref(tab:tab-type1-cr2) makes the per-specification conservatism,
and its cluster-robust correction, explicit at the null cell.

```{r tab-type1-cr2, echo = FALSE}
suppressPackageStartupMessages(library(dplyr))
t1 <- readRDS(here::here(
  'analysis/scripts/carryover-sensitivity/output/diag-null-type1-cr2.rds'))
tab <- t1$summary |>
  filter(spec %in% spec_order) |>
  arrange(match(spec, spec_order)) |>
  transmute(
    Specification    = spec_factor(spec),
    `$\\kappa$ model` = sprintf('%.2f', mod_se_ratio),
    `Type I model`    = sprintf('%.3f', type1_mod),
    `$\\kappa$ CR2`    = sprintf('%.2f', cr2_se_ratio),
    `Type I CR2`      = sprintf('%.3f', type1_cr2))
knitr::kable(tab, booktabs = TRUE, escape = FALSE,
  caption = 'Conservatism of the model-based interaction test and its
    cluster-robust correction, at the null cell ($c_{bm} = 0$,
    exponential DGP, $t_{1/2} = 1.0$ weeks, Hybrid design, $N = 70$,
    the covariance-moderation architecture, 3{,}000 replicates). $\\kappa$ is the ratio of the
    mean standard error to the empirical standard deviation of the
    interaction estimate; $\\kappa > 1$ indicates a conservative
    test. All three specifications are conservative under the
    model-based test, most so Exposure-weighted ($\\kappa = 1.10$,
    Type I $= 0.029$); the CR2 bias-reduced cluster-robust standard
    error restores $\\kappa$ to near one and Type I to nominal (0.045
    to 0.049) for all three.')
```

```{r tab-s6, echo = FALSE}
suppressPackageStartupMessages({library(dplyr); library(tidyr)})
s6 <- readRDS(here::here(
  'analysis/scripts/carryover-sensitivity/output/diag-s6-cr2.rds'))
ord <- c('Reference (N=70)', 'Small N (N=35)',
         'High autocorr (rho=0.9)',
         'Half-life mis-spec (assume 0.25)', '30% MCAR dropout')
lab <- c('Reference, $N=70$', 'Small $N=35$',
         'High autocorr, $\\rho=0.9$', 'Half-life mis-spec',
         '30\\% MCAR dropout')
## Both gaps are computed directly from s6$summary, which carries
## per-specification power under each inference, rather than reused
## from the pre-computed s6$gaps object.
gaps <- s6$summary |>
  filter(spec %in% spec_order) |>
  select(cell, spec, power_mod, power_cr2) |>
  pivot_wider(names_from = spec,
              values_from = c(power_mod, power_cr2)) |>
  transmute(cell,
            gap_mod_ew = power_mod_E2 - power_mod_E1,
            gap_cr2_ew = power_cr2_E2 - power_cr2_E1,
            gap_mod_la = power_mod_E3 - power_mod_E1,
            gap_cr2_la = power_cr2_E3 - power_cr2_E1)
tab <- s6$summary |>
  filter(spec == 'E2') |>
  left_join(gaps, by = 'cell') |>
  mutate(o = match(cell, ord)) |>
  arrange(o) |>
  transmute(
    Cell = lab[o],
    `$\\kappa$ model` = sprintf('%.2f', mod_se_ratio),
    `$\\kappa$ CR2`   = sprintf('%.2f', cr2_se_ratio),
    `EW model`        = sprintf('%.3f', power_mod),
    `EW CR2`          = sprintf('%.3f', power_cr2),
    `EW$-$U model`    = sprintf('%+.3f', gap_mod_ew),
    `EW$-$U CR2`      = sprintf('%+.3f', gap_cr2_ew),
    `LA$-$U model`    = sprintf('%+.3f', gap_mod_la),
    `LA$-$U CR2`      = sprintf('%+.3f', gap_cr2_la),
    `CR2 df`          = sprintf('%.0f', cr2_df))
knitr::kable(tab, booktabs = TRUE, escape = FALSE,
  caption = 'Cluster-robust recalibration at the reference cell and
    four stress cells (500 replicates, $c_{bm} = 0.45$,
    the covariance-moderation architecture). EW is Exposure-weighted,
    LA is Lag-adjusted, U is Unadjusted. $\\kappa$ is the ratio of the mean standard
    error to the empirical standard deviation of the interaction
    estimate for Exposure-weighted; a value above one is a
    conservative test. The two gap pairs give each alternative
    against the unadjusted benchmark under each inference. CR2 df is
    the mean Satterthwaite degrees of freedom for the cluster-robust
    test on Exposure-weighted. Computed with
    \\texttt{clubSandwich}.') |>
  kableExtra::kable_styling(latex_options = "scale_down")
```

```{r fig-type1-cr2, echo = FALSE, fig.cap = "Empirical Type I error at the null ($c_{bm} = 0$) under model-based and CR2 cluster-robust inference, across the five S6 stress cells and the three specifications (the covariance-moderation architecture). The model-based test (open) is conservative (near 0.03); CR2 (filled) is restored to the nominal 0.05 (dashed), within the binomial Monte Carlo band (gray). 500 null replicates per cell.", out.width = "95%"}
knitr::include_graphics("../../figures/02xs-type1-cr2.pdf")
```

Figure \@ref(fig:fig-type1-cr2) shows the effect of recalibration on
test size directly. The model-based null rejection rate sits below
nominal across all five S6 cells and all three specifications, while
the CR2 test restores it to within the Monte Carlo band around 0.05.
This matters because a test that is not correctly sized cannot
fairly be preferred on power alone. In the information-rich cells,
the calibration ratio $\kappa$ runs from 1.07 to 1.26 under the
model-based test and is restored to about one under CR2, and the
correctly sized test recovers two to five points of power in the
process.

Table \@ref(tab:tab-s6) reports both alternatives against the
unadjusted benchmark. The Exposure-weighted advantage is held or
slightly widened under CR2 at the reference cell ($+0.080$ to
$+0.092$), at the small-$N$ cell ($+0.066$ to $+0.086$), and at the
half-life mis-specification cell ($+0.010$ to $+0.024$); it narrows
under high autocorrelation ($+0.052$ to $+0.032$, though it still
leads); and at 30% MCAR dropout it is small under either inference
($+0.010$). We read the dropout cell as a restatement, from a
different angle, of the S3 finding below: the advantage is driven by
information carried in the off-drug observations, and once dropout
has removed most of that information there is little left to
distinguish. The Lag-adjusted gap, by contrast, is within $\pm 0.004$
of zero at every cell under the model-based test and within
$\pm 0.010$ under CR2, which corroborates the null of Section 3.3
under a second inference procedure and at four stress cells the main
grid does not cover. This recalibration was
validated only in the covariance-moderation architecture, Hybrid, $N = 70$ cells, and it
should not be extended to the crossover design, where
Exposure-weighted does not lead in the first place.

## Sensitivity analyses

Four Tier 2 blocks perturb one axis at a time around a common
reference cell (the covariance-moderation architecture, Hybrid, $N = 70$, $c_{bm} = 0.45$,
exponential DGP, $t_{1/2} = 1.0$), holding the remaining factors
fixed. The blocks are reported separately here and are not pooled
with Tier 1 or with each other.

### S1: within-factor autocorrelation

```{r fig-sens-s1, echo = FALSE, fig.cap = "Power by analysis specification across AR(1) autocorrelation $\\rho \\in \\{0.5, 0.7, 0.8, 0.9\\}$, the covariance-moderation architecture.", out.width = "60%"}
knitr::include_graphics("../../figures/02xs-sens-S1.pdf")
```

All three specifications gain power monotonically with $\rho$. At
$\rho = 0.5$, Exposure-weighted gives $0.798 \pm 0.018$ against
$0.738 \pm 0.020$ for both Unadjusted and Lag-adjusted, which are
identical to three decimal places at this level; at $\rho = 0.9$,
Exposure-weighted reaches $0.956 \pm 0.009$ against $0.906 \pm 0.013$
and $0.910 \pm 0.013$ (point estimate plus or minus MCSE,
$n_{sim} = 500$). The Exposure-weighted advantage, roughly 0.05 to
0.08 in absolute power throughout, is preserved at every
autocorrelation level we examined, which suggests that
autocorrelation strength and the choice of specification act
additively rather than interacting. Lag-adjusted tracks Unadjusted
across the whole range, never differing by more than 0.008.

### S2: analyst-truth half-life mismatch

```{r fig-sens-s2, echo = FALSE, fig.cap = "Power as the analyst's assumed half-life departs from the true DGP half-life. Only Exposure-weighted consumes the assumed half-life; Unadjusted and Lag-adjusted are invariant to it by construction and appear as flat reference lines.", out.width = "80%"}
knitr::include_graphics("../../figures/02xs-sens-S2.pdf")
```

Exposure-weighted appears fairly robust to a mismatch between the
analyst's assumed half-life and the true data-generating half-life.
Across the eight S2 cells (DGP $t_{1/2} \in \{0.5, 1.0\}$ weeks
crossed with analyst-assumed $t_{1/2} \in \{0.25, 0.5, 1.0, 2.0\}$
weeks) it averages 0.868, ranging from 0.766 to 0.964, with per-cell
MCSE between 0.008 and 0.019. Neither of the other two consumes the
half-life parameter, so both are invariant to the analyst's
assumption by construction; they average 0.846 and 0.841 across the
same eight cells and serve here as flat reference lines. The
exposure-weighted predictor's curvature is gentle enough that
mis-specifying the assumed half-life by a factor of two does not
erode its advantage at either true half-life. It would
appear, then, that it remains a defensible choice even when the true
pharmacokinetic half-life is known only approximately at the design
stage. Read together with the limiting-case relationship noted in
Section 2.3, this sweep also shows that the penalty for shortening
the assumed half-life toward zero, which drives the specification
continuously toward Unadjusted, is modest over the range examined.

### S3: dropout

```{r fig-sens-s3, echo = FALSE, fig.cap = "Power at dropout rates $\\in \\{0, 0.1, 0.2, 0.3\\}$ under MCAR and MAR mechanisms, the covariance-moderation architecture.", out.width = "80%"}
knitr::include_graphics("../../figures/02xs-sens-S3.pdf")
```

All three specifications lose power monotonically as the dropout rate
increases. At the MCAR baseline ($N = 70$, no dropout),
Exposure-weighted gives $0.866 \pm 0.015$ while Unadjusted and
Lag-adjusted both give $0.764 \pm 0.019$; at an MCAR dropout
rate of 0.30, these collapse to $0.148 \pm 0.016$, $0.122 \pm 0.015$
and $0.124 \pm 0.015$ respectively. The Exposure-weighted advantage
narrows steadily as dropout worsens: the absolute gap over
Unadjusted runs about 0.10 at zero dropout, 0.07 at a rate of 0.10,
0.04 at 0.20, and 0.03 at 0.30 under MCAR, narrowing further under
MAR dropout at the highest rate examined. This pattern is consistent
with that advantage originating in the exposure-weighted contrast
contributed by off-drug observations; when those observations are
preferentially lost, the edge contracts. At low to
moderate dropout, that is, at a rate of 0.20 or below, the
advantage is preserved. Lag-adjusted again tracks Unadjusted at every
dropout rate and under both mechanisms, the largest discrepancy being
0.010.

### S4: biomarker-moderation effect-size curve

```{r fig-sens-s4, echo = FALSE, fig.cap = "Power across the biomarker-moderation effect-size range $c_{bm} \\in \\{0, 0.10, 0.20, 0.30, 0.45, 0.60\\}$ under the covariance-moderation architecture. The $c_{bm} = 0$ point is the null for Type I verification.", out.width = "60%"}
knitr::include_graphics("../../figures/02xs-sens-S4.pdf")
```

The power curve is sigmoidal for all three specifications, as
expected. At $c_{bm} = 0$, empirical Type I error sits below nominal
for each, at $0.026 \pm 0.007$, $0.030 \pm 0.008$ and
$0.028 \pm 0.007$ for Unadjusted, Lag-adjusted and Exposure-weighted
respectively; power then rises through the mid-range and saturates
once $c_{bm} \geq 0.45$, reaching $0.916 \pm 0.012$,
$0.916 \pm 0.012$ and $0.958 \pm 0.009$ at
$c_{bm} = 0.6$. The Exposure-weighted advantage over the unadjusted
benchmark is largest at intermediate effect
sizes, $+0.082$ absolute at $c_{bm} = 0.30$ and $+0.102$ at
$c_{bm} = 0.45$, shrinking to $+0.042$ at $c_{bm} = 0.60$ as all
three approach the ceiling. The practical implication is
that the advantage matters most in trials targeting a moderate
biomarker-treatment interaction, and matters least when the true
effect is either very small or comfortably large. Lag-adjusted and
Unadjusted coincide across the entire effect-size range, differing by
at most 0.008 at any point on the curve.

## A higher-precision check

Because a number of the Tier 1 gaps are small, we reran a focused
24-cell grid (OL+BDC design, $N \in \{35, 70\}$, true half-life
$\in \{0.5, 1.0\}$) at 600 trials per cell, with all trials
converging. Because the specifications are fit to the same trials, we
can compare them with a paired McNemar test, which is considerably
more sensitive than treating them as independent groups. Type I error
across the twelve null cells of the rerun fell in
$[0.003, 0.053]$, so nominal alpha holds at the sample sizes and
half-lives examined.

Paired McNemar tests place Exposure-weighted ahead of the unadjusted
benchmark at all four alternative cells, by $+2.3$ to $+13.7$
percentage points, with the largest gap at the highest-leverage cell
($N = 70$, true $t_{1/2} = 1.0$ weeks, $p = 6.6 \times 10^{-17}$)
and $p \leq 0.015$ at the remaining three. This is a considerably
sharper confirmation than the Tier 1 grid alone provides, and it is
the comparison this rerun is best suited to make.

One scope caveat governs how the rerun's third arm should be read. It
was produced by the package's `lme_analysis()` routine rather than by
this study's driver, and that routine's carryover option appends time
since discontinuation as a linear main effect rather than the lagged
indicator $L_{it}$ used in the Tier 1 grid. Its third arm is
therefore not the Lag-adjusted specification defined in Section 2.3,
and we do not report it as such. That alternative is a reasonable
specification in its own right, and one we have not evaluated; we
return to it in Section 4.5.

# Discussion

## The lagged-treatment term does no work

The clearest result of this study is a negative one. Adding the
classical lagged-treatment indicator to the analysis model changed
nothing measurable, anywhere in the grid. Across all 216 cells the
mean power difference against the unadjusted benchmark was 0.003
against a Monte Carlo standard error of 0.018; bias, mean squared
error and coverage agreed to three decimal places; and the equality
held in every sensitivity block, at every autocorrelation level,
every dropout rate and every effect size, and at all five
cluster-robust stress cells under both inference procedures.

The strength of this conclusion rests on the design rather than on
the size of the grid. The two specifications estimate the same
coefficient, were fitted to the same simulated datasets under common
random numbers, and differ by exactly one term. There is no
confounding from a change of estimand, from Monte Carlo variation
between arms, or from differential test size, since Section 3.5 shows
both to be conservative to the same degree. What is left is the term
itself, and it contributed nothing.

The mechanism follows from where the term sits. Carryover damages the
interaction test on two fronts, as set out in Section 1: it attenuates
the coefficient, because contaminated occasions are coded as
drug-free, and it inflates the residual variance, because those
occasions are heterogeneous in a way the model does not represent. The
lagged indicator enters only as a nuisance main effect, with no
$\text{bm:}L_{it}$ product term, so the interaction it tests remains
$\text{bm:}D_{it}$ and the attenuation is untouched. It can in
principle absorb some of the second effect, and in these designs it
evidently absorbs too little of it to matter: the indicator marks a
single occasion per randomization path and is constant thereafter, so
it cannot describe a graded decline across a washout that runs two to
four occasions.

For practitioners the implication is direct. A lagged-treatment term
of this form should not be expected to recover power lost to
carryover, and reporting one does not constitute an adjustment for
carryover in any effective sense.

## The exposure-weighted advantage, and where it fails

The exposure-weighted predictor does help, but conditionally. At its
best cell (Hybrid design, $N = 70$, $c_{bm} = 0.45$,
$t_{1/2} = 1.0$) it exceeds the unadjusted benchmark by about
10 percentage points of power (0.860 versus 0.766), a difference of
roughly five Monte Carlo standard errors and therefore not
attributable to chance, and the high-precision rerun of Section 3.7
confirms an advantage at all four of its cells. On the other hand,
under the crossover design it performs considerably worse (0.488
versus 0.830). The reason for this reversal appears straightforward:
in a crossover trial, the within-subject correlation already carries
most of the signal that the decaying predictor was designed to
recover, so it gains little from its more elaborate representation of
exposure, and its down-weighting of off-drug observations works
against it. The recommendation therefore holds only in
the two non-crossover designs studied here, and should not be
extended to a crossover trial, where the unadjusted binary indicator
is materially better.

That the one specification which addresses attenuation is also the
only one to gain power, while the one which addresses only residual
variance gains nothing, suggests that attenuation is the mechanism
that matters in this setting. We offer that as an interpretation
consistent with the results rather than as a finding, since the
design does not isolate the two channels independently.

## What the robustness checks add

Three checks probe whether the exposure-weighted advantage, where it
exists, is fragile. On decay shape (Figure
\@ref(fig:fig-heatmap-mismatch)):
even when the analyst incorrectly assumes exponential decay and the
true form is Weibull, it still does not fall below the unadjusted
benchmark, within the
range of decay shapes we examined. On the assumed half-life (Figure
\@ref(fig:fig-sens-s2)): the other two specifications consume no
half-life and are unaffected, while this one is, yet at a true
half-life of one week mis-specifying it by a factor of four moves
power from 0.832 down to 0.766 at worst, a penalty of 6.6 percentage
points that still leaves it at or above the alternatives. On
autocorrelation (Figure
\@ref(fig:fig-sens-s1)): stronger serial correlation raises all three
specifications by roughly the same amount, so the lead of five to
eight points holds regardless of correlation strength. Taken together,
these checks suggest that where the advantage exists, it is not a
fragile artifact of one particular combination of nuisance
parameters.

One caveat qualifies the Tier 1 figures throughout. Outside block S2
the analysis-side half-life is set equal to the data-generating one,
so the exposure-weighted specification is granted an oracle no real
analyst possesses. Its Tier 1 margins are therefore upper bounds, and
block S2 gives the more realistic figure.

## Dropout matters more than the method

The finding with the largest practical consequence in this study
concerns missing data, not carryover. Unfortunately, when patients
drop out (Figure \@ref(fig:fig-sens-s3)), power collapses for both
specifications together: at 30% dropout, both fall to somewhere
between 0.12 and 0.25 and become essentially indistinguishable. No
specification offers protection against this loss; each can only
organize whatever data happen to remain. Dropout tied to symptom
severity
(MAR) is somewhat worse than dropout occurring at random (MCAR),
because the sickest patients, who carry the most signal, tend to
leave first. The practical implication is direct: retaining patients
in the trial yields considerably more statistical power than any
choice among these three specifications ever could.

## Scope, related work, and limits

This manuscript deliberately narrows the full-scope grid to
the covariance-moderation architecture, to the non-linear decay forms, and to the
covariance-moderation architecture, in the interest of a self-contained
evaluation of manageable length. The archived full-scope drafts
(`archive/report.Rmd` and `archive/report-slim.Rmd`)
show that the exposure-weighted advantage documented here does not
carry over to
the mean-moderation architecture, where the biomarker moderates the mean response
directly and it is marginally dominated in every design examined.
Readers who need the cross-architecture comparison should consult
that manuscript rather than extrapolate from this one.

We make no claim of numerical agreement with the originally published
power values, for the reason set out in Section 3.2: the reference
implementation's analysis-side carryover parameter is independent of
its data-generating one, and the drivers reproducing the original
figures leave the former at its default of zero, so the reference
analysis is the unadjusted specification rather than the
exposure-weighted one. Establishing what the published figures were
computed under is outstanding work.

The crossover
tradition surveyed by Jones and Kenward [-@jones2014crossover] would
lead one to expect that a shape-free lagged term should
outperform a parametric predictor once the decay shape is
mis-specified. Our results do not support that expectation over
the range we examined, for two separate reasons. The
exposure-weighted penalty for an incorrect half-life is too small to
be overtaken (Section 3.6), and more fundamentally the lagged term
does not improve on ignoring carryover at all (Section 4.1), so there
is no advantage available for decay-shape mis-specification to erode.

Two specifications adjacent to those studied here were not evaluated,
and both bound the negative result of Section 4.1. The first is a
lagged term interacted with the biomarker,
`Sx ~ bm + t + Db + L + bm:Db + bm:L`. Because our lagged term enters
as a main effect only, our finding is that a carryover *main effect*
does not recover the attenuated signal; it does not establish that no
shape-free specification can. The second is the alternative the
reference implementation itself provides, which appends time since
discontinuation as a linear main effect rather than a binary lag
indicator (Section 3.7). That term is indexed by elapsed time rather
than by position in the visit sequence, so unlike the lagged
indicator it can describe a graded decline across the whole washout,
and it is therefore the more promising of the two. Evaluating both is
the obvious next step.

Two further limitations bound what we can conclude. First, the
numbers presented here come from a single parameter set, calibrated
to a prazosin and post-traumatic stress disorder trajectory, so the
absolute power values should be regarded as illustrative rather than
general; the ranking is the more portable part of the result, and
even the ranking is conditional on the trial design, as we have
shown above. Second, the sensitivity checks vary one factor at a
time and can therefore miss genuine interactions between factors,
for instance between high autocorrelation and heavy dropout; the one
joint check available to us, reported in the full-scope manuscript
as block S5, revealed no such interaction.

# Conclusions

1. **The lagged-treatment term does nothing.** Adding the classical
   lagged just-off-drug indicator never improved on ignoring
   carryover altogether, on any measure, in any design, at any
   half-life, and in every sensitivity block. Because the two
   specifications share an estimand, were fitted to the same
   datasets, and differ by a single term, this is a clean null rather
   than an underpowered comparison. We trace it to the term entering
   as a nuisance main effect, which leaves the attenuation of the
   interaction untouched. A lagged term of this form should not be
   reported as an adjustment for carryover.

2. **Exposure weighting helps, but conditionally.** The
   exposure-weighted predictor
   achieves the highest power, about 10 points over the unadjusted
   benchmark, only in the
   Hybrid and OL+BDC designs, and its lead survives an incorrectly
   assumed decay shape and an incorrectly assumed half-life within
   those designs. Under the crossover design it is markedly worse
   than the binary indicator (0.488 against 0.830) and should not be
   used. Where it is used, we
   recommend reporting it with a CR2 cluster-robust standard error
   (Block S6).

3. **An approximate half-life is adequate where exposure weighting is
   used.** At a true half-life of one week, an assumed value wrong by
   a factor of four costs 6.6 percentage points of power, so a rough
   pharmacokinetic estimate is sufficient in practice. Note that
   outside block S2 the analyst is granted the true half-life, so the
   Tier 1 margins are upper bounds.

4. **Dropout matters more than the choice of method.** Power falls
   sharply
   as dropout increases, roughly halving at a dropout rate of 20%,
   and this loss affects all three specifications equally. Preventing
   dropout is a more effective use of design effort than any choice
   among them.

5. **The interaction test is somewhat conservative, and the
   conservatism can be corrected.** All three specifications reject
   slightly less often than the nominal 5% under the null, and a CR2
   cluster-robust standard error restores the correct rate while
   recovering a modest amount
   of power. The exposure-weighted advantage over the benchmark is
   held or widened under CR2 at
   four of the five S6 stress cells, and is small under 30% dropout,
   consistent with the fourth conclusion above
   (Block S6).

6. **This is a narrowed slice of a larger study.** The
   mean-moderation architecture
   and linear decay fall outside
   the scope of this manuscript; the full-scope grid reported in
   the archived drafts `archive/report.Rmd` and
   `archive/report-slim.Rmd` is the authority on those comparisons,
   and should be consulted before generalizing beyond
   the covariance-moderation architecture. We also make no claim of
   numerical agreement with the originally published power values
   (Section 3.2).

# Conflict of interest

The authors declare no conflicts of interest.

# Funding

This work received no specific funding.

# Data availability statement

The data and code supporting the findings of this study are
available in the pmsimstats-ng project repository. Per-replicate
Monte Carlo outputs and ADEMP-compliant cell-level summaries are
versioned under `analysis/data/`; simulation drivers are at
`analysis/scripts/carryover-sensitivity/`. Exact paths for this
manuscript are listed in the Reproducibility section below.

# Reproducibility

This manuscript was rendered with R 4.6.0 inside the project's
zzcollab Docker container; the production simulation itself ran
under R 4.5.3, as recorded in §3.6, with the image hash recorded in
`Dockerfile` and the package set captured in `renv.lock`. Principal
packages used by the simulation pipeline are `nlme` (continuous-time
linear-mixed analysis with `corCAR1` residual correlation),
`parallel` (8-core `mclapply` for parameter-grid sweeps),
`data.table` (the `original-extended` simulation collection), `furrr`
and `purrr` (the `tidyverse` simulation collection), and `rmarkdown`
plus `bookdown` for the manuscript build. No simulation code runs
during knitting; the manuscript loads pre-computed checkpoints only.

**Random-number generator.** At the top of the replicate loop, each
simulation driver sets the master seed to 20260507 and uses
`parallel::mc.set.seed` inside `mclapply`. Per-replicate seeds are
derived from the master seed as \texttt{master + rep\_idx}.

**Data and code availability.** All paths below are relative to the
repository root. Production-grade simulation outputs are
`01-factorial.rds` (principal factorial) and `04-sensitivity.rds`
(sensitivity blocks S1 through S5), both under the driver output
directory `analysis/scripts/` `carryover-sensitivity/output/`.
Cell-level summaries are `02-grid-summary.rds` and
`02-sensitivity-summary.rds` under `analysis/data/`. Driver scripts
are `01-run-factorial.R`, `04-run-sensitivity-blocks.R`, and
`02-summarise-grid.R`; the covariance-moderation architecture filtered figures used here
are produced by `03-render-figures-extra-slim.R`,
`04-sensitivity-figures-` `extra-slim.R`, and
`07-type1-cr2-figure-extra-slim.R`. All seven live in
`analysis/scripts/` `carryover-sensitivity/`, alongside
`simulation-grid-plan.md`, which documents the grid design and
serves as the Morris ADEMP pre-registration. High-precision
prototype rerun outputs are `02-carryover-replicates.rds`
(per-replicate) and `02-carryover-summary.txt` (Morris ADEMP
cell-level summary), under `analysis/data/quick-sim/`. The
manuscript source, its bibliography (`references.bib`), and the SIM
CSL file are in `analysis/report/` `02-carryover-sensitivity/`;
superseded wider-scope drafts are retained under `archive/` in that
directory and are documented in its `README.md`.

**Specification labels.** The three specifications are recorded as
`E1`, `E2`, `E3` in the simulation output and in the drivers, and are
displayed in this manuscript as Unadjusted, Lag-adjusted and
Exposure-weighted respectively. The mapping is applied at display time
only, by the single helper
`analysis/scripts/carryover-sensitivity/spec-labels.R`, which the
setup chunk and every figure script source; the stored values are
never rewritten, so the archived drafts and all `.rds` summaries
remain readable unchanged. Output archived
before 2026-07-31 stores the specifications as `A1`, `A2`, `A3`; the
migration script and the pre-migration backups are noted in
`analysis/report/NOTATION.md`.

# References

::: {#refs}
:::

<!--
  No in-document footer chunk. The page footer (source path, render
  time, git version) is injected at render time by tools/stamp.tex
  via the knit: hook above, which wraps long paths correctly.
-->
