---
title: "Carryover representation and inference for biomarker-treatment
  interaction tests in aggregated N-of-1 trials under covariance
  moderation: a factorial simulation comparison of nine analysis
  specifications"
author: 'pmsimstats team'
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: true
    toc_depth: 2
    number_sections: true
    keep_tex: true
    highlight: tango
    includes:
      in_header: ../sim-preamble.tex
fontsize: 11pt
linestretch: 1.1
geometry: 'left=3cm,right=3cm,top=2.5cm,bottom=2.5cm'
bibliography: references.bib
csl: statistics-in-medicine.csl
link-citations: true
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{float}
  - \usepackage[mathlines]{lineno}
  - \linenumbers
knit: (function(input, ...) { d <- dirname(input); while (!file.exists(file.path(d, 'tools', 'stamp-render.R')) && d != dirname(d)) d <- dirname(d); source(file.path(d, 'tools', 'stamp-render.R'))$value(input) })
---

<!--
  Build pipeline for this manuscript
  ==================================

  This is the single master source for this manuscript. Its scope is
  the covariance-moderation architecture (MVN differential correlation), the exponential and
  Weibull decay forms, and all three analysis specifications
  (stored E1/E3/E2; displayed Unadjusted, Lag-adjusted,
  Exposure-weighted via spec-labels.R).
  Earlier, wider-scope drafts are retained under archive/ and are no
  longer maintained; see README.md in this directory for the lineage
  and for what each archived draft uniquely contains.

  This Rmd does NOT run any simulation code. It loads pre-computed
  summaries and figures produced by the scripts below. Re-run the
  relevant steps when the DGP, analysis specs, or parameter grid
  change. All paths are relative to the repo root.

  1. Tier 1 principal factorial (heaviest step, ~85 min on 8 cores):
       Rscript analysis/scripts/carryover-sensitivity/01-run-factorial.R
     Writes: analysis/scripts/carryover-sensitivity/output/01-factorial.rds

  2. Tier 2 sensitivity blocks S1..S5 (~4 min on 8 cores; requires
     Tier 1 not to be running concurrently - both request all cores):
       Rscript analysis/scripts/carryover-sensitivity/04-run-sensitivity-blocks.R
     Writes: analysis/scripts/carryover-sensitivity/output/04-sensitivity.rds

  3. Summarize both tiers into cell-level power/bias/convergence
     (~seconds; re-run after either step 1 or step 2 changes):
       Rscript analysis/scripts/carryover-sensitivity/02-summarise-grid.R
     Writes:
       analysis/data/02-grid-summary.rds        (from Tier 1)
       analysis/data/02-sensitivity-summary.rds (from Tier 2)

  4. Filtered Tier 1 figures (no linear decay, all
     three specifications; writes analysis/figures/02xs-*.pdf; the
     matched-vs-mismatched panel (Figure 5) now reads
     02-decay-shape-sensitivity.rds, step 4d, not 02-grid-summary.rds
     directly - requires step 4d to have run first):
       Rscript analysis/scripts/carryover-sensitivity/03-render-figures-extra-slim.R

  4b. Annotated Tier 1 grid heatmap (writes
      analysis/figures/02xs-heatmap-hendrickson-{a,b,c}.pdf; Panel D
      is produced but no longer referenced by this document -
      Figures 1-2 (Panels A/B) were expanded to G1-G9 and superseded
      the Panel D / boxplot Type I check, see step 4c; Panel C
      (Figure 3) now reads 02-decay-shape-sensitivity.rds, step 4d,
      not 02-grid-summary.rds - requires step 4d to have run first):
       Rscript analysis/scripts/carryover-sensitivity/12-hendrickson-heatmap.R

  4c. Tier 1 grid Panels A/B (Figures 1-2), G1-G9 extension (writes
      analysis/data/02-grid-summary-hendrickson-g9.rds; 30 cells,
      exponential DGP, Architecture B only, common random numbers
      within each cell; ~3300 sec at n_reps = 500 on 8 cores):
       Rscript analysis/scripts/carryover-sensitivity/19-run-tier1-hendrickson-g9.R --reps 500

  4d. Decay-shape sensitivity, Figures 3 and 5, at a more extreme
      Weibull range (k = 0.25, 0.5, 2.0, 4.0, extended from an
      initial k = 0.5, 2.0, itself replacing an earlier k = 0.7, 1.5;
      k = 1.0 dropped as numerically identical to Exponential and so
      not a distinct condition) than 01-run-factorial.R's shared grid
      still uses; a standalone script rather than a change to that
      shared driver, since its grid is also consumed by the archived
      full-scope drafts and other manuscripts (writes
      analysis/data/02-decay-shape-sensitivity.rds; 30 cells,
      Architecture B only, G1-G3 only, common random numbers within
      each cell; ~770 sec at n_reps = 500 on 8 cores):
       Rscript analysis/scripts/carryover-sensitivity/23-run-decay-shape-sensitivity.R --reps 500

  4e. Decay-shape sensitivity, Architecture A counterpart (Supplement
      S12 Panel C), same k values as step 4d (writes
      analysis/data/02-decay-shape-sensitivity-archA.rds; 30 cells,
      G1-G3 only, common random numbers within each cell; n_reps to
      match S12 Panels A/B's precision; ~790 sec at n_reps = 500 on
      8 cores):
       Rscript analysis/scripts/carryover-sensitivity/24-run-decay-shape-sensitivity-archA.R --reps 500

  5. Filtered Tier 2 sensitivity figures (writes
     analysis/figures/02xs-sens-S1..S5.pdf):
       Rscript analysis/scripts/carryover-sensitivity/04-sensitivity-figures-extra-slim.R

  5f. Annotated Tier 2 heatmaps for S1-S4, S7, S8, S9 (writes
      analysis/figures/02xs-heatmap-sens-{S1,S2,S3,S4,S7,S8-a,S8-b,S9}.pdf;
      requires steps 3, 5b, 5d, 5e, 5g, 5h, 5i to have run first):
       Rscript analysis/scripts/carryover-sensitivity/13-hendrickson-heatmap-tier2.R

  5g. S9 alternative-paradigm specifications under the CO design
      (analysis/scripts/carryover-sensitivity/output/14-sensitivity-s9.rds),
      2 cells (t1/2) x 3 specs, 500 reps each (~8 min):
       Rscript analysis/scripts/carryover-sensitivity/14-run-sensitivity-s9.R

  5h. S10 G1-G9 CR2 comparison at the Hybrid reference cell (writes
      analysis/scripts/carryover-sensitivity/output/15-sensitivity-s10-g.rds;
      requires 5b to have run first, for G8): 1 cell, 8 specs
      (E1,E2,E3,E7,E9,E1cr2,E3cr2,E7cr2), 500 reps (~2 min):
       Rscript analysis/scripts/carryover-sensitivity/15-run-sensitivity-s10.R

  5i. S11 Type I error for G1-G5, G9 at the null (writes
      analysis/scripts/carryover-sensitivity/output/16-sensitivity-s11.rds):
      1 cell (c_bm = 0), 6 specs (E1,E2,E3,E7,E9,E7cr2), 2000 reps
      (~7 min):
       Rscript analysis/scripts/carryover-sensitivity/16-run-sensitivity-s11.R

  5b. S6 cluster-robust recalibration data
      (analysis/scripts/carryover-sensitivity/output/diag-s6-cr2.rds)
      is produced by 06-cr2-recalibration.R; no re-run needed for a
      plain re-knit.

  5c. Filtered S6 Type I figure (writes
      analysis/figures/02xs-type1-cr2.pdf; no longer referenced by
      this document as of the Figure 1/2 G1-G9 expansion - retained
      only because 06-cr2-recalibration.R's diag-s6-cr2.rds still
      backs Table tab-s6, the power-gap table):
       Rscript analysis/scripts/carryover-sensitivity/07-type1-cr2-figure-extra-slim.R

  5j. S1-S4 G1-G9 extension (writes
      analysis/data/02-sensitivity-summary-g9.rds; 25 cells, common
      random numbers within each cell; ~13800 sec at n_reps = 500,
      well above linear scaling from the 100-rep run - not further
      diagnosed):
       Rscript analysis/scripts/carryover-sensitivity/17-run-sensitivity-s1234-g9.R --reps 500

  5k. S6 G1-G9 extension (writes
      analysis/scripts/carryover-sensitivity/output/18-sensitivity-s6-g9.rds;
      10 cells, common random numbers within each cell; ~1210 sec at
      n_reps = 500; no longer referenced by this document as of the
      Figure 1/2 G1-G9 expansion, retained for archival/re-analysis):
       Rscript analysis/scripts/carryover-sensitivity/18-run-sensitivity-s6-g9.R --reps 500

  The unsuffixed figure scripts (03-render-figures.R,
  04-sensitivity-figures.R, 07-type1-cr2-figure.R) write the
  full-scope 02-*.pdf figures consumed only by the archived drafts.
  They are retained for that reason and are not part of this
  manuscript's build.

  5d. S7 biomarker-interacted carryover terms
      (analysis/scripts/carryover-sensitivity/output/09-sensitivity-s7.rds),
      one reference cell, 500 reps (~40 sec):
       Rscript analysis/scripts/carryover-sensitivity/09-run-sensitivity-s7.R

  5e. S8 architecture sensitivity of the specification ranking
      (analysis/scripts/carryover-sensitivity/output/10-sensitivity-s8.rds),
      2 architectures x 3 half-lives x 5 specs, 500 reps each (~5.5 min):
       Rscript analysis/scripts/carryover-sensitivity/10-run-sensitivity-s8.R

  6. Knit this Rmd (this step; ~1 min; no simulation code runs):
       Rscript -e "rmarkdown::render('analysis/report/02-carryover-sensitivity/report.Rmd')"

  6b. Decay-curve illustration (writes
      analysis/figures/02xs-decay-curves.pdf; no simulation, closed-
      form only; <1 sec):
       Rscript analysis/scripts/carryover-sensitivity/20-decay-curve-figure.R

  7. Supplement (same-directory companion document, added
      2026-08-19 to keep the main text to a working length; moves
      Blocks S1, S4, S10, S11 in full and S2/S8's cell-by-cell detail
      out of the main text, keeping only a one-paragraph summary of
      each there; also carries an exploratory Section S12, an
      Architecture A counterpart to Figures 1-3, outside the main
      manuscript's Architecture B scope). Requires steps 3, 5d, 5e,
      5f/5j (for the S1/S2/S4 G1-G9 heatmaps), 5h (S10), 5i (S11), 5b
      (S8's diag-s6-cr2.rds dependency for the G8 row in Table S10),
      and 5l/5m (below, for S12) to have run first:
       Rscript -e "rmarkdown::render('analysis/report/02-carryover-sensitivity/supplement.Rmd')"

  5l. S12 Tier 1 Panels A/B, Architecture A counterpart (writes
      analysis/data/02-grid-summary-hendrickson-g9-archA.rds; 30
      cells, exponential DGP, Architecture A only, n_sim = 500;
      ~55 min on 8 cores):
       Rscript analysis/scripts/carryover-sensitivity/21-run-tier1-hendrickson-g9-archA.R --reps 500

  5m. S12 heatmaps (writes analysis/figures/
      02xs-heatmap-hendrickson-{a,b,c}-archA.pdf; Panel C reads the
      existing 02-grid-summary.rds directly, no new simulation):
       Rscript analysis/scripts/carryover-sensitivity/22-hendrickson-heatmap-archA.R

  Grid design and rationale (full grid, of which this manuscript is
  a slice):
    analysis/scripts/carryover-sensitivity/simulation-grid-plan.md
-->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.path = "../../figures/02-cache-",
  cache.path = "_cache/"
)
library(tibble)
library(dplyr)
library(knitr)

## Display names for the three analysis specifications. Stored values
## stay E1/E2/E3 everywhere; only the display layer maps them, so the
## archived drafts and every .rds under analysis/data/ remain readable
## without regeneration. Display order is Unadjusted, Lag-adjusted,
## Exposure-weighted: the first two share the estimand bm:Db and are
## comparable on every measure, while the third targets bm:Dbc.
source(here::here(
  'analysis/scripts/carryover-sensitivity/spec-labels.R'))
```

# Abstract {-}

**Background.** Aggregated N-of-1 trials pool repeated within-patient
on/off contrasts to validate predictive biomarkers. Under covariance
moderation, a multivariate-normal differential-correlation
architecture in which the biomarker-treatment interaction lives in a
treatment-state-dependent correlation, a companion study found that
carryover costs up to roughly 31% of power in relative terms. The
analyst must still decide how to represent residual exposure in the
linear predictor, and it is not established which representation is
most robust to the ordinarily unknown pharmacokinetic decay shape.

**Methods.** We compare nine analysis-model specifications for the
drug-exposure regressor, in two tiers. The Tier 1 comparison, our
primary evidence base, covers three specifications across a full
factorial grid. Unadjusted is the binary on-drug indicator, ignoring
carryover altogether. Lag-adjusted augments that indicator with a
lagged just-off-drug term estimating carryover non-parametrically,
the classical crossover device. Exposure-weighted is a continuous
indicator committing to a parametric decay and half-life, the
current package default. We embed these three in an ADEMP-structured
factorial simulation [@morris2019simulation] crossing two
data-generating decay forms (exponential, Weibull) with three trial
designs, two sample sizes, and three interaction strengths (216
cells, 500 replicates each), plus sensitivity blocks and a
cluster-robust recalibration. All three are fitted to the same
simulated datasets, so every contrast is paired, and each
performance measure is reported with its Monte Carlo standard error
(MCSE). A Tier 2 extension adds six further specifications at a
single reference cell: an AIC-selected half-life variant, a
paired-difference regression with no repeated-measures model at all,
and CR2 cluster-robust variants of four of the above. This second
tier is a more targeted comparison, not a full grid.

**Results.** Three findings stand out. First, the lagged-treatment
term buys nothing. Lag-adjusted and Unadjusted are statistically
indistinguishable across all 216 Tier 1 cells, differing by a mean of
0.003 in power against an MCSE of 0.018, and agreeing to three
decimal places on bias, mean squared error, and coverage. Because
the two share an estimand and differ by exactly one nuisance term,
this is a clean null. Second, Exposure-weighted attains the highest
Tier 1 power only in the two designs with a blinded-discontinuation
phase (Hybrid and OL+BDC), where its advantage is about 10
percentage points (0.860 vs 0.766 for Unadjusted at the reference
cell). Under the classical crossover (CO) design it is markedly
inferior (0.488 vs 0.830). Where it leads, the lead is robust to
half-life mis-specification and to autocorrelation strength, and
widens under a light-tailed decay-shape mis-specification, but
narrows to statistical insignificance under the two most
heavy-tailed Weibull shapes examined. The model-based interaction
test is mildly conservative
for all three (mean null rejection 0.036 to 0.039), an artifact of
the working covariance that a CR2 cluster-robust standard error
removes without disturbing the ranking. Third, at the Hybrid
reference cell the Tier 2 extension finds that combining the
AIC-selected half-life with a CR2 standard error outperforms every
other specification tested, including the CR2-corrected
Exposure-weighted predictor, with Type I error held at nominal
despite the post-selection risk the half-life selection step raises
in principle. This result held at only that one cell, so we report
it as a promising direction rather than a validated recommendation.

**Conclusions.** Four points warrant emphasis. First, under the
covariance-moderation architecture we recommend the exposure-weighted
predictor, reported with a cluster-robust standard error, for the
Hybrid and OL+BDC designs. Under a crossover design it should not be
used, and the unadjusted binary indicator is materially better.
Second, the classical lagged-treatment remedy cannot be recommended
in any design examined here, since it never improved on ignoring
carryover altogether. We trace this to its entering the model as a
nuisance main effect, which leaves the attenuation of the interaction
untouched. Third, across all conditions anticipated dropout governs
power far more than the choice among specifications, and should
dominate design effort accordingly. Fourth, the AIC-selected
half-life specification with a CR2 correction merits further
validation as a candidate default, having outperformed every other
specification at the one cell where all nine were compared directly,
but it has not yet been tested against the same robustness battery
the Tier 1 three have. This manuscript restricts the grid to the
covariance-moderation architecture and to the exponential and
Weibull decay forms. The mean-moderation architecture and linear
decay are evaluated in the archived companion drafts
`archive/report.Rmd` and `archive/report-slim.Rmd`.

# Introduction

In a single N-of-1 trial, one patient is crossed repeatedly between
active treatment and a comparator, and the within-patient contrast
between on-drug and off-drug periods estimates that patient's
individual treatment effect. An aggregated N-of-1 trial pools many
such series into a mixed model, trading some within-patient
resolution for population-level inference at sample sizes
considerably smaller than a parallel-group trial would require
[@hendrickson2020]. The estimand is the biomarker-treatment
interaction, whether a baseline covariate $B_i$ modifies the
treatment effect. We examine three trial designs that differ in how
on-drug and off-drug occasions are arranged, a classical crossover
(CO), an open-label design followed by blinded discontinuation
(OL+BDC), and a Hybrid N-of-1 design, each defined in Section 2.2.

The inferential complication specific to this setting is carryover.
A drug's pharmacological effect does not switch off abruptly at
discontinuation but decays over days or weeks, so nominally off-drug
observations still carry residual exposure. An exposure regressor
that ignores this is a mis-measured version of what the patient
actually received, and the damage is twofold. The interaction
contrast is diluted, because occasions still carrying a substantial
fraction of the drug effect are coded identically to occasions
carrying none, biasing the biomarker-by-exposure coefficient toward
zero as any covariate measured with error would. And those same
occasions are heterogeneous in a way the model does not represent,
so the unexplained variation inflates the standard error of the
coefficient that has just been attenuated. The test loses on both
counts, and power falls accordingly.

This is a loss of power, not of validity. Under the null, $c_{bm} =
0$, no biomarker-response association exists at any level of
residual exposure, so a mis-coded regressor has no interaction to
distort and test size is unaffected by exposure coding. The mild
conservatism observed in Section 3.6 has a separate origin in the
working covariance and applies equally across specifications.
Carryover mis-specification is therefore a question of efficiency,
and we frame the study in terms of recovered power.

Three answers are available for how to represent residual exposure
in the linear predictor, corresponding to three positions an analyst
can take toward the unknown decay. The analyst may deny it, coding
exposure strictly on or off, may estimate it, adding a nuisance term
whose magnitude the data determine, or may assume it, committing to a
decay curve and a half-life. A companion study [@pmsimstats_dgp_architectures2026],
building on Hendrickson [-@hendrickson2020], shows that under the
covariance-moderation architecture, in which the interaction lives
in the treatment-state-dependent covariance $\text{Cor}(B_i,
BR_{it}) = c_{bm} \cdot \phi(t_{sd})$, carryover erodes power
directly and design-dependently. At a half-life of one week, the
relative loss is roughly 31% under OL+BDC but only 3 to 5% under
crossover and Hybrid, against 1 to 3% under an alternative
architecture where the biomarker moderates the mean response
directly. We restrict attention here to covariance moderation, where
the carryover problem is largest and Hendrickson's original results
were obtained. That earlier work fixed the decay shape at
exponential and evaluated only one analysis specification, leaving
open which representation of residual exposure performs best when
the true decay shape is not known precisely, as is ordinarily the
case.

We compare three specifications as our primary evidence base.
Unadjusted is the binary on-drug indicator, ignoring carryover
entirely and serving as the benchmark any remedy must beat.
Lag-adjusted augments it with a lagged just-off-drug term that
estimates carryover magnitude without assuming a functional form,
following the classical crossover-trial device of Jones and Kenward
[-@jones2014crossover]. Exposure-weighted is a continuous indicator
committing to a parametric decay form and half-life, the current
pmsimstats-ng default. Unadjusted and Lag-adjusted estimate the same
coefficient and are fitted to the same simulated datasets, differing
by exactly one nuisance term, so their contrast isolates that term's
contribution alone. Exposure-weighted differs on a second axis as
well, since it changes the regressor carrying the interaction. We
cross all three against exponential and Weibull data-generating
decay forms and ask whether one can be recommended outright, or
whether the choice depends on trial design, following the ADEMP
reporting framework of Morris, White, and Crowther
[-@morris2019simulation] throughout. Six further specifications, an
AIC-selected half-life variant, a paired-difference regression, and
CR2 cluster-robust variants of four base specifications, are added later at
a single reference cell (Section 2.5) once the results from these
first three raise specific new questions the Tier 1 grid alone
cannot answer. Section 2 describes the simulation design in ADEMP
order, Section 3 gives the results, and Section 4 takes up the
practical implications.

# Simulation design

The study simulates complete aggregated N-of-1 trials and analyzes
each one under every specification. A single replicate proceeds as
follows. Patients are allocated across the randomization paths of the
chosen trial design (Section 2.2). For each patient we draw a
pre-treatment biomarker together with the three latent response
components, namely the pharmacological response, the natural-history
trend and the placebo-belief component, jointly across all eight
measurement occasions from a multivariate normal distribution whose
means follow modified Gompertz trajectories and whose covariance
carries within-component AR(1) autocorrelation, cross-component
correlations, and the treatment-state-dependent biomarker-response
correlation that encodes the interaction (Section 2.3). The observed
symptom score at each occasion is the patient's baseline severity
less the sum of the three components, so that a beneficial effect
lowers the score. Carryover enters through a decay function of time
since discontinuation, which keeps part of the pharmacological effect
present at nominally off-drug occasions. All three analysis
specifications (Section 2.4) are then fitted to that same simulated
dataset, and we record the interaction coefficient, its standard
error and its $p$-value from each. This is repeated 500 times for
every parameter combination in the grid of Section 2.7.

Fitting all specifications to a common dataset is deliberate. It
makes every contrast among them paired, so that differences are
estimated with substantially less Monte Carlo variance than
independent draws would give. Throughout, $N$ denotes the **total**
number of patients in a trial, allocated as evenly as possible across
the design's randomization paths, so a design with $P$ paths assigns
about $N/P$ patients to each. At $N = 70$ this is 35 per path under
the two-path designs and 17 or 18 per path under the four-path Hybrid
design.

## Aims

The aim is to quantify, under covariance moderation, the cost of
mis-specifying the carryover representation in the analysis model,
and to determine which of the three candidate strategies performs
better across the grid of data-generating decay forms and trial
designs. A second aim, addressed in the Tier 2 sensitivity blocks of
Section 3.7, is to check whether the resulting ranking survives
perturbation of within-subject autocorrelation, dropout rate,
half-life accuracy, and interaction size.

## Trial designs

The three designs differ in how on-drug and off-drug occasions are
arranged, and so in how much post-discontinuation information each
generates. All three schedule eight measurement occasions per
patient across randomization paths, each a fixed on-drug/off-drug
schedule, and each design carries an expectancy weight (one during
open-label periods, one half during blinded periods) scaling the
placebo-belief component.

**Crossover (CO).** A traditional two-period within-patient
crossover, blinded throughout, occasions spaced 2.5 weeks apart from
2.5 to 20 weeks. One path is active for the first four occasions
then comparator, and the other reverses that order and so never
discontinues, contributing no post-discontinuation observations.

**Open-label followed by blinded discontinuation (OL+BDC).**
Open-label titration at 4, 8, 12, and 16 weeks, then weekly follow-up
to week 20 with a blinded late switch to placebo. The two paths
discontinue after week 17 or 18, so only two or three occasions
follow.

**Hybrid.** The Hendrickson design and this program's reference.
Open-label titration at weeks 4 and 8 is followed by blinded
discontinuation weekly from week 9 to 12, then a brief two-occasion
crossover at weeks 16 and 20. The four paths vary discontinuation
timing and crossover-arm order. It is the richest of the three in
on-off transitions and the only one allowing more than one
discontinuation per patient.

One consequence bears directly on the results. The first
post-discontinuation occasion falls one week after switching under
OL+BDC and Hybrid, but 2.5 weeks under CO. At the half-lives examined
here, roughly half the pharmacological effect therefore persists
into the first off-drug occasion under the two non-crossover
designs, against less than a fifth under CO, so the designs differ
substantially in how much carryover contamination they can
accumulate at all (Section 4.2).

## Data-generating mechanism

Under covariance moderation, the biomarker and the biological
response are drawn jointly from a multivariate normal distribution
with a treatment-state-dependent correlation, $\text{Cor}(B_i,
BR_{it}) = c_{bm} \cdot \phi(t_{sd})$, where $\phi$ is the decay
form and $t_{sd}$ the time since drug discontinuation. Mean
moderation and the combined dual-channel architecture fall outside
this manuscript's scope. The full cross-architecture comparison is
in the archived full-scope drafts (`archive/report.Rmd`,
`archive/report-slim.Rmd`). The decay function $\phi :
\mathbb{R}_+ \to [0, 1]$ maps $t_{sd}$ to a residual-effect
multiplier, and we evaluate two forms here, each matched to a common
half-life $t_{1/2}$. A third, linear decay, appears in the
full-scope manuscripts but not here.

### Exponential decay

$$
\phi_{\text{exp}}(t_{sd}) =
  \exp\left(-\lambda t_{sd}\right), \qquad
  \lambda = \frac{\ln 2}{t_{1/2}}
$$

This is the decay form used throughout the companion manuscript and
implemented in `functions.R` under `implementations/tidyverse/`. It
corresponds to classical first-order elimination kinetics, the
default pharmacokinetic assumption, and serves as the reference
against which Weibull is compared.

### Weibull decay

$$
\phi_{\text{weib}}(t_{sd}) =
  \exp\left(-\left(\lambda_w t_{sd}\right)^{k}\right), \qquad
  \lambda_w = \frac{(\ln 2)^{1/k}}{t_{1/2}}
$$

A shape parameter $k$ governs whether the elimination tail is
heavier than exponential ($k < 1$, slower elimination) or lighter
($k > 1$, faster clearance), recovering the exponential form exactly
at $k = 1$. We report four values, $k \in \{0.25, 0.5, 2.0, 4.0\}$,
two on each side of exponential: $k = 0.5$ and the more extreme
$k = 0.25$ (progressively more prolonged low-concentration tails),
and $k = 2.0$ and the more extreme $k = 4.0$ (progressively faster
clearance once concentration falls below some threshold). This is a
wider sensitivity range than an earlier version of this grid used
($k = 0.7, 1.5$ only). $k = 1.0$ is not a fifth Weibull condition: it
is numerically indistinguishable from Exponential and so is not
reported separately, giving five decay-shape levels in the grid
rather than six.

```{r fig-decay-curves, echo = FALSE, fig.cap = "The carryover decay multiplier $\\phi(t_{sd})$ at the reference half-life ($t_{1/2} = 1.0$ week) for exponential decay and the four Weibull shapes. The dotted reference lines mark the defining half-life point ($\\phi = 0.5$ at $t_{sd} = t_{1/2}$), which all five curves pass through by construction; the shape parameter governs only how quickly $\\phi$ falls before that point and how slowly it decays after.", out.width = "70%"}
knitr::include_graphics("../../figures/02xs-decay-curves.pdf")
```

Figure \@ref(fig:fig-decay-curves) makes the shape difference
concrete. The $k = 0.25$ and $k = 0.5$ curves have progressively
heavier tails, and $\phi$ stays well above the exponential curve for
every $t_{sd} > t_{1/2}$, most markedly at $k = 0.25$. In practical
terms, the analysis-side predictor picks up substantially more
residual carryover during a prolonged washout than the half-life
alone would suggest, and increasingly so as $k$ falls further below
one. The $k = 2.0$ and $k = 4.0$ curves have progressively lighter
tails and clear sharply faster once concentration falls below the
half-maximum threshold. All five curves pass through the same
half-life-defining point ($\phi = 0.5$ at $t_{sd} = 1.0$), so the
five forms remain comparable at a fixed nominal half-life even
though their shapes diverge substantially on either side of it.

## Estimand and analysis specifications

The interaction estimand is the on-drug biomarker-response
correlation $c_{bm}$, a dimensionless quantity. Bias and empirical
SE are reported on a calibrated effect-size scale (the expected
difference in on-drug response per SD of the biomarker at
$t_{1/2}=0$, per `docs/19-mean-moderation-implementation-notes.tex`).
Power and Type I error are on the native test-statistic scale, with
the null estimand $c_{bm} = 0$. The analysis model is
`nlme::lme(Sx ~ bm * X + t, random = ~1 | ptID, correlation = corCAR1(form = ~ t | ptID))`
for the Tier 1 three, which differ only in how $X$ is constructed and
whether a nuisance term is added.
Unadjusted and Lag-adjusted both test the coefficient on
$\text{bm:}D_{it}$, so every performance measure is comparable
between them. Exposure-weighted tests $\text{bm:}D_{bc,it}$, a
different quantity. Power and Type I error remain comparable across
all three, since they are properties of the test, but bias, MSE, and
coverage are not comparable across that divide (Section 3.1). Nine
specifications are evaluated in total across this manuscript: the
Tier 1 three below, plus six Tier 2 additions introduced in the
'Choice of analysis specifications' section that follows and tested
in Blocks S7-S10 (Section 3.7). Internally, and in the code and data
files described in the Reproducibility section, each specification
is stored under an `E`-prefixed code (`E1`, `E2`, ...); this
manuscript's prose refers to all nine uniformly by the compact
`G1`-`G9` code defined in `spec-labels.R`, or by name, per Table
\@ref(tab:tab-spec-summary) below.

```{r tab-spec-summary, echo = FALSE}
tab_spec <- tibble::tribble(
  ~Code, ~Stored, ~Name, ~Formula, ~Target,
  'G1', 'E1', 'Unadjusted', 'Sx ~ bm + t + Db + bm:Db', 'bm:Db',
  'G2', 'E3', 'Lag-adjusted', 'Sx ~ bm + t + Db + bm:Db + L', 'bm:Db',
  'G3', 'E2', 'Exposure-weighted', 'Sx ~ bm + t + Dbc + bm:Dbc', 'bm:Dbc',
  'G4', 'E7', 'AIC-selected t1/2', 'G3 formula, t1/2 by AIC over \\{0.25,0.5,1.0,2.0\\}', 'bm:Dbc',
  'G5', 'E9', 'Paired-difference', 'diff ~ bm (patient-level OLS)', 'bm slope',
  'G6', 'E1cr2', 'Unadjusted + CR2', 'G1 formula, CR2 SE', 'bm:Db',
  'G7', 'E3cr2', 'Lag-adjusted + CR2', 'G2 formula, CR2 SE', 'bm:Db',
  'G8', 'E2cr2', 'Exposure-weighted + CR2', 'G3 formula, CR2 SE', 'bm:Dbc',
  'G9', 'E7cr2', 'AIC-selected t1/2 + CR2', 'G4 formula, CR2 SE', 'bm:Dbc'
)
knitr::kable(tab_spec, booktabs = TRUE, escape = FALSE,
  col.names = c('Code', 'Stored', 'Name', 'Formula', 'Target coefficient'),
  caption = 'All nine analysis specifications evaluated in this
    manuscript. G1-G3 are the Tier 1 headline three (Section 2.4);
    G4-G5 are Tier 2 additions replacing the retired E4/E5/E6
    (Blocks S7-S9, Section 3.7); G6-G9 are CR2 cluster-robust
    variants of G1, G2, G4, and G3 respectively (Block S10). G10
    (G5+CR2) does not apply and is omitted: G5 collapses to one row
    per patient before fitting, leaving no repeated-measures cluster
    structure for CR2 to correct.') |>
  kableExtra::kable_styling(latex_options = "scale_down", font_size = 9)
```

### Unadjusted (G1): binary on-drug indicator

Here $X = D_{it}$, the binary on-drug indicator (one on drug, zero
off). Carryover is not represented at all. An off-drug occasion
still carrying most of the pharmacological effect is therefore coded
identically to one carrying none. This is the default an analyst
obtains without considering carryover, and we treat it as the
benchmark the other specifications must justify their additional
structure against. It consumes no half-life, so it is invariant to
the analyst's pharmacokinetic assumptions.

### Lag-adjusted (G2): binary indicator plus estimated carryover term

We retain $X = D_{it}$ and augment the model with a lagged-treatment
indicator, $L_{it} = D_{i,t-1}(1 - D_{it})$, marking off-drug
timepoints that immediately follow an on-drug one. This is the
classical crossover-trial device of Jones and Kenward
[-@jones2014crossover]. Carryover magnitude is estimated from the
data rather than assumed, so, like Unadjusted, this specification is
invariant to the assumed half-life. $L_{it}$ enters only as a
nuisance main effect; there is no $\text{bm:}L_{it}$ product, so the
interaction actually tested remains $\text{bm:}D_{it}$. Lag-adjusted
can absorb mean-level carryover, but it cannot repair the
interaction's attenuation (Section 1). We think of it as Unadjusted
with a carryover nuisance control added, not as a fundamentally
different model of the interaction (Section 4.1).

### Exposure-weighted (G3): continuous decayed predictor

Here $X = \text{Dbc}_{it}$, a continuous exposure-decayed predictor
matched to the analyst's assumed decay form and half-life; this is
the current pmsimstats-ng default. At on-drug timepoints
$D_{bc,it} = 1$. Off drug it decays as
$D_{bc,it} = (1/2)^{t_{sd,it}/\hat{t}_{1/2}}$, losing half its value
every $\hat{t}_{1/2}$ weeks after discontinuation. When the assumed
decay form matches the true one, the predictor is well matched. When
it does not, it is mis-specified, and we probe the consequence
directly in Block S2. The half-life assumption enters only through
the predictor values handed to `lme`, not through the model formula,
so Exposure-weighted is in effect a different model for each assumed
half-life. It is the only one of the three original specifications
that consumes a half-life at all, and this is what the S2
sensitivity contrast (Section 3.7) is designed to probe. A limiting
relationship connects Exposure-weighted back to Unadjusted. As
$\hat{t}_{1/2} \rightarrow 0$, $D_{bc,it} \rightarrow D_{it}$, so
Unadjusted is exactly the boundary member of the Exposure-weighted
family at zero assumed half-life. The S2 sweep is therefore a
traverse anchored at Unadjusted, not an independent check.

### AIC-selected half-life (G4)

This specification uses Exposure-weighted's formula,
$X = \text{Dbc}_{it}$, but chooses $\hat{t}_{1/2}$ from the data
rather than assuming it. Writing $M(t_{1/2})$ for the `lme` fit of
Section 2.4 with $D_{bc,it}$ constructed at a given candidate
half-life, the selected half-life is

$$
\hat{t}_{1/2} = \operatorname*{arg\,min}_{t_{1/2} \,\in\, \mathcal{T}}
  \text{AIC}\big(M(t_{1/2})\big), \qquad
  \mathcal{T} = \{0.25,\ 0.5,\ 1.0,\ 2.0\}\ \text{weeks},
$$

with $\text{AIC}(M) = -2\log L(M) + 2p$, $L(M)$ the fitted model's
likelihood and $p$ its number of estimated parameters
[-@akaike1974new]. Reported power, the interaction coefficient, and
its standard error are then read from $M(\hat{t}_{1/2})$ alone, the
single winning fit, not pooled or averaged across $\mathcal{T}$. This
mirrors `characterize_carryover()` (`R/carryover_analysis.R`), the
package's own real-data half-life-selection routine. Like
Exposure-weighted, it targets $\text{bm:}D_{bc,it}$. Unlike
Exposure-weighted, it requires no half-life from the analyst at all.
The cost is a post-selection inference risk [-@berk2013valid]: $M(\hat
t_{1/2})$'s reported standard error and $p$-value treat $\hat
t_{1/2}$ as though it were fixed in advance, when the same data in
fact chose it, so the nominal inference does not formally account
for the selection step. Block S11 (Section 3.7) checks whether this
risk manifests as Type I error inflation in practice.

### Paired-difference (G5)

This is a two-stage, correlation-structure-agnostic alternative,
rather than an elaboration of the mixed-model family above. For each
patient $i$ with $n_i^{\text{on}} > 0$ on-drug and
$n_i^{\text{off}} > 0$ off-drug occasions, define the patient's own
on-drug and off-drug means,

$$
\bar{S}_{x,i}^{\text{on}} = \frac{1}{n_i^{\text{on}}}
  \sum_{t:\, D_{it}=1} S_{x,it}, \qquad
\bar{S}_{x,i}^{\text{off}} = \frac{1}{n_i^{\text{off}}}
  \sum_{t:\, D_{it}=0} S_{x,it},
$$

and their difference $\Delta_i = \bar{S}_{x,i}^{\text{on}} -
\bar{S}_{x,i}^{\text{off}}$. The second stage regresses this single
per-patient value on the (centered) biomarker across the $N'
\leq N$ patients contributing a difference,

$$
\Delta_i = \beta_0 + \beta_1\, bm_i + \varepsilon_i, \qquad
i = 1, \dots, N',
$$

by plain OLS, with the interaction test on $\beta_1$. There is no
repeated-measures model, no carryover representation, and no
correlation structure of any kind: $\Delta_i$ collapses each
patient's eight occasions to one number before $bm$ ever enters the
model. We view it as a direct biomarker-interaction extension of the
paired t-test that Senn, Julious and Araujo [-@senn2014four] found to
outperform carryover-adjusted mixed models for ordinary N-of-1
treatment-effect estimation. Patients contributing only on-drug or
only off-drug observations (for instance, CO's never-discontinuing
path) have $n_i^{\text{on}} = 0$ or $n_i^{\text{off}} = 0$, cannot
form $\Delta_i$, and are dropped, so $N' < N$ whenever such a path
exists in the design.

### CR2 cluster-robust variants (G6-G9)

Section 3.6 (Block S6) shows that the model-based interaction test is
mildly conservative, and that a CR2 bias-reduced cluster-robust
standard error restores nominal size without disturbing the ranking
among G1-G3. We use the estimator of Bell and McCaffrey
[-@bell2002bias], via `clubSandwich`, paired with Satterthwaite
degrees of freedom. That correction, `lme+CR2` in Section 2.7, was
applied only to G3 there. Block S10 (Section 3.7) extends it to G1,
G2, and G4 as well, replacing only the standard error, reference
distribution, and $p$-value on each specification's existing point
estimate: G6 (G1+CR2), G7 (G2+CR2), G9 (G4+CR2). G8 (G3+CR2) is the
same specification already introduced in Section 2.7 and Block S6,
and we carry it into the nine-way G1-G9 comparison at Block S10
without rerunning it. A tenth combination, G5+CR2, does not apply.
G5 already collapses to one row per patient before fitting, so there
is no repeated-measures cluster structure left for CR2 to correct.

## Choice of analysis specifications

Three considerations bounded which specifications were tested, rather
than surveying the wider space of models used in the N-of-1 and
crossover-trial literature.

The Tier 1 three (Unadjusted, Lag-adjusted, Exposure-weighted, above)
were chosen because each represents a distinct, established position
an analyst can take toward an unmeasured or assumed pharmacokinetic
decay, not because they exhaust that space. Unadjusted is the default
any analyst obtains without considering carryover at all, and so is
the necessary benchmark for every alternative. Lag-adjusted follows
Jones and Kenward's [-@jones2014crossover] classical crossover-trial
device, the shape-free remedy the pre-mixed-model crossover literature
has long recommended. Exposure-weighted is the pmsimstats-ng package's
own established default, so it is the specification any user of the
package is actually running today, and the manuscript's practical
stake in getting its behavior right is correspondingly higher than
for the other two.

This left a wide range of alternative model classes untested by
design: Bayesian hierarchical N-of-1 models; GEE and other marginal
models, compared systematically rather than as a robustness check on
one specification's standard error (Section 2.7); randomization and
permutation tests; functional or spline-based flexible decay curves;
state-space or Kalman-filter treatments of drug concentration; and
two-stage 'series of N-of-1 trials' meta-analytic combination. Each
is established in the N-of-1 or crossover-trial literature, and none
is evaluated here. This is a scoping decision, not an oversight: the
manuscript's aim is to compare carryover *representations* within a
single, fixed conditional mixed-model framework, matching
Hendrickson's [-@hendrickson2020] original analysis model, not to
conduct a comprehensive methods bake-off. A wider survey belongs to
future work (Section 4.6).

Blocks S7 through S9 (Section 3.7) revisit this scoping decision once,
for a specific reason. Section 4.1 shows that no carryover term
tested so far can repair the interaction's attenuation, no matter
how it enters, as long as it enters only as a nuisance main effect.
The natural next model interacts the carryover term with the
biomarker directly.
An earlier version of Block S7 tested two such specifications, `E5`
(a lagged term crossed with `bm`) and `E6` (a continuous elapsed-time
term crossed with `bm`). Both underperformed Unadjusted at every cell
examined across three blocks, two trial designs, two architectures,
and three carryover half-lives (Sections 3.6, 4.5), traced to a
structural cause: the data-generating process examined throughout
this manuscript has a single true decay parameter, so crossing a
second carryover term with the biomarker adds a free parameter with
no counterpart in the truth being estimated, and two collinear
regressors chasing one true degree of freedom lose power to a
correctly specified single-parameter model almost by construction.

We expect a further variant of the same interacted-term family to
fail on that same structural argument, so rather than test one we
replaced `E5`/`E6` with two specifications that answer different
open questions about the same underlying problem, the analyst's
unknown true decay half-life. `G4` asks whether estimating the
half-life from the data recovers the power a correctly assumed
half-life would give. It selects the half-life by AIC over a
candidate grid, reusing the logic of `characterize_carryover()`
(`R/carryover_analysis.R`), already used for real-data analysis in
this package. This addresses Block S2's question from the opposite
direction: instead of asking how much a wrong assumption costs, we
ask whether estimation avoids needing the assumption at all. `G5`
asks a more basic question raised directly by the wider literature:
is a conditional mixed model even necessary? Senn, Julious and
Araujo [-@senn2014four] compared four analysis strategies for
ordinary (no biomarker interaction) N-of-1 trials. They found that a
simple paired t-test on within-patient treatment differences
achieved near-nominal Type I error, higher power, and comparable
bias and MSE relative to a carryover-adjusted mixed model, and
recommended the mixed model only when carryover is specifically
suspected. `G5` is a direct biomarker-interaction extension of their
preferred method: a two-stage, correlation-structure-agnostic
regression of each patient's own mean on-drug-minus-off-drug
difference on the biomarker across patients, with no
repeated-measures model and no carryover representation at all.

This remains a bounded comparison. `G4` and `G5` were chosen because
each is implementable within, or trivially adjacent to, the existing
analysis pipeline (Section 2.7) and each answers a specific question
this manuscript's own results raised, not because they close the
wider literature gap identified above. The Bayesian,
GEE-as-primary-model, randomization-test, and two-stage meta-analytic
alternatives remain untested and are discussed as future work
(Section 4.6).

## Factorial grid and parameter settings

The principal comparison crosses biomarker moderation, carryover
half-life, trial design, and sample size against the three analysis
specifications, under covariance moderation and the exponential DGP.

| Parameter | Values |
|---|---|
| Biomarker moderation ($c_{bm}$) | 0.0, 0.30, 0.45 |
| Carryover half-life ($t_{1/2}$) | 0.0, 0.5, 1.0 weeks |
| Sample size ($N$) | 35, 70 |
| Trial design | CO, N-of-1 (Hybrid), OL+BDC |
| DGP architecture | B (covariance moderation) only |
| Replicates per cell | 500 (target), 50 for development |

Three specifications, three half-lives, three designs, two sample
sizes, and three biomarker levels give
$3 \times 3 \times 3 \times 2 \times 3 = 162$ cells. The $c_{bm} = 0$
row is the Type I error check against nominal 5%, and the
$t_{1/2} = 0$ row, with no carryover, gives a best-case power
reference at each combination of $N$, design, and $c_{bm}$ against
which the cost of nonzero carryover can be read off directly.

Decay-shape sensitivity (Figures 3 and 5) is evaluated separately,
at the reference cell only ($c_{bm} = 0.45$, $t_{1/2} = 1.0$ weeks),
rather than across the full grid above: exponential decay and four
Weibull shapes ($k \in \{0.25, 0.5, 2.0, 4.0\}$, Section 2.3) crossed
with the three specifications, three designs, and two sample sizes
give $5 \times 3 \times 3 \times 2 = 90$ further cells
(`23-run-decay-shape-sensitivity.R`). We restrict the decay-shape
axis to this one cell because it is the axis every other result in
this manuscript holds fixed at exponential; a full crossing of decay
shape with the grid above would multiply the simulation cost several
times over for a comparison this manuscript uses in only two
figures. $k = 1.0$ is not a fifth Weibull condition here, or anywhere
in this manuscript: it recovers the exponential form exactly
(Section 2.3) and so is not a distinct decay-shape condition.

## Inference and the cluster-robust correction {#sec-cr2}

The interaction test described so far is model-based. The standard
error of $\beta_{bm:X}$ is read from the fitted `lme` object, valid
only if the assumed residual covariance is correct. That assumption
is not innocuous here. The working covariance combines a patient
random intercept with a `corCAR1` AR(1) residual correlation, and
the companion calibration study [@pmsimstats_calibration2026] shows
this does not match the covariance the data-generating mechanism
actually induces, inflating the standard error by roughly six to ten
percent and depressing the rejection rate correspondingly. We
quantify this with the calibration ratio

$$
\kappa = \frac{\overline{\widehat{SE}}}{SD\big(\hat\beta_{bm:X}\big)}
       = \frac{\frac{1}{n_{sim}}\sum_{r=1}^{n_{sim}} \widehat{SE}_r}
              {\sqrt{\frac{1}{n_{sim}-1}\sum_{r=1}^{n_{sim}}
                \big(\hat\beta_{bm:X,r} - \bar{\hat\beta}_{bm:X}\big)^2}},
$$

the mean model-reported standard error across the $n_{sim}$
simulation replicates, divided by the empirical standard deviation of
the replicate-level point estimates $\hat\beta_{bm:X,r}$ themselves
($\kappa = 1$ is correctly calibrated, since the model's own reported
uncertainty then matches the estimator's actual sampling variability;
$\kappa > 1$ means the model overstates that uncertainty, a
conservative test). At the null cell $\kappa$ runs 1.05 to 1.10
across specifications, with empirical Type I error 0.029 to 0.039
against nominal 0.05.

The remedy is a cluster-robust (sandwich) variance estimator, which
derives the variance of $\hat\beta_{bm:X}$ from observed
between-cluster variation rather than the assumed covariance, and so
remains valid under mis-specification. Clusters are patients (eight
occasions each, unique across randomization paths). Writing
$X_i$ for patient $i$'s rows of the fixed-effects design matrix and
$\hat{e}_i = y_i - X_i\hat\beta$ for that patient's residual vector
from the fitted model ($N$ patients, indexed $i = 1, \dots, N$), the
naive (Liang-Zeger) sandwich estimator is

$$
\widehat{\text{Var}}_{\text{sandwich}}(\hat\beta) =
  \left(\sum_{i=1}^{N} X_i' \hat{V}_i^{-1} X_i\right)^{-1}
  \left(\sum_{i=1}^{N} X_i' \hat{V}_i^{-1} \hat{e}_i \hat{e}_i'
    \hat{V}_i^{-1} X_i\right)
  \left(\sum_{i=1}^{N} X_i' \hat{V}_i^{-1} X_i\right)^{-1},
$$

with $\hat{V}_i$ the working covariance from the `lme` fit. This
formula is exactly the source of the standard-error inflation
described above, reversed: it substitutes each patient's own,
directly observed $\hat{e}_i\hat{e}_i'$ for the single assumed
$\hat{V}_i$ everywhere it appears in the middle term, so a mismatched
working covariance no longer biases the variance estimate, at the
cost of needing enough clusters for that substitution to be stable.
With only $N = 35$ or $70$ clusters, the naive form in
$\widehat{\text{Var}}_{\text{sandwich}}$ is itself markedly
downward-biased, so we use the CR2 bias-reduced estimator of Bell
and McCaffrey [-@bell2002bias]. CR2 replaces each raw residual
$\hat{e}_i$ above with an adjusted residual $\hat{e}_i^{\text{CR2}} =
A_i \hat{e}_i$, where $A_i$ is a patient-specific matrix chosen so
that $\widehat{\text{Var}}_{\text{sandwich}}$, recomputed with
$\hat{e}_i^{\text{CR2}}$ in place of $\hat{e}_i$, is exactly unbiased
under the working model (the construction of $A_i$ is given in full
in Bell and McCaffrey [-@bell2002bias]; we do not reproduce it here).
We pair this with Satterthwaite degrees of freedom (both via
`clubSandwich`), replacing only the standard error, reference
distribution, and $p$-value, leaving point estimates untouched.

We apply the correction in two places. A null-cell diagnostic
($c_{bm} = 0$, exponential DGP, $t_{1/2} = 1.0$ weeks, Hybrid,
$N = 70$, 3,000 replicates) establishes whether CR2 restores nominal
size (Section 3.6). A five-cell block, the reference cell plus four
stress cells (small $N$, high autocorrelation, half-life
mis-specification, 30% MCAR dropout, 500 replicates each, labeled
S6 in this manuscript's block numbering), then asks whether
restoring size disturbs the ranking (same section). Blocks S1
through S4 are the Tier 2 sensitivity analyses of Section 3.7. Block
S5, an exploratory autocorrelation-by-carryover interaction, is
outside the present scope and reported only in the archived
full-scope drafts.

## Performance measures and computing environment

For each cell we report Power, Type I error, Bias, Empirical SE
(calibrated scale), Coverage of the nominal 95% Wald CI, MSE,
non-convergence rate, and the MCSE of each, following Morris et al.
[-@morris2019simulation, Tables 5-6]. We chose $n_{sim} = 500$ so
the MCSE on power at $p = 0.5$ would not exceed 0.022. No simulation
runs for this manuscript. We filter the same 540-cell production
grid summary used in the archived full-scope drafts down to the
216-cell slice defined by covariance moderation and the two
non-linear decay forms. All three specifications are fit to the same
simulated dataset within each cell (common random numbers), so
every pairwise contrast, especially Lag-adjusted minus Unadjusted,
carries reduced Monte Carlo variance relative to independent draws.
The Tier 2 sensitivity blocks and S6 recalibration are filtered from
the same shared production outputs. Exact file paths and versions
are given in the Reproducibility section.

# Results

All results reported below derive from the same 500-replicate
production run used in the full-scope manuscripts, filtered to
the covariance-moderation architecture and the exponential and
Weibull decay forms. The Monte Carlo standard error on power does
not exceed 0.022 at $p = 0.5$, and differences smaller than this
should not be over-interpreted. Coverage of the nominal 95% Wald
confidence interval is reported for the reference cell, and
non-convergence rates are reported per cell in the tables that
follow.

## Headline performance table {#sec-headline}

Table \@ref(tab:headline) reports performance at the reference cell
(exponential DGP, Hybrid design, $N = 70$, $c_{bm} = 0.45$), where
non-convergence was zero throughout. The bias, MSE, and coverage
columns are computed against Exposure-weighted's calibrated target,
$\theta_{\text{true}} = -c_{bm}\sigma_{BR} = -3.6$, while Unadjusted
and Lag-adjusted estimate a different coefficient,
$\text{bm:}D_{it}$. Their apparent bias and under-coverage reflect
that estimand mismatch, not a defect in the estimators, so those
columns are comparable within the Unadjusted/Lag-adjusted pair but
not across to Exposure-weighted. For the three-way comparison we
rely on power and Type I error, properties of the test rather than
the coefficient tested.

```{r headline, echo = FALSE}
suppressPackageStartupMessages({library(dplyr)})
hd <- readRDS(here::here(
  'analysis/data/02-grid-summary.rds'))$summary |>
  filter(carryover_form == 'exponential', dgp_arch == 'mvn',
         design == 'Hybrid', N == 70, c_bm == 0.45,
         spec %in% spec_order) |>
  mutate(
    Specification = spec_factor(spec),
    `t1/2` = t1half,
    `Power (MCSE)` = sprintf('%.3f (%.3f)', power, mc_se_power),
    `Bias (MCSE)` = sprintf('%+.3f (%.3f)', bias, mc_se_bias),
    `Empirical SE` = sprintf('%.3f', empirical_se),
    `Coverage (MCSE)` = sprintf('%.3f (%.3f)',
                                  coverage, mc_se_coverage),
    `MSE (MCSE)` = sprintf('%.3f (%.3f)', mse, mc_se_mse),
    `Non-conv` = sprintf('%.3f', non_convergence_rate)
  ) |>
  arrange(`t1/2`, Specification) |>
  select(`t1/2`, Specification, `Power (MCSE)`, `Bias (MCSE)`,
         `Empirical SE`, `Coverage (MCSE)`, `MSE (MCSE)`,
         `Non-conv`)

knitr::kable(
  hd, booktabs = TRUE,
  caption = 'Headline performance at the reference cell
    (exponential DGP, Hybrid design, $N = 70$,
    $c_{bm} = 0.45$, $n_{sim} = 500$). Power is the rejection rate of
    $H_0: \\beta_{bm:X} = 0$ at $\\alpha = 0.05$, the primary outcome
    of this manuscript. Bias is mean(estimate $-$
    $\\theta_{\\text{true}}$) on the calibrated effect-size scale,
    $\\theta_{\\text{true}} = -c_{bm} \\cdot \\sigma_{BR} = -3.6$
    (per docs/19 calibration; see the estimand-mismatch caveat above
    for Unadjusted/Lag-adjusted). Empirical SE is the standard
    deviation of the point estimates across replicates, the actual
    sampling variability of the estimator, as distinct from
    whatever SE any single fit reports (the comparison the
    calibration ratio $\\kappa$ of Section 2.7 is built on). Coverage
    is the proportion of replicates whose nominal 95\\% Wald CI
    contains $\\theta_{\\text{true}}$; the low Unadjusted/Lag-adjusted
    values reflect the same estimand mismatch as Bias, not
    miscalibrated inference. MSE is bias$^2$ + Empirical SE$^2$, an
    overall accuracy summary. Non-conv is the fraction of replicates
    where `nlme::lme()` failed to converge (zero throughout at this
    cell). MCSEs are reported per Morris et al. (2019, Table 6).'
) |>
  kableExtra::kable_styling(latex_options = "scale_down")
```

## Tier 1 grid overview

Figures \@ref(fig:fig-hendrickson-a) through
\@ref(fig:fig-hendrickson-c) give a full-grid view of the Tier 1
results as annotated heatmaps, with trial design as column facets,
$N$ as row-block facets, analysis specification as the row axis
within each block, and a blue (low power) to red (high power) fill
scale. Each panel fixes two of the grid's remaining dimensions and
varies a third, since no single two-dimensional heatmap can display
all four simultaneously. Panels A and B report all nine G1-G9
specifications (a dedicated G1-G9 simulation restricted to the
exponential DGP and Architecture B, $n_{sim} = 500$);
Panel C retains the original three (G1-G3), since it is the only
panel that varies DGP decay form, an axis the G1-G9 extension did
not re-run.

```{r fig-hendrickson-a, echo = FALSE, fig.cap = "Tier 1 grid, biomarker-effect axis, all nine G1-G9 specifications: power by trial design (columns), $N$ (row blocks), and analysis specification (rows within block), across $c_{bm} \\in \\{0, 0.30, 0.45\\}$ at $t_{1/2} = 1.0$ weeks, exponential DGP, Architecture B ($n_{sim} = 500$).", out.width = "95%"}
knitr::include_graphics("../../figures/02xs-heatmap-hendrickson-a.pdf")
```

Two features are visible at a glance in Panel A that the headline
table makes only cell by cell. First, within every design by $N$
block, `G1` and `G2` (Unadjusted, Lag-adjusted) track each other
closely at every biomarker level, differing by at most 0.03, within
Monte Carlo noise at $n_{sim} = 500$. This matches
the higher-precision result of Section 3.4, where the two are
numerically identical to two decimal places. Second, the
Exposure-weighted family (`G3`, `G4`, and their CR2 variants `G8`,
`G9`) visibly separates from `G1`/`G2`/`G6`/`G7` only in the Hybrid
and OL+BDC columns. In the CO column the whole family underperforms
at every biomarker level past zero. At the reference cell
($c_{bm} = 0.45$, $t_{1/2} = 1.0$, $N = 70$), for example, `G4` reaches
0.52 against 0.85 to 0.87 for `G1`/`G2`/`G6`/`G7`. This is the
design-dependence of Section 3.4, read directly off the grid now
across all nine specifications rather than three. Whether that
separation depends on carryover being present at all, Panel A alone
cannot show, since it fixes $t_{1/2} = 1.0$ throughout.

```{r fig-hendrickson-b, echo = FALSE, fig.cap = "Tier 1 grid, carryover-axis, all nine G1-G9 specifications: power by trial design (columns), $N$ (row blocks), and analysis specification (rows within block), across $t_{1/2} \\in \\{0, 0.5, 1.0\\}$ weeks at $c_{bm} = 0.45$, exponential DGP, Architecture B ($n_{sim} = 500$).", out.width = "95%"}
knitr::include_graphics("../../figures/02xs-heatmap-hendrickson-b.pdf")
```

Panel B answers that, for Hybrid and OL+BDC. At $t_{1/2} = 0$ the
Exposure-weighted family's separation from `G1`/`G2`/`G6`/`G7`
disappears in both designs (all nine specifications sit at
0.94-1.00), and only opens up as the half-life increases, confirming
the separation in Panel A is specifically a carryover effect there
(the $t_{1/2} = 1.0$ column is shared with Panel A's reference cell).
Under CO the picture is different: `G4` and `G9` are already
collapsed at $t_{1/2} = 0$ (0.54 and 0.53, against 0.89-0.90 for the
rest), before carryover enters the picture at all. This is the AIC
mis-selection pathology of Block S9 read directly off the grid, not
a carryover effect: `G4`'s half-life selection drifts to the largest
candidate under CO's sparse off-drug sampling regardless of the true
half-life, so its CO weakness predates $t_{1/2} > 0$ rather than
growing from it. Panel B also adds a
ranking result Panel A alone does not show. `G9` and `G8` lead, in a
near-tie, at the Hybrid and OL+BDC reference cells (both around 0.88
at Hybrid, 0.65 at OL+BDC). `G5`
(Paired-difference) is the strongest specification at CO that is not
`G1`/`G2` or a CR2 variant of them (0.73, versus 0.50 to 0.52 for
the Exposure-weighted family there). This is consistent with `G5`
discarding the carryover representation that costs the
Exposure-weighted family power under CO.

```{r fig-hendrickson-c, echo = FALSE, fig.cap = "Tier 1 grid, decay-shape axis: power by trial design (columns), $N$ (row blocks), and analysis specification (rows within block), across the five DGP decay-shape levels (exponential, Weibull $k \\in \\{0.25, 0.5, 2.0, 4.0\\}$) at $c_{bm} = 0.45$, $t_{1/2} = 1.0$ weeks.", out.width = "95%"}
knitr::include_graphics("../../figures/02xs-heatmap-hendrickson-c.pdf")
```

Panel C shows the Exposure-weighted row's advantage in the Hybrid
and OL+BDC columns is not uniform across decay shape. It widens
substantially on the light-tailed side ($k = 2.0$, $4.0$), reaching
$+0.14$ to $+0.21$ absolute over Unadjusted/Lag-adjusted at OL+BDC.
On the heavy-tailed side it narrows and, at the two most extreme
shapes examined ($k = 0.25$, $0.5$), disappears: the three
specifications are statistically indistinguishable at OL+BDC (gaps
of $0.00$ to $0.02$, well under one MCSE), and only barely separated
at Hybrid ($+0.01$ to $+0.04$). This is the decay-form sensitivity
of Section 3.5, shown here across the full grid rather than at one
cell.

## Matched-decay baseline

At the matched cell (exponential DGP, exposure-weighted analysis,
$N = 70$, $c_{bm} = 0.45$, Hybrid design), power
declines from $0.988 \pm 0.005$ at $t_{1/2} = 0$ to
$0.948 \pm 0.010$ at $t_{1/2} = 0.5$ to $0.860 \pm 0.016$ at
$t_{1/2} = 1.0$ weeks (point estimate plus or minus MCSE,
$n_{sim} = 500$). All 500 replicates converged at every half-life in
this cell.

We had previously read the value at $t_{1/2} = 1.0$ as agreeing with
the power of 0.82 reported by Hendrickson et al. for a comparable
cell [@hendrickson2020]. Unfortunately, we must withdraw that
comparison. In the reference implementation, the analysis-side and
data-generating carryover parameters are set independently. The
drivers that reproduce the original figures supply a half-life to
the data-generating step only, leaving the analysis-side half-life
at its default of zero. At a zero half-life, the exposure-decayed
predictor collapses onto the binary on-drug indicator. The reference
analysis is therefore Unadjusted, not Exposure-weighted, so the
appropriate comparator is our unadjusted power at the same cell,
$0.766$. The
grids differ further on interaction strength and half-life, so 'the
same cell' was approximate in any case. We have not resolved
what the published figure was computed under, and make no claim of
agreement or disagreement here. Nothing else in this manuscript
depends on that comparison, since the analysis specifications are
compared against one another within a single simulation under common
random numbers.

## Power by analysis specification

```{r fig-power-by-spec, echo = FALSE, fig.cap = "Power versus carryover half-life under the exponential DGP, by analysis specification and trial design.", out.width = "95%"}
knitr::include_graphics("../../figures/02xs-power-by-spec.pdf")
```

Two features of Figure \@ref(fig:fig-power-by-spec) carry the
substance of this study. First, the Unadjusted and Lag-adjusted
curves lie on top of one another in every panel and at every
half-life. At the reference cell they give $0.766$ and $0.770$ at
$t_{1/2} = 1.0$ weeks, and across all 216 cells the mean difference
is 0.003, with mean bias, MSE, and coverage agreeing to three
decimal places. Since the two share an estimand, are fitted to the
same simulated datasets, and differ by exactly one nuisance term,
this is as clean a null as the design can produce. The
lagged-treatment term contributes nothing on any measure (Section
4.1). Second, the three specifications give nearly identical power
at $t_{1/2} = 0$, where there is no carryover to model. They diverge
as the half-life increases, with Exposure-weighted separating from
the other two. It retains a modest but consistent advantage in the
Hybrid and OL+BDC designs from $t_{1/2} = 0.5$ weeks onward. Under
the crossover design, though, it is markedly inferior to both
alternatives ($0.488$ against $0.830$ for Unadjusted). We think this
happens because the within-subject correlation in a crossover trial
already carries most of the signal the decaying predictor targets.
It gains little there, while its down-weighting of off-drug
observations works against it (Section 4).

## Decay-form mis-specification

```{r fig-heatmap-mismatch, echo = FALSE, fig.cap = "Power at the critical cell ($t_{1/2} = 1.0$ weeks, Hybrid, $N = 70$, $c_{bm} = 0.45$) across five DGP decay forms (rows) and specifications (columns). Fill scale is stretched to the observed range, not [0,1], and not comparable to other figures.", out.width = "85%"}
knitr::include_graphics("../../figures/02xs-heatmap-matched-vs-mismatched.pdf")
```

The heatmap cell at the intersection of the exponential DGP row and
the Exposure-weighted column is the matched reference, with power
$0.850 \pm 0.016$ in this figure's own simulation (a dedicated
decay-shape run, Section 2.6; the $0.860$ reported for the same
nominal cell in Table \@ref(tab:headline) and Figure
\@ref(fig:fig-power-by-spec) comes from an independent draw of the
same DGP, and the two agree within one MCSE). The Unadjusted and
Lag-adjusted columns, invariant to the assumed decay form by
construction, are near-identical throughout, as in Figure
\@ref(fig:fig-power-by-spec), so variation down those columns is
data-generating decay and Monte Carlo noise. The remaining
Exposure-weighted row gives the cost of assuming exponential decay
when the truth is Weibull, and it is not symmetric. On the
light-tailed side the advantage widens: at $k = 2.0$ it reaches
$0.880$ against $0.788$ for both alternatives ($+0.092$), and at
$k = 4.0$, $0.930$ against $0.786$/$0.788$ ($+0.142$ to $+0.144$).
On the heavy-tailed side it narrows and, at the two most extreme
shapes examined, is no longer distinguishable from noise: at
$k = 0.5$, power $0.708$ against $0.676$ and $0.672$ ($+0.032$ to
$+0.036$, roughly 1.5 MCSE), and at $k = 0.25$, $0.568$ against
$0.558$ and $0.554$ ($+0.010$ to $+0.014$, well under one MCSE). We
read this as the advantage eroding to nothing under a sufficiently
heavy-tailed mis-specification, not reversing (Section 4).

## Type I error control and cluster-robust recalibration

Empirical Type I error at the $c_{bm} = 0$ cells of Figure
\@ref(fig:fig-hendrickson-a) sits uniformly below the nominal 0.05
level for every one of the nine specifications, across every design
and $N$ examined there. This is not sampling noise but the
conservatism of the model-based test (Section 2.7), from a working
covariance that does not match the data-generating one. Because the
conservatism is uniform across specifications, it does not disturb
the comparison among them, and the cluster-robust recalibration
below confirms that a CR2 standard error restores nominal rejection
without changing which specification leads. This also establishes
that the null result of Section 3.4 is not an artifact of
differential test size, since Unadjusted and Lag-adjusted are
conservative to the same degree. Table \@ref(tab:tab-type1-cr2)
makes this explicit at the null cell, at a higher replicate count
(3,000) than the Tier 1 grid affords.

```{r tab-type1-cr2, echo = FALSE}
suppressPackageStartupMessages(library(dplyr))
t1 <- readRDS(here::here(
  'analysis/scripts/carryover-sensitivity/output/diag-null-type1-cr2.rds'))
tab <- t1$summary |>
  filter(spec %in% spec_order) |>
  arrange(match(spec, spec_order)) |>
  transmute(
    Specification    = spec_factor(spec),
    `$\\kappa$ model` = sprintf('%.2f', mod_se_ratio),
    `Type I model`    = sprintf('%.3f', type1_mod),
    `$\\kappa$ CR2`    = sprintf('%.2f', cr2_se_ratio),
    `Type I CR2`      = sprintf('%.3f', type1_cr2))
knitr::kable(tab, booktabs = TRUE, escape = FALSE,
  caption = 'Conservatism of the model-based interaction test and its
    cluster-robust correction, at the null cell ($c_{bm} = 0$,
    exponential DGP, $t_{1/2} = 1.0$ weeks, Hybrid design, $N = 70$,
    3{,}000 replicates).
    $\\kappa$ is the calibration ratio defined in Section 2.7.
    $\\kappa > 1$ indicates a conservative
    test. All three specifications are conservative under the
    model-based test, most so Exposure-weighted ($\\kappa = 1.10$,
    Type I $= 0.029$). The CR2 bias-reduced cluster-robust standard
    error restores $\\kappa$ to near one and Type I to nominal (0.045
    to 0.049) for all three.')
```

```{r tab-s6, echo = FALSE}
suppressPackageStartupMessages({library(dplyr); library(tidyr)})
s6 <- readRDS(here::here(
  'analysis/scripts/carryover-sensitivity/output/diag-s6-cr2.rds'))
ord <- c('Reference (N=70)', 'Small N (N=35)',
         'High autocorr (rho=0.9)',
         'Half-life mis-spec (assume 0.25)', '30% MCAR dropout')
lab <- c('Reference, $N=70$', 'Small $N=35$',
         'High autocorr, $\\rho=0.9$', 'Half-life mis-spec',
         '30\\% MCAR dropout')
## Both gaps are computed directly from s6$summary, which carries
## per-specification power under each inference, rather than reused
## from the pre-computed s6$gaps object.
gaps <- s6$summary |>
  filter(spec %in% spec_order) |>
  select(cell, spec, power_mod, power_cr2) |>
  pivot_wider(names_from = spec,
              values_from = c(power_mod, power_cr2)) |>
  transmute(cell,
            gap_mod_ew = power_mod_E2 - power_mod_E1,
            gap_cr2_ew = power_cr2_E2 - power_cr2_E1,
            gap_mod_la = power_mod_E3 - power_mod_E1,
            gap_cr2_la = power_cr2_E3 - power_cr2_E1)
tab <- s6$summary |>
  filter(spec == 'E2') |>
  left_join(gaps, by = 'cell') |>
  mutate(o = match(cell, ord)) |>
  arrange(o) |>
  transmute(
    Cell = lab[o],
    `$\\kappa$ model` = sprintf('%.2f', mod_se_ratio),
    `$\\kappa$ CR2`   = sprintf('%.2f', cr2_se_ratio),
    `EW model`        = sprintf('%.3f', power_mod),
    `EW CR2`          = sprintf('%.3f', power_cr2),
    `EW$-$U model`    = sprintf('%+.3f', gap_mod_ew),
    `EW$-$U CR2`      = sprintf('%+.3f', gap_cr2_ew),
    `LA$-$U model`    = sprintf('%+.3f', gap_mod_la),
    `LA$-$U CR2`      = sprintf('%+.3f', gap_cr2_la),
    `CR2 df`          = sprintf('%.0f', cr2_df))
knitr::kable(tab, booktabs = TRUE, escape = FALSE,
  caption = 'Cluster-robust recalibration at the reference cell and
    four stress cells (500 replicates, $c_{bm} = 0.45$). EW is
    Exposure-weighted,
    LA is Lag-adjusted, U is Unadjusted. $\\kappa$ is the calibration
    ratio of Section 2.7, here for Exposure-weighted. A value above
    one is a
    conservative test. The two gap pairs give each alternative
    against the unadjusted benchmark under each inference. CR2 df is
    the mean Satterthwaite degrees of freedom for the cluster-robust
    test on Exposure-weighted. Computed with
    \\texttt{clubSandwich}.') |>
  kableExtra::kable_styling(latex_options = "scale_down")
```

Table \@ref(tab:tab-s6) confirms this directly. The model-based
rejection rate sits below nominal across all five S6 cells and all
three specifications, while CR2 restores it to within the Monte
Carlo band around 0.05. In the information-rich cells the
calibration ratio $\kappa$ runs from 1.07 to 1.26 under the
model-based test, restored to about one under CR2, recovering two to
five points of power. The table reports both
alternatives against the unadjusted benchmark. The Exposure-weighted
advantage is held or slightly widened under CR2 at the reference
cell ($+0.080$ to $+0.092$), the small-$N$ cell ($+0.066$ to
$+0.086$), and the half-life mis-specification cell ($+0.010$ to
$+0.024$). It narrows under high autocorrelation ($+0.052$ to
$+0.032$, though it still leads), and is small under 30% MCAR
dropout ($+0.010$), which we read as the S3 finding restated, since
dropout removes the off-drug information the advantage depends on.
The Lag-adjusted gap, by contrast, is within $\pm 0.010$ of zero at
every cell under either inference, corroborating the Section 3.4
null under a second procedure and at four stress cells the main grid
does not cover. This recalibration was validated only in the Hybrid,
$N = 70$ cells, and should not be extended to the crossover design,
where Exposure-weighted does not lead in the first place.

## Sensitivity analyses

Nine Tier 2 blocks (S1-S4, S7, S8, S9, S10, S11) perturb one axis at
a time around a common reference cell (Hybrid, $N = 70$,
$c_{bm} = 0.45$, exponential DGP, $t_{1/2} = 1.0$), holding the
remaining factors fixed. Two exceptions are S9, which changes the
trial design itself, and S11, which sets $c_{bm} = 0$. A tenth block,
S6, applies the cluster-robust correction of Section 3.6 at that
same cell plus four stress cells, and an eleventh, S5, is
exploratory and reported only in the archived full-scope drafts. The
blocks are reported separately here and are not pooled with Tier 1
or with each other. Four blocks
(S1, S4, S10, S11) are reported here only as a one-paragraph summary,
with full cell-by-cell detail, tables, and figures in the
Supplement; two more (S2, S8) keep their headline finding and a
summary paragraph here, with the remainder of their detail likewise
moved to the Supplement.

### S1: within-factor autocorrelation

All nine specifications gain power monotonically with autocorrelation
$\rho \in \{0.5, 0.7, 0.8, 0.9\}$. The Exposure-weighted advantage
over Unadjusted and Lag-adjusted, roughly 0.05 to 0.08 in absolute
power, is preserved at every level examined. This suggests that
autocorrelation strength and specification choice act additively
rather than interacting. Lag-adjusted tracks Unadjusted throughout
(Supplement, Section S1).

### S2: analyst-truth half-life mismatch

Block S2 crosses two true half-lives ($t_{1/2} \in \{0.5, 1.0\}$
weeks) with four analyst-assumed values ($\hat{t}_{1/2} \in \{0.25,
0.5, 1.0, 2.0\}$ weeks). The penalty for an incorrect assumption is
modest throughout (Exposure-weighted stays within about 7 points of
its correctly-specified power at every cell but one), with an
asymmetry: under-assuming the half-life is safer than over-assuming
it. This follows from the limiting-case relationship of Section 2.4,
since shrinking $\hat{t}_{1/2}$ drives the predictor toward the
binary indicator, which remains serviceable, whereas inflating it
drives the predictor toward a constant, at which the on-off contrast
and the estimand itself degenerate. Full cell-by-cell detail,
figures, and the nine-specification extension are in the Supplement,
Section S2.

### S3: dropout

```{r fig-sens-s3, echo = FALSE, fig.cap = "Power at dropout rates $\\in \\{0, 0.1, 0.2, 0.3\\}$ under MCAR and MAR mechanisms.", out.width = "80%"}
knitr::include_graphics("../../figures/02xs-sens-S3.pdf")
```

```{r fig-heatmap-sens-s3, echo = FALSE, fig.cap = "Block S3: power by dropout rate and analysis specification, faceted by dropout mechanism, same data as the line plot above. The MAR arm has no $0\\%$ cell (Section 3.7); it shares the MCAR reference cell.", out.width = "80%"}
knitr::include_graphics("../../figures/02xs-heatmap-sens-S3.pdf")
```

```{r fig-heatmap-sens-s3-g9, echo = FALSE, fig.cap = "Block S3 extended to all nine G1-G9 specifications ($n_{sim} = 500$).", out.width = "90%"}
knitr::include_graphics("../../figures/02xs-heatmap-sens-S3-g9.pdf")
```

All three specifications lose power monotonically as dropout
increases. At the MCAR baseline ($N = 70$, no dropout),
Exposure-weighted gives $0.866 \pm 0.015$ against $0.764 \pm 0.019$
for the other two. At 30% MCAR dropout, all three collapse to
$0.122$ to $0.148$. The Exposure-weighted advantage narrows steadily
as dropout worsens, from about 0.10 absolute at zero dropout to 0.07
at 10%, 0.04 at 20%, and 0.03 at 30% under MCAR, narrowing further
under MAR at the highest rate. This is consistent with the advantage
originating in the exposure-weighted contrast from off-drug
observations, which contracts as those observations are
preferentially lost. The advantage is preserved at dropout of 20% or
below. Lag-adjusted again tracks Unadjusted throughout, the largest
discrepancy being 0.010.

### S4: biomarker-moderation effect-size curve

The power curve is sigmoidal for all three original specifications
across $c_{bm} \in \{0, 0.10, 0.20, 0.30, 0.45, 0.60\}$, saturating
once $c_{bm} \geq 0.45$. The Exposure-weighted advantage over the
unadjusted benchmark is largest at intermediate effect sizes
($+0.10$ absolute at $c_{bm} = 0.45$), shrinking as all three
approach the ceiling, so the advantage matters most for trials
targeting a moderate biomarker-treatment interaction (Supplement,
Section S4, full curve and nine-specification extension).

### S7: alternative-paradigm specifications

An earlier version of this block tested two carryover terms crossed
with the biomarker, `E5` ($\text{Sx} \sim bm + t + D_{it} +
\text{bm:}D_{it} + L_{it} + \text{bm:}L_{it}$) and `E6` ($\text{Sx}
\sim bm + t + D_{it} + \text{bm:}D_{it} + t_{sd} +
\text{bm:}t_{sd}$). This was motivated by Section 4.1's observation
that a carryover term entering only as a nuisance main effect cannot
repair the interaction's attenuation. Both underperformed Unadjusted
at every cell examined, evaluated by a joint 2-df Wald test since
each adds a second interaction coefficient. We traced this to a
structural cause discussed in Section 4.6: this manuscript's DGP has
a single true decay parameter, so a second, DGP-unmotivated
interaction coefficient mostly adds estimation noise. `E5`/`E6` were replaced by two
specifications addressing the same open problem, the analyst's
unknown decay half-life, from different angles (Section 2.4): `G4`
(AIC-selected half-life) and `G5` (paired-difference regression).
Both are single-coefficient tests, unlike `E5`/`E6`.

```{r tab-s7, echo = FALSE}
suppressPackageStartupMessages(library(dplyr))
s7 <- readRDS(here::here(
  'analysis/scripts/carryover-sensitivity/output/09-sensitivity-s7.rds'))
lab7 <- c(E1 = 'G1: Unadjusted', E3 = 'G2: Lag-adjusted',
          E2 = 'G3: Exposure-weighted', E7 = 'G4: AIC-selected t1/2',
          E9 = 'G5: Paired-difference')
tab <- s7$summary |>
  mutate(o = match(spec, names(lab7))) |>
  arrange(o) |>
  transmute(
    Specification = lab7[spec],
    `Power (MCSE)` = sprintf('%.3f (%.3f)', power, mc_se_power),
    `Mean selected t1/2` = ifelse(is.na(mean_best_t_half), '--',
                                   sprintf('%.2f', mean_best_t_half))
  )
knitr::kable(tab, booktabs = TRUE, escape = FALSE,
  caption = 'Block S7: all five G1-G5 specifications at the
    reference cell (Hybrid, $N = 70$, $c_{bm} = 0.45$,
    $t_{1/2} = 1.0$, exponential DGP, $n_{sim} = 500$, all
    replicates converged). G4 targets bm:Dbc at the AIC-selected
    half-life (candidate grid 0.25, 0.5, 1.0, 2.0); G5 targets the
    slope of the patient-level diff~bm regression. Neither is the
    same estimand as bm:Db (G1/G2), so only power is comparable
    across rows. Mean selected t1/2 is the average AIC-chosen
    half-life for G4 across replicates, against a true half-life of
    1.0.') |>
  kableExtra::kable_styling(latex_options = "scale_down")
```

```{r fig-heatmap-sens-s7, echo = FALSE, fig.cap = "Block S7: power by analysis specification, same figures as the table above.", out.width = "75%"}
knitr::include_graphics("../../figures/02xs-heatmap-sens-S7.pdf")
```

Table \@ref(tab:tab-s7) gives a genuinely different result from the
retired `E5`/`E6` comparison. Both `G4` and `G5` beat Unadjusted at
this cell, and `G4` edges out `G3` (Exposure-weighted) as well. `G4`
leads overall ($0.832$), `G3` follows ($0.826$), then `G5` ($0.804$,
the two-stage paired-difference regression with no repeated-measures
model at all), and `G1`/`G2` tie at the back ($0.748$ each, the
already-established Section 3.4 null replicated once more).

`G4`'s AIC selection is imperfect but usefully informative. The mean
selected half-life is $1.31$ weeks against a true value of $1.0$,
and the selection distribution concentrates on the true value and
its immediate neighbor (`r round(100 * mean(readRDS(here::here('analysis/scripts/carryover-sensitivity/output/09-sensitivity-s7.rds'))$results$best_t_half[readRDS(here::here('analysis/scripts/carryover-sensitivity/output/09-sensitivity-s7.rds'))$results$spec == 'E7'] %in% c(1.0, 2.0), na.rm = TRUE))`% of replicates choose $1.0$ or $2.0$ weeks). This is close
enough to the truth that `G4` slightly exceeds `G3`'s own
oracle-half-life power at this cell.

We see a more notable result in `G5`. It discards the
repeated-measures structure, the correlation model, and any
carryover representation entirely, and still beats `G1`/`G2`. This
echoes Senn, Julious and Araujo's [-@senn2014four] finding that
simplicity is not obviously costly for N-of-1 treatment-effect
estimation. None of these results should be read as general,
however. Block S9, next, tests whether the ranking survives a
different trial design, and it does not for every specification
equally.

### S8: architecture sensitivity of the specification ranking

Every result reported to this point, Tier 1 and blocks S1-S7 alike,
is restricted to Architecture B (covariance moderation), the
manuscript's stated scope (Section 2.3). Block S8 asks a narrower
question than a full architecture comparison, which is the purpose
of the companion manuscript on DGP architectures
[@pmsimstats_dgp_architectures2026]: does the ranking among analysis
specifications established above under Architecture B hold under
Architecture A (mean moderation) as well, or is the choice of
carryover representation itself architecture-dependent? We refit the
five specifications of block S7's table at the same reference cell
under both architectures, at three half-lives
($t_{1/2} \in \{0, 0.5, 1.0\}$).

The S7 result (`G4` and `G5` both beating `G1`) turns out to be
architecture-specific, joining Exposure-weighted in a pattern that
grows with carryover severity rather than switching on at some
threshold. Under Architecture B, `G4` leads at $t_{1/2} = 1.0$
($0.850$), ahead of `G5` ($0.842$), `G3` ($0.836$), and `G1`/`G2`
($0.760$). Under Architecture A the pattern reverses for `G4` and
`G3` alike: `G1`/`G2` lead ($0.972$/$0.976$), `G5` trails only
slightly ($0.964$), while `G3` ($0.946$) and `G4` ($0.950$) trail by
2 to 3 points, a real if modest reversal. This matches the archived
full-scope drafts' finding that Exposure-weighted is marginally
dominated under mean moderation (Section 4.6). `G4` inherits the
same sensitivity, since it shares Exposure-weighted's formula. `G5`
is comparatively more robust to architecture, never reversing to a
large disadvantage either way. Lag-adjusted (`G2`) remains
statistically indistinguishable from Unadjusted (`G1`) under both
architectures at every half-life, extending Section 3.4's clean
null to Architecture A as well. The one specification recommendation
this manuscript makes, exposure weighting in the Hybrid and OL+BDC
designs (Conclusions), is therefore itself conditional on
Architecture B and grows more important as carryover worsens. The
same qualification now extends to `G4`. Full cell-by-cell detail,
the table, and both architecture panels are in the Supplement,
Section S8.

### S9: does the reference-cell ranking survive under CO?

Blocks S7 and S8 found `G4` and `G5` both beat Unadjusted at the
Hybrid reference cell under Architecture B, with `G4` slightly
exceeding Exposure-weighted (`G3`) itself. Block S9 asks whether that
ranking is specific to Hybrid's short, compressed washout, by
refitting all five G1-G5 specifications under the CO design (off-drug
arm spanning 2.5 to 20 weeks post-discontinuation, Section 2.2) at
two true half-lives.

```{r fig-heatmap-sens-s9, echo = FALSE, fig.cap = "Block S9: power under the CO design at two true half-lives, all five G1-G5 specifications.", out.width = "80%"}
knitr::include_graphics("../../figures/02xs-heatmap-sens-S9.pdf")
```

It is not, and including `G3` in this block changes the diagnosis
reached from `G4` alone. At $t_{1/2} = 1.0$, `G1`/`G2` lead
($0.834$/$0.838$); `G5` trails moderately ($0.750$); `G3` and `G4`
both collapse, to $0.482$ and $0.474$ respectively, statistically
indistinguishable from each other (gap $0.008$, well under one MCSE).
At $t_{1/2} = 0.5$ the same pattern holds ($0.866$/$0.864$ for
`G1`/`G2`, $0.778$ for `G5`, $0.546$/$0.552$ for `G3`/`G4`, again
indistinguishable). Because `G3` uses the *true* half-life, not an
estimated one, its near-identical collapse shows that `G4`'s failure
under CO is inherited almost entirely from Exposure-weighted's own
established weakness under this design (Section 3.4: $0.488$ versus
$0.830$ in the Tier 1 grid), not created by the AIC-selection step.
The selection step does have a real, separately confirmed pathology
under CO: it concentrates overwhelmingly on the largest candidate
half-life in the grid ($2.0$ weeks, chosen in 483 of 500 replicates
at $t_{1/2} = 1.0$ and 435 of 500 at $t_{1/2} = 0.5$, regardless of
which true half-life generated the data), consistent with a weakly
informative likelihood surface across candidates when off-drug
occasions are few and widely spaced. But that mis-selection costs
`G4` essentially nothing beyond what `G3` already loses under CO
with the correct half-life in hand. The dominant story is the
continuous decayed predictor's mismatch with CO's design generally
(Section 3.4). AIC selection is an identified but practically minor
addition to that story here, not the primary cause it appeared to be
from `G4` in isolation.

`G5`'s milder reversal has no comparable mechanical explanation. It
simply reflects that CO's own within-subject contrast (Section 3.4)
already captures most of the signal a between-patient
paired-difference regression is built to extract, so discarding the
repeated-measures structure costs more there than it does under
Hybrid. None of `G3`, `G4`, or `G5`'s Hybrid-cell behavior is
design-general, then. The design-specific guidance in Section 4.5
('What analysis is best in which situations, in practice') should be
read accordingly: `G3`/`G4` should not be used under CO regardless
of whether the half-life is assumed or estimated.

### S10: does a CR2 correction change the ranking among G1-G9?

Section 3.6 (Block S6) showed the model-based interaction test is
mildly conservative, and that a CR2 cluster-robust standard error
recovers two to five points of power without disturbing the ranking
among `G1`, `G3`, and `G2`. That correction had not been applied to
`G4` or `G5`. This block asks whether CR2 helps `G4` and `G1`/`G2`
the way it already helps `G3`, and whether it changes which
specification leads at the Hybrid, Architecture B reference cell. We
use the compact `G1`-`G9` naming layer for the resulting nine-way
comparison: `G6`-`G9` cross CR2 with `G1`, `G2`, `G4`, and `G3`. `G5`
has no CR2 variant, since its paired-difference regression collapses
to one row per patient before fitting, leaving no repeated-measures
cluster structure for CR2 to correct.

CR2 helps every specification it was applied to: `G1`/`G2`
(Unadjusted and Lag-adjusted, both $0.748$ model-based) rise to `G6`
$0.776$ and `G7` $0.764$. The ranking among the five base
specifications is unchanged by these gains, since `G1`/`G2` remain
well below `G3`-`G5` regardless. What does change the ranking is
`G9`: CR2 lifts `G4` (AIC-selected half-life, $0.832$ model-based) to
$0.870$, a $3.8$-point gain, the largest CR2 correction observed
anywhere in this manuscript, enough for `G9` to overtake every other
specification tested, including `G8` ($0.863$). This did not come
with a Type I error cost (Section S11 of the Supplement checks this
directly): every replicate converged under `G9`, and the power gain
is at least consistent with CR2 correcting real conservatism rather
than papering over a different problem. Full detail, the reference-
cell table, and the heatmap are in the Supplement, Section S10.

### S11: is G4's Type I error controlled?

`G4` selects its half-life by AIC and then tests the interaction
coefficient from the selected model, a post-selection inference
problem: the same data drive both the selection and the test, which
can in principle inflate Type I error beyond nominal. Block S11
checks this directly, refitting `G1`-`G5` and `G9` at the null
($c_{bm} = 0$) at the Hybrid reference cell, with $n_{sim} = 2000$
for tighter precision on a rate expected near $0.05$.

The concern does not materialize. `G4`'s Type I error is $0.038$,
conservative and in the same range as `G1`-`G3` ($0.034$-$0.036$),
not inflated by the selection step. `G9` (`G4`+CR2) is $0.052$,
indistinguishable from nominal $0.05$. Read together with Section
3.6's power results, `G9`'s status as the best-performing
specification (Section 3.7, Block S10) is not an artifact of an
uncontrolled test: it wins on power while holding Type I error at
nominal, the same standard `G3`+CR2 (`G8`) is held to. This does not
generalize beyond the single cell tested. Full detail and the
heatmap are in the Supplement, Section S11.

## A higher-precision check

Because a number of the Tier 1 gaps are small, we reran a focused
24-cell grid (OL+BDC design, $N \in \{35, 70\}$, true half-life
$\in \{0.5, 1.0\}$) at 600 trials per cell, all converging. We
compared specifications with a paired McNemar test, considerably
more sensitive than an independent-groups comparison since all three
are fit to the same trials. Type I error across the twelve null
cells fell in $[0.003, 0.053]$, so nominal alpha holds. Paired
McNemar tests place Exposure-weighted ahead of the unadjusted
benchmark at all four alternative cells, by $+2.3$ to $+13.7$
percentage points. The largest gap is at the highest-leverage cell
($N = 70$, true $t_{1/2} = 1.0$ weeks, $p = 6.6 \times 10^{-17}$),
with $p \leq 0.015$ at the remaining three, a sharper confirmation
than the Tier 1 grid alone provides.

One scope caveat governs the rerun's third arm. It was produced by
the package's `lme_analysis()` routine, whose carryover option
appends time since discontinuation as a linear main effect rather
than the lagged indicator $L_{it}$ used in the Tier 1 grid. It is
therefore not the Lag-adjusted specification of Section 2.4 and is
not reported as such here. That alternative remains a reasonable
specification we have not evaluated (Section 4.6).

# Discussion

## The lagged-treatment term does no work

The clearest result of this study is a negative one. Adding the
classical lagged-treatment indicator to the analysis model changed
nothing measurable, anywhere in the grid. Across all 216 cells the
mean power difference against the unadjusted benchmark was 0.003
against an MCSE of 0.018. Bias, MSE, and coverage agreed to three
decimal places, and the equality held in every sensitivity block and
all five cluster-robust stress cells under both inference
procedures.

The strength of this conclusion rests on the design rather than the
grid size. The two specifications estimate the same coefficient,
were fitted to the same simulated datasets under common random
numbers, and differ by exactly one term. There is therefore no
confounding from a change of estimand, from Monte Carlo variation
between arms, or from differential test size (both are conservative
to the same degree, Section 3.6). What is left is the term itself,
and it contributed nothing.

The mechanism follows from where the term sits. Carryover damages
the interaction test on two fronts (Section 1). It attenuates the
coefficient, because contaminated occasions are coded as drug-free,
and it inflates the residual variance, because those occasions are
heterogeneous in a way the model does not represent. The lagged
indicator enters only as a nuisance main effect, with no
$\text{bm:}L_{it}$ product term, so the tested interaction remains
$\text{bm:}D_{it}$ and the attenuation is untouched. It can in
principle absorb some of the second effect, but evidently absorbs
too little to matter, since it marks a single occasion per
randomization path and is constant thereafter, unable to describe a
graded decline across a two- to four-occasion washout.

A more capable, elapsed-time-indexed nuisance term might succeed
where the crude lagged indicator fails. We therefore ran a targeted
check on the alternative the reference implementation provides,
`Sx ~ bm + t + Db + bm:Db + tsd`, called the washout-adjusted
specification (`E4`), across the three designs at true half-lives of
0.5 and 1.0 weeks ($N = 70$, $c_{bm} = 0.45$, exponential DGP, 500
replicates). It does slightly better, and the improvement is
instructive rather than material. Type I error is at nominal for
both (mean 0.050 versus 0.045). Power exceeds the benchmark in all
six cells by a mean of 1.3 points (sign test $p = 0.031$), reaching
significance under crossover at both half-lives ($+1.8$, $p=0.008$;
$+2.2$, $p=0.003$) and OL+BDC at the longer one ($+2.2$, $p=0.015$),
but not Hybrid ($+1.0$, $p=0.27$). The mechanism is exactly the one
Section 2.4 predicts. Point estimates are unchanged, so nothing
about the attenuation is repaired. The entire effect runs through
the standard error, which only a long enough washout gives it
something to describe. The crossover schedule places
off-drug occasions 2.5 to 10 weeks after discontinuation, against
one to two weeks for Hybrid. The practical ceiling on what any
carryover nuisance term can deliver is therefore about two
percentage points, and only in long-washout designs, against the
nine to eleven points exposure weighting delivers where it works.
This check covers only the exponential DGP at a single $N$ and
effect size, so it is a probe rather than a fourth arm. It shares
every specification's structural limitation of entering carryover as
a main effect only (Section 4.6). For practitioners the implication
is direct. A
carryover nuisance term of either form should not be expected to
recover meaningful power lost to carryover, and reporting one does
not constitute an effective adjustment for it.

## The exposure-weighted advantage, and where it fails

The exposure-weighted predictor does help, but conditionally. At its
best cell (Hybrid, $N = 70$, $c_{bm} = 0.45$, $t_{1/2} = 1.0$) it
exceeds the unadjusted benchmark by about 10 points of power (0.860
versus 0.766, roughly five MCSEs and so not attributable to chance),
confirmed at all four cells of the Section 3.8 high-precision rerun.
Under the crossover design it performs considerably worse (0.488
versus 0.830). We think this happens because the within-subject
correlation there already carries most of the signal the decaying
predictor was designed to recover, so it gains little while its
down-weighting of off-drug observations works against it. The
recommendation therefore holds only in the two non-crossover
designs, and should not be extended to a crossover trial, where the
unadjusted binary indicator is materially better.

That the one specification addressing attenuation is also the only
one to gain power, while the one addressing only residual variance
gains nothing, suggests attenuation is the mechanism that matters
here. We read this as an interpretation consistent with the results,
not as a finding the design isolates directly.

## What the robustness checks add

The Section 3.7 sensitivity blocks show the exposure-weighted
advantage, where it exists, is not uniformly fragile, but it is not
unconditionally robust either. Under incorrect decay-shape
assumptions (Figure \@ref(fig:fig-heatmap-mismatch), Section 3.5) it
widens on the light-tailed side and narrows to statistical
insignificance on the heavy-tailed side, disappearing entirely at
the two most extreme Weibull shapes examined ($k = 0.25$, $0.5$); it
survives an incorrectly assumed half-life, short of a large
over-assumption at a short true half-life (Block S2, Supplement
Section S2); and it holds its five- to eight-point lead across the
autocorrelation range examined (Block S1, Supplement Section S1).
Decay-shape mis-specification is therefore the one robustness axis
examined here that can erode the advantage away entirely, rather
than only narrowing it.

One caveat qualifies the Tier 1 figures throughout. Outside block
S2, the analysis-side half-life is set equal to the data-generating
one. This grants Exposure-weighted an oracle no real analyst
possesses, so its Tier 1 margins are upper bounds, and block S2
gives the more realistic figure.

## Dropout matters more than the method

The finding with the largest practical consequence concerns missing
data, not carryover. When patients drop out (Figure
\@ref(fig:fig-sens-s3)), power collapses for all three
specifications together, to between 0.12 and 0.25 at 30% dropout,
essentially indistinguishable. No specification offers protection,
since each can only organize whatever data remain. MAR dropout is
somewhat worse than MCAR, because the sickest patients, who carry
the most signal, tend to leave first. Retaining patients in the
trial yields considerably more power than any choice among these
three specifications ever could.

## What analysis is best in which situations, in practice

Collecting the findings above into a decision rule rather than a set
of pairwise comparisons:

- **Covariance moderation (Architecture B), Hybrid or OL+BDC
  design.** Use Exposure-weighted with a CR2 cluster-robust standard
  error. The advantage holds under half-life mis-specification short
  of a large over-assumption (Section 3.7, Block S2), and across the
  autocorrelation range examined (Block S1). It holds and widens
  under decay-shape mis-specification on the light-tailed side, but
  narrows to statistical insignificance at the two most extreme
  heavy-tailed shapes examined (Section 3.5): this recommendation
  should not be read as holding under an arbitrarily heavy-tailed
  decay form. CR2 restores nominal Type I error without disturbing the
  ranking (Block S6, validated at $N \in \{35, 70\}$ in the Hybrid
  design specifically; OL+BDC was not itself stress-tested under
  CR2). This remains the recommended default: it has been through
  the full robustness battery `G4`/`G5` have not (next point).
- **`G4` (AIC-selected half-life) and `G5` (paired-difference) are
  promising at this same cell, not yet validated defaults.** Both
  beat Exposure-weighted itself at the Hybrid, Architecture B
  reference cell (Blocks S7-S8: `G4` $0.850$, `G5` $0.842$, `G3`
  $0.836$, at $t_{1/2} = 1.0$), and `G4` in particular removes the
  need to assume a half-life at all. Neither has been tested against
  decay-shape mis-specification, dropout, or under CR2 the way `G3`
  has, so this is grounds to investigate them further, not to
  displace the established default yet. Adding a CR2 correction to
  `G4` (`G9` in the nine-way naming layer) goes further still: at
  this same cell it outperforms every one of the other eight
  specifications, including `G3`+CR2, with no accompanying Type I
  error inflation (Section 3.7, Blocks S10-S11). This is the single
  best result anywhere in this manuscript, and also the least
  validated. It rests on one cell, with none of `G4`'s own
  robustness gaps yet closed.
- **Crossover (CO) design, any architecture.** Use Unadjusted.
  Exposure-weighted is markedly worse under CO (Section 3.4), and
  Block S9 shows `G4` and `G5` fail here too, `G4` catastrophically.
  `G4`'s AIC half-life selection systematically drifts to the largest
  candidate in its grid under CO's sparse, widely spaced off-drug
  sampling (chosen in over 85% of replicates regardless of the true
  half-life), collapsing its power to roughly half of Unadjusted's.
  It should specifically be avoided under any design with few,
  widely spaced off-drug observations. `G5` also underperforms
  Unadjusted under CO, though far more mildly. The classical
  lagged-treatment remedy adds nothing here either (next point),
  including in the one design where its washout is long enough to
  give it a chance (Section 4.1).
- **Architecture uncertain, or known to be mean moderation
  (Architecture A).** Use Unadjusted or Lag-adjusted, not
  Exposure-weighted or `G4`. Both specifications' advantage under
  covariance moderation reverses, marginally, under mean moderation
  (Block S8). `G5` is the one specification whose performance does
  not depend much on which architecture generated the data, small
  under B, smaller under A, but it has been tested at only one design
  and reference cell outside the Hybrid comparison (Block S9).
- **Never rely on Lag-adjusted as a carryover remedy.** It is
  statistically indistinguishable from Unadjusted everywhere tested
  in this manuscript, Tier 1 and every Tier 2 block alike (Section
  4.1). If a reviewer or protocol expects some carryover term to be
  present, including it costs nothing measurable, but it should not
  be reported as having adjusted for carryover.
- **Minimize dropout before optimizing specification choice.** It is
  the single largest lever in the entire grid (Section 4.4), and no
  specification, including `G4`/`G5`, has been shown to offer any
  protection against it; their behavior under dropout is untested.

### Why G4 and G5 succeed at Hybrid but fail under CO

Both new specifications' Hybrid, Architecture B advantage evaporates
under the CO design (Block S9), for two different, design-specific
reasons rather than a shared structural flaw of the kind that sank
`E5`/`E6` (Section 4.6). `G4`'s failure is mechanical. With only a
few, widely spaced off-drug occasions, the likelihood surface across
candidate half-lives is weakly informative, and AIC selection drifts
to the boundary of its candidate grid rather than locating the true
value, systematically overestimating the half-life. Section 3.7's S2
block already showed that over-assumption is the more damaging
direction of half-life error when the half-life is fixed by the
analyst. Under CO, `G4` reproduces that same damaging error through
the estimation procedure itself, on nearly every replicate.

This is a concrete, testable prediction: `G4` should be reliable
whenever a design gives several off-drug occasions across a graded
range of elapsed time, as Hybrid does despite its short overall
washout, and unreliable whenever off-drug sampling is sparse,
however long the washout runs in calendar time.

`G5`'s failure has no comparable mechanical account and is milder.
The most plausible explanation is that CO's own within-subject
contrast (Section 3.4) already carries most of the signal a
between-patient paired-difference regression is built to extract, so
discarding the repeated-measures and correlation structure costs more
under CO than it does under Hybrid, where that structure is doing
more work. This was not tested directly and would require comparing
`G5` against the within-cell decomposition of `G3`'s advantage to
confirm.

Two implications follow. First, these results reinforce rather than
undercut the earlier `E5`/`E6` diagnosis (Section 4.6). The more
fundamental unresolved question is not which design an alternative
specification is tested under, but whether the underlying
data-generating process rewards that specification's added structure
at all, and neither `G4` nor `G5` was tested against a DGP built to
favor them specifically. Second, the practical recommendation is
narrower than 'estimate the half-life' or 'use a simpler method':
both `G4` and `G5` are candidates worth validating further within
the Hybrid/OL+BDC, Architecture B cell where this manuscript's other
recommendations already concentrate, not general-purpose
replacements for Exposure-weighted or Unadjusted.

## Scope, related work, and limits

This manuscript deliberately narrows the full-scope grid to the
covariance-moderation architecture and to the non-linear decay
forms, in the interest of a self-contained evaluation of manageable
length. The archived full-scope drafts (`archive/report.Rmd` and
`archive/report-slim.Rmd`) show that the exposure-weighted advantage
documented here does not carry over to mean moderation, where the
biomarker moderates the mean response directly and Exposure-weighted
is marginally dominated in every design examined. Readers who need
that cross-architecture comparison should consult those drafts
rather than extrapolate from this one. Establishing what the
originally published power values (Section 3.3) were computed under
remains outstanding work.

The crossover tradition surveyed by Jones and Kenward
[-@jones2014crossover] would lead one to expect a shape-free lagged
term to outperform a parametric predictor once the decay shape is
mis-specified. Our results do not support that expectation over the
range examined, for two reasons. The exposure-weighted penalty for
an incorrect half-life is too small to be overtaken (Section 3.7),
and more fundamentally the lagged term does not improve on ignoring
carryover at all (Section 4.1), leaving no advantage for
decay-shape mis-specification to erode.

One adjacent specification bounds the negative result of Section
4.1. Every carryover term considered there, lagged and
washout-adjusted alike, enters as a nuisance main effect with no
product against the biomarker. That finding is therefore that a
carryover *main effect* does not recover the attenuated signal, not
that no shape-free specification can. The natural candidate is a
carryover term interacted with the biomarker, either
`Sx ~ bm + t + Db + bm:Db + L + bm:L` or its washout-adjusted
counterpart with `bm:tsd`, letting the biomarker effect decline
across the washout without committing to a rate.

An earlier version of Block S7 tested exactly these two
specifications and found a clean negative result at every cell
examined. We traced this to a structural cause: this manuscript's
DGP has a single true decay parameter, so a second, interacted
carryover coefficient adds estimation noise without a corresponding
signal to recover. They were replaced by two specifications
addressing the same open problem, the analyst's unknown decay
half-life, from different angles: `G4` (AIC-selected half-life) and
`G5` (a paired-difference regression with no carryover
representation at all).

Both beat Unadjusted at the Hybrid, Architecture B reference cell,
`G4` even edging out Exposure-weighted itself (Block S7). Neither
survives a change of trial design or architecture (Blocks S8-S9).
`G4`'s AIC selection fails specifically under sparse off-drug
sampling (Section 4.5), and `G5`'s advantage appears to depend on
the repeated-measures structure it discards being otherwise
underused, which is design-dependent. Block S8 shows the
Lag-adjusted null of Section 3.4 replicates under Architecture A, so
it is not an artifact specific to covariance moderation.
Exposure-weighted's advantage, and now `G4`'s, are
architecture-sensitive, reversing in direction, though not
substantially, under Architecture A (Section 3.7, Block S8).

Three further N-of-1 analysis approaches sit outside the G1-G9
family and are not evaluated here, because each requires substantial
independent methodological development rather than a drop-in
specification within our existing pipeline.

Tang and Landes [-@tang2020ttests] propose formula-based $t$-tests
that correct a single-subject paired-difference test's standard
error and degrees of freedom for an estimated first-order serial
correlation, the natural refinement `G5` lacks (`G5` discards
within-patient correlation entirely). Their tests have no
biomarker-moderation term, though, and would need a `diff ~ bm`
extension with a serial-correlation-aware variance to enter our
comparison.

Schmid and Yang [-@schmid2022bayesian] describe a Bayesian
hierarchical alternative to the frequentist mixed-model and
sandwich-estimator family spanning `G1`-`G9`, with partial pooling
of individual-level treatment-effect, trend, and autocorrelation
parameters across patients. They note that carryover is tractable
only in aggregated, multi-individual analyses like ours, but a
biomarker-moderation extension of their model is a separate
undertaking.

Daza [-@daza2018causal] frames N-of-1 inference as a counterfactual,
time-varying-treatment problem, estimated via g-computation or
inverse-probability weighting. This is a structurally different
paradigm from `G1`-`G9`'s explicit exponential-decay or lag
covariates: carryover there is confounding to be controlled rather
than a predictor to be modeled.

None of the three is a candidate for a tenth specification within
this manuscript's scope, but each identifies a direction future work
on this estimand should consider.

Two further limitations bound what we can conclude. The numbers
presented here come from a single parameter set calibrated to a
prazosin and PTSD trajectory, so absolute power values should be
read as illustrative and the ranking, itself conditional on trial
design, as the more portable result. And the sensitivity checks vary
one factor at a time and can therefore miss genuine interactions
between factors, for instance between high autocorrelation and heavy
dropout. The one joint check available to us, block S5 in the
full-scope manuscript, revealed no such interaction.

# Conclusions

1. **The lagged-treatment term does nothing.** It never improved on
   ignoring carryover altogether, on any measure, in any design, at
   any half-life, and in every sensitivity block, a clean null given
   the two specifications share an estimand and differ by a single
   term. The term enters as a nuisance main effect only, leaving the
   attenuation of the interaction untouched, and should not be
   reported as an adjustment for carryover.

2. **Exposure weighting helps, but conditionally.** It achieves the
   highest power, about 10 points over the unadjusted benchmark, only
   in the Hybrid and OL+BDC designs. Its lead survives mis-specified
   half-life there, and widens under a light-tailed decay-shape
   mis-specification, but narrows to statistical insignificance at
   the two most heavy-tailed Weibull shapes examined; it is not
   robust to an arbitrarily heavy decay tail. Under crossover it
   is markedly worse than the binary indicator (0.488 against 0.830)
   and should not be used. Where used, we recommend a CR2
   cluster-robust standard error (Block S6).

3. **An approximate half-life is adequate, but err short.** A
   fourfold error at a true half-life of one week costs 6.6 points of
   power, so a rough pharmacokinetic estimate suffices in practice.
   The error is asymmetric. Under-assuming drives the predictor
   toward the still-serviceable binary indicator, while over-assuming
   erodes the advantage to a tie. Outside block S2 the analyst is
   granted the true half-life, so the Tier 1 margins are upper
   bounds.

4. **Dropout matters more than the choice of method.** Power falls
   sharply as dropout increases, roughly halving at 20%, equally
   across all three specifications. Preventing dropout is a more
   effective use of design effort than any choice among them.

5. **The interaction test is somewhat conservative, and correctably
   so.** All three specifications reject slightly less often than
   nominal under the null. A CR2 cluster-robust standard error
   restores the correct rate while recovering modest power, and the
   exposure-weighted advantage is held or widened under CR2 at four
   of five S6 stress cells (Block S6).

6. **This is a narrowed slice of a larger study.** Mean moderation
   and linear decay fall outside this manuscript's scope. The
   archived drafts `archive/report.Rmd` and `archive/report-slim.Rmd`
   are the authority on those comparisons and should be consulted
   before generalizing beyond covariance moderation. We also make no
   claim of numerical agreement with the originally published power
   values (Section 3.3).

7. **Carryover terms interacted with the biomarker do not help; two
   alternative-paradigm specifications do, but only at the reference
   cell.** Crossing a lagged or washout carryover term with the
   biomarker gave lower power than the unadjusted benchmark at every
   cell examined (Block S7's original version). We traced this to a
   structural cause: this manuscript's DGP has a single true decay
   parameter, so a second, interacted coefficient adds estimation
   noise without a corresponding signal to recover (item 1's
   mechanism, restated). They were replaced by two specifications
   answering the same open question, the analyst's unknown decay
   half-life, differently. `G4` estimates it by AIC selection; `G5`
   discards the decay representation, and the mixed model itself,
   for a two-stage paired-difference regression following Senn,
   Julious and Araujo [-@senn2014four]. Both beat Exposure-weighted
   at the Hybrid, Architecture B reference cell (Blocks S7-S8; `G4`
   $0.850$, `G5` $0.842$, `G3` $0.836$ at $t_{1/2} = 1.0$). Neither
   result generalizes: Block S8 shows both reverse, like
   Exposure-weighted itself, under Architecture A, and Block S9
   shows both fail under the crossover design, `G5` mildly and `G4`
   catastrophically. `G4`'s CO failure has a clear mechanism: AIC
   selection systematically drifts to the boundary of its candidate
   half-life grid when off-drug sampling is sparse and widely
   spaced, reproducing the S2 finding that over-assuming the
   half-life is the more damaging error, now through the estimation
   procedure itself rather than a fixed wrong assumption. Both
   specifications should be read as promising candidates for further
   validation at the one cell type where this manuscript already
   recommends Exposure-weighted, not as general-purpose replacements
   for it or for Unadjusted.

8. **Adding CR2 to the AIC-selected specification gives the single
   best result in this manuscript, at one cell only.** `G9` (`G4`
   plus a CR2 cluster-robust standard error) outperforms all eight
   other specifications at the Hybrid, Architecture B reference
   cell, including `G3`+CR2, and its Type I error holds at nominal
   despite the post-selection risk `G4`'s half-life selection step
   raises in principle (Section 3.7, Blocks S10-S11). We read this
   as a promising direction, not a recommendation: `G9` has not been
   tested against decay-shape mis-specification, dropout, a
   crossover design, or an alternative architecture, all of which
   item 2's specification (`G3`+CR2) has already been checked
   against.

# Conflict of interest

The authors declare no conflicts of interest.

# Funding

This work received no specific funding.

# Data availability statement

The data and code supporting the findings of this study are
available in the pmsimstats-ng project repository. Per-replicate
Monte Carlo outputs and ADEMP-compliant cell-level summaries are
versioned under `analysis/data/`. Simulation drivers are at
`analysis/scripts/carryover-sensitivity/`. Exact paths for this
manuscript are listed in the Reproducibility section below.

# Reproducibility

This manuscript was rendered with R 4.6.0 inside the project's
zzcollab Docker container. The production simulation ran under R
4.5.3 (§3.6), with the image hash in `Dockerfile` and the package
set in `renv.lock`. The simulation study uses `nlme` for the continuous-time
linear-mixed analysis, `parallel` for the 8-core parameter-grid
sweep, and `data.table` or `furrr`/`purrr` in the two simulation
collections. No simulation code runs during knitting, which loads
pre-computed checkpoints only. Each driver sets a master seed
(20260507) and derives per-replicate seeds from it via
`parallel::mc.set.seed`.

**Data and code availability.** All simulation outputs, cell-level
summaries, and driver scripts are versioned under
`analysis/scripts/carryover-sensitivity/` and `analysis/data/` in
the pmsimstats-ng repository. This includes the principal factorial,
the S1 to S5 sensitivity blocks, the S6 cluster-robust recalibration,
the washout-adjusted (E4) check of Section 4.1, and the
high-precision prototype rerun of Section 3.7. It also includes the
S7 alternative-paradigm block
(`09-run-sensitivity-s7.R`, `output/09-sensitivity-s7.rds`); the
S8 architecture-by-half-life block
(`10-run-sensitivity-s8.R`, `output/10-sensitivity-s8.rds`); the
S9 CO-design block
(`14-run-sensitivity-s9.R`, `output/14-sensitivity-s9.rds`); the
S10 CR2 comparison block
(`15-run-sensitivity-s10.R`, `output/15-sensitivity-s10-g.rds`); and
the S11 Type I error block
(`16-run-sensitivity-s11.R`, `output/16-sensitivity-s11.rds`). All
five were added 2026-08-18/19 (master seed 20260415, 500 replicates
per cell except S11's 2000; Hybrid reference cell except S9, which
uses CO, and S11, which sets $c_{bm} = 0$). `E5`/`E6`
(biomarker-interacted carryover terms) were tested in an earlier
version of S7-S9 and are documented in Section 4.6; they were
replaced by `G4` (AIC-selected half-life, reusing the candidate-grid
logic of `characterize_carryover()`, `R/carryover_analysis.R`) and
`G5` (paired-difference regression). The `E1`/`E3`/`E2`, `E7`, `E9`
fitters, and the `E1cr2`/`E3cr2`/`E7cr2` CR2 variants used by S10
(`cr2_extract()`, `fit_spec_s7`, `simulate_cell_s7`), are appended to
`simulation-core.R` alongside the Tier 1 `fit_spec`/`fit_all_specs`
pair, which they do not modify. `spec-labels.R` additionally defines
the `G1`-`G9` compact display-code layer (`g_code`, `g_labels`) used
in S10's table and figure; `G8` (`G3`+CR2, stored as `E2cr2`) is reused directly from
Block S6 rather than rerun, so it does not share common random
numbers with the rest of S10's comparison. Figure
\@ref(fig:fig-hendrickson-c), and the decay-form mismatch panel
(Figure \@ref(fig:fig-heatmap-mismatch)), retain the original three
specifications (G1-G3), since decay shape is the only axis those two
figures vary. Both were regenerated from a dedicated 30-cell
simulation at a more extreme Weibull shape range
($k = 0.25, 0.5, 2.0, 4.0$, extended from an initial $k = 0.5, 2.0$,
itself replacing an earlier $k = 0.7, 1.5$; $k = 1.0$ is
not a separate condition, since it duplicates Exponential exactly)
than the shared production grid (`02-grid-summary.rds`) still uses
(`23-run-decay-shape-sensitivity.R`,
`02-decay-shape-sensitivity.rds`, $n_{sim} = 500$, Architecture B
only). This is a standalone script rather than a change to
`01-run-factorial.R`'s shared grid, since that grid is also consumed
by the archived full-scope drafts and other manuscripts. Figures
\@ref(fig:fig-hendrickson-a) and
\@ref(fig:fig-hendrickson-b) were later expanded to all nine G1-G9
specifications, from a dedicated 30-cell simulation restricted to
the exponential DGP and Architecture B
(`19-run-tier1-hendrickson-g9.R`,
`02-grid-summary-hendrickson-g9.rds`, $n_{sim} = 500$).
This replaced the original three-specification version of both
panels rather than supplementing it, and superseded the separate
Type I error checks the manuscript previously reported at Figures
6-9 (the Tier 1 Type I boxplot/heatmap and the S6 G1-G3/G1-G9 Type I
dot plot/heatmaps), since the $c_{bm} = 0$ column of Panel A now
reports Type I error for all nine specifications directly. Those
retired figures' underlying data and scripts
(`03-render-figures-extra-slim.R`,
`07-type1-cr2-figure-extra-slim.R`,
`18-run-sensitivity-s6-g9.R`) are retained for archival purposes but
no longer referenced by this document. The corresponding Tier 2
heatmaps (S1-S4, S6, S7, S8, S9, S10, S11) are produced by
`13-hendrickson-heatmap-tier2.R` from `02-sensitivity-summary.rds`
and the S7/S8/S9/S10/S11 output `.rds` files (plus `diag-s6-cr2.rds`
for `G8` and Table \@ref(tab:tab-s6)); these Tier 2 figures are
additional to, not replacements for, the existing S1-S4 line plots
and the S7/S8 kable tables. The grid design and
Morris ADEMP pre-registration are in
`simulation-grid-plan.md` in the same directory. Superseded
wider-scope drafts are retained under `archive/`, documented in its
`README.md`.

**Specification labels.** All nine specifications are stored under
`E`-prefixed codes (`E1`, `E2`, `E3`, `E7`, `E9`, `E1cr2`, `E2cr2`,
`E3cr2`, `E7cr2`) in every data file and script, and this manuscript
displays them uniformly as `G1`-`G9`, or by name, at render time
only. The mapping is applied by `spec-labels.R`, so the stored
values and archived outputs are unaffected: what a reader sees as
`G3` (Exposure-weighted) in the prose above is stored as `E2` in
every `.rds` file this manuscript reads. Output archived before
2026-07-31 stores the specifications as `A1`, `A2`, `A3`. The
migration is documented in `analysis/report/NOTATION.md`.

# References

::: {#refs}
:::

<!--
  No in-document footer chunk. The page footer (source path, render
  time, git version) is injected at render time by tools/stamp.tex
  via the knit: hook above, which wraps long paths correctly.
-->
