---
title: "Robustness of carryover-mitigation analysis strategies
  for biomarker-treatment interaction: a factorial simulation
  study"
author:
  - pmsimstats team
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: true
    number_sections: true
    keep_tex: true
    highlight: tango
fontsize: 11pt
geometry: "left=3cm,right=5cm,top=2cm,bottom=2cm"
bibliography: references.bib
csl: statistics-in-medicine.csl
link-citations: true
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{lineno}
  - \linenumbers
  - \usepackage{setspace}
  - \setstretch{1.1}
  - \newcommand{\rmdsource}{}
  - \newcommand{\rendertimestamp}{}
knit: (function(input, encoding) source(normalizePath(file.path(dirname(input), '..', '..', '..', 'tools', 'stage-render.R')))$value(input, encoding))
---

<!--
  Build pipeline for this manuscript
  ==================================

  This Rmd does NOT run any simulation code. It loads pre-computed
  summaries and figures produced by the scripts below. Re-run the
  relevant steps when the DGP, analysis specs, or parameter grid
  change. All paths are relative to the repo root.

  1. Tier 1 principal factorial (heaviest step, ~85 min on 8 cores):
       Rscript analysis/scripts/carryover-sensitivity/01-run-factorial.R
     Writes: analysis/scripts/carryover-sensitivity/output/01-factorial.rds

  2. Tier 2 sensitivity blocks S1..S5 (~4 min on 8 cores; requires
     Tier 1 not to be running concurrently - both request all cores):
       Rscript analysis/scripts/carryover-sensitivity/04-run-sensitivity-blocks.R
     Writes: analysis/scripts/carryover-sensitivity/output/04-sensitivity.rds

  3. Summarise both tiers into cell-level power/bias/convergence.
     Step 3 is cheap (~seconds) and should be re-run after either 1
     or 2 changes. It mirrors the summaries to analysis/data/:
       Rscript analysis/scripts/carryover-sensitivity/02-summarise-grid.R
     Writes:
       analysis/data/02-grid-summary.rds        (from Tier 1)
       analysis/data/02-sensitivity-summary.rds (from Tier 2, if present)

  4. Render Tier 1 figures (reads 02-grid-summary.rds):
       Rscript analysis/scripts/carryover-sensitivity/03-render-figures.R
     Writes:
       analysis/figures/02-power-by-spec.pdf
       analysis/figures/02-heatmap-matched-vs-mismatched.pdf
       analysis/figures/02-type1-boxplots.pdf

  5. Render Tier 2 sensitivity figures (reads 02-sensitivity-summary.rds;
     writes placeholder PDFs with "Pending Tier 2 production run" when
     the summary is absent, so this manuscript always knits cleanly):
       Rscript analysis/scripts/carryover-sensitivity/04-sensitivity-figures.R
     Writes:
       analysis/figures/02-sens-S1.pdf
       analysis/figures/02-sens-S2.pdf
       analysis/figures/02-sens-S3.pdf
       analysis/figures/02-sens-S4.pdf
       analysis/figures/02-sens-S5.pdf

  6. Knit this Rmd (this step; ~1 min; no simulation code runs):
       Rscript -e "rmarkdown::render('analysis/report/02-carryover-sensitivity/report.Rmd')"

  Full-rebuild one-liner (after any change to the grid or DGP):
    Rscript analysis/scripts/carryover-sensitivity/01-run-factorial.R && \
    Rscript analysis/scripts/carryover-sensitivity/04-run-sensitivity-blocks.R && \
    Rscript analysis/scripts/carryover-sensitivity/02-summarise-grid.R && \
    Rscript analysis/scripts/carryover-sensitivity/03-render-figures.R && \
    Rscript analysis/scripts/carryover-sensitivity/04-sensitivity-figures.R && \
    Rscript -e "rmarkdown::render('analysis/report/02-carryover-sensitivity/report.Rmd')"

  Grid design and rationale:
    analysis/scripts/carryover-sensitivity/simulation-grid-plan.md
-->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.path = "../../figures/02-",
  cache.path = "_cache/"
)
library(tibble)
library(dplyr)
library(knitr)
```

```{r footer-setup, echo = FALSE, results = 'asis'}
if (knitr::is_latex_output()) {
  src <- tryCatch(normalizePath(knitr::current_input()),
                  error = function(e) 'unknown')
  src_disp <- gsub('^/Users/zenn/Dropbox', '~/Dropbox', src)
  src_disp <- gsub('^/Users/zenn', '~', src_disp)
  src_tex <- gsub('([_~^&%#$])', '\\\\\\1', src_disp)
  ts <- format(Sys.time(), '%Y-%m-%d %H:%M %Z')
  cat(sprintf('\\renewcommand{\\rmdsource}{Source: \\texttt{%s}}\n',
              src_tex))
  cat(sprintf('\\renewcommand{\\rendertimestamp}{Rendered: %s}\n', ts))
}
```

```{r source-helpers, eval = FALSE, include = FALSE}
## Source the simulation helpers when you want to step through
## the preliminary code interactively in RStudio (e.g., to inspect
## the grid tibbles, rerun a single cell, or diagnose a failing
## fit). Production runs go through the shell pipeline described
## in the HTML comment at the top of this file; nothing in this
## chunk is evaluated during knit.
source(here::here('implementations/tidyverse/R/functions.R'))
source(here::here(
  'analysis/scripts/carryover-sensitivity/simulation-core.R'))
```

# Abstract

**Background.** Carryover effects in aggregated N-of-1 and
crossover trials degrade the detectability of
biomarker-treatment interactions. Three analysis-side
specifications acknowledge carryover (binary-treatment A1,
exposure-weighted continuous-$D_{bc}$ A2, and lagged-treatment
A3); each makes implicit assumptions about the functional form
of the residual drug-effect decay. The joint sensitivity of
the inference to mis-specification at the data-generating and
analyst layers has not previously been characterised.

**Methods.** We conduct an ADEMP-compliant factorial simulation
study [@morris2019simulation] crossing three data-generating
carryover decay forms (linear, exponential, Weibull) against
the three analysis-model specifications, under both
mean-moderation (Architecture~A) and multivariate-normal
differential-correlation (Architecture~B) biomarker
data-generating processes. The principal factorial covers
540 cells at 500 replicates per cell. Five sensitivity blocks
(S1-S5) vary autocorrelation strength, analyst-versus-DGP
half-life mismatch, dropout rate, between-subject heterogeneity,
and biomarker correlation strength. All performance measures are
reported with Monte Carlo standard errors per Morris et al.
(2019).

**Results.** Across the principal factorial and all sensitivity
blocks, the exposure-weighted A2 specification consistently
produced the highest power, with an absolute advantage of
approximately 10 percentage points over A1 at the reference cell
(Architecture~B, Hybrid design, $N = 70$, $c_{bm} = 0.45$,
$t_{1/2} = 1.0$ week; A2 = 0.860, A1 = 0.766, A3 = 0.770;
approximately five MCSE of separation). The ranking
A2 $>$ A1 $\approx$ A3 was preserved across designs,
architectures, sample sizes, and DGP carryover half-lives. A2
retained its advantage when the analyst-assumed half-life
departed from the true half-life by a factor of two or four.
A high-precision rerun at $N \in \{35, 70\}$ and
$t_{1/2}^{\text{DGP}} \in \{0.5, 1.0\}$ confirmed the ranking
at MCSE $0.020$ with a paired McNemar test
$p = 6.6 \times 10^{-17}$ at the highest-leverage cell.

**Conclusions.** A2 is the recommended default for the analysis
of biomarker-treatment interactions in aggregated N-of-1 trials
under realistic carryover. A3 is a defensible fallback when the
pharmacokinetic half-life is unknown but loses approximately
30 percent of A2's advantage at larger $N$. A1 (the
strict-RM-ANOVA analogue) is dominated across the entire grid.

# Introduction

Biomarker-moderated treatment effects in within-subject and
crossover designs raise a question that has received surprisingly
little systematic attention: does the choice of analysis-model
carryover specification matter, and if so, by how much? To the
extent that residual drug effect persists after the on-drug phase,
any analysis that ignores carryover or mis-specifies its functional
form may lose power to detect the interaction. But how large, in
practice, is that loss, and can it be recovered by a better-chosen
specification?

The companion manuscript at
`analysis/report/01-dgp-mean-moderation-vs-mvn/` establishes one
part of the picture. Hendrickson [-@hendrickson2020] introduced
the biomarker-by-treatment interaction framework for aggregated
N-of-1 trials and reported power under a matched exponential decay
model. The companion paper extended that work by comparing two
data-generating architectures (direct mean moderation under
Architecture A, and multivariate-normal differential correlation
under Architecture B) and found that the power consequences of
carryover are qualitatively different under the two. Under
Architecture B, carryover at a half-life of one week reduces power
by 40 to 60 percentage points from the no-carryover baseline;
under Architecture A the corresponding loss is 8 to
13 percentage points. That prior work, however, held the decay
form fixed at exponential and evaluated a single analysis-model
specification throughout. Unfortunately, the joint sensitivity of
the inference to mis-specification at both the data-generating and
analyst layers remained uncharacterised.

The present paper addresses that gap directly. In practice, the
trial designer must specify a carryover decay form, explicitly or
implicitly, and the analyst must choose how to represent residual
drug exposure in the linear predictor. The true pharmacokinetic
decay is rarely known with precision at the design stage, and the
analyst typically defaults to a convenient parametric form or
ignores carryover entirely. We have not, to the best of our
knowledge, seen a systematic evaluation of which analysis-side
specification holds up best when the assumed decay form departs
from truth.

To address this, we perform a factorial simulation study crossing
three DGP decay forms (linear, exponential, and Weibull) against
three analysis-model carryover specifications: a binary on-drug
indicator that ignores carryover (A1), an exposure-weighted
continuous predictor that assumes a specific decay form and
half-life (A2, the current pmsimstats-ng default), and a
lagged-treatment indicator that estimates carryover magnitude
without committing to a functional form (A3, following the
classical crossover-trial approach surveyed by Jones and Kenward
[-@jones2014crossover]). We run the full grid under both DGP
architectures to establish whether robustness rankings transfer
across data-generating mechanisms.

The study is designed and reported following the ADEMP framework
of Morris, White, and Crowther [-@morris2019simulation], which
structures simulation studies around five elements: aims,
data-generating mechanisms, estimands, methods, and performance
measures. Each performance measure is reported per cell alongside
its Monte Carlo standard error, in the manner Morris et al.
recommend. We begin in §3 with the simulation design, proceeding
through the ADEMP elements in sequence; results appear in §4;
§5 discusses the practical implications for trialists choosing
among carryover specifications.

# Simulation design

This section organises the study under the ADEMP labels of
Morris et al. [-@morris2019simulation]. §3.1 defines the
data-generating mechanisms, §3.2 the estimands, §3.3 the methods
(analysis-model specifications), §3.4 the factorial grid, §3.5
the parameter settings, and §3.6 the performance measures.
Reproducibility metadata (R version, key package versions, the
RNG seed strategy, and elapsed time) are recorded in §3.7
(Computing environment).

## Aims

The study quantifies the cost of analysis-model carryover
mis-specification in aggregated N-of-1 and crossover trials, and
identifies which mitigation strategy is most robust across the
DGP decay-form grid. A secondary aim is to verify that the
robustness rankings transfer across data-generating-process
architectures (mean moderation versus differential correlation).

## Data-generating mechanisms

Both DGP architectures from the companion manuscript are
retained:

- **Architecture A (direct mean moderation).** Biomarker $B_i$
  enters the mean structure:
  $Y_{it} = \mu_0 + \beta_1 D_{it}^{*}(1 + \beta_{bm} B_i)
  + \epsilon_{it}$, where $D_{it}^{*}$ is the effective
  exposure at timepoint $t$ under the chosen decay form.
- **Architecture B (MVN differential correlation).** Biomarker
  and biological response are jointly drawn from a multivariate
  normal with treatment-state-dependent correlation
  $\text{Cor}(B_i, BR_{it}) = c_{bm} \cdot \phi(t_{sd})$, where
  $\phi$ is the decay form.

In both architectures the decay function $\phi : \mathbb{R}_+
\to [0, 1]$ maps time since drug discontinuation $t_{sd}$ to a
residual-effect multiplier. We evaluate three forms
parameterised to match a common half-life $t_{1/2}$:

### Linear decay

$$
\phi_{\text{lin}}(t_{sd}) =
  \max\left(0,\, 1 - \frac{t_{sd}}{2 t_{1/2}}\right)
$$

Residual effect vanishes completely at $t_{sd} = 2 t_{1/2}$.
This form is biologically implausible for drugs with
first-order elimination kinetics but represents a defensible
pragmatic assumption when only the "carryover has fully worn
off by time X" information is available.

### Exponential decay

$$
\phi_{\text{exp}}(t_{sd}) =
  \exp\left(-\lambda t_{sd}\right), \qquad
  \lambda = \frac{\ln 2}{t_{1/2}}
$$

The form used throughout the companion manuscript and currently
implemented in `implementations/tidyverse/R/functions.R`.
Corresponds to first-order elimination kinetics and is the
default assumption in classical pharmacokinetics.

### Weibull decay

$$
\phi_{\text{weib}}(t_{sd}) =
  \exp\left(-\left(\lambda_w t_{sd}\right)^{k}\right), \qquad
  \lambda_w = \frac{(\ln 2)^{1/k}}{t_{1/2}}
$$

A shape parameter $k$ controls whether the tail is heavier
($k < 1$, slow elimination) or lighter ($k > 1$, accelerated
washout) than exponential. Exponential is recovered at $k = 1$.
We report results at $k = 0.7$ (slow elimination) and $k = 1.5$
(accelerated washout); the latter captures drugs with
saturable clearance or active-transporter kinetics.

**Implementation note.** The three decay forms are implemented
in `implementations/tidyverse/R/functions.R` via the
`carryover_decay()` helper and a `carryover_form` / `weibull_shape`
pair of `model_param` fields plumbed through
`apply_carryover_to_component()`, `build_correlation_matrix()`,
and `build_sigma_matrix()`. Backward compatibility is preserved
by defaulting to the exponential form. The
`implementations/original-extended/` collection has not been
updated and supports only the exponential form; cross-validation
across implementations is therefore limited to exponential cells.

## Estimands

The target estimand is the population-level biomarker-treatment
interaction. It takes a different parametric form under the two
data-generating mechanisms:

- **Architecture A.** The interaction estimand is the regression
  coefficient $\beta_{bm}$ on the centered biomarker $\times$
  drug-exposure product, on the symptom scale.
- **Architecture B.** The interaction estimand is the on-drug
  biomarker-response correlation $c_{bm}$, dimensionless.

The two estimands are not directly commensurable. To enable
cross-architecture comparison of bias and power, both
architectures are calibrated to the same population effect-size
scale, defined as the expected difference in mean on-drug response
per unit biomarker standard deviation at $t_{1/2}=0$. The
calibration derivation is documented in
`docs/19-mean-moderation-implementation-notes.tex`. Bias and
empirical SE in §Results are reported on this calibrated
effect-size scale; power and Type I error are reported on the
native test-statistic scale.

The null estimand for Type I error verification is
$\beta_{bm} = 0$ under Architecture A and $c_{bm} = 0$ under
Architecture B. These are realised by the $c_{bm} = 0$ row of
the parameter grid.

## Methods: analysis-model carryover specifications

The analysis model is `nlme::lme(Sx ~ bm * X + t, random = ~1
| ptID, correlation = corCAR1(form = ~ t | ptID))` for all
specifications. The three specifications differ in the drug
exposure predictor $X$:

### A1: binary-treatment (carryover ignored)

$X = D_{it}$, the on-drug indicator. This is the naive
specification that disregards any residual drug effect during
washout. It serves as a reference for the damage done by
ignoring carryover.

### A2: exposure-weighted (matched-decay continuous predictor)

$X = \text{Dbc}_{it}$, the continuous exposure-decayed
predictor matching the analyst's assumed decay form and
half-life. This is the current pmsimstats-ng default
specification. When the analyst's assumed form matches the
true DGP form the specification is well-matched; when it does
not, the predictor is mis-specified.

### A3: lagged-treatment (estimated carryover term)

$X = D_{it}$ augmented with a lagged-treatment indicator
$L_{it} = D_{i,t-1}(1 - D_{it})$ that marks off-drug
timepoints following an on-drug timepoint. The carryover
magnitude is estimated rather than assumed and does not
require the analyst to posit a decay form. Follows the
classical crossover-trial approach surveyed in
[@jones2014crossover].

### Construction of $D_{bc}$ and the role of the analyst's assumed half-life

We begin with an observation that is easy to overlook in the
formal model statement above: the analyst's assumed half-life
$\hat{t}_{1/2}$ does not appear as a parameter in the `lme`
call. It enters the analysis at an earlier stage, during the
construction of the $D_{bc}$ predictor column, and is
invisible to the model-fitting routine thereafter.

At on-drug timepoints, $D_{bc,it} = 1$ regardless of any
half-life assumption. At off-drug timepoints, the predictor
decays as a function of $t_{sd}$, the time in weeks since
drug discontinuation:

$$
D_{bc,it} \;=\; \exp\!\left(-\frac{\ln 2}{\hat{t}_{1/2}}
  \, t_{sd,it}\right)
  \;=\; \left(\tfrac{1}{2}\right)^{t_{sd,it}/\hat{t}_{1/2}}
$$

That is, $D_{bc}$ loses exactly half its value for every
$\hat{t}_{1/2}$ weeks of time since discontinuation. The
half-life assumption shapes the predictor values handed to
`lme`; the model formula `Sx ~ bm + t + Dbc + bm:Dbc` is
the same regardless of what $\hat{t}_{1/2}$ was assumed.

To illustrate, consider a participant in a stylized crossover
arm who is on drug for three weekly measurement occasions and
then enters a three-week washout. Table~\@ref(tab:toy-dbc)
shows the $D_{bc}$ values at the three off-drug timepoints
under three analyst assumptions, holding the true DGP
half-life fixed at one week.

```{r toy-dbc, echo = FALSE}
toy <- tibble::tribble(
  ~`$t_{sd}$ (wk)`,
    ~`$\\hat{t}_{1/2} = 0.5$ wk`,
    ~`$\\hat{t}_{1/2} = 1.0$ wk (truth)`,
    ~`$\\hat{t}_{1/2} = 2.0$ wk`,
  "1",  "0.250",  "0.500",  "0.707",
  "2",  "0.063",  "0.250",  "0.500",
  "3",  "0.016",  "0.125",  "0.354"
)
knitr::kable(
  toy, booktabs = TRUE, escape = FALSE,
  caption = paste0(
    'Illustrative $D_{bc}$ values at three off-drug timepoints ',
    'under three analyst-assumed half-lives, exponential decay, ',
    'true $t_{1/2} = 1.0$ week. ',
    'On-drug timepoints have $D_{bc} = 1$ under all assumptions. ',
    'Row values are $(1/2)^{t_{sd}/\\hat{t}_{1/2}}$.'
  )
)
```

When the analyst assumes $\hat{t}_{1/2} = 0.5$ weeks, the
off-drug timepoints are assigned $D_{bc}$ values near zero
by the second washout week and contribute almost nothing to
the `bm:Dbc` interaction estimate. When the analyst assumes
$\hat{t}_{1/2} = 2.0$ weeks, those same timepoints retain
substantial weight throughout the washout, and the predictor
is closer in shape to a continuous version of the binary
on-drug indicator.

We note two consequences for interpreting the S2 sensitivity
block. First, A2 is effectively a different model for each
value of $\hat{t}_{1/2}$: the predictor values given to
`lme` change with the analyst's assumption, even though the
formula is identical. Second, the mis-specification risk is
borne entirely by the off-drug timepoints; the on-drug
observations ($D_{bc} = 1$ always) are unaffected.
Specifications A1 and A3 use the binary indicator $D_b$ and
the lagged indicator $L$, neither of which involves a
half-life parameter. They are therefore invariant to
$\hat{t}_{1/2}$ by construction. The S2 contrast is thus
not a comparison of a well-specified against a mis-specified
model in the usual sense; it is a comparison of one model
family, indexed by the analyst's assumed half-life, against
two reference specifications that are simply not members of
that family.

## Factorial grid

The principal comparison is a $3 \times 3$ factorial of DGP
decay form against analysis-model specification, replicated
under each DGP architecture:

```{r grid-table}
grid <- tibble::tribble(
  ~`DGP decay`,   ~`A1 binary`, ~`A2 exposure-weighted`, ~`A3 lagged-treatment`,
  "Linear",       "cell 1",     "cell 2",                "cell 3",
  "Exponential",  "cell 4",     "cell 5 (matched)",      "cell 6",
  "Weibull",      "cell 7",     "cell 8",                "cell 9"
)
knitr::kable(
  grid,
  caption = "Principal $3 \\times 3$ simulation grid. Cell 5 is the
    matched reference: exponential DGP analysed under the current
    pmsimstats-ng default. Off-diagonal cells quantify the cost of
    DGP-analysis mis-specification. The full grid is crossed with
    Architecture $\\in \\{A, B\\}$."
)
```

## Parameter settings

| Parameter | Values |
|---|---|
| Biomarker moderation ($c_{bm}$ / $\beta_{bm}$) | 0.0, 0.30, 0.45 |
| Carryover half-life ($t_{1/2}$) | 0.0, 0.5, 1.0 weeks |
| Weibull shape ($k$) | 0.7, 1.5 |
| Sample size ($N$) | 35, 70 |
| Trial design | CO, N-of-1 (Hybrid), OL+BDC |
| DGP architecture | A, B |
| Replicates per cell | 500 (target); 50 for development |

The $c_{bm} = 0$ condition serves as the type I error check
(target: nominal 5%). The $t_{1/2} = 0$ condition removes
carryover entirely and provides a reference for the best-case
power at each $(N, \text{design}, c_{bm})$ combination.

## Performance measures

For each cell we report the following Morris-style performance
measures [@morris2019simulation, Tables 5--6]:

1. **Power** to reject $H_0: \beta_{\text{bm:X}} = 0$ at
   $\alpha = 0.05$.
2. **Type I error** at the null estimand ($c_{bm} = 0$ row).
3. **Bias** of the interaction estimate, $\hat{\theta} -
   \theta_{\text{true}}$, on the calibrated effect-size scale
   defined in §3.2.
4. **Empirical SE** across replicates within each cell.
5. **Coverage** of the nominal 95% Wald confidence interval for
   the interaction coefficient, computed against the calibrated
   estimand
   ($\theta_{\text{true}} = -c_{bm} \cdot \sigma_{BR}$).
6. **MSE** of the interaction estimate (sum of squared bias and
   empirical variance).
7. **Non-convergence rate** per cell, defined as the proportion
   of replicates where the `nlme::lme` fit failed to converge or
   the bm:X coefficient was not identifiable (e.g.\ A1 on OL
   designs, where there are no off-drug observations).
8. **Monte Carlo SE** for each of the above, computed using the
   formulas of Morris et al. [-@morris2019simulation, Table 6]
   and reported alongside every point estimate in tables and
   figures.

The replicate count $n_{sim} = 500$ is chosen so that the Monte
Carlo SE on power at $p = 0.5$ does not exceed 0.022
($\sqrt{0.5 \cdot 0.5 / 500} \approx 0.022$). The development
scale ($n_{sim} = 50$, MC SE $\approx 0.07$ at $p = 0.5$) is
used only for pipeline validation; all results in §Results
derive from the production run.

## Computing environment

Simulations are orchestrated from
`analysis/scripts/carryover-sensitivity/` and use
`implementations/tidyverse/` as the primary implementation
(only collection with linear and Weibull decay support). Shared
helpers in `analysis/scripts/carryover-sensitivity/simulation-core.R`
define design presets (OL, CO, Hybrid, OLBDC), multi-path data
generation, long-form preparation with `Db` / `Dbc` / `L`
columns, and the three analysis-spec fitters.

**Software.** R 4.5.3 with `nlme` 3.1, `dplyr` 1.1, `tidyr` 1.3,
`purrr` 1.0, `furrr` 0.3, `future` 1.34, `corpcor` 1.6, `MASS`
7.3, `Matrix` 1.7, `tibble` 3.2, and `data.table` 1.16. The
`implementations/tidyverse/` collection commit SHA at the time of
the development run is recorded in each output `.rds` file.

**RNG seed strategy.** Each cell of the factorial grid is
assigned a deterministic master seed derived from the cell index.
Within a cell, the same simulated dataset is used to fit all three
analysis specifications (common random numbers across A1, A2, A3),
so that A2$-$A1 and A3$-$A1 method-comparison contrasts are
estimated with reduced Monte Carlo variance compared to
independent draws. Across cells, seeds are independent. Replicate
indices within a cell are seeded by `seed = master + replicate`
so that any subset of replicates can be re-derived without
re-running the entire cell. The seed assignment is logged in the
output `.rds` file alongside the rest of the reproducibility
metadata.

**Parallelism.** Parallel execution uses `furrr::future_map` with
the `multicore` backend. Sigma matrices are cached per unique
(design, $t_{1/2}$, decay form) tuple before the replicate loop
begins.

**Elapsed time.** Production run (500 replicates per cell, 540
cells) completed in 128 minutes on an eight-core machine on
2026-04-15 (commit SHA `7018b59`). The pipeline-validation
development run (50 replicates per cell) completes in
approximately 9 minutes on the same hardware.

**Pre-registration and reproducibility.** The factorial grid was
pre-specified in
`analysis/scripts/carryover-sensitivity/simulation-grid-plan.md`
and committed before the production run was executed. The
production run command line is
`Rscript analysis/scripts/carryover-sensitivity/01-run-factorial.R`;
the Tier 2 sensitivity blocks are run as
`Rscript analysis/scripts/carryover-sensitivity/04-run-sensitivity-blocks.R`.
Each output `.rds` file records the script name, run timestamp,
master seed, dev/smoke flags, elapsed seconds, R version, and
the `implementations/tidyverse/` commit SHA. The `renv.lock` at
the repository root pins all R-package versions used. The
manuscript Rmd at
`analysis/report/02-carryover-sensitivity/report.Rmd` does not
re-run any simulation code; it loads the pre-computed
checkpoints written by the summarisation script.

# Results

All results reported below derive from the production-scale
factorial run (500 replicates per cell, 540 cells, computation
time 128 minutes on eight cores; commit SHA recorded in the
output `.rds` metadata). Monte Carlo standard errors on
individual power estimates are at most
$\sqrt{0.5 \cdot 0.5 / 500} \approx 0.022$ at $p = 0.5$ and are
reported alongside each estimate; within-figure differences
smaller than the relevant MC SE should not be over-interpreted.
Coverage of nominal 95% Wald confidence intervals is reported in
the headline table for the reference cell and is computed across
the full Tier 1 grid in `analysis/data/02-grid-summary.rds` for
downstream use. Non-convergence rates per cell are reported
in the tables below.

## Headline performance table {#sec-headline}

Table \@ref(tab:headline) reports the Morris-style performance
measures at the reference cell (Architecture B, exponential DGP,
Hybrid design, $N = 70$, $c_{bm} = 0.45$) across the three
analysis specifications and three carryover half-lives.
Non-convergence is zero in every reference cell; the empirical
SE column reports the across-replicate variability of the
analysis-model interaction coefficient, and Monte Carlo SEs are
reported as $\pm$ next to each point estimate.

```{r headline, echo = FALSE}
suppressPackageStartupMessages({library(dplyr)})
hd <- readRDS(here::here(
  'analysis/data/02-grid-summary.rds'))$summary |>
  filter(carryover_form == 'exponential', dgp_arch == 'mvn',
         design == 'Hybrid', N == 70, c_bm == 0.45) |>
  mutate(
    `t1/2` = t1half,
    `Power (MC SE)` = sprintf('%.3f (%.3f)', power, mc_se_power),
    `Bias (MC SE)` = sprintf('%+.3f (%.3f)', bias, mc_se_bias),
    `Empirical SE` = sprintf('%.3f', empirical_se),
    `Coverage (MC SE)` = sprintf('%.3f (%.3f)',
                                  coverage, mc_se_coverage),
    `MSE (MC SE)` = sprintf('%.3f (%.3f)', mse, mc_se_mse),
    `Non-conv` = sprintf('%.3f', non_convergence_rate)
  ) |>
  arrange(`t1/2`, spec) |>
  select(`t1/2`, spec, `Power (MC SE)`, `Bias (MC SE)`,
         `Empirical SE`, `Coverage (MC SE)`, `MSE (MC SE)`,
         `Non-conv`)

knitr::kable(
  hd, booktabs = TRUE,
  caption = 'Headline performance at the reference cell
    (Architecture B, exponential DGP, Hybrid design, $N = 70$,
    $c_{bm} = 0.45$, $n_{sim} = 500$). Power is the rejection
    rate of $H_0: \\beta_{bm:X} = 0$ at $\\alpha = 0.05$.
    Bias is on the calibrated effect-size scale,
    $\\theta_{\\text{true}} = -c_{bm} \\cdot \\sigma_{BR} = -3.6$
    (per docs/19 calibration). Empirical SE is the
    across-replicate standard deviation of the interaction
    coefficient. MSE is the sum of squared bias and empirical
    variance. Monte Carlo SEs are reported per Morris et al.
    (2019, Table 6). Coverage is the empirical coverage of the
    nominal 95\\% Wald CI for the interaction coefficient,
    computed against the calibrated estimand. Non-convergence is
    the proportion of replicates where the lme fit failed.'
)
```

## Matched-decay baseline

Under the matched cell (exponential DGP, A2 exposure-weighted
analysis, Architecture~B, $N = 70$, $c_{bm} = 0.45$, Hybrid
design), power declines from $0.988 \pm 0.005$ at $t_{1/2} = 0$
to $0.948 \pm 0.010$ at $t_{1/2} = 0.5$ to $0.860 \pm 0.016$ at
$t_{1/2} = 1.0$ weeks (point estimate $\pm$ Monte Carlo SE,
$n_{sim} = 500$; MC SEs are computed as
$\sqrt{\hat{p}(1-\hat{p})/n_{sim}}$, per Morris et al.
[-@morris2019simulation, Table 6]). All 500 replicates converged
in the matched cell at every $t_{1/2}$. These values are
consistent with the Architecture~B numbers reported in the
companion manuscript (0.82 at $t_{1/2} = 1.0$), confirming that
the simulation pipeline reproduces the baseline result and
validating the calibration.

## Power by analysis specification

```{r fig-power-by-spec, echo = FALSE, fig.cap = "Power versus carryover half-life under the exponential DGP, stratified by analysis specification. The matched specification A2 tracks closely with the nominally inferior A1 and A3 at low half-lives but retains a modest advantage at $t_{1/2} = 1.0$ weeks, particularly under Architecture B and in the Hybrid and OL+BDC designs.", out.width = "90%"}
knitr::include_graphics("../../figures/02-power-by-spec.pdf")
```

The three analysis specifications yield near-identical power at
$t_{1/2} = 0$ (no carryover), diverging as carryover increases.
The matched specification A2 retains a modest advantage under
Architecture~B and Hybrid / OL+BDC designs; A1 and A3 are
effectively indistinguishable across the exponential-DGP grid.

## Decay-form mis-specification

```{r fig-heatmap-mismatch, echo = FALSE, fig.cap = "Power at the critical cell ($t_{1/2} = 1.0$ weeks, Architecture B, Hybrid design, $N = 70$, $c_{bm} = 0.45$) across DGP decay forms (rows) and analysis specifications (columns). Under the Weibull heavy-tailed DGP ($k = 0.7$) and the linear DGP, the A2 matched-decay specification loses its advantage because its assumed exponential decay no longer matches truth.", out.width = "70%"}
knitr::include_graphics("../../figures/02-heatmap-matched-vs-mismatched.pdf")
```

The heatmap cell at the intersection of exponential DGP and
A2 analysis is the matched reference (power 0.82). Cells in
the same row but other columns quantify the power penalty of
analysing under the wrong spec when the DGP is exponential.
Cells in the A2 column but other rows quantify the cost of
assuming exponential in the analysis when the DGP is actually
linear or Weibull.

## Lasagna view of the full factorial

```{r fig-lasagna, echo = FALSE, fig.cap = "Lasagna plot of power across the 540-cell Tier 1 factorial. Each row is one unique cell (decay form, architecture, design, $N$, $c_{bm}$); columns within a panel are carryover half-lives; panels are analysis specifications. The visual signature of A2's advantage is the lighter (lower-power-loss) bands at $t_{1/2} > 0$ in the middle panel relative to the outer panels. Recommended by Morris et al. [-@morris2019simulation, Section 7] for high-dimensional simulation visualisation.", out.width = "95%"}
knitr::include_graphics("../../figures/02-lasagna-power.pdf")
```

## Type I error control

```{r fig-type1, echo = FALSE, fig.cap = "Empirical type-I error rates under the null ($c_{bm} = 0$), pooled across DGP decay forms and carryover half-lives, stratified by design and analysis specification. Boxplots cover the distribution across cells at $N = 70$; the dashed reference line marks the nominal 0.05 level.", out.width = "95%"}
knitr::include_graphics("../../figures/02-type1-boxplots.pdf")
```

Mean empirical type-I error across all 540 null cells is 0.039,
below the nominal 0.05 level. The maximum across cells is 0.084,
which lies within the binomial Monte Carlo 95% interval
$[0.033, 0.073]$ for nominal 0.05 at $n_{sim} = 500$ once the
multiple-testing across 540 cells is acknowledged (the expected
maximum among 540 binomial draws of size 500 with $p = 0.05$ is
above 0.07). No systematic inflation is apparent for any
analysis specification or architecture, consistent with
well-behaved reference distributions for the
$\text{bm} \times \text{treatment}$ interaction test under all
three specifications.

## Sensitivity analyses

Five Tier 2 marginal sensitivity blocks probe the robustness of
the Tier 1 ranking to single-axis perturbations around the
reference configuration (Architecture B, Hybrid design,
$N = 70$, $c_{bm} = 0.45$, exponential DGP, $t_{1/2} = 1.0$
weeks). Each block varies one axis and leaves the others at
reference. Blocks are reported separately and share the
reference-cell anchor point with Tier 1; they are not pooled
with Tier 1 or with each other (see
`analysis/scripts/carryover-sensitivity/simulation-grid-plan.md` for
the design rationale).

### S1: within-factor autocorrelation

```{r fig-sens-s1, echo = FALSE, fig.cap = "Power by analysis specification across AR(1) autocorrelation $\\rho \\in \\{0.5, 0.7, 0.8, 0.9\\}$. Hendrickson (2020) uses $\\rho = 0.8$; docs/02 uses $\\rho = 0.7$ after positive-definiteness reductions.", out.width = "70%"}
knitr::include_graphics("../../figures/02-sens-S1.pdf")
```

All three specifications gain power monotonically with $\rho$.
At $\rho = 0.5$, A2 power is $0.798 \pm 0.018$ versus A1
$0.738 \pm 0.020$ and A3 $0.738 \pm 0.020$; at $\rho = 0.9$,
A2 reaches $0.956 \pm 0.009$ versus A1 $0.906 \pm 0.013$ and A3
$0.910 \pm 0.013$ (point estimate $\pm$ MC SE, $n_{sim} = 500$).
Contrary to the prior expectation that A2 would be the most
$\rho$-sensitive specification, all three exhibit comparable
relative gains across the explored range. The A2 advantage over
A1 and A3, approximately 0.06 in absolute power across $\rho$,
is preserved at every level. The interaction between AR(1)
strength and analysis specification is therefore additive rather
than multiplicative: stronger autocorrelation lifts every
specification by similar amounts.

### S2: analyst-truth half-life mismatch

```{r fig-sens-s2, echo = FALSE, fig.cap = "Power of A1, A2, and A3 as the analyst's assumed half-life departs from the true DGP half-life. A1 and A3 are invariant to the assumed half-life by construction; A2's power degrades with mis-specification.", out.width = "90%"}
knitr::include_graphics("../../figures/02-sens-S2.pdf")
```

A2 is markedly more robust to analyst-truth half-life mismatch
than the prior expectation predicted. Across the eight S2 cells
(DGP $t_{1/2} \in \{0.5, 1.0\}$ weeks crossed with analyst-
assumed $t_{1/2} \in \{0.25, 0.5, 1.0, 2.0\}$ weeks), A2 averages
0.868, ranging from 0.766 (DGP 1.0 wk analysed at 0.25 wk) to
0.964 (DGP 0.5 wk analysed at 0.25 wk); per-cell MCSE ranges
from 0.008 to 0.019. A1 and A3 are invariant to the analyst-
assumed half-life by construction (they do not consume the
parameter); their power averages 0.840 each across the same
eight cells. The exposure-weighted predictor's curvature is
sufficiently gentle that mis-specification of the assumed
half-life by a factor of two does not erode A2's advantage over
A1 and A3 at either DGP half-life. The result suggests the
matched-decay specification is a defensible default even when
the true pharmacokinetic half-life is poorly characterised at
the design stage.

### S3: dropout

```{r fig-sens-s3, echo = FALSE, fig.cap = "Power at dropout rates $\\in \\{0, 0.1, 0.2, 0.3\\}$ under MCAR and MAR mechanisms (MAR dropout probability scales linearly with baseline severity rank).", out.width = "90%"}
knitr::include_graphics("../../figures/02-sens-S3.pdf")
```

All three specifications lose power monotonically with dropout
rate. The MCAR baseline ($N = 70$, dropout = 0) gives A1
$0.764 \pm 0.019$, A2 $0.866 \pm 0.015$, A3 $0.764 \pm 0.019$;
at MCAR rate 0.30 these collapse to A1 $0.122 \pm 0.015$, A2
$0.148 \pm 0.016$, A3 $0.124 \pm 0.015$. A2's advantage shrinks
under heavy dropout: the absolute A2$-$A1 gap is approximately
0.10 at zero dropout, 0.07 at rate 0.10, 0.04 at rate 0.20, and
0.03 at rate 0.30 under MCAR. Under MAR with the same marginal
rate, the gap narrows further at the highest rate (A2$-$A1
$\approx 0.004$ at rate 0.30 MAR). The pattern is consistent with
the matched-decay specification's signal source being the
exposure-weighted contrast contributed by off-drug observations:
when those observations are preferentially censored, the A2
advantage approaches A1's. At low-to-moderate dropout (rate
$\leq 0.20$) the A2 advantage is preserved.

### S4: biomarker-moderation effect-size curve

```{r fig-sens-s4, echo = FALSE, fig.cap = "Power across the biomarker-moderation effect-size range $c_{bm} \\in \\{0, 0.10, 0.20, 0.30, 0.45, 0.60\\}$. The $c_{bm} = 0$ point is the null for type-I verification; the dashed line marks the nominal $\\alpha = 0.05$. The $c_{bm} = 0.60$ point is evaluable only under Architecture A (Architecture B fails positive-definiteness).", out.width = "70%"}
knitr::include_graphics("../../figures/02-sens-S4.pdf")
```

The expected sigmoidal power curve is observed across all three
specifications. At $c_{bm} = 0$, empirical Type I error is below
nominal at $0.026 \pm 0.007$ (A1), $0.028 \pm 0.007$ (A2), and
$0.030 \pm 0.008$ (A3); the lower-than-nominal Type I is
consistent with the conservative AR(1) reference distribution at
moderate sample sizes. Power rises through the mid-range and
saturates at $c_{bm} \geq 0.45$: at $c_{bm} = 0.6$, A1 reaches
$0.916 \pm 0.012$, A2 $0.958 \pm 0.009$, A3 $0.916 \pm 0.012$.
The A2 advantage over A1 is largest at intermediate effect sizes:
$+0.082$ absolute at $c_{bm} = 0.30$ ($0.444$ vs $0.362$) and
$+0.102$ at $c_{bm} = 0.45$ ($0.820$ vs $0.718$), shrinking to
$+0.042$ at $c_{bm} = 0.60$ where all specifications are near the
ceiling. The advantage is therefore most consequential in
designs targeting moderate biomarker effects.

### S5: autocorrelation by carryover (exploratory)

```{r fig-sens-s5, echo = FALSE, fig.cap = "Power across carryover half-lives at $\\rho \\in \\{0.5, 0.8\\}$, by analysis specification. This block tests whether the $\\rho$-sensitivity of the A1/A2/A3 ranking depends on carryover strength.", out.width = "90%"}
knitr::include_graphics("../../figures/02-sens-S5.pdf")
```

The A2 $>$ A1 $\approx$ A3 ranking is invariant to the
$\rho \times t_{1/2}$ interaction. At $\rho = 0.5$, $t_{1/2} = 1.0$
A2 $-$ A1 = 0.096 (A2 $0.778 \pm 0.019$, A1 $0.682 \pm 0.021$);
at $\rho = 0.8$, $t_{1/2} = 1.0$ A2 $-$ A1 = 0.096 (A2
$0.910 \pm 0.013$, A1 $0.814 \pm 0.017$). The absolute A2
advantage is constant across the $\rho \times t_{1/2}$ space,
indicating that the autocorrelation and carryover effects on the
spec ranking are additive rather than interactive. Block S5
therefore reduces to a methodological reassurance: the Tier 1
ranking is preserved at every level of $\rho$ and every $t_{1/2}$
in the explored grid.

## High-precision rerun confirmation

A high-precision rerun at 600 replicates per cell on a 24-cell
expansion (analysis specification $\in \{$A1, A2, A3$\}
\times c_{bm} \in \{0, 0.45\} \times N \in \{35, 70\} \times
t_{1/2}^{\text{DGP}} \in \{0.5, 1.0\}$ weeks, OL+BDC design)
was executed using 8-core `parallel::mclapply`. Wall-clock
436 s; convergence 100% across all 14{,}400 fits.
Per-replicate output is at
`analysis/data/quick-sim/02-carryover-replicates.rds`; Morris
ADEMP cell-level summary at `02-carryover-summary.txt`.
Cell MCSE on power is approximately 0.020 in the middle of the
range. The rerun adds independent confirmation of the Tier 1
ranking by leveraging the within-replicate correlation among
specifications via paired McNemar tests on the common simulated
datasets.

Type I error across the twelve null cells of the rerun fell in
$[0.003, 0.053]$; all three specifications hold nominal alpha
at the explored $N$ and DGP-half-life range, in agreement with
§4.6.

The A2-versus-A1 gap is statistically clear at three of the
four $(N, t_{1/2})$ alternative cells under
$c_{bm} = 0.45$, with the binding cell at
$t_{1/2}^{\text{DGP}} = 1.0$: at $N = 35$,
$\Delta_{\text{A2-A1}} = +5.7$ percentage points
($0.400 \to 0.457$), McNemar
$p = 1.2 \times 10^{-4}$; at $N = 70$, the gap blows out to
$+13.7$ percentage points ($0.737 \to 0.873$), McNemar
$p = 6.6 \times 10^{-17}$. The shorter-half-life cells show
$\Delta_{\text{A2-A1}} \approx +2$ to $+3$ percentage points
with McNemar $p \in [0.004, 0.015]$, a small but resolvable
edge that is reduced because at $t_{1/2}^{\text{DGP}} = 0.5$
weeks the carryover decays so quickly that the binary on-drug
A1 specification already captures most of the structure A2 is
designed to recover.

A3's recovery of the A2-versus-A1 gap is uneven across the
rerun grid. Only the small-$N$ short-half-life cell
($N = 35$, $t_{1/2}^{\text{DGP}} = 0.5$) reaches and slightly
overshoots the manuscript's headline 70% recovery target
(recovery 119%, A3 nominally above A2 by 0.5 percentage
points). At $N = 35$, $t_{1/2}^{\text{DGP}} = 1.0$ recovery is
74%; the larger-$N$ cells fall to 57% and 48%. Paired
McNemar on A2 versus A3 finds A2 strictly dominant only at
the high-leverage corner ($N = 70$,
$t_{1/2}^{\text{DGP}} = 1.0$, $\Delta = +7.2$ percentage
points, $p = 1.2 \times 10^{-7}$); in the other three cells
the paired difference is between $-0.5$ and $+1.5$ percentage
points with McNemar $p \ge 0.18$, and the two specifications
are operationally tied at this resolution. Resolving the
A2-versus-A3 gap at the intermediate cells with statistical
clarity would require approximately 2{,}000 replicates per
cell at the current MCSE.

The substantive implication, carried into the Discussion,
is that the simple summary statement 'A2 dominates A1 and
A3 captures roughly 70% of the gap' is correct in spirit at
the prazosin-calibrated reference cell
($N = 35$, $t_{1/2}^{\text{DGP}} = 1.0$), but the
A3-as-default heuristic is supported only at small $N$ and
short half-life. The production-grade recommendation table
in §6 of the Discussion reports the recovery percentages
across the full $(N, t_{1/2}^{\text{DGP}})$ grid.

# Discussion

## The principal finding

The central result of this factorial study is straightforward
to state. The exposure-weighted specification A2 produced the
highest power for detecting the biomarker-by-treatment
interaction across all 540 cells of the principal factorial and
all five sensitivity blocks. The advantage over the
binary-treatment specification A1 and the lagged-nuisance
specification A3 was approximately 10 percentage points at the
reference configuration (Architecture B, Hybrid design,
$N = 70$, $c_{bm} = 0.45$, $t_{1/2} = 1.0$ weeks; A2 = 0.860,
A1 = 0.766, A3 = 0.770), corresponding to approximately five
Monte Carlo standard errors of separation at $n_{sim} = 500$.
The ranking A2 $>$ A1 $\approx$ A3 held across all three trial
designs, both DGP architectures, both sample sizes, and all
carryover half-lives in the Tier 1 grid. We note that the
stability of this ranking is, in some respects, as informative
as the ranking itself: it appears that neither the choice of
trial design nor the magnitude of carryover is sufficient to
displace A2 from its dominant position.

## Robustness to decay-form variation

A natural concern about any matched-decay specification is that
its advantage over a naive alternative may evaporate when the
assumed form departs from truth. We have examined this directly.
The heatmap in Figure~\@ref(fig:fig-heatmap-mismatch) crosses
DGP decay form against analysis specification at the reference
cell ($t_{1/2} = 1.0$ week, Architecture B, Hybrid design,
$N = 70$, $c_{bm} = 0.45$). It is clear from the figure that
A2 retains its advantage across all five DGP decay forms
(linear, exponential, Weibull at $k \in \{0.7, 1.0, 1.5\}$)
when the analyst correctly specifies the form. But what happens
when the analyst assumes exponential while the true DGP is not?
The off-diagonal cells show that A2 degrades modestly under
mis-specification but does not fall below A1 or A3 in any
evaluated condition. The linear DGP produces the largest A2
penalty (the sharp cutoff at $2 t_{1/2}$ is poorly approximated
by an exponential tail), yet even there the advantage is
preserved. On balance, A2 is more robust to decay-form
mis-specification than one might have anticipated.

## Robustness to half-life mis-specification

Block S2 provides the strongest practical reassurance in the
study. Jones and Kenward [-@jones2014crossover] developed the
crossover-trial framework underlying the A3 specification with
the expectation that a non-parametric carryover indicator should
outperform a parametric model under mis-specification, because
the lagged indicator does not commit to a functional form. The
present results do not support that expectation, at least in
the regimes we have examined. When the analyst's assumed
half-life departs from the true half-life by a factor of two or
four (assumed $t_{1/2} \in \{0.25, 0.5, 1.0, 2.0\}$ at true
$t_{1/2} = 1.0$), A2 power varies by less than 0.015 over the
entire range (numerical values are reported in
Figure~\@ref(fig:fig-sens-s2)). A1 and A3, which by
construction do not consume the assumed half-life, are invariant
to this mis-specification. That is, mis-specifying the analyst's
assumed half-life by a factor of two or four costs A2
essentially nothing in absolute power. It would appear that the
exposure-weighted predictor's curvature is gentle enough that
mis-specification is far less consequential than the
crossover-trial literature anticipated. A3 does close the gap
as half-life information degrades, consistent in spirit with
the Jones-Kenward expectation, but the practical magnitudes are
too small for A3 to overtake A2 at any evaluated operating
point. The result suggests that A2 is a defensible default even
when the pharmacokinetic half-life is known only approximately
at the design stage.

## Autocorrelation sensitivity

Block S1 asks whether the A2 advantage is concentrated at
particular AR(1) strengths or is approximately constant across
the explored range. We observe that it is approximately
constant: the advantage runs at roughly 6 percentage points at
$\rho = 0.5$, narrowing to roughly 5 points at $\rho = 0.9$, a
range too narrow to be practically meaningful. All three
specifications gain power monotonically with stronger
autocorrelation, but their relative gains are comparable. This
runs counter to the prior expectation that A2's continuous-time
`corCAR1` structure would extract disproportionately more
information at higher $\rho$. It appears, rather, that the
autocorrelation and carryover effects on the specification
ranking are additive: stronger serial correlation lifts every
specification by roughly the same amount, leaving the
A2 $>$ A1 $\approx$ A3 ordering undisturbed.

## Dropout and the limits of specification refinement

Block S3 is perhaps the most sobering result in the study. On
the one hand, the three specifications are meaningfully
distinguished at low-to-moderate dropout: at 10% MCAR,
power drops from approximately 0.86 to approximately 0.56
across all specifications, and A2 retains a modest absolute
advantage throughout. On the other hand, at 30% dropout all
specifications are near floor (0.12 to 0.25) and are
effectively indistinguishable. A2 provides no protection
against the power loss itself.

Under MAR dropout biased by baseline severity, the pattern is
qualitatively similar but shifted slightly: MAR dropout at 10%
is comparable to MCAR dropout at approximately 12%. This is
expected because MAR dropout preferentially censors
high-severity participants, who contribute more to the
interaction signal under Architecture B.

The practical implication is direct. Dropout prevention (shorter
trial durations, participant engagement, adaptive scheduling) is
far more consequential for the biomarker-interaction test than
the choice of analysis-model carryover specification. From our
perspective, trialists who invest in refining the carryover
specification while tolerating high dropout rates have the
priorities somewhat reversed.

## Architecture dependence

The companion manuscript establishes that Architecture B (MVN
differential correlation) loses substantially more power under
carryover than Architecture A (direct mean moderation). The
present results confirm that this qualitative gap is robust to
decay-form variation: Architecture B is more sensitive to
carryover under all three decay forms and at all five Weibull
shape values examined. We note that the A2 advantage is larger
under Architecture B than under Architecture A, which is
consistent with the mechanism: under Architecture B, carryover
erodes the biomarker-response correlation directly, and the
continuous exposure-weighted predictor recovers more of that
signal than either the binary on-drug indicator or the
lagged-nuisance term. The A2 $>$ A1 $\approx$ A3 ranking holds
under both architectures, though the magnitude of the benefit
warrants closer attention in trials where the data-generating
mechanism is believed to follow Architecture A.

## The latent-subtype question

A reviewer of the companion manuscript broached the question of
whether Architecture B's covariance-based interaction is
adequately modelled by the MVN differential-correlation
parameterisation, or whether an explicit finite-mixture DGP
would be more faithful to the latent-subtype hypothesis. The
question merits acknowledgment here because it bears on the
present results. A true mixture DGP would produce
biomarker-response associations that are qualitatively different
from those generated by a continuous correlation: discrete
responder/non-responder subgroups rather than a graded
covariance. Under a mixture DGP, the lagged-treatment
specification A3 might gain an advantage over A2, because A3
does not assume a specific decay form and may be more tolerant
of the non-smooth response profiles that mixture data can
produce. The present simulation does not evaluate a mixture DGP
and cannot address this question directly. We have not been
able to determine, as yet, whether the A2 dominance we observe
extends to genuinely discrete-subtype data-generating processes;
this is a direction that merits investigation in subsequent
work.

## Comparison with prior simulation studies

Four prior simulation programmes provide direct quantitative
benchmarks for the present results. We discuss each in turn,
focusing on cells where the prior parameter ranges overlap with
ours and noting where the design targets differ.

Hendrickson [-@hendrickson2020] introduced the
biomarker-by-treatment interaction question for aggregated
N-of-1 trials and reported power for the matched-decay (A2)
analysis under exponential carryover at $t_{1/2} = 1$ week,
$N = 70$, $c_{bm} = 0.45$, Hybrid design. The reported power
was 0.82 at 1{,}000 replicates (single architecture). The
present 500-replicate factorial reproduces this result at the
same cell to $0.860 \pm 0.016$ (Architecture B), consistent
with Hendrickson within sampling uncertainty. The principal
methodological extension here is the addition of A1 and A3 as
comparators; Hendrickson did not contrast analysis-side
specifications, so the A2-vs-A1 advantage of approximately
10 percentage points is a new contribution rather than a
replication.

Jones and Kenward [-@jones2014crossover] developed the
crossover-trial framework underlying the A3 specification with
the expectation, noted above, that the non-parametric
lagged-on indicator should outperform a parametric carryover
model under decay-form mis-specification. The present S2
sensitivity block directly tests this expectation and finds it
is not supported in the regimes we examined: A2's power varies
by less than 0.015 across a four-fold half-life
mis-specification, so A3 does not become preferred at any
evaluated mismatch level. The Jones-Kenward expectation is
preserved in spirit (A3 closes the gap as half-life information
degrades), but the practical magnitudes are too small for A3 to
overtake A2 in the explored window. The disagreement is
informative: it appears that the exposure-weighted predictor's
curvature is gentle enough that mis-specification is less
costly than the crossover-trial literature anticipates, at
least at the prazosin-calibrated parameters.

Senn [-@senn2016mastering] established the
variance-decomposition foundation for within-subject designs,
predicting that within-subject contrasts are disproportionately
advantageous when between-patient variance
$\sigma^2_{\text{between}}$ is large relative to within-patient
variance. The present simulation does not vary
$\sigma^2_{\text{between}}$ as a primary axis, but the S1
autocorrelation sensitivity block bears on the related question
of how within-subject information accumulates with serial
correlation. Our finding that A2's advantage is approximately
constant across $\rho \in [0.5, 0.9]$ is qualitatively
consistent with Senn's prediction that within-subject methods
extract roughly proportional information across correlation
regimes; the absence of a strong $\rho$ interaction with the
analysis specification is the new result.

Sturdevant and Lumley [-@sturdevant2021carryover] developed the
two-tier reporting convention we have adopted here: a principal
factorial covering the main contrast, supplemented by marginal
sensitivity blocks anchored at a reference cell. Their
convention was designed for simulation studies whose main claim
is a ranking or contrast rather than a precise power estimate,
and our adoption of it is documented in the
simulation-grid-plan.md design document. We note that the
two-tier structure proved useful in practice: the Tier 1
factorial establishes the A2 $>$ A1 $\approx$ A3 ranking, while
the Tier 2 blocks probe its robustness to single-axis
perturbations. The extension to the biomarker-interaction
estimand is, to the best of our knowledge, new.

The pmsimstats-ng companion manuscripts at
`analysis/report/01-dgp-mean-moderation-vs-mvn/` and
`analysis/report/05-nof1-design-sensitivity/` provide
complementary benchmarks: paper 01 establishes the
architecture-dependence of carryover-induced power loss (the
property confirmed here in the architecture-dependence
subsection), and paper 05 reports the effect of carryover on
the Hybrid N-of-1 design's saturation profile (the property
confirmed in §4.4).

## Limitations

Several limitations should be noted when interpreting these
results.

First, the simulation draws its response-trajectory parameters
from a single configuration calibrated to the Hendrickson (2020)
prazosin-PTSD framework (Gompertz maxima, rates, and
dispersions). The absolute power values are specific to this
parameterisation. The ranking of analysis specifications is
expected to be more portable, but we have not been able to
verify this across alternative response profiles. Trialists
working in therapeutic areas with substantially different
trajectory shapes should regard the power estimates as
illustrative rather than definitive.

Second, the three analysis specifications evaluated here do not
exhaust the space of possible carryover-aware analysis models.
Weighted analysis, within-subject contrasts, two-stage random
slopes, and exclusion of contaminated observations (discussed
in the companion manuscript's Section 6) are not evaluated. We
have endeavored to choose the three specifications that
represent the natural practical alternatives for the N-of-1
context, but a more comprehensive comparison would cross the
present grid with these additional strategies, and others may
merit consideration.

Third, the sensitivity blocks are one-factor-at-a-time
perturbations around a single reference configuration.
Unfortunately, joint perturbations, for example high $\rho$
combined with high dropout, may produce interactions that the
marginal blocks do not capture. The exploratory block S5
($\rho \times$ carryover) provides one such joint check and
does not reveal a qualitative interaction, which is reassuring,
but the space of possible joint combinations is large and our
exploration of it is necessarily incomplete.

Fourth, the Tier 1 grid omits the open-label (OL) design because
OL lacks the off-drug contrast required by the A1 and A3
specifications. For trialists considering an OL design, the
analysis-model question is moot: only A2 (or the `bm:t`
trajectory-based test from the companion manuscript) is
applicable.

Fifth, dropout is modelled as monotone: once a participant
misses a timepoint, all subsequent timepoints are missing. In
practice, intermittent missingness is common in N-of-1 trials,
particularly when participants have good and bad symptom days
that influence measurement compliance. Intermittent missingness
would likely be less destructive than monotone dropout because
it preserves some late-trial observations that contribute to
the treatment contrast. We note this as a realistic departure
from the modelled mechanism that future work should address.

The headline Tier 1 ranking is further supported by the
high-precision rerun reported at the end of §4, which resolves
the A2-vs-A1 gap to MCSE 0.020 with paired McNemar
$p < 10^{-3}$ at the headline cells.

# Conclusions

1. **A2 (exposure-weighted) is the recommended default.** Across
   the full factorial and all sensitivity perturbations, the
   continuous exposure-weighted predictor `Dbc` with an assumed
   exponential decay consistently produces the highest power for
   detecting the biomarker-by-treatment interaction under
   carryover. The advantage over the binary-treatment (A1) and
   lagged-nuisance (A3) alternatives is approximately 10
   percentage points at the reference configuration and is
   robust to DGP decay-form variation, within-factor
   autocorrelation, and analyst-truth half-life mismatch.

2. **Half-life mis-specification is essentially harmless.** A2
   power loses less than 0.02 in absolute terms across a
   four-fold mis-specification of the analyst's assumed half-life
   (Block S2). Trialists who have only an approximate
   pharmacokinetic estimate can adopt A2 with confidence that
   the choice of assumed half-life is not load-bearing.

3. **The lagged-treatment specification does not dominate under
   mis-specification.** Contrary to the expectation from the
   crossover-trial literature, A3 does not become preferable when
   the analyst's decay-form assumption is wrong. A3's
   model-free character provides robustness to assumed half-life
   but not enough efficiency gain to overtake A2 at any evaluated
   operating point.

4. **Dropout is more consequential than the carryover
   specification.** Power degrades sharply with dropout rate
   under all specifications. At 20% dropout, all three
   specifications lose approximately half their power relative to
   the no-dropout condition. Trialists should prioritise dropout
   prevention over carryover-specification refinement.

5. **The DGP architecture modulates the magnitude but not the
   direction of the ranking.** Architecture B (MVN differential
   correlation) amplifies the benefit of A2 because
   carryover-induced correlation erosion makes the continuous
   predictor's information advantage more consequential. Under
   Architecture A (direct mean moderation), A2's advantage is
   smaller but still positive.

6. **Reporting recommendations.** Simulation studies evaluating
   biomarker-moderated treatment effects under carryover should
   report: (a) the analysis-model carryover specification used,
   (b) the assumed decay form and half-life, (c) sensitivity of
   power estimates to at least one alternative specification,
   and (d) the anticipated dropout rate, which is the dominant
   source of power loss in the evaluated grid.

7. **Type-I error is well-controlled.** All three specifications
   maintain nominal or conservative type-I error rates across the
   full Tier 1 grid and the S4 effect-size curve, with no
   evidence of inflation under any DGP decay form, architecture,
   or carryover level.

# Conflict of interest

The authors declare no conflicts of interest.

# Funding

This work received no specific funding.

# Data availability statement

The data and code that support the findings of this study are
openly available in the pmsimstats-ng project repository.
Per-replicate Monte Carlo outputs and ADEMP-compliant cell-level
summaries are versioned under `analysis/data/`; simulation
drivers are at `analysis/scripts/`. Exact paths for this
manuscript are listed in the Reproducibility section below.

# Reproducibility

This manuscript was rendered with R 4.4.x inside the project's
zzcollab Docker container (image hash recorded in `Dockerfile`;
package set captured in `renv.lock`). Principal packages used
by the simulation pipeline are `nlme` (continuous-time
linear-mixed analysis with `corCAR1` residual correlation),
`parallel` (8-core `mclapply` for parameter-grid sweeps),
`data.table` (the `original-extended` simulation collection),
`furrr` and `purrr` (the `tidyverse` simulation collection),
and `rmarkdown` plus `bookdown` for the manuscript build.

**Random-number generator.** Each simulation driver sets
`set.seed(20260507)` at the top of the replicate loop and uses
`parallel::mc.set.seed` inside `mclapply`. Per-replicate seeds
derive from the master seed by `+ rep_idx`.

**Data and code availability.** Production-grade simulation
outputs are at
`analysis/scripts/carryover-sensitivity/output/01-factorial.rds`
(principal factorial) and `04-sensitivity.rds` (sensitivity
blocks S1-S5). Cell-level summaries are at
`analysis/data/02-grid-summary.rds` and
`analysis/data/02-sensitivity-summary.rds`. Driver scripts are
`01-run-factorial.R`, `04-run-sensitivity-blocks.R`, and
`02-summarise-grid.R` under
`analysis/scripts/carryover-sensitivity/`. The simulation grid
plan and design rationale are documented at
`analysis/scripts/carryover-sensitivity/simulation-grid-plan.md`,
serving as a Morris ADEMP pre-registration. The high-precision
prototype rerun outputs are at
`analysis/data/quick-sim/02-carryover-replicates.rds`
(per-replicate) and `02-carryover-summary.txt` (Morris ADEMP
cell-level summary). The manuscript source, paper-specific
bibliography (`references.bib`), and SIM CSL file are at
`analysis/report/02-carryover-sensitivity/`.

# References
