---
title: "Exposure-weighted versus lagged-treatment carryover
  specifications under differential-correlation biomarker
  interactions: a factorial simulation study"
author: 'pmsimstats team'
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: true
    toc_depth: 2
    number_sections: true
    keep_tex: true
    highlight: tango
    includes:
      in_header: ../sim-preamble.tex
bibliography: references.bib
csl: statistics-in-medicine.csl
link-citations: true
linestretch: 1.2
fontsize: 12pt
geometry: "left=2.5cm,right=6cm,top=2.5cm,bottom=2.5cm"
header-includes:
  - \usepackage{amssymb}
knit: (function(input, ...) { d <- dirname(input); while (!file.exists(file.path(d, 'tools', 'stamp-render.R')) && d != dirname(d)) d <- dirname(d); source(file.path(d, 'tools', 'stamp-render.R'))$value(input) })
---

<!--
  Build pipeline for this manuscript
  ==================================

  This is the single master source for this manuscript. Its scope is
  Architecture B (MVN differential correlation), the exponential and
  Weibull decay forms, and the E2 versus E3 analysis specifications.
  Earlier, wider-scope drafts are retained under archive/ and are no
  longer maintained; see README.md in this directory for the lineage
  and for what each archived draft uniquely contains.

  This Rmd does NOT run any simulation code. It loads pre-computed
  summaries and figures produced by the scripts below. Re-run the
  relevant steps when the DGP, analysis specs, or parameter grid
  change. All paths are relative to the repo root.

  1. Tier 1 principal factorial (heaviest step, ~85 min on 8 cores):
       Rscript analysis/scripts/carryover-sensitivity/01-run-factorial.R
     Writes: analysis/scripts/carryover-sensitivity/output/01-factorial.rds

  2. Tier 2 sensitivity blocks S1..S5 (~4 min on 8 cores; requires
     Tier 1 not to be running concurrently - both request all cores):
       Rscript analysis/scripts/carryover-sensitivity/04-run-sensitivity-blocks.R
     Writes: analysis/scripts/carryover-sensitivity/output/04-sensitivity.rds

  3. Summarize both tiers into cell-level power/bias/convergence
     (~seconds; re-run after either step 1 or step 2 changes):
       Rscript analysis/scripts/carryover-sensitivity/02-summarise-grid.R
     Writes:
       analysis/data/02-grid-summary.rds        (from Tier 1)
       analysis/data/02-sensitivity-summary.rds (from Tier 2)

  4. Filtered Tier 1 figures (Architecture B, no linear decay, E2/E3
     only; writes analysis/figures/02xs-*.pdf):
       Rscript analysis/scripts/carryover-sensitivity/03-render-figures-extra-slim.R

  5. Filtered Tier 2 sensitivity figures (writes
     analysis/figures/02xs-sens-S1..S5.pdf):
       Rscript analysis/scripts/carryover-sensitivity/04-sensitivity-figures-extra-slim.R

  5b. S6 cluster-robust recalibration data
      (analysis/scripts/carryover-sensitivity/output/diag-s6-cr2.rds)
      is produced by 06-cr2-recalibration.R; no re-run needed for a
      plain re-knit.

  5c. Filtered S6 Type I figure (writes
      analysis/figures/02xs-type1-cr2.pdf):
       Rscript analysis/scripts/carryover-sensitivity/07-type1-cr2-figure-extra-slim.R

  The unsuffixed figure scripts (03-render-figures.R,
  04-sensitivity-figures.R, 07-type1-cr2-figure.R) write the
  full-scope 02-*.pdf figures consumed only by the archived drafts.
  They are retained for that reason and are not part of this
  manuscript's build.

  6. Knit this Rmd (this step; ~1 min; no simulation code runs):
       Rscript -e "rmarkdown::render('analysis/report/02-carryover-sensitivity/report.Rmd')"

  Grid design and rationale (full grid, of which this manuscript is
  a slice):
    analysis/scripts/carryover-sensitivity/simulation-grid-plan.md
-->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.path = "../../figures/02-cache-",
  cache.path = "_cache/"
)
library(tibble)
library(dplyr)
library(knitr)

## The archived simulation output labels the three analysis
## specifications A1/A2/A3. The manuscript reports them as E1/E2/E3 so
## that the letters A/B/C stay reserved for DGP architectures. Data
## values are never rewritten; they are mapped at display time only.
spec_display <- function(x) {
  m <- c(A1 = 'E1', A2 = 'E2', A3 = 'E3')
  x <- as.character(x)
  ifelse(x %in% names(m), unname(m[x]), x)
}
```

# Abstract {-}

**Background.** Aggregated N-of-1 trials pool repeated within-patient
on/off contrasts to validate predictive biomarkers. Under the
multivariate-normal (MVN) differential-correlation architecture
(Architecture B), in which the biomarker-treatment interaction is
encoded in a treatment-state-dependent correlation, a companion study
found that carryover, that is, residual drug effect that decays after
discontinuation, costs up to roughly 31% of power in relative terms.
The analyst must still decide how to represent that residual exposure
in the linear predictor. Unfortunately, it is not established which
representation is most robust to the ordinarily unknown
pharmacokinetic decay shape.

**Methods.** We compare two analysis-model specifications for the
drug-exposure regressor under Architecture B: an exposure-weighted
continuous indicator that assumes a parametric decay and half-life
(E2, the current default), and the binary on-drug indicator augmented
by a lagged just-off-drug term that estimates carryover
non-parametrically (E3, the classical crossover device). We embed
them in an ADEMP-structured factorial simulation
[@morris2019simulation] crossing two data-generating process (DGP)
decay forms (exponential, Weibull) with the two specifications, under
three trial designs, two sample sizes, and three interaction
strengths (216 cells, 500 replicates each), together with four
single-axis sensitivity blocks and a cluster-robust recalibration. We
report every performance measure with its Monte Carlo standard error
(MCSE).

**Results.** E2 attains the highest power only in the two designs
with a blinded-discontinuation phase (Hybrid and OL+BDC), where its
advantage over E3 is about 10 percentage points (0.860 vs 0.770 at
the reference cell). Conversely, under the classical crossover (CO)
design E2 is markedly inferior (0.488 vs 0.830 for E3, which tracks
the binary indicator closely). Where E2 leads, the lead appears
robust to decay-shape and half-life mis-specification and to
autocorrelation strength, and a paired-McNemar high-precision rerun
confirms it at the highest-leverage cell ($+7.2$ points,
$p = 1.2 \times 10^{-7}$),
though the two specifications are statistically tied at three of
four other rerun cells. The model-based interaction test is mildly
conservative (mean null rejection 0.039), an artifact of the working
covariance that a CR2 cluster-robust standard error removes without
disturbing the ranking.

**Conclusions.** Three points warrant emphasis. First, under
Architecture B we recommend the exposure-weighted predictor E2,
reported with a cluster-robust standard error, for the Hybrid and
OL+BDC designs; the lagged-treatment specification E3 is preferable,
or at least not worse, under a crossover design. Second, across all
conditions anticipated dropout governs power far more than the choice
between the two specifications, and should dominate design effort
accordingly. Third, we do not claim to have settled the wider
comparison: this manuscript restricts the grid to Architecture B and
to the exponential and Weibull decay forms, and does not evaluate the
binary on-drug specification (E1) or Architecture A (mean
moderation); the full-scope evaluation is retained in this
repository as the archived companion drafts
`archive/report.Rmd` and `archive/report-slim.Rmd`.

# Introduction

In a single N-of-1 trial, one patient is crossed repeatedly between
active treatment and a comparator over a sequence of measurement
occasions, and the within-patient contrast between the on-drug and
off-drug periods estimates that patient's individual treatment
effect. An aggregated N-of-1 trial pools many such single-patient
series into a mixed model, trading some of the resolution available
within a single patient for population-level inference and for
sample sizes considerably smaller than a parallel-group trial would
require [@hendrickson2020]. The estimand of interest is the
biomarker-treatment interaction, that is, whether a baseline
covariate $B_i$ modifies the treatment effect. We examine three
trial designs that differ in how the on-drug and off-drug occasions
are arranged, a classical crossover (CO), an open-label design
followed by blinded discontinuation (OL+BDC), and a Hybrid N-of-1
design; each is defined in Section 2.

The inferential complication specific to this setting is carryover.
When a drug is discontinued, its pharmacological effect does not
switch off abruptly but decays over days or weeks, so that
observations recorded during the nominally off-drug period still
carry residual exposure. Treating those observations as though they
were cleanly off-drug mis-specifies the design matrix, and the
consequences for the interaction test are worth setting out
explicitly.

The exposure regressor the analyst writes down is, in that case, a
mis-measured version of the exposure the patient actually received,
and the damage is of two kinds. First, the contrast that the
interaction term is meant to estimate is diluted. Occasions that
still carry a substantial fraction of the drug effect are entered
with the same value as occasions that carry none, so the coefficient
on the biomarker-by-exposure product is biased toward zero, in the
familiar manner of a covariate measured with error. Second, those
same occasions are heterogeneous in a way the model does not
represent, and the unexplained variation is absorbed into the
residual variance, inflating the standard error of the very
coefficient that has just been attenuated. The test statistic loses
on both counts, its numerator shrinking while its denominator grows,
and power falls accordingly.

We note that this is a loss of power and not a loss of validity.
Under the null, $c_{bm} = 0$, Architecture B generates no
biomarker-response association at any level of residual exposure, so
a mis-coded exposure regressor has no interaction to distort, and the
size of the test is unaffected by the choice of exposure coding. The
mild conservatism we do observe in Section 3.5 has a separate origin
in the working covariance and applies equally to both specifications.
Carryover mis-specification is therefore a question of efficiency
rather than a source of spurious findings, and it is on that basis
that we frame what follows in terms of recovered power.

But how should residual exposure be represented in the linear
predictor? Two answers are in common use, and the question we take up
here is which of them recovers more of the power that carryover would
otherwise cost.

A companion study [@pmsimstats_dgp_architectures2026], building on
Hendrickson [-@hendrickson2020], shows that under what we call
Architecture B, a multivariate-normal differential-correlation
formulation in which the interaction lives in the
treatment-state-dependent covariance
$\text{Cor}(B_i, BR_{it}) = c_{bm} \cdot \phi(t_{sd})$, carryover
erodes power directly, and the loss is strongly design-dependent: at
a half-life of one week, the relative loss is roughly 31% under
OL+BDC, but only 3 to 5% under the crossover and Hybrid designs.
Under an alternative architecture, in which the biomarker instead
moderates the mean response directly, the corresponding loss is only
1 to 3% in every design; that architecture, together with a third
combined architecture, falls outside the scope of the present
manuscript,
which restricts attention to Architecture B, where the carryover
problem is largest and where Hendrickson's original results were
obtained. Unfortunately, that earlier work fixed the decay shape at
exponential and evaluated only a single analysis specification,
leaving open the question of which representation of residual
exposure performs best when the true decay shape is not known with
any precision, as is ordinarily the case in practice.

We compare two ways of representing drug exposure in the linear
predictor: an exposure-weighted continuous indicator that commits to
a parametric decay form and half-life, which we call E2 and which is
the current default in the pmsimstats-ng package, and the binary
on-drug indicator augmented by a lagged just-off-drug term that
estimates the magnitude of carryover without assuming any particular
functional form, which we call E3, following the classical
crossover-trial device described by Jones and Kenward
[-@jones2014crossover]. We cross these two specifications against
exponential and Weibull data-generating decay forms and ask whether
one specification can be recommended outright, or whether the choice
depends on the trial design. The study follows the ADEMP reporting
framework of Morris, White, and Crowther
[-@morris2019simulation], and every performance measure we report
carries its Monte Carlo standard error. Section 2 describes the
simulation design in ADEMP order, Section 3 gives the results, and
Section 4 takes up the practical implications for trialists.

# Simulation design

## Aims

The aim of this study is to quantify, under Architecture B, the cost
of mis-specifying the carryover representation in the analysis
model, and to determine which of the two candidate strategies, E2 or
E3, performs better across the grid of data-generating decay forms
and trial designs. A second and related aim, addressed in the Tier 2
sensitivity blocks of Section 3.6, is to check whether the resulting
ranking survives perturbation of the within-subject autocorrelation,
the dropout rate, the accuracy of the assumed half-life, and the
size of the biomarker-treatment interaction itself.

## Data-generating mechanism

Under Architecture B, the biomarker and the biological response are
drawn jointly from a multivariate normal distribution with a
treatment-state-dependent correlation,
$\text{Cor}(B_i, BR_{it}) = c_{bm} \cdot \phi(t_{sd})$, where $\phi$
denotes the decay form and $t_{sd}$ is the time elapsed since drug
discontinuation. Architecture A, in which the biomarker instead
moderates the mean response directly, and a combined Architecture C,
fall outside the scope of the present manuscript; the full
comparison across architectures is reported in the archived
full-scope drafts (`archive/report.Rmd`,
`archive/report-slim.Rmd`). In both architectures,
the decay function $\phi : \mathbb{R}_+ \to [0, 1]$ maps $t_{sd}$ to
a residual-effect multiplier. We evaluate two such forms here, each
parameterized to match a common half-life $t_{1/2}$. A third form,
linear decay, appears in the full-scope manuscripts but is not
considered further here.

### Exponential decay

$$
\phi_{\text{exp}}(t_{sd}) =
  \exp\left(-\lambda t_{sd}\right), \qquad
  \lambda = \frac{\ln 2}{t_{1/2}}
$$

This is the decay form used throughout the companion manuscript, and
it is the one currently implemented in `functions.R` under
`implementations/tidyverse/`. It corresponds to
classical first-order elimination kinetics and is the default
assumption in pharmacokinetics; we retain it here as the reference
against which the Weibull form is compared.

### Weibull decay

$$
\phi_{\text{weib}}(t_{sd}) =
  \exp\left(-\left(\lambda_w t_{sd}\right)^{k}\right), \qquad
  \lambda_w = \frac{(\ln 2)^{1/k}}{t_{1/2}}
$$

A shape parameter $k$ governs whether the elimination tail is
heavier than exponential ($k < 1$, slower elimination) or lighter
($k > 1$, faster off-drug clearance); the exponential form is
recovered exactly at $k = 1$. We report results at $k = 0.7$,
representing a drug with a prolonged low-concentration tail, and at
$k = 1.5$, representing a drug cleared more rapidly once
concentration falls below some saturation threshold, perhaps through
active transport. Together with $k = 1.0$, which is numerically
indistinguishable from the exponential form, this gives four
decay-shape levels in the grid.

## Estimand and analysis specifications

Under Architecture B, the interaction estimand is the on-drug
biomarker-response correlation $c_{bm}$, a dimensionless quantity.
Bias and empirical standard error are reported on a calibrated
effect-size scale, namely the expected difference in on-drug
response per standard deviation of the biomarker at $t_{1/2} = 0$,
as defined in `docs/19-mean-moderation-implementation-notes.tex`;
power and Type I error are reported on the native test-statistic
scale. The null estimand used for Type I error verification is
$c_{bm} = 0$, realized by the corresponding row of the parameter
grid.

The analysis model is `nlme::lme(Sx ~ bm * X + t, random = ~1
| ptID, correlation = corCAR1(form = ~ t | ptID))` for both
specifications considered here; they differ only in how the
drug-exposure predictor $X$ is constructed.

### E2: exposure-weighted (continuous predictor)

Under E2, $X = \text{Dbc}_{it}$, a continuous exposure-decayed
predictor constructed to match the analyst's assumed decay form and
half-life; this is the current pmsimstats-ng default. At on-drug
timepoints $D_{bc,it} = 1$ regardless of the assumed half-life. At
off-drug timepoints the predictor decays as
$D_{bc,it} = \exp(-(\ln 2/\hat{t}_{1/2})\,t_{sd,it}) =
(1/2)^{t_{sd,it}/\hat{t}_{1/2}}$, losing exactly half its value for
every $\hat{t}_{1/2}$ weeks that elapse after discontinuation. When
the analyst's assumed decay form matches the true data-generating
form, the predictor is well matched; when it does not, it is
mis-specified in a way we probe directly in Tier 2 block S2. We note
that the half-life assumption enters only through the predictor
values handed to `lme`, not through the model formula, which is
identical regardless of $\hat{t}_{1/2}$. That is, E2 is in effect a
different model for every value of the assumed half-life, even
though it is written the same way each time.

### E3: lagged-treatment (estimated carryover term)

Under E3, $X = D_{it}$ is augmented with a lagged-treatment indicator
$L_{it} = D_{i,t-1}(1 - D_{it})$, which marks the off-drug timepoints
immediately following an on-drug timepoint. The magnitude of
carryover is estimated from the data rather than assumed in advance,
so this specification does not require the analyst to posit any
particular decay form. The device follows the classical
crossover-trial tradition surveyed by Jones and Kenward
[-@jones2014crossover]. One structural feature of E3 bears directly
on how the results below should be read: the lagged term $L_{it}$
enters the model only as a nuisance main effect, and the interaction
actually tested under E3 remains $\text{bm:}D_{it}$, the same
coefficient tested under the binary on-drug specification (which we
call E1 and do not evaluate further in this manuscript). E3 is best
understood, then, as E1 with a carryover nuisance control added,
rather than as a fundamentally different model of the interaction.
Because it carries no half-life parameter, E3 is invariant to the
analyst's assumed half-life by construction, in contrast to E2. This
asymmetry, one specification indexed continuously by
$\hat{t}_{1/2}$ and the other not a member of that family at all, is
what the Tier 2 block S2 sensitivity contrast in Section 3.6 is
designed to probe.

## Factorial grid and parameter settings

The principal comparison is a factorial of DGP decay form (four
levels: exponential; Weibull at $k \in \{0.7, 1.0, 1.5\}$) against
analysis-model specification (E2, E3), under Architecture B only:

```{r grid-table}
grid <- tibble::tribble(
  ~`DGP decay`,          ~`E2 exposure-weighted`, ~`E3 lagged-treatment`,
  "Exponential",         "cell 1 (matched)",       "cell 2",
  "Weibull (k = 0.7)",   "cell 3",                 "cell 4",
  "Weibull (k = 1.0)",   "cell 5",                 "cell 6",
  "Weibull (k = 1.5)",   "cell 7",                 "cell 8"
)
knitr::kable(
  grid,
  caption = "Principal simulation grid (Architecture B only). Cell 1
    is the matched reference: exponential DGP analyzed under the
    current pmsimstats-ng default. Off-diagonal cells quantify the
    cost of DGP-analysis mis-specification."
)
```

| Parameter | Values |
|---|---|
| Biomarker moderation ($c_{bm}$) | 0.0, 0.30, 0.45 |
| Carryover half-life ($t_{1/2}$) | 0.0, 0.5, 1.0 weeks |
| Weibull shape ($k$) | 0.7, 1.0, 1.5 (Weibull form only) |
| Sample size ($N$) | 35, 70 |
| Trial design | CO, N-of-1 (Hybrid), OL+BDC |
| DGP architecture | B (MVN differential correlation) only |
| Replicates per cell | 500 (target); 50 for development |

The 216-cell count follows the same nesting logic used in the
full-scope manuscript, with the architecture dimension fixed at B
and the linear decay form removed. The decay-shape combinations
number four (exponential, and Weibull at $k \in \{0.7, 1.0, 1.5\}$),
so the grid comes to $4 \times 3 \times 3 \times 2 \times 3 = 216$
cells, the factors being decay shape, carryover half-life, trial
design, sample size, and biomarker level. The $c_{bm} = 0$ row
serves as the Type I error check, targeting the nominal 5% level,
and the $t_{1/2} = 0$ row, in which carryover is absent altogether,
gives a best-case power reference at each combination of sample
size, design, and $c_{bm}$, against which the power cost of nonzero
carryover can be read off directly.

## Performance measures and computing environment

For each cell we report Power, Type I error (at $c_{bm} = 0$), Bias
and Empirical SE of the interaction estimate (on the calibrated
scale), Coverage of the nominal 95% Wald CI, MSE, the
non-convergence rate, and the MCSE of each measure, following Morris
et al. [-@morris2019simulation, Tables 5-6].

We chose $n_{sim} = 500$ so that the Monte Carlo standard error on
power at $p = 0.5$ would not exceed 0.022
($\sqrt{0.5 \cdot 0.5 / 500} \approx 0.022$). No simulation is run
for this manuscript; we instead filter the same production-scale
grid summary used in the archived full-scope drafts (540 cells,
`analysis/data/02-grid-summary.rds`,
computed with `implementations/tidyverse/` as the primary
implementation under R 4.5.3, 128 minutes on eight cores, commit SHA
`7018b59`, dated 2026-04-15) down to the 216-cell slice defined by
Architecture B, the two non-linear decay forms, and the E2 and E3
specifications. Within each cell, both remaining specifications were
fit to the same simulated dataset (common random numbers), so the
E2 minus E3 contrast is estimated with reduced Monte Carlo variance
relative to what independent draws would give. The Tier 2
sensitivity blocks and the S6 cluster-robust recalibration are
filtered from the same shared outputs,
`02-sensitivity-summary.rds` under `analysis/data/` and
`diag-s6-cr2.rds` under the driver output directory; their designs
were already anchored at Architecture B. The filtered figures used
here are produced by three filtered figure scripts, listed with
their full paths in the Reproducibility section.

# Results

All results reported below derive from the same 500-replicate
production run used in the full-scope manuscripts, filtered to
Architecture B, the exponential and Weibull decay forms, and the E2
and E3 specifications. The Monte Carlo standard error on power does
not exceed 0.022 at $p = 0.5$, and differences smaller than this
should not be over-interpreted. Coverage of the nominal 95% Wald
confidence interval is reported for the reference cell, and
non-convergence rates are reported per cell in the tables that
follow.

## Headline performance table {#sec-headline}

Table \@ref(tab:headline) reports performance at the reference cell
(Architecture B, exponential DGP, Hybrid design, $N = 70$,
$c_{bm} = 0.45$), where non-convergence was zero throughout. A
caveat applies to the bias, mean squared error, and coverage
columns: these are computed against E2's calibrated target,
$\theta_{\text{true}} = -c_{bm}\sigma_{BR} = -3.6$, while E3
estimates a different coefficient, $\text{bm:}D_{it}$. E3's apparent
bias and under-coverage in this table reflect that mismatch in
estimand rather than any defect in the estimator, and for that
reason we compare only power and Type I error across the two
specifications; the bias and coverage columns are reported for
completeness but are not directly comparable between E2 and E3.

```{r headline, echo = FALSE}
suppressPackageStartupMessages({library(dplyr)})
hd <- readRDS(here::here(
  'analysis/data/02-grid-summary.rds'))$summary |>
  filter(carryover_form == 'exponential', dgp_arch == 'mvn',
         design == 'Hybrid', N == 70, c_bm == 0.45,
         spec %in% c('A2', 'A3')) |>
  mutate(
    `t1/2` = t1half,
    `Power (MCSE)` = sprintf('%.3f (%.3f)', power, mc_se_power),
    `Bias (MCSE)` = sprintf('%+.3f (%.3f)', bias, mc_se_bias),
    `Empirical SE` = sprintf('%.3f', empirical_se),
    `Coverage (MCSE)` = sprintf('%.3f (%.3f)',
                                  coverage, mc_se_coverage),
    `MSE (MCSE)` = sprintf('%.3f (%.3f)', mse, mc_se_mse),
    `Non-conv` = sprintf('%.3f', non_convergence_rate)
  ) |>
  arrange(`t1/2`, spec) |>
  mutate(spec = spec_display(spec)) |>
  select(`t1/2`, spec, `Power (MCSE)`, `Bias (MCSE)`,
         `Empirical SE`, `Coverage (MCSE)`, `MSE (MCSE)`,
         `Non-conv`)

knitr::kable(
  hd, booktabs = TRUE,
  caption = 'Headline performance at the reference cell
    (Architecture B, exponential DGP, Hybrid design, $N = 70$,
    $c_{bm} = 0.45$, $n_{sim} = 500$). Power is the rejection rate of
    $H_0: \\beta_{bm:X} = 0$ at $\\alpha = 0.05$. Bias is on the
    calibrated effect-size scale,
    $\\theta_{\\text{true}} = -c_{bm} \\cdot \\sigma_{BR} = -3.6$
    (per docs/19 calibration). MCSEs are reported per Morris et al.
    (2019, Table 6).'
) |>
  kableExtra::kable_styling(latex_options = "scale_down")
```

## Matched-decay baseline

At the matched cell (exponential DGP, E2 exposure-weighted analysis,
Architecture B, $N = 70$, $c_{bm} = 0.45$, Hybrid design), power
declines from $0.988 \pm 0.005$ at $t_{1/2} = 0$ to
$0.948 \pm 0.010$ at $t_{1/2} = 0.5$ to $0.860 \pm 0.016$ at
$t_{1/2} = 1.0$ weeks (point estimate plus or minus MCSE,
$n_{sim} = 500$). All 500 replicates converged at every half-life in
this cell. The value at $t_{1/2} = 1.0$ agrees with Hendrickson's
reported power of 0.82 for the same cell [@hendrickson2020], which
we take as reassurance that the simulation pipeline reproduces the
baseline result correctly.

## Power by analysis specification

```{r fig-power-by-spec, echo = FALSE, fig.cap = "Power versus carryover half-life under the exponential DGP, Architecture B, stratified by analysis specification. E2 retains a modest advantage over E3 in the Hybrid and OL+BDC designs at $t_{1/2} = 1.0$ weeks, but is dominated by E3 under the crossover (CO) design.", out.width = "95%"}
knitr::include_graphics("../../figures/02xs-power-by-spec.pdf")
```

The two specifications give nearly identical power at $t_{1/2} = 0$,
where there is no carryover to model, and diverge progressively as
the half-life increases. E2 retains a modest but consistent
advantage in the Hybrid and OL+BDC designs, an advantage that first
appears at $t_{1/2} = 0.5$ weeks and grows through $t_{1/2} = 1.0$
weeks. This advantage does not appear under the crossover design,
where E2 is in fact dominated, a point we return to in Section 4.
The mechanism appears straightforward: in a crossover trial, the
within-subject correlation already carries most of the signal that
E2's decaying predictor was designed to recover, so E2 has little
left to gain, and its down-weighting of the off-drug observations
only works against it.

## Decay-form mis-specification

```{r fig-heatmap-mismatch, echo = FALSE, fig.cap = "Power at the critical cell ($t_{1/2} = 1.0$ weeks, Architecture B, Hybrid design, $N = 70$, $c_{bm} = 0.45$) across DGP decay forms (rows) and analysis specifications (columns). Under the heavy-tailed Weibull DGP ($k = 0.7$), E2's advantage over E3 narrows because its assumed exponential decay no longer matches truth. The fill scale is stretched to the observed power range (0.66-0.87), not the full [0,1] range, so that these cell-to-cell differences are visible; the absolute shading should not be compared to other figures in this manuscript.", out.width = "70%"}
knitr::include_graphics("../../figures/02xs-heatmap-matched-vs-mismatched.pdf")
```

The heatmap cell at the intersection of the exponential DGP row and
the E2 column is the matched reference, with power
$0.860 \pm 0.016$. The same row's E3 column gives the power penalty
of the lagged specification when the data-generating decay is
exponential, while the other rows of the E2 column give the cost of
assuming exponential decay in the analysis when the true decay form
is Weibull. Even under the heaviest-tailed Weibull data-generating
process we examined ($k = 0.7$), E2's advantage over E3 narrows but
does not reverse. We take up the practical interpretation of this in
Section 4.

## Type I error control and cluster-robust recalibration

```{r fig-type1, echo = FALSE, fig.cap = "Empirical Type I error rates under the null ($c_{bm} = 0$), Architecture B, pooled across DGP decay forms and carryover half-lives, stratified by design and analysis specification. Boxplots cover the distribution across cells at $N = 70$; the dashed reference line marks the nominal 0.05 level.", out.width = "80%"}
knitr::include_graphics("../../figures/02xs-type1-boxplots.pdf")
```

Across the null cells, empirical Type I error sits uniformly below
the nominal 0.05 level for both specifications and in all three
designs. This shortfall is not sampling noise but a genuine
conservatism of the model-based interaction test, which the
companion calibration study [@pmsimstats_calibration2026] traces to
the corCAR1 working covariance: the estimated standard error runs
six to ten percent too large, and the realized rejection rate falls
correspondingly. Because the conservatism is uniform across
specifications, it does not disturb the E2-versus-E3 comparison, and
the cluster-robust recalibration reported below (block S6) confirms
that a CR2 cluster-robust standard error restores the rejection rate
to nominal without changing which specification leads. Table
\@ref(tab:tab-type1-cr2) makes the per-specification conservatism,
and its cluster-robust correction, explicit at the null cell.

```{r tab-type1-cr2, echo = FALSE}
suppressPackageStartupMessages(library(dplyr))
t1 <- readRDS(here::here(
  'analysis/scripts/carryover-sensitivity/output/diag-null-type1-cr2.rds'))
tab <- t1$summary |>
  filter(spec %in% c('A2', 'A3')) |>
  transmute(
    Specification    = spec_display(spec),
    `$\\kappa$ model` = sprintf('%.2f', mod_se_ratio),
    `Type I model`    = sprintf('%.3f', type1_mod),
    `$\\kappa$ CR2`    = sprintf('%.2f', cr2_se_ratio),
    `Type I CR2`      = sprintf('%.3f', type1_cr2))
knitr::kable(tab, booktabs = TRUE, escape = FALSE,
  caption = 'Conservatism of the model-based interaction test and its
    cluster-robust correction, at the null cell ($c_{bm} = 0$,
    exponential DGP, $t_{1/2} = 1.0$ weeks, Hybrid design, $N = 70$,
    Architecture B, 3{,}000 replicates). $\\kappa$ is the ratio of the
    mean standard error to the empirical standard deviation of the
    interaction estimate; $\\kappa > 1$ indicates a conservative
    test. Both specifications are conservative under the model-based
    test, most so the exposure-weighted E2 ($\\kappa = 1.10$, Type I
    $= 0.029$); the CR2 bias-reduced cluster-robust standard error
    restores $\\kappa$ to near one and Type I to nominal (0.045 to
    0.049) for both specifications.')
```

```{r tab-s6, echo = FALSE}
suppressPackageStartupMessages({library(dplyr); library(tidyr)})
s6 <- readRDS(here::here(
  'analysis/scripts/carryover-sensitivity/output/diag-s6-cr2.rds'))
ord <- c('Reference (N=70)', 'Small N (N=35)',
         'High autocorr (rho=0.9)',
         'Half-life mis-spec (assume 0.25)', '30% MCAR dropout')
lab <- c('Reference, $N=70$', 'Small $N=35$',
         'High autocorr, $\\rho=0.9$', 'Half-life mis-spec',
         '30\\% MCAR dropout')
## The pre-computed s6$gaps object is the E2-E1 gap inherited from
## the full-scope (archive/report.Rmd) diagnostic run; this manuscript's
## comparison is E2 vs E3, so the E2-E3 gap is computed directly
## from s6$summary (which carries per-spec power under both
## inferences) rather than reused from s6$gaps.
a3_gap <- s6$summary |>
  filter(spec %in% c('A2', 'A3')) |>
  select(cell, spec, power_mod, power_cr2) |>
  pivot_wider(names_from = spec, values_from = c(power_mod, power_cr2)) |>
  transmute(cell,
            gap_mod_a3 = power_mod_A2 - power_mod_A3,
            gap_cr2_a3 = power_cr2_A2 - power_cr2_A3)
tab <- s6$summary |>
  filter(spec == 'A2') |>
  left_join(a3_gap, by = 'cell') |>
  mutate(o = match(cell, ord)) |>
  arrange(o) |>
  transmute(
    Cell = lab[o],
    `$\\kappa$ model` = sprintf('%.2f', mod_se_ratio),
    `$\\kappa$ CR2`   = sprintf('%.2f', cr2_se_ratio),
    `Power model`     = sprintf('%.3f', power_mod),
    `Power CR2`       = sprintf('%.3f', power_cr2),
    `E2$-$E3 model`   = sprintf('%+.3f', gap_mod_a3),
    `E2$-$E3 CR2`     = sprintf('%+.3f', gap_cr2_a3),
    `CR2 df`          = sprintf('%.0f', cr2_df))
knitr::kable(tab, booktabs = TRUE, escape = FALSE,
  caption = 'Cluster-robust recalibration of E2 (Power columns) and
    of the E2$-$E3 ranking (last two columns), at the reference cell
    and four stress cells (500 replicates, $c_{bm} = 0.45$,
    Architecture B). $\\kappa$ is the ratio of the mean standard
    error to the empirical standard deviation of the interaction
    estimate for E2; a value above one is a conservative test. The
    E2$-$E3 columns are computed directly from the per-specification
    power at this cell under each inference, so they are a genuine
    check of this manuscript\'s own comparison rather than the
    E2-vs-E1 contrast reported for the full-scope grid. CR2 df is
    the mean Satterthwaite degrees of freedom for the cluster-robust
    test on E2. Computed with \\texttt{clubSandwich}.') |>
  kableExtra::kable_styling(latex_options = "scale_down")
```

```{r fig-type1-cr2, echo = FALSE, fig.cap = "Empirical Type I error at the null ($c_{bm} = 0$) under model-based and CR2 cluster-robust inference, across the five S6 stress cells and the two specifications (Architecture B). The model-based test (open) is conservative (near 0.03); CR2 (filled) is restored to the nominal 0.05 (dashed), within the binomial Monte Carlo band (gray). 500 null replicates per cell.", out.width = "80%"}
knitr::include_graphics("../../figures/02xs-type1-cr2.pdf")
```

Figure \@ref(fig:fig-type1-cr2) shows the effect of recalibration on
test size directly. The model-based null rejection rate sits below
nominal across all five S6 cells and both specifications, while the
CR2 test restores it to within the Monte Carlo band around 0.05.
This matters because a test that is not correctly sized cannot
fairly be preferred on power alone. In the information-rich cells,
the calibration ratio $\kappa$ runs from 1.07 to 1.26 under the
model-based test and is restored to about one under CR2, and the
correctly sized test recovers two to five points of power in the
process. Table \@ref(tab:tab-s6) reports the E2 minus E3 gap, which
is this manuscript's own comparison, computed directly from the
per-specification power at each S6 cell rather than borrowed from
the E2-versus-E1 contrast reported in the full-scope grid. That gap
is held or slightly widened under CR2 at the reference cell
($+0.080$ to $+0.092$), at the small-$N$ cell ($+0.068$ to
$+0.086$), and at the half-life mis-specification cell ($+0.012$ to
$+0.022$); it narrows somewhat under high autocorrelation ($+0.052$
to $+0.036$, though E2 still leads); and at 30% MCAR dropout it
collapses from a small E2 lead under the model-based test
($+0.014$) to an essentially null difference under CR2
($-0.0002$), which we read as a statistical tie. We do not regard
this as a contradiction of the ranking so much as a restatement,
from a different angle, of the dropout finding in block S3 below:
E2's advantage is driven by information carried in the off-drug
observations, and once dropout has removed most of that information
there is not much left for either the model-based test or its
cluster-robust correction to distinguish. This recalibration was
validated only in the Architecture B, Hybrid, $N = 70$ cells, and it
should not be extended to the crossover design, where E2 does not
lead in the first place.

## Sensitivity analyses

Four Tier 2 blocks perturb one axis at a time around a common
reference cell (Architecture B, Hybrid, $N = 70$, $c_{bm} = 0.45$,
exponential DGP, $t_{1/2} = 1.0$), holding the remaining factors
fixed. The blocks are reported separately here and are not pooled
with Tier 1 or with each other.

### S1: within-factor autocorrelation

```{r fig-sens-s1, echo = FALSE, fig.cap = "Power by analysis specification across AR(1) autocorrelation $\\rho \\in \\{0.5, 0.7, 0.8, 0.9\\}$, Architecture B.", out.width = "60%"}
knitr::include_graphics("../../figures/02xs-sens-S1.pdf")
```

Both specifications gain power monotonically with $\rho$. At
$\rho = 0.5$, E2 power is $0.798 \pm 0.018$ against E3's
$0.738 \pm 0.020$; at $\rho = 0.9$, E2 reaches $0.956 \pm 0.009$
against E3's $0.910 \pm 0.013$ (point estimate plus or minus MCSE,
$n_{sim} = 500$). The E2 advantage over E3, roughly 0.06 in absolute
power throughout, is preserved at every autocorrelation level we
examined, which suggests that autocorrelation strength and the
choice of specification act additively rather than interacting with
one another.

### S2: analyst-truth half-life mismatch

```{r fig-sens-s2, echo = FALSE, fig.cap = "Power of E2 and E3 as the analyst's assumed half-life departs from the true DGP half-life. E3 is invariant to the assumed half-life by construction; E2's power degrades with mis-specification.", out.width = "80%"}
knitr::include_graphics("../../figures/02xs-sens-S2.pdf")
```

E2 appears fairly robust to a mismatch between the analyst's assumed
half-life and the true data-generating half-life. Across the
eight S2 cells (DGP $t_{1/2} \in \{0.5, 1.0\}$ weeks crossed with
analyst-assumed $t_{1/2} \in \{0.25, 0.5, 1.0, 2.0\}$ weeks), E2
averages 0.868, ranging from 0.766 to 0.964, with per-cell MCSE
between 0.008 and 0.019. E3, which does not consume the half-life
parameter, is invariant to the analyst's assumption by construction,
and its power averages 0.840 across the same eight cells. The
exposure-weighted predictor's curvature is gentle enough that
mis-specifying the assumed half-life by a factor of two does not
erode E2's advantage over E3 at either true half-life. It would
appear, then, that E2 remains a defensible choice even when the true
pharmacokinetic half-life is known only approximately at the design
stage.

### S3: dropout

```{r fig-sens-s3, echo = FALSE, fig.cap = "Power at dropout rates $\\in \\{0, 0.1, 0.2, 0.3\\}$ under MCAR and MAR mechanisms, Architecture B.", out.width = "80%"}
knitr::include_graphics("../../figures/02xs-sens-S3.pdf")
```

Both specifications lose power monotonically as the dropout rate
increases. At the MCAR baseline ($N = 70$, no dropout), E2 gives
$0.866 \pm 0.015$ and E3 gives $0.764 \pm 0.019$; at an MCAR dropout
rate of 0.30, these collapse to $0.148 \pm 0.016$ and
$0.124 \pm 0.015$ respectively. E2's advantage narrows steadily as
dropout worsens: the absolute E2 minus E3 gap runs about 0.10 at
zero dropout, 0.07 at a rate of 0.10, 0.04 at 0.20, and 0.03 at 0.30
under MCAR, narrowing further under MAR dropout at the highest rate
examined (about 0.004 at rate 0.30). This pattern is consistent with
E2's advantage originating in the exposure-weighted contrast
contributed by off-drug observations; when those observations are
preferentially lost, E2's edge contracts toward E3's. At low to
moderate dropout, that is, at a rate of 0.20 or below, the E2
advantage is preserved.

### S4: biomarker-moderation effect-size curve

```{r fig-sens-s4, echo = FALSE, fig.cap = "Power across the biomarker-moderation effect-size range $c_{bm} \\in \\{0, 0.10, 0.20, 0.30, 0.45, 0.60\\}$ under Architecture B. The $c_{bm} = 0$ point is the null for Type I verification.", out.width = "60%"}
knitr::include_graphics("../../figures/02xs-sens-S4.pdf")
```

The power curve is sigmoidal for both specifications, as expected.
At $c_{bm} = 0$, empirical Type I error sits below nominal, at
$0.028 \pm 0.007$ for E2 and $0.030 \pm 0.008$ for E3; power then
rises through the mid-range and saturates once $c_{bm} \geq 0.45$,
reaching $0.958 \pm 0.009$ for E2 and $0.916 \pm 0.012$ for E3 at
$c_{bm} = 0.6$. The E2 advantage is largest at intermediate effect
sizes, $+0.082$ absolute at $c_{bm} = 0.30$ and $+0.102$ at
$c_{bm} = 0.45$, shrinking to $+0.042$ at $c_{bm} = 0.60$ as both
specifications approach the ceiling. The practical implication is
that the advantage matters most in trials targeting a moderate
biomarker-treatment interaction, and matters least when the true
effect is either very small or comfortably large.

## A higher-precision check

Because a number of the Tier 1 gaps are small, we reran a focused
24-cell grid (OL+BDC design, $N \in \{35, 70\}$, true half-life
$\in \{0.5, 1.0\}$) at 600 trials per cell, with all trials
converging. Because E2 and E3 are fit to the same trials, we can
compare them with a paired McNemar test, which is considerably more
sensitive than treating the two specifications as independent
groups. Type I error across the twelve null cells of the rerun fell
in $[0.003, 0.053]$, so both specifications hold nominal alpha at
the sample sizes and half-lives examined. Paired McNemar tests on E2
versus E3 find E2 strictly ahead only at the highest-leverage cell
($N = 70$, true $t_{1/2} = 1.0$ weeks, $\Delta = +7.2$ percentage
points, $p = 1.2 \times 10^{-7}$); in the remaining three cells the
paired difference lies between $-0.5$ and $+1.5$ percentage points,
with $p \geq 0.18$, and the two specifications are operationally
indistinguishable at this level of precision. Resolving the
E2-versus-E3 gap at the intermediate cells with statistical clarity
would require on the order of 2,000 replicates per cell, given the
current Monte Carlo standard error.

# Discussion

## The principal finding, and the conditions under which it fails

The principal finding is that E2's advantage over E3 is real, but
conditional. On the one hand, at its best cell (Hybrid design,
$N = 70$, $c_{bm} = 0.45$, $t_{1/2} = 1.0$), E2 exceeds E3 by about
10 percentage points of power (0.860 versus 0.770), a difference of
roughly five Monte Carlo standard errors and therefore not
attributable to chance. On the other hand, under the crossover
design E2 performs considerably worse (0.488 versus 0.830 for E3).
The reason for this reversal appears straightforward: in a crossover
trial, the within-subject correlation already carries most of the
signal that E2's decaying predictor was designed to recover, so E2
gains little from its more elaborate representation of exposure, and
its down-weighting of off-drug observations works against it. The
recommendation that E2 is to be preferred therefore holds only in
the two non-crossover designs studied here, and should not be
extended to a crossover trial.

## What the robustness checks add

Three checks probe whether E2's advantage, where it exists, is
fragile. On decay shape (Figure \@ref(fig:fig-heatmap-mismatch)):
even when the analyst incorrectly assumes exponential decay and the
true form is Weibull, E2 still does not fall below E3, within the
range of decay shapes we examined. On the assumed half-life (Figure
\@ref(fig:fig-sens-s2)): E3 does not use a half-life and is
therefore unaffected, while E2 does, yet mis-specifying the
half-life by a factor of two or four moves E2's power only from
about 0.83 down to 0.77, never below E3. On autocorrelation (Figure
\@ref(fig:fig-sens-s1)): stronger serial correlation raises both
specifications by roughly the same amount, so E2's lead of five to
six points holds regardless of correlation strength. Taken together,
these checks suggest that where E2's advantage exists, it is not a
fragile artifact of one particular combination of nuisance
parameters.

## Dropout matters more than the method

The finding with the largest practical consequence in this study
concerns missing data, not carryover. Unfortunately, when patients
drop out (Figure \@ref(fig:fig-sens-s3)), power collapses for both
specifications together: at 30% dropout, both fall to somewhere
between 0.12 and 0.25 and become essentially indistinguishable. E2
offers no protection against this loss; it can only organize
whatever data happen to remain. Dropout tied to symptom severity
(MAR) is somewhat worse than dropout occurring at random (MCAR),
because the sickest patients, who carry the most signal, tend to
leave first. The practical implication is direct: retaining patients
in the trial yields considerably more statistical power than the
choice between E2 and E3 ever could.

## Scope, related work, and limits

This manuscript deliberately narrows the full-scope grid to
Architecture B, to the non-linear decay forms, and to the
E2-versus-E3 contrast, in the interest of a self-contained
evaluation of manageable length. The archived full-scope drafts
(`archive/report.Rmd` and `archive/report-slim.Rmd`)
show that the E2 advantage documented here does not carry over to
Architecture A, where the biomarker moderates the mean response
directly and E2 is marginally dominated in every design we examined;
nor is the binary on-drug specification E1 dominated by E2 outside
the Hybrid and OL+BDC cells under Architecture B considered here.
Readers who need the cross-architecture comparison, or the
comparison against E1, should consult that manuscript rather than
extrapolate from this one.

Within the narrower scope adopted here, our matched-decay power at
the headline cell, $0.860 \pm 0.016$, agrees with Hendrickson's
[-@hendrickson2020] reported 0.82 for the same cell. The crossover
tradition surveyed by Jones and Kenward [-@jones2014crossover] would
lead one to expect that a shape-free lagged term such as E3 should
outperform a parametric predictor such as E2 once the decay shape is
mis-specified. Our S2 result does not support that expectation over
the range we examined, because E2's penalty for an incorrect
half-life appears too small for E3 to overtake it.

Two further limitations bound what we can conclude. First, the
numbers presented here come from a single parameter set, calibrated
to a prazosin and post-traumatic stress disorder trajectory, so the
absolute power values should be regarded as illustrative rather than
general; the ranking is the more portable part of the result, and
even the ranking is conditional on the trial design, as we have
shown above. Second, the sensitivity checks vary one factor at a
time and can therefore miss genuine interactions between factors,
for instance between high autocorrelation and heavy dropout; the one
joint check available to us, reported in the full-scope manuscript
as block S5, revealed no such interaction.

# Conclusions

1. **E2 helps, but conditionally.** The exposure-weighted predictor
   achieves the highest power, about 10 points over E3, only in the
   Hybrid and OL+BDC designs, and its lead survives an incorrectly
   assumed decay shape and an incorrectly assumed half-life within
   those designs. Under the crossover design, E2 confers no
   advantage, and E3 is at least as good. Where E2 is used, we
   recommend reporting it with a CR2 cluster-robust standard error
   (Block S6).

2. **An approximate half-life is adequate, where E2 is used.**
   Mis-specifying the half-life by a factor of two to four moves
   E2's power only modestly, so a rough pharmacokinetic estimate is
   sufficient in practice.

3. **Dropout matters more than either method.** Power falls sharply
   as dropout increases, roughly halving at a dropout rate of 20%,
   and this loss affects both specifications equally. Preventing
   dropout is a more effective use of design effort than choosing
   between E2 and E3.

4. **The interaction test is somewhat conservative, and the
   conservatism can be corrected.** Both specifications reject
   slightly less often than the nominal 5% under the null, and a CR2
   cluster-robust standard error restores the correct rate while
   recovering a modest amount
   of power. The E2-versus-E3 gap is held or widened under CR2 at
   four of the five S6 stress cells, but collapses to a statistical
   tie under 30% dropout, consistent with the third conclusion above
   (Block S6).

5. **This is a narrowed slice of a larger study.** Architecture A,
   the binary specification E1, and linear decay all fall outside
   the scope of this manuscript; the full-scope grid reported in
   the archived drafts `archive/report.Rmd` and
   `archive/report-slim.Rmd` is the authority on those comparisons,
   and should be consulted before generalizing beyond
   Architecture B.

# Conflict of interest

The authors declare no conflicts of interest.

# Funding

This work received no specific funding.

# Data availability statement

The data and code supporting the findings of this study are
available in the pmsimstats-ng project repository. Per-replicate
Monte Carlo outputs and ADEMP-compliant cell-level summaries are
versioned under `analysis/data/`; simulation drivers are at
`analysis/scripts/carryover-sensitivity/`. Exact paths for this
manuscript are listed in the Reproducibility section below.

# Reproducibility

This manuscript was rendered with R 4.6.0 inside the project's
zzcollab Docker container; the production simulation itself ran
under R 4.5.3, as recorded in §3.6, with the image hash recorded in
`Dockerfile` and the package set captured in `renv.lock`. Principal
packages used by the simulation pipeline are `nlme` (continuous-time
linear-mixed analysis with `corCAR1` residual correlation),
`parallel` (8-core `mclapply` for parameter-grid sweeps),
`data.table` (the `original-extended` simulation collection), `furrr`
and `purrr` (the `tidyverse` simulation collection), and `rmarkdown`
plus `bookdown` for the manuscript build. No simulation code runs
during knitting; the manuscript loads pre-computed checkpoints only.

**Random-number generator.** At the top of the replicate loop, each
simulation driver sets the master seed to 20260507 and uses
`parallel::mc.set.seed` inside `mclapply`. Per-replicate seeds are
derived from the master seed as \texttt{master + rep\_idx}.

**Data and code availability.** All paths below are relative to the
repository root. Production-grade simulation outputs are
`01-factorial.rds` (principal factorial) and `04-sensitivity.rds`
(sensitivity blocks S1 through S5), both under the driver output
directory `analysis/scripts/carryover-sensitivity/output/`.
Cell-level summaries are `02-grid-summary.rds` and
`02-sensitivity-summary.rds` under `analysis/data/`. Six scripts
under `analysis/scripts/carryover-sensitivity/` generate everything
used here:

- `01-run-factorial.R` (Tier 1 principal factorial),
- `04-run-sensitivity-blocks.R` (Tier 2 sensitivity blocks),
- `02-summarise-grid.R` (cell-level summaries),
- `03-render-figures-extra-slim.R`,
- `04-sensitivity-figures-extra-slim.R`, and
- `07-type1-cr2-figure-extra-slim.R` (the three Architecture B
  filtered figure scripts).

The same directory holds `simulation-grid-plan.md`, which documents
the grid design and serves as the Morris ADEMP
pre-registration. High-precision
prototype rerun outputs are `02-carryover-replicates.rds`
(per-replicate) and `02-carryover-summary.txt` (Morris ADEMP
cell-level summary), under `analysis/data/quick-sim/`. The
manuscript source, its bibliography (`references.bib`), and the SIM
CSL file are in `analysis/report/` `02-carryover-sensitivity/`;
superseded wider-scope drafts are retained under `archive/` in that
directory and are documented in its `README.md`.

**Specification labels.** The three specifications are recorded as
`A1`, `A2`, `A3` in the archived simulation output and are reported
here as E1, E2, E3; the mapping is applied at display time and the
stored values are unchanged.

# References

::: {#refs}
:::

<!--
  No in-document footer chunk. The page footer (source path, render
  time, git version) is injected at render time by tools/stamp.tex
  via the knit: hook above, which wraps long paths correctly.
-->
