---
title: "Three data-generating architectures for biomarker-treatment
  interactions in aggregated N-of-1 trials, and why the choice
  changes power under carryover: a tutorial"
author: 'pmsimstats team'
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  pdf_document:
    latex_engine: xelatex
    toc: false
    number_sections: true
    keep_tex: true
    highlight: tango
    includes:
      in_header: claudecode.tex
fontsize: 11pt
geometry: "left=2.4cm,right=2.4cm,top=2.2cm,bottom=2.2cm"
bibliography: references.bib
csl: statistics-in-medicine.csl
link-citations: true
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{booktabs}
  - \usepackage{setspace}
  - \setstretch{1.0}
knit: (function(input, ...) { d <- dirname(input); while (!file.exists(file.path(d, 'tools', 'stamp-render.R')) && d != dirname(d)) d <- dirname(d); source(file.path(d, 'tools', 'stamp-render.R'))$value(input) })
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)
```

\section{Abstract}

\textbf{Background.} Simulation is the standard tool for sizing
aggregated N-of-1 trials -- designs in which each participant is
repeatedly crossed on and off treatment and many single-patient
series are pooled -- intended to validate predictive biomarkers.
Any such simulation must encode how the biomarker moderates the
treatment effect in its data-generating process (DGP), and the
encoding is a substantive choice. This tutorial, aimed at applied
statisticians who may be new to the N-of-1 setting, formalizes
\emph{three} architectures and shows that the choice among them,
usually treated as an implementation detail, materially changes the
estimated power loss under carryover.

\textbf{Methods.} We define three DGP architectures.
\emph{Architecture A} (direct mean moderation): the biomarker
scales the treatment effect in the conditional mean.
\emph{Architecture B} (differential correlation, the
multivariate normal (MVN) approach): the interaction is encoded as
treatment-state-dependent correlation in the joint distribution.
\emph{Architecture C} (combined): both channels are active at once,
with independent parameters $c_{bm,A}$ (mean) and $c_{bm,B}$
(correlation), so that A and B are its single-channel special cases.
Holding the analysis model fixed (a linear mixed model with
continuous-time AR(1) residual correlation), we quantify by Monte
Carlo simulation the power to detect the interaction as carryover
increases, across three within-subject designs, calibrated to a
prazosin/PTSD reference application.

\textbf{Results.} Under Architecture A power is nearly preserved
because the biomarker-effect proportionality is maintained at every
occasion: at the reference cell it falls only from 0.751 to 0.730
(2.8\% relative) as the carryover half-life grows to one week.
Under Architecture B the same carryover erodes the differential
correlation that carries the signal, and power falls from 0.777 to
0.539 (30.6\% relative), roughly eleven times the Architecture A
loss and far beyond Monte Carlo error
($z = 7.64$, $p < 10^{-13}$). The discrepancy is strongly
design-dependent: large in designs with a discontinuation phase,
negligible in a classical crossover whose within-subject AR(1)
structure already dominates the correlation channel. Architecture C
behaves intermediately, its carryover sensitivity rising with the
weight $c_{bm,B}$ placed on the correlation channel.

\textbf{Conclusions.} The DGP architecture is a substantive
scientific assumption, not a parameterization convenience: it
determines how severely carryover appears to reduce power. We
recommend that biomarker-validation simulation studies state which
of the three architectures they adopt and, when the mechanism is
uncertain, report power across the relevant architectures.

\bigskip\noindent\rule{\textwidth}{0.4pt}\bigskip

\section{1. Introduction}

::: {.bullets data-latex=""}
\textbf{Power for these designs is estimated by simulation, and the
simulation must encode the biomarker interaction in its DGP -- in
more than one possible way.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item The target is the biomarker-treatment interaction.
\item Closed-form power exists only in simplified cases
[@pmsimstats-paper08]; for the designs and model here it is obtained by
Monte Carlo simulation.
\item How the interaction is built into the DGP is a modeling
choice, the subject of this paper.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
Precision-medicine trials increasingly ask whether a baseline
**biomarker** modifies the treatment effect -- a
biomarker-treatment interaction. For simplified special cases, such
as a balanced strict RM-ANOVA under sphericity, the power of the
interaction test does have a closed form, and a companion paper
[@pmsimstats-paper08] derives it together with its equivalence to a
two-sample $t$-test. For the continuous-time mixed model and the
N-of-1 designs considered here, however, no such closed form is
available, so power is estimated by Monte Carlo simulation: many
trials are generated, analyzed, and the rejection rate recorded. The point
that motivates this paper is that the simulation must itself encode
the interaction in its data-generating process, and there is more
than one way to do so.
:::

::: {.bullets data-latex=""}
\textbf{There are three architectures: A (mean moderation), B
(differential correlation), and C, which activates both channels at
once.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item Architecture A: an interaction term scales the treatment
effect in the mean; standard across trial-simulation frameworks.
\item Architecture B: the biomarker changes the response-outcome
correlation, encoding the interaction in the joint distribution
(Hendrickson's N-of-1 approach).
\item Architecture C: both channels act together, with A and B as
its single-channel special cases.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
We distinguish three architectures. The common one, **direct mean
moderation** (Architecture A), adds an interaction term to the
outcome equation so the biomarker scales the size of the treatment
effect; essentially every standard trial-simulation framework uses
it [@simon2010; @freidlin2014]. The second, **differential
correlation** (Architecture B), used by Hendrickson et al.
[@hendrickson2020] for N-of-1 trials, instead makes the biomarker
change the correlation between treatment response and outcome, so
the interaction is encoded in the joint distribution rather than in
the mean; on this reading it is the reduced-form second-moment
encoding of a latent responder-class mechanism (Section 4;
[@pmsimstats-paper03]). The third, **combined** (Architecture C), activates both
channels simultaneously through two independent parameters; A and B
are exactly its single-channel limits. We treat all three as
co-equal data-generating architectures, defined precisely in
Section 2.

In plain language, without the latent-variable vocabulary: suppose
patients come in two kinds, responders and non-responders, but we
cannot see which kind a given patient is. The biomarker is only an
imperfect clue to that hidden status, so two patients with the same
biomarker value can still respond differently. Architecture A cannot
represent this, because it assumes the biomarker fixes each
patient's drug effect exactly; Architecture B represents it by
letting biomarker and response track each other only while the drug
acts, as a tendency rather than a fixed rule. That tendency fades
once the drug washes out, which is why carryover costs more power
under Architecture B than under Architecture A (Section 4 gives
clinical examples).
:::

::: {.bullets data-latex=""}
\textbf{The two encodings, found in two implementations of the same
software, gave different carryover stories; this paper explains
why.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item The original code used Architecture B; a re-implementation
used Architecture A.
\item The two disagreed about how much carryover hurts power.
\item We formalize both and give guidance on which to use.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
We ran into both while building the pmsimstats software: the
original code used Architecture B, while a re-implementation had
quietly switched to Architecture A, and the two gave different
stories about carryover. This paper pins down why. We define the
two architectures precisely, show by simulation that they react
very differently to carryover, and give simple guidance on which to
use.
:::

::: {.bullets data-latex=""}
\textbf{Primer: aggregated N-of-1 trials pool repeated
within-patient on/off contrasts, and carryover is their
distinguishing nuisance.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item A single N-of-1 trial repeatedly crosses one patient on/off;
pooling many gives population inference at small $N$.
\item Carryover is residual drug effect decaying over days to weeks,
contaminating off-drug occasions.
\item Three within-subject designs (crossover (CO), open-label
with blinded discontinuation (OL+BDC), Hybrid) differ in
how on/off occasions are arranged.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
\textbf{The N-of-1 setting, for readers new to it.} In a single
N-of-1 trial one participant is repeatedly crossed between active
treatment and a comparator over a sequence of measurement
occasions, so the within-participant on-versus-off contrast
estimates that participant's individual treatment effect; an
\emph{aggregated} N-of-1 trial pools many such series in a mixed
model to obtain population-level inference at far smaller sample
sizes than a parallel-group design. The feature that distinguishes
these designs analytically is \emph{carryover}: residual drug effect
that decays over days to weeks after discontinuation, contaminating
off-drug occasions. We study three within-subject designs that
differ in how on- and off-drug occasions are arranged -- a
traditional crossover (CO), an open-label-plus-blinded-
discontinuation design (OL+BDC), and a Hybrid N-of-1 design --
defined in Section 3.
:::

\bigskip\noindent\rule{\textwidth}{0.4pt}\bigskip

\section{2. Three DGP Architectures}

::: {.bullets data-latex=""}
\textbf{Notation: $B_i$ biomarker, $D_{it}$ on-drug indicator,
$BR_{it}$ biological response, with \texttt{Dbc} the analysis-layer
exposure.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item $B_i$, $D_{it}$, $BR_{it}$ describe the data-generating layer.
\item The analysis layer uses a continuous exposure \texttt{Dbc}
(Section 3.1).
\item Typewriter identifiers denote code/formula variables;
math symbols are used otherwise.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
\textbf{Notation.} Throughout this paper, $B_i$ denotes the
biomarker value of participant $i$, $D_{it}$ the binary on-drug
indicator at timepoint $t$ (1 on drug, 0 off drug), and
$BR_{it}$ the biological response component of the outcome.
These three symbols describe the data-generating-process layer.
The analysis-model layer uses a related but distinct continuous
drug exposure variable, \texttt{Dbc}, defined in Section 3.1.
Code-style identifiers in typewriter font (\texttt{bm},
\texttt{Dbc}, \texttt{BR}) refer to specific variables in the
pmsimstats codebase or to R formula syntax; mathematical
symbols are used otherwise.
:::

\subsection{2.1 Architecture A: Direct Mean Moderation}

In this approach the biomarker enters the mean structure of the
outcome directly. The treatment effect for participant $i$ at
timepoint $t$ is:

\[Y_{it} = \mu_0 + \beta_1 \cdot D_{it} \cdot (1 + \beta_{bm}
\cdot B_i) + \epsilon_{it}\]

::: {.bullets data-latex=""}
\textbf{Under Architecture A the interaction is explicit in the
mean: a higher biomarker gives a proportionally larger treatment
effect.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item $\beta_{bm}$ is the biomarker moderation parameter on the
centered biomarker.
\item The effect scales with $B_i$ deterministically in the
conditional mean.
\item The signal lives in the first moment of the outcome.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
where $D_{it}$ is the binary on-drug indicator (or a continuous
measure of drug exposure), $B_i$ is the participant's biomarker
value (typically centered), and $\beta_{bm}$ is the biomarker
moderation parameter. The interaction is explicit in the mean:
participants with higher biomarker values have proportionally
larger treatment effects.
:::

Key properties:

\begin{itemize}
\item The interaction is \textbf{deterministic} given the biomarker value
  and treatment status.
\item The biomarker moderation $\beta_{bm}$ is a \textbf{population-level
  parameter} that applies uniformly to all participants.
\item The signal for the interaction is in the \textbf{first moment}
  (conditional mean) of the outcome distribution.
\item Adding noise, random effects, or carryover to $D_{it}$ scales
  the entire treatment effect (including its biomarker-dependent
  component) proportionally. The \textbf{ratio} of drug effect between
  high-biomarker and low-biomarker participants is preserved.
\end{itemize}

::: {.bullets data-latex=""}
\textbf{The multiplicative form maps onto the standard additive
effect-modeling regression, so both are instances of Architecture
A.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item The additive interaction $\beta_3$ equals
$\beta_1 \cdot \beta_{bm}$ in the multiplicative form.
\item The additive form adds a prognostic main effect $\beta_2$ the
multiplicative form sets to zero.
\item For interaction power the two are interchangeable up to that
prognostic distinction.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
We note that the multiplicative form above maps directly onto
the standard additive regression form for predictive biomarker
effects used in the trial methodology literature, specifically
the effect modeling equation of Kent et al. [@kent2020path]:
\[Y_{it} = \beta_0 + \beta_1 D_{it} + \beta_2 B_i + \beta_3
D_{it} B_i + \epsilon_{it}\]
The interaction coefficient $\beta_3$ in the additive form
corresponds to the product $\beta_1 \cdot \beta_{bm}$ in the
multiplicative form. The additive form additionally
accommodates a prognostic main effect $\beta_2$ for the
biomarker, which the multiplicative form sets to zero by
construction. For the purpose of power calculations on the
interaction coefficient, the two forms are interchangeable up
to this prognostic-effect distinction, and both fall within
Architecture A.
:::

This architecture is used in the pedagogical simulation
(\texttt{implementations/simple/simulation.R}) and in the
exploratory carryover analyses
(\texttt{simulation\_carryover\_spectrum.R}) of the
pmsimstats project.

\subsection{2.2 Architecture B: Differential Correlation (MVN)}

In this approach the biomarker and the biological response (BR)
component are jointly drawn from a multivariate normal
distribution with treatment-state-dependent correlation:

\[\begin{pmatrix} B_i \\ BR_{it} \end{pmatrix} \sim
\text{MVN}\left(\begin{pmatrix} \mu_B \\ \mu_{BR}(t, D_{it})
\end{pmatrix}, \Sigma(D_{it})\right)\]

where the covariance matrix $\Sigma$ has the following
treatment-state-dependent BM-BR correlation:

\[\text{Cor}(B_i, BR_{it}) = \begin{cases}
c_{bm} & \text{if } D_{it} = 1 \text{ (on drug)} \\
c_{bm} \cdot e^{-\lambda \cdot t_{sd}} &
\text{if } D_{it} = 0 \text{ (off drug)}
\end{cases}\]

where $t_{sd}$ is time since drug discontinuation. The
exponential-decay form is the default; the limiting cases
$\lambda = 0$ (no decay; correlation persists indefinitely
during the off-drug phase) and $\lambda \to \infty$ (immediate
decay; correlation drops to zero the instant the drug is
discontinued) recover the two no-carryover model variants.

The interaction is not in the mean structure: the population
mean of $BR_{it}$ is the same for all participants with the
same treatment history. Instead it emerges from the
\textbf{conditional distribution}: given a participant's
biomarker value $B_i = b$, the conditional expectation of
their biological response is:

\[E[BR_{it} | B_i = b, D_{it} = 1] = \mu_{BR}(t) + c_{bm}
\cdot \frac{\sigma_{BR}}{\sigma_B} \cdot (b - \mu_B)\]

::: {.bullets data-latex=""}
\textbf{Under Architecture B the interaction is in the second
moment: the biomarker modulates BR through a correlation that
decays during the off-drug period.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item The conditional mean of BR rises with $b$ via the correlation,
not a mean parameter.
\item Effective correlation is $c_{bm}$ on-drug, $c_{bm}
e^{-\lambda t_{sd}}$ off-drug.
\item Moderation persists with attenuated strength off-drug and
vanishes only at complete decay.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
This conditional expectation has an interaction-like structure:
the biomarker value $b$ modulates the expected BR. The
effective correlation is $c_{bm}$ on drug and $c_{bm} \cdot
e^{-\lambda t_{sd}}$ during the off-drug period; the moderation
therefore persists with attenuated strength during off-drug
periods and vanishes only in the limit of complete decay.
:::

Key properties:

\begin{itemize}
\item The interaction is \textbf{probabilistic}: it manifests as a
  tendency, not a deterministic scaling.
\item The biomarker moderation $c_{bm}$ is a \textbf{correlation
  parameter} that governs the strength of the association.
\item The signal for the interaction is in the \textbf{second moment}
  (covariance structure) of the joint distribution.
\item Carryover effects in the DGP inflate the BR means during
  off-drug periods, making them resemble on-drug means. This
  reduces the \textbf{differential} in the correlation structure
  between treatment states, weakening the signal that the
  analysis model detects.
\end{itemize}

This architecture is used in the Hendrickson et al. [@hendrickson2020]
publication code and the audited \texttt{R/} package in
pmsimstats-ng.

\subsection{2.3 Architecture C: Combined Mean Moderation and
Differential Correlation}

Architecture C (combined) applies both interaction mechanisms in
sequence. First, the MVN covariance matrix is constructed with
BM-BR differential correlation weighted by \texttt{c.bm\_b}, exactly
as under Architecture B. Second, after the MVN draw, the BR
component is additively shifted at on-drug timepoints by

\[
\Delta BR_{it} = c_{bm,A} \cdot z_{B,i} \cdot \sigma_{BR}
  \cdot \mathbf{1}[\text{on-drug}]
\]

where $z_{B,i} = (B_i - \mu_B)/\sigma_B$ is the standardized
biomarker for participant $i$, exactly as under Architecture A.

The two parameters \texttt{c.bm\_a} and \texttt{c.bm\_b} are
independent. Setting $c_{bm,B} = 0$ removes the covariance channel
and recovers pure Architecture A; setting $c_{bm,A} = 0$ removes
the mean-shift channel and recovers pure Architecture B. The
boundary cases are enforced algebraically rather than by convention,
so the special cases serve as internal validation gates for any
simulation study using Architecture C.

We note that when both parameters are set to their full values
($c_{bm,A} = c_{bm,B} = 0.45$), the total interaction effect
on the BR mean is approximately double that of either pure
architecture at $c_{bm} = 0.45$. A researcher wishing to hold total
effect size constant across architectures may instead parameterize
as $c_{bm,A} = (1-\alpha) \cdot c$ and
$c_{bm,B} = \alpha \cdot c$ for a mixing weight $\alpha \in [0,1]$
and total weight $c$. We do not adopt this constraint in the
simulations below, preferring the factorial parameterization that
spans the full $(c_{bm,A}, c_{bm,B})$ space.

::: {.bullets data-latex=""}
\textbf{The two channels are mechanistically distinct: A is a
pharmacodynamic mean shift, B a co-regulatory coupling in the
covariance.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item Channel A: the biomarker governs effective dose/receptor
occupancy, amplifying the mean drug effect.
\item Channel B: on-drug periods couple biomarker and response, a
coupling absent in the off-drug period.
\item They need not scale in proportion, motivating the combined
Architecture C.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
The two channels are mechanistically distinct and need not scale
in proportion. The mean-shift channel (Architecture A) reflects a
pharmacodynamic relationship: the drug's biological effect on BR
is amplified in participants whose biomarker is above the
population mean, because the biomarker governs the effective dose
or receptor occupancy. This mechanism operates at the level of the
mean trajectory for each participant.

The covariance channel (Architecture B) reflects a different
process: on-drug periods impose a co-regulatory coupling between
biomarker expression and treatment response, perhaps because the
drug activates a pathway whose output is gated by the biomarker
signal. This coupling is absent during off-drug periods, producing
the differential correlation structure that Architecture B encodes
in the MVN covariance matrix.
:::

A given drug could activate one, both, or neither channel. The
simulation practitioner who is uncertain which channel predominates
should regard Architecture C with a range of $\alpha$ values as
the natural sensitivity framework.

\subsection{2.4 Mathematical Comparison}

We begin the comparison by considering the expected outcome
difference between two participants with biomarker values
$b_1$ and $b_2$ at the same drug exposure level $D$ under
Architecture A:

\[E[Y \mid b_1, D] - E[Y \mid b_2, D] = \beta_1 \cdot \beta_{bm}
\cdot (b_1 - b_2) \cdot D\]

The absolute magnitude of this difference scales linearly with
$D$ and therefore shrinks during off-drug periods when $D$ is
replaced by an exposure-decayed value $D \cdot \text{carryover}$.
We note, however, that the \textbf{ratio} of the drug effect
between high-biomarker and low-biomarker participants is
preserved at every level of exposure. The
biomarker-treatment signal contracts in amplitude but does
not lose its identifiability under carryover.

Under Architecture B, the analogous quantity is:

\[E[BR | b_1, D=1] - E[BR | b_2, D=1] = c_{bm} \cdot
\frac{\sigma_{BR}}{\sigma_B} \cdot (b_1 - b_2)\]

During off-drug periods with carryover:

\[E[BR | b, D=0, t_{sd}] = \mu_{BR,\text{off}}(t_{sd}) +
c_{bm} \cdot e^{-\lambda t_{sd}} \cdot
\frac{\sigma_{BR}}{\sigma_B} \cdot (b - \mu_B)\]

Here the biomarker-dependent component decays exponentially
with time off drug. As $t_{sd}$ increases, the off-drug
conditional expectation converges to $\mu_{BR,\text{off}}$
for all biomarker values, eliminating the differential signal
entirely.

Under Architecture C both channels are active. The
analysis model's interaction coefficient depends on the covariance
between $B_i$ and $D_{bc}$-weighted outcomes. The mean-shift
channel contributes a term that degrades exactly as under
Architecture A: the \texttt{Dbc} contrast is compressed during
carryover but the ratio of the drug effect between biomarker groups
is preserved at every exposure level. The covariance channel
contributes a term that degrades as under Architecture B: the
BM-BR correlation generating the second-moment signal decays
exponentially off-drug. The two channels are additive at the
level of the total biomarker-by-drug covariance, so the net power
loss under Architecture C lies strictly between the pure-architecture
extremes when $c_{bm,A}, c_{bm,B} > 0$. As either parameter
approaches zero the corresponding boundary case is recovered.

\bigskip\noindent\rule{\textwidth}{0.4pt}\bigskip

\section{3. Consequences for Statistical Power Under Carryover}

\subsection{3.1 Empirical demonstration}

We ran identical simulation configurations under both
architectures using the pmsimstats-ng framework, matching the
DGP parameters as closely as possible between the two
approaches. We note that the non-biologic response components
(time-varying response, placebo-belief accumulation, and
residual noise) are generated from the same multivariate
normal draw with identical means, variances, within-factor
AR(1) autocorrelation, and cross-factor correlations in both
architectures. The two DGPs differ \emph{only} in how the
biomarker-to-biological-response (BM-BR) linkage is encoded:
Architecture~B embeds it in the off-diagonal $c_{bm}$ entries of
$\Sigma$, whereas Architecture~A omits those entries and adds a
post-draw shift $c_{bm} \cdot B_i^{*} \cdot \sigma_{BR}$ to
on-drug BR columns (see
\texttt{19-mean-moderation-implementation-notes.tex} for the
scaling derivation). The divergent carryover behavior reported
below is therefore attributable to the BM-BR parameterization,
not to any difference in how the placebo, time-varying, or
residual-noise components are generated. The analysis model was
the same in both cases: an \texttt{nlme::lme} fit with a
\texttt{corCAR1} continuous-time autocorrelation structure and
a single biomarker-by-drug interaction term, written
\texttt{bm:Dbc} in R formula syntax.

The analysis-model drug exposure variable \texttt{Dbc} is the
\emph{continuous} analogue of the binary DGP indicator $D_{it}$.
It equals $1$ during on-drug periods and $\exp(-\lambda t_{sd})$
during off-drug periods, where $t_{sd}$ is time since drug
discontinuation and $\lambda$ is the analysis-model carryover
decay rate (typically derived from the drug's pharmacokinetic
half-life as $\lambda = \ln 2 / t_{1/2}$, in which case it
matches the DGP decay rate of Section 2.2 by construction).
Whereas $D_{it}$ flips abruptly between 0 and 1, \texttt{Dbc}
captures the smooth shrinkage of residual drug effect during
the off-drug period. The interaction coefficient on \texttt{bm:Dbc} is
therefore the test of whether the biomarker moderates the
exposure-decayed drug effect.

We follow the ADEMP framework of Morris, White, and Crowther
(2019): Aims, Data-generating mechanisms, Estimands, Methods,
Performance measures.

\textbf{Setup:}

\begin{itemize}
\item Trial designs: CO (traditional crossover), Hybrid (N-of-1
  Hybrid), OL+BDC (open-label with blinded discontinuation)
\item Sample size: $N = 35$ per randomization path (effective
  analysis $N = 70$ for CO and OL+BDC, 140 for Hybrid, after the
  paths are concatenated)
\item Biomarker correlation / moderation: $c_{bm} = 0.45$ /
  $\beta_{bm} = 0.45$
\item DGP carryover half-life: $t_{1/2} \in \{0, 0.5, 1.0\}$ weeks
\item Analysis: matched \texttt{Dbc} with same half-life
\item Replicates: 1{,}000 per cell (the Monte Carlo standard
  error (MCSE)
  on a binomial power estimate near $0.5$ is approximately
  $\sqrt{0.5 \cdot 0.5 / 1{,}000} \approx 0.016$)
\item Implementation: 8-core \texttt{parallel::mclapply}; full
  factorial 2 architectures $\times$ 3 designs $\times$ 3
  $t_{1/2}$ $\times$ 2 $c_{bm}$ levels = 36 cells, 36{,}000
  fits, 100\% convergence
\end{itemize}

The $N = 35$ choice is the smallest of the prazosin-PTSD-
calibrated sample sizes used in the published methodological
literature; it is a regime in which neither architecture
saturates power at the no-carryover cells, so any
carryover-induced erosion is mechanically visible. Type I error
under the $c_{bm} = 0$ null cells sat in $[0.010, 0.074]$
across the 18 null cells. These rates below the nominal 0.05 are
not architecture-specific; they reflect a known conservatism of the
standard \texttt{corCAR1} Wald test, whose model-based standard
error is over-stated by roughly 6 to 10 percent, characterized in a
companion calibration study [@pmsimstats-paper10] together with a
cluster-robust remedy. Being common to both architectures, the
conservatism does not affect the architecture contrast.

\textbf{Results under Architecture A (direct mean moderation):}

\begin{table}[htbp]
\centering
\begin{tabular}{lccc}
\toprule
Design & $t_{1/2} = 0$ & $t_{1/2} = 0.5$ & $t_{1/2} = 1.0$ \\
\midrule
CO & 0.739 & 0.739 & 0.725 \\
Hybrid & 0.992 & 0.987 & 0.978 \\
OL+BDC & 0.751 & 0.748 & 0.730 \\
\bottomrule
\end{tabular}
\end{table}

::: {.bullets data-latex=""}
\textbf{Under Architecture A the carryover-induced power loss is
modest in every design (1-3\% relative).}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item Relative loss to $t_{1/2} = 1$: 1.9\% (CO), 1.4\% (Hybrid),
2.8\% (OL+BDC).
\item Below the 8-13\% range earlier exploratory runs suggested.
\item Power is essentially intact under carryover.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
Power declines modestly under Architecture A: relative loss
from $t_{1/2} = 0$ to $t_{1/2} = 1$ is 1.9\% (CO), 1.4\%
(Hybrid), and 2.8\% (OL+BDC). We note that all three values
fall well below the 8-13\% relative-loss range suggested by
earlier exploratory runs in this project for
direct-mean-moderation DGPs, though they reproduce the qualitative
finding of those runs, namely a modest erosion that leaves power
essentially intact.
:::

\textbf{Results under Architecture B (differential correlation,
MVN):}

\begin{table}[htbp]
\centering
\begin{tabular}{lccc}
\toprule
Design & $t_{1/2} = 0$ & $t_{1/2} = 0.5$ & $t_{1/2} = 1.0$ \\
\midrule
CO & 0.745 & 0.754 & 0.723 \\
Hybrid & 0.995 & 0.980 & 0.944 \\
OL+BDC & 0.777 & 0.694 & 0.539 \\
\bottomrule
\end{tabular}
\end{table}

::: {.bullets data-latex=""}
\textbf{Under Architecture B the loss is much larger and strongly
design-dependent, reaching 30.6\% at OL+BDC but absent at CO.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item Relative loss: 3.0\% (CO, not significant), 5.1\% (Hybrid),
30.6\% (OL+BDC) -- about 11x the Architecture A loss.
\item The CO cell fluctuates within MC error, consistent with no
architecture-by-carryover interaction there.
\item Below the 40-60\% range earlier runs suggested, but confirming
substantial, design-dependent erosion.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
Power declines more sharply under Architecture B, and most
markedly so in the OL+BDC design: relative loss from
$t_{1/2} = 0$ to $t_{1/2} = 1$ is 3.0\% (CO; not significantly
different from Architecture A's CO loss), 5.1\% (Hybrid), and
\textbf{30.6\% (OL+BDC)}. The CO cell shows a small,
non-monotone fluctuation ($0.745 \to 0.754 \to 0.723$), which
is within one MCSE and is consistent with
the null finding of no significant architecture-by-carryover
interaction in that design. The OL+BDC loss is approximately
11 times the matched Architecture-A OL+BDC loss (2.8\%). It
falls below the 40-60\% relative-loss range suggested by earlier
exploratory runs in this project for the MVN
differential-correlation DGP, but it confirms the qualitative
finding of those runs, namely a substantial and strongly
design-dependent erosion.
:::

We compare the carryover-induced power loss between architectures
using a difference-of-differences $z$-statistic. Let
$\Delta_X = \hat{\pi}_{X,0} - \hat{\pi}_{X,1}$ denote the
within-architecture loss from $t_{1/2} = 0$ to $t_{1/2} = 1.0$
for architecture $X \in \{A, B\}$, estimated from $n = 1{,}000$
independent replicates at each cell. The B-versus-A gap is
$\Delta_B - \Delta_A$, with standard error

\[
\text{SE} = \sqrt{
  \frac{\hat{\pi}_{B,0}(1-\hat{\pi}_{B,0})}{n}
+ \frac{\hat{\pi}_{B,1}(1-\hat{\pi}_{B,1})}{n}
+ \frac{\hat{\pi}_{A,0}(1-\hat{\pi}_{A,0})}{n}
+ \frac{\hat{\pi}_{A,1}(1-\hat{\pi}_{A,1})}{n}}
\]

since the four proportions come from independent replicate sets.
All tests are two-sided. At OL+BDC this gives $z = 7.64$
($p < 10^{-13}$); at Hybrid, $z = 3.96$ ($p < 10^{-4}$); the
CO contrast is not statistically distinguishable from zero
($z = 0.29$) because the CO design's within-subject AR(1)
structure dominates the carryover-mediated correlation decay
channel that Architecture B uses to encode the
biomarker-treatment interaction.

\textbf{Worked reading of the contrast.} At OL+BDC, the
within-architecture carryover loss is $\Delta_A = 0.751 - 0.730 =
0.021$ (2.8\% relative) and $\Delta_B = 0.777 - 0.539 = 0.238$
(30.6\%). The quantity of interest is the difference-of-differences
$\Delta_B - \Delta_A = 0.217$ (about 22 percentage points), and the
$z$-statistic above scales it by its standard error to give
$z = 7.64$ ($p < 10^{-13}$): the architectures' carryover behavior
differs well beyond Monte Carlo error. The same contrast is clearly
positive for the Hybrid design and indistinguishable from zero for
the crossover design.

\subsection{3.2 Why the architectures diverge}

We turn now to a mechanistic account of why the two
architectures produce such different power losses under
carryover.

::: {.bullets data-latex=""}
\textbf{Under Architecture A, carryover only compresses the
treatment contrast; the biomarker signal-to-noise ratio is nearly
preserved.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item Carryover shrinks \texttt{Dbc} and inflates off-drug BR means
(shared with Architecture B).
\item But $\beta_{bm}$ is a fixed mean parameter with drug-state-
independent noise.
\item So the interaction variance falls but the signal-to-noise
ratio degrades only modestly.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
\textbf{Under Architecture A}, carryover reduces the magnitude
of the drug exposure variable seen by the analysis model
(\texttt{Dbc} drops below 1 during the off-drug period), and the mean BR at
off-drug timepoints is inflated by the residual drug effect
exactly as in Architecture B. Both mechanisms compress the
between-state treatment contrast and are shared by the two
architectures. What Architecture A lacks is the second channel
of signal loss: the biomarker moderation coefficient
$\beta_{bm}$ is a fixed population parameter, the biomarker
enters the mean structure directly rather than through a
correlation, and the noise structure is independent of drug
state. The biomarker-by-drug interaction term therefore has
reduced variance (because \texttt{Dbc} has less contrast between
drug states), but the signal-to-noise ratio degrades only
modestly.
:::

\textbf{Under Architecture B}, carryover operates on two levels
simultaneously:

\begin{enumerate}
\item \textbf{Mean blurring}: The BR means during off-drug periods are
  inflated by residual carryover, making them resemble on-drug
  BR means. This reduces the treatment contrast that the analysis
  model uses to distinguish drug effect from no-drug. This channel
  is shared with Architecture A.

\item \textbf{Correlation erosion}: The BM-BR correlation during off-drug
  periods decays exponentially with $t_{sd}$. This is not just a
  statistical modeling choice; it is a structural consequence
  of the DGP. The correlation between biomarker and BR reflects
  the degree to which the drug effect is active; as the drug
  washes out, the correlation weakens because the drug-mediated
  component of BR variance shrinks relative to the
  drug-independent component. This channel is \emph{unique} to
  Architecture B, because only here does the biomarker signal
  live in the second moment of the joint distribution. Under
  Architecture A the drug-mediated and drug-independent
  components of BR variance move in tandem (both are scaled by
  the same \texttt{Dbc}-weighted drug effect), so their ratio,
  and hence the identifiability of the interaction, is preserved
  even as amplitudes shrink.
\end{enumerate}

::: {.bullets data-latex=""}
\textbf{Architecture B's interaction signal is attacked from both
sides, which Architecture A escapes -- hence the divergence.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item The \texttt{bm:Dbc} coefficient depends on the covariance of
$B_i$ with \texttt{Dbc}-weighted outcomes.
\item Under B both the \texttt{Dbc} contrast and the BM-BR
correlation decay together.
\item Under A the second attack is absent, so power is largely
preserved.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
The analysis model's interaction coefficient (the coefficient
on \texttt{bm:Dbc}) depends on the covariance between $B_i$ and
\texttt{Dbc}-weighted outcomes. Under Architecture B, this
covariance is being attacked from both sides: the treatment
contrast in \texttt{Dbc} is compressed, and the BM-BR
correlation that generates the covariance signal is
simultaneously decaying. That Architecture A escapes the second
attack entirely accounts for the divergence in power loss
documented above.
:::

\subsection{3.3 Robustness check at $N = 70$}

::: {.bullets data-latex=""}
\textbf{An $N = 70$ confirmation run checks that the architecture
ordering survives larger samples.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item Section 3.1 uses $N = 35$, where the loss is mechanically
visible (no ceiling saturation).
\item The check runs 12 cells at $N = 70$, 800 replicates each, on
the OL+BDC design.
\item 100\% convergence across 9{,}600 fits.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
The Section~3.1 production results are at $N = 35$, where
neither architecture saturates power and the carryover-induced
loss is mechanically visible. We conducted a confirmation run
at $N = 70$ to check whether the qualitative ordering survives
the saturation that comes with the larger sample size. An
800-replicate-per-cell Monte Carlo run was executed at $N = 70$
across 12 cells (architecture $\in \{\text{MVN},
\text{mean moderation}\} \times c_{bm} \in \{0, 0.45\} \times
t_{1/2} \in \{0, 0.5, 1.0\}$ weeks), all on the OL+BDC
design, using 8-core \texttt{parallel::mclapply}. Wall-clock
483~s; convergence 100\% across 9{,}600 fits.
:::

At $c_{bm} = 0.45$:

\begin{table}[htbp]
\centering
\begin{tabular}{lccc}
\toprule
Architecture        & $t_{1/2} = 0$ & $t_{1/2} = 0.5$ & $t_{1/2} = 1.0$ \\
\midrule
MVN (B)             & 0.978         & 0.945           & 0.885 \\
Mean moderation (A) & 0.976         & 0.973           & 0.959 \\
\bottomrule
\end{tabular}
\end{table}

::: {.bullets data-latex=""}
\textbf{At $N = 70$ the B-versus-A ordering persists, though
absolute magnitudes are compressed by ceiling saturation.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item Architecture B loses 9.3 points ($z = 7.4$); Architecture A
1.7 points ($z = 2.0$).
\item Both sit within 2-5 points of the ceiling at $t_{1/2} = 0$,
truncating the absorbable loss.
\item The $N = 35$ results remain the canonical reference for the
narrative magnitude.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
Within-architecture losses from $t_{1/2} = 0$ to $t_{1/2} = 1$
are 9.3 percentage points (Architecture~B; two-proportion
$z$-test on the paired cells, $z = 7.4$, $p < 10^{-12}$) and
1.7 percentage points (Architecture~A; $z = 2.0$,
$p \approx 0.05$), where each $z$-statistic uses the formula
$(\hat{\pi}_0 - \hat{\pi}_1) /
\sqrt{\hat{\pi}_0(1-\hat{\pi}_0)/n + \hat{\pi}_1(1-\hat{\pi}_1)/n}$
with $n = 800$ replicates per cell. The qualitative ordering matches
Section~3.1's $N = 35$ result. The absolute magnitudes are
compressed because $N = 70$ at the prazosin-PTSD calibration
places both architectures within 2-5 percentage points of the
unit ceiling at $t_{1/2} = 0$, mechanically truncating the
loss either architecture's curve can absorb. We note that
reading the narrative loss-magnitude claim against the $N = 70$
robustness numbers (3\% and 9\% relative loss for
Architecture~B versus 0.4\% and 1.8\% for Architecture~A)
requires acknowledging this saturation effect; the $N = 35$
production results in Section~3.1 are the canonical reference
for the narrative magnitude.
:::

\textbf{Type~I error.} Across all 18 null cells of the $N = 35$
production run, Type~I error sat in $[0.010, 0.074]$, with
mean 0.043 and median 0.042. We observe no
architecture-specific inflation. The uniformly sub-nominal rates
are consistent with the \texttt{corCAR1} working-covariance
miscalibration characterized in the companion calibration study
[@pmsimstats-paper10]; because it affects both architectures alike,
it is common-mode with respect to the architecture contrast.
Architecture~B's
$t_{1/2} = 1.0$ OL+BDC null ($p = 0.010$, slightly
conservative) is the smallest cell; the corresponding
alternative ($p = 0.539$) is the most informative cell of the
Section~3.1 narrative.

\textbf{Estimator bias.} Mean estimated
$\hat\beta_{bm{:}D_{bc}}$ at $c_{bm} = 0.45$ ranged from
$-0.231$ to $-0.281$ across the 18 alternative cells, with the
$N = 35$ Architecture-A OL+BDC cell at $t_{1/2} = 1.0$
producing the largest absolute mean ($-0.281$); the
Architecture~B cells maintained mean near $-0.234$ across
all $t_{1/2}$ values, consistent with the Architecture-B
information channel being eroded through correlation decay
rather than through coefficient drift. Empirical SE of
$\hat\beta_{bm{:}D_{bc}}$ runs 0.05 to 0.10 across cells; the
within-replicate model SE matches at scale, indicating
nominal-95\% Wald coverage holds across the grid.

\subsection{3.4 Architecture C: combined-channel power results}

We ran a factorial simulation on the Architecture~C parameter grid.
The design factors are CO, Hybrid, and OL+BDC; $N = 35$ total
subjects per cell (randomized equally across paths);
$t_{1/2} \in \{0, 0.5, 1.0\}$ weeks; 1000 replicates per cell;
A2 analysis specification. The parameter panel is $c_{bm,A} \times
c_{bm,B} \in \{0, 0.22, 0.45\}^2$ (nine cells per
design--carryover combination). By construction, the boundary
cells ($c_{bm,A} = 0.45$, $c_{bm,B} = 0$) reduce the combined
DGP to the pure Architecture~A functional form, and
($c_{bm,A} = 0$, $c_{bm,B} = 0.45$) reduce it to the pure
Architecture~B form. These boundary cells confirm that the
combined parameterization recovers each limiting case
algebraically. They are not, however, numerical checks against the
power values of \S3.1: the Architecture~C grid here uses $N = 35$
as the total sample size, whereas \S3.1 uses $N = 35$ per
randomization path (effective $N = 70$/140). The boundary-cell
power values are therefore lower than, and not directly comparable
to, the matched pure-architecture cells of \S3.1.

\begin{table}[h]
\small\centering
\caption{Architecture C empirical power: $3 \times 3$ parameter
panel. $N = 35$, carryover $t_{1/2} = 1.0$ week, A2 analysis
specification, 1000 replicates per cell. Here $N = 35$ is the
total sample size, unlike \S3.1 where $N = 35$ is per path
(effective $N = 70$/140), so these power values are not directly
comparable to the \S3.1 tables. Rows index $c_{bm,A}$
(mean-shift channel); columns index $c_{bm,B}$ (covariance
channel). By construction, the boundary column $c_{bm,B} = 0$
corresponds to the pure Architecture A form; the boundary row
$c_{bm,A} = 0$ corresponds to the pure Architecture B form.}
\label{tab:arch-c-power}
\begin{tabular}{lrrr}
\toprule
& \multicolumn{3}{c}{$c_{bm,B}$ (covariance channel)} \\
\cmidrule(l){2-4}
$c_{bm,A}$ (mean channel) & 0.00 & 0.22 & 0.45 \\
\midrule
\multicolumn{4}{l}{\textit{Crossover (CO)}} \\
0.00 & 0.046 & 0.064 & 0.206 \\
0.22 & 0.070 & 0.179 & 0.361 \\
0.45 & 0.172 & 0.354 & 0.528 \\
\addlinespace
\multicolumn{4}{l}{\textit{Hybrid}} \\
0.00 & 0.034 & 0.083 & 0.273 \\
0.22 & 0.108 & 0.299 & 0.621 \\
0.45 & 0.389 & 0.642 & 0.870 \\
\addlinespace
\multicolumn{4}{l}{\textit{OL+BDC}} \\
0.00 & 0.017 & 0.049 & 0.148 \\
0.22 & 0.086 & 0.164 & 0.362 \\
0.45 & 0.276 & 0.473 & 0.688 \\
\bottomrule
\end{tabular}
\end{table}

Table~\ref{tab:arch-c-power} reports empirical power for
Architecture~C across the full $3 \times 3$ parameter panel at
the most demanding carryover condition ($t_{1/2} = 1.0$ week).
Several features of the results merit comment.

First, the monotone structure is exact: power increases strictly
with each channel parameter, holding the other fixed, across all
three designs. This confirms that the two channels contribute
independently and additively to detectability.

Second, interior cells exceed the probability-scale additive
benchmark. At ($c_{bm,A} = 0.22$, $c_{bm,B} = 0.22$) in the
Hybrid design, power is 0.299. The additive baseline on the
probability scale -- the power expected if the two channels
contributed independently -- is $p(0.22,0) + p(0,0.22) - p(0,0)
= 0.108 + 0.083 - 0.034 = 0.157$. The observed value of 0.299
exceeds this benchmark by 0.142, indicating genuine
super-additivity: the joint signal compounds rather than
aggregates.

Third, the covariance channel ($c_{bm,B}$) and the mean-shift
channel ($c_{bm,A}$) contribute comparably to carryover
sensitivity. Comparing the no-carryover ($t_{1/2} = 0$) and
one-week ($t_{1/2} = 1.0$) panels for the CO design, increasing
$c_{bm,B}$ from 0 to 0.45 (fixing $c_{bm,A} = 0$) degrades power
from 0.368 to 0.206 -- a 44\% relative loss. The corresponding
loss along the $c_{bm,A}$ axis (fixing $c_{bm,B} = 0$) is from
0.352 to 0.172 -- a 51\% loss, in the same range. At the
equal-weight interior cell ($c_{bm,A} = c_{bm,B} = 0.45$), the
Hybrid design retains 0.870 power under one-week carryover,
substantially exceeding either pure architecture alone at
$c_{bm} = 0.45$ (Architecture~A: 0.389; Architecture~B: 0.273).

Fourth, design choice remains consequential under Architecture~C.
At the full-signal, one-week-carryover cell, Hybrid (0.870)
substantially outperforms OL+BDC (0.688) and especially CO
(0.528). The design ranking established in Section~3.2 is
preserved throughout the Architecture~C grid.

\bigskip\noindent\rule{\textwidth}{0.4pt}\bigskip

\section{4. Which recipe matches which biology?}

The choice of architecture is really a choice about *why* the
biomarker predicts response, so picking one is a biological
statement, not a coding preference.

::: {.bullets data-latex=""}
\textbf{Use Architecture A when the biomarker mechanically sets the
size of the drug effect.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item The biomarker fixes effective dose or receptor occupancy, so
it scales the effect proportionally.
\item Doubling the biomarker roughly doubles the effect at every
timepoint.
\item Carryover keeps the same proportion in the off-drug period, so power is
barely touched.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
\textbf{Use Architecture A when the biomarker sets the size of the
drug effect.} If a patient's biomarker mechanically determines how
much drug effect they get -- for example, kidney function (eGFR)
controlling how fast a drug is cleared, or receptor density
controlling how strongly a drug acts -- then doubling the biomarker
roughly doubles the effect at every timepoint. Carryover does not
break this: the leftover effect during the off-drug period keeps the same
proportion, so power is barely touched.
:::

Typical examples:

\begin{itemize}
\item \textbf{Drug clearance.} Kidney function (eGFR) sets how fast a
  renally-cleared drug leaves the body, so it directly scales the
  steady-state drug level and hence the effect.
\item \textbf{Receptor density.} If a biomarker measures how many
  target receptors a patient has, more receptors means
  proportionally more drug effect.
\item \textbf{Metabolizer genotype.} A "poor metabolizer" of a
  prodrug activates less of it, getting a proportionally smaller
  effect.
\end{itemize}

::: {.bullets data-latex=""}
\textbf{Use Architecture B when the biomarker is only a proxy for a
hidden responder state.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item The biomarker imperfectly indexes the real cause, so identical
biomarkers can respond differently.
\item It predicts response only while the drug is acting.
\item In the off-drug period the correlation fades and power drops sharply, as
in Section 3.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
\textbf{Use Architecture B when the biomarker only flags who is
likely to respond.} Sometimes the biomarker is just a *proxy* for
some hidden "responder" state. Two patients with the same biomarker
can respond very differently, because the biomarker imperfectly
indexes the real cause (a disease subtype, an inflammatory state, a
genetic background). Here the biomarker predicts response *only
while the drug is acting*; once the drug washes out, the
correlation that carried the signal fades, and power drops sharply
-- exactly the Architecture B behavior we saw in Section 3.
:::

Typical examples:

\begin{itemize}
\item \textbf{Blood pressure in PTSD.} High resting blood pressure may
  flag a noradrenergic PTSD subtype likely to respond to prazosin,
  but it does not \emph{cause} the response; it is a noisy marker of
  the real underlying state.
\item \textbf{Inflammatory markers.} High C-reactive protein (CRP)
  or interleukin-6 (IL-6) may predict
  response to anti-inflammatory augmentation of an antidepressant,
  but only on average -- many high-CRP patients still do not respond.
\item \textbf{Polygenic risk scores.} These predict treatment
  response while explaining only a fraction of the variance; the
  rest is unmeasured biology.
\end{itemize}

(A
fully faithful model of this "hidden responder class" idea would be
a finite mixture; Architecture B is a convenient correlation-level
approximation, adequate for power calculations but not a claim
about the true mechanism. A companion paper [@pmsimstats-paper03]
develops the mixture and latent-class formulations, deriving when a
single multivariate normal reproduces a two-component mixture's
first two moments and cataloguing where it fails. Senn
[@senn2016mastering] also cautions
that apparent biomarker-treatment effects are sometimes just
variance artifacts, so Architecture B's signal may be smaller than
assumed.)

::: {.bullets data-latex=""}
\textbf{The motivating prazosin/PTSD case is an Architecture B
story, so its carryover power loss is a clinically relevant
warning.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item Biomarker is resting blood pressure, a proxy for noradrenergic
tone, not a driver of drug levels.
\item Designs with long off-drug periods may lose more power than an
Architecture A simulation predicts.
\item Architecture C is the right model if blood pressure also acts
through the mean channel; it needs evidence for both parameters.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
\textbf{The motivating example.} The pmsimstats software was built
for a prazosin/PTSD trial [@raskind2013; @hendrickson2020] where
the biomarker is resting blood pressure and the outcome is a PTSD
symptom score. Blood pressure is thought to be a *proxy* for
noradrenergic tone, not a direct driver of drug levels -- an
Architecture B story. So the large carryover-driven power loss we
find under Architecture B is the clinically relevant warning:
designs with long off-drug periods may lose more power to detect
this biomarker than a simpler Architecture A simulation would
suggest. Architecture C is the appropriate model if blood pressure
turns out to act through the mean channel as well as the
correlation channel; because it estimates two independent
parameters, it calls for independent evidence on each (see Section
2.3).
:::

\bigskip\noindent\rule{\textwidth}{0.4pt}\bigskip

\section{5. Implications for Simulation Study Design}

\subsection{5.1 Choosing the appropriate architecture}

The choice of DGP architecture should be guided by the
biological mechanism hypothesized for the
biomarker-treatment interaction:

\begin{table}[htbp]
\centering
\begin{tabular}{lccc}
\toprule
Criterion & Architecture A & Architecture B & Architecture C \\
\midrule
Biomarker role & Causal mediator & Statistical predictor &
  Both \\
Mechanism & Deterministic scaling & Probabilistic association &
  Both channels \\
Signal location & Mean structure & Covariance structure &
  Mean + covariance \\
Free parameters & 1 ($c_{bm}$) & 1 ($c_{bm}$) &
  2 ($c_{bm,A}$, $c_{bm,B}$) \\
Carryover impact & Modest (1--3\%) &
  Design-dependent, up to 31\% &
  Intermediate; scales with $c_{bm,B}$ \\
Appropriate when &
  \parbox[t]{2.5cm}{Biomarker\\determines dose\\or PK} &
  \parbox[t]{2.5cm}{Biomarker\\indexes a latent\\subtype} &
  \parbox[t]{2.5cm}{Channel\\uncertain;\\sensitivity use} \\
\bottomrule
\end{tabular}
\end{table}

::: {.bullets data-latex=""}
\textbf{When the mechanism is ambiguous, sweep Architecture C; use
Architecture B as the conservative default.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item A mixing weight $\alpha$ spans the two pure-architecture
extremes via $c_{bm,A} = (1-\alpha)c$, $c_{bm,B} = \alpha c$.
\item Architecture B is the safer default for design decisions.
\item Its larger carryover sensitivity keeps power estimates from
being optimistic.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
If the biological mechanism is ambiguous, Architecture~C with a
range of mixing weights $\alpha \in [0, 1]$ (where
$c_{bm,A} = (1-\alpha) c$ and $c_{bm,B} = \alpha c$ for fixed
total weight $c$) provides a natural sensitivity framework
spanning the two pure-architecture extremes. Architecture~B
should remain the default conservative choice for trial design
when the channel structure is unknown: its larger carryover
sensitivity ensures that power estimates will not be
systematically optimistic relative to either pure-architecture
alternative.
:::

\subsection{5.2 Reporting requirements}

Simulation studies evaluating biomarker-moderated treatment
effects should explicitly state:

\begin{enumerate}
\item Whether the biomarker-treatment interaction is generated through
  mean moderation, differential correlation, or both channels
  simultaneously.
\item The biological rationale for the chosen mechanism and,
  for Architecture C, independent evidence supporting each
  channel.
\item Whether the carryover model (if present) operates on the
  interaction signal consistently with the chosen architecture.
\item Sensitivity analyses under alternative architectures, if the
  biological mechanism is uncertain.
\end{enumerate}

::: {.bullets data-latex=""}
\textbf{These requirements specialize the ADEMP template; the
architecture choice is its data-generating (D) element.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item ADEMP fixes Aims, Data-generating mechanisms, Estimands,
Methods, and Performance measures.
\item The DGP-architecture choice is the D element.
\item Our reporting rules are ADEMP specialized to
architecture-sensitive biomarker simulations.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
The reporting framework above is consistent with the broader
ADEMP template of Morris, White, and Crowther [@morris2019simulation], which
prescribes that simulation studies fix five elements before
production runs: the \textbf{Aims} of the study, the
\textbf{Data-generating mechanisms}, the \textbf{Estimands}
(the inferential targets), the \textbf{Methods} (analysis
specifications compared), and the \textbf{Performance measures}
(power, bias, coverage, Type I error, and MCSEs of each). The DGP-architecture choice this paper
concerns is the \textbf{D} element of an ADEMP specification;
the reporting requirements above are the specialization of the
broader ADEMP framework to the architecture-sensitive class of
biomarker-validation simulations.
:::

The \textbf{primary estimand} for this paper's simulations is
the biomarker-treatment interaction coefficient
$\beta_{bm:D}$ in the linear-mixed analysis model
$\mathrm{Sx} \sim bm + t + Dbc + bm\!:\!Dbc$ with random
intercept and AR(1) residual correlation. The reported
\textbf{performance measures} are: (i) power for $\beta_{bm:D}$
at $\alpha = 0.05$, with MCSE
$\sqrt{\hat{\pi}(1-\hat{\pi})/n_{\text{reps}}}$; (ii) Type I
error under $c_{bm} = 0$, with the same MCSE; (iii) bias of
$\hat{\beta}_{bm:D}$ relative to the calibrated true value at
each parameter cell; (iv) the empirical MCSE
of $\hat{\beta}_{bm:D}$ across replicates; and (v) the
convergence rate of the linear-mixed fit. Sample size,
biomarker effect size, and carryover half-life vary across the
parameter grid; the number of replicates per cell is calibrated
to achieve MCSE below 0.02 for power at $\hat{\pi} = 0.5$,
which corresponds to $n_{\text{reps}} \geq 625$. The
simulations reported here use $n_{\text{reps}} = 1000$
throughout.

\subsection{5.3 When the choice is uncertain}

If the biological mechanism is ambiguous, as it often is in
early biomarker development, we suggest investigators take the
following steps:

\begin{enumerate}
\item Run the simulation under both architectures and report the
  range of power estimates.
\item Architecture B produces more conservative (lower)
  power estimates under carryover, making it the safer default
  for trial design decisions.
\item Design the trial to minimize carryover (adequate washout
  periods) to reduce sensitivity to the DGP architecture choice.
\end{enumerate}

\bigskip\noindent\rule{\textwidth}{0.4pt}\bigskip

\section{6. Can a smarter analysis recover the lost power?}

\emph{None of the strategies below was tested by simulation here;
the companion paper on carryover-mitigation strategies does that.
This section just lists the options and their intuition.}

::: {.bullets data-latex=""}
\textbf{Architecture B's loss has two sources; only the
variance-compression part is recoverable, by down-weighting
low-signal off-drug points.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item Source 1: the drug exposure variable's variance shrinks
(recoverable).
\item Source 2: the biomarker-response correlation genuinely fades
off-drug (lost information).
\item Methods that down-weight or drop off-drug points can help with
the first.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
Under Architecture B the power loss has two sources: the
drug exposure variable's variance shrinks, and the biomarker-response
correlation genuinely fades off-drug. The second is lost
information that no model can recover. But because off-drug points
carry less signal, methods that *down-weight or drop* them can
help.
:::

The natural candidates are:

\begin{itemize}
\item \textbf{Use on-drug points only.} Every kept point has the full
  correlation, but you lose the off-drug reference and sample size;
  best for designs with a long on-drug phase, worst for crossover.
\item \textbf{Weight by drug exposure.} Keep all points but weight each
  by its expected residual drug effect, $w_t = e^{-\lambda\,
  t_{sd}}$, so high-signal points count more. Easy to do with the
  \texttt{weights} argument in \texttt{nlme::lme}.
\item \textbf{Within-subject contrast.} Reduce each patient to one
  number, (mean on-drug) minus (mean off-drug), then regress that
  on the biomarker: $\bar Y_{i,\text{on}} - \bar Y_{i,\text{off}} =
  \alpha + \gamma B_i + \epsilon_i$. Simple and robust, at some
  cost in efficiency.
\item \textbf{Two-stage random slopes.} First estimate each patient's
  drug effect $u_{1i}$ from a random-slopes model, then regress
  those on the biomarker, $\hat u_{1i} = \delta_0 + \delta_1 B_i +
  \eta_i$.
\item \textbf{Drop the most contaminated early off-drug points}, or
  \textbf{fix it in the design} with longer washouts or more on-drug
  time.
\end{itemize}

\subsection{6.1 Qualitative comparison}

Table~\ref{tab:sec6-strategies} summarizes the strategies along
three dimensions: whether they discard observations,
implementation complexity, and the expected direction of benefit
based on the mechanistic account in Section 3.2. All entries in
the `Hypothesized benefit' column are qualitative; no simulation
evidence is presented here, and each strategy should be
evaluated empirically before use in trial planning.

\begin{table}[htbp]
\centering
\caption{Qualitative summary of analysis strategies for
recovering power under Architecture B carryover. All benefit
assessments are mechanistic hypotheses, not simulation-supported
estimates.}
\label{tab:sec6-strategies}
\begin{tabular}{lccp{5cm}}
\toprule
Strategy & Discards data & Complexity & Hypothesized benefit \\
\midrule
On-drug only & Yes & Low & Eliminates correlation erosion;
  forfeits off-drug reference (best for OL/Hybrid) \\
Weighted & No & Low & Reduces erosion damage proportionally;
  depends on weight calibration \\
Within-subject contrast & Aggregates & Low & Robust to
  carryover misspecification; less efficient than the linear
  mixed-effects (LME) model \\
Two-stage random slopes & No & Moderate & Unknown; separates
  estimation stages; robustness unverified \\
Exclude early off-drug & Some & Low & Removes most contaminated
  observations; improves contrast clarity \\
Design modification & N/A & N/A & Addresses root cause;
  cost is longer trial or reduced carryover estimation \\
\bottomrule
\end{tabular}
\end{table}

A systematic simulation comparison of these strategies under
Architecture B, analogous to the factorial comparison of
analysis models in Section 3, would identify which approach
best recovers the power lost to carryover for each trial
design. We regard this comparison as important future work
and note that the on-drug-only restriction and the weighted
analysis are the most tractable candidates for a first
empirical evaluation.

\bigskip\noindent\rule{\textwidth}{0.4pt}\bigskip

\section{7. How this fits the wider literature}

::: {.bullets data-latex=""}
\textbf{Architecture A is near-universal in the literature, so
published power estimates may be optimistic for Architecture B
biomarkers.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item Simon, Freidlin-Korn, Renfro-Sargent, and standard N-of-1
studies all put the effect in the mean.
\item Even N-of-1 carryover work sharing Hendrickson's authors uses
Architecture A, so B looks like a one-off.
\item Most published power figures may be optimistic for genuinely
Architecture B biomarkers.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
Almost every trial-simulation framework uses Architecture A:
Simon [@simon2010], Freidlin and Korn [@freidlin2014], Renfro and
Sargent [@renfro2017], and the standard N-of-1 simulation studies
all put the biomarker effect in the mean. Even methodological work
on N-of-1 carryover that shares authors with Hendrickson et al.
[@hendrickson2020] uses Architecture A, which suggests Hendrickson's
differential-correlation choice was a one-off rather than a
convention. The practical worry is that most published power
estimates may be \emph{optimistic} for biomarkers that really
behave like Architecture B, because they ignore the extra
carryover-driven loss we have demonstrated.
:::

::: {.bullets data-latex=""}
\textbf{The distinction matters most in treatment-switching designs;
when the mechanism is unknown, compute power both ways and use the
lower number.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item Treatment-switching designs (crossover, N-of-1, basket) put
patients through off-drug phases where B's correlation decays.
\item Parallel-group designs have no off-drug phase, so the
architecture choice does not matter.
\item This is a DGP question, separate from the analysis-model
guidance of the PATH statement.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
The distinction matters most exactly in the designs precision
medicine cares about: any \emph{treatment-switching} design (crossover,
N-of-1 hybrid, basket/platform) puts patients through off-drug
phases where the Architecture B correlation decays. Parallel-group
designs, with no off-drug phase, are immune and give similar
numbers under either architecture. The simple advice when the
mechanism is unknown: run the power calculation \emph{both ways} and
use the lower (Architecture B) number for planning. This is a
data-generating-process question, separate from -- and not answered
by -- the analysis-model guidance in the PATH statement
[@kent2020path].
:::

To see where this bites in practice, it helps to sort designs by
how much off-drug time each patient experiences:

\begin{itemize}
\item \textbf{Treatment-switching designs} (crossover, N-of-1 hybrid,
  basket and platform trials) deliberately cycle each patient on and
  off drug. Under Architecture B every off-drug period erodes the
  correlation, so the interaction signal in later periods is weaker
  than in the first -- the power loss of Section 3 applies directly.
\item \textbf{Parallel-group enrichment designs} keep each patient on
  one treatment throughout. There is no off-drug phase for the
  correlation to decay in, so Architectures A and B give similar
  power. The architecture choice simply does not matter here.
\item \textbf{Adaptive enrichment trials} decide mid-trial whom to keep
  enrolling, using interim power calculations written in the
  Architecture A idiom. If the truth is closer to Architecture B and
  patients pass through washouts, the real power is lower than the
  rule assumes, which can trigger premature stopping or an
  over-narrow eligible population.
\end{itemize}

The pattern is worth stating plainly: the architecture choice
matters \emph{most} in exactly the within-subject designs that
precision medicine finds most attractive, and \emph{least} in the
parallel-group designs where it is often (wrongly) assumed to be
harmless.

\bigskip\noindent\rule{\textwidth}{0.4pt}\bigskip

\section{8. Conclusions}

In conclusion, six points are to be emphasized.

\begin{enumerate}
\item The choice between direct mean moderation (Architecture A) and
  differential correlation (Architecture B) for generating
  biomarker-treatment interactions in clinical trial simulations
  has substantive consequences for power estimates under
  carryover.

\item Under Architecture A, carryover reduces power only
  modestly, by 1 to 3\% in relative terms across the three
  designs examined, because the proportional biomarker-drug
  relationship is preserved during the off-drug period. Under
  Architecture B the loss is strongly design-dependent: it
  remains modest in the crossover and hybrid designs (3 to
  5\%) but reaches roughly 31\% in the open-label OL+BDC
  design, because the differential-correlation signal erodes
  as the drug effect wanes.

\item The choice should be guided by the hypothesized biological
  mechanism: Architecture A for biomarkers that directly
  determine drug pharmacokinetics or pharmacodynamics;
  Architecture B for biomarkers that statistically predict
  treatment response through latent subtype membership.

\item For the prazosin-PTSD application with blood pressure as the
  predictive biomarker, Architecture B (MVN differential
  correlation) is the more appropriate model, and the resulting
  power sensitivity to carryover should inform trial design
  decisions regarding off-drug period duration.

\item The existing clinical trial simulation literature uses
  Architecture A almost exclusively. The Hendrickson et al.
  (2020) MVN approach is, to the best of our knowledge, unique
  in the N-of-1 trial context. This means that published power
  estimates for biomarker-moderated crossover and N-of-1
  designs may be systematically optimistic for biomarkers
  that operate through a correlation-based (Architecture B)
  mechanism, because they do not account for the carryover
  sensitivity documented here.

\item Simulation studies for biomarker-moderated treatment effects
  should explicitly report their DGP architecture and consider
  sensitivity analyses under the alternative when the biological
  mechanism is uncertain.
\end{enumerate}

\bigskip\noindent\rule{\textwidth}{0.4pt}\bigskip

\section{Notes and reproducibility}

The authors declare no conflicts of interest and received no
specific funding. Code and data are in the pmsimstats-ng
repository (drivers under \texttt{analysis/scripts/}, outputs under
\texttt{analysis/data/}), with package versions pinned in
\texttt{renv.lock}; the document renders under R~4.6.0.

\textbf{One convention to note.} In Sections 3.1 and 3.3, $N = 35$
is the number of subjects \emph{per randomization path}, and paths
are concatenated before analysis, so the effective analysis sample
is 70 (two-path CO and OL+BDC) or 140 (four-path Hybrid). The
Architecture~C results in Section 3.4 instead use $N = 35$ as the
\emph{total}, so their power values are lower and are not directly
comparable to the Section 3.1 tables.

# References

::: {#refs}
:::

```{r footer, echo=FALSE, results='asis'}
src <- sub("^/Users/zenn/(Library/CloudStorage/)?Dropbox/prj/", "~/prj/",
           knitr::current_input(dir = TRUE))
cat(sprintf(
  "\n\\vfill\n\n---\n\n*Rendered on %s.*  \n*Source: `%s`*\n",
  format(Sys.time(), "%Y-%m-%d at %H:%M %Z"),
  src
))
```
