---
title: "A closed-form approximation to the biomarker-treatment
  interaction estimator in aggregated N-of-1 trials, and what it
  implies for design efficiency under a mandatory open-label
  lead-in"
author: 'pmsimstats team'
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: true
    toc_depth: 2
    number_sections: true
    keep_tex: true
    highlight: tango
    includes:
      in_header: claudecode.tex
fontsize: 11pt
linestretch: 1.1
geometry: 'left=3cm,right=3cm,top=2.5cm,bottom=2.5cm'
bibliography: references.bib
csl: statistics-in-medicine.csl
link-citations: true
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{float}
  - \usepackage[mathlines]{lineno}
  - \linenumbers
knit: (function(input, ...) { d <- dirname(input); while (!file.exists(file.path(d, 'tools', 'stamp-render.R')) && d != dirname(d)) d <- dirname(d); source(file.path(d, 'tools', 'stamp-render.R'))$value(input) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

cf_rel <- file.path('analysis', 'data', 'nof1-design-efficiency',
                    'closed-form-validation.csv')
cf_path <- if (requireNamespace('here', quietly = TRUE) &&
               file.exists(here::here(cf_rel))) {
  here::here(cf_rel)
} else {
  file.path('..', '..', 'data', 'nof1-design-efficiency',
            'closed-form-validation.csv')
}

cf <- read.csv(cf_path, stringsAsFactors = FALSE)
cf$design_lab <- c(CO = 'CO', Hybrid = 'Hybrid',
                   OLBDC = 'OL+BDC')[cf$design]
cf <- cf[order(cf$t1half,
               match(cf$design, c('CO', 'Hybrid', 'OLBDC')),
               cf$N), ]

fit_var    <- lm(I(sd_beta^2 * N) ~ 0 + v_sum, data = cf)
sigma2_hat <- unname(coef(fit_var)[1])
r2_var     <- summary(fit_var)$r.squared

n_cells    <- nrow(cf)
mae_power  <- mean(abs(cf$power_err))
rmse_power <- sqrt(mean(cf$power_err^2))
cor_power  <- cor(cf$power, cf$power_pred)

cell <- function(dsg, n, t) {
  cf[cf$design == dsg & cf$N == n & cf$t1half == t, ]
}

co_sd_ratio <- cell('CO', 35, 0)$sd_beta / cell('CO', 70, 0)$sd_beta

pct_range <- function(v) sprintf('%.0f-%.0f', min(v), max(v))
co_under  <- pct_range(100 * (cf$ratio_sd[cf$design == 'CO'] - 1))
oth_over  <- pct_range(100 * (1 - cf$ratio_sd[cf$design != 'CO']))

hyb_atten_pp <- 100 * (1 - cell('Hybrid', 70, 1)$A)
hyb_b0       <- cell('Hybrid', 70, 0)$mean_beta
hyb_b1       <- cell('Hybrid', 70, 1)$mean_beta
hyb_beta_pct <- 100 * (1 - abs(hyb_b1) / abs(hyb_b0))

mean_abs_beta <- mean(abs(cf$mean_beta))

fmt <- function(x, d = 3) sprintf(paste0('%.', d, 'f'), x)

# {-} keeps the minus typeset as an ordinary symbol inside |.|.
fmt_signed <- function(x, d = 4) {
  sprintf(paste0('%s%.', d, 'f'), ifelse(x < 0, '{-}', ''), abs(x))
}

# Labels and placement are emitted verbatim so the raw \ref{} cross
# references in the prose keep resolving.
latex_table <- function(x, col_names, align, caption, label) {
  tab <- knitr::kable(x, format = 'latex', booktabs = TRUE,
                      col.names = col_names, align = align,
                      row.names = FALSE, escape = FALSE,
                      linesep = '')
  cat('\\begin{table}[H]\n\\centering\n',
      '\\caption{', caption, '}\n',
      '\\label{', label, '}\n',
      tab, '\n\\end{table}\n', sep = '')
}
```

**Short communication.**

# Introduction

Every design question this project's companion manuscripts have
addressed, that is, how many randomization paths, what sample
size, how long to wait before a discontinuation-phase assessment,
and whether a brief crossover tail is worth its extra visits, has
so far been answered by Monte Carlo simulation
[@pmsimstats-paper01; @pmsimstats-paper02; @pmsimstats-paper05;
@pmsimstats-paper09; @pmsimstats-paper11]. That is appropriate for
the compendium's headline estimand, which combines a Gompertz
response trajectory, a dual-channel biomarker-moderation
architecture, exponential or Weibull carryover, and a continuous-time
AR(1) mixed model into a system with no obvious closed form. But
it leaves a practical gap. A trialist choosing among two or three
concrete design options today cannot get even an
order-of-magnitude power estimate without running a fresh
simulation. This stands in contrast to the treatment main
effect, for which Senn's and Zucker's classical variance formulas
[@sennSampleSizeConsiderations2019; @zucker2010] have long
supplied exactly this kind of design intuition. It also stands
in contrast to the dichotomized-biomarker case, for which the
companion test-procedure manuscript derives a closed-form
repeated-measures ANOVA power formula [@pmsimstats-paper10].
Neither covers the
compendium's actual estimand: a continuous biomarker, a continuous
exposure predictor $D_{bc}$, and a mixed model fit by
`nlme::lme` with `corCAR1` residual correlation.

This short communication was motivated by a concrete instance of the
gap: a trial whose first phase must be open-label, for recruitment
reasons that are not themselves a design choice. Given that fixed
constraint, what should the analyst prioritize among the remaining
design parameters to make the biomarker-treatment interaction test as
efficient as possible? We first give a qualitative answer, drawn
directly from the compendium's existing simulation results. We then
ask whether a genuine closed-form approximation to the interaction
estimator's sampling variance and power can be derived, and, if so,
how far it can be trusted. We derive such an approximation from two
components: a within-subject two-group contrast reduction of the
analysis model, and the exact variance of a mean of AR(1)-correlated,
unequally spaced observations. We validate both components against
Monte Carlo output already on disk, produced for a companion
manuscript, rather than commissioning a new simulation study,
which is the appropriate scope for a short communication. We
report the validation honestly. The variance and power
approximation performs well. The attenuation-factor component
captures the correct qualitative ranking of designs by carryover
vulnerability but not its magnitude. We close
with quantitative, closed-form-grounded design guidance for the
open-label-first scenario.

# What the compendium already establishes

Before deriving anything new, it is worth stating precisely what the
existing simulation evidence already supports, since a closed form is
only useful to the extent that it reproduces conclusions already
established with more care. Four results from the compendium bear
directly on the open-label-first design question.

First, design topology after the open-label phase is the largest
lever available. At matched total sample size $N=70$, the Hybrid
design (open-label, then blinded discontinuation, then a brief
crossover; four randomization paths) dominates open-label-plus-
blinded-discontinuation alone (OL+BDC; two paths) in absolute power
at every carryover half-life tested, $0.842$ versus $0.751$ at
$t_{1/2}=0$ and $0.784$ versus $0.730$ at $t_{1/2}=1.0$ week
[@pmsimstats-paper01]. Adding the crossover tail is worth its extra
visits.

Second, matching total sample size $N$ across designs being compared
is not optional bookkeeping: it changed the compendium's own
conclusions twice. An unmatched comparison inflated Hybrid's apparent
power ceiling in the architecture manuscript and manufactured an
apparent dropout-robustness advantage for Hybrid in the
informative-dropout manuscript that a matched re-run showed was a
ceiling effect [@pmsimstats-paper01; @pmsimstats-paper09].

Third, the carryover decay's functional *shape* (exponential
versus Weibull, across shape parameters $k \in \{0.25, 0.5, 2,
4\}$) barely moves the compendium's substantive conclusions. The
*analysis specification*, that is, whether the exposure predictor
is a binary indicator, an exposure-weighted continuous decay, or
an AIC-selected-half-life variant, matters considerably more
[@pmsimstats-paper02].

Fourth, misspecifying the residual correlation structure (fitting an
exchangeable or naive model when the truth is AR(1)) is a larger
anti-conservatism risk than misspecifying carryover decay shape
[@pmsimstats-paper10].

These four results already answer "what to prioritize" at a
qualitative level. What they do not supply is a way to move from a
verbal ranking to a number: how much does adding the crossover tail
buy, quantitatively, as a function of the design's own visit
schedule, without running a new simulation for every candidate
design? That is the question the rest of this communication addresses.

# A closed-form approximation

## Setup

We work under the mean-moderation architecture
[@pmsimstats-paper01], in which the biomarker $B_i$ scales the
treatment effect directly in the outcome's conditional mean rather
than through a covariance channel, and we approximate the analysis
specification E1 (binary on-drug indicator $D_{it}$, carryover
unmodeled at the analysis stage; @pmsimstats-paper02) by a simple
two-group within-subject contrast. For participant $i$ with $k_{on}$
on-drug and $k_{off}$ off-drug assessment timepoints at fixed
calendar times, define
\begin{equation}
\Delta_i = \overline{Y}_{i}^{\,on} - \overline{Y}_{i}^{\,off},
(\#eq:contrast)
\end{equation}
the mean on-drug outcome minus the mean off-drug outcome. Because the
participant random intercept $u_i$ is common to both means, it
cancels in $\Delta_i$ exactly (no random slope is present in the
compendium's baseline model), leaving
\begin{equation}
\Delta_i = \beta_1\big(1 + \beta_{bm} B_i\big)\cdot\bar D_{bc,i}^{\,on}
  - \beta_1\big(1+\beta_{bm}B_i\big)\cdot\bar D_{bc,i}^{\,off} + \eta_i,
(\#eq:delta-decomp)
\end{equation}
where $\bar D_{bc,i}^{\,on}=1$ by construction and $\bar
D_{bc,i}^{\,off} = \overline{D_{bc}}_i^{\,off}$ is the mean of the
continuous exposure-decayed indicator across the off-drug assessment
times, which is less than one whenever residual carryover has not
fully decayed by the time of assessment, and $\eta_i$ is the
difference of the two residual means. A second-stage ordinary least
squares regression of $\Delta_i$ on the standardized biomarker $b_i$
then recovers the interaction coefficient as its slope,
\begin{equation}
\Delta_i = \gamma_0 + \beta_{bm:D}\, b_i + \eta_i, \qquad
  \beta_{bm:D} = \beta_1\,\beta_{bm}\cdot A_i,
(\#eq:e9-like)
\end{equation}
where $A_i = 1 - \overline{D_{bc}}_i^{\,off}$ is a per-participant
**carryover attenuation factor**. This construction is exactly the
two-stage paired-difference specification E9 of the companion
carryover manuscript [@pmsimstats-paper02], specialized to a strict
on/off split rather than E9's general on/off means, and is offered
here as an approximation to E1 (which is fit by the full mixed model
on all timepoints jointly, not by this two-group reduction) rather
than as an exact restatement of it.

## Sampling variance under AR(1) correlation

The variance of the mean of $k$ observations drawn from a stationary
process with autocovariance $\gamma(h) = \sigma^2\rho^{|h|}$ at
arbitrary, possibly unequally spaced times $t_1,\dots,t_k$ has the
exact closed form
\begin{equation}
V(\mathbf{t}, \sigma^2, \rho) = \frac{\sigma^2}{k^2}
  \sum_{j=1}^{k}\sum_{l=1}^{k} \rho^{|t_j - t_l|}.
(\#eq:var-mean-ar1)
\end{equation}
Applying Equation \@ref(eq:var-mean-ar1) separately to a
participant's on-drug and off-drug visit times and summing (treating
the two period means as approximately independent, an approximation
we return to in the Discussion) gives the variance of the
within-subject contrast,
\begin{equation}
\mathrm{Var}(\eta_i) \approx V(\mathbf{t}^{on}, \sigma^2, \rho) +
  V(\mathbf{t}^{off}, \sigma^2, \rho) \;\equiv\; v(\mathrm{design},
  \rho),
(\#eq:v-sum)
\end{equation}
a quantity computable directly from a design's visit schedule and an
assumed AR(1) parameter $\rho$, with no simulation required. Pooling
across $N$ participants (assuming a standardized biomarker, so
$\mathrm{Var}(b_i)=1$) and, for a multi-path design, averaging
$v(\mathrm{design},\rho)$ across the design's randomization paths,
the closed-form approximation to the interaction estimator's sampling
variance is
\begin{equation}
\widehat{\mathrm{Var}}\big(\hat\beta_{bm:D}\big) \approx
  \frac{\sigma^2\, v_1(\mathrm{design},\rho)}{N},
(\#eq:var-closed-form)
\end{equation}
where $v_1$ is $v(\mathrm{design},\rho)$ evaluated at $\sigma^2=1$, so
that $\sigma^2$ enters as a single free scale parameter separating
design-and-correlation structure ($v_1$, closed-form) from residual
noise magnitude ($\sigma^2$, an empirical quantity like any other
nuisance variance component).

## Predicted power

Given Equations \@ref(eq:e9-like) and \@ref(eq:var-closed-form), a
Wald-normal approximation to power at level $\alpha$ follows
immediately, in the same spirit as the general four-channel
decomposition of the companion calibration manuscript
[@pmsimstats-paper10, its $\kappa=1,\nu=\infty$ special case]:
\begin{equation}
z = \frac{|\beta_{bm:D}\cdot A(t_{1/2})|}
  {\sqrt{\sigma^2\, v_1(\mathrm{design},\rho)/N}}, \qquad
  \hat\pi \approx \Phi(z - z_{1-\alpha/2}) + \Phi(-z - z_{1-\alpha/2}),
(\#eq:power-closed-form)
\end{equation}
with $A(t_{1/2})$ the design-averaged attenuation factor of
Equation \@ref(eq:e9-like) and $\beta_{bm:D}$ the interaction
coefficient's magnitude in the absence of carryover. Every quantity
on the right-hand side except $\sigma^2$, $\rho$, and $\beta_{bm:D}$
itself is computable directly from the candidate design's visit
schedule and a candidate carryover half-life, with no simulation.

# Validation against existing Monte Carlo output

## Data and procedure

We validate Equations \@ref(eq:var-closed-form) and
\@ref(eq:power-closed-form) against Monte Carlo output already
produced for the dual-channel-architecture companion manuscript
[@pmsimstats-paper11], rather than running a new simulation. That
manuscript's driver
(`analysis/scripts/dgp-combined/01-run-combined-factorial.R`)
generates data under a combined architecture with independent
mean-channel and covariance-channel weights $c_{bm,A}$ and
$c_{bm,B}$; its $(c_{bm,A}=0.45,\, c_{bm,B}=0)$ boundary is, by
construction, exactly the pure mean-moderation architecture this
communication's derivation assumes. Its stored summary
(`analysis/data/dgp-combined/01-combined-summary.rds`) already
includes matched total sample sizes $N \in \{35, 70\}$, three designs
(crossover, CO; open-label-plus-blinded-discontinuation, OL+BDC;
Hybrid), three carryover half-lives $t_{1/2}\in\{0, 0.5, 1.0\}$
weeks, and analysis specification E1, at $1{,}000$ replicates per
cell. We compute $v_1(\mathrm{design}, \rho{=}0.7)$ and
$A(\mathrm{design}, t_{1/2})$ directly from the same visit schedules
the driver itself uses [@pmsimstats-paper02, §2 for the design
definitions, reproduced with driver-matching timepoints in
`analysis/scripts/nof1-design-efficiency/01-closed-form-validation.R`],
fit the single free scale parameter $\sigma^2$ by ordinary least
squares of $\widehat{\mathrm{sd}}(\hat\beta)^2 \cdot N$ on $v_1$
across all 18 design-by-half-life-by-$N$ cells, and then compare the
resulting closed-form power predictions to the cells' simulated
power. $\rho=0.7$ is the AR(1) parameter the driver itself uses by
default and is not separately estimated.

## The variance-scaling component performs well

With a single fitted scale parameter ($\hat\sigma^2 =
`r fmt(sigma2_hat)`$), the
closed form explains $R^2 = `r fmt(r2_var)`$ of the variation in
$\widehat{\mathrm{sd}}(\hat\beta)^2 \cdot N$ across the
`r n_cells` cells
(Table \ref{tab:variance-fit}). Two structural predictions of
Equation \@ref(eq:var-closed-form) are confirmed directly. First,
the sampling variance scales as $1/N$ almost exactly: the CO
design's empirical standard deviation ratio between $N=35$ and
$N=70$ is $`r fmt(co_sd_ratio)`$, against a predicted
$\sqrt{70/35}=`r fmt(sqrt(2))`$. Second, the design ranking by
$v_1$ (OL+BDC highest, Hybrid intermediate, CO lowest, at fixed
$\rho$) is reproduced in direction, though not exactly in
magnitude, by the empirical standard deviations. A systematic
residual pattern is visible and is reported rather than concealed:
the closed form under-predicts CO's empirical standard deviation by
roughly `r co_under`% and over-predicts Hybrid's and OL+BDC's by
roughly `r oth_over`%. The most likely source is Equation
\@ref(eq:v-sum)'s
independence approximation between a participant's on-drug and
off-drug period means, which is exact only when no AR(1) correlation
crosses the on/off boundary; CO's two long, contiguous four-visit
blocks per path plausibly violate that approximation more than the
other designs' more fragmented schedules.

```{r variance-fit, results='asis'}
d_var <- cf[cf$t1half %in% c(0, 1), ]
latex_table(
  data.frame(
    design = d_var$design_lab,
    N      = d_var$N,
    t1half = sprintf('%.1f', d_var$t1half),
    sd     = sprintf('%.3f', d_var$sd_beta),
    pred   = sprintf('%.3f', d_var$pred_sd_beta),
    ratio  = sprintf('%.2f', d_var$ratio_sd)
  ),
  col_names = c('Design', '$N$', '$t_{1/2}$ (wk)',
                '$\\mathrm{sd}(\\hat\\beta)$',
                '$\\widehat{\\mathrm{sd}}(\\hat\\beta)$', 'Ratio'),
  align = 'lrrrrr',
  caption = sprintf(paste0(
    'Closed-form sampling-standard-deviation predictions against ',
    'Monte Carlo output, at $\\hat\\sigma^2 = %.3f$ (fitted once, ',
    'across all %d cells). $\\mathrm{sd}(\\hat\\beta)$ is the ',
    'empirical standard deviation of the fitted \\texttt{bm:Dbc} ',
    'coefficient across 1{,}000 replicates; ',
    '$\\widehat{\\mathrm{sd}}(\\hat\\beta)$ is the closed-form ',
    'prediction of Equation \\ref{eq:var-closed-form}.'),
    sigma2_hat, nrow(cf)),
  label = 'tab:variance-fit'
)
```

The $t_{1/2}=0.5$ rows, omitted from Table \ref{tab:variance-fit}
for space, follow the same pattern and are included in the full
18-row output of
`analysis/data/nof1-design-efficiency/closed-form-validation.csv`.

## Power prediction

Using the fitted $\hat\sigma^2$ and, for each cell, that cell's own
empirical $\overline{\hat\beta}$ as the effective true effect in
Equation \@ref(eq:power-closed-form), a choice discussed in
\S\ref{limitations}, the closed form predicts simulated power with
a mean absolute error of $`r fmt(mae_power)`$, a root-mean-square
error of $`r fmt(rmse_power)`$, and a correlation of
$`r fmt(cor_power)`$ across the `r n_cells` cells (Table
\ref{tab:power-fit} gives the $t_{1/2}=0$ and $t_{1/2}=1.0$ rows in
full). For a first-pass closed form validated against a single
architecture at one $\rho$ value, this is a substantively useful
level of agreement: a trialist screening candidate designs would
reach the same qualitative decision from the closed form as from the
full Monte Carlo simulation in the great majority of cells inspected
here.

```{r power-fit, results='asis'}
d_pow <- cf[cf$t1half %in% c(0, 1), ]
latex_table(
  data.frame(
    design = d_pow$design_lab,
    N      = d_pow$N,
    t1half = sprintf('%.1f', d_pow$t1half),
    sim    = sprintf('%.3f', d_pow$power),
    pred   = sprintf('%.3f', d_pow$power_pred),
    err    = sprintf('$%+.3f$', d_pow$power_err)
  ),
  col_names = c('Design', '$N$', '$t_{1/2}$ (wk)',
                'Power (simulated)', 'Power (closed form)', 'Error'),
  align = 'lrrrrr',
  caption = paste0(
    'Closed-form power predictions against simulated power ',
    '(1{,}000 replicates per cell, $\\alpha=0.05$ two-sided).'),
  label = 'tab:power-fit'
)
```

The closed form correctly identifies Hybrid as the best-powered
design at both sample sizes and both half-lives shown, matching the
compendium's own headline finding [@pmsimstats-paper01]. It does not
correctly order CO against OL+BDC: the closed form ranks CO above
OL+BDC at $N=70$ (predicted power
$`r fmt(cell('CO', 70, 0)$power_pred)`$ versus
$`r fmt(cell('OLBDC', 70, 0)$power_pred)`$ at
$t_{1/2}=0$), while the simulation ranks them the other way
($`r fmt(cell('CO', 70, 0)$power)`$
versus $`r fmt(cell('OLBDC', 70, 0)$power)`$). This is a genuine
limitation, not a rounding
artefact, and is reported as such rather than smoothed over: the
closed form is reliable for identifying the best design among
candidates that differ substantially in $v_1$, and unreliable for
close calls between designs whose $v_1$ values are similar.

## The attenuation factor: correct ranking, overstated magnitude

Table \ref{tab:attenuation} compares the closed-form attenuation
factor $A(t_{1/2})$, computed purely from each design's visit
schedule, to the actual relative change in the fitted mixed-model
coefficient's magnitude, $|\overline{\hat\beta}(t_{1/2}{=}1.0)| \,/\,
|\overline{\hat\beta}(t_{1/2}{=}0)|$, from the same Monte Carlo cells.
The direction is right in every case, and the ranking across designs
is right: CO's off-drug visits are, on one randomization path, a
pre-drug baseline period that carryover cannot contaminate at all, so
CO shows the least closed-form attenuation of the three designs,
exactly matching the compendium's own qualitative finding that CO's
carryover-induced power loss is the smallest of the designs compared
[@pmsimstats-paper01]. But the *magnitude* is badly overstated
throughout: the closed form predicts a
$`r fmt(hyb_atten_pp, 1)`$-percentage-point
attenuation for Hybrid between $t_{1/2}=0$ and $t_{1/2}=1.0$ weeks,
while the fitted coefficient's magnitude falls by only about
$`r fmt(hyb_beta_pct, 1)`\%$
over the same range ($N=70$: $|`r fmt_signed(hyb_b0)`|$ to
$|`r fmt_signed(hyb_b1)`|$).

```{r attenuation, results='asis'}
d_att <- do.call(rbind, lapply(c('CO', 'Hybrid', 'OLBDC'),
  function(dsg) {
    r0 <- cf[cf$design == dsg & cf$N == 70 & cf$t1half == 0, ]
    r1 <- cf[cf$design == dsg & cf$N == 70 & cf$t1half == 1, ]
    data.frame(
      design = r1$design_lab,
      A      = sprintf('%.3f', r1$A),
      ratio  = sprintf('%.3f', abs(r1$mean_beta / r0$mean_beta))
    )
  }))
latex_table(
  d_att,
  col_names = c('Design', '$A(t_{1/2}{=}1.0)$ (closed form)',
                paste0('$|\\bar{\\hat\\beta}(1.0)/',
                       '\\bar{\\hat\\beta}(0)|$ (simulated)')),
  align = 'lrr',
  caption = paste0(
    'Closed-form attenuation factor versus the actual relative ',
    "change in the fitted interaction coefficient's magnitude, ",
    '$N=70$.'),
  label = 'tab:attenuation'
)
```

The explanation is structural, not a fitting failure. The closed
form's attenuation factor comes from a crude two-group mean split
that discards each observation's individual timing; the fitted mixed
model instead includes an explicit time covariate $t$ alongside
`bm:Dbc`, and recovers gradient information within the off-drug
period (how the outcome trends as $D_{bc}$ decays across successive
off-drug visits) that the two-group contrast simply averages away.
This is a genuine, useful negative result. It means the
attenuation factor of Equation \@ref(eq:e9-like) should be read
as a *design screening heuristic*, in that it will correctly tell
an analyst which of two designs loses more information to a
given carryover half-life, and not as a *quantitative bias
estimate* for the coefficient a properly specified mixed-model
analysis will actually report.

# Design guidance for a mandatory open-label lead-in

Given a fixed open-label first phase, the validated components above
translate the compendium's qualitative guidance into a design
procedure. First, compute $v_1(\mathrm{design}, \rho)$ from
Equation \@ref(eq:v-sum) for every design under consideration, using
each candidate's actual visit schedule; this can be done for a
design that has never been simulated, in seconds, without touching
the simulation pipeline. Second, use $v_1$'s ranking to eliminate
designs that are clearly dominated (a large $v_1$ relative to the
alternatives) and to identify a small shortlist worth a full Monte
Carlo comparison; Table \ref{tab:variance-fit}'s validation shows
this ranking is trustworthy for well-separated candidates and should
not be trusted to resolve close calls, which still warrant
simulation. Third, use $A(t_{1/2})$ from Equation \@ref(eq:e9-like)
only as a relative screen across designs at a shared assumed
half-life, not as a quantitative power adjustment; §4.4 shows it
substantially overstates the actual attenuation a `corCAR1` mixed
model with a time covariate will recover from. Fourth, since adding
visits after the mandatory open-label phase is the largest lever in
$v_1$'s design-dependence (Table \ref{tab:variance-fit}), and since
the compendium's own matched-$N$ comparison shows a brief crossover
tail after blinded discontinuation is worth its cost in the one
design pairing checked here [@pmsimstats-paper01], a Hybrid-style
continuation of the mandatory open-label phase should be the default
starting point for the design shortlist, to be displaced only by a
full simulation showing a specific alternative outperforms it for
the trial's actual parameters.

# Limitations {#limitations}

Several scope limitations bound how far this closed form should be
trusted. It is derived and validated under the mean-moderation
architecture only; the covariance-moderation architecture's
interaction lives in $\partial\,\mathrm{Cor}(BM,BR)/\partial D_{bc}$
rather than in the conditional mean [@pmsimstats-paper01], and the
two-group contrast reduction of Equation \@ref(eq:contrast) does not
apply to it without a separate derivation. The validation uses the
Monte Carlo cells' own empirical $\overline{\hat\beta}$ as the
effective true effect in Equation \@ref(eq:power-closed-form) rather
than the architecture's nominal calibration target
$c_{bm,A}\cdot\sigma_{BR}$: the two differ by roughly an order of
magnitude in the data inspected here ($3.6$ nominal against
approximately $`r fmt(mean_abs_beta, 2)`$ observed), a discrepancy
this communication
does not resolve and flags as an open calibration question for the
companion architecture manuscripts rather than papering over with an
ad hoc rescaling. The variance formula's on/off independence
approximation (Equation \@ref(eq:v-sum)) is shown, by its residual
pattern in Table \ref{tab:variance-fit}, to be imperfect for designs
with long contiguous within-phase blocks. Validation is against a
single $\rho$ value ($0.7$, the driver default) and a single pair of
sample sizes; whether the closed form's accuracy holds at other
correlation strengths or substantially larger or smaller $N$ is not
established here. Finally, the closed form omits the placebo-belief
and natural-history components of the outcome decomposition
entirely [@pmsimstats-paper06]; a biomarker correlated with placebo
responsiveness will bias the interaction estimate in a way no version
of this closed form, restricted to the pharmacological channel alone,
can detect.

# Conclusion

We have presented a closed-form approximation to the
biomarker-treatment interaction estimator's sampling variance,
built from a within-subject two-group contrast and the exact
variance of a mean of AR(1)-correlated observations. It
reproduces `r fmt(100 * r2_var, 0)`% of the variation in the
interaction estimator's Monte Carlo sampling variance and
predicts simulated power to within a mean absolute error of
`r fmt(mae_power, 2)`, using a single fitted nuisance-variance
parameter and no new simulation. A companion attenuation factor
correctly ranks trial designs by their vulnerability to carryover
but should not be read as a quantitative bias estimate, because
it discards timing information a properly specified mixed-model
analysis recovers. Together these give a concrete, quantitative
starting point for the design question that motivated this
communication, namely what to prioritize given a mandatory
open-label first phase, without displacing the full Monte Carlo
comparison the compendium's other manuscripts continue to supply
for final design decisions.

# Reproducibility

No new Monte Carlo simulation was run for this manuscript. All
numbers reported here are produced by
`analysis/scripts/nof1-design-efficiency/01-closed-form-validation.R`,
which reads the existing output of
`analysis/scripts/dgp-combined/01-run-combined-factorial.R`
(`analysis/data/dgp-combined/01-combined-summary.rds`) and writes
`analysis/data/nof1-design-efficiency/closed-form-validation.{rds,csv}`.

# References
