---
title: "Dissecting Treatment Response: The BR-ER-TV Framework
  and Gompertz Trajectory Modeling in N-of-1 Trial Design"
subtitle: "A Clinician's Guide to the Hendrickson et al. (2020)
  Simulation, with Worked Examples from the Raskind et al.
  (2013) Prazosin Trial"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
    number_sections: true
  pdf_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.width = 8, fig.height = 4.5
)
library(data.table)
library(ggplot2)
library(minpack.lm)
```

# Why This Document Exists

When a patient improves on a medication, the natural
clinical question is straightforward: did the drug work?
The answer, however, is rarely simple. Clinical improvement
is a composite of at least three distinct processes operating
simultaneously, and standard parallel-group randomized
controlled trials (PG-RCTs) are poorly equipped to
disentangle them at the individual level.

Hendrickson et al. (2020) proposed an aggregated N-of-1
trial design that addresses this limitation directly. Their
simulation study, published in *Frontiers in Digital Health*,
compared the statistical power of four trial architectures for
validating a predictive biomarker -- standing systolic blood
pressure -- against treatment-specific response to the PTSD
pharmacotherapeutic prazosin. The simulation parameters were
grounded in data from the Raskind et al. (2013) parallel-group
RCT of prazosin in active-duty soldiers with combat-related
PTSD (*Am J Psychiatry* 170:1003--1010).

This document serves two purposes. First, it provides an
accessible exposition of the three-factor decomposition
(BR, ER, TV) that underlies the Hendrickson simulation,
with concrete clinical examples drawn from the prazosin
trial. Second, it reconstructs the Gompertz trajectory
parameters from the published Raskind data, allowing the
reader to see exactly how abstract simulation inputs map
to observable clinical outcomes. The intended audience is
clinicians with moderate statistical training who wish to
understand the Hendrickson paper at a level deeper than the
abstract but without requiring facility in simulation
programming.


# The Clinical Problem: What Does 'Response' Actually Mean?

## A Familiar Scenario

Consider a soldier returning from deployment in Iraq with
severe combat-related PTSD. His Clinician-Administered PTSD
Scale (CAPS) total score is 82 at baseline -- well above the
50-point threshold used by Raskind et al. (2013) for trial
eligibility. He is enrolled in a 15-week double-blind RCT of
prazosin, an alpha-1 adrenoreceptor antagonist originally
developed for hypertension, now repurposed for PTSD
nightmares. His standing systolic blood pressure at baseline
is 130 mmHg.

After 15 weeks on prazosin, his CAPS score drops to 55 -- a
27-point improvement. Is this the drug working?

The honest answer is: partly. His improvement is an
inseparable mixture of at least three processes:

1. **The drug itself** reduced his hyperarousal and
   nightmares by blocking alpha-1 adrenergic receptors
   in the CNS (the *biological response*).

2. **His expectation of benefit** -- the knowledge that he
   was enrolled in a trial and might be receiving active
   medication -- produced neurobiologically real symptom
   relief through dopaminergic and anti-inflammatory
   pathways (the *expectancy-related response*).

3. **The passage of time** -- the structure of regular
   clinical contact, regression to the mean, and natural
   fluctuation in PTSD severity -- contributed its own
   trajectory of change (the *time-varying response*).

A standard PG-RCT can estimate the *average* drug effect
across the group (prazosin minus placebo), but it cannot
tell this individual soldier how much of his 27-point
improvement was pharmacological and how much was
non-specific. More critically for precision medicine, it
cannot determine whether his baseline blood pressure of
130 mmHg *predicted* his drug-specific response.

## Why the Standard RCT Falls Short

The Raskind et al. (2013) trial enrolled 67 soldiers
(32 prazosin, 35 placebo) and found a statistically
significant 11.3-point advantage for prazosin on the CAPS
total score (p = 0.02). That group-level finding is
valuable but limited in two important ways.

First, participants randomized to placebo provide no
information about what their drug-specific response would
have been. Second, participants randomized to prazosin who
improve cannot be distinguished from participants who would
have improved equally on placebo. As Hendrickson et al.
(2020) note, this makes PG-RCTs 'poorly optimized for
quantifying the relationship of a biomarker measured at
baseline to a treatment-specific response.'

The key insight of the N-of-1 approach is that each
participant serves as his or her own control, cycling
through periods on active drug and placebo. This within-person
contrast isolates the biological response from the
non-specific components far more efficiently than
between-group comparisons.


# The Three-Factor Decomposition

## Overview

The Hendrickson simulation models clinical symptom change
over time as the linear sum of three factors:

$$Y_{it} = \text{BL}_i + \text{BR}_{it} + \text{ER}_{it} +
\text{TV}_{it} + \varepsilon_{it}$$

where $i$ indexes participants, $t$ indexes timepoints,
$\text{BL}_i$ is the baseline symptom level, and
$\varepsilon_{it}$ is measurement noise. Each of the three
response factors captures a distinct causal mechanism.

## Biological Response (BR)

**Definition.** BR is the direct pharmacological effect of
the drug on the patient's physiology. It depends on drug
exposure and individual pharmacokinetics, and is
independent of the patient's beliefs or expectations.

**Clinical example.** Prazosin is an alpha-1 adrenoreceptor
antagonist that reduces norepinephrine effects at CNS
alpha-1 adrenoreceptors (Raskind et al. 2013). Enhanced
CNS adrenergic activity contributes to PTSD
pathophysiology, particularly the hyperarousal symptoms
assessed by the CAPS. When a soldier takes prazosin, the
drug accumulates over the 6-week titration period,
blocking these receptors with increasing effectiveness.
His hyperarousal symptoms decline -- not because he
believes they will, but because the receptor is physically
occupied.

Critically, Raskind et al. (2013) note that prazosin has a
'short duration of action (6 to 10 hours),' necessitating
twice-daily dosing. This means BR is pharmacokinetically
tied to recent drug exposure. When prazosin is withdrawn,
the drug washes out within days and BR returns to zero.
This on-off property is precisely what makes crossover and
N-of-1 designs informative: the contrast between on-drug
and off-drug periods within the same person isolates BR
from everything else.

**What BR looks like over time.** BR does not increase
linearly. Each additional day of drug exposure produces
diminishing returns as receptors approach saturation. The
trajectory follows a saturating curve: rapid initial
improvement, then a gradual leveling off toward a
biological ceiling. In the Raskind data, the
prazosin-minus-placebo difference grew quickly through
week 7 and then began to plateau, consistent with this
pattern.

## Expectancy-Related Response (ER)

**Definition.** ER is the component of improvement
attributable to the patient's belief or expectation that
he or she is receiving effective treatment. This is the
placebo effect in its neurobiological sense -- not a mere
artifact, but a measurable physiological response driven
by expectation.

**Clinical example.** When our soldier is told he will
receive either prazosin or placebo, the mere act of
enrollment and clinical attention activates hope and
expectation. Research in placebo neuroscience has
demonstrated that expectation of benefit can activate
endogenous opioid and dopamine systems, reduce
inflammatory markers, and produce measurable symptom
relief. In the Raskind trial, placebo participants improved
by 13.8 CAPS points over 15 weeks -- a substantial change
that cannot be attributed to any pharmacological mechanism.

**Why ER depends on trial phase.** The magnitude of ER
varies with the participant's awareness of treatment
assignment:

| Trial phase                  | Expectancy | Reasoning |
|:-----------------------------|:----------:|:----------|
| Open-label (on active drug)  | 1.0        | Full knowledge of active treatment |
| Blinded (drug vs. placebo)   | 0.5        | Uncertain whether receiving drug |
| Known placebo or washout     | ~0         | No expectation of drug benefit |

This is why the Hendrickson N-of-1 design (their Design 4)
begins with an 8-week open-label prazosin phase. During this
period, both BR and ER are fully engaged, producing maximum
clinical improvement and allowing acutely symptomatic
patients to stabilize before entering the blinded phases.
The subsequent blinded discontinuation and crossover blocks
then manipulate drug exposure while holding expectancy at
the intermediate level (0.5), enabling the analysis to
separate BR from ER.

**Why ER must be modeled separately.** If ER is absorbed
into a generic 'error' term, the model conflates
pharmacological and psychological sources of improvement.
A patient whose 20-point improvement comprises 8 points of
BR and 12 points of ER would be indistinguishable from a
patient whose 20-point improvement comprises 15 points of
BR and 5 points of ER. Yet these two patients have
fundamentally different relationships between their
biomarker and their drug-specific response. Collapsing BR
and ER obscures precisely the signal that biomarker
validation requires.

## Time-Varying Response (TV)

**Definition.** TV is the component of symptom change that
occurs independently of any treatment -- the natural
disease trajectory. It encompasses regression to the mean,
the therapeutic effect of regular clinical contact, and
the patient's evolving life circumstances.

**Clinical example.** Even without prazosin, our soldier's
PTSD severity is not static. Over the 15 weeks of the
trial, he has regular visits with a clinician, keeps a
structured schedule, and may experience natural fluctuations
in symptom severity related to the calendar (proximity to
deployment anniversary), employment, and social support. Some
of the 13.8-point improvement seen in the Raskind placebo
group reflects TV: structured clinical contact and the
passage of time both contribute to symptom change.

**Why TV cannot be absorbed into a time covariate.** A
simple linear time term assumes that symptoms change at a
constant rate throughout the trial. In practice, the
trajectory is typically nonlinear: rapid early improvement
(regression to the mean, engagement with clinical
structure) that gradually slows. The Gompertz function
captures this S-shaped trajectory. Moreover, TV is
individual-specific. Two patients may have very different
natural trajectories -- one improving steadily, another
worsening due to external stressors -- and a single time
covariate cannot accommodate this heterogeneity.

## Putting the Three Factors Together

Consider our soldier's trajectory through the Hendrickson
N-of-1 design (Design 4):

**Weeks 1--8 (open-label prazosin):**

- BR = 9 points (prazosin accumulates, hyperarousal declines)
- ER = 6 points (full expectancy: he knows he is on drug)
- TV = 4 points (clinical structure, regression to mean)
- Total improvement = 19 points. He feels substantially better.

**Weeks 9--12 (blinded discontinuation -- randomly assigned
to placebo without his knowledge):**

- BR = 0 (prazosin washes out within days; no drug, no biological effect)
- ER = 3 points (expectancy at 0.5: he is told he 'may or may not' be on drug)
- TV = 5 points (natural trajectory continues to plateau)
- Total improvement from baseline = 8 points. He notices a setback.

**Weeks 13--16 (crossover period 1 -- on active drug,
blinded):**

- BR = 8 points (prazosin restored; biological effect returns)
- ER = 3 points (still blinded; expectancy remains at 0.5)
- TV = 5 points (natural trajectory at plateau)
- Total improvement from baseline = 16 points.

**What the N-of-1 design reveals:** The contrast between
the blinded discontinuation (total = 8) and the crossover
period on drug (total = 16) isolates the biological
response: 16 - 8 = 8 points, which is the within-person
BR estimate. This is far more informative than the
group-level 11.3-point difference from the PG-RCT, because
it is specific to this patient and uncontaminated by
between-person variability in ER and TV.

**What a one-component model would conclude:** 'Improvement
varied from 8 to 19 points across phases -- noisy, hard to
interpret.'

**What the three-component model reveals:** 'BR is
consistent at 8--9 points when on drug and 0 when off. ER
contributes 3--6 points depending on expectancy. TV
plateaus at 4--5 points. The drug works for this patient;
expectations modulate the total but do not account for the
drug-specific effect.'


# The Biomarker Question

## From Group-Level Efficacy to Individual Prediction

The Raskind et al. (2013) trial established that prazosin
is, on average, more effective than placebo for PTSD. But
the precision medicine question is different: *Can we
predict, before treatment begins, which individual patients
will benefit most from prazosin?*

A separate post-hoc analysis by Raskind et al. (2016,
*Biol Psychiatry* 80:736--742) found that higher
pretreatment standing systolic blood pressure was
associated with greater PTSD symptom reduction in soldiers
treated with prazosin. This makes mechanistic sense:
elevated standing blood pressure may reflect higher
baseline CNS adrenergic tone, and patients with more
adrenergic dysregulation have more 'room' for prazosin to
work.

## Why Biomarker Validation Requires Decomposition

The statistical test that Hendrickson et al. (2020) simulate
is not 'Does prazosin work?' but rather 'Does the baseline
biomarker (blood pressure) predict the magnitude of the
drug-specific response (BR)?'

This distinction matters because the biomarker should
predict *only* BR, not ER or TV. If elevated blood pressure
predicted expectancy-related improvement, that would be a
confound, not a biomarker-treatment interaction. And if it
predicted the time-varying trajectory, the association would
be spurious. The three-factor decomposition isolates the
target quantity (BR) from the noise (ER + TV), concentrating
statistical power on the clinically relevant signal.

The analysis model from Hendrickson et al. (2020) formalizes
this as:

$$S_{i,t} = \beta_{i,0} + \beta_1 \cdot bm + \beta_2
\cdot Db + \beta_3 \cdot t + \beta_4 \cdot bm \cdot Db$$

where $bm$ is the baseline biomarker, $Db$ is drug exposure,
$t$ is time, and the interaction term $\beta_4$ tests
whether the biomarker predicts drug-specific response. A
non-zero $\beta_4$ indicates that the biomarker is
predictive -- the precision medicine finding.


# The Modified Gompertz Function

## Why Trajectories Are Not Linear

Each of the three response factors follows a characteristic
time course. The improvement does not accumulate at a
constant rate; instead, it follows a saturating pattern:

- **Early phase:** Rapid change as the mechanism engages
  (drug accumulates, expectancy forms, regression to mean
  operates).
- **Middle phase:** Improvement continues but decelerates.
- **Late phase:** The factor approaches its biological or
  psychological ceiling.

A linear model would predict indefinite improvement -- 2
CAPS points per week, every week, forever. After 50 weeks,
that would imply a 100-point improvement on a scale with a
maximum of about 136. This is biologically implausible.

The modified Gompertz function captures the saturating
pattern with three interpretable parameters:

$$f(t) = \text{max} \cdot e^{-\text{disp}
\cdot e^{-\text{rate} \cdot t}}$$

with an offset-rescaling to ensure $f(0) = 0$ and
$f(\infty) = \text{max}$.

```{r}
modgompertz <- function(t, maxr, disp, rate) {
  y <- maxr * exp(-disp * exp(-rate * t))
  vert_offset <- maxr * exp(-disp * exp(-rate * 0))
  y <- y - vert_offset
  y <- y * (maxr / (maxr - vert_offset))
  y
}
```

## Parameter Interpretation

| Parameter | Symbol       | Clinical meaning | Example |
|:----------|:-------------|:-----------------|:--------|
| Maximum   | $\text{max}$ | Ceiling effect: the most this factor can contribute | BR max = 11: prazosin can reduce CAPS by at most 11 points |
| Rate      | $\text{rate}$ | Speed of onset: how quickly the factor reaches half its ceiling | rate = 0.42: half-maximum BR reached by approximately week 4 |
| Displacement | $\text{disp}$ | Curve shape: controls the inflection point and steepness | disp = 5: produces a delayed onset followed by rapid rise |

The displacement parameter deserves particular comment. With
$\text{disp} = 5$, the Gompertz curve has a pronounced
sigmoidal shape: near-zero improvement in the first 1--2
weeks, then a rapid rise, then a plateau. This matches the
clinical observation from the Raskind trial that prazosin
required a 6-week titration period before reaching
therapeutic doses, and that most improvement occurred
between weeks 2 and 10.


# Reconstructing the Raskind Data

## Extracting CAPS Trajectories

The Raskind et al. (2013) trial assessed CAPS total scores
at baseline and weeks 7, 11, and 15. Table 3 of that paper
provides adjusted means at baseline and week 15. Weeks 7
and 11 values are interpolated from Figure 1D, which plots
the full CAPS total score trajectory for both groups.

```{r}
weeks <- c(0, 7, 11, 15)

caps_prazosin <- c(77.3, 60, 54, 52.4)
caps_placebo  <- c(85.7, 76, 74, 71.9)

trial_data <- data.table(
  week = rep(weeks, 2),
  caps = c(caps_prazosin, caps_placebo),
  group = rep(c("Prazosin", "Placebo"), each = 4)
)

trial_data[, baseline := caps[1], by = group]
trial_data[, improvement := baseline - caps]
```

The baseline CAPS scores differed between groups (prazosin
77.3, placebo 85.7), reflecting the expected variability
in a trial of this size (N = 67). The Hendrickson simulation
uses the pooled baseline mean of 83.07 (SD = 18.48),
averaged across both groups.

```{r fig.height=4}
ggplot(trial_data, aes(x = week, y = improvement,
                       color = group, shape = group)) +
  geom_point(size = 3) +
  geom_line(linewidth = 0.8) +
  labs(
    x = "Week",
    y = "Improvement from baseline (CAPS points)",
    title = "CAPS Score Improvement: Raskind et al. (2013)",
    subtitle = "Reconstructed from Table 3 and Figure 1D"
  ) +
  scale_color_manual(values = c(
    "Placebo" = "firebrick", "Prazosin" = "steelblue"
  )) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "right")
```

Both groups improve over time, but the prazosin group
improves more. The gap between the curves widens most
between baseline and week 7 (during the titration period),
then stabilizes. This widening gap is the observable
signature of BR.

## Decomposing the Trajectories into BR and TV+ER

The logic of the decomposition follows directly from the
factor definitions:

- **Placebo group improvement = TV + ER.** These patients
  receive no drug, so BR = 0 at all timepoints. Their
  entire improvement is non-specific: some combination of
  natural trajectory (TV) and expectancy (ER). Because this
  was a double-blind trial, placebo participants believed
  they might be receiving active drug, so ER is non-zero.

- **BR = Prazosin improvement minus Placebo improvement.**
  Subtracting the placebo trajectory from the prazosin
  trajectory removes the non-specific components,
  isolating the drug-specific effect.

```{r}
praz <- trial_data[group == "Prazosin"]
plac <- trial_data[group == "Placebo"]

br_improvement <- praz$improvement - plac$improvement
tver_improvement <- plac$improvement

factor_data <- data.table(
  week = weeks,
  BR = br_improvement,
  TV_ER = tver_improvement
)

knitr::kable(
  factor_data, digits = 2,
  caption = "Factor decomposition of CAPS improvement",
  col.names = c("Week", "BR (drug-specific)",
                "TV+ER (non-specific)")
)
```

```{r}
factor_long <- melt(
  factor_data, id.vars = "week",
  variable.name = "factor",
  value.name = "improvement"
)

ggplot(factor_long, aes(x = week, y = improvement,
                        color = factor, shape = factor)) +
  geom_point(size = 3) +
  geom_line(linewidth = 0.8) +
  labs(
    x = "Week",
    y = "Improvement (CAPS points)",
    title = "Factor Decomposition of Treatment Response",
    subtitle = paste(
      "BR = prazosin minus placebo;",
      "TV+ER = placebo trajectory"
    )
  ) +
  scale_color_manual(
    values = c("BR" = "steelblue", "TV_ER" = "sienna"),
    labels = c("BR (drug-specific)",
               "TV+ER (non-specific)")
  ) +
  theme_minimal(base_size = 12)
```

Several features of this decomposition merit attention.
First, the TV+ER component is substantial -- comparable in
magnitude to BR. By week 15, the placebo group improved by
approximately 14 CAPS points, while the drug-specific
difference was approximately 11 points. This means that
roughly half of the prazosin group's total improvement was
non-specific. This is not unusual in psychiatric trials; it
is precisely why isolating BR is important for biomarker
validation.

Second, the Raskind trial was double-blinded throughout, so
the expectancy component is at an intermediate level
(participants knew they had a 50% chance of receiving
prazosin). In the Hendrickson N-of-1 design, the open-label
phase would produce a larger ER contribution, followed by a
reduction during blinded phases.

## The 'Tabula Rasa' Assumption

The Raskind data provide a single placebo trajectory that
represents TV + ER combined. There is no empirical way to
separate TV from ER using these data alone, because every
placebo participant simultaneously experienced both the
passage of time (TV) and the expectation of possible benefit
(ER).

Hendrickson et al. (2020) address this by adopting a
'tabula rasa' (TaRa) parameter set in which TV and ER are
assigned identical Gompertz parameters: max = 6.506,
disp = 5, rate = 0.35, SD = 10 for each. The combined
TV + ER trajectory is simply split equally between the two
factors. This is a modeling convenience, not an empirical
claim. The sensitivity analyses in Hendrickson Figures 5A
and 5B explore what happens when TV and ER parameters
diverge.

For the clinician, the practical implication is this: the
exact partition of non-specific improvement into TV and ER
matters less than the fact that both are separated from BR.
The biomarker validation test targets $\beta_4$ (the
biomarker-by-drug interaction), which depends on BR alone.


# Fitting the Gompertz Model to the Raskind Data

## Fitting BR (Drug-Specific Response)

We fit the modified Gompertz function to the BR trajectory
(prazosin minus placebo improvement at each timepoint)
using nonlinear least squares with starting values informed
by the Hendrickson publication parameters.

```{r}
br_fit_data <- data.table(
  t = factor_data$week,
  y = factor_data$BR
)

fit_br <- nlsLM(
  y ~ modgompertz(t, maxr, disp, rate),
  data = br_fit_data,
  start = list(maxr = 12, disp = 5, rate = 0.3),
  lower = c(maxr = 1, disp = 0.1, rate = 0.01),
  upper = c(maxr = 50, disp = 20, rate = 2)
)

br_params <- coef(fit_br)
summary(fit_br)
```

**Fitted BR parameters:**
$\text{max} = `r sprintf("%.3f", br_params["maxr"])`$,
$\text{disp} = `r sprintf("%.3f", br_params["disp"])`$,
$\text{rate} = `r sprintf("%.3f", br_params["rate"])`$

**Hendrickson publication parameters:**
$\text{max} = 10.986$,
$\text{disp} = 5.000$,
$\text{rate} = 0.420$

The fitted parameters should approximate the Hendrickson
values, since both are derived from the same underlying
Raskind data. Any discrepancy reflects the difference
between fitting group-level summary statistics (our
approach, using 4 timepoints read from published figures)
versus fitting individual-level data (the Hendrickson
authors had access to the original dataset).

## Fitting TV+ER (Non-Specific Response)

```{r}
tver_fit_data <- data.table(
  t = factor_data$week,
  y = factor_data$TV_ER
)

fit_tver <- nlsLM(
  y ~ modgompertz(t, maxr, disp, rate),
  data = tver_fit_data,
  start = list(maxr = 15, disp = 5, rate = 0.3),
  lower = c(maxr = 1, disp = 0.1, rate = 0.01),
  upper = c(maxr = 50, disp = 20, rate = 2)
)

tver_params <- coef(fit_tver)
summary(fit_tver)
```

**Fitted TV+ER parameters:**
$\text{max} = `r sprintf("%.3f", tver_params["maxr"])`$,
$\text{disp} = `r sprintf("%.3f", tver_params["disp"])`$,
$\text{rate} = `r sprintf("%.3f", tver_params["rate"])`$

**Hendrickson publication parameters (each of TV and ER):**
$\text{max} = 6.506$,
$\text{disp} = 5.000$,
$\text{rate} = 0.350$

Note that the Hendrickson max of 6.506 applies to each of
TV and ER separately. Under the tabula rasa assumption,
TV + ER combined has an effective maximum of approximately
$2 \times 6.506 = 13.0$ CAPS points. The fitted TV+ER
maximum from the Raskind placebo data should approximate
this value.

## Comparing Fitted and Publication Curves

```{r fig.width=10, fig.height=5}
t_smooth <- seq(0, 20, by = 0.1)

br_curve <- data.table(
  t = t_smooth,
  y = modgompertz(t_smooth, br_params["maxr"],
                  br_params["disp"], br_params["rate"]),
  source = "Fitted BR"
)
tver_curve <- data.table(
  t = t_smooth,
  y = modgompertz(t_smooth, tver_params["maxr"],
                  tver_params["disp"], tver_params["rate"]),
  source = "Fitted TV+ER"
)

br_hend <- data.table(
  t = t_smooth,
  y = modgompertz(t_smooth, 10.986, 5, 0.42),
  source = "Hendrickson BR"
)
tver_hend <- data.table(
  t = t_smooth,
  y = modgompertz(t_smooth, 6.506, 5, 0.35),
  source = "Hendrickson TV+ER (single factor)"
)

curves <- rbind(br_curve, tver_curve, br_hend, tver_hend)

obs <- rbind(
  data.table(t = weeks, y = factor_data$BR,
             source = "Observed BR"),
  data.table(t = weeks, y = factor_data$TV_ER,
             source = "Observed TV+ER")
)

ggplot() +
  geom_line(
    data = curves,
    aes(x = t, y = y, color = source, linetype = source),
    linewidth = 0.8
  ) +
  geom_point(
    data = obs,
    aes(x = t, y = y, color = source),
    size = 3
  ) +
  scale_color_manual(values = c(
    "Fitted BR" = "steelblue",
    "Hendrickson BR" = "steelblue",
    "Fitted TV+ER" = "sienna",
    "Hendrickson TV+ER (single factor)" = "sienna",
    "Observed BR" = "steelblue",
    "Observed TV+ER" = "sienna"
  )) +
  scale_linetype_manual(values = c(
    "Fitted BR" = "solid",
    "Hendrickson BR" = "dashed",
    "Fitted TV+ER" = "solid",
    "Hendrickson TV+ER (single factor)" = "dashed"
  )) +
  labs(
    x = "Week",
    y = "Improvement (CAPS points)",
    title = paste(
      "Gompertz Fits: Raskind Data vs.",
      "Hendrickson Publication Parameters"
    ),
    subtitle = paste(
      "Solid = fitted to Raskind group means;",
      "Dashed = Hendrickson simulation values"
    )
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "right")
```

The dashed Hendrickson TV+ER curve represents a *single*
factor (either TV or ER alone, each with max = 6.506). The
solid fitted TV+ER curve represents the *combined*
non-specific trajectory from the placebo group. The
Hendrickson curve should therefore lie below the fitted
curve, roughly at half its height.


# Variance and the Covariance Structure

## Between-Person Variability in Each Factor

The Gompertz parameters describe the *mean* trajectory.
But patients vary: one soldier's BR might be 12 CAPS
points while another's is only 4. This between-person
variability is captured by the standard deviation of each
factor, which enters the simulation through the covariance
matrix used to generate individual trajectories.

From Raskind Table 3, we can recover the raw CAPS total
score SD from the 95% confidence intervals:

- Prazosin: mean = 77.3, 95% CI = 69.1--85.5,
  giving $\text{SE} = 4.1$ and
  $\text{SD} = 4.1 \times \sqrt{32} = 23.2$
- Placebo: mean = 85.7, 95% CI = 78.0--93.3,
  giving $\text{SE} = 3.9$ and
  $\text{SD} = 3.9 \times \sqrt{35} = 23.1$

For the change-from-baseline SD, the reported difference
in change is 11.3 points (95% CI = 2.0--20.7). The SE of
the difference is $(20.7 - 2.0) / (2 \times 1.96) = 4.77$.
Assuming equal SDs across groups:

$$4.77 = \text{SD} \times \sqrt{1/32 + 1/35}
       = \text{SD} \times 0.247$$

This yields an SD of change of approximately 19.3 CAPS
points.

The Hendrickson factor SDs are substantially smaller than
this total SD because they represent the variability of
each individual factor, not the total outcome. The total
outcome variability is the sum of variances across all
factors plus baseline variability, partially offset by
cross-factor correlations.

**Hendrickson factor standard deviations:**

| Factor              |  SD  |
|:--------------------|-----:|
| BR (drug-specific)  |    8 |
| TV (time-variant)   |   10 |
| ER (expectancy)     |   10 |
| Baseline (BL)       | 18.5 |

```{r}
total_var <- 18.5^2 + 10^2 + 10^2 + 8^2
```

The sum of marginal variances is
`r sprintf("%.0f", total_var)`, yielding an implied total
SD of `r sprintf("%.1f", sqrt(total_var))`. The observed
CAPS SD is approximately 23, which is lower than this
implied value. This is expected: the factors are not
independent, and cross-correlations (particularly between
baseline severity and the response factors) reduce the
total variance below the sum of its components.

## Why Separate Variances Matter

Each factor has its own variance because the underlying
mechanisms operate on different timescales and are subject
to different sources of individual variation:

- **BR variance** (SD = 8) reflects individual differences
  in pharmacokinetics: metabolism rate, receptor density,
  drug absorption. Two soldiers on identical prazosin doses
  may achieve different CNS drug concentrations.

- **TV variance** (SD = 10) reflects heterogeneity in
  natural disease trajectory: life circumstances,
  comorbidities, social support, proximity to trauma cues.

- **ER variance** (SD = 10) reflects individual differences
  in suggestibility, prior treatment experience, and the
  strength of the therapeutic alliance.

The simulation generates correlated factor values from a
multivariate normal distribution with a covariance matrix
that encodes both the marginal variances and the
between-factor correlations (including the correlation
between baseline biomarker and BR, which is the target of
the validation test).


# Parameter Comparison

```{r}
comparison <- data.table(
  Factor = c("BR", "TV+ER (combined)",
             "TV (Hendrickson)", "ER (Hendrickson)"),
  max_fitted = c(br_params["maxr"], tver_params["maxr"],
                 NA, NA),
  max_hendrickson = c(10.986, NA, 6.506, 6.506),
  disp_fitted = c(br_params["disp"], tver_params["disp"],
                   NA, NA),
  disp_hendrickson = c(5, NA, 5, 5),
  rate_fitted = c(br_params["rate"], tver_params["rate"],
                   NA, NA),
  rate_hendrickson = c(0.42, NA, 0.35, 0.35),
  sd_hendrickson = c(8, NA, 10, 10)
)

knitr::kable(
  comparison, digits = 3,
  caption = paste(
    "Comparison of parameters fitted to Raskind",
    "group means vs. Hendrickson publication values"
  )
)
```


# How Trial Design Affects Power: Lessons from the Simulation

## The Four Designs Compared

Hendrickson et al. (2020) compared four trial architectures,
all matched in total duration and number of assessment
points (see their Figure 2):

1. **Open-label (OL):** 20 weeks of open-label prazosin.
   All participants know they are on drug. BR and ER are
   fully confounded; the biomarker-by-drug interaction
   cannot be tested directly.

2. **Open-label + blinded discontinuation (OL+BDC):**
   16 weeks of open-label prazosin, then a 4-week blinded
   discontinuation. The discontinuation block introduces
   a within-person contrast but is brief.

3. **Traditional crossover (CO):** Two 10-week periods,
   randomized to drug-then-placebo or placebo-then-drug.
   Both periods are blinded. Provides within-person
   contrast but no open-label stabilization phase.

4. **Proposed N-of-1 design:** 8 weeks open-label, 4 weeks
   blinded discontinuation, then two 4-week crossover
   blocks. This design combines the clinical advantages
   of open-label stabilization with the statistical
   advantages of multiple within-person contrasts.

## Key Findings

The simulation results (Hendrickson Figure 4A) demonstrate
that:

- **Design 4 (N-of-1) had the highest power** for detecting
  the biomarker-treatment interaction under the tabula rasa
  parameter set, outperforming both the open-label and
  OL+BDC designs. With 70 subjects and a biomarker-BR
  correlation of 0.6, the N-of-1 design achieved power of
  approximately 0.99 without censoring.

- **The traditional crossover was comparable** in power to
  the N-of-1 design (also approximately 0.99 with 70
  subjects), but lacked the clinical advantage of beginning
  with open-label stabilization.

- **The open-label design performed poorly** (power of
  approximately 0.25 with 70 subjects), because it cannot
  separate BR from ER.

- **Carryover was devastating to power** (Hendrickson
  Figure 4B). Even a short carryover half-life of 0.1 weeks
  reduced N-of-1 power substantially, because it blurs the
  on-drug/off-drug contrast that the design depends on.
  For prazosin, with its 6--10 hour duration of action, the
  pharmacokinetic half-life is approximately 0.02--0.03
  weeks, so carryover should be negligible between weekly
  assessments.

## Clinical Implications

For the clinician considering an N-of-1 trial to validate
a predictive biomarker, the Hendrickson results suggest
several principles:

- **Include an open-label phase** to stabilize acutely
  symptomatic patients and maximize enrollment. This is
  ethically important: patients referred for PTSD treatment
  are unlikely to accept initial randomization to placebo.

- **Choose drugs with short half-lives** (or at least
  short pharmacodynamic durations). Carryover effects
  erode the power advantage of crossover designs. Prazosin,
  with its 6--10 hour duration of action, is well suited
  to this approach.

- **Maximize the number of within-person contrasts.** The
  N-of-1 design's advantage comes from having multiple
  on-drug and off-drug periods within the same patient,
  which averages out the noise from any single transition.

- **The signal-to-noise ratio matters more than sample
  size.** Figure 5A of Hendrickson et al. shows that when
  the BR maximum is large relative to the TR and ER
  maximums, power approaches 1.0 even with modest sample
  sizes. The practical implication is that drugs with
  strong biological effects and short durations of action
  are ideal candidates for N-of-1 biomarker validation.


# Sensitivity to Gompertz Parameters: What Figure 5 Tells Us

The preceding sections establish that the simulation's power
depends on the trial design and the biomarker-BR correlation.
But a question of equal practical importance is: *How
sensitive are these power estimates to the Gompertz
parameters themselves?* If the true trajectory parameters
differ from the tabula rasa values, do the design
recommendations still hold?

Hendrickson et al. (2020) address this directly in their
Figure 5, which systematically varies the maximum response
values (Figure 5A) and the rate parameters (Figure 5B)
across all three factors and all four trial designs. These
results have important implications for clinicians
considering which drugs and which outcome measures are
amenable to N-of-1 biomarker validation.

## Effect of Maximum Response Values (Figure 5A)

Figure 5A presents a 3 x 3 grid of heat maps for each
trial design. The rows correspond to the BR maximum (5, 8,
or 11 CAPS points), and the columns correspond to the ER
maximum (5, 8, or 11), with each cell within a panel
varying the TR maximum (5, 8, or 11). The biomarker-BR
correlation is fixed at 0.6 and N = 35.

The results reveal a clear and clinically interpretable
pattern:

**Increasing BR max improves power across all designs.**
This is the signal: a larger drug-specific ceiling means
more separation between high-biomarker and low-biomarker
patients, making the interaction easier to detect. When
BR max = 11 (roughly equivalent to the tabula rasa value
of 10.986), power exceeds 0.90 for the crossover and
N-of-1 designs regardless of the ER and TR maximums.
When BR max drops to 5 (a weak drug effect), power falls
below 0.20 for the open-label design and remains marginal
even for the crossover.

*Clinical translation:* A drug that produces at most a
5-point improvement on the CAPS (out of a maximum possible
score of approximately 136) is a poor candidate for
biomarker validation via any design. The drug effect is
simply too small relative to the noise floor. Prazosin,
with its estimated BR max of approximately 11 CAPS points,
sits in a favorable region of this parameter space.

**Increasing ER max decreases power, especially for
designs with expectancy variation.** Larger non-specific
expectancy effects add variance to the outcome without
contributing to the biomarker-BR signal. This dilution is
most pronounced in the OL+BDC and N-of-1 designs, which
include open-label phases where expectancy is at its
maximum (1.0) followed by blinded phases where it drops to
0.5. The resulting within-person ER change adds noise to
the treatment contrast.

*Clinical translation:* Conditions with large placebo
responses (e.g., depression, chronic pain) will require
larger sample sizes or stronger drug effects to achieve
adequate power. Conditions where the placebo response is
modest (e.g., neurodegenerative diseases with objective
biomarkers) may be more favorable for N-of-1 designs.

**Increasing TR max also decreases power.** The
time-varying component is pure noise from the perspective
of biomarker validation. It affects all participants
equally regardless of drug assignment, so it does not
bias the interaction estimate, but it inflates the
residual variance and reduces the signal-to-noise ratio.

*Clinical translation:* Diseases with highly variable
natural trajectories will require larger samples. One
strategy to mitigate this is to shorten the trial (reducing
the window over which TV accumulates), but this must be
balanced against the need for sufficient treatment exposure
to observe BR.

**The ratio of BR max to non-specific maximums is the
key quantity.** When BR max = 11 and both ER max and TR
max = 5 (a favorable signal-to-noise ratio), even the
open-label design achieves power above 0.60. When the
ratio reverses (BR max = 5, ER max = TR max = 11), power
is uniformly poor. This highlights that the choice of
drug and outcome measure matters as much as the choice of
trial design.

## Effect of Rate Parameters (Figure 5B)

Figure 5B uses the same grid layout but varies the rate
parameters (0.2, 0.35, or 0.5 per week) for each factor,
holding the maximums at their tabula rasa values. The
biomarker-BR correlation is fixed at 0.3 (lower than in
Figure 5A), making the power landscape more conservative.

The rate parameter controls how quickly each factor
approaches its ceiling. A rate of 0.5 means the factor
reaches half-maximum in approximately 2--3 weeks; a rate
of 0.2 means this takes 5--7 weeks. The results are less
straightforward than for the maximums:

**For designs with blinded discontinuation (OL+BDC and
N-of-1), faster BR rates generally improve power.** A
higher BR rate means the drug effect ramps up quickly
during on-drug periods and (because the offset-rescaling
enforces $f(0) = 0$) disappears quickly when the drug is
removed. This sharpens the within-person contrast between
on-drug and off-drug assessments.

**For the traditional crossover, faster BR rates
paradoxically *decrease* power across much of the
parameter space.** This occurs because in the crossover
design, each treatment period is 10 weeks long. When BR
reaches its ceiling quickly (say by week 3), the remaining
7 weeks of the period provide assessment points that are
all at the same plateau value. These 'ceiling' observations
contain no additional information about the rate of onset
but do accumulate TV and ER variance. In effect, the
crossover design wastes assessment points when the drug
effect saturates early.

*Clinical translation:* The relationship between drug
onset speed and trial design is not monotonic. For drugs
with very fast onset (analogous to high BR rate), the
N-of-1 design with its shorter treatment blocks (4 weeks)
is more efficient than the crossover with its longer blocks
(10 weeks). For drugs with slower onset, longer treatment
blocks may be necessary to observe the full BR trajectory,
favoring the crossover design.

**Faster ER rates decrease power for the OL+BDC and N-of-1
designs.** When the expectancy response ramps up quickly,
it reaches its ceiling during the open-label phase and is
fully established before the blinded transition. This means
the ER drop upon entering the blinded phase is large and
abrupt, adding a substantial within-person change that is
unrelated to BR. The analysis model must absorb this
ER-driven variance, reducing the precision of the
biomarker-by-drug interaction estimate.

**TR rate has minimal impact on power** across most of the
parameter space and most designs. This is because the
time-varying component is continuous and gradual: whether
it reaches its plateau by week 5 or week 10 has little
effect on the within-person contrasts that drive power in
the crossover and N-of-1 designs.

## Practical Synthesis

The Figure 5 results can be distilled into a set of
practical guidelines for clinicians and trialists:

1. **Screen candidate drugs by their signal-to-noise
   profile.** Before committing to an N-of-1 design,
   estimate the likely BR max from prior RCT data (as we
   did from the Raskind trial) and compare it to the
   expected ER and TR maximums. If BR max is less than half
   of ER max + TR max, power will be poor regardless of
   design.

2. **Match block length to drug onset speed.** Fast-onset
   drugs (high BR rate) benefit from shorter treatment
   blocks with more transitions (the N-of-1 design).
   Slow-onset drugs may require longer blocks to observe
   the full BR trajectory (favoring the crossover).

3. **Account for the placebo response in power
   calculations.** Conditions with large expectancy effects
   require larger sample sizes. The tabula rasa assumption
   (ER max = TR max = 6.5) may be optimistic for conditions
   like depression, where placebo response rates often
   exceed 30%.

4. **The Gompertz parameters are not nuisance quantities.**
   They directly affect the feasibility of biomarker
   validation. Reporting and estimating these parameters
   from pilot data should be a standard part of N-of-1
   trial planning.


# Limitations of This Reconstruction

Several caveats apply to the Gompertz parameter estimates
derived here:

- **Only four timepoints** (weeks 0, 7, 11, 15) are
  available from the Raskind publication, providing minimal
  constraint on a three-parameter model. With four data
  points and three parameters, there is only one residual
  degree of freedom.

- **Weeks 7 and 11 values were read from Figure 1D** of
  Raskind et al. (2013) and are approximate. The published
  table provides exact values only at baseline and week 15.

- **We are fitting group means, not individual data.** The
  Hendrickson authors had access to the original
  individual-level data from the Raskind trial, which
  would provide far more constraint on the Gompertz fit.
  Our reconstruction is a pedagogical exercise, not a
  replication.

- **The displacement parameter is poorly constrained.**
  With only four timepoints, the data cannot distinguish
  between different values of $\text{disp}$ that produce
  similar curves over the observed time window. The
  Hendrickson choice of $\text{disp} = 5$ for all factors
  was likely guided by the individual-level data and by
  theoretical considerations about the shape of treatment
  response curves.


# Summary

The Hendrickson et al. (2020) simulation makes a compelling
case for aggregated N-of-1 designs in predictive biomarker
validation. The key conceptual ingredients are:

1. **Decomposition of treatment response** into biological
   (BR), expectancy (ER), and time-varying (TV) components,
   each following a Gompertz trajectory with parameters
   derived from the Raskind et al. (2013) prazosin trial.

2. **Isolation of the target signal** (the biomarker-by-BR
   interaction) from the noise (ER + TV), which
   concentrates statistical power on the precision medicine
   question.

3. **Within-person contrasts** achieved through blinded
   discontinuation and crossover phases, which provide
   far more efficient estimation of individual-level BR
   than between-group comparisons.

4. **Practical clinical design** that begins with
   open-label stabilization to address ethical and
   enrollment concerns, while preserving statistical power
   through subsequent blinded phases.

The Gompertz parameters are not arbitrary simulation inputs.
They are grounded in observable clinical data: the shape and
magnitude of CAPS score trajectories in a real trial of
prazosin for combat-related PTSD. Understanding where these
parameters come from -- and what they represent clinically --
is essential for interpreting the simulation results and
for designing future N-of-1 trials in precision psychiatry.


# References

- Hendrickson RC, Thomas RG, Schork NJ, Raskind MA (2020).
  Optimizing aggregated N-of-1 trial designs for predictive
  biomarker validation: statistical methods and theoretical
  findings. *Front Digit Health* 2:13.

- Raskind MA, Peterson K, Williams T, et al. (2013). A
  trial of prazosin for combat trauma PTSD with nightmares
  in active-duty soldiers returned from Iraq and
  Afghanistan. *Am J Psychiatry* 170:1003--1010.

- Raskind MA, Peskind ER, Chow B, et al. (2016). Trial of
  prazosin for post-traumatic stress disorder in military
  veterans. *Biol Psychiatry* 80:736--742.


*Session info:* `r sessionInfo()$R.version$version.string`
