---
title: "Gompertz Model Fitting: Reverse-Engineering the Hendrickson
  Parameters from the Raskind et al. (2013) Trial Data"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(data.table)
library(ggplot2)
library(minpack.lm)  # for nlsLM (robust nonlinear least squares)
```

# Background

The Hendrickson et al. (2020) simulation uses a modified Gompertz
function to model the expected trajectory of each response factor
(BR, ER, TR) over time. The parameters for this function were
extracted from the Raskind et al. (2013) parallel-group RCT of
prazosin for PTSD in active-duty soldiers (Am J Psychiatry
170:1003-1010).

This document reverse-engineers the data from the published
figures and tables, fits the modified Gompertz model, and
compares the fitted parameters with those used in the simulation.

# Summary of Raskind et al. (2013)

**Trial design:** 15-week double-blind RCT, N=67 (32 prazosin,
35 placebo), with 6-week titration period. Assessments at
baseline, weeks 7, 11, and 15.

**The biomarker:** Standing systolic blood pressure (2 min after
standing). Table 3 shows baseline standing systolic BP: prazosin
group mean=125, placebo group mean=121 mmHg. Blood pressure did
not significantly differ over time or between groups (p=0.61).

**The outcome:** CAPS total score. Baseline: prazosin 77.3,
placebo 85.7. Week 15: prazosin 52.4, placebo 71.9. Difference
in change from baseline: 11.3 points (p=0.02).

**Where the Hendrickson simulation parameters came from:**

- Baseline CAPS mean (83.07) and SD (18.48): averaged from
  both groups
- Baseline BP mean (124.33) and SD (15.36): averaged from
  both groups
- BR trajectory (max=10.99, rate=0.42): fit to the
  prazosin-minus-placebo difference curve (Figure 1D shows
  the divergence)
- TV+ER trajectory (max=6.51, rate=0.35): fit to the placebo
  group trajectory

**The biomarker-treatment interaction finding:** This paper does
not test it directly. Hendrickson et al. (2020) cites a separate
post-hoc analysis (Raskind et al. 2016, Biol Psychiatry
80:736-42) that found 'higher pretreatment blood pressure is
associated with greater PTSD symptom reduction in soldiers
treated with prazosin.' That finding motivated the N-of-1 trial
design to validate the biomarker prospectively.

**Prazosin pharmacokinetics:** The paper notes prazosin has a
'short duration of action (6 to 10 hours),' which is why
twice-daily dosing was used. This maps to a carryover half-life
of roughly 3-5 hours (0.02-0.03 weeks) -- much shorter than
even the Hendrickson publication's 0.1 weeks. At these true
pharmacokinetic half-lives, the carryover residual between
weekly assessments is negligible.

# The Modified Gompertz Function

The function used in pmsimstats is:

$$f(t) = \text{max} \cdot e^{-\text{disp} \cdot e^{-\text{rate} \cdot t}}$$

with an offset-rescaling to ensure $f(0) = 0$ and
$f(\infty) = \text{max}$:

```{r}
modgompertz <- function(t, maxr, disp, rate) {
  y <- maxr * exp(-disp * exp(-rate * t))
  vert_offset <- maxr * exp(-disp * exp(-rate * 0))
  y <- y - vert_offset
  y <- y * (maxr / (maxr - vert_offset))
  y
}
```

# Extracting Data from Raskind et al. (2013)

## CAPS Total Score Trajectories

From Table 3 and Figure 1D of Raskind et al. (2013):

```{r}
# Assessment timepoints (weeks)
weeks <- c(0, 7, 11, 15)

# CAPS total scores (adjusted means from Table 3)
caps_prazosin <- c(77.3, NA, NA, 52.4)
caps_placebo  <- c(85.7, NA, NA, 71.9)

# Interpolate weeks 7 and 11 from Figure 1D
# (read approximately from the plotted curves)
caps_prazosin <- c(77.3, 60, 54, 52.4)
caps_placebo  <- c(85.7, 76, 74, 71.9)

trial_data <- data.table(
  week = rep(weeks, 2),
  caps = c(caps_prazosin, caps_placebo),
  group = rep(c("Prazosin", "Placebo"), each = 4)
)

# Compute change from baseline (improvement is positive)
trial_data[, baseline := caps[1], by = group]
trial_data[, improvement := baseline - caps]
```

## Visualize the Raw Trajectories

```{r fig.width=8, fig.height=4}
ggplot(trial_data, aes(x = week, y = improvement,
                       color = group, shape = group)) +
  geom_point(size = 3) +
  geom_line(linewidth = 0.8) +
  labs(x = "Week", y = "Improvement from baseline (CAPS points)",
       title = "CAPS Score Improvement: Raskind et al. (2013)",
       subtitle = "Reconstructed from Table 3 and Figure 1D") +
  scale_color_manual(values = c("Placebo" = "red",
                                "Prazosin" = "blue")) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "right")
```

## Decompose into Three Factors

Following Hendrickson et al. (2020):

- **TV + ER** (combined): the placebo group trajectory.
  Since placebo participants receive no drug, their
  improvement is entirely non-specific (time-variant +
  expectancy).
- **BR**: the difference between prazosin and placebo
  trajectories. This isolates the drug-specific effect.

```{r}
# Separate by group
praz <- trial_data[group == "Prazosin"]
plac <- trial_data[group == "Placebo"]

# BR = prazosin improvement - placebo improvement
br_improvement <- praz$improvement - plac$improvement

# TV+ER = placebo improvement
tver_improvement <- plac$improvement

factor_data <- data.table(
  week = weeks,
  BR = br_improvement,
  TV_ER = tver_improvement
)

knitr::kable(factor_data, digits = 2,
             caption = "Factor decomposition")
```

```{r fig.width=8, fig.height=4}
factor_long <- melt(factor_data, id.vars = "week",
                    variable.name = "factor",
                    value.name = "improvement")

ggplot(factor_long, aes(x = week, y = improvement,
                        color = factor, shape = factor)) +
  geom_point(size = 3) +
  geom_line(linewidth = 0.8) +
  labs(x = "Week",
       y = "Improvement (CAPS points)",
       title = "Factor Decomposition",
       subtitle = "BR = drug-specific; TV+ER = non-specific") +
  scale_color_manual(
    values = c("BR" = "blue", "TV_ER" = "brown"),
    labels = c("BR (drug-specific)",
               "TV+ER (non-specific)")) +
  theme_minimal(base_size = 12)
```

# Fitting the Modified Gompertz

## Fit to BR (Drug-Specific Response)

```{r}
# Remove baseline (t=0, improvement=0) for fitting
# but it is implicitly enforced by the offset-rescaling
br_fit_data <- data.table(
  t = factor_data$week,
  y = factor_data$BR
)

# Fit using nonlinear least squares
# Start values informed by Hendrickson parameters
fit_br <- nlsLM(
  y ~ modgompertz(t, maxr, disp, rate),
  data = br_fit_data,
  start = list(maxr = 12, disp = 5, rate = 0.3),
  lower = c(maxr = 1, disp = 0.1, rate = 0.01),
  upper = c(maxr = 50, disp = 20, rate = 2)
)

br_params <- coef(fit_br)
summary(fit_br)
```

**Fitted parameters:**
$\text{max} = `r sprintf("%.3f", br_params["maxr"])`$,
$\text{disp} = `r sprintf("%.3f", br_params["disp"])`$,
$\text{rate} = `r sprintf("%.3f", br_params["rate"])`$

**Hendrickson parameters:**
$\text{max} = 10.986$, $\text{disp} = 5.000$, $\text{rate} = 0.420$

## Fit to TV+ER (Non-Specific Response)

```{r}
tver_fit_data <- data.table(
  t = factor_data$week,
  y = factor_data$TV_ER
)

fit_tver <- nlsLM(
  y ~ modgompertz(t, maxr, disp, rate),
  data = tver_fit_data,
  start = list(maxr = 15, disp = 5, rate = 0.3),
  lower = c(maxr = 1, disp = 0.1, rate = 0.01),
  upper = c(maxr = 50, disp = 20, rate = 2)
)

tver_params <- coef(fit_tver)
summary(fit_tver)
```

**Fitted parameters:**
$\text{max} = `r sprintf("%.3f", tver_params["maxr"])`$,
$\text{disp} = `r sprintf("%.3f", tver_params["disp"])`$,
$\text{rate} = `r sprintf("%.3f", tver_params["rate"])`$

**Hendrickson parameters (TV = ER = 'tabula rasa'):**
$\text{max} = 6.506$, $\text{disp} = 5.000$, $\text{rate} = 0.350$

## Visualize the Fits

```{r fig.width=10, fig.height=5}
t_smooth <- seq(0, 20, by = 0.1)

# Fitted curves
br_curve <- data.table(
  t = t_smooth,
  y = modgompertz(t_smooth, br_params["maxr"],
                  br_params["disp"], br_params["rate"]),
  source = "Fitted BR"
)
tver_curve <- data.table(
  t = t_smooth,
  y = modgompertz(t_smooth, tver_params["maxr"],
                  tver_params["disp"], tver_params["rate"]),
  source = "Fitted TV+ER"
)

# Hendrickson curves
br_hend <- data.table(
  t = t_smooth,
  y = modgompertz(t_smooth, 10.986, 5, 0.42),
  source = "Hendrickson BR"
)
tver_hend <- data.table(
  t = t_smooth,
  y = modgompertz(t_smooth, 6.506, 5, 0.35),
  source = "Hendrickson TV+ER"
)

curves <- rbind(br_curve, tver_curve, br_hend, tver_hend)

# Observed data points
obs <- rbind(
  data.table(t = weeks, y = factor_data$BR,
             source = "Observed BR"),
  data.table(t = weeks, y = factor_data$TV_ER,
             source = "Observed TV+ER")
)

ggplot() +
  geom_line(data = curves, aes(x = t, y = y,
            color = source, linetype = source),
            linewidth = 0.8) +
  geom_point(data = obs, aes(x = t, y = y,
             color = source), size = 3) +
  scale_color_manual(values = c(
    "Fitted BR" = "blue",
    "Hendrickson BR" = "blue",
    "Fitted TV+ER" = "brown",
    "Hendrickson TV+ER" = "brown",
    "Observed BR" = "blue",
    "Observed TV+ER" = "brown")) +
  scale_linetype_manual(values = c(
    "Fitted BR" = "solid",
    "Hendrickson BR" = "dashed",
    "Fitted TV+ER" = "solid",
    "Hendrickson TV+ER" = "dashed")) +
  labs(x = "Week",
       y = "Improvement (CAPS points)",
       title = "Gompertz Fits: Observed Data vs Hendrickson Parameters",
       subtitle = paste("Solid = fitted to Raskind data;",
                        "Dashed = Hendrickson publication values")) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "right")
```

# Variance Estimation

The between-person SD of each factor is needed for the
covariance matrix. Hendrickson used SD = 8 for BR and
SD = 10 for TV+ER.

```{r}
# From Raskind Table 3:
# CAPS total score baseline SD (from 95% CI width):
# Prazosin: 77.3, 95% CI = 69.1-85.5 -> SE = 4.1, SD = 4.1*sqrt(32) = 23.2
# Placebo:  85.7, 95% CI = 78.0-93.3 -> SE = 3.9, SD = 3.9*sqrt(35) = 23.1

# Change from baseline SD:
# Difference in change: 11.3, 95% CI = 2.0-20.7
# SE of difference = (20.7-2.0)/(2*1.96) = 4.77
# Since SE = sqrt(SD_praz^2/n_praz + SD_plac^2/n_plac)
# and if SD_praz ~ SD_plac ~ SD:
# 4.77 = SD * sqrt(1/32 + 1/35) = SD * 0.247
# SD of change ~ 19.3

# The SD of 8 (BR) and 10 (TV+ER) in Hendrickson are
# lower than the raw CAPS SD because they represent the
# SD of each FACTOR, not the total outcome. The total
# outcome SD includes all three factors plus baseline
# variability.

total_var <- 18.5^2 + 10^2 + 10^2 + 8^2
```

Raw CAPS total score SD from the trial is approximately 23
points, which includes baseline variability plus all three
factors.

**Hendrickson factor SDs:**

| Factor              |  SD  |
|:--------------------|-----:|
| BR (drug-specific)  |    8 |
| TV (time-variant)   |   10 |
| ER (expectancy)     |   10 |
| Baseline (BL)       | 18.5 |

Sum of marginal variances:
`r sprintf("%.0f", total_var)`,
yielding an implied total SD of
`r sprintf("%.1f", sqrt(total_var))`.
The observed CAPS SD is approximately 23.
Cross-correlations reduce the total below the sum of
marginal variances.

# Parameter Comparison Table

```{r}
comparison <- data.table(
  Factor = c("BR", "TV+ER", "TV (Hendrickson)",
             "ER (Hendrickson)"),
  max_fitted = c(br_params["maxr"], tver_params["maxr"],
                 NA, NA),
  max_hendrickson = c(10.986, NA, 6.506, 6.506),
  disp_fitted = c(br_params["disp"], tver_params["disp"],
                   NA, NA),
  disp_hendrickson = c(5, NA, 5, 5),
  rate_fitted = c(br_params["rate"], tver_params["rate"],
                   NA, NA),
  rate_hendrickson = c(0.42, NA, 0.35, 0.35),
  sd_hendrickson = c(8, NA, 10, 10)
)

knitr::kable(comparison, digits = 3,
             caption = "Parameter comparison")
```

# Discussion

## Key Observations

1. **The BR parameters** (drug-specific response) should
   closely match Hendrickson's values (max=10.986,
   disp=5, rate=0.42) because they were derived from the
   same Raskind data.

2. **The TV+ER parameters** were set equal in Hendrickson
   ('tabula rasa' assumption) because the placebo trajectory
   cannot separate time-variant from expectancy-related
   improvement. The combined placebo trajectory has a single
   Gompertz fit, which Hendrickson assigned identically to
   both TV and ER.

3. **The displacement parameter** (disp=5) was the same for
   all factors. This controls the inflection point of the
   Gompertz curve and is the least well-constrained parameter
   given only 4 timepoints.

4. **The SD values** (8 for BR, 10 for TV/ER) represent
   between-person variability in each factor, not the total
   outcome variability. They were likely derived from the
   variance decomposition of the mixed-effects model used
   in the Raskind analysis.

## Limitations of This Reconstruction

- Only 4 timepoints are available (weeks 0, 7, 11, 15),
  providing limited constraint on a 3-parameter model.
- Weeks 7 and 11 values were interpolated from Figure 1D
  (approximate).
- The Gompertz fit is underdetermined with 4 points and
  3 parameters (only 1 degree of freedom).
- The published Hendrickson parameters may have used
  individual-level data (available to the authors but not
  to us) rather than group means.

## Implications for the Simulation

The Gompertz parameters control the mean trajectory of
each factor. They determine:

- **How quickly** each factor reaches its plateau (rate)
- **How large** the plateau is (max)
- **The shape** of the approach to plateau (disp)

The simulation's power estimates are sensitive to the
ratio of BR max to TV+ER max (signal-to-noise ratio) and
to the rates (faster BR means more drug effect is captured
within the assessment window). Figure 5 of Hendrickson
et al. (2020) explores this sensitivity systematically.

*Session info:* `r sessionInfo()$R.version$version.string`
