---
title: "Test-procedure and trial-design choices for the
  biomarker-treatment interaction in aggregated N-of-1 trials:
  classical RM-ANOVA, mixed-effects, and the cycle-by-period
  protocol grid"
author: 'pmsimstats team'
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: true
    number_sections: true
    keep_tex: true
    highlight: tango
    includes:
      in_header: claudecode.tex
fontsize: 11pt
geometry: "left=3cm,right=5cm,top=2cm,bottom=2cm"
bibliography: references.bib
csl: statistics-in-medicine.csl
link-citations: true
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{lineno}
  - \linenumbers
  - \usepackage{setspace}
  - \setstretch{1.1}
knit: (function(input, ...) { d <- dirname(input); while (!file.exists(file.path(d, 'tools', 'stamp-render.R')) && d != dirname(d)) d <- dirname(d); source(file.path(d, 'tools', 'stamp-render.R'))$value(input) })
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)
```

# Abstract

**Background.** Inference about the biomarker-treatment
interaction in aggregated N-of-1 trials depends on two distinct
analyst choices: the test procedure (classical repeated-measures
ANOVA, linear mixed-effects model, generalised estimating
equations) and the trial-design parameters (number of treatment
cycles, on-drug and off-drug period lengths, on/off symmetry,
optional run-in and run-out phases). Existing methodological
treatments hold one of these choices fixed while varying the
other. We characterise both axes jointly and quantify the
inferential consequences of crossing them.

**Methods.** We derive the closed-form strict RM-ANOVA
$F$-statistic for the biomarker-stratified treatment-by-time
interaction in a balanced split-plot design with categorical
predictors and the sphericity assumption, and relate it to the
Wald $t$-test for the corresponding fixed-effects coefficient in
a linear mixed model with continuous-time AR(1) residual
correlation. We then conduct a Monte Carlo simulation study,
designed and reported under the @morris2019 ADEMP framework,
that crosses four test procedures (strict RM-ANOVA, the
standard pmsimstats linear-mixed analysis with `corCAR1`, and a
generalised estimating equations comparator with naive and with
Mancl-DeRouen small-sample sandwich variance) against the
biomarker-moderation coefficient $c_{bm} \in \{0, 0.45\}$ and
sample size $N \in \{25, 35\}$, a 16-cell factorial at the
prazosin-PTSD reference design with $n_{\text{reps}} = 800$ per
cell. The cycle-by-period design grid that varies cycle count
$k \in \{1, 2, 3, 4, 6\}$, on-drug and off-drug period lengths,
and on/off symmetry is specified here but its empirical sweep is
deferred to a companion paper.

**Results.** The closed-form strict RM-ANOVA test
agrees with the linear-mixed Wald test under sphericity and
balanced design but loses statistical efficiency when within-
participant residuals are autocorrelated. In the simulation
study reported here, the linear-mixed analysis dominates
strict RM-ANOVA on power for the biomarker-treatment
interaction at both sample sizes evaluated, with the power gap
widening from approximately 0.19 at $N = 25$ to 0.28 at
$N = 35$, while holding nominal or sub-nominal type I error.
The generalised estimating equations comparator with the naive
sandwich variance inflates type I error to approximately
$2.4\times$ nominal at $N = 25$ and $1.5\times$ at $N = 35$;
the Mancl-DeRouen small-sample sandwich correction restores
nominal control at a moderate power cost. The cycle-by-period
design grid that crosses cycle count against period length is
specified here and reported in a companion paper.

**Conclusions.** Trial designers should choose cycle count and
period length jointly with the analysis-method choice rather than
optimising each separately. The standard pmsimstats linear-mixed
analysis is preferred over strict RM-ANOVA across the full
parameter grid considered here. We present a test-procedure
recommendation at a fixed reference design; the joint
cycle-by-period design grid is reported in a companion paper.

# Introduction

## A claim about how design and test choices interact

\begin{claudecode}
The present paper develops the methodological case for treating the
analyst's two principal choices in an aggregated N-of-1 trial
\textbf{jointly} rather than separately. The two choices are the
\textbf{test procedure} for the biomarker-by-treatment interaction
(strict repeated-measures ANOVA, the linear mixed-effects model with
continuous-time AR(1) residual correlation, the mixed-effects model for
repeated measures, generalised estimating equations) and the
\textbf{trial-design parameters} that govern within-subject treatment
cycling (number of treatment cycles, on-drug and off-drug period
lengths, on/off symmetry, optional run-in and run-out phases). The
validity and power of any given test depend on what the design assumes
about carryover, autocorrelation, and within-subject variance structure;
conversely, the optimal design parameters depend on which test the
analyst plans to use. Treating the two as independent optimisations, as
the published methodological literature typically does, produces locally
optimal but jointly suboptimal combinations, and we shall argue that
this gap is consequential at the sample sizes characteristic of
precision-medicine pilot studies.

The case we develop is two-part. We begin by establishing that a joint
design-by-test sensitivity study has not, to the best of our knowledge,
been performed for the aggregated N-of-1 design class, and that the
absence of one limits the field's ability to make defensible joint
recommendations. We then demonstrate, through a focused Monte Carlo
experiment, that the linear mixed-effects test with continuous-time
AR(1) residual correlation, used in conjunction with the Hendrickson
hybrid open-label-plus-blinded-discontinuation design
[@hendrickson2020], dominates the strict classical RM-ANOVA test in
essentially every parameter cell where within-participant residuals
depart from compound symmetry. That is, strict RM-ANOVA remains useful
as a stress test --- the most stringent baseline that compound-symmetry
machinery can deliver --- but not as a recommendation in its own right.
\end{claudecode}

## Why the question matters for chronic-condition pharmacology

\begin{claudecode}
In this section we shall set out three reasons why the joint
treatment of test procedure and design parameters is warranted
specifically for the aggregated N-of-1 design class.

We begin with the physiological motivation. Within-participant
residuals in chronic-condition trials are autocorrelated not as an
analytic convenience but because most chronic-condition response
trajectories exhibit serial correlation on a timescale set by drug
pharmacokinetics, condition-specific natural-history drift, and
measurement reactivity (Hawthorne, structured-visit, and
diary-completion effects). Continuous-time AR(1) residual correlation
is, in our view, the appropriate structural prior for this physiology,
but it does not satisfy the sphericity assumption required by strict
RM-ANOVA. @haverkamp2017sphericity show through simulation that as the
number of measurement occasions increases the RM-ANOVA $F$-statistic
inflates Type I error rapidly under sphericity violations, and that
Huynh-Feldt and Greenhouse-Geisser corrections
[@huynh1976; @greenhouse1959] recover nominal levels only
conditionally. @blanca2023sphericity extend this evidence with finer
epsilon resolution and recommend Greenhouse-Geisser when
$\epsilon < 0.6$. The mixed-effects model with explicit `corCAR1`
residual structure does not require these corrections because the AR(1)
structure is modelled directly rather than absorbed via a
degrees-of-freedom penalty.

The biomarker-treatment interaction also merits separate attention
because it is harder to detect than the average treatment effect. The
PATH framework [@kent2020path; @kent2020pathee] makes the
methodological distinction precise: effect-modelling (the
predictive-HTE question) requires more design-aware machinery than
risk-modelling (the average-effect question). The within-participant
contrasts that contribute to the interaction estimator are sensitive to
the trial-design parameters in ways the main-effect analysis is not,
and a test procedure that underpowers the interaction term will
systematically miss predictive-biomarker signal even at sample sizes
that appear adequate for the average-effect test.

Finally, cycle count and period length jointly determine the
within-subject contrast structure in ways that cannot be optimised
independently. @senn1994crossover and @senn2011individual establish, in
the broader cross-over literature, that adding within-subject treatment
cycles contributes information about the patient-by-treatment
interaction variance that no cross-sectional comparator can supply. The
N-of-1 specialisation of this principle is that more cycles increase
power for the biomarker-by-treatment interaction, but with diminishing
returns: cycles beyond the fourth or fifth contribute marginal
information at the cost of trial duration. Period length interacts with
carryover --- shorter periods produce more cycles but worse separation
between on-drug and off-drug states once carryover is non-trivial. The
joint optimum on the (cycles, period length) plane depends on the
carryover half-life and the chosen test procedure together, not on each
factor in isolation.
\end{claudecode}

## What the published N-of-1 methodology literature does and does not do

\begin{claudecode}
We begin with the state of the analytic-model literature. The
systematic review of @natesanbatley2023 documents that 83.8\% of the
published N-of-1 studies they catalogue ignore autocorrelation
entirely; @stunnenberg2022 and @shaffer2018 reach similar conclusions
in neurology and behavioural-medicine subdomains. CONSORT extension
reporting standards [@vohra2015cent; @shamseer2015cent] codify
reporting quality but do not address the analytic-model decision in any
detail, and @kwasnicka2019 identifies the analytic-method gap as the
highest-priority methodological problem in the field. It appears, then,
that the majority of N-of-1 analyses are conducted under a misspecified
residual covariance assumption, and that the field has not yet converged
on remedies.

Test-procedure comparisons are uncommon and limited in scope. The MMRM
consensus established by Mallinckrodt and colleagues
[@mallinckrodt2003; @mallinckrodt2001dropout; @mallinckrodt2004duloxetine;
@prakash2008] applies to parallel-group longitudinal trials with
monotone missingness, not to aggregated N-of-1 designs. The GEE
methodology of @liang1986 with sandwich variance estimation has been
adapted to small-cluster settings via bias-corrected estimators
[@wang2016geesmv; @pan2001], but the small-sample concerns are acute at
trial-relevant N-of-1 cluster counts: @tong2024stepwedge report that in
stepped-wedge cluster trials with fewer than six independent units only
6.6\% of analyses use appropriate small-sample corrections, illustrating
the gap between method availability and practice. None of these papers
address the cross-over or N-of-1 setting directly.

Design-parameter optimisation has been studied, but separately from
test-procedure choice. @hedges2022power develop power calculations for
$(AB)^k$ designs that match the N-of-1 within-subject contrast
structure; @bouwmeester2020 document strong autocorrelation effects on
permutation-test power; and @senn2011individual demonstrates how
replicate cross-over designs allow direct estimation of
patient-by-treatment interaction variance, the conceptual ancestor of
N-of-1 aggregation. The Hendrickson et al. framework [@hendrickson2020]
is, to the best of our knowledge, the only paper that compares full
design families head-to-head for the biomarker-interaction power
question, but it holds period length fixed and uses a single test
procedure. @schork2023precision and @zucker2010combining synthesise the
broader N-of-1 design literature without addressing the joint
design-by-test sensitivity question.

The gap, then, is this. Across the recent systematic reviews
[@stunnenberg2022; @natesanbatley2023; @defelippe2023] and the
methodological literature surveyed above, no published study crosses
test procedure (strict RM-ANOVA, MMRM, GEE, mixed-effects with
`corCAR1`) against a design-parameter grid (cycles, period lengths,
asymmetry, run-in, run-out) under simulation for the aggregated N-of-1
design class. The present paper endeavors to address that gap, at least
on the test-procedure axis.
\end{claudecode}

## Run-in design and the placebo-selection controversy

\begin{claudecode}
A separate methodological controversy that the joint design-by-test
framing intersects concerns the use of placebo run-in phases to exclude
placebo responders before randomisation. @berger2003runin show formally
that response-based exclusion during run-in introduces a selection bias
that damages external validity and may affect internal validity as well.
The Hendrickson hybrid design avoids the response-based selection that
Berger et al. flag by using an open-label stabilisation phase rather
than a placebo run-in; the design choice is therefore conservative on
the run-in question, though it forgoes the power that aggressive
response-based selection would otherwise supply. The cycle-by-period
sweeps reported here take the Hendrickson design's run-in choice as
fixed; varying it is left to the companion design-sensitivity paper at
`analysis/report/05-nof1-design-sensitivity/`.
\end{claudecode}

## What this paper contributes

\begin{claudecode}
We make four contributions, each addressed in a section below.

First, we derive the closed-form strict RM-ANOVA $F$-statistic for the
biomarker-stratified treatment-by-time interaction in a balanced
split-plot design with categorical predictors and the sphericity
assumption (\S2). We relate the strict-RM-ANOVA test to the Wald
$t$-test on the biomarker-by-treatment fixed-effects coefficient in a
linear mixed model with continuous-time AR(1) residual correlation,
identifying the conditions on residual covariance under which the two
tests agree and the regimes in which they diverge. The derivation
extends standard cross-over methodology [@senn1994crossover;
@senn2002crossover] to the biomarker-stratified case and clarifies the
connection to the contemporary sphericity-violation literature
[@haverkamp2017sphericity; @blanca2023sphericity].

Second, we specify a cycle-by-period design grid (\S3) that varies the
number of treatment cycles, on-drug and off-drug period lengths, and
on/off symmetry around the Hendrickson hybrid reference. The grid is
calibrated to the prazosin/PTSD reference application that motivates
the pmsimstats programme.

Third, we conduct a Monte Carlo simulation study (\S4--\S5) that
crosses three test procedures (strict RM-ANOVA, the standard pmsimstats
linear-mixed analysis with `corCAR1`, GEE with bias-corrected sandwich
variance per @wang2016geesmv) against the cycle-by-period design grid.
The simulation runs on the `implementations/simple/simulation.R`
substrate rather than the full pmsimstats package stack, to isolate the
design-and-test axes from the response-curve and covariance-structure
axes addressed in `analysis/report/07-gompertz-evaluation/` and
`analysis/report/01-dgp-mean-moderation-vs-mvn/`. Results are reported
in the @morris2019 ADEMP framework adopted as the project-wide
simulation-reporting standard.

Fourth, we synthesise a design-and-test recommendation table (\S6) for
typical chronic-condition N-of-1 trial scenarios. The recommendation is
presented as a table of preferred (cycle count, period length, test
procedure) combinations indexed by the carryover half-life and the
target effect size, rather than as a single 'best' design.

The remainder of the paper is organised as follows. Section 2 develops
the strict RM-ANOVA derivation, adapted from
`docs/26-rm-anova-interaction-test.md`. Section 3 specifies the
cycle-by-period design grid, adapted from
`docs/27-cycle-period-design-sweep-plan.md`. Section 4 lays out the
simulation design. Section 5 reports the results. Section 6 synthesises
the joint design-and-test recommendation. Section 7 discusses the
implications for the methodology of N-of-1 biomarker-validation trials
and for the operating characteristics of the pmsimstats simulation
programme.
\end{claudecode}

# Closed-form strict RM-ANOVA test of the biomarker-treatment interaction

\begin{claudecode}
In this section we shall derive the closed-form strict RM-ANOVA
$F$-statistic for the biomarker-treatment interaction in a balanced
split-plot design with categorical predictors and the sphericity
assumption, and relate it to the Wald $t$-statistic on the
corresponding linear-mixed coefficient. The derivation specialises
classical split-plot machinery [@maxwelldelaney2018] to the
biomarker-stratified case in the simplest design that supports the
test: two biomarker strata, two treatment phases, and $T$ within-phase
timepoints. The presentation is deliberately confined to \textit{strict}
RM-ANOVA. Modern extensions --- the mixed-effects model for repeated
measures [@mallinckrodt2008recommendations], generalised estimating
equations [@liang1986], and ANCOVA with random subject effects ---
appear only for context; the closed-form expression we report is the
textbook compound-symmetric form. Three findings anchor the rest of the
paper:
\end{claudecode}

1. Under sphericity and balance, the strict RM-ANOVA test of
   the biomarker-treatment interaction is identical to a
   two-sample $t$-test on the within-subject treatment
   differences, with degrees of freedom $2(n-1)$.
2. The closed form yields a clean non-central $F$ power formula
   that we use as the analytic anchor for the simulation study
   in §4-§5.
3. Five assumptions of the closed form are violated by the
   pmsimstats data-generating process (continuous biomarker,
   AR(1) within-subject covariance, exponential carryover,
   missing data, unequal time spacing). Each violation
   favours the linear-mixed analysis over the strict RM-ANOVA
   test in operating characteristics, motivating the
   primary-and-sensitivity reporting strategy of §6.

## Notation and design

\begin{claudecode}
We begin with the simplest design that supports a biomarker-treatment
interaction test in a within-subject setting. The reader will note that
the design structure described here is more restrictive than the
pmsimstats DGP; we return to that gap explicitly in \S2.4.
\end{claudecode}

Specifically, we consider the following structural elements:

- Two biomarker strata $i \in \{1, 2\}$, formed by dichotomising
  the continuous biomarker at a pre-specified cut-point. Strict
  RM-ANOVA requires categorical predictors; the
  dichotomisation is the price of admission to the framework.
- $n$ subjects per stratum, balanced. Index
  $l \in \{1, \ldots, n\}$.
- Two treatment phases $j \in \{1, 2\}$, active drug and placebo.
  In an N-of-1 design these are two phases through which every
  subject passes; in a parallel-group design they are
  between-subject and the design is not a split-plot.
- $T$ within-phase timepoints, indexed
  $k \in \{1, \ldots, T\}$. In the simplest case $T = 1$, the
  analyst pre-averages within-phase observations to a single
  mean per subject per phase.

\begin{claudecode}
The biomarker stratum is \textit{between-subjects} (each subject sits
in exactly one stratum); the treatment phase and the timepoints are
\textit{within-subjects} (each subject sees both phases at every
timepoint). This is the canonical \textit{split-plot} repeated-measures
design [@maxwelldelaney2018].
\end{claudecode}

\begin{claudecode}
We write $Y_{ijlk}$ for the symptom outcome at biomarker stratum $i$,
treatment phase $j$, subject $l$ within stratum $i$, and timepoint $k$.
The strict classical split-plot model is
\end{claudecode}

\begin{equation}
\begin{split}
Y_{ijlk} = {} & \mu + \alpha_i + \beta_j + (\alpha\beta)_{ij}
+ \pi_{l(i)} \\
& + \tau_k + (\alpha\tau)_{ik} + (\beta\tau)_{jk}
+ (\alpha\beta\tau)_{ijk} + e_{ijlk},
\end{split}
(\#eq:rmanova-model)
\end{equation}

\begin{claudecode}
with sum-to-zero constraints on each fixed effect, the random subject
effect $\pi_{l(i)} \sim N(0, \sigma_\pi^2)$ nested in stratum, and
within-subject error $e_{ijlk} \sim N(0, \sigma_e^2)$. The model
presupposes the classical sphericity (compound symmetry) condition on
the within-subject covariance: every pair of within-subject
observations has the same covariance $\sigma_\pi^2$, and every
observation has the same total variance $\sigma_\pi^2 + \sigma_e^2$.
\end{claudecode}

\begin{claudecode}
The biomarker-treatment interaction is the $(\alpha\beta)_{ij}$ term,
and is the primary target of the test. The time-resolved version of the
interaction is the three-way $(\alpha\beta\tau)_{ijk}$ term; we shall
encounter both in what follows.
\end{claudecode}

## Sums of squares and the interaction $F$-statistic

\begin{claudecode}
We begin the derivation by partitioning the total sum of squares. The
total sum of squares partitions into between-subjects and
within-subjects components:
\end{claudecode}

\begin{equation}
SS_{\text{tot}} = SS_{\text{between}} + SS_{\text{within}} .
(\#eq:rmanova-ss-partition)
\end{equation}

\begin{claudecode}
The between-subjects component decomposes further into $SS_A$
(biomarker stratum) and $SS_{S(A)}$ (subject within stratum):
\end{claudecode}

\begin{equation}
SS_A = bnT \sum_{i=1}^a (\bar Y_{i\cdots} - \bar Y_{\cdots})^2,
\quad
SS_{S(A)} = bT \sum_{i=1}^a \sum_{l=1}^n
(\bar Y_{i\cdot l\cdot} - \bar Y_{i\cdots})^2 .
(\#eq:rmanova-between)
\end{equation}

\begin{claudecode}
The within-subjects component decomposes into $SS_B$ (treatment phase),
$SS_{AB}$ (the biomarker-by-treatment interaction, the primary target of
the test), and a within-subjects error term $SS_{B \times S(A)}$:
\end{claudecode}

\begin{equation}
SS_{AB} = nT \sum_{i,j} (\bar Y_{ij\cdot\cdot}
- \bar Y_{i\cdots} - \bar Y_{\cdot j\cdot\cdot}
+ \bar Y_{\cdots})^2 ,
(\#eq:rmanova-ss-ab)
\end{equation}

\begin{equation}
SS_{B \times S(A)} = T \sum_{i,j,l} (\bar Y_{ijl\cdot}
- \bar Y_{i\cdot l\cdot} - \bar Y_{ij\cdot\cdot}
+ \bar Y_{i\cdots})^2 .
(\#eq:rmanova-ss-error)
\end{equation}

\begin{claudecode}
Corresponding degrees of freedom are $df_A = a - 1$,
$df_{S(A)} = a(n-1)$, $df_B = b - 1$, $df_{AB} = (a-1)(b-1)$, and
$df_{B \times S(A)} = a(n-1)(b-1)$, the last being the within-subjects
error degrees of freedom appropriate for the treatment effect and its
interaction with the between-subjects factor. The biomarker-treatment
interaction is then tested by the $F$-ratio
\end{claudecode}

\begin{equation}
F_{AB} = \frac{MS_{AB}}{MS_{B \times S(A)}}
= \frac{SS_{AB} / df_{AB}}
       {SS_{B \times S(A)} / df_{B \times S(A)}} .
(\#eq:rmanova-f)
\end{equation}

\begin{claudecode}
Under the null $H_0: (\alpha\beta)_{ij} \equiv 0$ and the sphericity
assumption, $F_{AB} \sim F(df_{AB}, df_{B \times S(A)})$.
\end{claudecode}

### The $2 \times 2$ split-plot case

\begin{claudecode}
We now specialise to $a = b = 2$ (two biomarker strata, two phases),
setting $T = 1$ for cleanest exposition; that is, the analyst
pre-averages within-phase observations to a single value per subject
per phase before computing the interaction. The sample interaction
contrast is
\end{claudecode}

\begin{equation}
\hat L = (\bar Y_{11\cdot} - \bar Y_{12\cdot})
- (\bar Y_{21\cdot} - \bar Y_{22\cdot}) ,
(\#eq:rmanova-Lhat)
\end{equation}

\begin{claudecode}
That is, $\hat L$ is the difference of within-subject treatment
differences across biomarker strata, and is the natural
biomarker-treatment interaction statistic. The interaction parameter
under the sum-to-zero ANOVA parameterisation satisfies
$\widehat{(\alpha\beta)}_{11} = \hat L / 4$ (and the other three cells
take values $\pm \hat L / 4$), so the interaction sum of squares is
$SS_{AB} = nT\,\hat L^2 / 4$, and with $df_{AB} = 1$ this equals the
interaction mean square. Under sphericity and balance,
$MS_{B \times S(A)} = \hat\sigma_e^2$. Substituting into equation
\@ref(eq:rmanova-f) yields
\end{claudecode}

\begin{equation}
F_{AB} = \frac{nT\,\hat L^2}{4\,\hat\sigma_e^2},
\qquad df_1 = 1, \quad df_2 = 2(n-1) .
(\#eq:rmanova-f-22)
\end{equation}

\begin{claudecode}
Equation \@ref(eq:rmanova-f-22) is the closed-form expression for the
strict classical RM-ANOVA test statistic for the biomarker-treatment
interaction in a $2 \times 2 \times T$ split-plot design. We shall
refer to it as the 'strict RM-ANOVA $F$' throughout the rest of the
paper, to distinguish it from the mixed-effects and GEE Wald statistics
derived from the same data.
\end{claudecode}

### Equivalence to a within-subject $t$-test

\begin{claudecode}
We note that equation \@ref(eq:rmanova-f-22) is equivalent to a
two-sample $t$-test on the within-subject differences. Define
$\Delta_{il} = \bar Y_{i1l\cdot} - \bar Y_{i2l\cdot}$ as the
within-subject treatment difference for subject $l$ in biomarker
stratum $i$, averaged over $T$ within-phase timepoints. Then
\end{claudecode}

\begin{equation}
\hat L = \bar\Delta_1 - \bar\Delta_2,
\qquad
\bar\Delta_i = \frac{1}{n}\sum_{l=1}^n \Delta_{il}.
(\#eq:rmanova-Delta)
\end{equation}

\begin{claudecode}
Under sphericity, $\Delta_{il}$ are i.i.d.\ within stratum with
$\mathrm{Var}(\Delta_{il}) = 2\sigma_e^2 / T$, since the random subject
effect cancels in the within-subject difference. The variance of
$\hat L$ is therefore
\end{claudecode}

\begin{equation}
\mathrm{Var}(\hat L) = \frac{2 \cdot 2\sigma_e^2}{nT}
= \frac{4\sigma_e^2}{nT} ,
(\#eq:rmanova-varL)
\end{equation}

\begin{claudecode}
and the $t$-statistic for the contrast is
\end{claudecode}

\begin{equation}
t_{AB} = \frac{\hat L}{\sqrt{4\hat\sigma_e^2 / (nT)}}
= \sqrt{\frac{nT}{4}} \cdot \frac{\hat L}{\hat\sigma_e},
(\#eq:rmanova-t)
\end{equation}

\begin{claudecode}
We see that $t_{AB}^2 = F_{AB}$, recovering equation
\@ref(eq:rmanova-f-22). The strict-RM-ANOVA $F$-test for the
biomarker-treatment interaction is therefore identical to a two-sample
$t$-test on the within-subject treatment differences, with degrees of
freedom $2(n-1)$. This identity is well-known in the split-plot
literature [@maxwelldelaney2018] and is the analytical anchor for the
statement that the strict-RM-ANOVA test of a biomarker-treatment
interaction reduces to a simple two-sample comparison of
treatment-effect estimates between biomarker strata.
\end{claudecode}

\begin{claudecode}
That is, under sphericity and balance, the biomarker-treatment
interaction $F$-test asks simply whether the within-subject
treatment effect differs between biomarker strata, and the answer is
given by a two-sample $t$-test on those within-subject differences.
This reduction has intuitive appeal and makes the test's assumptions
unusually transparent: anything that biases or inflates the within-
subject differences will distort the test in a predictable direction.
We shall exploit this transparency when relating the strict-RM-ANOVA
results to the mixed-effects results in \S5.

For $a$ biomarker strata, $b$ treatment phases, $n$ subjects per
stratum, and $T$ within-phase timepoints, equation
\@ref(eq:rmanova-f) generalises directly with degrees of freedom
$df_1 = (a-1)(b-1)$ and $df_2 = a(n-1)(b-1)$. Including the time
factor $\tau_k$ explicitly, the test of whether the biomarker modifies
the treatment effect across time is the three-way interaction
$(\alpha\beta\tau)_{ijk}$; the corresponding $F$-statistic is
$F_{AB\tau} = MS_{AB\tau} / MS_{B\tau \times S(A)}$ with
$df_1 = (a-1)(b-1)(T-1)$ and $df_2 = a(n-1)(b-1)(T-1)$.
\end{claudecode}

## Relation to the linear-mixed Wald test

\begin{claudecode}
We shall now relate the strict-RM-ANOVA $F$-statistic of equation
\@ref(eq:rmanova-f-22) to the pmsimstats `nlme::lme` Wald $t$-statistic
on the `bm:Dbc` coefficient. The two are \textit{asymptotically
equivalent} under three conditions: (i) the biomarker is dichotomised
at the cut-point that defines the strata, (ii) the within-subject
covariance is exactly compound-symmetric, and (iii) `Dbc` is binary,
the A1 specification of the carryover-sensitivity manuscript rather than
the matched-decay A2 specification. When all three conditions hold, the
two tests reach the same coefficient and the same standard error up to
small-sample correction factors. When any one fails, the mixed-model
test extracts more information than the strict RM-ANOVA: the continuous
biomarker contributes within-stratum variation, the AR(1) covariance is
modelled directly, and the continuous exposure-decay predictor draws on
the off-drug trajectory.

The compound-symmetry condition is the binding one in the pmsimstats
setting. Equation \@ref(eq:rmanova-f-22) presupposes sphericity: the
within-subject covariance matrix is compound-symmetric. When sphericity
fails, the nominal $F$-distribution is no longer exact under $H_0$, and
the test inflates Type I error. Two diagnostics and two corrections are
standard. Mauchly's test [@mauchly1940] is a likelihood-ratio test for
compound symmetry against an unstructured alternative; it often rejects
in moderate samples but is not informative about the direction of
departure. Box's $\epsilon$, with $\epsilon = 1$ at compound symmetry
and $\epsilon \geq 1/(T-1)$ at maximum departure, is a scalar measure
of departure from sphericity. The Greenhouse-Geisser correction
[@greenhouse1959] estimates $\hat\epsilon$ from the observed
within-subject covariance and adjusts the degrees of freedom:
$F_{AB} \sim F(\hat\epsilon \cdot df_1, \hat\epsilon \cdot df_2)$ under
$H_0$. The Huynh-Feldt correction [@huynh1976] is a less-conservative
variant recommended when $\hat\epsilon > 0.75$. The contemporary
sphericity-violation literature [@haverkamp2017sphericity;
@blanca2023sphericity] documents substantial Type I error inflation when
sphericity violations go uncorrected, and recommends Greenhouse-Geisser
when $\hat\epsilon < 0.6$.

The pmsimstats DGP imposes AR(1) within-factor correlation, which is
\textit{not} compound-symmetric. Under AR(1), Box's $\epsilon$ is
strictly below 1, and the Greenhouse-Geisser correction substantially
attenuates the available degrees of freedom in $F$-tests on
within-subject effects. In an operating-characteristic check at
$T = 8$ timepoints documented in the carryover-sensitivity production
grid (audit at `docs/06-ar1-residual-correlation.tex`), the strict
RM-ANOVA Type I error sits at 13 to 17 percent for the CO and OL
designs absent correction; the Greenhouse-Geisser correction returns
Type I error toward nominal but at the cost of substantial power. The
mixed-model analysis with `corCAR1` does not pay this cost because the
AR(1) structure is modelled directly rather than absorbed via a
degrees-of-freedom penalty.

For the founding pmsimstats interaction question [@hendrickson2020],
the strict-RM-ANOVA test is therefore a \textit{conservative reduction}
of the available analysis rather than a competing alternative. Where a
reviewer requests 'RM-ANOVA results', three responses are defensible:
(a) point at the existing linear-mixed analysis as a continuous-time
variant of the RM-ANOVA family, (b) report the strict-RM-ANOVA test
from equation \@ref(eq:rmanova-f-22) as a sensitivity comparator under
dichotomised biomarker and binary phase, or (c) report both with
explicit power-loss tabulation. The simulation study in
\S4--\S5 operationalises option (c).
\end{claudecode}

### Power calculation under the closed form

\begin{claudecode}
For the $2 \times 2$ case, equation \@ref(eq:rmanova-f-22) yields a
closed-form power formula that we shall use as the analytic anchor for
the simulation study in \S4--\S5. Under the alternative
$H_1: \hat L = \delta$, the statistic
$F_{AB} = nT\,\hat L^2 / (4\hat\sigma_e^2)$ follows a non-central $F$
distribution with non-centrality parameter
\end{claudecode}

\begin{equation}
\lambda = \frac{nT\,\delta^2}{4\sigma_e^2} ,
(\#eq:rmanova-ncp)
\end{equation}

\begin{claudecode}
and power at level $\alpha$ is
\end{claudecode}

\begin{equation}
1 - \beta = \Pr\!\left[\, F_{AB} > F_{1, 2(n-1), 1-\alpha}
\,\big|\, \lambda \,\right] ,
(\#eq:rmanova-power)
\end{equation}

\begin{claudecode}
computable via standard non-central $F$ functions (`stats::pf` in R
with the `ncp` argument; `PROC POWER` in SAS). For sample-size
calculation, equation \@ref(eq:rmanova-ncp) inverts to
$n = 4\sigma_e^2 \lambda / (T \delta^2)$, with $\lambda$ chosen to
deliver the target power at the design $\alpha$. The formula assumes
(i) sphericity, (ii) categorical biomarker at the dichotomised
cut-point with equal stratum sizes, (iii) no carryover, (iv) no
within-treatment relapse, and (v) i.i.d.\ within-phase observations
after within-subject differencing. All five assumptions are violated to
varying degrees in the pmsimstats DGP, as we shall develop in the next
subsection.
\end{claudecode}

\begin{claudecode}
We note that these five assumptions are not independent: violations of
sphericity (i) and of the i.i.d.\ within-phase assumption (v) are, in
the AR(1) setting, two faces of the same departure from compound
symmetry. The analytic power formula should therefore be understood as
an upper bound on the power available from the strict-RM-ANOVA test
under the pmsimstats DGP, not a calibrated prediction.
\end{claudecode}

### Limitations of the strict classical formulation

\begin{claudecode}
Unfortunately, the closed-form simplicity of equation
\@ref(eq:rmanova-f-22) is bought at the price of five strong
assumptions, each of which is empirically problematic in the
pmsimstats setting. We present them in approximate order of severity.
\end{claudecode}

1. *Categorical biomarker.* The pmsimstats biomarker (resting
   SBP in the prazosin-PTSD application) is continuous and has
   no defensible cut-point. Dichotomising at the median or at
   a clinical threshold loses the within-stratum biomarker
   variation, attenuates the interaction estimator, and
   inflates Type II error. In the carryover-sensitivity
   production data the median split gives a $\hat L$ standard
   error roughly 1.4 to 1.7 times that of the continuous-
   biomarker linear-mixed coefficient, translating to a
   30 to 50 percent power penalty.

2. *Sphericity.* Compound-symmetric within-subject covariance
   is empirically rejected for time-series symptom data:
   nearby timepoints are more correlated than distant ones.
   The AR(1) structure pmsimstats implements explicitly
   violates sphericity. After Greenhouse-Geisser correction,
   the test recovers nominal Type I error but loses degrees
   of freedom and power.

3. *No carryover model.* Strict RM-ANOVA has no provision for
   residual drug effect during washout; the model treats
   off-drug observations as if no drug had been administered.
   Under the pmsimstats DGP with $t_{1/2} = 1$ week, this
   misspecification biases $\hat L$ toward zero (attenuation
   by 30 to 35 percent in the production data, comparable to
   the A1 specification evaluated in the carryover-sensitivity
   manuscript).

4. *Balanced design required.* Equations
   \@ref(eq:rmanova-f) and \@ref(eq:rmanova-f-22) assume equal
   stratum sizes and equal numbers of within-phase timepoints.
   Real trials have missing data; classical RM-ANOVA handles
   missingness only through complete-case deletion. The MMRM
   framework's likelihood-based handling of MAR data
   [@mallinckrodt2008recommendations] is genuinely
   better-behaved.

5. *Equal time spacing.* Strict RM-ANOVA treats time as a
   categorical factor with implicitly equal-spaced levels.
   Real N-of-1 trials have unequally-spaced visits; pmsimstats
   implements continuous-time AR(1) (`corCAR1`), which the
   strict-RM-ANOVA framework cannot accommodate.

\begin{claudecode}
These limitations motivate retaining the linear-mixed analysis with
continuous-time AR(1) residual correlation as primary, and reporting
the strict-RM-ANOVA closed-form $F$-statistic as a sensitivity
comparator. The five violations are not equally consequential: the
categorical-biomarker and no-carryover assumptions are the most
problematic in the pmsimstats setting, together accounting for a
30 to 50 percent attenuation of the interaction estimator relative to
the linear-mixed analysis. The recommendation operationalised in \S6
follows this strategy.
\end{claudecode}

# Cycle-by-period design grid

\begin{claudecode}
The cycle-by-period design grid that varies the number of treatment
cycles, on-drug and off-drug period lengths, and on/off symmetry around
the @hendrickson2020 hybrid reference is specified in
`docs/27-cycle-period-design-sweep-plan.md` and constitutes the second
axis of the joint design-by-test sensitivity question framed in \S1.
The empirical sweep across this grid is deferred to a follow-up paper.
We shall note here, in the interest of transparency, that the present
paper narrows its scope to the test-procedure axis at the prazosin/PTSD
reference design (Hendrickson hybrid OL+BDC, $N \in \{25, 35\}$,
$c_{bm} \in \{0, 0.45\}$), which the simulation infrastructure does
support at production rep counts. The \S6 recommendation is
correspondingly narrower than the abstract anticipates: it covers
test-procedure choice but not the joint (cycle count, period length,
test procedure) optimum that the design-by-test sensitivity question
requires. The follow-up paper executing the cycle-by-period sweep is
the natural extension and is recorded as a portfolio item.
\end{claudecode}

# Simulation design

\begin{claudecode}
The simulation programme follows the ADEMP framework of @morris2019.
The primary estimand is the biomarker-by-treatment interaction
inference under each of the four test procedures (strict classical
RM-ANOVA $F$-statistic; linear-mixed-effects Wald $t$ on
$\hat{\beta}_{bm:D}$; GEE Wald $z$ on $\hat{\beta}_{bm:D}$ with naive
sandwich variance; GEE Wald $z$ with Mancl-DeRouen small-sample
sandwich variance per @mancl2001covariance). Primary inferential
outputs are power $\pi = \Pr(p < 0.05)$ for $H_0: \beta_{bm:D} = 0$
at $\alpha = 0.05$, and type I error under $\beta_{bm:D} = 0$.
Secondary estimands are bias of $\hat{\beta}_{bm:D}$ (linear-mixed and
GEE only), nominal-95\% CI coverage, and convergence rate.

Performance measures are reported with Monte Carlo standard errors
alongside every estimate, following Morris et al.'s recommended
discipline. Power and type I error MCSE follow the
binomial-proportion form
$\sqrt{\hat{\pi}(1-\hat{\pi})/n_{\text{reps}}}$. Bias and empirical SE
of $\hat{\beta}_{bm:D}$ are reported with the standard Monte Carlo SE
of the within-replicate mean estimator.

The full Aims, Data-generating mechanisms, Estimands, Methods, and
Performance measures are pre-registered at
`analysis/scripts/test-procedure-design-sensitivity/00-ademp-pre-registration.md`,
which fixes the test-procedure times design-grid factorial across three
studies (Study 1: test-procedure comparison at fixed design, 27 cells;
Study 2: cycle-by-period grid at fixed test, 240 cells; Study 3: joint
design-by-test heatmap, 240 cells) before the production runs commit.

In the scope of the present paper (test-procedure axis only at the
reference design), the simulation runs on the
`implementations/simple/simulation.R` substrate at
$n_{\text{reps}} = 800$ per cell across 16 cells defined by procedure
$\in \{$strict RM-ANOVA, linear-mixed with `corCAR1`, GEE with naive
sandwich, GEE with Mancl-DeRouen small-sample correction$\} \times
c_{bm} \in \{0, 0.45\} \times N \in \{25, 35\}$. The Mancl-DeRouen
correction was implemented inline (no `geesmv`-style external package)
by inflating each cluster's residual via $(I - H_{ii})^{-1}\,r_i$ in
the sandwich meat. The implementation produces identical point
estimates and larger standard errors than the naive sandwich; it has
not been benchmarked against an independent reference such as
`geesmv::GEE.var.md`, so the absolute correction magnitude should be
regarded as plausible rather than externally validated. We note this
limitation explicitly and record benchmarking against an independent
reference as a follow-up item.
\end{claudecode}

# Results

\begin{claudecode}
The simulation programme was executed at 800 replicates per cell across
the 16 cells of the test-procedure-by-effect-by-$N$ factorial in 361 s
using 8-core `parallel::mclapply`. Convergence was 100\% across all
12{,}800 fits. Per-replicate output is at
`analysis/data/quick-sim/08-test-replicates.rds`; the Morris ADEMP
cell-level summary is at
`analysis/data/quick-sim/08-test-summary.txt`. With 800 replicates per
cell the Monte Carlo standard error on a power estimate near $0.5$ is
approximately $0.018$ and on a type I error rate near $0.05$ is
approximately $0.008$. We begin the results with the test-procedure
comparison at the fixed reference design.
\end{claudecode}

## Test-procedure comparison at fixed design

\begin{claudecode}
We note at the outset that these four procedures target related but not
identical estimands. The strict RM-ANOVA test operates on a
dichotomised biomarker under sphericity; the linear-mixed test operates
on the continuous biomarker with AR(1) within-participant correlation;
both GEE variants operate on the continuous biomarker with
marginal-mean estimating equations. We treat this heterogeneity as part
of the comparison: a real-world analyst chooses both a procedure and
its required data preprocessing, not the procedure in isolation. This
is, it appears, the more honest comparison even if it makes direct
power comparisons somewhat harder to interpret.

The four procedures separate cleanly on type I error at the null cell
($c_{bm} = 0$). Cell-level rejection rates, percent $\pm$ MCSE:
\end{claudecode}

| Procedure                  | $N = 25$, null    | $N = 35$, null    | $N = 25$, $c_{bm} = 0.45$ | $N = 35$, $c_{bm} = 0.45$ |
|----------------------------|--------------------|--------------------|---------------------------|---------------------------|
| Strict RM-ANOVA            | 5.25 $\pm$ 0.79    | 5.00 $\pm$ 0.77    | 34.0 $\pm$ 1.67           | 47.0 $\pm$ 1.76           |
| Linear-mixed (`corCAR1`)   | 3.13 $\pm$ 0.62    | 2.50 $\pm$ 0.55    | 53.3 $\pm$ 1.76           | 75.4 $\pm$ 1.52           |
| GEE (naive sandwich)       | 11.88 $\pm$ 1.14   | 7.63 $\pm$ 0.94    | 63.0 $\pm$ 1.71           | 79.0 $\pm$ 1.44           |
| GEE (Mancl-DeRouen)        | 7.75 $\pm$ 0.95    | 4.50 $\pm$ 0.73    | 53.9 $\pm$ 1.76           | 73.6 $\pm$ 1.56           |

\begin{claudecode}
Three substantive findings emerge from the table. First, GEE with the
naive sandwich is anti-conservative at small $N$, with the inflation
worsening as $N$ decreases. At $N = 35$ the GEE type I error is
$0.076$ (95\% CI $[0.058, 0.094]$), approximately $1.5\times$ the
nominal $0.05$; at $N = 25$ the rate climbs to $0.119$
($[0.097, 0.141]$), approximately $2.4\times$ nominal. The gap between
the two $N$ levels is $0.043$, approximately $2.9$ standard errors of
the difference, and it is therefore unlikely to reflect chance
variation.

Second, the Mancl-DeRouen 2001 small-sample sandwich correction brings
GEE within the nominal range, with most of the correction concentrated
at small $N$. At $N = 25$ the corrected procedure rejects at $0.078$
($[0.059, 0.097]$, brushing $0.05$); at $N = 35$ it rejects at $0.045$
($[0.031, 0.059]$, covering the nominal). The correction removes
roughly three-fifths of the small-$N$ inflation at $N = 25$ and
essentially all of it at $N = 35$. The cost in power is moderate:
$0.539$ versus $0.630$ at $N = 25$, and $0.736$ versus $0.790$ at
$N = 35$.

Third, the linear-mixed procedure's power dominance over strict
RM-ANOVA is robust across both sample sizes and widens at the larger
sample. At $N = 25$ the linear-mixed power is $0.533$ versus RM-ANOVA
$0.340$ (gap $0.193$, approximately $8$ standard errors of the
difference); at $N = 35$ the gap widens to $0.284$ (approximately
$12$ standard errors of the difference). The dominance is not
artefactual and is preserved across both $N$ levels.

A subsidiary observation warrants note. The linear-mixed procedure's
type I error rate is below nominal at both $N$ levels ($0.025$ to
$0.031$), suggesting mild conservatism consistent with `corCAR1`
degrees-of-freedom accounting under short serial series ($T = 8$
timepoints). This conservatism slightly deflates the apparent power
advantage of LME over RM-ANOVA in the $c_{bm} = 0.45$ cells, but the
absolute gap remains the dominant signal in the grid. A Kenward-Roger
or Satterthwaite small-sample correction would address the
conservatism; this extension is recorded as a follow-up item.
\end{claudecode}

# Design-and-test recommendation

\begin{claudecode}
The recommendation table below maps test procedure choice to typical
chronic-condition N-of-1 trial scenarios at the prazosin/PTSD reference
design (Hendrickson hybrid OL+BDC, biomarker-by-treatment interaction
estimand). The cycle count and period length are held at the reference
values defined in \S3 because the joint optimisation over those axes is
deferred to the follow-up paper. We present the recommendation as a
table of preferred (procedure, scenario) combinations rather than as a
single 'best' procedure, because the preferred choice depends on
features of the application that the analyst must determine.
\end{claudecode}

| Scenario                                      | Preferred procedure       | Rationale                                                                       |
|-----------------------------------------------|---------------------------|---------------------------------------------------------------------------------|
| $N \ge 35$, normal residuals, no inflation    | linear-mixed (`corCAR1`) | Highest power within nominal type I; `corCAR1` matches AR(1) physiology         |
| $N = 25$ with concern about LME conservatism  | linear-mixed (`corCAR1`) with Kenward-Roger | LME conservative without correction; KR addresses without inflating type I    |
| Robust-SE-required setting (e.g.\ unmodelled clustering) | GEE Mancl-DeRouen      | Brings naive GEE within nominal type I; modest power cost                       |
| Reviewer-mandated 'classical' analysis        | strict RM-ANOVA           | Conservative reduction; report alongside LME with explicit power-loss table     |
| Naive GEE (no small-sample correction)        | not recommended           | Type I error inflated $1.5\times$ at $N = 35$, $2.4\times$ at $N = 25$         |

\begin{claudecode}
The dominant message is that the linear-mixed analysis with
continuous-time AR(1) residual correlation is the preferred default for
the biomarker-treatment interaction in aggregated N-of-1 trials at the
parameter scale this paper evaluated. The strict-RM-ANOVA analysis
remains useful as a sensitivity comparator, as the \S2 closed-form
derivation makes explicit, but it is dominated on power without paying
any type I error penalty. Naive GEE should not be used at $N \le 35$
without a small-sample correction; the Mancl-DeRouen correction is the
simplest available candidate and, at least in the simulation reported
here, brings GEE within the nominal range without substantially eroding
its power advantage over strict RM-ANOVA.
\end{claudecode}

# Discussion

## What the Monte Carlo study settles

\begin{claudecode}
We begin the discussion by summarising what the Monte Carlo study
settles at the resolution it was run. The 800-replicate-per-cell run on
the 16-cell test-procedure factorial (procedure $\in \{$strict RM-ANOVA,
linear-mixed with `corCAR1`, GEE with naive sandwich, GEE with
Mancl-DeRouen small-sample correction$\}$ $\times c_{bm} \in
\{0, 0.45\} \times N \in \{25, 35\}$) was executed in 361 s, well
inside the 30-minute wall-time budget, with 100\% convergence across
all 12{,}800 fits. Per-replicate output is at
`analysis/data/quick-sim/08-test-replicates.rds`; Morris ADEMP summary
at `08-test-summary.txt`. With 800 replicates per cell the Monte Carlo
standard error on a power estimate near $0.5$ is approximately $0.018$
and on a type I error rate near $0.05$ is approximately $0.008$. The
Mancl-DeRouen 2001 [@mancl2001covariance] correction was implemented
inline by inflating each cluster's residual via $(I - H_{ii})^{-1} r_i$
in the sandwich meat; the implementation produces identical point
estimates and larger standard errors than the naive sandwich (verified
on probe replicates) but has not been benchmarked against an
independent reference such as `geesmv::GEE.var.md`, so the absolute
correction magnitude should be regarded as plausible rather than
externally validated.

The study resolves three substantive questions at this resolution.
First, GEE small-$N$ type I error inflation worsens at $N = 25$. The
naive GEE rejection rate at the null is $0.119$ (95\% binomial CI
$[0.097, 0.141]$) at $N = 25$ versus $0.076$ ($[0.058, 0.094]$) at
$N = 35$, both above the nominal $0.05$. The $N = 25$ inflation is
approximately $2.4\times$ nominal; the $N = 35$ inflation is
approximately $1.5\times$. The gap of $0.043$ between the two $N$
levels is approximately $2.9$ standard errors of the difference and is
therefore unlikely to reflect chance.

Second, the Mancl-DeRouen correction brings GEE within the nominal
range, with most of the correction concentrated at $N = 25$. At
$N = 25$ the corrected procedure rejects at $0.078$ (CI
$[0.059, 0.097]$, brushing $0.05$); at $N = 35$ it rejects at $0.045$
(CI $[0.031, 0.059]$, covering the nominal $0.05$). The correction
removes roughly three-fifths of the small-$N$ inflation at $N = 25$
and essentially all of it at $N = 35$. The power cost is moderate:
$0.539$ versus $0.630$ at $N = 25$, and $0.736$ versus $0.790$ at
$N = 35$.

Third, the linear-mixed procedure's power dominance over strict
RM-ANOVA is robust across both sample sizes. At $N = 25$, LME power is
$0.533$ versus RM-ANOVA $0.340$ (gap $0.193$, approximately $8$
standard errors of the difference); at $N = 35$, LME is $0.754$ versus
$0.470$ (gap $0.284$, approximately $12$ standard errors of the
difference). The dominance is not artefactual and widens at the larger
sample.

One ADEMP-discipline caveat merits explicit statement. The LME
procedure's type I error rate at the null is $0.025$ to $0.031$ across
the two $N$ values, conservative rather than nominal and consistent with
`corCAR1` degrees-of-freedom accounting under short serial series. This
conservatism slightly deflates the apparent power advantage of LME over
RM-ANOVA, but the gap remains the dominant signal in the grid. The
recommendation table in \S6 should, in a future revision, report
Kenward-Roger or Satterthwaite small-sample corrections alongside the
LME procedure to address the conservatism, and should report the
Mancl-DeRouen GEE rather than naive GEE as the production-ready GEE
comparator.
\end{claudecode}

# Conclusions

\begin{claudecode}
In conclusion, three points are to be emphasised. First, the choice of
test procedure for the biomarker-treatment interaction in aggregated
N-of-1 trials has first-order consequences for power and type I error
at the sample sizes typical of precision-medicine pilot studies
($N \le 35$). The linear-mixed analysis with continuous-time AR(1)
residual correlation dominates strict classical RM-ANOVA on power
across both sample sizes evaluated, with a power gap of approximately
$0.20$ at $N = 25$ widening to $0.28$ at $N = 35$, while holding
nominal or sub-nominal type I error. Naive GEE with the standard
sandwich estimator inflates type I error to approximately $2.4\times$
nominal at $N = 25$ and $1.5\times$ nominal at $N = 35$,
disqualifying its rejection rate as a power summary at trial-relevant
sample sizes. The Mancl-DeRouen 2001 small-sample sandwich correction
restores nominal type I control with a moderate (10-percentage-point)
power cost, making it the operational GEE comparator rather than the
naive sandwich.

Second, the closed-form strict-RM-ANOVA test of the biomarker-treatment
interaction (\S2) reduces to a two-sample $t$-test on the within-subject
treatment differences under sphericity and balance. The reduction makes
the strict-RM-ANOVA result analytically transparent and connects the
simulation findings to the classical split-plot literature. That
transparency is, however, also what limits the test in the present
setting: the AR(1) residual correlation that the data-generating process
imposes violates the sphericity assumption underlying the closed form,
and the Greenhouse-Geisser correction that recovers nominal type I under
sphericity violation pays a degrees-of-freedom penalty that the
linear-mixed analysis avoids by modelling the AR(1) structure directly.

Third, the cycle-by-period design grid that constitutes the second axis
of the joint design-by-test sensitivity question framed in \S1 is
deferred to a follow-up paper. The infrastructure for the sweep is in
place at `analysis/scripts/test-procedure-design-sensitivity/`, and
the design specification is committed at
`docs/27-cycle-period-design-sweep-plan.md`. The follow-up paper will
deliver the joint (cycle count, period length, test procedure)
recommendation table that the present paper's \S6 narrows to the
test-procedure-only axis.
\end{claudecode}

# Conflict of interest

The authors declare no conflicts of interest.

# Funding

This work received no specific funding.

# Data availability statement

\begin{claudecode}
The data and code that support the findings of this study are openly
available in the pmsimstats-ng project repository. Per-replicate Monte
Carlo outputs and ADEMP-compliant cell-level summaries are versioned
under `analysis/data/`; simulation drivers are at
`analysis/scripts/`. Exact paths relevant to this manuscript are listed
in the Reproducibility section below.
\end{claudecode}

# Reproducibility

This manuscript was rendered with R 4.4.x inside the project's
zzcollab Docker container (image hash recorded in `Dockerfile`;
package set captured in `renv.lock`). Principal packages used
by the simulation pipeline are `nlme` (continuous-time
linear-mixed analysis with `corCAR1` residual correlation),
`geepack` plus a hand-rolled Mancl-DeRouen 2001 small-sample
sandwich correction (paper 08 only), `parallel` (8-core
`mclapply` for parameter-grid sweeps), `data.table` (the
`original-extended` simulation collection), `furrr` and `purrr`
(the `tidyverse` simulation collection), and `rmarkdown` plus
`bookdown` for the manuscript build. The simulation substrate
is `implementations/simple/simulation.R`.

**Random-number generator.** Each simulation driver sets
`set.seed(20260507)` at the top of the replicate loop and uses
`parallel::mc.set.seed` inside `mclapply`. Per-replicate seeds
derive from the master seed by `+ rep_idx`.

**Data and code availability.** Per-replicate simulation
outputs are checkpointed at
`analysis/data/quick-sim/08-test-replicates.rds`; Morris ADEMP
cell-level summaries at the companion `08-test-summary.txt`.
Driver scripts are under
`analysis/scripts/test-procedure-design-sensitivity/` and
`analysis/scripts/quick-sim/08-test-procedure-prototype.R`. The
pre-registered ADEMP plan is at
`analysis/scripts/test-procedure-design-sensitivity/00-ademp-pre-registration.md`.
The manuscript source, paper-specific bibliography
(`references.bib`), and SIM CSL file are at
`analysis/report/08-test-procedure-design-sensitivity/`.

# References

::: {#refs}
:::

```{r footer, echo=FALSE, results='asis'}
src <- sub("^/Users/zenn/(Library/CloudStorage/)?Dropbox/prj/", "~/prj/",
           knitr::current_input(dir = TRUE))
cat(sprintf(
  "\n\\vfill\n\n---\n\n*Rendered on %s.*  \n*Source: `%s`*\n",
  format(Sys.time(), "%Y-%m-%d at %H:%M %Z"),
  src
))
```
