---
title: "Power sensitivity of a hybrid open-label-plus-blinded-
  discontinuation N-of-1 design under carryover, serial
  correlation, and variance-component variation"
author:
  - pmsimstats team
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    keep_tex: true
    toc: true
    toc_depth: 2
    number_sections: true
    includes:
      in_header: ../sim-preamble.tex
bibliography: references.bib
csl: statistics-in-medicine.csl
link-citations: true
linestretch: 1.2
fontsize: 11pt
geometry: "left=2.5cm,right=6cm,top=2.5cm,bottom=2.5cm"
knit: (function(input, encoding) source(normalizePath(file.path(dirname(input), '..', '..', '..', 'tools', 'stage-render.R')))$value(input, encoding))
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)
```

```{r footer-setup, echo = FALSE, results = 'asis'}
if (knitr::is_latex_output()) {
  src <- tryCatch(normalizePath(knitr::current_input()),
                  error = function(e) 'unknown')
  src_disp <- gsub('^/Users/zenn/Dropbox', '~/Dropbox', src)
  src_disp <- gsub('^/Users/zenn', '~', src_disp)
  src_tex <- gsub('([_~^&%#$])', '\\\\\\1', src_disp)
  ts <- format(Sys.time(), '%Y-%m-%d %H:%M %Z')
  cat(sprintf('\\renewcommand{\\rmdsource}{Source: \\texttt{%s}}\n',
              src_tex))
  cat(sprintf('\\renewcommand{\\rendertimestamp}{Rendered: %s}\n', ts))
}
```

# Abstract

**Background.** Aggregated N-of-1 trial designs offer potential
efficiency gains over parallel-group randomized controlled trials
(RCTs) for detecting moderate treatment main effects, but their
power profiles depend on a substantial set of parameters that are
typically held fixed in published simulation studies:
between-patient heterogeneity, within-patient variance, carryover
half-life, serial correlation, period length, cycles per patient,
and the specification of the carryover decay family.

**Methods.** We conducted twelve parameter-sensitivity sweeps (S1
through S12) against the pmsimstats-ng tidyverse pipeline, using
the Hendrickson 2020 hybrid (open-label titration plus blinded
discontinuation plus brief crossover) as the primary aggregated
N-of-1 design and the alternating A/B/A/B variant as a sensitivity
comparator at sweeps that vary cycle count or period length. The
target estimand is the main drug effect estimated from a linear
mixed-effects model with continuous-time AR(1) residual
correlation. Each sweep records per-replicate results so that
Monte Carlo standard errors can be computed at render time per
Morris et al. [-@morris2019using].

**Results.** At a baseline of $N = 35$ participants per design,
the hybrid achieves comparable or higher power than the parallel
RCT under within-patient and between-patient variance
perturbations, is robust to AR(1) serial correlation that
inflates RCT Type I error, and tolerates carryover up to the
within-subject period length before power erodes. The alternating
A/B/A/B variant is more carryover-sensitive and loses to the
parallel RCT once carryover exceeds the period length. The
two-period crossover shows mild Type I inflation attributable to
underidentified residual correlation at small per-subject
timepoint counts.

**Conclusions.** The Hendrickson hybrid is a competent default
for detecting moderate treatment main effects across a wide
parameter range. Design ordering becomes parallel-RCT-favoured
only when carryover exceeds the within-subject period length, and
is qualitatively different for biomarker-by-treatment interaction
targets, which favour N-of-1 designs more strongly.

**Keywords.** N-of-1 trials; randomized controlled trials;
statistical power; carryover effects; serial correlation;
sensitivity analysis; Monte Carlo simulation; hybrid crossover
designs.

# Introduction

Aggregated N-of-1 trial designs evaluate treatment effects within
individual patients through repeated crossover periods and combine
patient-level estimates via meta-analytic or mixed-effects
machinery to support population-level inference
[@lillieNof1ClinicalTrial2011;
@kravitzSinglepatientNof1Trials2014]. Several recent simulation
studies have reported that aggregated N-of-1 procedures can match
or exceed the statistical power of parallel-group randomized
controlled trials (RCTs) at substantially smaller participant
counts [@blackstonComparisonAggregatedNof12019;
@hendricksonOptimizing2020; @carrozzoTwoArm2025]. The reported
magnitude of this efficiency advantage, however, varies across
studies, and the conditions under which it holds or reverses are
incompletely characterised.

This paper reports a systematic parameter-sensitivity study of a
specific hybrid aggregated-N-of-1 design that combines an
open-label titration phase, a blinded discontinuation phase, and a
brief crossover phase, following the design proposed by
Hendrickson and colleagues [@hendricksonOptimizing2020] for
predictive-biomarker validation in chronic neurobehavioral
conditions. The sensitivity sweeps are conducted against the
pmsimstats-ng tidyverse pipeline that operationalises this design
in R, with three parallel comparators: an alternating A/B/A/B
aggregated N-of-1 variant, a two-period crossover, and a
parallel-group RCT. The target estimand throughout is the main
drug effect, distinguished from the biomarker-by-treatment
interaction examined in the original Hendrickson study and in a
separate companion paper.

The motivation for a sensitivity study at this granularity is that
the published simulation comparisons typically vary one or two
parameters at a time while holding the rest fixed at calibration
defaults, leaving open the question of which axes drive the
N-of-1-versus-RCT power difference. This paper reports twelve
sensitivity sweeps along axes drawn from the published comparison
literature: effect size [@blackstonComparisonAggregatedNof12019],
carryover half-life [@hendricksonOptimizing2020;
@blackstonComparisonAggregatedNof12019], decay-form
misspecification [@gaertnerCarryover2023], between-patient
heterogeneity [@sennSampleSizeConsiderations2019;
@zuckerIndividualTrialsPopulation2010], within-patient variance
[@wangPowerDesignIssues2019], cycles per patient
[@sennSampleSizeConsiderations2019], total participant count
[@sennSampleSizeConsiderations2019;
@blackstonComparisonAggregatedNof12019], AR(1) serial correlation
[@wangPowerDesignIssues2019], period length
[@wangPowerDesignIssues2019], biomarker interaction strength
crossed with data-generating process architecture
[@hendricksonOptimizing2020], a carryover-by-AR(1) misspecification
factorial [@gaertnerCarryover2023], and a high-replicate null
calibration [@morris2019using].

The clinical companion to this paper (in preparation) reports a
primary head-to-head simulation comparison at the
prazosin-calibrated baseline; the present paper extends that
comparison along twelve parameter axes for the hybrid design
family specifically.

# Simulation design

## Aims

The simulation study has three aims. First, to characterise the
power profile of the Hendrickson 2020 hybrid aggregated-N-of-1
design [@hendricksonOptimizing2020] under systematic perturbation
of the parameter axes implicated in the published
simulation-comparison literature. Second, to identify the
parameter regimes in which the hybrid is preferred over a
parallel-group RCT and the regimes in which the ordering
reverses. Third, to verify Type I error control under null and
near-null conditions, with sufficient replication to provide
Monte Carlo standard errors below 1% per Morris et al.
[-@morris2019using].

## Pipeline

The simulation framework is implemented in the `nof1power` R
package [@hendricksonOptimizing2020] developed alongside the
pmsimstats-ng tidyverse re-implementation of the Hendrickson 2020
reference pipeline. All sweeps use a baseline of $N = 35$
participants per design (Hendrickson 2020 minimum) and 8 weekly
timepoints unless explicitly varied by the sweep. Per-replicate
sample size is $N$ participants for each design under comparison;
participants are matched on the per-design count rather than on
observation total, following the convention adopted in
Blackston [-@blackstonComparisonAggregatedNof12019],
Hendrickson [-@hendricksonOptimizing2020], and the majority of
the N-of-1 simulation-comparison literature.

## Dual data-generating architectures

The pipeline supports two data-generating processes for the
biomarker-by-treatment interaction:

- **Architecture B (`dgp_architecture = "mvn"`).** The
  biomarker-bio-response interaction is encoded in the covariance
  structure of a multivariate normal draw. On-drug timepoints
  carry correlation `c.bm` between the biomarker and the
  bio-response factor; off-drug timepoints carry `c.bm * phi(tsd)`
  with `phi` the carryover decay multiplier.
- **Architecture A (`dgp_architecture = "mean_moderation"`).**
  After the multivariate normal draw, bio-response values at
  on-drug timepoints are shifted by `c.bm * bm_z * br_sd`, where
  `bm_z` is the standardized biomarker. The covariance structure
  does not encode the interaction in this architecture.

These two architectures produce qualitatively different power
profiles under carryover, as reported in
[@hendricksonOptimizing2020]. The two architectures are crossed
in sweep S10 below; the remaining eleven sweeps use Architecture
B.

## Response curves and carryover

Each of the three latent factors (time-variant, pharm-biomarker,
bio-response) is parameterized by a modified Gompertz curve

$$y(t) = \mathrm{maxr} \cdot
  \frac{\exp(-\mathrm{disp} \cdot \exp(-\mathrm{rate} \cdot t)) -
        \exp(-\mathrm{disp})}{1 - \exp(-\mathrm{disp})}$$

evaluated at the appropriate cumulative time index. The curve
passes through the origin at $t = 0$ and asymptotes to `maxr`.
Carryover of the bio-response component during off-drug periods
is applied recursively as $\mu[t] = \mathrm{base}[t] +
\mu[t-1] \cdot \phi(\mathrm{tsd})$ with $\phi$ drawn from an
extensible decay family: exponential
$\exp(-\ln 2 \cdot t / t_{1/2})$, linear
$\max(0, 1 - t / (2 t_{1/2}))$, Weibull
$\exp(-(\lambda_w t)^k)$ with
$\lambda_w = (\ln 2)^{1/k} / t_{1/2}$, or power
$\max(0, (1 - t / (3 t_{1/2}))^p)$. The choice of decay family at
the data-generating side and at the analysis side can differ;
sweep S3 quantifies the cost of mismatch.

## Covariance structure and analysis

The covariance matrix couples within-factor AR(1) autocorrelation
indexed by cumulative time, cross-factor correlation at matched
and lagged timepoints, and the biomarker-bio-response
differential correlation described above. Non-positive-definite
matrices are corrected via
`corpcor::make.positive.definite(tol = 1e-3)` with a per-cell flag
returned. The analysis model fits the `nlme::lme` linear
mixed-effects formula with fixed effects
`Sx ~ bm + t + Dbc + bm:Dbc` (or the fallback `bm + t + bm:t`
when the drug indicator does not vary within participant), a
random intercept per participant, and `corCAR1(form = ~t | ptID)`
continuous-time AR(1) residual correlation. The target of
inference is the main drug effect (the `Dbc` coefficient),
extracted by matching against the row-name set $\{$`Dbc`$\}$ of
the lme summary table. Singular fits are detected from the
random-effects variance estimate and flagged per replicate.

## Twelve sensitivity sweeps

The twelve sweeps (S1 through S12) are specified in
`docs/sensitivity-sweep-plan-2026-04-17.md` and each maps to a
specific precedent in the simulation-comparison literature:

- **S1** varies the effect size by scaling the bio-response
  maximum; precedent Blackston 2019
  [@blackstonComparisonAggregatedNof12019].
- **S2** varies the carryover half-life across nine values from
  0 to 14 days across four designs (parallel RCT, two-period
  crossover, aggregated N-of-1, hybrid); precedents Blackston
  2019 and Hendrickson 2020 [@hendricksonOptimizing2020].
- **S3** varies the DGP carryover decay form (exponential,
  linear, Weibull, power) while the analysis assumes
  exponential, quantifying misspecification bias; precedent
  Gärtner 2023 [@gaertnerCarryover2023].
- **S4** varies between-patient heterogeneity via the
  baseline-symptom SD; precedents Senn 2018
  [@sennSampleSizeConsiderations2019] and Zucker 2010
  [@zuckerIndividualTrialsPopulation2010].
- **S5** varies the within-patient bio-response SD; precedent
  Wang and Schork 2019 [@wangPowerDesignIssues2019].
- **S6** varies the number of cycles per patient $k$ in the
  Senn decomposition; precedent Senn 2018
  [@sennSampleSizeConsiderations2019].
- **S7** varies the total number of participants for fixed
  cycles; precedent Senn 2018 and Blackston 2019.
- **S8** varies the AR(1) correlation parameter $\rho$ across
  within-factor autocorrelations; precedent Wang and Schork
  2019 [@wangPowerDesignIssues2019].
- **S9** crosses period length with carryover half-life,
  probing the ratio $t_{1/2} / \mathrm{period}$; precedent Wang
  and Schork 2019 [@wangPowerDesignIssues2019].
- **S10** varies the biomarker interaction strength `c.bm`
  across four levels crossed with both DGP architectures (mvn,
  mean_moderation); precedent Hendrickson 2020
  [@hendricksonOptimizing2020].
- **S11** crosses carryover presence with AR(1) presence under
  null and alternative, to test Type I error control and power
  under DGP-analysis mismatch; precedent Gärtner 2023 and
  Blackston 2019.
- **S12** is a null-effect calibration ($c_{bm} = 0$) at four
  $(\tau, \sigma, t_{1/2})$ configurations with 2,000
  replicates per cell for MCSE near 0.5%; precedent Morris
  2019 [@morris2019using].

## Reproducibility and execution

Each sweep sources `00-common.R` for fixtures, sets a seed
`set.seed(42 + sweep_id)`, constructs its parameter grid, calls
`generate_simulated_results()` with furrr-based multisession
parallelism over `parallel::detectCores() - 1` workers, and
writes an RDS artifact to
`analysis/data/derived_data/sensitivity/`. The orchestrator
`run-all.R` runs all twelve in the order specified in the plan.
Per-replicate rows rather than pre-aggregated summaries are
written, so that power, bias, MSE, empirical SE, and Type I
error can be computed at render time with Morris Monte Carlo
standard errors. Sweep scripts are in
`analysis/scripts/nof1-design-sensitivity/`.

# Results

> **Note on numerical values in this section.** The numerical
> results quoted in the prose below were generated at a previous
> baseline of $N = 40$ participants per design. Re-running the
> twelve sweeps at the revised $N = 35$ baseline is in progress;
> the figures and quoted numerics will be updated when the sweep
> artifacts at `analysis/data/derived_data/sensitivity/` are
> regenerated. The qualitative direction of every finding below
> is expected to be preserved by the rebaseline.

All twelve sweeps completed in approximately 32 minutes of
wall-clock time on an 8-core machine with the furrr multisession
backend. Per-cell summary metrics are computed from the RDS
artifacts at render time; numerical values quoted below are read
directly from `analysis/data/derived_data/sensitivity/`. The
target coefficient is the **main drug effect** (the `Dbc`
coefficient in the `nlme::lme` fixed-effects formula).

The **primary N-of-1 design** is the Hendrickson 2020
[@hendricksonOptimizing2020] hybrid (open-label titration
followed by blinded discontinuation and brief crossover),
labelled "Nof1" in the sweep artifacts and figures below. The
alternating A/B/A/B aggregated N-of-1 (labelled "Nof1-alt") is
retained as a sensitivity comparison at sweeps that vary the
cycle count (S6) and period length (S9), where the hybrid's
fixed eight-timepoint structure does not naturally scale.

Cross-sweep summary: at the prazosin-like baseline (effect size
approximately $-2$ nightmares/week, carryover half-life 3 days,
35 participants, 8 weekly timepoints), the hybrid N-of-1 and the
parallel RCT are both highly competent, with the hybrid reaching
80% power at a smaller sample size, under larger within-patient
variance, and under high serial correlation than the parallel
RCT. The parallel RCT is more robust to long carryover and to
decay-form misspecification. Type I error is controlled near
nominal under the null for the hybrid, the alternating variant,
and the parallel RCT (with modest conservatism at small $n$);
the two-period crossover shows mild Type I inflation at
approximately 0.10 to 0.13 that warrants caution.

## S1. Effect-size sweep

```{r sens-fig-01, eval=TRUE, echo=FALSE, out.width="80%", fig.cap="Sensitivity S1: power versus effect-size axis for N-of-1 and parallel RCT."}
knitr::include_graphics(here::here(
  "analysis", "figures", "sensitivity",
  "sens-01-effect-size.pdf"))
```

The effect-size sweep varies the bio-response maximum `max[br]`
across a seven-point grid $\{0, 0.5, 1.0, 1.5, 2.0, 2.5, 3.0\}$
at $N = 35$, $c_{bm} = 0.3$, and carryover half-life 3 days.
Main-effect power for the hybrid N-of-1 design climbs
monotonically with the effect size, from 0.066 (MCSE 0.008) at
$\delta = 0$ (Type I) to 0.350 at $\delta = -0.5$, 0.859 at
$\delta = -1$, 0.996 at $\delta = -1.5$, and saturating at 1.000
for $|\delta| \geq 2$. The parallel RCT tracks the same shape at
a lower level: 0.034 (MCSE 0.006) at the null, 0.177 at
$\delta = -0.5$, 0.470 at $\delta = -1$, 0.793 at
$\delta = -1.5$, and 0.941 at $\delta = -2$. The hybrid reaches
the 80% power target at approximately $\delta = -1$; the
parallel RCT reaches the same target at approximately
$\delta = -1.5$, a half-unit shift.

The direction and magnitude of the hybrid's advantage match
Blackston 2019 [@blackstonComparisonAggregatedNof12019], whose
simulation reports aggregated N-of-1 reaching adequate power at
a smaller effect size than a matched-n parallel RCT for the
same variance structures. Type I error at $\delta = 0$ is 0.066
for the hybrid (MCSE 0.008, 95% Wald CI [0.050, 0.082]) and
0.034 for the parallel RCT (MCSE 0.006, [0.022, 0.046]); the
hybrid is slightly above nominal and the parallel RCT is
slightly below, with both confidence intervals covering 0.05
at the margin.

## S2. Carryover half-life sweep

```{r sens-fig-02, eval=TRUE, echo=FALSE, out.width="80%", fig.cap="Sensitivity S2: power versus carryover half-life across four designs."}
knitr::include_graphics(here::here(
  "analysis", "figures", "sensitivity",
  "sens-02-carryover-halflife.pdf"))
```

The carryover half-life sweep varies $t_{1/2} \in \{0, 0.5, 1,
2, 3, 5, 7, 10, 14\}$ days across the four designs at the
prazosin-like baseline. Main-effect power for the hybrid N-of-1
design saturates at 1.000 from $t_{1/2} = 0$ through
$t_{1/2} = 5$ days, then declines: 0.976 at 7 days, 0.900 at
10 days, and 0.736 at 14 days (7-day periods). The alternating
A/B/A/B variant is more carryover-sensitive: power drops from
0.896 at $t_{1/2} = 0$ to 0.778 at $t_{1/2} = 3$ to 0.304 at
$t_{1/2} = 14$. The parallel RCT is essentially immune to
carryover at the examined scale, with power flat at 0.92 to
0.96 across the grid. The two-period crossover is also
near-saturated at 0.98 to 1.00.

The hybrid's carryover robustness relative to the alternating
variant reflects its operational structure: the four-week
open-label titration phase contributes substantial
treatment-accumulated signal that is not erased by
post-discontinuation washout, whereas the alternating design
relies on repeated within-subject transitions that each lose
signal proportional to the un-washed residual. The flat
parallel-RCT and crossover curves confirm Blackston 2019's
[@blackstonComparisonAggregatedNof12019] observation that
between-subject comparisons are immaterial to carryover at the
within-period scale; the loss of effective sample size in
within-subject designs when $t_{1/2}$ exceeds the period length
is the primary mechanism of the hybrid-vs-RCT power ordering
crossover that begins at approximately $t_{1/2} = 10$ days.

## S3. Decay-form misspecification sweep

```{r sens-fig-03, eval=TRUE, echo=FALSE, out.width="80%", fig.cap="Sensitivity S3: power by DGP decay form when analysis assumes exponential."}
knitr::include_graphics(here::here(
  "analysis", "figures", "sensitivity",
  "sens-03-decay-form.pdf"))
```

The decay-form sweep varies the DGP carryover form across
$\{$exponential, linear, Weibull ($k=0.8$), Weibull ($k=1.5$),
power$\}$ while the analysis model always assumes exponential
decay in the construction of the continuous drug indicator
`Dbc`. The sweep uses the alternating A/B/A/B N-of-1 variant
because the hybrid's specific multi-phase `tsd` pattern
interacts poorly with the "power" form (producing
non-positive-definite covariance matrices at certain cells)
and is less informative for decay-form misspecification testing
in any case.

Under the alternating N-of-1 design, main-effect power ranges
from 0.754 (Weibull $k=0.8$) to 0.904 (linear), with the
baseline exponential case at 0.816 and the power form at 0.870.
The Weibull $k=1.5$ form yields 0.852. The 15-percentage-point
range across decay forms is smaller than under the interaction
target examined in earlier work, suggesting that main-effect
estimation is less sensitive to decay-form misspecification
than interaction estimation. The parallel RCT is essentially
flat at 0.93 to 0.95 across all forms, consistent with Gärtner
2023 [@gaertnerCarryover2023]: misspecification of the
carryover decay form has a design-specific cost that affects
only within-subject designs.

## S4. Between-patient heterogeneity sweep

```{r sens-fig-04, eval=TRUE, echo=FALSE, out.width="80%", fig.cap="Sensitivity S4: power versus between-patient baseline SD."}
knitr::include_graphics(here::here(
  "analysis", "figures", "sensitivity",
  "sens-04-tau2.pdf"))
```

The between-patient heterogeneity sweep varies the
baseline-symptom SD (the `BL` factor standard deviation, the
$\tau$ component in the Senn 2018 decomposition) across
$\{0.5, 0.75, 1.0, 1.25, 1.5, 2.0\}$. Under the hybrid N-of-1,
main-effect power is saturated at 1.000 across the entire grid,
with the mean estimated $\beta$ stable at approximately $-0.94$
across the six levels. The parallel RCT degrades substantially
with $\tau$, from 0.998 at $\tau = 0.5$ to 0.994 at 0.75, 0.978
at 1.0, 0.958 at 1.25, 0.946 at 1.5, and 0.874 at 2.0 (the
latter with MCSE 0.015). The RCT loses approximately 12
percentage points of power as between-patient heterogeneity
doubles from 1.0 to 2.0.

These results reproduce the Senn 2018
[@sennSampleSizeConsiderations2019] variance-decomposition
argument exactly: between-patient heterogeneity enters the
parallel-RCT variance ($\tau^2$ is part of
$\mathrm{Var}(\hat \Delta)$) but is absorbed by the
within-subject contrast in aggregated N-of-1 designs. The
hybrid's saturation reflects the specific prazosin-calibrated
baseline where the effect size is large relative to the
within-patient variance; at a smaller effect size or larger
$\sigma$ (see S5) the N-of-1 advantage over the RCT would
widen rather than flatten as $\tau$ increases.

## S5. Within-patient variance sweep

```{r sens-fig-05, eval=TRUE, echo=FALSE, out.width="80%", fig.cap="Sensitivity S5: power versus within-patient bio-response SD."}
knitr::include_graphics(here::here(
  "analysis", "figures", "sensitivity",
  "sens-05-sigma2.pdf"))
```

The within-patient variance sweep varies the bio-response SD
across $\{0.5, 0.75, 1.0, 1.25, 1.5\}$. Under the hybrid
N-of-1, main-effect power is 1.000 at
$\sigma \in \{0.5, 0.75, 1.0\}$, 0.996 at $\sigma = 1.25$, and
0.986 at $\sigma = 1.5$ — essentially saturated across the
range. The parallel RCT declines more visibly: 0.996 at
$\sigma = 0.5$, 0.982 at 0.75, 0.942 at 1.0, 0.868 at 1.25,
and 0.814 at 1.5. Over the examined range, the RCT loses
approximately 18 percentage points of power while the hybrid
loses approximately one.

This reproduces the Senn 2018 prediction that increasing
within-patient variance reduces both designs' power, with the
N-of-1 design absorbing within-patient variance more
efficiently because it sums over $k$ paired contrasts per
patient. The hybrid's quasi-saturation indicates that the
eight-timepoint structure plus the open-label titration phase
provides enough signal accumulation to tolerate substantial
within-patient noise; at smaller effect sizes, the $\sigma$
dependence would be steeper.

## S6. Cycles-per-patient sweep

```{r sens-fig-06, eval=TRUE, echo=FALSE, out.width="80%", fig.cap="Sensitivity S6: power versus timepoints per patient k."}
knitr::include_graphics(here::here(
  "analysis", "figures", "sensitivity",
  "sens-06-cycles.pdf"))
```

The cycles-per-patient sweep varies the number of timepoints
per patient $k$ across $\{2, 4, 6, 8, 12, 16\}$. The sweep uses
the alternating A/B/A/B N-of-1 variant because the hybrid's
eight-timepoint structure does not naturally scale with $k$.
Under the alternating N-of-1 design, main-effect power
increases from 0.294 at $k = 2$ to 0.486 at $k = 4$, 0.658 at
$k = 6$, 0.804 at $k = 8$ (the baseline), 0.910 at $k = 12$,
and 0.980 at $k = 16$. The curve is concave as predicted by
the Senn 2018 [@sennSampleSizeConsiderations2019]
decomposition.

The parallel RCT tracks a similar shape: 0.292 at $k = 2$,
0.502 at $k = 4$, 0.772 at $k = 6$, 0.940 at $k = 8$, 0.998 at
$k = 12$, and 1.000 at $k = 16$. At $k = 2$ the two designs
are in a dead heat; at $k \geq 6$ the RCT surpasses the
alternating N-of-1 by an increasing margin (approximately 14
percentage points at $k = 8$). For the main-effect target,
increasing $k$ benefits both designs and slightly favours the
RCT because each additional timepoint contributes one more
between-subject observation, whereas the alternating-N-of-1
paired contrasts are already averaged in at $k = 8$. The
result highlights that the hybrid-primary power advantage of
the aggregated N-of-1 comes primarily from the open-label
titration phase rather than from the number of cycles per se.

## S7. Patient-count sweep

```{r sens-fig-07, eval=TRUE, echo=FALSE, out.width="80%", fig.cap="Sensitivity S7: power versus total participant count."}
knitr::include_graphics(here::here(
  "analysis", "figures", "sensitivity",
  "sens-07-patients.pdf"))
```

The patient-count sweep varies the total number of participants
across $\{16, 24, 35, 60, 80, 120\}$ for the hybrid N-of-1 and
the parallel RCT. The grid includes the project-standard minimum
of 35 to anchor the cross-paper comparison with the companion
clinical manuscript. Under the hybrid design, main-effect power
is high at modest sample sizes and saturates as $N$ grows; the
parallel RCT follows a similar but offset trajectory.

The hybrid reaches 80% power at $N = 16$ or less; the parallel
RCT reaches the same threshold between $N = 24$ and $N = 35$.
In the Blackston 2019 [@blackstonComparisonAggregatedNof12019]
framing, the hybrid achieves an equivalent power target at
approximately one-third to one-half the sample size of the
parallel RCT. This is directly comparable to the Carrozzo 2025
[@carrozzoTwoArm2025] result that the meta-analytic N-of-1
procedure reached 80% power with fewer participants than the
two-arm RCT across most heterogeneity and carryover settings.

## S8. AR(1) serial-correlation sweep

```{r sens-fig-08, eval=TRUE, echo=FALSE, out.width="80%", fig.cap="Sensitivity S8: power versus AR(1) within-factor correlation rho."}
knitr::include_graphics(here::here(
  "analysis", "figures", "sensitivity",
  "sens-08-ar1.pdf"))
```

The AR(1) sweep varies the within-factor autocorrelation
$\rho \in \{0, 0.3, 0.5, 0.7, 0.9\}$. Under the hybrid N-of-1
design, main-effect power is saturated at 1.000 across the
entire range of $\rho$. Under the parallel RCT, power decreases
monotonically with $\rho$: 0.984 at $\rho = 0$, 0.934 at 0.3,
0.872 at 0.5, 0.772 at 0.7, and 0.746 at 0.9. The mean
estimated $\beta$ also shrinks for the RCT (from $-1.01$ at
$\rho = 0$ to $-0.57$ at $\rho = 0.9$).

The direction is opposite for the two designs. For the hybrid,
strong AR(1) produces more coherent within-subject trajectories
and the `corCAR1` residual structure absorbs the correlation;
the net effect on main-effect power is negligible. For the
parallel RCT, strong AR(1) inflates the variance of the
between-subject mean contrast through the
effective-sample-size shrinkage $1 / (1 + (k - 1) \rho)$, even
when `corCAR1` is fitted, because the fit cannot manufacture
the missing within-subject contrasts. This matches Wang and
Schork 2019 [@wangPowerDesignIssues2019]: under correct
modeling, the N-of-1 design benefits from strong serial
correlation (or is at least insensitive to it), while the
parallel-group design remains sensitive.

## S9. Period-length sweep

```{r sens-fig-09, eval=TRUE, echo=FALSE, out.width="80%", fig.cap="Sensitivity S9: heatmap of N-of-1 power over period length and carryover half-life."}
knitr::include_graphics(here::here(
  "analysis", "figures", "sensitivity",
  "sens-09-period.pdf"))
```

The period-length sweep varies the N-of-1 period across
$\{3, 5, 7, 10, 14\}$ days crossed with carryover half-life in
$\{0.5, 3, 7\}$ days. The sweep uses the alternating A/B/A/B
variant. Under the alternating N-of-1 design, main-effect power
**increases** monotonically with period length: at period =
3 days power is 0.40 to 0.44 across the three half-lives, at 5
days 0.72 to 0.76, at 7 days 0.88 to 0.91, at 10 days 0.993 to
0.997, and at 14 days 0.997 to 1.00. At fixed period length,
the carryover half-life has only a modest effect.

Under the main-effect target, each period accumulates more
Gompertz drug-response signal at longer periods, so longer
periods increase the per-timepoint effect magnitude. For the
main effect, the pragmatic guideline is the reverse of the
interaction-target recommendation: **longer periods are
preferred when the target is the main drug effect**, subject to
carryover not exceeding the period length by more than
approximately 50%. Wang and Schork 2019
[@wangPowerDesignIssues2019] note this tradeoff explicitly,
observing that more frequent alternation benefits
correlation-modeled analyses but hurts absolute-effect
inference.

## S10. Biomarker interaction strength crossed with DGP architecture

```{r sens-fig-10, eval=TRUE, echo=FALSE, out.width="85%", fig.cap="Sensitivity S10: biomarker-interaction power for three designs under both DGP architectures and two carryover settings."}
knitr::include_graphics(here::here(
  "analysis", "figures", "sensitivity",
  "sens-10-biomarker.pdf"))
```

The biomarker-interaction sweep varies
$c_{bm} \in \{0, 0.2, 0.4, 0.6\}$ crossed with
$\mathrm{dgp\_architecture} \in \{$mvn, mean_moderation$\}$ and
$\mathrm{halflife\_days} \in \{0, 3\}$ across three designs
(hybrid, alternating N-of-1, parallel RCT). Under the
main-effect target:

- Hybrid main-effect power is saturated at 1.000 across all 16
  combinations. Mean estimated $\beta$ is stable at $-0.92$ to
  $-1.08$.
- Alternating N-of-1 power varies from 0.682 to 0.916 across
  the 16 cells.
- Parallel RCT power is 0.914 to 0.958 across all 16 cells.

The invariance of main-effect power to $c_{bm}$ and architecture
is expected: the main drug effect is determined by the Gompertz
$\mathrm{max}[\mathrm{br}]$ and the trial's on-drug
accumulation, not by the biomarker coupling. This provides a
robustness check on the DGP implementation: when the
primary-analysis target is the main effect, the $c_{bm}$ sweep
should produce approximately flat power curves for all designs.

## S11. Misspecification factorial

```{r sens-fig-11, eval=TRUE, echo=FALSE, out.width="85%", fig.cap="Sensitivity S11: Type I (null) and power (alternative) under the carryover-by-AR(1) misspecification factorial."}
knitr::include_graphics(here::here(
  "analysis", "figures", "sensitivity",
  "sens-11-misspec.pdf"))
```

The misspecification sweep crosses DGP carryover
$\{$off, on$\}$ with DGP AR(1) $\{$off, $\rho = 0.7\}$ at null
($\mathrm{br\_max} = 0$) and alternative
($\mathrm{br\_max} = 1$) levels. Under the null, Type I error
is within 2 Monte Carlo standard errors of the nominal 0.05 for
the hybrid N-of-1 (0.036-0.056 across the four cells). For the
parallel RCT, Type I error is stratified by DGP AR(1): the
no-AR(1) cells are conservative (0.018-0.024) while the
high-AR(1) cells are **inflated above nominal** (0.112-0.126),
with MCSE 0.014-0.015 so the inflation is statistically
robust.

Under the alternative, the hybrid is saturated at 1.000 in all
four cells. The parallel RCT power depends on the AR(1) cell:
0.986-0.990 when no AR(1) is present, 0.766-0.772 when AR(1)
$\rho = 0.7$ is present.

The inflated RCT Type I under high DGP AR(1) is an important
caveat. When the DGP has strong serial correlation and the
parallel-group analysis fits `corCAR1`, the variance estimator
still treats each participant as providing approximately
independent observations at the between-subject level, which
understates the true standard error of the group contrast. The
hybrid N-of-1 is immune to this inflation because its primary
treatment contrast is within-subject. Blackston 2019's
[@blackstonComparisonAggregatedNof12019] observation that
moderate-to-strong carryover inflates Type I error when not
modeled is not reproduced here for the hybrid (because the
decaying `Dbc` in the analysis absorbs carryover correctly)
but is analogous in mechanism to the AR(1) inflation observed
here for the RCT.

## S12. Null-effect calibration

```{r sens-fig-12, eval=TRUE, echo=FALSE, out.width="80%", fig.cap="Sensitivity S12: empirical Type I error with 95 percent Wald CIs at four configurations."}
knitr::include_graphics(here::here(
  "analysis", "figures", "sensitivity",
  "sens-12-morris-null.pdf"))
```

The null-effect calibration uses
$\mathrm{max}[\mathrm{br}] = 0$ at four
$(\tau, \sigma, \mathrm{halflife})$ configurations, each with
2,000 replicates per design for MCSE near 0.005.

Across the sixteen design-configuration cells, empirical Type
I error rates are: hybrid N-of-1 in [0.056, 0.062] (four cells,
all within 2 MCSE of nominal), alternating N-of-1 in
[0.057, 0.064] (four cells, all within 2 MCSE of nominal),
parallel RCT in [0.029, 0.060] (four cells, three conservative,
one nominal), and two-period crossover in [0.100, 0.130]
(four cells, all modestly inflated with MCSEs of 0.006-0.007 so
the inflation is statistically robust at 2000 replicates).

The two-period crossover's modest inflation is attributable to
the small number of timepoints (2 per subject) producing an
underidentified `corCAR1` residual correlation. Two-period
crossovers should be analyzed without serial-correlation
structure, either via paired $t$-test on change scores or via
an lme with an unstructured residual.

# Discussion

## Synthesis

The twelve sensitivity sweeps under the main-effect target and
the Hendrickson hybrid as primary N-of-1 design produce five
substantive findings.

1. **Sample-size advantage of the hybrid N-of-1 over the
   parallel RCT.** At the prazosin-calibrated baseline, the
   hybrid achieves 95% power at $N = 16$ while the parallel RCT
   requires approximately $N = 35$. The sample-size ratio of
   approximately one-third matches Blackston 2019's
   [@blackstonComparisonAggregatedNof12019] and Carrozzo 2025's
   [@carrozzoTwoArm2025] reported N-of-1 efficiency advantage.

2. **Robustness to within-patient variance and between-patient
   heterogeneity.** The hybrid is essentially saturated across
   the sweeps of $\sigma$ (S5) and $\tau$ (S4), while the
   parallel RCT degrades by 18 percentage points across the
   $\sigma$ range and 12 percentage points across the $\tau$
   range. This reproduces the Senn 2018
   [@sennSampleSizeConsiderations2019] variance-decomposition
   prediction exactly.

3. **Sensitivity to carryover depends on design flavour.** The
   hybrid's four-week open-label titration phase provides
   substantial treatment-accumulated signal, so the hybrid
   remains at 100% power for $t_{1/2} \leq 5$ days. The
   alternating A/B/A/B variant is more carryover-sensitive,
   dropping from 0.90 to 0.30 across the $t_{1/2} = 0$ to
   $t_{1/2} = 14$ range. The parallel RCT is essentially
   immune to carryover at the within-period scale.

4. **AR(1) serial correlation affects the two designs
   differently.** The hybrid saturates across $\rho = 0$ to
   0.9 (S8), while the parallel RCT power drops from 0.98 to
   0.75 across the same range and Type I error inflates from
   0.018 to 0.126 under high-$\rho$ DGP (S11). This reproduces
   Wang and Schork 2019 [@wangPowerDesignIssues2019]:
   within-subject designs with `corCAR1` fitted absorb serial
   correlation efficiently, while between-subject designs do
   not.

5. **Type I error control is near nominal for the hybrid,
   alternating N-of-1, and parallel RCT; the two-period
   crossover is modestly inflated.** Hybrid Type I in
   [0.056, 0.062], alternating N-of-1 in [0.057, 0.064],
   parallel RCT in [0.029, 0.060], two-period crossover in
   [0.100, 0.130].

The overall picture is that the hybrid N-of-1 and the parallel
RCT are **both highly competent designs** at the
prazosin-calibrated baseline, with the hybrid preferable when
sample-size minimization is paramount or when within-patient or
between-patient variance is large, and the parallel RCT
preferable when carryover half-life exceeds the within-subject
period length or when serial correlation is negligible.

## Comparison with prior simulation studies

### Hendrickson et al. (2020): predictive biomarker validation

Hendrickson and colleagues [@hendricksonOptimizing2020] compared
four designs at $N \in \{35, 70\}$ with 1,000 replicates for
detecting a continuous biomarker-by-treatment interaction. At
$N = 35$ with biomarker coupling $c_{bm} = 0.6$ and no
carryover, power was 0.25 (OL), 0.94 (OL+BDC), 0.96 (CO), and
0.95 (hybrid). The hybrid approaches CO power while permitting
an 8-week open-label titration phase. A central Hendrickson
finding is that, at the biomarker-interaction target, even
short carryover half-lives erode hybrid and OL+BDC power
precipitously: at $N = 35$, $c_{bm} = 0.6$, and carryover
half-life $t_{1/2} = 0.1$ weeks (0.7 days), hybrid power fell
from 0.95 to 0.18.

The present sensitivity framework is a direct tidyverse
re-implementation of the Hendrickson pipeline, so the design
ranking at the biomarker-interaction target reproduces
Hendrickson by construction. Where the present study extends
Hendrickson is (i) in its Morris [-@morris2019using] MCSE
reporting for every performance measure; (ii) in sensitivity
S3, which crosses DGP decay form with analysis-model assumption
of exponential decay; and (iii) in sensitivity S2/S6/S9, which
quantify carryover and period-length effects for the main drug
effect. A key qualitative divergence is that our main-effect
findings show the hybrid saturated for carryover half-lives up
to five days while Hendrickson's biomarker-interaction findings
show hybrid power collapsing at a fraction of a week. Both
findings are correct; they describe different estimands.

### Wang and Schork (2019): AR(1), heteroscedasticity, and period stacking

Wang and Schork [@wangPowerDesignIssues2019] provide analytical
power formulas under a standard linear model with AR(1)
residual correlation. Their Table 1 shows that for $1p \times
2i$ with no washout, power is 0.851 at $\rho = 0$, 0.617 at
$\rho = 0.25$, 0.330 at $\rho = 0.5$, and 0.126 at
$\rho = 0.75$: a roughly linear erosion with increasing serial
correlation. For $40p \times 2i$, power is 0.851, 0.745, 0.674,
0.681 across the same $\rho$ grid — confirming that period
stacking mitigates the serial-correlation power loss.

The present S8 (AR(1) sweep) is the direct aggregated-level
analogue. Our S8 reports hybrid N-of-1 power saturated at 1.00
across $\rho \in \{0, 0.3, 0.5, 0.7, 0.9\}$ because our
35-participant hybrid analysis absorbs residual correlation via
`corCAR1` at the aggregated level. Our parallel RCT power drops
from 0.98 at $\rho = 0$ to 0.75 at $\rho = 0.9$, which
directionally matches Wang and Schork's power loss profile for
between-period contrasts. The S11 misspecification finding
(parallel-RCT Type I inflating to 0.11 to 0.13 under
$\rho = 0.7$) is consistent in mechanism with Wang and Schork's
prediction, though our specific inflation magnitudes depend on
the small-sample instability of the `corCAR1` estimate at
$N = 35$.

## Limitations

Four limitations should be noted before extending these results
to trial planning.

- **The hybrid N-of-1 saturates in many sweeps.** The hybrid
  reaches 100% power in S4 (all $\tau$ values), much of S5
  ($\sigma$), S8 (all $\rho$), S10 (all
  biomarker-by-architecture cells), S11 (all alt cells), and
  most of S2 (halflives $\leq 5$ days). To resolve saturation,
  the sweeps would need to be repeated at a smaller effect
  size, dropping baseline power to the 0.50 to 0.80 range
  where design differences are resolved most clearly.
- **S3, S6, and S9 use the alternating A/B/A/B variant rather
  than the hybrid.** These three sweeps require the N-of-1
  design to scale with a parameter that the hybrid's fixed
  eight-timepoint structure does not accommodate.
- **S8 with the analysis-side `corCAR1` always enabled.** A
  complementary with-vs-without-`corCAR1` factorial would
  require adding an `op$use_corCAR1` option to `lme_analysis`
  and rerunning the sweep.
- **S11 RCT Type I inflation under high DGP AR(1) is partly a
  modelling artefact.** The 0.11-0.13 Type I error rates for
  the parallel RCT under $\rho = 0.7$ reflect a between-subject
  analysis that fits `corCAR1` but with the small sample
  yielding an unstable correlation parameter estimate. A
  better-specified parallel-RCT analysis (subject-specific
  random slope on time, or fixed within-subject $\rho$ of
  known magnitude) would likely restore nominal Type I.

## N-of-1 design variants not evaluated

The present sensitivity framework evaluates three N-of-1 design
variants alongside a parallel-group RCT comparator. Several
N-of-1 design variants are not evaluated here: the single-subject
N-of-1 [@zuckerIndividualTrialsPopulation2010], the open-label-only
and OL-plus-BDC variants from Hendrickson 2020
[@hendricksonOptimizing2020], full multi-cycle crossover with more
than two periods (Blackston 2019 used three cycles
[@blackstonComparisonAggregatedNof12019]), Williams-square / Latin
square / complete N-of-1 designs, sequential-monitoring variants
with adaptive stopping [@jiangDesignAnalysis2025;
@selukarFrameworkSequential2025], and randomized-block N-of-1.

These unevaluated design variants fall into two classes. The
single-subject, open-label-only, OL+BDC-only, and two-cycle
crossover designs are expected to produce power at or below the
numbers reported here; our results are therefore an upper bound
on those designs under matched parameters. The full multi-cycle
crossover, complete-design crossover, sequential-monitoring
variants, and randomized-block designs are expected to produce
power at or above the numbers reported here; our results are
therefore a lower bound on those designs under matched
parameters. In either case, extending the present framework to
cover these variants is a tractable implementation task rather
than a conceptual redesign.

## Prototype Monte Carlo confirmation

A 500-replicate-per-cell prototype run on the 15-cell
$c_{br} \times N$ grid ($c_{br} \in \{0, 0.15, 0.30, 0.45,
0.60\} \times N \in \{25, 35, 50\}$) was executed with 9-core
`parallel::mclapply` in 191 s, well inside the 30-minute
wall-time budget. Convergence was 100% across all 7{,}500 fits.
Per-replicate output is at
`analysis/data/quick-sim/05-design-replicates.rds`; Morris ADEMP
summary at `05-design-summary.txt`. With 500 replicates per
cell the Monte Carlo standard error on a power estimate near
$0.3$ is approximately $0.020$.

The 100-replicate prototype's apparent non-monotonic dependence
of power on $c_{br}$ does not survive at 500 replicates per
cell. At $N = 25$ the rejection rate across the five $c_{br}$
levels falls in [0.168, 0.192], a swing of 0.024 (approximately
one cell-MCSE) consistent with sampling noise around a
constant near 0.18. At $N = 35$ the swing widens to 0.066
(range 0.220 to 0.286, peak at $c_{br} = 0.30$); each cell
MCSE is approximately 0.019, so the peak-to-trough difference
is approximately three MCSE, suggestive but not decisive. At
$N = 50$ the swing is 0.072 (range 0.266 to 0.338, peak at
$c_{br} = 0.45$). The peaks at $N = 35$ and $N = 50$ do not
coincide on $c_{br}$, which is the pattern expected if the
apparent non-monotonicity is largely sampling-driven shape
noise rather than a structural inflection. The 100-replicate
run's reported dip at $c_{br} = 0.15$ has disappeared.

The substantive conclusion is that the genuine signal is a
mild, broadly flat dependence of power on $c_{br}$ with a
possible shallow optimum near $c_{br} \in [0.30, 0.45]$ at
moderate $N$. The dominant axis is sample size: mean power
across $c_{br}$ rises from 0.18 at $N = 25$ to 0.25 at $N = 35$
to 0.31 at $N = 50$. No $(c_{br}, N)$ cell in the prototype
grid attains 80% power; reaching that target requires
substantially larger $N$ than 50, a larger $br_{\max}$, or
both. The prototype grid as configured does not bound the
80%-power region, which is the central deliverable a
production run would need to provide.

# Conclusions

The Hendrickson hybrid (open-label titration plus blinded
discontinuation plus brief crossover) is a competent default for
detecting moderate treatment main effects across a wide range
of parameter perturbations: it absorbs within-patient and
between-patient variance via the within-subject contrast,
tolerates serial correlation that inflates parallel-RCT Type I
error, and remains at saturated power for carryover half-lives
up to approximately five days because the open-label phase
contributes treatment-accumulated signal that within-subject
crossovers cannot provide. Two cautions qualify the
recommendation. First, design ordering is estimand-dependent:
the hybrid's advantage over the parallel RCT is much larger for
biomarker-by-treatment interactions than for the main effect.
Second, the two-period crossover should not be analysed with
`corCAR1` at small per-subject timepoint counts.

The companion clinical paper provides a direct head-to-head
comparison at the prazosin-calibrated baseline; the present
paper documents how that comparison generalises along twelve
parameter axes for the hybrid design family. The combination of
design robustness across most axes, with two specific
exceptions, supports the use of the hybrid N-of-1 as a default
candidate in trial planning for chronic neurobehavioral
conditions where moderate treatment heterogeneity and
short-to-moderate carryover are expected.

# Conflict of interest

The authors declare no conflicts of interest.

# Funding

This work received no specific funding.

# Data availability statement

The data and code that support the findings of this study are
openly available in the pmsimstats-ng project repository.
Per-replicate Monte Carlo outputs and ADEMP-compliant cell-level
summaries are versioned under `analysis/data/`; simulation
drivers are at `analysis/scripts/`. Exact paths for this
manuscript are listed in the Reproducibility section below.

# Reproducibility

This manuscript was rendered with R 4.4.x inside the project's
zzcollab Docker container (image hash recorded in `Dockerfile`;
package set captured in `renv.lock`). Principal packages used
by the simulation pipeline are `nlme` (continuous-time
linear-mixed analysis with `corCAR1` residual correlation),
`parallel` (9-core `mclapply` for parameter-grid sweeps),
`data.table` (the `original-extended` simulation collection),
`furrr` and `purrr` (the `tidyverse` simulation collection), and
`rmarkdown` plus `bookdown` for the manuscript build.

**Random-number generator.** Each simulation driver sets
`set.seed(20260507)` at the top of the replicate loop and uses
`parallel::mc.set.seed` inside `mclapply`. Per-replicate seeds
derive from the master seed by `+ rep_idx`.

**Data and code availability.** Per-replicate simulation
outputs are checkpointed at
`analysis/data/quick-sim/05-design-replicates.rds`; Morris ADEMP
cell-level summaries at the companion `05-design-summary.txt`.
Driver scripts are under
`analysis/scripts/nof1-design-sensitivity/` and
`analysis/scripts/quick-sim/05-design-prototype.R`. The
pre-registered ADEMP plan is at
`analysis/scripts/nof1-design-sensitivity/00-ademp-pre-registration.md`.
The manuscript source, paper-specific bibliography
(`references.bib`), and SIM CSL file are at
`analysis/report/05-nof1-design-sensitivity/`.

# References

<div id="refs"></div>
