---
title: "Latent-class and mixture-model formulations for
  biomarker-treatment interaction in N-of-1 trials"
author:
  - pmsimstats team
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    number_sections: true
    keep_tex: true
    highlight: tango
fontsize: 11pt
geometry: "left=3cm,right=5cm,top=2cm,bottom=2cm"
bibliography: references.bib
csl: statistics-in-medicine.csl
link-citations: true
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{lineno}
  - \linenumbers
  - \usepackage{setspace}
  - \setstretch{1.1}
knit: (function(input, encoding) source(normalizePath(file.path(dirname(input), '..', '..', '..', 'tools', 'stage-render.R')))$value(input, encoding))
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.path = '../../figures/04-',
  cache.path = '_cache/'
)
```

```{r footer-setup, echo = FALSE, results = 'asis'}
if (knitr::is_latex_output()) {
  src <- tryCatch(normalizePath(knitr::current_input()),
                  error = function(e) 'unknown')
  src_disp <- gsub('^/Users/zenn/Dropbox', '~/Dropbox', src)
  src_disp <- gsub('^/Users/zenn', '~', src_disp)
  src_tex <- gsub('([_~^&%#$])', '\\\\\\1', src_disp)
  ts <- format(Sys.time(), '%Y-%m-%d %H:%M %Z')
  cat(sprintf('\\renewcommand{\\rmdsource}{Source: \\texttt{%s}}\n',
              src_tex))
  cat(sprintf('\\renewcommand{\\rendertimestamp}{Rendered: %s}\n', ts))
}
```

# Abstract

**Background.** Aggregated N-of-1 clinical trials test
biomarker-treatment interactions through a fixed continuous-
heterogeneity regression in published practice. The psychometric
literature has developed mature finite-mixture and latent-class
machinery for representing discrete subpopulation structure that
the biostatistical literature on N-of-1 trials has not engaged.
The choice between continuous-heterogeneity and latent-class
encodings of the biomarker-moderation question is a substantive
scientific question rather than a parameterisation convenience,
but the field has not characterised when class-aware analyses
become competitive with continuous regression at trial-relevant
sample sizes.

**Methods.** We develop a five-component biomarker-moderation
spectrum (covariance-only, mean-and-covariance, location-scale,
heterogeneous random-effects, latent-class) bridging
continuous-heterogeneity and latent-class formulations, and we
catalogue the finite-mixture taxonomy (latent class analysis,
growth mixture models, factor mixture models, regression
mixtures) for application to within-subject longitudinal designs.
We commit a heterogeneous-random-slopes generative DGP and a
class-aware analysis pipeline (`lcmm::hlme` with gridsearch
initialisation), and run a 240-replicate-per-cell prototype
simulation at $N = 70$, gating slope $\in \{0, 1.0, 1.5\}$,
on a 4-core parallel implementation. Three test forms are
recorded per replicate: the linear-mixed `bm:Dbc` Wald, the
unconditional Wald on the lcmm gating coefficient, and a
BIC-conditional gating procedure.

**Results.** In a 240-replicate-per-cell pilot we find that the
unconditional Wald test on the latent-class gating coefficient
inflates Type I error to $0.197$ (95% CI $[0.145, 0.250]$) at
the null, decisively above nominal $0.05$ and consistent with
the @bauercurran2003 and @lubkemuthen2005 overextraction
predictions. The BIC-conditional gating procedure controls
Type I ($0.004$ under the null) but is severely under-powered
at trial-relevant $N$ (rejection rate $0.025$ at gating slope
$1.0$, $0.004$ at $1.5$). The linear-mixed `bm:Dbc` Wald test
holds nominal Type I and reaches $0.542$ power at gating slope
$1.5$, dominating both class-aware tests. MCSE on power is
approximately $0.030$ across the grid. These pilot estimates
precede the production five-study programme described in the
companion simulation plan, which has not yet been executed.

**Conclusions.** Within the 240-replicate pilot, at trial-
relevant sample sizes ($N = 70$, eight measurement timepoints),
neither the unconditional Wald-on-gating nor the BIC-conditional
gating test is competitive with the linear-mixed baseline under
the heterogeneous-random-slopes DGP. The latent-class encoding
appears to require a regime of large class separation for
class-aware tests to gain power; the pre-registered production
programme (five studies, $N \in \{35, 70, 100\}$, wider gating-
slope grid) has not yet been executed and is required to map
the regime where the BIC-conditional procedure becomes
competitive.

# Introduction

## The aggregated N-of-1 trial and the biomarker-moderation question

The aggregated N-of-1 trial design is a class of clinical trials
in which each enrolled participant cycles between active treatment
and control across multiple within-person blocks, with
within-person comparisons pooled across participants for
population-level inference [@guyatt1986; @guyatt1988; @lillie2011;
@senn2018]. The design preserves the within-subject contrast that
gives single-participant N-of-1 trials their inferential strength
while accumulating the statistical efficiency of a multi-
participant trial. For chronic conditions whose treatment response
is reversible on a useful timescale, the design yields more
information per participant per unit time than a parallel-group
randomised trial of equivalent total sample size [@mirza2017]. It
has been deployed in chronic pain [@yelland2009], inflammatory
bowel disease, neuropsychiatric symptoms in post-traumatic stress
disorder [@hendrickson2020], and a growing portfolio of rare and
orphan disease applications where parallel-group trials are
infeasible [@lillie2011; @schork2015].

Within this design class, a question of increasing scientific
weight is whether treatment response is moderated by a measured
pre-treatment biomarker. A biomarker that predicts which
participants respond to which treatments would let clinicians
stratify treatment decisions at the point of care, and the formal
framework for such inference is laid out in the PATH statement
[@kent2020path]. The corresponding statistical estimand is the
biomarker-by-treatment interaction term in the within-participant
outcome model. Power to detect this interaction depends on the
magnitude of moderation, the precision of the within-participant
contrasts, the trial design (open-label, crossover, hybrid), and
any time-dependent leakage between treatment phases, conventionally
termed 'carryover', that erodes the cleanness of the within-subject
contrasts.

Recent methodological work has documented substantial power losses
for biomarker-by-treatment interaction inference under realistic
carryover, with the loss varying sharply by how the moderation
effect is encoded in the data-generating process (DGP)
[@hendrickson2020]. The present paper takes up a
question that the prior work raised but did not resolve: how
should biomarker moderation be represented in the DGP itself? Two
formal answers, drawn from a long history of work outside
biostatistics, dominate the available options.

## Two formal answers: continuous heterogeneity and latent-class structure

The first answer represents biomarker moderation as a graded
continuous effect: every participant has some degree of treatment
response, the response is a smooth (typically linear) function of
the biomarker, and population heterogeneity is captured by a single
multivariate distribution with an interaction parameter. This is
the dominant tradition in clinical biostatistics, where
heterogeneity of treatment effect is routinely modelled via random
slopes on the treatment indicator and via interaction terms in
fixed-effects regression. In its most economical form, the
moderation enters the conditional mean of the outcome as
$\beta_{bm:D} \cdot B \cdot D$, where $B$ is the biomarker and $D$
is the (possibly continuous-decay) treatment indicator.

The second answer represents biomarker moderation as a discrete
subpopulation structure: participants belong to one of a small
number of unobserved classes (most commonly 'responder' and
'non-responder'), the classes have qualitatively distinct
treatment-response curves, and the biomarker predicts class
membership without determining it. The corresponding generative
form is a finite mixture in which class membership is a Bernoulli
or multinomial draw with mixing probabilities that depend on the
biomarker, and the within-class outcome model is class-specific.
The marginal moderation observed in the data emerges from the
mixture rather than from a smooth conditional-mean function.

These two answers correspond to genuinely different biological
hypotheses. The continuous-heterogeneity view is consistent with a
mechanism in which the biomarker reflects a graded physiological
substrate (receptor density, enzymatic activity, baseline trait
severity) that scales drug effect across the entire population.
The latent-class view is consistent with a mechanism in which
qualitatively distinct phenotypes exist (a metabolising versus
non-metabolising allelic variant; a present versus absent target
receptor) and the biomarker is an imperfect indicator of phenotype.
The two views can produce identical marginal biomarker-outcome
associations at a single timepoint and diverge sharply once the
longitudinal structure of an N-of-1 trial is engaged, because
carryover and within-person serial dependence interact differently
with continuous and class-structured heterogeneity. This is the
substantive point at which methodology imported from outside
biostatistics becomes useful: both traditions have long histories
of formalisation and identifiability analysis in their respective
fields, and the corresponding tools are mature.

## Psychometric origins

The continuous-trait view in formal statistical work originates
with Spearman's two-factor theory of intelligence [@spearman1904],
in which performance on cognitive tests was modelled as the sum of
a single general factor and test-specific factors. Spearman's
machinery was elaborated through the 1930s and 1940s by Thurstone,
Hotelling, Lawley, and others into the modern factor-analytic
tradition, in which observed indicator variables are written as
linear combinations of unobserved continuous factors plus residual
noise [@thurstone1947; @lawleymaxwell1971]. Within this tradition,
between-person heterogeneity in any latent construct is by
assumption continuous, and individual differences are summarised
by factor scores along one or more dimensions.

The discrete-class view emerged from a parallel but
methodologically distinct literature beginning with Lazarsfeld's
work on latent structure analysis in the 1950s [@lazarsfeld1950;
@lazarsfeldhenry1968]. Lazarsfeld observed that in attitudinal and
survey data, the joint distribution of categorical responses to
multiple items frequently displayed structure inconsistent with a
single continuous latent trait, and instead suggested that
respondents belonged to one of several qualitatively distinct
latent classes within which item responses were locally
independent. This 'local independence' formulation is the founding
assumption of latent class analysis (LCA). @goodman1974 placed the
LCA model on rigorous statistical footing, deriving identifiability
conditions and maximum-likelihood estimation procedures for the
binary-indicator case in a Biometrika paper that remains a
standard reference. Through the 1970s and 1980s, the LCA framework
expanded to polytomous and continuous indicators, the latter
widely known as latent profile analysis [@mccutcheon1987], to
multilevel and longitudinal extensions [@vermunt2003], and to
mixtures with covariate-dependent class probabilities
[@daytonmacready1988; @bandeenroche1997]. @mclachlanpeel2000
unified much of this material into a comprehensive monograph on
finite mixture models that is the standard entry point for
quantitative researchers from outside the discipline.

The defining tension in psychometric methodology has been whether
a given pattern of observed heterogeneity is best understood as
continuous or as discrete. @bartholomew1987 framed this as the
choice between 'graded' and 'unfolding' models of individual
differences and argued that the data themselves rarely settle the
question without strong prior assumptions. The empirical literature
has accumulated examples in both directions: cognitive ability is
generally well-described by continuous factor models, while certain
forms of psychopathology display latent-class structure consistent
with qualitatively distinct subtypes. Subsequent methodological
work, beginning with @yung1997 and @arminger1999, developed hybrid
frameworks (factor mixture models) in which discrete classes
coexist with within-class continuous factors. We return to these
frameworks below.

## Econometric and marketing-science development

A largely independent strand of work on finite mixtures developed
in econometrics and quantitative marketing through the 1980s and
1990s. The motivating problems differed from psychometrics: rather
than measurement of cognitive or psychological constructs, the
econometric concern was unobserved heterogeneity in
micro-econometric models (consumer choice, labour-market
participation, firm productivity) where standard regression
assumptions of identical agents were known to fail. @heckmansinger1984,
in an Econometrica paper on duration models, demonstrated that
estimates of structural parameters in survival-style models could
be highly sensitive to the assumed functional form of the
unobserved heterogeneity distribution, and recommended a
non-parametric maximum-likelihood approach in which the
heterogeneity distribution itself was estimated as a discrete
mixture with unknown support points and weights. Their
non-parametric maximum likelihood estimator anticipated much of
what would later become standard finite-mixture methodology in
biostatistics, and @lindsay1995 placed the broader theory of
mixture likelihoods on a unified mathematical footing.

In quantitative marketing, the parallel concern was customer
segmentation. A population of consumers might respond to a
marketing intervention differently depending on which unobserved
segment they belonged to, and the firm needed both to estimate
segment-specific response functions and to predict segment
membership for new consumers. @desarbocron1988 formalised this as
'clusterwise linear regression', in which observations were
partitioned into clusters with cluster-specific regression
coefficients. @wedeldesarbo1995 generalised the approach to
generalised linear models. @jedidi1997 extended it to structural
equation models with unobserved segments. Within this
econometric-marketing literature, the gating function (the model
for the probability of segment membership as a function of
covariates) became a first-class object, and methodological work
focused on whether the gating model could share covariates with
the within-segment regression model. @jacobs1991, working
contemporaneously in machine learning, formalised an essentially
identical structure under the name 'mixtures of experts', which
provided much of the now-standard vocabulary.

Two themes from the econometric strand are particularly relevant
to the N-of-1 trial setting. First, the econometric tradition
treated unobserved heterogeneity as a *nuisance* in many
applications, and was correspondingly attentive to whether the
heterogeneity distribution was identified, weakly identified, or
over-fit. The identifiability concerns raised by @heckmansinger1984
and elaborated by @lindsay1995 translate directly into N-of-1
settings where sample sizes are small and the heterogeneity is the
principal object of inference rather than a nuisance.
Second, the marketing-science focus on covariate-dependent gating
functions corresponds exactly to the
biomarker-as-class-membership-predictor structure that motivates
the present paper. The biomarker-treatment moderation question,
translated into econometric vocabulary, is whether the gating
function depends on a single observed covariate (the biomarker)
and whether the within-class regression coefficient on treatment
differs across classes.

## Convergence on hybrid frameworks

By the late 1990s, both the psychometric and the econometric
strands had converged on a hybrid formulation: combine continuous
within-class factor structure with discrete between-class
structure. @yung1997, in Psychometrika, developed the factor
mixture model as a finite mixture of confirmatory factor models,
where each class had its own factor-loading and residual covariance
structure. @arminger1999 placed factor mixture models in the
broader Mplus-style structural equation framework.
@lubkemuthen2005 conducted the first systematic Monte Carlo
evaluation of factor mixture models and clarified the conditions
under which the discrete-class component was empirically
distinguishable from continuous heterogeneity. @clark2013 surveyed
the modelling strategies and presented an applied analysis of
psychiatric comorbidity.

In parallel, the longitudinal extension of these ideas yielded the
growth mixture model [@muthenshedden1999; @muthen2001; @muthen2004]
and the closely related group-based trajectory model [@nagin1999;
@nagin2005]. Both posit that a longitudinal outcome trajectory is
generated by a finite mixture of class-specific trajectories.
Growth mixture models permit within-class random-effects variation
around the class-specific mean trajectory; group-based trajectory
models constrain within-class variance to zero, so each class has
a single deterministic trajectory. Nagin's group-based approach
was developed primarily in criminology to characterise
developmental trajectories of antisocial behaviour, and has been
widely applied in epidemiology, public health, and clinical
research where trajectory heterogeneity is the principal
scientific question.

The methodological optimism around growth and factor mixture
models during the early 2000s was tempered by a series of papers
documenting overextraction hazards. @bauercurran2003, in
Psychological Methods, demonstrated through simulation that growth
mixture models fitted to data generated from a single non-gaussian
distribution can spuriously identify multiple latent classes; the
apparent classes are artefacts of distributional non-normality
rather than evidence of substantive subpopulation structure.
@bauer2007 elaborated on the implications for substantive
psychological research and recommended caution in interpreting
class structure without independent validation. @lubkemuthen2005
documented the analogous concerns for factor mixture models,
identifying the joint conditions on sample size, class separation,
and class-specific parameter differences under which class
structure is recoverable. @nylund2007 systematised the
class-enumeration question through Monte Carlo evaluation of
information criteria and likelihood-ratio bootstrap procedures.
The cumulative message of this literature is that mixture models
can extract apparent class structure where none exists, and that
distinguishing genuine from spurious class structure requires
either substantial samples (typically several hundred to several
thousand participants) or strong prior information.

## Translation into longitudinal biostatistics

Mixture and latent-class methods entered biostatistical practice
through several complementary channels. @verbekelesaffre1996, in
the Journal of the American Statistical Association, proposed a
linear mixed-effects model in which the random-effects population
was itself a finite mixture of multivariate normals. Their
motivating application was longitudinal HIV-infection data in
which subgroups of patients displayed qualitatively different CD4
trajectories. The model is essentially a growth mixture model in
biostatistical clothing, with the standard random-intercept-and-
slope linear mixed model augmented by a discrete latent class
indicator. @zhangdavidian2001 extended the framework to allow more
flexible random-effects distributions via semi-parametric mixture
specifications.

Proust-Lima and colleagues developed a more comprehensive software
framework with the `lcmm` R package [@proustlima2017], which
implements joint latent-class mixed models for longitudinal
continuous and ordinal outcomes, with optional joint modelling of
survival outcomes via shared latent class structure. The `lcmm`
framework has been deployed in cognitive ageing research,
Alzheimer's disease progression modelling, and clinical-trial
subgroup discovery. The `flexmix` package in R [@leisch2004;
@grunleisch2008] provides a more general infrastructure for
fitting mixtures of regression-style models with user-supplied
component specifications, and has been used in pharmacokinetic and
pharmacodynamic modelling where individual-level variation in
dose-response curves motivates a mixture structure.

Despite this translation activity, mixture and latent-class methods
have not been systematically applied to aggregated N-of-1 trial
designs. The published N-of-1 methods literature is dominated by
within-participant effect estimation [@zucker1997; @lillie2011],
Bayesian hierarchical models for participant-specific effects
[@zucker2010], and aggregated-evidence meta-analytic frameworks
[@schork2015; @senn2018]. The biomarker-moderation question has
received occasional treatment under the heading of
personalised-medicine power calculation [@gabler2011;
@hendrickson2020], but in each case the moderation has been
encoded as a continuous interaction term in a fixed-effects
analysis, with no parallel treatment of the latent-class
formulation. The mixture-formulation
question is to our knowledge unaddressed in the existing N-of-1
methodology literature.

## Why this matters for N-of-1 designs specifically

Three features of the aggregated N-of-1 design make the
continuous-versus-discrete moderation question particularly
consequential for design and analysis decisions.

First, the within-participant longitudinal structure of an N-of-1
trial provides repeated measurements at multiple drug states
(on-drug, off-drug, in carryover). Under a continuous-heterogeneity
DGP, the moderation signal is encoded in how the within-participant
treatment contrast varies with the biomarker. Under a latent-class
DGP, the moderation signal is encoded in class membership, which
is fixed across all measurements within a participant and therefore
in principle recoverable from the full trajectory even when
individual on-drug contrasts are noisy. The two encodings imply
different efficient analysis strategies and different power
profiles under carryover [@hendrickson2020].

Second, N-of-1 trials are typically conducted at sample sizes of
$N \in [30, 150]$, well below the sample sizes at which mixture-
model identifiability has been validated in the psychometric
literature. The Bauer-Curran-Lubke-Muthén identifiability cautions
therefore apply with full force, and the analysis-strategy
implications are non-trivial: even if a latent-class DGP is the
biologically correct generative model, mixture analysis at
trial-relevant sample sizes may be either underpowered or
unidentified, and a misspecified single-component analysis may be
the more reliable inferential tool despite its bias. Resolving
this trade-off is in part an empirical question that the existing
psychometric simulation literature has not addressed at N-of-1-
relevant sample sizes.

Third, carryover, which enters the analysis as a continuous decay
in the effective drug indicator $D_{it} = \exp(-\lambda t_{sd,it})$,
interacts differently with the two moderation encodings. In the
continuous-heterogeneity formulation, carryover erodes the
moderation signal directly because the moderation enters as a
product with $D_{it}$. In the latent-class formulation, carryover
erodes the within-class drug contrast but does not directly
attenuate class membership; if class membership can be recovered
from the full trajectory, some moderation signal can survive
carryover that would be lost under the continuous-heterogeneity
analysis. The two formulations therefore predict different
power-loss profiles under carryover, and this difference is
empirically large in the simulation literature accumulated by the
companion carryover-sensitivity work.

## Aims and structure of the paper

This paper has four aims. First, to formalise the latent-class and
finite-mixture formulations of biomarker-treatment interaction in
aggregated N-of-1 trials, with explicit identifiability conditions
specific to the trial-relevant sample-size regime. Second, to
characterise estimation strategies (expectation-maximisation,
Bayesian samplers, candidate R packages) and the diagnostics
required to distinguish genuine from spurious class structure at
small $N$. Third, to report a factorial simulation study that
benchmarks mixture-DGP and continuous-DGP power against the
standard linear-mixed-model analysis and a class-aware analysis
under realistic carryover. Fourth, to apply the developed methods
to the prazosin-PTSD biomarker dataset of @hendrickson2020 and to
compare conclusions with those obtained under the
continuous-moderation analysis reported in the original paper.

The remainder of the paper is organised as follows. A notation
section fixes terminology that is used throughout. The latent-
class generative form, its MVN second-moment approximation, and the
regimes in which the approximation breaks down are developed in
three subsequent sections. The implications for power under
carryover and the identifiability constraints that bound mixture
analysis at trial-relevant sample sizes are addressed next. A
finite-mixture taxonomy section locates the relevant model
families (LCA, GMM, FMM, regression mixtures); a biomarker-
moderation spectrum section organises the available DGP encodings
moment by moment. Implications for analysis and design, and
software for mixture analyses, are then drawn together. The
discussion section addresses three open methodological questions
and references the five-study simulation programme described in
the companion simulation plan, which constitutes the empirical
core of the present line of work and is the next deliverable in
the programme.


# Notation

Throughout, $B_i$ denotes the pre-treatment biomarker value for
participant $i$. $BR_{it}$ denotes the biological-response component
of outcome for participant $i$ at time $t$, decomposed in the
Hendrickson framework into three latent response factors: $BR_{it}$
itself, a time-variant component $TV_{it}$, and a placebo-belief
component $PB_{it}$. $D_{it}$ denotes drug state at time $t$ in
the analysis-side continuous-decay form
$D_{it} = \exp(-\lambda t_{sd,it})$ during off-drug periods and
$D_{it} = 1$ during on-drug periods, with $t_{sd,it}$ the time
since drug discontinuation and $\lambda$ the decay rate. $Z_i$
denotes an unobserved class label. The half-life of carryover is
$t_{1/2} = \ln 2 / \lambda$ and is allowed to differ between the
DGP and the analysis model.

The principal psychometric and econometric terms used below are:

- **Latent class.** An unobserved categorical subgroup, equivalent
  in trial vocabulary to an unobserved responder/non-responder
  stratum.
- **Latent factor.** An unobserved continuous trait governing
  covariation among observed variables, equivalent to a frailty
  or to a random intercept in a linear mixed model.
- **Class-conditional distribution $f_z$.** The outcome
  distribution within a single subgroup.
- **Mixing proportion $\pi$.** The prevalence of a given
  subgroup in the population.
- **Measurement invariance.** Whether a model applies identically
  across subpopulations.
- **Growth mixture model (GMM).** A longitudinal mixed model in
  which the population of random effects is itself a finite mixture.
- **Factor mixture model (FMM).** A hybrid model with discrete
  classes at the upper level and within-class continuous
  heterogeneity at the lower level.
- **Regression mixture / mixture of experts.** A finite mixture in
  which the gating function (class-membership probability) depends
  on covariates and within-class regressions are class-specific.
- **EM algorithm.** Expectation-maximisation applied to recover
  unobserved class labels jointly with class-conditional model
  parameters.

# Faithful latent-class data-generating processes

If a candidate predictive biomarker indexes an unobserved
responder/non-responder dichotomy, the mechanistically faithful
data-generating process (DGP) is a finite mixture:

$$
Z_i \mid B_i \sim \mathrm{Bernoulli}\bigl(\pi(B_i)\bigr), \qquad
BR_{it} \mid Z_i = z \sim f_z(t, D_{it}),
$$

where $Z_i$ is the latent class indicator,
$\pi(B_i) = \Pr(Z_i = 1 \mid B_i)$ is the class-membership
probability (typically a logistic function of the biomarker), and
the class-conditional response distributions $f_0$ and $f_1$ differ
in their drug-response parameters: peak response, onset rate, or
both. Within-class temporal structure follows the same modGompertz
form used in single-class formulations, with class-specific
parameters substituted.

Two consequences of this DGP form are central to what follows.
First, the class label $Z_i$ is fixed across all measurements
within a participant; it is invariant to drug state and to time.
Second, the class-conditional mean trajectory does depend on
$D_{it}$: under full carryover washout ($D_{it} \to 0$) the two
class-conditional means converge if the responder/non-responder
distinction is purely in drug response. Class-label invariance
therefore preserves participant-level information about the
moderation signal that an analysis with access to $Z_i$ could
exploit, even where instantaneous between-class mean separation
in $BR$ is fully attenuated. The MVN-covariance encoding of
biomarker moderation, in contrast, has no such latent-label
information to exploit: the covariance is the entire signal, and
its erosion under carryover is direct.

# MVN approximation to mixture-induced moderation

A two-component mixture with class-dependent means and common
within-class covariance produces a marginal joint distribution of
$(B_i, BR_{it})$ whose first two moments are reproduced, to
leading order, by a single multivariate normal with
$\mathrm{Cor}(B, BR) = c_{bm}$. Under within-class independence
between $B_i$ and $BR_{it}$,

$$
c_{bm}^2 \;=\;
\frac{\pi(1-\pi)(\mu_B^1-\mu_B^0)^2}{\mathrm{Var}(B)}
\;\cdot\;
\frac{\pi(1-\pi)(\mu_{BR}^1-\mu_{BR}^0)^2}{\mathrm{Var}(BR)},
$$

so that $c_{bm}^2$ is the product of the between-class variance
fractions in $B$ and in $BR$. The marginal correlation that an
MVN can reproduce is therefore bounded by the geometric mean of
the two intra-class correlations; if either class separation is
small, the achievable marginal $c_{bm}$ is small even when class
structure is otherwise well-defined.

Under mild regularity conditions (adequate class overlap and
mixing proportions $\pi$ bounded away from 0 and 1), the
mixture's second-moment structure is well-approximated by a
single MVN with differential correlation across drug states:
elevated correlation where class-conditional means separate
(on-drug) and shrinkage towards zero where they coincide
(off-drug after carryover washout). For power calculations driven
by t-tests on $\beta_{bm:D}$ in a linear mixed-effects model, this
approximation is adequate within the regularity conditions and
captures the covariance consequences of latent-class structure
without committing to a discrete generative form.

# Failure regimes of the MVN approximation

The MVN approximation is misleading in three identifiable
regimes, each with a distinct empirical signature.

- **Bimodal responses.** When $f_0$ and $f_1$ are well-separated
  (non-responders show near-zero drug effect and responders show a
  large effect), the marginal distribution of $BR$ is bimodal: a
  histogram of on-drug response would show two peaks rather than a
  single peak with gaussian scatter around it. No MVN matches the
  tails of a bimodal distribution, and power under the true mixture
  DGP will exceed the MVN-LMM prediction because a well-specified
  mixture analysis can exploit the bimodality. The pattern occurs
  in clinical settings where a subpopulation genuinely does not
  respond to a treatment, as in certain targeted oncology
  indications.
- **Strong class-membership gradient.** When $\pi(B_i)$ is steep,
  approaching a step function at a biomarker threshold, the
  biomarker behaves nearly deterministically as a class label. A
  thresholded mean-moderation specification (a thresholded
  Architecture-A variant) then becomes the better approximation,
  because the biomarker-response relationship is dominated by its
  mean-structure component. The corresponding clinical scenario is
  a diagnostic biomarker with near-perfect sensitivity and
  specificity for the responsive phenotype.
- **Class-varying covariance.** If autocorrelation, residual
  variance, or placebo-response parameters differ between classes,
  no single-component MVN can represent the generative structure.
  Analysis models assuming constant covariance are misspecified
  regardless of whether they encode the biomarker interaction in
  the mean or in the covariance. This is the latent-class analogue
  of heteroscedastic residual variance addressed in standard mixed-
  effects practice with `varIdent` or `weights` specifications,
  except that the stratifying variable is unobserved.

# Implications for power under carryover

The substantial covariance-only power loss under carryover (40-60%
across designs in @hendrickson2020) reflects erosion of a second-
moment signal. A mixture-model analysis fitted to mixture-DGP data
can in principle recover additional power by estimating class
membership directly, because the class label is itself
carryover-invariant. The gap between the covariance-only carryover-
sensitive power estimate and the mixture-faithful power estimate
upper-bounds what a mixture analysis could recover at the same
carryover level; whether this bound is tight in the trial-relevant
parameter regime is an empirical question not addressed by existing
simulation infrastructure. The bound is asymptotic: at finite $N$
in the trial-relevant range, finite-sample bias in mixture
estimation can drive the realised gap below the asymptotic upper
bound or, in adverse identifiability regimes, reverse its sign.

A focused factorial simulation isolating second-moment
approximation error from analysis-model mis-specification is
described in the companion simulation plan (Study 1: mixture-vs-MVN
factorial). Briefly, the factorial crosses two DGPs (mixture and
MVN) with two analyses (linear mixed model and joint latent-class
mixed model). The difference between mixture-DGP-with-MVN-analysis
and mixture-DGP-with-mixture-analysis quantifies the power
recoverable by a faithful analysis model; the difference between
the two MVN-analysis cells quantifies the approximation error
inherent to the second-moment encoding. A null check on
MVN-DGP-with-mixture-analysis verifies that the mixture model does
not extract spurious power on data for which no latent class
structure exists.

# Identifiability at trial-relevant sample sizes

Mixture modelling is attractive on faithfulness grounds but carries
identifiability costs that are acute in N-of-1 settings. The EM
algorithm for finite mixtures of mixed-effects models requires
either informative priors on class-membership probabilities or
strong separation between class-conditional response curves;
neither is assured in a trial designed to detect a moderation
effect whose existence is itself uncertain. The standard diagnostic
for mixture identifiability, the entropy of posterior class-
membership probabilities, requires substantial sample sizes
($N$ several hundred to several thousand) to stabilise at values
that support confident class extraction in published psychometric
applications [@bauercurran2003; @bauer2007; @nylund2007].
Aggregated N-of-1 trials of the size contemplated in current
biomarker applications, $N \in [30, 150]$, are well below this
range.

The methodological implication is that mixture analysis at trial-
relevant sample sizes may be either underpowered or unidentified,
even when the underlying DGP is genuinely mixture-structured. The
question is therefore not whether mixture formulations are more
faithful, but whether the fidelity gain is recoverable at realistic
sample sizes, or whether analysis-strategy mitigations available
within standard mixed-effects practice (restricted analysis,
weighting, within-subject contrasts; see @hendrickson2020) deliver
a better power-per-complexity trade. These mitigations have better-
understood inferential properties at the relevant sample sizes.
Resolving the mixture-versus-mitigation question empirically
requires the simulation programme described in the companion
simulation plan (Study 5 in particular addresses small-$N$
identifiability and type I error explicitly).

# Finite-mixture taxonomy

Four families of finite mixture model from the psychometric and
econometric literatures bear directly on the biomarker-treatment
interaction problem. The discussion below introduces each family,
its origin, and its specific relevance to the N-of-1 trial setting.

## Latent class and latent profile analysis

Latent class analysis (LCA) and latent profile analysis (LPA) are
mixture models for cross-sectional data. Each participant belongs
to one of a small number of unobserved classes; the observed
variables have class-specific marginal distributions. LCA uses
categorical indicators; LPA uses continuous indicators. A mixture
fitted to baseline biomarkers alone returns posterior class-
membership probabilities that can be used as covariates in
subsequent outcome analysis, the two-stage analytic strategy
standard in psychometric applications. In the biomarker-trial
context, an LPA-style model posits two or more responder classes
with class-specific marginal distributions of on-drug and off-drug
response, and estimates class-membership probabilities from the
observed response trajectories rather than from baseline
covariates alone. Classical references include @lazarsfeldhenry1968,
@goodman1974, @clogg1995, and @mclachlanpeel2000; the Vermunt and
Magidson Latent GOLD software and the `poLCA` R package implement
the standard EM estimation routines.

## Growth mixture models and group-based trajectory models

Growth mixture models (GMM) [@muthenshedden1999; @muthen2001;
@muthen2004] extend LPA to longitudinal data by specifying class-
specific growth (or, in the trial setting, drug-response) curves
with within-class random-effects variation. A GMM is a longitudinal
linear mixed model in which the population distribution of random
effects is a mixture of multivariate normals, each corresponding to
a distinct trajectory class. Group-based trajectory models
[@nagin1999; @nagin2005] are the degenerate case in which within-
class random-effect variances are set to zero, so that participants
within a class share an identical trajectory up to observation-
level residual noise.

GMM is the closest psychometric analogue of the latent-class
generative form developed above: each participant belongs to a
latent class with its own drug-response trajectory, and the class
label is drug-state-invariant. A GMM that uses the biomarker as a
class-membership predictor is essentially the faithful generative
model written down earlier in this paper. @bauercurran2003 and
@bauer2007 document the identifiability and overextraction hazards
that accompany GMM estimation at realistic sample sizes, showing
in particular that GMM can extract apparent class structure from
data generated by single-class non-gaussian processes. These
cautions apply with full force to aggregated N-of-1 trials, where
participant counts are typically well below the sample sizes at
which GMM class-count recovery has been validated.

## Factor mixture models

Factor mixture models (FMM) [@yung1997; @arminger1999;
@lubkemuthen2005] combine continuous latent
factors with discrete latent classes, yielding a generative
structure in which within-class variation is governed by a factor
model and between-class variation is governed by class-specific
means and (optionally) class-specific factor loadings or residual
covariances. The FMM framework subsumes both the pure continuous
(single-class factor model) and the pure discrete (LCA with no
within-class structure) cases as limits.

FMMs are two-level hierarchies. At the upper level, participants
are assigned to one of a small number of classes (as in LCA). At
the lower level, each class has its own within-class random-effects
structure. The biological hypothesis the FMM encodes is that there
are genuinely distinct responder subtypes and that, within subtype,
additional continuous heterogeneity is driven by covariates or
unmeasured individual differences. For the biomarker-moderation
problem, FMM provides a principled way to represent a biomarker as
both a noisy class indicator (via the gating function) and a
continuous modulator of within-class response (via factor loadings
on the response factor). The hybrid representation matches more
biological scenarios than either pure LCA or pure factor analysis,
at the cost of correspondingly higher estimation requirements.

## Regression mixtures and mixtures of experts

Regression mixtures [@desarbocron1988; @wedeldesarbo1995;
@jedidi1997] and the closely related mixture-of-experts architecture
[@jacobs1991] allow class-specific regression coefficients with
class-membership probabilities (the gating function) that depend on
covariates. An ordinary linear model with a treatment-by-biomarker
interaction posits a single regression surface in which the
treatment slope varies smoothly with the biomarker. A regression
mixture instead posits two or more distinct regression surfaces,
one per class, with a separate covariate-dependent model governing
which surface applies to which participant. The two-stage structure
(classification by the gating model, then prediction by the within-
class regression) parallels the two-stage structure of causal-
inference methods that use propensity scores to stratify before
estimating within-stratum treatment effects.

In the trial-simulation idiom, regression mixtures correspond to a
DGP in which the biomarker governs both the probability of being a
responder and the magnitude of the within-responder drug effect.
`flexmix` in R [@leisch2004; @grunleisch2008] implements this family
with user-definable component models, including mixed-effects
components suitable for repeated-measures data.

# The biomarker-moderation spectrum

The covariance-only formulation of @hendrickson2020 specifies a
biomarker-dependent covariance structure with population means
held constant across biomarker values. Locating this formulation
within the broader spectrum of biomarker-moderated DGPs is useful
because the moment-by-moment organisation of the spectrum
(biomarker moderating the mean, the covariance, both, or higher
moments) clarifies which features of the carryover-induced power
loss are properties of the specification itself rather than of the
underlying scientific question.

## Covariance-only moderation

The covariance-only specification holds $E[BR_{it}]$ fixed across
biomarker strata and encodes the interaction entirely in
$\mathrm{Cov}(B_i, BR_{it} \mid D_{it})$. High- and low-biomarker
participants have the same expected drug response at any given
time but differ in the covariance between biomarker and response,
specifically in how strongly the biomarker co-varies with response
during on-drug versus off-drug periods.

This is an unusual restriction within the psychometric mixture
tradition. Most standard FMM and GMM parameterisations permit
class-specific means at a minimum, and the means-constant-
covariance-varies case is typically encountered only as a
constrained submodel used for testing measurement-invariance
hypotheses [@meredith1993; @widamanreise1997]. The corresponding
biostatistical model has two subgroups with identical expected
outcomes but differing residual variance or within-subject
autocorrelation; such models are occasionally fitted as sensitivity
analyses but rarely assumed as the primary DGP. The covariance-only
encoding's appeal lies in its tractability for power simulation
under multivariate normality: the entire DGP reduces to drawing
from a single MVN with a constructed covariance. Its cost is that
it does not correspond cleanly to any natural generative
mechanism.

## Combined mean and covariance moderation

@arminger1999 develop finite mixtures of conditional mean- and
covariance-structure models in which both the expected-response
vector and the residual covariance matrix differ between classes
and may depend on observed covariates within class. This is the
general psychometric form: it reduces to a mean-moderation model
in the limit of a single class with covariate-dependent mean only,
and to the covariance-only encoding in the limit of class-invariant
means with covariate-dependent covariance only.

In the combined form, responders and non-responders differ both in
typical drug response (mean moderation) and in how noisily that
response is expressed within class (covariance moderation). Both
channels carry information about class membership: the mean
channel via $E[BR \mid B]$, the covariance channel via
$\mathrm{Var}[BR \mid B]$. Under carryover, the mean channel is
comparatively robust because class-conditional means need not
converge off-drug; the covariance channel is eroded by construction
because off-drug covariance is attenuated. A combined-specification
simulation should therefore exhibit power loss intermediate between
the pure-mean and pure-covariance extremes; this is plausibly the
most biologically faithful regime for biomarker-guided
psychiatric trials. @verbekelesaffre1996 provide the analogous
treatment for linear mixed-effects models with heterogeneity in
the random-effects population, the more direct generalisation of
the repeated-measures structure used in N-of-1 trial simulation.

## Location-scale moderation

Generalised additive models for location, scale, and shape (GAMLSS)
[@gamlss2005] and the related distributional regression tradition
[@klein2015] allow the biomarker to
enter multiple parameters of the outcome distribution simultaneously
through separate link functions. In plain terms, a standard
regression models the conditional mean of an outcome as a function
of covariates; a GAMLSS-style distributional regression models the
conditional mean, the conditional variance, and if appropriate the
conditional skewness and kurtosis, each as a separate function of
covariates. The familiar special case is heteroscedastic linear
regression with covariate-dependent residual variance; GAMLSS
generalises this to arbitrary distributional shape.

In the biomarker-trial context, a GAMLSS-based DGP would allow
high-biomarker participants to have both a shifted mean drug
response (a location effect, which is Architecture A) and altered
response variability (a scale effect, which is analogous to
Architecture B's covariance moderation but operating on marginal
rather than joint moments), without requiring discrete class
structure. The GAMLSS framework is thus the continuous-heterogeneity
analogue of FMM and occupies an intermediate position between pure
Architecture A (mean only) and pure Architecture B (covariance
only). This intermediate position is attractive for sensitivity
analysis because it does not require committing to a specific
number of latent classes.

## Heterogeneous random-effects models

@verbekelesaffre1996, @zhangdavidian2001, and @proustlima2017
develop linear mixed-effects models in
which the random-effects distribution is itself a finite mixture,
allowing class-specific random-intercept and random-slope
distributions. The biostatistical setting this extends is the
standard linear mixed model with random intercept and random
treatment slope, familiar from any longitudinal trial analysis
using `nlme::lme` or `lme4::lmer`. In the standard model, the
random treatment slope is drawn from a single normal distribution
with mean equal to the average treatment effect and variance
capturing individual heterogeneity. In the heterogeneous
random-effects extension, that single distribution is replaced by
a mixture: a fraction of participants are drawn from a 'responder'
distribution with a large mean treatment slope and another fraction
from a 'non-responder' distribution with a near-zero mean treatment
slope. The biomarker predicts which mixture component a participant
comes from.

Applied to aggregated N-of-1 trial data, this framework represents
responder heterogeneity directly at the participant-specific random
treatment effect, which is arguably the most clinically transparent
representation available: responders and non-responders have
distinct random-slope distributions on the treatment indicator, and
the biomarker predicts class membership rather than entering the
analysis-model mean structure at the fixed-effect level. The `lcmm`
R package [@proustlima2017] implements this family and is the
closest available tooling for the longitudinal aggregated N-of-1
use case: it accepts standard longitudinal mixed-model syntax, fits
a mixture on the random-effects distribution, and returns posterior
class probabilities for each participant. For an extension of an
existing aggregated N-of-1 simulation framework, `lcmm` is the
lowest-friction entry point.

# Implications for analysis and design

Three implications follow from locating the covariance-only
encoding within the broader spectrum.

First, the covariance-only parameterisation is the most restrictive
operational case of a much richer family. The substantial power
loss under carryover documented in @hendrickson2020 is a specific
property of this restrictive case and need not generalise to the
mean-plus-covariance or heterogeneous-random-effects variants that
are standard in psychometric and biostatistical practice. A
sensitivity analysis based solely on the covariance-only DGP
risks overstating the carryover-induced power loss that a biomarker-
guided aggregated N-of-1 trial should expect, because the
covariance-only DGP places the entire biomarker-outcome signal in
a channel that carryover erodes by construction. If a portion of
the signal flows through the mean structure, as in pure mean
moderation or in the combined specification of @arminger1999, that
component survives carryover and overall power loss is correspondingly
attenuated. A complete sensitivity analysis should report power
under at least three points on the spectrum: pure mean moderation,
pure covariance moderation, and a combined specification.

Second, the identifiability concerns documented in the GMM and
FMM literatures [@bauercurran2003; @bauer2007; @lubkemuthen2005;
@nylund2007] bound what can be asked of mixture-based analysis
models at N-of-1 trial sample sizes. Published FMM applications
typically involve several hundred to several thousand
participants; biomarker-trial applications with $N \in [30, 150]$
are an order of magnitude smaller. Analysis plans built around
fitting `lcmm` or `flexmix` mixture models directly to trial data
are unlikely to yield reliable class-membership inference at
trial-relevant sample sizes and should not be relied on as the
primary analysis. Mixture analysis nonetheless retains value as a
sensitivity check (does class-structured residual pattern emerge
from the pre-specified analysis?) and as a DGP for power simulation
(does the primary analysis remain robust to class-structured
departures from gaussianity?).

Third, if N-of-1 simulation infrastructure pursues a richer DGP in
subsequent work, the natural target is the heterogeneous random-
slopes form rather than a full FMM. The random-slopes
parameterisation preserves the analysis model's compatibility with
the existing `nlme::lme` machinery (the analysis model continues to
be fitted as a single linear mixed model with participant-specific
random treatment slopes) while admitting the biologically faithful
latent-class generative structure at the DGP level. The simulation
DGP is extended in a direction internal to the standard mixed-
effects framework, and the analysis model remains the pre-
specified linear mixed-effects fit; this preserves the audit chain
of the existing framework and permits direct comparison against
the covariance-only baseline under matched effect-size calibration.

# Software for mixture analyses

R packages relevant to estimating the models surveyed above, in
approximate order of fit to the longitudinal aggregated N-of-1 use
case, are as follows.

- `lcmm` [@proustlima2017]. Joint latent-class mixed models for
  longitudinal continuous and ordinal outcomes; the closest
  available tooling for heterogeneous random-effects specification.
  Syntax is closely related to `nlme::lme` with a latent-class
  extension.
- `flexmix` [@leisch2004; @grunleisch2008]. General regression-
  mixture framework with user-definable component models; suitable
  for regression-mixture and mixture-of-experts specifications.
  More general than `lcmm` but requires more configuration.
- `mclust` [@mclust2016]. Gaussian finite mixture models with a
  range of class-specific covariance parameterisations; supports
  @arminger1999 mean-plus-covariance specifications non-
  longitudinally. Useful for cross-sectional latent-profile analyses
  of baseline biomarker panels but does not natively handle
  repeated-measures structure.
- `gamlss` [@gamlss2005]. Distributional regression with
  covariate-dependent location, scale, and shape parameters.
- `poLCA` [@linzerlewis2011]. Polytomous latent class analysis for
  the categorical-indicator case; useful if baseline class
  indicators (e.g., symptom-checklist items) are incorporated into
  class assignment.
- `OpenMx` and `lavaan` with mixture extensions. Structural-
  equation mixture modelling for FMM-style specifications; more
  general but with steeper configuration cost than `lcmm` or
  `flexmix`.

The reference implementation outside R is Mplus [@mplus], which
remains the standard environment for GMM, FMM, and related mixture-
SEM specifications. Much of the published psychometric simulation
literature on mixture-model identifiability and overextraction
hazards [@bauercurran2003; @lubkemuthen2005; @nylund2007] uses
Mplus, and direct comparison with that literature occasionally
requires access to it.

# Discussion

The biomarker-treatment moderation problem in aggregated N-of-1
trials sits at the intersection of two methodological traditions
that have, until now, only weakly engaged with each other. The
biostatistical tradition has developed efficient estimators for
within-participant treatment contrasts and the meta-analytic
machinery to combine them, but has typically encoded biomarker
moderation as a continuous interaction in fixed-effects regression.
The psychometric tradition has developed mature finite-mixture and
latent-class machinery for representing discrete subpopulation
structure but has applied it primarily to cross-sectional or
observational settings with sample sizes well above the N-of-1
range. The argument of this paper is that the choice between
continuous-heterogeneity and latent-class encodings is a
substantive scientific question, not a parameterisation
convenience, and that recognising this choice opens both a
sensitivity-analysis programme (the spectrum of biomarker-
moderation specifications above) and an analysis-strategy programme
(class-aware mixture analyses where identifiability permits, and
analysis-strategy mitigations where it does not).

Two limitations bound what the paper currently provides. First,
the simulation programme that would empirically resolve the
mixture-versus-mitigation efficiency question at trial-relevant
sample sizes is described in the companion simulation plan but
not yet executed. The five-study factorial outlined there
(mixture-vs-MVN, the mean-covariance-combined triptych,
heterogeneous random slopes, the three-regime breakdown grid, and
the small-$N$ identifiability and type I error pre-flight)
constitutes the empirical core of the present line of work and is
the next deliverable in the programme. Second, the application to
real biomarker-guided trial data has not been carried out. The
prazosin-PTSD analysis of @hendrickson2020 provides the most
immediate target; the simulation programme should run in parallel
with re-analysis of that dataset under at least three biomarker-
moderation encodings and a class-aware analysis where convergence
permits.

Three open methodological questions remain. The first is whether
class-aware analyses are usable as primary analyses at any
realistic N-of-1 sample size, or only as sensitivity checks. The
identifiability evidence from the broader mixture-model literature
suggests the latter; only direct simulation in the trial-relevant
parameter regime can settle this. The second is how analysis
half-life should be specified when the DGP is a mixture rather
than a single MVN. The continuous-decay drug indicator
$D_{it} = \exp(-\lambda t_{sd,it})$ has a clean interpretation
under a single-component generative model; its interpretation when
the underlying generative process is class-structured and class-
conditional carryover dynamics differ between classes is less
obvious and may require analysis-side mixture extension. The third
is whether the trial design itself should respond to the
moderation-encoding question. If a latent-class generative form is
biologically more plausible than a continuous one, designs that
maximise the per-participant contrast and minimise carryover may
gain power that the standard hybrid design surrenders.

The five-study simulation programme described in the companion
plan is structured to address each of these questions
sequentially. The factorial designs are pre-registered through
ADEMP [@morris2019] tables committed before production runs; the
class-aware analyses are wrapped uniformly so that simulation
results are directly comparable across studies; and the
infrastructure required to add the necessary DGPs to the existing
simulation framework is enumerated, scoped, and sequenced to
support both this paper's revisions and the companion carryover
and biomarker-interaction papers in the wider research programme.

## Why the analysis model is simpler than the data-generating process

A natural reaction to the simulation programme above is that the
DGP is structured (a three-component decomposition into
biological response, placebo-belief response, and time-variant
natural history under the standard pmsimstats parameterisation;
or a latent-class mixture under the alternatives in this paper)
while the analysis model is a four-term linear mixed model
`Sx ~ bm + t + Dbc + bm:Dbc` with random intercept and AR(1)
residual correlation. A reader newly encountering this asymmetry
is reasonable to ask why the analysis does not mirror the DGP
component-by-component, and whether doing so would improve
inference about the moderation estimand. The short answer, with
brief justification, is the following.

A component-matched analysis (a non-linear mixed-effects model
fitting Gompertz BR, PB, and TV components with participant-
specific random effects) is impractical for trial-relevant N
for four reasons: the trial design must supply the contrasts
that identify each component (open-label versus blinded for PB,
on-drug versus blinded-placebo for BR, off-drug timepoints for
TV), and many designs supply only some of them; each Gompertz
component costs roughly three to four parameters at the population
level and additional random-effect parameters per participant,
which at $N \in [30, 150]$ inflates the variance of the moderation
estimand more than it reduces residual confounding; the non-
linear fitting machinery has higher convergence-failure rates and
is more sensitive to starting values than the linear-mixed
analogue; and the moderation estimand $\beta_{bm:D}$ is robust to
the BR-versus-PB attribution for the validation question, even
though it is not robust to the larger questions of whether the
moderation is class-structured (the present paper) or covariance-
versus-mean-encoded (the dgp-mean-moderation-vs-mvn companion).

The smallest extension that retains tractability while improving
the BR-versus-PB sensitivity of the inference is a phase indicator
plus its interaction with the drug indicator, schematically
`Sx ~ bm + t + Dbc + phase + Dbc:phase + bm:Dbc + bm:Dbc:phase
+ (1 | ptID)`. This adds three to four parameters and tests
two things directly: whether the apparent drug effect shrinks
under blinding (a signature that the open-label drug effect was
partly PB-mediated) and whether the moderation itself shrinks
under blinding (a signature that the biomarker-by-treatment
interaction was partly biomarker-by-PB rather than biomarker-by-
BR). For papers where pharmacological-versus-placebo
disentanglement is a secondary scientific question, this
extension is the right step; for papers where the primary
estimand is the marginal moderation, the simple `Sx ~ bm + t +
Dbc + bm:Dbc` analysis is adequate. The simulation programme
fixed in the companion plan uses the simple form throughout; the
phase-indicator extension is a candidate sensitivity analysis to
add to Study 1 (the mixture-vs-MVN factorial) once the primary
cells are run. The pedagogy doc at
`docs/24-component-decomposition-pedagogy.md` develops this point
in more detail with worked numerical examples.

## The latent-class DGP as a generator of apparent subadditivity

A connection worth flagging explicitly: the latent-class DGP
specified in Study 1 is a natural mechanism for generating
**apparent subadditivity** in a downstream one-component
analysis, even when no direct $BR \times PB$ interaction is
specified in the class-conditional generative model. The
companion methods paper at
`analysis/report/06-component-decomposition/` defines
subadditivity as a non-zero coefficient $\gamma$ on a
$BR \times PB$ interaction term in the conditional-mean
specification of the additive decomposition, and notes that
$\gamma > 0$ in a fitted additive-extension analysis is
ambiguous between two distinct underlying mechanisms.

The first mechanism is **direct mechanistic interaction**:
shared neural substrate, ceiling effects on the symptom
scale, or belief-dynamics adaptation each produce a true
non-additive coupling between drug and placebo response.
The second mechanism is **emergent subadditivity from latent-
class structure**: if the trial population is a mixture of
responder and non-responder classes (the formulation in §3 of
the present paper) and class membership correlates with both
$BR$ and $PB$ strength (responder-class participants have both
higher $BR$ and higher $PB$ on average), then the marginal
expected value of $BR \cdot PB$ over the unobserved class label
has a non-zero covariance term whose magnitude depends on the
gating function $\pi(B)$. From the additive-extension analysis
this looks exactly like a $\gamma$-weighted interaction.

The Study 1 pre-registration (see
`analysis/scripts/latent-class-mixture-application/00-ademp-
pre-registration.md`) accordingly includes a
`pb_class_correlation` axis that varies $\mathrm{Cor}(BR, PB)$
across class labels: a PB-class-independent cell (matching the
original specification) and a PB-class-correlated cell
(generating apparent subadditivity). The Study 1 simulation
output therefore lets the analyst distinguish the two
mechanisms empirically: if the additive-extension $\hat{\gamma}$
is large in the PB-class-correlated cell but small in the
PB-class-independent cell at matched marginal effect size, the
mechanism is emergent rather than mechanistic. The cross-paper
synthesis with the BR-PB-TV decomposition paper is therefore
not just methodological convenience but a direct route to
falsifying or confirming a class-aware versus mechanistic-
interaction interpretation of any apparent subadditivity
finding.

## Progress to date

The five-study production programme is pre-registered under
the ADEMP framework of Morris, White, and Crowther (2019)
[@morris2019] in
`analysis/scripts/latent-class-mixture-application/00-ademp-pre-registration.md`;
the present manuscript reports a Study 5 pilot at 240
replicates per cell.

Phase 1 of the simulation programme has reached the
infrastructure-and-pilot milestone. Three components are
committed and exercised end-to-end on the host R session.

**Pre-registration.** ADEMP tables for all five studies
[@morris2019] are committed at
`analysis/scripts/latent-class-mixture-application/
00-ademp-pre-registration.md`, with cell counts (~2,300 cells
across the programme), seeded random-number control, an effect-
size calibration sub-study, and a deviation log
specification. Subsequent production runs are diagnosed against
these pre-registered estimands rather than against any post-hoc
wrapper choice.

**Class-aware analysis wrapper.** An `lcmm_analysis()` wrapper
around `lcmm::hlme` is committed at `01-lcmm-wrapper.R` with the
same input-output contract as the existing `lme_analysis()`. The
wrapper uses a two-stage fit (a single-class `hlme` is fitted
first to anchor starting values for a `gridsearch`-based
multi-start `hlme(ng=2)` fit) and extracts two moderation
signals: the gating-coefficient z-test on the biomarker (the
primary class-aware moderation test, identifying whether the
biomarker predicts class membership) and the within-class
$\beta_{bm:D}$ t-test (the secondary signal, retained for
inferential symmetry with the linear-mixed baseline).

**Heterogeneous-random-slopes DGP.** A wrapper around
`generateData()` that adds a Bernoulli class-membership channel
with logistic gating in the standardised biomarker and class-
specific multipliers on the BR factor at on-drug timepoints is
committed at `02-heterogeneous-slopes-dgp.R`. On a single-
replicate smoke test ($N = 70$, gating slope 1.5, responder
multiplier 1.0, non-responder multiplier 0.0) the DGP produces
46% responders and a marginal $\mathrm{Cor}(B, BR) = 0.63$ that
the linear-mixed bm:Dbc test detects ($p = 0.002$) without
class awareness; the lcmm fit returns a posterior class-
assignment entropy of 0.86 and a BIC 170 lower than the matched
single-component MVN-DGP fit, indicating that the mixture model
discriminates the two DGPs at this parameter cell.

**Study 5 pre-flight pilot.** A 100-replicate pilot under the
single-component null DGP (Architecture B with $c_{bm} = 0$,
$N = 70$, hybrid OL design) is reported at
`output/03-study5-pilot.rds`. Per-replicate timing is 0.05 s
for the linear-mixed baseline and 1.57 s for the class-aware
analysis (32:1 cost ratio, consistent with the smoke-test
estimate). lcmm convergence rate is 96%; the four non-converging
replicates encountered numerical issues during the gridsearch
initialisation.

Type I error at $\alpha = 0.05$ (Monte Carlo standard error
0.022) varies sharply across candidate test forms, motivating
the choice of the primary class-aware moderation test:

| Test                                     | Rejection rate |
|------------------------------------------|----------------|
| lme `bm:Dbc` Wald t-test                 | 0.000          |
| lcmm gating-on-`bm` Wald z (unconditional) | **0.250**    |
| lcmm within-class `bm:Dbc` Wald z        | 0.000          |
| BIC selects ng=2 over ng=1               | 0.010          |
| LRT chi-sq(df=4), naive reference        | 0.146          |
| BIC-conditional gating procedure         | 0.000          |

The unconditional Wald-on-gating test rejects at five times
the nominal rate. Diagnosis of the rejecting replicates shows
that the empirical standard deviation of the gating coefficient
across reps (sd = 63.9) is roughly fifty times the median
within-replicate Wald SE (0.86), and the empirical sd of the
z-statistic is 1.47 against a nominal 1.0. The gating
coefficient is centred at zero (median -0.10, sign balance
46/54) and shows no systematic bias; the inflation is driven by
the conditional Wald SE under-estimating the true variability
of $\hat{\beta}_{\text{gating}}$ at trial-relevant $N$. This is
a finite-sample identifiability artifact predicted by the
@bauercurran2003 and @lubkemuthen2005 overextraction literature:
when the data have no class structure but `hlme` is forced to
fit `ng=2`, the optimiser latches onto random patterns to
define a class boundary, and the resulting Wald SE does not
absorb the labelling uncertainty.

The naive LRT against a chi-squared(df=4) reference is
similarly anti-conservative (rejection 0.146), because the
true LRT distribution at the ng=1 boundary is a chi-bar-squared
mixture rather than a chi-squared (@mclachlanpeel2000, ch. 6).
The bootstrapped LRT (BLRT) of @nylund2007 is the standard
calibration but is computationally expensive at the per-fit
budget of a class-aware simulation.

The pragmatic class-aware moderation test is therefore a
**BIC-conditional gating procedure**: fit ng=1 and ng=2,
prefer ng=2 only if BIC supports it ($\Delta_{\text{BIC}} > 0$
or, more conservatively, $> 6$ per Raftery 1995), and only then
test the gating coefficient. At $N = 70$ under the null DGP,
BIC selects ng=2 in 1% of replicates, and zero of those reps
also reject the gating Wald, giving the conditional procedure
a Type I error of 0 at the pilot's resolution. The conditional
procedure is therefore usable as a primary moderation test at
trial-relevant $N$. Whether the same conditional procedure has
adequate power to detect a true two-class DGP is the central
question for the rest of Study 5; pilots at non-null cells
will measure this directly.

The diagnostic distinction matters for the paper's argument.
The earlier proposition that 'class-aware analyses are not
usable as primary at trial-relevant $N$' is too strong as
stated: it applies to the unconditional Wald-on-gating test
but not to the BIC-conditional procedure. The corrected
proposition is that the unconditional Wald test on a single
mixture-model coefficient is unreliable at small $N$ and
requires a model-comparison wrapper (BIC or BLRT) to control
Type I error.

**Production-budget implication.** Extrapolated to the pre-
registered cell counts (Study 1 has 768 cells × 1,000
replicates = 768,000 fits per analysis arm), the class-aware
arm of Study 1 is approximately 340 hours of serial CPU time at
the pilot's mean fit time. Parallelisation across 8-16 cores
reduces this to 22-44 hours per arm. Studies 2 and 3 require
similar budgets; the Study 5 production run (5,000 replicates
× three sample sizes × three class counts × three DGP families)
is the largest single block at approximately 60 hours per arm
on 8 cores. The full programme is feasible on a single
workstation over a week of background compute, with the
calibration sub-study and Study 5 pre-flight running first to
finalise effect-size axes and identifiability cuts before
committing to Studies 1, 2, and 3.

**Open question raised by the pilot.** With Type I error of the
BIC-conditional procedure controlled at the null at $N = 70$,
the next-most-pressing identifiability question is whether the
BIC selection step has enough sensitivity to detect a real
two-class DGP at trial-relevant $N$. A class-aware procedure
that is conservative under the null but rarely fires under
true class structure is no more useful than a procedure that
fires too often. Pilot 4 (planned for the next iteration) will
fit the same BIC-conditional procedure to the heterogeneous-
random-slopes DGP from Study 3 (with several values of the
gating slope and class-separation $\Delta$) and report the
power profile of the conditional procedure alongside that of
the linear-mixed baseline. The expected pattern is that the
BIC-conditional procedure outperforms the linear-mixed bm:Dbc
test at large class separations (where ng=2 BIC consistently
wins) but underperforms it at moderate class separations (where
BIC is undecided and the procedure falls back to a default that
does not exploit the latent-class structure).

## Pilot results (Study 5, 240 replicates)

The Study 5 prototype was rerun at 240 replicates per cell on
4-core `parallel::mclapply` with three gating-slope levels
(gs $\in \{0, 1.0, 1.5\}$) and three test forms (lme `bm:Dbc`
Wald, lcmm gating Wald unconditional, BIC-conditional gating).
Wall time 558 s; convergence 92.9% under the null and 98.3 to
98.8% at non-null cells. Per-replicate output at
`analysis/data/quick-sim/03-latent-replicates.rds`; Morris
summary at `03-latent-summary.txt`. Monte Carlo standard error
at the working power near $0.5$ is approximately $0.030$.

The unconditional Wald-on-gating type I error of $0.197$
(95% binomial CI $[0.145, 0.250]$) is now decisively
distinguishable from the nominal $0.05$ (more than five MCSE
of separation) and is statistically consistent with the
100-replicate pilot's $0.250$ point estimate. Power at
gs = $1.0$ ($0.304$) and gs = $1.5$ ($0.360$) sits below the
lme baseline ($0.325$ and $0.542$ respectively), so the
unconditional Wald buys nothing in exchange for the type I
inflation. The BIC-conditional procedure does control type I
($0.004$ under the null, with upper CI near $0.013$), but its
power is $0.025$ at gs = $1.0$ and $0.004$ at gs = $1.5$:
under the heterogeneous-random-slopes DGP at $N = 70$, the
BIC step selects ng = 1 in essentially every replicate and the
conditional procedure approximates a never-reject rule. The
non-monotonicity (gs = $1.5$ below gs = $1.0$) suggests that
as the heterogeneous-slopes signal strengthens, a single-class
lme captures it well enough that BIC continues to favor ng = 1
and the conditional procedure does not fire.

The lme `bm:Dbc` test holds nominal type I (0 of 240 under
the null) and reaches $0.542$ power at gs = $1.5$, dominating
both lcmm tests at this sample size and DGP regime. The
proposition the manuscript should carry forward is that, at
trial-relevant $N$ and under the heterogeneous-random-slopes
generating mechanism examined here, neither the unconditional
Wald-on-gating nor the BIC-conditional gating test is
competitive with the linear-mixed baseline: the unconditional
test is severely anti-conservative, and the BIC-conditional
test is severely under-powered. A canonical run at 1{,}000
replicates per cell with a wider gating-slope grid (and at
$N \in \{35, 70, 100\}$) is required to map the regime where
the BIC-conditional procedure does become competitive, likely
at large class separations rather than heterogeneous random
slopes.

# Conflict of interest

The authors declare no conflicts of interest.

# Funding

This work received no specific funding.

# Data availability statement

The data and code that support the findings of this study are
openly available in the pmsimstats-ng project repository.
Per-replicate Monte Carlo outputs and ADEMP-compliant cell-level
summaries are versioned under `analysis/data/`; simulation
drivers are at `analysis/scripts/`. Exact paths for this
manuscript are listed in the Reproducibility section below.

# Reproducibility

This manuscript was rendered with R 4.4.x inside the project's
zzcollab Docker container (image hash recorded in `Dockerfile`;
package set captured in `renv.lock`). Principal packages used
by the simulation pipeline are `nlme` (continuous-time
linear-mixed analysis with `corCAR1` residual correlation),
`parallel` (4-core `mclapply` for the lcmm-bottlenecked run),
`data.table` (the `original-extended` simulation collection),
`furrr` and `purrr` (the `tidyverse` simulation collection),
and `rmarkdown` plus `bookdown` for the manuscript build. The
class-aware analyses additionally use `lcmm` with `gridsearch`
initialisation.

**Random-number generator.** Each simulation driver sets
`set.seed(20260507)` at the top of the replicate loop and uses
`parallel::mc.set.seed` inside `mclapply`. Per-replicate seeds
derive from the master seed by `+ rep_idx`.

**Data and code availability.** Per-replicate simulation
outputs from the 240-replicate Study 5 pilot are checkpointed
at `analysis/data/quick-sim/03-latent-replicates.rds`; Morris
ADEMP cell-level summaries at
`analysis/data/quick-sim/03-latent-summary.txt`; the effect-
size calibration table at
`analysis/data/quick-sim/03-calibration-table.rds`. Driver
scripts are under
`analysis/scripts/latent-class-mixture-application/`,
including the pilot driver
`03-study5-pilot.R` (executed) and the production driver
`10-study5-production.R` (committed; the five-study production
programme has not yet been run). The pre-registered ADEMP plan
is at
`analysis/scripts/latent-class-mixture-application/00-ademp-pre-registration.md`.
The manuscript source, paper-specific bibliography
(`references.bib`), and SIM CSL file are at
`analysis/report/03-latent-class-mixture-application/`.
