---
title: "Type-I calibration of fixed-effect interaction tests in
  linear mixed models: a staged diagnosis of working-covariance
  mis-specification and a cluster-robust remedy"
author:
  - pmsimstats team
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: true
    number_sections: true
    keep_tex: false
    highlight: tango
fontsize: 11pt
geometry: "left=3cm,right=3cm,top=2.5cm,bottom=2.5cm"
bibliography: references.bib
csl: statistics-in-medicine.csl
link-citations: true
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{lineno}
  - \linenumbers
  - \usepackage{setspace}
  - \setstretch{1.1}
---

<!--
  Build pipeline for this manuscript
  ==================================

  This Rmd loads the diagnostic outputs produced by the staged
  examination of type-I conservatism in the carryover-sensitivity
  study. It does not itself run any simulation. Regenerate the
  inputs with, from the repository root:

    Rscript analysis/scripts/carryover-sensitivity/\
      check-null-architecture-equivalence.R
    Rscript analysis/scripts/carryover-sensitivity/\
      diagnose-null-type1.R
    Rscript analysis/scripts/carryover-sensitivity/\
      diagnose-null-type1-stages46.R --stage 4
    Rscript analysis/scripts/carryover-sensitivity/\
      diagnose-null-type1-stages46.R --stage 6
    Rscript analysis/scripts/carryover-sensitivity/\
      diagnose-null-type1-cr2.R
    Rscript analysis/scripts/carryover-sensitivity/\
      diagnose-alt-cell-cr2.R

  These write the five .rds files read below into
  analysis/scripts/carryover-sensitivity/output/. Then knit:

    Rscript -e "rmarkdown::render(
      'analysis/report/10-interaction-test-calibration/report.Rmd')"

  The examination plan that preceded this manuscript is retained
  alongside it as type1-examination-plan.md.
-->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  fig.align = 'center'
)
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(tibble)
  library(ggplot2)
  library(knitr)
})

out_dir <- here::here(
  'analysis/scripts/carryover-sensitivity/output')
d13 <- readRDS(file.path(out_dir, 'diag-null-type1.rds'))
d4  <- readRDS(file.path(out_dir, 'diag-null-type1-stage4.rds'))
d6  <- readRDS(file.path(out_dir, 'diag-null-type1-stage6.rds'))
dcr <- readRDS(file.path(out_dir, 'diag-null-type1-cr2.rds'))
dalt <- readRDS(file.path(out_dir, 'diag-alt-cell-cr2.rds'))

theme_paper <- theme_bw(base_size = 10) +
  theme(panel.grid.minor = element_blank(),
        strip.background = element_rect(fill = 'grey92',
                                        colour = NA),
        legend.position = 'top')

spec_cols <- c(A1 = '#1f78b4', A2 = '#33a02c', A3 = '#e31a1c')
```

# Abstract {-}

**Background.** The linear mixed model is the standard tool for
testing a fixed-effect interaction in longitudinal and clustered
data. The model-based Wald test of that interaction is exact only
when the working within-cluster covariance is correctly specified.
In practice an analyst chooses a convenient parametric residual
structure, a continuous-time autoregressive correlation, compound
symmetry, or an unstructured form, and then trusts the model-based
standard error. The direction and the magnitude of the
mis-calibration that follows when the working covariance is wrong
are rarely checked.

**Methods.** We set out a staged diagnostic protocol, a
four-channel decomposition of the test statistic, a
working-covariance model ladder, and a sample-size gradient, and
apply it to a worked example: the test of a biomarker-by-treatment
interaction in an aggregated N-of-1 carryover simulation, analysed
with an `lme` model carrying a random intercept and a corCAR1
residual correlation.

**Results.** At the null the model-based standard error of the
interaction is over-estimated by 6 to 10 percent, and the realised
type-I error is approximately 0.03 against a nominal 0.05. The
degrees-of-freedom rule, the positive-definiteness correction, and
point-estimate bias are excluded as causes. The model ladder
localises the mis-calibration to the corCAR1 residual layer: a
random intercept alone is well calibrated. The sample-size gradient
shows the bias is asymptotic, not a moderate-sample artefact: it
does not attenuate from $N = 35$ to $N = 280$. A CR2 bias-reduced
cluster-robust standard error restores the standard-error ratio to
approximately one and the type-I error to approximately 0.05. At an
alternative cell with a non-null effect the recalibration raises
power by a near-uniform increment and leaves the ranking of
analysis specifications unchanged.

**Conclusions.** A mis-specified parametric working covariance can
bias the model-based standard error of a mixed-model interaction
test asymptotically, in either direction. Cluster-robust standard
errors are a general remedy, and the staged diagnostic is a
reusable instrument for detecting the problem.

# Introduction

The linear mixed model is, by a wide margin, the most common
vehicle for testing whether an effect is moderated by a covariate
in longitudinal or clustered data [@laird1982; @diggle2002]. The
quantity of interest is a fixed-effect interaction coefficient, and
the inference almost always proceeds through a Wald test: the
coefficient estimate is divided by a model-based standard error and
referred to a $t$ distribution. That procedure is exact only under
a condition that is easy to state and easy to forget. The working
covariance, the random-effects structure together with the
parametric residual correlation the analyst has assumed, must be
correct.

It has been understood since the work of Eicker, Huber, and White
that this condition is more demanding than it appears
[@eicker1967; @huber1967; @white1980]. When the working covariance
is mis-specified, the model-based standard error is no longer a
consistent estimator of the true sampling variability of the
fixed-effect estimate. Liang and Zeger [-@liang1986] carried this
insight into longitudinal data analysis: their generalized
estimating equations decouple the point estimate, computed under a
convenient working correlation, from the variance estimate,
computed by an empirical sandwich that remains consistent whatever
the working correlation. The sandwich, or cluster-robust, variance
estimator has since become a standard instrument, and a substantial
literature has refined it for the small-sample case
[@mancl2001; @bell2002; @cameron2015; @pustejovsky2018].

Yet in routine mixed-model practice the working covariance is still
trusted. An analyst fitting longitudinal data with `nlme` or a
comparable package will choose a residual correlation, very often a
continuous-time first-order autoregressive form, because it is
pharmacokinetically or biologically plausible
[@pinheiro2000], and will then report the model-based standard
error of the interaction without asking what happens if the chosen
form is not exactly the form the data follow. This is not
negligence so much as a reasonable working assumption that is
seldom audited. But how large is the error when the assumption
fails? Is the resulting test conservative or anti-conservative? Does
the mis-calibration diminish as the sample grows, so that it can be
dismissed as a small-sample nuisance, or does it persist? These
questions do not have a single answer, because the answer depends
on the interaction being tested and on the way the working
covariance departs from the truth. They have to be asked of a
particular analysis.

This paper does exactly that for one analysis, and in doing so
offers two contributions that we believe are more general than the
single case. The first is a diagnostic protocol. We decompose the
Wald statistic into four channels through which mis-calibration can
enter, and we add two further instruments, a working-covariance
model ladder that localises the cause and a sample-size gradient
that separates a finite-sample artefact from an asymptotic one. The
protocol is not specific to the example and can be applied to any
mixed-model interaction test whose calibration is in doubt. The
second contribution is the finding the protocol produced. In our
worked example the model-based interaction test is conservative,
not anti-conservative as the sandwich literature's emphasis on
under-estimated standard errors might lead one to expect; the
conservatism is material, on the order of ten percent in
standard-error terms; and it is asymptotic, persisting undiminished
across an eightfold range of sample size. We trace it to a single
component of the working covariance, and we show that a
cluster-robust standard error removes it.

The worked example is drawn from the pmsimstats-ng project, a
simulation study of power to detect biomarker-treatment
interactions in aggregated N-of-1 trials under carryover
[@hendrickson2020]. The example is incidental to the argument. The
phenomenon, a parametric working covariance that biases a
fixed-effect interaction test asymptotically, and the remedy are
not properties of that simulation; they are properties of the
mixed-model Wald test, and the example serves only to make them
concrete and quantified. We begin in Section 2 with the diagnostic
protocol. Section 3 describes the worked example. Section 4 reports
the staged diagnosis and the cluster-robust remedy. Section 5
discusses the generality of the result and its limits.

# A staged diagnostic protocol

Consider a fixed-effect interaction coefficient $\theta$ estimated
by $\hat\theta$, with a model-based standard error
$\widehat{\mathrm{SE}}$ and an assigned degrees of freedom $\nu$.
The Wald test rejects the null $H_0: \theta = 0$ at level $\alpha$
when the statistic $T = \hat\theta / \widehat{\mathrm{SE}}$ exceeds
the critical value $t_{\nu, 1-\alpha/2}$ in absolute value. The
test is correctly calibrated when the realised rejection rate under
$H_0$ equals $\alpha$. When it does not, the discrepancy must enter
through one of exactly four channels, and the channels are
separable by simulation.

## The four-channel decomposition

1. **Point-estimate bias.** If $\mathrm{E}(\hat\theta) \neq 0$
   under $H_0$, the statistic is decentred. We measure this by the
   mean of $\hat\theta$ across null replicates.
2. **Standard-error calibration.** If the model-based
   $\widehat{\mathrm{SE}}$ does not estimate the true sampling
   standard deviation of $\hat\theta$, the statistic is rescaled.
   We measure this by the ratio
   $$
   r \;=\; \frac{\overline{\widehat{\mathrm{SE}}}}
                 {\mathrm{SD}(\hat\theta)},
   $$
   the mean model-reported standard error over the empirical
   standard deviation of the estimate across replicates. A ratio
   above one indicates an over-stated standard error and a
   conservative test; a ratio below one the reverse.
3. **Degrees of freedom.** If $\nu$ is too small, the $t$
   reference has tails too heavy and the test is conservative;
   if too large, the reverse. We measure this by recomputing the
   p-values against a normal reference and comparing the realised
   type-I error.
4. **Reference-distribution shape.** Even with $\hat\theta$
   unbiased, $\widehat{\mathrm{SE}}$ calibrated, and $\nu$ correct,
   the statistic may not follow a $t$ distribution. We measure this
   by the standard deviation of $T$ under $H_0$, which should be
   approximately one, and by the empirical distribution function of
   the p-values, which should be Uniform$(0,1)$.

The four channels are not merely a checklist. They are close to
exhaustive: a Wald test is a centred statistic, a scale, a
reference distribution, and a degrees-of-freedom choice, and
nothing else. Locating the discrepancy in one channel rather than
another is the first and most informative step of a diagnosis,
because the channels have different causes and different remedies.

## The working-covariance model ladder

The four-channel decomposition says *how* a test is mis-calibrated.
It does not say *why*. When the decomposition points to the
standard-error channel, the natural follow-up is to ask which part
of the fitted model is responsible. We answer this with a model
ladder: the same datasets are refitted under a graded sequence of
working covariances, from the simplest to the production model,
and the standard-error ratio is tracked along the sequence. The
rung at which the ratio departs from one identifies the component
that introduces the mis-calibration. Because every rung is fitted
to the same simulated data, the comparison is free of Monte Carlo
noise between models.

## The sample-size gradient

A standard-error mis-calibration may be one of two qualitatively
different things. It may be a finite-sample artefact, in which case
it attenuates as the sample grows and the working-covariance
parameters are estimated more sharply; the small-sample corrections
of Kenward and Roger [-@kenward1997; -@kenward2009] address
mis-calibration of this kind. Or it may be asymptotic, a
consequence of a working covariance that does not match the
data-generating covariance no matter how much data is collected, in
which case it persists. The two are distinguished by running the
decomposition across a range of sample sizes. Attenuation toward
nominal calibration marks the finite-sample case; a flat profile
marks the asymptotic case. The distinction matters, because a
finite-sample artefact at a realistic sample size may be tolerable,
whereas an asymptotic bias is a property of the method and will not
be outgrown.

# Worked example: an interaction test under carryover

The example is the test of a biomarker-by-treatment interaction in
the carryover-sensitivity simulation study of the pmsimstats-ng
project. Aggregated N-of-1 trials repeatedly cross each participant
between an active treatment and a control condition, and the
scientific question is whether a baseline biomarker moderates the
treatment effect. Carryover, the persistence of drug effect into a
washout period, complicates the analysis, and the study compares
analysis-model strategies for accommodating it
[@morris2019simulation]. The reporting follows the ADEMP structure
of Morris, White, and Crowther [-@morris2019simulation].

## The analysis model

For a symptom score $S_{x,it}$ on participant $i$ at timepoint $t$,
the analysis model is the linear mixed model
$$
S_{x,it} \;=\; \beta_0 + \beta_1\,\mathrm{bm}_i
  + \beta_2\,t + \beta_3\,X_{it}
  + \theta\,(\mathrm{bm}_i \times X_{it}) + b_i + \varepsilon_{it},
$$
fitted by `nlme::lme` with a participant random intercept
$b_i \sim \mathrm{N}(0, \sigma_b^2)$ and a continuous-time
first-order autoregressive residual correlation,
$\mathrm{corCAR1}(\mathrm{form} = \sim t \mid \mathrm{ptID})$. Here
$\mathrm{bm}_i$ is the standardised biomarker, $X_{it}$ a
drug-exposure predictor, and $\theta$ the interaction coefficient
of interest. The continuous-time autoregressive residual was adopted
in the project for its clinical realism, nearby measurements being
more strongly correlated than distant ones, and for the numerical
range it affords [@pinheiro2000]. The Wald test of $\theta$ is the
object of this paper.

## The null configuration

Calibration is a null-hypothesis property, so the diagnosis is run
where the biomarker exerts no moderation, $c_{bm} = 0$, and the
true interaction is therefore $\theta = 0$. The primary cell is the
reference configuration of the parent study: the Hybrid N-of-1
design, $N = 70$ participants, an exponential carryover form with a
one-week half-life, and an autoregressive parameter of 0.7. A
preliminary check established that at $c_{bm} = 0$ the study's two
data-generating architectures collapse to a bit-identical process,
so the diagnosis need not stratify by architecture.

The feature of the example that matters for the argument is the
relationship between the working covariance and the truth. The
simulated symptom score is a sum of three latent factors, each
following its own autoregressive process, with timepoint-dependent
variances, plus a participant-level baseline term. The working
covariance of the analysis model, one random intercept and one
single-parameter corCAR1 residual, cannot reproduce that structure
exactly. The example is therefore an instance of the general
situation the protocol is designed for: a parametric working
covariance that is plausible, conventional, and not exactly
correct.

# Results

The diagnosis was carried out at 5000 null replicates for the
four-channel decomposition and the model ladder, and at 1500
replicates per point for the sample-size gradient. The
cluster-robust remedy was verified at 3000 replicates. All
replicate-level outputs are retained for inspection.

## The conservatism is standard-error over-estimation

```{r tab-decomp}
t1 <- d13$decomp |>
  left_join(d13$stage3 |> select(spec, mean_df, type1_05_z),
            by = 'spec') |>
  transmute(
    Spec = spec,
    `Mean estimate` = sprintf('%+.3f', mean_est),
    `SE ratio $r$`  = sprintf('%.3f', se_ratio),
    `SD($T$)`       = sprintf('%.3f', sd_tstat),
    `Type I ($t$)`  = sprintf('%.3f', type1_05),
    `Type I ($z$)`  = sprintf('%.3f', type1_05_z),
    `Mean DF`       = sprintf('%.0f', mean_df))
kable(t1, booktabs = TRUE, escape = FALSE,
  caption = 'Four-channel decomposition at the primary null cell
    (Hybrid design, $N = 70$, $c_{bm} = 0$, 5000 replicates), for
    the three analysis specifications A1, A2, A3. The interaction
    estimate is effectively unbiased; the SE ratio exceeds one; the
    statistic is correspondingly compressed; the $t$ and $z$
    references coincide because the degrees of freedom are large.')
```

Table \@ref(tab:tab-decomp) reports the decomposition for the
three analysis specifications carried by the parent study, which
differ only in the drug-exposure predictor $X_{it}$ and are treated
here as three instances of the same interaction test. The reading
is consistent across all three. The interaction estimate is
effectively unbiased: the null mean is approximately $+0.03$ on a
coefficient whose sampling standard deviation is near unity, a
displacement of about 0.04 of a standard deviation. A displacement
of that size is statistically detectable at 5000 replicates but is
negligible in magnitude, and in any case it points toward rejection
and so cannot produce a conservative test. The degrees-of-freedom
channel is excluded directly: the model assigns approximately 487
degrees of freedom to the interaction term, and recomputing the
p-values against a normal reference moves the realised type-I error
by about 0.001, as the near-identical $t$- and $z$-reference
columns show.

The discrepancy is in the standard-error channel. The model-based
standard error exceeds the empirical sampling standard deviation of
the estimate by 6 to 10 percent. This single fact accounts for the
whole of the conservatism: the standard deviation of the test
statistic equals the reciprocal of the standard-error ratio to
three significant figures, so the statistic is compressed by
exactly the factor by which the standard error is inflated, and no
residual reference-distribution anomaly remains to be explained.
The realised type-I error at $\alpha = 0.05$ is 0.029 to 0.035
against the nominal 0.05.

```{r fig-ecdf, fig.width = 5.6, fig.height = 5.2, out.width = "62%", fig.cap = "Empirical distribution function of the null p-values at the primary cell, against the Uniform(0,1) diagonal. All three specifications bow uniformly below the diagonal, the signature of a global scale compression rather than a tail anomaly."}
d13$results |>
  filter(converged) |>
  ggplot(aes(p_value, colour = spec)) +
  stat_ecdf(geom = 'step', linewidth = 0.6) +
  geom_abline(slope = 1, intercept = 0, linetype = 'dashed',
              colour = 'grey50') +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  scale_colour_manual(values = spec_cols) +
  labs(x = 'Null p-value', y = 'Empirical CDF', colour = 'Spec') +
  theme_paper
```

Figure \@ref(fig:fig-ecdf) confirms the nature of the
mis-calibration. The empirical distribution function of the null
p-values lies below the diagonal across the whole unit interval and
for all three specifications. A test that failed only in its tails
would depart from the diagonal only near zero; the uniform bowing
seen here is the signature of a global scale compression, which is
what an inflated standard error produces. The positive-definiteness
correction available to the data generator was not triggered in any
of the cell's paths and is excluded as a contributor.

## The corCAR1 residual layer is the cause

```{r tab-ladder}
lab <- c(M0 = 'OLS, no within-cluster correlation',
         M1 = 'Random intercept only',
         M2 = 'corCAR1 residual only',
         M3 = 'Random intercept + corCAR1 (production)')
t2 <- d4$summary |>
  select(spec, model, se_ratio, type1_05) |>
  pivot_wider(names_from = spec,
              values_from = c(se_ratio, type1_05)) |>
  arrange(model) |>
  transmute(
    Model = as.character(model),
    Structure = lab[as.character(model)],
    `$r$ (A1)` = sprintf('%.3f', se_ratio_A1),
    `$r$ (A2)` = sprintf('%.3f', se_ratio_A2),
    `Type I (A1)` = sprintf('%.3f', type1_05_A1),
    `Type I (A2)` = sprintf('%.3f', type1_05_A2))
kable(t2, booktabs = TRUE, escape = FALSE,
  caption = 'Working-covariance model ladder at the primary null
    cell (5000 replicates), for specifications A1 and A2. The same
    datasets are refitted under each structure. M1, a random
    intercept alone, is well calibrated; M3, the production model,
    adds a corCAR1 residual and is conservative.')
```

The decomposition says the standard error is over-estimated. The
model ladder, Table \@ref(tab:tab-ladder), says which part of the
model does the over-estimating. Four working covariances were
fitted to the identical null datasets: ordinary least squares with
no within-cluster term (M0), a random intercept alone (M1), a
corCAR1 residual alone (M2), and the production combination of a
random intercept with a corCAR1 residual (M3).

The result is unambiguous. M1, the random intercept with no
residual correlation, is well calibrated: its standard-error ratio
sits between 0.97 and 0.99 and its type-I error between 0.049 and
0.055. M3, which adds the corCAR1 residual on top of that random
intercept, is conservative: the ratio rises to between 1.07 and
1.10 and the type-I error falls to between 0.029 and 0.033. The
corCAR1 residual-correlation layer is therefore the proximate cause
of the conservatism. It is worth being precise about what this does
and does not say. It does not say that the corCAR1 residual is
wrong in any absolute sense; the simulated factors are genuinely
autocorrelated, so a residual correlation term is capturing a real
feature of the data. It says that adding that term to a model that
already carries a random intercept degrades the calibration of the
fixed-effect interaction test.

```{r fig-ladder, fig.width = 7, fig.height = 3.5, out.width = "100%", fig.cap = "Interaction standard-error calibration across the model ladder. Left, the ratio of model-reported to empirical standard error; right, the realised type-I error. The dashed lines mark perfect calibration. The random intercept alone (M1) sits on both reference lines; the production model (M3) does not."}
ref <- tibble(metric = c('se_ratio', 'type1_05'),
              yint = c(1.0, 0.05))
d4$summary |>
  select(spec, model, se_ratio, type1_05) |>
  pivot_longer(c(se_ratio, type1_05), names_to = 'metric',
               values_to = 'value') |>
  ggplot(aes(model, value, colour = spec, group = spec)) +
  geom_hline(data = ref, aes(yintercept = yint),
             linetype = 'dashed', colour = 'grey50') +
  geom_line(linewidth = 0.6) + geom_point(size = 2) +
  facet_wrap(~ metric, scales = 'free_y') +
  scale_colour_manual(values = spec_cols) +
  labs(x = 'Working covariance', y = NULL, colour = 'Spec') +
  theme_paper
```

Two further features of the ladder, visible in
Figure \@ref(fig:fig-ladder), are worth recording. The first is
that M0, ordinary least squares, is not anti-conservative, as the
familiar warning that ignoring positive within-cluster correlation
under-states standard errors would suggest. It is strongly
conservative, with a standard-error ratio near 1.3. The reason is
instructive: the interaction here is a within-participant contrast,
and ordinary least squares charges it the full residual variance,
between-participant and within-participant combined, when the error
relevant to a within-participant contrast is only the
within-participant part. The familiar warning concerns
between-cluster effects; for a within-cluster effect the sign of
the OLS error is reversed. The second feature is that the corCAR1
residual without a random intercept (M2) is intermediate. Taken
together the ladder shows that calibration is recovered by the
random intercept alone and removed by the corCAR1 addition, and
that the magnitude and even the sign of the mis-calibration is a
property of the working covariance, not a universal constant.

## The conservatism is asymptotic, not a small-sample artefact

```{r tab-grad}
t3 <- d6$summary |>
  select(N, spec, se_ratio, type1_05) |>
  pivot_wider(names_from = spec,
              values_from = c(se_ratio, type1_05)) |>
  arrange(N) |>
  transmute(
    `$N$` = N,
    `$r$ A1` = sprintf('%.3f', se_ratio_A1),
    `$r$ A2` = sprintf('%.3f', se_ratio_A2),
    `$r$ A3` = sprintf('%.3f', se_ratio_A3),
    `T1 A1` = sprintf('%.3f', type1_05_A1),
    `T1 A2` = sprintf('%.3f', type1_05_A2),
    `T1 A3` = sprintf('%.3f', type1_05_A3))
kable(t3, booktabs = TRUE, escape = FALSE,
  caption = 'Sample-size gradient under the production model M3
    (1500 replicates per point). The standard-error ratio $r$ and
    the type-I error (T1) are shown for each specification. Neither
    trends toward its nominal value as $N$ increases.')
```

If the conservatism were a finite-sample artefact, it would
diminish as the sample grew and the working-covariance parameters
were estimated more precisely. Table \@ref(tab:tab-grad) and
Figure \@ref(fig:fig-grad) show that it does not. Across an
eightfold range of sample size, from $N = 35$ to $N = 280$
participants, the standard-error ratio holds between 1.05 and 1.10
with no trend toward one, and the type-I error holds between 0.028
and 0.045 with no trend toward 0.05. At the largest sample size the
production specification still rejects the true null at 0.028. The
conservatism is asymptotic. It is a property of the M3 working
covariance and will not be outgrown.

```{r fig-grad, fig.width = 7, fig.height = 3.5, out.width = "100%", fig.cap = "Type-I error and standard-error ratio against sample size under the production model M3. Neither panel trends toward its dashed reference line as the number of participants increases from 35 to 280."}
ref2 <- tibble(metric = c('se_ratio', 'type1_05'),
               yint = c(1.0, 0.05))
d6$summary |>
  select(N, spec, se_ratio, type1_05) |>
  pivot_longer(c(se_ratio, type1_05), names_to = 'metric',
               values_to = 'value') |>
  ggplot(aes(N, value, colour = spec, group = spec)) +
  geom_hline(data = ref2, aes(yintercept = yint),
             linetype = 'dashed', colour = 'grey50') +
  geom_line(linewidth = 0.6) + geom_point(size = 2) +
  facet_wrap(~ metric, scales = 'free_y') +
  scale_x_continuous(breaks = c(35, 70, 140, 280)) +
  scale_colour_manual(values = spec_cols) +
  labs(x = 'Sample size N', y = NULL, colour = 'Spec') +
  theme_paper
```

The explanation that is consistent with all three results is a
working-covariance mismatch rather than a small-sample effect. The
simulated symptom score is a mixture of autoregressive factors with
unequal, timepoint-dependent variances, plus a participant baseline
term. The M3 working covariance, a single random intercept and a
single-parameter corCAR1 residual, cannot match a mixture of
autoregressive processes exactly. Increasing the number of
participants estimates the parameters of that working covariance
ever more sharply, but it does not change the fact that the
covariance being estimated is the wrong one, and so the
model-based standard error converges to a biased target. The bias
persists because it is a bias of the model, not of the estimate of
the model.

## A cluster-robust standard error restores calibration

```{r tab-cr2}
t4 <- dcr$summary |>
  transmute(
    Spec = spec,
    `$r$ model-based` = sprintf('%.3f', mod_se_ratio),
    `$r$ CR2`         = sprintf('%.3f', cr2_se_ratio),
    `Type I model`    = sprintf('%.3f', type1_mod),
    `Type I CR2`      = sprintf('%.3f', type1_cr2),
    `CR2 df`          = sprintf('%.1f', cr2_df))
kable(t4, booktabs = TRUE, escape = FALSE,
  caption = 'Cluster-robust remedy at the primary null cell (3000
    replicates). The model-based standard error of the production
    model is replaced by a CR2 bias-reduced cluster-robust standard
    error clustered on participant, with Satterthwaite degrees of
    freedom. The CR2 standard-error ratio is approximately one and
    the type-I error approximately 0.05.')
```

The diagnosis points to a remedy that does not require abandoning
the corCAR1 model. If the model-based standard error is biased
because the working covariance is wrong, the natural replacement is
a standard error that is consistent whatever the working covariance
[@white1980; @liang1986]. We replace the model-based standard error
of the interaction with a cluster-robust sandwich standard error,
clustered on participant, in the bias-reduced CR2 form of Bell and
McCaffrey [-@bell2002], with the Satterthwaite degrees of freedom
of Pustejovsky and Tipton [-@pustejovsky2018; @satterthwaite1946].
The CR2 adjustment and its associated small-sample test were
designed for precisely the regime here, a moderate number of
clusters, and are implemented for `lme` objects in the
`clubSandwich` package [@tipton2015; @tipton2015moderators].

Table \@ref(tab:tab-cr2) reports the verification. The model-based
standard-error ratio reproduces the conservatism, between 1.05 and
1.10. The CR2 cluster-robust ratio is between 0.98 and 1.00, and
the realised type-I error rises from approximately 0.03 under the
model-based test to between 0.045 and 0.049 under the CR2 test,
within Monte Carlo error of the nominal 0.05. The Satterthwaite
degrees of freedom for the cluster-robust test are approximately
24, far below the model-based 487; the small-sample reference is
doing real work, and a naive normal reference for the sandwich
statistic would itself be mildly anti-conservative
[@cameron2015; @long2000].

```{r fig-cr2, fig.width = 7.5, fig.height = 3.4, out.width = "100%", fig.cap = "Null p-value distribution functions under the model-based and CR2 cluster-robust tests, by specification. The model-based curves bow below the diagonal; the CR2 curves lie on it."}
dcr$results |>
  filter(converged, !is.na(cr2_p)) |>
  select(spec, mod_p, cr2_p) |>
  pivot_longer(c(mod_p, cr2_p), names_to = 'source',
               values_to = 'p') |>
  mutate(source = recode(source,
    mod_p = 'Model-based Wald', cr2_p = 'CR2 cluster-robust')) |>
  ggplot(aes(p, colour = source)) +
  stat_ecdf(geom = 'step', linewidth = 0.6) +
  geom_abline(slope = 1, intercept = 0, linetype = 'dashed',
              colour = 'grey50') +
  facet_wrap(~ spec) +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  scale_colour_manual(values = c(
    'Model-based Wald' = '#e31a1c',
    'CR2 cluster-robust' = '#1f78b4')) +
  labs(x = 'Null p-value', y = 'Empirical CDF', colour = NULL) +
  theme_paper
```

Figure \@ref(fig:fig-cr2) shows the same result as a distribution
function. The model-based p-values bow below the diagonal, as in
Figure \@ref(fig:fig-ecdf); the cluster-robust p-values lie on it.
The point estimate and the corCAR1 model are unchanged; only the
standard error of the interaction, and the reference against which
it is tested, are replaced.

## The remedy at an alternative cell

The diagnosis to this point has been conducted under the null,
where calibration is defined. A reader of a comparative study,
however, will press a further question. The purpose of the parent
study is to rank analysis specifications by power, and if a
recalibration of the standard error raises the power of every
specification, does it also disturb their order? We checked this
directly at one alternative cell, the parent study's reference
configuration with a non-null biomarker moderation,
$c_{bm} = 0.45$, retaining the Architecture-B Hybrid design at
$N = 70$.

```{r tab-alt}
t5 <- dalt$summary |>
  transmute(
    Spec = spec,
    `$r$ model-based` = sprintf('%.3f', mod_se_ratio),
    `$r$ CR2`         = sprintf('%.3f', cr2_se_ratio),
    `Power model`     = sprintf('%.3f', power_mod),
    `Power CR2`       = sprintf('%.3f', power_cr2))
kable(t5, booktabs = TRUE, escape = FALSE,
  caption = 'The cluster-robust remedy at an alternative cell
    (Architecture B, Hybrid design, $N = 70$, $c_{bm} = 0.45$,
    exponential DGP, 3000 replicates). Power is the rejection rate
    of the interaction test; the standard-error ratio $r$ is as
    defined for the null. The model-based test is conservative
    here too, and the CR2 test recalibrates it.')
```

Three things follow from Table \@ref(tab:tab-alt). First, the
standard-error over-estimation is not confined to the null: at the
alternative cell the model-based ratio is 1.12 to 1.15, if anything
slightly larger than the 1.06 to 1.10 seen under the null. Second,
the CR2 cluster-robust standard error recalibrates the alternative
cell as it did the null, bringing the ratio to 0.99 for all three
specifications. Third, and most consequential for a comparative
study, the recalibration is close to a uniform level shift. It
raises the power of every specification by roughly three to four
percentage points, and the gaps between specifications are left
intact: the advantage of the leading specification A2 over A1 is
$+0.086$ under the model-based test and $+0.082$ under the
cluster-robust test, and over A3 it is $+0.088$ and $+0.089$. The
ranking, and the magnitude of the leading specification's
advantage, survive the recalibration unchanged.

```{r fig-alt, fig.width = 6, fig.height = 3.6, out.width = "75%", fig.cap = "Power of the three analysis specifications at the alternative cell, under the model-based and CR2 cluster-robust tests. The recalibration raises every specification by a near-uniform increment and preserves their order."}
dalt$summary |>
  select(spec, power_mod, power_cr2) |>
  pivot_longer(c(power_mod, power_cr2), names_to = 'method',
               values_to = 'power') |>
  mutate(method = recode(method,
    power_mod = 'Model-based Wald',
    power_cr2 = 'CR2 cluster-robust')) |>
  ggplot(aes(spec, power, fill = method)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_text(aes(label = sprintf('%.3f', power)),
            position = position_dodge(width = 0.7),
            vjust = -0.4, size = 3) +
  scale_fill_manual(values = c(
    'Model-based Wald' = '#e31a1c',
    'CR2 cluster-robust' = '#1f78b4')) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = 'Analysis specification', y = 'Power', fill = NULL) +
  theme_paper
```

The lesson generalises beyond the example. A cluster-robust
standard error corrects the absolute error rate, the type-I error
under the null and the power under the alternative alike, but
because the correction acts on the standard error and not on the
point estimate, and because the standard-error mis-calibration here
is similar across the specifications being compared, it shifts the
level without reordering the contrast. A comparative study whose
claim is a ranking is therefore robust to the recalibration, even
where its absolute power figures, as reported under the
model-based test, are mildly pessimistic.

# Discussion

## The result is not about autoregressive correlation

It would be a mistake to read this paper as a caution against the
continuous-time autoregressive residual in particular. The corCAR1
term is the component the model ladder happened to implicate in
this example, but the mechanism is general. The model ladder is
itself the evidence: it fitted four working covariances and
obtained four different standard-error ratios, from 1.3 under
ordinary least squares to 1.0 under a random intercept to near 1.1
under the production model. The mis-calibration is a property of
the match between the working covariance and the truth, and any
parametric residual or random-effects structure, compound symmetry,
an unstructured form, a random slope, a spatial correlation, is
capable of producing it. That a mis-specified working covariance
leaves the model-based fixed-effect standard error inconsistent is,
moreover, not a new claim; it is the Eicker-Huber-White result
carried into the mixed-model setting [@eicker1967; @huber1967;
@white1980; @freedman2006]. What the worked example contributes is
a quantified, decomposed instance in a realistic longitudinal
interaction test, with two features that are not obvious in
advance.

## Two features that are not obvious in advance

The first is the direction. The sandwich literature is usually
motivated by anti-conservative tests: ignored positive correlation
under-states standard errors, and the robust estimator corrects an
inflated rejection rate downward. Here the model-based test is
conservative, and the cluster-robust estimator corrects the
rejection rate upward. The ladder explains why the intuition can
mislead. The effect under test is a within-cluster contrast, and
for a within-cluster contrast the relationship between a
mis-specified covariance and the standard-error error need not have
the sign that the between-cluster intuition supplies. An analyst
who reasons that "my standard errors are conservative because I
have modelled the correlation" has the logic backward in this case:
modelling the correlation, through the corCAR1 term, is what
introduced the conservatism, and the less elaborate model was the
calibrated one.

The second is that the bias is asymptotic. It would have been
comfortable to attribute the conservatism to moderate sample size,
and an earlier informal reading of the parent study did so. The
sample-size gradient refutes that reading directly. This matters
for how the problem is classified. The small-sample corrections of
Kenward and Roger [-@kenward1997; -@kenward2009] adjust the
fixed-effect standard error and its degrees of freedom for the
finite-sample uncertainty in the estimated variance parameters;
they are the right instrument for a finite-sample artefact. They
are not the right instrument here, because the bias is not a
finite-sample artefact: it does not vanish as the variance
parameters are estimated with unlimited precision. A Kenward-Roger
correction to a mis-specified working covariance corrects the
finite-sample term and leaves the asymptotic mismatch untouched.
The cluster-robust standard error addresses the asymptotic mismatch
directly, which is why it is the appropriate remedy here.

## When the problem should be expected

The conditions for the problem are not exotic. They are a
fixed-effect interaction tested in a mixed model; a parametric
working covariance that is plausible but not known to be exactly
correct,
which is to say almost every parametric working covariance in
applied use; and an interaction that is partly or wholly a
within-cluster contrast. Wherever those conditions hold, the
calibration of the model-based Wald test is an open question, and
the staged protocol of Section 2 answers it at the cost of a
simulation. We would suggest, as a matter of routine practice, that
an interaction test reported from a mixed model with a non-trivial
working covariance be accompanied either by a calibration check of
this kind or by a cluster-robust standard error, which removes the
question by being agnostic to the working covariance.

## Limitations

Several limitations bound the scope of the evidence. The diagnosis
was carried out for one data-generating family, one estimand, and
one production working covariance; the asymptotic conservatism is
demonstrated for the corCAR1-plus-random-intercept model, and the
generality beyond that rests on the model ladder and on established
theory rather than on a sweep of every covariance structure. A
broader empirical base, fitting an unstructured or
compound-symmetry analysis to the same data, would strengthen the
general claim, and is a natural extension. The cluster-robust remedy is not
without cost: the Satterthwaite degrees of freedom for the CR2 test
are approximately 24 against a model-based 487, so the robust test
spends precision to buy validity, and at a genuinely small number
of clusters that price would rise. The principal diagnosis was
conducted at the null, where calibration is defined; the
consequence for power was checked at a single alternative cell
(Section 4.5), where the recalibration proved a near-uniform level
shift that preserved the ranking of specifications. Whether that
uniformity holds across the whole factorial, rather than at the one
cell examined, was not established. Finally, the small positive
point-estimate bias noted in Section 4.1, while negligible in
magnitude and irrelevant to the conservatism, is a genuine
finite-sample feature of the estimator that a separate examination
might characterise.

# Conclusions

1. The model-based Wald test of a fixed-effect interaction in a
   linear mixed model can be materially mis-calibrated when the
   parametric working covariance does not match the data-generating
   covariance. In the worked example the mis-calibration is a
   conservative one: the model-based standard error of the
   interaction is over-estimated by 6 to 10 percent and the
   realised type-I error is approximately 0.03 against a nominal
   0.05.
2. The four-channel decomposition localises the discrepancy to the
   standard-error channel and excludes point-estimate bias, the
   degrees-of-freedom rule, and the positive-definiteness
   correction.
3. The working-covariance model ladder localises the cause to the
   corCAR1 residual layer. A random intercept alone is well
   calibrated; adding the corCAR1 residual removes the calibration.
4. The sample-size gradient shows the mis-calibration is
   asymptotic. It does not attenuate from $N = 35$ to $N = 280$,
   and so it is a property of the working covariance and not a
   small-sample artefact amenable to a Kenward-Roger correction.
5. A CR2 bias-reduced cluster-robust standard error, clustered on
   participant and referred to a Satterthwaite distribution,
   restores the standard-error ratio to approximately one and the
   type-I error to approximately 0.05, without altering the point
   estimate or the analysis model. At an alternative cell the same
   recalibration is a near-uniform level shift in power that
   preserves the comparative ranking of specifications.
6. The staged protocol, the four-channel decomposition, the model
   ladder, and the sample-size gradient, is a reusable instrument
   for any mixed-model interaction test whose calibration is in
   doubt. We recommend that interaction tests reported from mixed
   models with a non-trivial working covariance be accompanied by a
   calibration check of this kind, or by a cluster-robust standard
   error that is agnostic to the working covariance.

# Reproducibility

The diagnosis is driven by five scripts under
`analysis/scripts/carryover-sensitivity/`:
`diagnose-null-type1.R` for the four-channel decomposition,
`diagnose-null-type1-stages46.R` for the model ladder and the
sample-size gradient, `diagnose-null-type1-cr2.R` for the
cluster-robust verification at the null, and
`diagnose-alt-cell-cr2.R` for the alternative-cell check, with
`check-null-architecture-equivalence.R` establishing the
architecture-invariance of the null. Each writes a replicate-level
`.rds` file into the `output/` subdirectory, and this manuscript
reads those files and regenerates every table and figure inline.
The cluster-robust standard errors use the `clubSandwich` package.
The examination plan that preceded the manuscript is retained
alongside it as `type1-examination-plan.md`.

# References
