---
title: "Robustness of carryover-mitigation analysis strategies
  for biomarker-treatment interaction: a factorial simulation
  study"
author: 'pmsimstats team'
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: true
    number_sections: true
    keep_tex: true
    highlight: tango
    includes:
      in_header: claudecode.tex
fontsize: 11pt
geometry: "left=3cm,right=5cm,top=2cm,bottom=2cm"
bibliography: references.bib
csl: statistics-in-medicine.csl
link-citations: true
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{lineno}
  - \linenumbers
  - \usepackage{setspace}
  - \setstretch{1.1}
knit: (function(input, ...) { d <- dirname(input); while (!file.exists(file.path(d, 'tools', 'stamp-render.R')) && d != dirname(d)) d <- dirname(d); source(file.path(d, 'tools', 'stamp-render.R'))$value(input) })
---

<!--
  Build pipeline for this manuscript
  ==================================

  This Rmd does NOT run any simulation code. It loads pre-computed
  summaries and figures produced by the scripts below. Re-run the
  relevant steps when the DGP, analysis specs, or parameter grid
  change. All paths are relative to the repo root.

  1. Tier 1 principal factorial (heaviest step, ~85 min on 8 cores):
       Rscript analysis/scripts/carryover-sensitivity/01-run-factorial.R
     Writes: analysis/scripts/carryover-sensitivity/output/01-factorial.rds

  2. Tier 2 sensitivity blocks S1..S5 (~4 min on 8 cores; requires
     Tier 1 not to be running concurrently - both request all cores):
       Rscript analysis/scripts/carryover-sensitivity/04-run-sensitivity-blocks.R
     Writes: analysis/scripts/carryover-sensitivity/output/04-sensitivity.rds

  3. Summarise both tiers into cell-level power/bias/convergence.
     Step 3 is cheap (~seconds) and should be re-run after either 1
     or 2 changes. It mirrors the summaries to analysis/data/:
       Rscript analysis/scripts/carryover-sensitivity/02-summarize-grid.R
     Writes:
       analysis/data/02-grid-summary.rds        (from Tier 1)
       analysis/data/02-sensitivity-summary.rds (from Tier 2, if present)

  4. Render Tier 1 figures (reads 02-grid-summary.rds):
       Rscript analysis/scripts/carryover-sensitivity/03-render-figures.R
     Writes:
       analysis/figures/02-power-by-spec.pdf
       analysis/figures/02-heatmap-matched-vs-mismatched.pdf
       analysis/figures/02-type1-boxplots.pdf

  5. Render Tier 2 sensitivity figures (reads 02-sensitivity-summary.rds;
     writes placeholder PDFs with "Pending Tier 2 production run" when
     the summary is absent, so this manuscript always knits cleanly):
       Rscript analysis/scripts/carryover-sensitivity/04-sensitivity-figures.R
     Writes:
       analysis/figures/02-sens-S1.pdf
       analysis/figures/02-sens-S2.pdf
       analysis/figures/02-sens-S3.pdf
       analysis/figures/02-sens-S4.pdf
       analysis/figures/02-sens-S5.pdf

  5b. Sensitivity block S6 (cluster-robust recalibration; ~18 min on
      8 cores; requires the clubSandwich package). Runs the five
      stress cells at c_bm=0.45 (power) AND their c_bm=0 mirrors
      (Type I), under both model-based and CR2 inference:
       Rscript analysis/scripts/carryover-sensitivity/06-cr2-recalibration.R
     Writes:
       analysis/scripts/carryover-sensitivity/output/diag-s6-cr2.rds
         ($summary = power/kappa, $type1 = null rejection rates)

  5c. S6 Type I figure (model-based vs CR2 at the null; ~seconds):
       Rscript analysis/scripts/carryover-sensitivity/07-type1-cr2-figure.R
     Writes:
       analysis/figures/02-type1-cr2.pdf

  6. Knit this Rmd (this step; ~1 min; no simulation code runs):
       Rscript -e "rmarkdown::render('analysis/report/02-carryover-sensitivity/report.Rmd')"

  Full-rebuild one-liner (after any change to the grid or DGP):
    Rscript analysis/scripts/carryover-sensitivity/01-run-factorial.R && \
    Rscript analysis/scripts/carryover-sensitivity/04-run-sensitivity-blocks.R && \
    Rscript analysis/scripts/carryover-sensitivity/02-summarize-grid.R && \
    Rscript analysis/scripts/carryover-sensitivity/03-render-figures.R && \
    Rscript analysis/scripts/carryover-sensitivity/04-sensitivity-figures.R && \
    Rscript -e "rmarkdown::render('analysis/report/02-carryover-sensitivity/report.Rmd')"

  Grid design and rationale:
    analysis/scripts/carryover-sensitivity/simulation-grid-plan.md
-->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.path = "../../figures/02-",
  cache.path = "_cache/"
)
library(tibble)
library(dplyr)
library(knitr)

## The archived simulation output labels the three analysis
## specifications A1/A2/A3. The manuscript reports them as M1/M2/M3 so
## that the letters A/B/C stay reserved for DGP architectures. Data
## values are never rewritten; they are mapped at display time only.
spec_display <- function(x) {
  m <- c(A1 = 'M1', A2 = 'M2', A3 = 'M3')
  x <- as.character(x)
  ifelse(x %in% names(m), unname(m[x]), x)
}
```

```{r source-helpers, eval = FALSE, include = FALSE}
## Source the simulation helpers when you want to step through
## the preliminary code interactively in RStudio (e.g., to inspect
## the grid tibbles, rerun a single cell, or diagnose a failing
## fit). Production runs go through the shell pipeline described
## in the HTML comment at the top of this file; nothing in this
## chunk is evaluated during knit.
source(here::here('implementations/tidyverse/R/functions.R'))
source(here::here(
  'analysis/scripts/carryover-sensitivity/simulation-core.R'))
```

# Abstract

**Background.** Carryover effects in aggregated N-of-1 and
crossover trials degrade the detectability of
biomarker-treatment interactions. Three analysis-side
specifications acknowledge carryover (binary-treatment M1,
exposure-weighted continuous-$D_{bc}$ M2, and lagged-treatment
M3); each makes implicit assumptions about the functional form
of the residual drug-effect decay. The joint sensitivity of
the inference to mis-specification at the data-generating and
analyst layers has not previously been characterized.

**Methods.** We conduct an ADEMP-compliant factorial simulation
study [@morris2019simulation] crossing three data-generating
carryover decay forms (linear, exponential, Weibull) against
the three analysis-model specifications, under both
mean-moderation (Architecture~A) and multivariate-normal (MVN)
differential-correlation (Architecture~B) biomarker
data-generating processes (DGP). The principal factorial covers
540 cells at 500 replicates per cell. Six sensitivity blocks
(S1-S6) vary autocorrelation strength, analyst-versus-DGP
half-life mismatch, dropout rate, between-subject heterogeneity,
and biomarker correlation strength, and recalibrate the test with
a cluster-robust standard error. All performance measures are
reported with Monte Carlo standard errors (MCSE) per Morris et al.
(2019).

**Results.** Under Architecture~B in the Hybrid and open-label
plus blinded-discontinuation (OL+BDC) designs, the
exposure-weighted M2 specification produced the
highest power, with an absolute advantage of approximately 10
percentage points over M1 at the most favorable cell
(Architecture~B, Hybrid design, $N = 70$, $c_{bm} = 0.45$,
$t_{1/2} = 1.0$ week; M2 = 0.860, M1 = 0.766, M3 = 0.770;
approximately five MCSE of separation). The advantage was not
universal: under the crossover (CO) design M2 was substantially
less powerful than M1 and M3 (for example M2 = 0.488 versus
M1 = 0.830 at the matched CO cell), and under Architecture~A
(direct mean moderation) M2 was marginally dominated in every
design. The ranking M2 $>$ M1 $\approx$ M3 therefore held only
under Architecture~B in the non-crossover designs; where it held,
it was preserved across sample sizes and DGP carryover half-lives,
and M2 retained its advantage when the analyst-assumed half-life
departed from the true half-life by a factor of two or four.
A high-precision rerun at $N \in \{35, 70\}$ and
$t_{1/2}^{\text{DGP}} \in \{0.5, 1.0\}$ confirmed the ranking
at MCSE $0.020$ with a paired McNemar test
$p = 6.6 \times 10^{-17}$ at the highest-leverage cell. The
model-based test is mildly conservative, a property the companion
calibration study characterizes; a CR2 cluster-robust standard
error (block S6) restores nominal size and recovers two to five
percentage points of power without altering the ranking.

**Conclusions.** Under Architecture~B with a Hybrid or OL+BDC
design, M2 reported with a cluster-robust standard error is the
recommended default for the analysis of biomarker-treatment
interactions in aggregated N-of-1 trials under realistic
carryover. This recommendation is conditional: under the crossover
design, or when the data-generating mechanism is assumed to follow
Architecture~A, M2 confers no advantage and the binary (M1) or
lagged (M3) specification is at least as powerful. Where M2 leads,
M3 is a defensible fallback when the pharmacokinetic half-life is
unknown, recovering roughly 70 percent of M2's advantage at small
$N$ and short half-life but only about half at larger $N$. M1 (the
strict-RM-ANOVA analogue) is dominated only within the
Architecture~B non-crossover cells where M2 leads; it is
competitive or superior elsewhere.

# Introduction

::: {.bullets data-latex=""}
\textbf{The joint sensitivity of biomarker-interaction power to
analysis-side carryover mis-specification has not previously been
characterized in the within-subject setting.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item Residual drug effect after the on-drug phase degrades the
  biomarker-treatment interaction test.
\item The practical magnitude of the loss and its recoverability
  are the motivating questions.
\item No systematic evaluation of this question exists for the
  within-subject biomarker-interaction setting.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
We begin with a question that has received surprisingly little
systematic attention in the biomarker-moderated N-of-1 trial
literature. When residual drug effect persists after the on-drug
phase, any analysis that ignores carryover or mis-specifies its
functional form will, in principle, lose power to detect the
biomarker-treatment interaction. But how large, in practice, is
that loss, and can it be recovered by a better-chosen analysis
specification? To the best of our knowledge, no systematic
evaluation of this question exists for the within-subject
biomarker-interaction setting.
:::

::: {.bullets data-latex=""}
\textbf{The companion manuscript shows Architecture B loses
substantially more power under carryover than Architecture A (up
to roughly 31\% in relative terms versus 1--3\%), but held the
decay form and analysis specification fixed.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item Hendrickson (2020) established the biomarker-interaction
  framework for aggregated N-of-1 trials with matched
  exponential decay.
\item Companion paper (Architecture A vs.\ B): under one week of
  carryover, Architecture B loses up to about 31\% of power in
  relative terms (design-dependent), Architecture A only 1--3\%.
\item Joint sensitivity to DGP-analyst mis-specification
  remained uncharacterized.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
The companion manuscript at
`analysis/report/01-dgp-mean-moderation-vs-mvn/` establishes one
part of the picture. Hendrickson [-@hendrickson2020] introduced
the biomarker-treatment interaction framework for aggregated
N-of-1 trials and reported power under a matched exponential decay
model. The companion paper extended that work by comparing two
data-generating architectures (direct mean moderation under
Architecture A, and multivariate-normal differential correlation
under Architecture B) and found that the power consequences of
carryover differ qualitatively between them. Under Architecture B,
carryover at a half-life of one week reduces power by up to roughly
31 percent in relative terms, the loss being strongly
design-dependent (largest under the OL+BDC design and negligible
under the crossover design); under Architecture A the
corresponding loss is only 1 to 3 percent. That prior work,
however, held the decay form fixed at
exponential and evaluated a single analysis-model specification
throughout. Unfortunately, the joint sensitivity of the inference
to mis-specification at both the data-generating and analyst
layers remained uncharacterized.
:::

::: {.bullets data-latex=""}
\textbf{The analyst must choose a decay-form assumption even when
the true pharmacokinetic decay is unknown; which specification is
most robust to that choice has not been evaluated.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item The carryover decay form is specified at the design stage,
  often without precise pharmacokinetic information.
\item The analyst typically defaults to a convenient parametric
  form or ignores carryover entirely.
\item Which analysis-side specification holds up best under
  decay-form mis-specification is the central question.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
The present paper addresses that gap directly. The trial designer
must specify a carryover decay form, explicitly or implicitly, and
the analyst must choose how to represent residual drug exposure in
the linear predictor. In practice, the true pharmacokinetic decay
is rarely known with precision at the design stage: the analyst
typically defaults to a convenient parametric form or ignores
carryover entirely. We have not, to the best of our knowledge,
seen a systematic evaluation of which analysis-side specification
holds up best when the assumed decay form departs from truth, and
it is that question we shall take up here.
:::

::: {.bullets data-latex=""}
\textbf{The study crosses three DGP decay forms against three
analysis specifications under both architectures to establish
whether robustness rankings are architecture-dependent.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item DGP decay forms: linear, exponential, Weibull.
\item Analysis specifications: M1 (binary), M2
  (exposure-weighted, pmsimstats-ng default), M3
  (lagged-treatment, Jones-Kenward).
\item Full grid run under both Architecture A and B to test
  whether rankings transfer across DGP mechanisms.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
To address this, we perform a factorial simulation study crossing
three DGP decay forms (linear, exponential, and Weibull) against
three analysis-model carryover specifications: a binary on-drug
indicator that ignores carryover altogether (M1), an
exposure-weighted continuous predictor that commits to a specific
decay form and half-life (M2, the current pmsimstats-ng default),
and a lagged-treatment indicator that estimates carryover magnitude
without positing a functional form (M3, following the classical
crossover-trial approach surveyed by Jones and Kenward
[-@jones2014crossover]). The full grid is run under both DGP
architectures to establish whether robustness rankings are
preserved across data-generating mechanisms or whether the
preferred specification is architecture-dependent.
:::

::: {.bullets data-latex=""}
\textbf{The study follows the ADEMP framework of Morris et al.
(2019) and reports every performance measure with its Monte Carlo
standard error.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item ADEMP structure: Aims, Data-generating mechanisms,
  Estimands, Methods, Performance measures.
\item MCSE reported alongside every point estimate per Morris
  et al. (2019).
\item §3 design; §4 results; §5 practical implications for
  trialists.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
The study is organized and reported following the ADEMP framework
of Morris, White, and Crowther [-@morris2019simulation], which
structures simulation studies around aims, data-generating
mechanisms, estimands, methods, and performance measures. Each
performance measure is reported per cell alongside its Monte Carlo
standard error, as Morris et al. recommend. We proceed in §3
through the simulation design in ADEMP order; results are
presented in §4; and §5 takes up the practical implications for
trialists choosing among the three carryover specifications.
:::

# Simulation design

::: {.bullets data-latex=""}
\textbf{This section organizes the study under ADEMP: DGP
(§3.1), estimands (§3.2), methods (§3.3), factorial grid (§3.4),
parameters and performance measures (§3.5), and computing
environment (§3.6).}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item ADEMP provides the structural scaffold for the simulation
  design description.
\item Reproducibility metadata (R version, packages, RNG seed,
  elapsed time) are in §3.6.
\item §3.4 defines the principal $3 \times 3$ factorial and its
  design rationale.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
This section organizes the study under the ADEMP labels of
Morris et al. [-@morris2019simulation]. We begin in §3.1 with the
data-generating mechanisms, then turn in §3.2 to the estimands,
§3.3 to the methods (analysis-model specifications), §3.4 to the
factorial grid, and §3.5 to the parameter settings and performance
measures. Reproducibility metadata, including the R version, key
package versions, the RNG seed strategy, and elapsed time, are
recorded in §3.6 (Computing environment).
:::

## Aims

::: {.bullets data-latex=""}
\textbf{The primary aim is to identify the most robust analysis
specification across the DGP decay-form grid; the secondary aim
is to verify that rankings transfer across architectures.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item Primary: quantify the cost of carryover mis-specification
  and identify the most robust analysis strategy.
\item Secondary: verify that robustness rankings are
  architecture-invariant (Architecture A vs.\ B).
\item Architecture-invariance would justify a single default
  recommendation regardless of assumed biomarker biology.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
The primary aim of this study is to quantify the cost of
analysis-model carryover mis-specification in aggregated N-of-1
and crossover trials, and to identify which mitigation strategy
is most robust across the DGP decay-form grid. A secondary aim
is to verify that the robustness rankings transfer across the two
data-generating-process architectures (mean moderation versus
differential correlation), so that a practitioner can be
reasonably confident that the recommendation does not depend on
which biomarker biology is assumed.
:::

## Data-generating mechanisms

::: {.bullets data-latex=""}
\textbf{The two single-channel architectures (A and B) from the
companion manuscript are retained; the combined Architecture C is
out of scope here. They differ only in how the decay function
$\phi$ enters the DGP.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item Architecture A: $\phi$ enters the mean structure via
  effective exposure $D_{it}^{*}$.
\item Architecture B: $\phi$ enters the biomarker-response
  correlation $\text{Cor}(B_i, BR_{it}) = c_{bm} \cdot
  \phi(t_{sd})$.
\item Three decay forms for $\phi$ are evaluated: linear,
  exponential, and Weibull.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
We retain the two single-channel DGP architectures from the
companion manuscript. A separate companion manuscript formalizes a
combined Architecture C, which activates both the mean-moderation
and differential-correlation channels simultaneously; Architecture C
is outside the scope of the present study, which evaluates the
carryover specifications under each single-channel architecture in
isolation. In Architecture A (direct mean moderation), the
biomarker $B_i$
enters the mean structure as
$Y_{it} = \mu_0 + \beta_1 D_{it}^{*}(1 + \beta_{bm} B_i)
+ \epsilon_{it}$, where $D_{it}^{*}$ is the effective exposure at
timepoint $t$ under the chosen decay form. In Architecture B
(MVN differential correlation), the biomarker and biological
response are jointly drawn from a multivariate normal with
treatment-state-dependent correlation
$\text{Cor}(B_i, BR_{it}) = c_{bm} \cdot \phi(t_{sd})$, where
$\phi$ is the decay form.
:::

In both architectures the decay function $\phi : \mathbb{R}_+
\to [0, 1]$ maps time since drug discontinuation $t_{sd}$ to a
residual-effect multiplier. We evaluate three forms,
each parameterized to match a common half-life $t_{1/2}$:

### Linear decay

$$
\phi_{\text{lin}}(t_{sd}) =
  \max\left(0,\, 1 - \frac{t_{sd}}{2 t_{1/2}}\right)
$$

::: {.bullets data-latex=""}
\textbf{The linear form is biologically implausible but
pragmatically defensible when only a complete off-drug clearance
time is known.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item Residual effect reaches zero exactly at $t_{sd} = 2
  t_{1/2}$, unlike the exponential tail.
\item Inconsistent with first-order elimination kinetics.
\item Represents the analyst's assumption that carryover has
  fully cleared by a known time horizon.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
Residual effect vanishes completely at $t_{sd} = 2 t_{1/2}$.
The linear form is biologically implausible for drugs with
first-order elimination kinetics, but it represents a defensible
pragmatic assumption when the only available information is that
carryover has fully worn off by some specified time $X$.
:::

### Exponential decay

$$
\phi_{\text{exp}}(t_{sd}) =
  \exp\left(-\lambda t_{sd}\right), \qquad
  \lambda = \frac{\ln 2}{t_{1/2}}
$$

::: {.bullets data-latex=""}
\textbf{The exponential form is the natural reference decay because
it corresponds to first-order elimination kinetics.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item Used throughout the companion manuscript and implemented in
  \texttt{implementations/tidyverse/R/functions.R}.
\item Corresponds to classical pharmacokinetic first-order
  elimination.
\item Retained as the reference against which linear and Weibull
  forms are compared.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
This is the form used throughout the companion manuscript and
currently implemented in `implementations/tidyverse/R/functions.R`.
It corresponds to first-order elimination kinetics and is the
default assumption in classical pharmacokinetics; we retain it
here as the natural reference against which the linear and Weibull
forms are compared.
:::

### Weibull decay

$$
\phi_{\text{weib}}(t_{sd}) =
  \exp\left(-\left(\lambda_w t_{sd}\right)^{k}\right), \qquad
  \lambda_w = \frac{(\ln 2)^{1/k}}{t_{1/2}}
$$

::: {.bullets data-latex=""}
\textbf{The Weibull shape parameter $k$ controls whether the
elimination tail is heavier or lighter than exponential.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item $k < 1$: slower elimination with a prolonged
  low-concentration tail (evaluated at $k = 0.7$).
\item $k > 1$: accelerated off-drug clearance via saturable clearance or
  active-transporter kinetics (evaluated at $k = 1.5$).
\item $k = 1$ recovers the exponential form exactly.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
A shape parameter $k$ governs whether the elimination tail is
heavier ($k < 1$, slow elimination) or lighter ($k > 1$,
accelerated off-drug clearance) than the exponential, which is recovered
exactly at $k = 1$. We report results at $k = 0.7$, representing
a drug with a prolonged low-concentration tail, and at $k = 1.5$,
which captures drugs with saturable clearance or active-transporter
kinetics that clear more rapidly once the concentration falls below
the saturation threshold.
:::

::: {.bullets data-latex=""}
\textbf{All three decay forms are implemented via
\texttt{carryover\_decay()} in the tidyverse collection, with
exponential as the backward-compatible default.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item \texttt{carryover\_form} and \texttt{weibull\_shape} fields
  are plumbed through \texttt{apply\_carryover\_to\_component()},
  \texttt{build\_correlation\_matrix()}, and
  \texttt{build\_sigma\_matrix()}.
\item The \texttt{original-extended} collection supports only
  exponential; cross-implementation validation is therefore
  limited to exponential cells.
\item Backward compatibility is preserved: absent
  \texttt{carryover\_form} defaults to exponential.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
**Implementation note.** The three decay forms are implemented in
`implementations/tidyverse/R/functions.R` via the
\texttt{carryover\_decay()} helper function, with a \texttt{carryover\_form} and
\texttt{weibull\_shape} pair of \texttt{model\_param} fields plumbed through
\texttt{apply\_carryover\_to\_component()}, \texttt{build\_correlation\_matrix()},
and \texttt{build\_sigma\_matrix()}. Backward compatibility is preserved
by defaulting to the exponential form when \texttt{carryover\_form} is
absent. We note that the `implementations/original-extended/`
collection has not been updated and supports only the exponential
form; cross-validation across implementations is therefore
limited to exponential cells, which is sufficient for pipeline
verification but not for the full decay-form comparison.
:::

## Estimands

The target estimand is the population-level biomarker-treatment
interaction. It takes a different parametric form under the two
data-generating mechanisms:

- **Architecture A.** The interaction estimand is the regression
  coefficient $\beta_{bm}$ on the centered biomarker $\times$
  drug-exposure product, on the symptom scale.
- **Architecture B.** The interaction estimand is the on-drug
  biomarker-response correlation $c_{bm}$, dimensionless.

::: {.bullets data-latex=""}
\textbf{The two architectures yield non-commensurable estimands
that are harmonised to a common calibrated effect-size scale.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item Architecture A estimand: regression coefficient $\beta_{bm}$
  on centered biomarker $\times$ exposure product.
\item Architecture B estimand: on-drug biomarker-response
  correlation $c_{bm}$, dimensionless.
\item Both calibrated to a common effect-size scale (expected
  response difference per SD biomarker at $t_{1/2} = 0$);
  derivation in \texttt{docs/19-mean-moderation-implementation-notes.tex}.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
The two estimands are not directly commensurable. To enable
cross-architecture comparison of bias and power, we calibrate
both architectures to a common population effect-size scale,
defined as the expected difference in mean on-drug response per
unit biomarker standard deviation at $t_{1/2} = 0$. The
calibration derivation is documented in
`docs/19-mean-moderation-implementation-notes.tex`. Bias and
empirical standard error in the Results section are reported on
this calibrated scale; power and Type I error are reported on
the native test-statistic scale.

The null estimand for Type I error verification is
$\beta_{bm} = 0$ under Architecture A and $c_{bm} = 0$ under
Architecture B, each realized by the $c_{bm} = 0$ row of the
parameter grid.
:::

## Methods: analysis-model carryover specifications

The analysis model is `nlme::lme(Sx ~ bm * X + t, random = ~1
| ptID, correlation = corCAR1(form = ~ t | ptID))` for all
specifications. The three specifications differ in the drug
exposure predictor $X$:

::: {.bullets data-latex=""}
\textbf{All three specifications share a common \texttt{nlme::lme}
formula and \texttt{corCAR1} residual structure; they differ only
in the drug exposure predictor $X$.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item Formula: \texttt{Sx \textasciitilde{} bm * X + t}, random
  intercept \texttt{\textasciitilde{}1 | ptID}.
\item Residual correlation: \texttt{corCAR1(form =
  \textasciitilde{} t | ptID)}.
\item The three specifications (M1, M2, M3) differ only in how
  $X$ is constructed.
\end{itemize}
:::

### M1: binary-treatment (carryover ignored)

::: {.bullets data-latex=""}
\textbf{M1 uses the binary on-drug indicator and ignores carryover
entirely, serving as the dominated reference specification.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item $X = D_{it}$: takes value 1 on-drug, 0 off-drug.
\item Makes no attempt to model residual drug effect during
  the off-drug period.
\item Serves as the lower-power reference against which M2 and
  M3 are compared.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
$X = D_{it}$, the on-drug indicator. This is the naive
specification that disregards any residual drug effect during
the off-drug period; it serves as the reference against which the power cost
of ignoring carryover is measured.
:::

### M2: exposure-weighted (continuous predictor)

::: {.bullets data-latex=""}
\textbf{M2 uses a continuous exposure-weighted predictor that
commits to a specific decay form and half-life, and is the current
pmsimstats-ng default.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item $X = D_{bc,it}$: decays off-drug according to the
  analyst's assumed form and half-life.
\item Well-matched when the analyst's assumed form coincides with
  the true DGP form; mis-specified otherwise.
\item Sensitivity of M2 to assumed-versus-true mismatch is
  evaluated in Tier 2 block S2.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
$X = D_{bc,it}$, the continuous exposure-decayed predictor
constructed to match the analyst's assumed decay form and
half-life. This is the current pmsimstats-ng default
specification. When the analyst's assumed form coincides with the
true DGP form the predictor is well-matched; when it does not,
the predictor is mis-specified in a way that the Tier 2 block S2
is designed to probe.
:::

### M3: lagged-treatment (estimated carryover term)

::: {.bullets data-latex=""}
\textbf{M3 augments the binary indicator with a lagged-treatment
term that estimates carryover magnitude without assuming a decay
form.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item $L_{it} = D_{i,t-1}(1 - D_{it})$ marks off-drug timepoints
  immediately following on-drug timepoints.
\item Carryover magnitude is estimated from the data; no
  functional form or half-life assumption is required.
\item Follows the classical crossover-trial framework of Jones
  and Kenward [-@jones2014crossover].
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
$X = D_{it}$ augmented with a lagged-treatment indicator
$L_{it} = D_{i,t-1}(1 - D_{it})$ that marks off-drug timepoints
following an on-drug timepoint. The carryover magnitude is
estimated from the data rather than assumed a priori, so the
specification does not require the analyst to posit a functional
decay form. This approach follows the classical crossover-trial
framework surveyed in [@jones2014crossover]. We note one structural
feature of this specification that bears on its interpretation: the
lagged term $L_{it}$ enters the model only as a main-effect
nuisance covariate, and the biomarker interaction tested under M3
remains $\text{bm:}D_{it}$, the same coefficient as under M1.
Consequently M3 differs from M1 only through the variance that
$L_{it}$ absorbs in the off-drug timepoints, and the two
specifications are expected to track closely; M3 is best understood
as M1 with a carryover nuisance control rather than as a
fundamentally different interaction model.
:::

### Construction of $D_{bc}$ and the role of the analyst's assumed half-life

::: {.bullets data-latex=""}
\textbf{$D_{bc}$ halves every $\hat{t}_{1/2}$ weeks off-drug; the
formula handed to \texttt{lme} is invariant to the analyst's
half-life assumption.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item $D_{bc,it} = (1/2)^{t_{sd}/\hat{t}_{1/2}}$ at off-drug
  timepoints.
\item The model formula \texttt{Sx \textasciitilde{} bm + t +
  Dbc + bm:Dbc} is the same regardless of $\hat{t}_{1/2}$.
\item The half-life assumption is encoded in the predictor
  values, not in the formula.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
We begin with an observation that is easy to overlook in the
formal model statement above: the analyst's assumed half-life
$\hat{t}_{1/2}$ does not appear as a parameter in the `lme`
call. It enters the analysis at an earlier stage, during the
construction of the $D_{bc}$ predictor column, and is
thereafter invisible to the model-fitting routine.

At on-drug timepoints, $D_{bc,it} = 1$ regardless of the assumed
half-life. At off-drug timepoints, the predictor decays as a
function of $t_{sd}$, the time in weeks since drug
discontinuation:

$$
D_{bc,it} \;=\; \exp\!\left(-\frac{\ln 2}{\hat{t}_{1/2}}
  \, t_{sd,it}\right)
  \;=\; \left(\tfrac{1}{2}\right)^{t_{sd,it}/\hat{t}_{1/2}}
$$

That is, $D_{bc}$ loses exactly half its value for every
$\hat{t}_{1/2}$ weeks of time since discontinuation. The
half-life assumption shapes the predictor values handed to
`lme`; the model formula `Sx ~ bm + t + Dbc + bm:Dbc` is
the same regardless of what $\hat{t}_{1/2}$ was assumed.
:::

To illustrate, consider a participant in a stylized crossover
arm who is on drug for three weekly measurement occasions and
then enters a three-week off-drug period. Table~\@ref(tab:toy-dbc)
shows the $D_{bc}$ values at the three off-drug timepoints
under three analyst assumptions, holding the true DGP
half-life fixed at one week.

```{r toy-dbc, echo = FALSE}
toy <- tibble::tribble(
  ~`$t_{sd}$ (wk)`,
    ~`$\\hat{t}_{1/2} = 0.5$ wk`,
    ~`$\\hat{t}_{1/2} = 1.0$ wk (truth)`,
    ~`$\\hat{t}_{1/2} = 2.0$ wk`,
  "1",  "0.250",  "0.500",  "0.707",
  "2",  "0.063",  "0.250",  "0.500",
  "3",  "0.016",  "0.125",  "0.354"
)
knitr::kable(
  toy, booktabs = TRUE, escape = FALSE,
  caption = paste0(
    'Illustrative $D_{bc}$ values at three off-drug timepoints ',
    'under three analyst-assumed half-lives, exponential decay, ',
    'true $t_{1/2} = 1.0$ week. ',
    'On-drug timepoints have $D_{bc} = 1$ under all assumptions. ',
    'Row values are $(1/2)^{t_{sd}/\\hat{t}_{1/2}}$.'
  )
)
```

::: {.bullets data-latex=""}
\textbf{The analyst's assumed half-life shapes the $D_{bc}$
predictor values handed to \texttt{lme} but is invisible to the
model-fitting routine itself.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item Under-assumed half-life: off-drug timepoints receive
  near-zero $D_{bc}$ and contribute almost nothing to the
  \texttt{bm:Dbc} estimate.
\item Over-assumed half-life: off-drug timepoints retain
  substantial weight, making $D_{bc}$ resemble a graded
  on-drug indicator.
\item M1 and M3 are invariant to $\hat{t}_{1/2}$ by
  construction; S2 contrasts a continuously indexed model family
  against two half-life-free reference specifications.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
When the analyst assumes $\hat{t}_{1/2} = 0.5$ weeks, the
off-drug timepoints are assigned $D_{bc}$ values near zero by
the second off-drug week and contribute almost nothing to the
`bm:Dbc` interaction estimate. When the analyst assumes
$\hat{t}_{1/2} = 2.0$ weeks, those same timepoints retain
substantial weight throughout the off-drug period, and the predictor
becomes closer in shape to a continuously graded version of the
binary on-drug indicator.

Two consequences follow for interpreting the S2 sensitivity
block. First, M2 is effectively a different model for each value
of $\hat{t}_{1/2}$: the predictor values handed to `lme` change
with the analyst's assumption, even though the formula is
identical. Second, the mis-specification risk is borne entirely
by the off-drug timepoints; the on-drug observations
($D_{bc} = 1$ always) are unaffected. Specifications M1 and M3
use the binary on-drug indicator $D_b$ and the lagged indicator
$L$ respectively, neither of which involves a half-life
parameter, so both are invariant to $\hat{t}_{1/2}$ by
construction. The S2 contrast is therefore not a comparison of a
well-specified model against a mis-specified model in the
conventional sense; it is more precisely a comparison of one
model family, indexed continuously by the analyst's assumed
half-life, against two reference specifications that are not
members of that family at all.
:::

## Factorial grid

The principal comparison is a $3 \times 3$ factorial of DGP
decay form against analysis-model specification, replicated
under each DGP architecture:

```{r grid-table}
grid <- tibble::tribble(
  ~`DGP decay`,   ~`M1 binary`, ~`M2 exposure-weighted`, ~`M3 lagged-treatment`,
  "Linear",       "cell 1",     "cell 2",                "cell 3",
  "Exponential",  "cell 4",     "cell 5 (matched)",      "cell 6",
  "Weibull",      "cell 7",     "cell 8",                "cell 9"
)
knitr::kable(
  grid,
  caption = "Principal $3 \\times 3$ simulation grid. Cell 5 is the
    matched reference: exponential DGP analyzed under the current
    pmsimstats-ng default. Off-diagonal cells quantify the cost of
    DGP-analysis mis-specification. The full grid is crossed with
    Architecture $\\in \\{A, B\\}$."
)
```

## Parameter settings

| Parameter | Values |
|---|---|
| Biomarker moderation ($c_{bm}$ / $\beta_{bm}$) | 0.0, 0.30, 0.45 |
| Carryover half-life ($t_{1/2}$) | 0.0, 0.5, 1.0 weeks |
| Weibull shape ($k$) | 0.7, 1.0, 1.5 (Weibull form only) |
| Sample size ($N$) | 35, 70 |
| Trial design | CO, N-of-1 (Hybrid), OL+BDC |
| DGP architecture | A, B |
| Replicates per cell | 500 (target); 50 for development |

The 540-cell count follows from these factors with one coupling: the
Weibull shape varies only within the Weibull decay form, because the
linear and exponential forms carry no shape parameter. The decay-form
and shape combinations therefore number five (linear; exponential;
Weibull at $k \in \{0.7, 1.0, 1.5\}$), and the principal grid is
$5 \times 3 \times 3 \times 2 \times 2 \times 3 = 540$ cells
(decay-shape $\times$ half-life $\times$ design $\times$ sample size
$\times$ architecture $\times$ biomarker level). A naive product that
crossed the three decay forms with three shapes would over-count; the
shape factor is nested within the Weibull form, not crossed with all
three.

::: {.bullets data-latex=""}
\textbf{Two anchor conditions in the parameter grid serve
diagnostic roles: $c_{bm} = 0$ for Type I error and
$t_{1/2} = 0$ for best-case power.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item $c_{bm} = 0$: null row for Type I error verification,
  targeting nominal 5\%.
\item $t_{1/2} = 0$: no-carryover reference providing best-case
  power at each $(N, \text{design}, c_{bm})$ cell.
\item Non-zero $t_{1/2}$ rows directly quantify the power cost
  of carryover relative to the no-carryover reference.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
The $c_{bm} = 0$ condition serves as the Type I error check,
targeting the nominal 5% level. The $t_{1/2} = 0$ condition
removes carryover entirely and provides a best-case power
reference at each $(N, \text{design}, c_{bm})$ combination,
against which the power cost of non-zero carryover can be
directly read off.
:::

## Performance measures

For each cell we report the following Morris-style performance
measures [@morris2019simulation, Tables 5--6]:

1. **Power** to reject $H_0: \beta_{\text{bm:X}} = 0$ at
   $\alpha = 0.05$.
2. **Type I error** at the null estimand ($c_{bm} = 0$ row).
3. **Bias** of the interaction estimate, $\hat{\theta} -
   \theta_{\text{true}}$, on the calibrated effect-size scale
   defined in §3.2.
4. **Empirical SE** across replicates within each cell.
5. **Coverage** of the nominal 95% Wald confidence interval for
   the interaction coefficient, computed against the calibrated
   estimand
   ($\theta_{\text{true}} = -c_{bm} \cdot \sigma_{BR}$).
6. **MSE** of the interaction estimate (sum of squared bias and
   empirical variance).
7. **Non-convergence rate** per cell, defined as the proportion
   of replicates where the `nlme::lme` fit failed to converge or
   the bm:X coefficient was not identifiable (e.g.\ M1 on
   open-label (OL) designs, where there are no off-drug observations).
8. **MCSE** for each of the above, computed using the
   formulas of Morris et al. [-@morris2019simulation, Table 6]
   and reported alongside every point estimate in tables and
   figures.

::: {.orig data-latex=""}
\textbf{Eight Morris-style performance measures are reported per
cell, each accompanied by its MCSE.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item Power, Type I error, bias, empirical SE, coverage, MSE,
  non-convergence rate, and MCSE for each measure.
\item Coverage is computed against the calibrated estimand
  $\theta_{\text{true}} = -c_{bm} \cdot \sigma_{BR}$.
\item MCSEs use the formulas of Morris et al. (2019, Table 6).
\end{itemize}
:::

::: {.bullets data-latex=""}
\textbf{$n_{sim} = 500$ is calibrated to keep MCSE on power
below 0.022, with $n_{sim} = 50$ reserved for pipeline
validation only.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item Production run: $n_{sim} = 500$, MCSE
  $\leq 0.022$ at $p = 0.5$.
\item Development run: $n_{sim} = 50$, MCSE
  $\approx 0.07$; pipeline validation only.
\item All reported results derive from the production-scale run.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
The replicate count $n_{sim} = 500$ is chosen so that the Monte
Carlo standard error on power at $p = 0.5$ does not exceed 0.022
($\sqrt{0.5 \cdot 0.5 / 500} \approx 0.022$). The development
scale ($n_{sim} = 50$, MCSE $\approx 0.07$ at $p = 0.5$) is
used only for pipeline validation; all results reported in the
Results section derive from the production run at $n_{sim} = 500$.
:::

## Computing environment

::: {.bullets data-latex=""}
\textbf{The tidyverse collection is the only implementation
supporting all three decay forms; common-random-number seeding
enables paired contrasts across M1, M2, and M3.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item All three decay forms are supported only in
  \texttt{implementations/tidyverse/}; cross-implementation
  parity checks are limited to exponential cells.
\item Within-cell common random numbers reduce MC variance for
  the M2$-$M1 and M3$-$M1 paired contrasts.
\item Production run (540 cells, $n_{sim} = 500$): 128 minutes
  on eight cores; commit SHA \texttt{7018b59}, 2026-04-15.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
Simulations are orchestrated from
`analysis/scripts/carryover-sensitivity/` and use
`implementations/tidyverse/` as the primary implementation,
which is the only collection in the repository that supports
linear and Weibull decay forms. Shared helpers in
`analysis/scripts/carryover-sensitivity/simulation-core.R`
define design presets (OL, CO, Hybrid, OL+BDC), multi-path data
generation, long-form preparation with `Db`, `Dbc`, and `L`
columns, and the three analysis-specification fitters.

**Software.** R 4.5.3 with `nlme` 3.1, `dplyr` 1.1, `tidyr` 1.3,
`purrr` 1.0, `furrr` 0.3, `future` 1.34, `corpcor` 1.6, `MASS`
7.3, `Matrix` 1.7, `tibble` 3.2, and `data.table` 1.16. The
`implementations/tidyverse/` collection commit SHA at the time of
the production run is recorded in each output `.rds` file.

**RNG seed strategy.** Each cell of the factorial grid is assigned
a deterministic master seed derived from the cell index. Within a
cell, the same simulated dataset is used to fit all three analysis
specifications (common random numbers across M1, M2, and M3), so
that the M2$-$M1 and M3$-$M1 contrasts are estimated with reduced
Monte Carlo variance relative to independent draws. Seeds are
independent across cells. Replicate indices within a cell are
seeded as \texttt{master + replicate\_index}, so that any subset of
replicates can be re-derived without re-running the entire cell.
The seed assignment is logged in the output `.rds` file alongside
the remaining reproducibility metadata.

**Parallelism.** Parallel execution uses \texttt{furrr::future\_map} with
the `multicore` backend. Sigma matrices are cached per unique
(design, $t_{1/2}$, decay form) tuple before the replicate loop
begins, avoiding redundant matrix construction across workers.

**Elapsed time.** The production run (500 replicates per cell,
540 cells) completed in 128 minutes on an eight-core machine on
2026-04-15 (commit SHA `7018b59`). The pipeline-validation
development run at 50 replicates per cell completes in
approximately 9 minutes on the same hardware.

**Pre-registration and reproducibility.** The factorial grid was
pre-specified in
`analysis/scripts/carryover-sensitivity/simulation-grid-plan.md`
and committed before the production run was executed. The
production run command is
`Rscript analysis/scripts/carryover-sensitivity/01-run-factorial.R`;
the Tier 2 sensitivity blocks are produced by
`Rscript analysis/scripts/carryover-sensitivity/04-run-sensitivity-blocks.R`.
Each output `.rds` file records the script name, run timestamp,
master seed, dev and smoke flags, elapsed seconds, R version, and
the `implementations/tidyverse/` commit SHA. The `renv.lock` at
the repository root pins all R-package versions. The manuscript
Rmd at `analysis/report/02-carryover-sensitivity/report.Rmd` does
not re-execute any simulation code; it loads the pre-computed
checkpoints written by the summarization script.
:::

# Results

::: {.bullets data-latex=""}
\textbf{All reported results derive from the 540-cell
production-scale run at $n_{sim} = 500$; MCSE bounds apply
throughout.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item MCSE on power $\leq 0.022$ at $p = 0.5$; differences
  smaller than the relevant MCSE should not be
  over-interpreted.
\item Coverage of nominal 95\% Wald CIs is reported for the
  reference cell and computed across the full Tier 1 grid.
\item Non-convergence rates are reported per cell in the tables
  that follow.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
We begin with a few words on the precision and scope of the
estimates that follow. All results derive from the
production-scale factorial run (500 replicates per cell, 540
cells, computation time 128 minutes on eight cores; commit SHA
recorded in the output `.rds` metadata). Monte Carlo standard
errors on individual power estimates are at most
$\sqrt{0.5 \cdot 0.5 / 500} \approx 0.022$ at $p = 0.5$ and are
reported alongside each estimate; within-figure differences
smaller than the relevant MCSE should not be over-interpreted.
Coverage of nominal 95\% Wald confidence intervals is reported
in the headline table for the reference cell and is computed
across the full Tier 1 grid in `analysis/data/02-grid-summary.rds`
for downstream use. Non-convergence rates per cell are reported
in the tables below.
:::

## Headline performance table {#sec-headline}

::: {.bullets data-latex=""}
\textbf{The headline table provides the complete Morris-style
performance profile for the reference cell across all three
specifications and carryover levels.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item Reference cell: Architecture B, exponential DGP, Hybrid
  design, $N = 70$, $c_{bm} = 0.45$.
\item Non-convergence is zero at every reference cell.
\item MCSEs are reported as $\pm$ beside each point estimate
  throughout.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
Table \@ref(tab:headline) reports the Morris-style performance
measures at the reference cell (Architecture B, exponential DGP,
Hybrid design, $N = 70$, $c_{bm} = 0.45$) across the three
analysis specifications and three carryover half-lives.
Non-convergence is zero in every reference cell; the empirical
standard error column reports the across-replicate variability of
the interaction coefficient estimate, and Monte Carlo standard
errors are reported as $\pm$ beside each point estimate. One
caveat applies to the bias, MSE, and coverage columns: these are
computed against a single calibrated target,
$\theta_{\text{true}} = -c_{bm} \cdot \sigma_{BR} = -3.6$, which is
the population value of the M2 (continuous-$D_{bc}$) interaction
coefficient. Specifications M1 and M3 estimate the coefficient on
$\text{bm:}D_{it}$, a different parameter, so their apparent bias
and under-coverage in this table reflect the mismatch between their
estimand and M2's target rather than a defect of the M1/M3
estimators. The bias, MSE, and coverage columns are therefore not
directly comparable across specifications; only power and Type I
error, which test each specification's own coefficient against
zero, support a like-for-like comparison.
:::

```{r headline, echo = FALSE}
suppressPackageStartupMessages({library(dplyr)})
hd <- readRDS(here::here(
  'analysis/data/02-grid-summary.rds'))$summary |>
  filter(carryover_form == 'exponential', dgp_arch == 'mvn',
         design == 'Hybrid', N == 70, c_bm == 0.45) |>
  mutate(
    `t1/2` = t1half,
    `Power (MCSE)` = sprintf('%.3f (%.3f)', power, mc_se_power),
    `Bias (MCSE)` = sprintf('%+.3f (%.3f)', bias, mc_se_bias),
    `Empirical SE` = sprintf('%.3f', empirical_se),
    `Coverage (MCSE)` = sprintf('%.3f (%.3f)',
                                  coverage, mc_se_coverage),
    `MSE (MCSE)` = sprintf('%.3f (%.3f)', mse, mc_se_mse),
    `Non-conv` = sprintf('%.3f', non_convergence_rate)
  ) |>
  arrange(`t1/2`, spec) |>
  mutate(spec = spec_display(spec)) |>
  select(`t1/2`, spec, `Power (MCSE)`, `Bias (MCSE)`,
         `Empirical SE`, `Coverage (MCSE)`, `MSE (MCSE)`,
         `Non-conv`)

knitr::kable(
  hd, booktabs = TRUE,
  caption = 'Headline performance at the reference cell
    (Architecture B, exponential DGP, Hybrid design, $N = 70$,
    $c_{bm} = 0.45$, $n_{sim} = 500$). Power is the rejection
    rate of $H_0: \\beta_{bm:X} = 0$ at $\\alpha = 0.05$.
    Bias is on the calibrated effect-size scale,
    $\\theta_{\\text{true}} = -c_{bm} \\cdot \\sigma_{BR} = -3.6$
    (per docs/19 calibration). Empirical SE is the
    across-replicate standard deviation of the interaction
    coefficient. MSE is the sum of squared bias and empirical
    variance. MCSEs are reported per Morris et al.
    (2019, Table 6). Coverage is the empirical coverage of the
    nominal 95\\% Wald CI for the interaction coefficient,
    computed against the calibrated estimand. Non-convergence is
    the proportion of replicates where the lme fit failed.'
)
```

## Matched-decay baseline

::: {.bullets data-latex=""}
\textbf{The matched-decay baseline confirms pipeline fidelity and
shows a power decline from 0.988 at no carryover to 0.860 at
$t_{1/2} = 1.0$ weeks.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item Matched cell (exponential DGP, M2, Architecture B,
  $N = 70$, $c_{bm} = 0.45$, Hybrid): power
  $0.988 \to 0.948 \to 0.860$ across $t_{1/2}
  \in \{0, 0.5, 1.0\}$ weeks.
\item All 500 replicates converged at every $t_{1/2}$.
\item Consistent with companion manuscript's Architecture B
  result (0.82 at $t_{1/2} = 1.0$ weeks).
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
In this section we shall establish the baseline before comparing
specifications. Under the matched cell (exponential DGP, M2
exposure-weighted analysis, Architecture~B, $N = 70$,
$c_{bm} = 0.45$, Hybrid design), power declines from
$0.988 \pm 0.005$ at $t_{1/2} = 0$ to $0.948 \pm 0.010$ at
$t_{1/2} = 0.5$ to $0.860 \pm 0.016$ at $t_{1/2} = 1.0$ weeks
(point estimate $\pm$ MCSE, $n_{sim} = 500$; MCSEs
computed as $\sqrt{\hat{p}(1-\hat{p})/n_{sim}}$ per Morris et
al. [-@morris2019simulation, Table 6]). All 500 replicates
converged in the matched cell at every $t_{1/2}$. The value at
$t_{1/2} = 1.0$ is consistent with Hendrickson's reported power of
0.82 for this cell [@hendrickson2020], confirming that the
simulation pipeline reproduces the baseline result and that the
calibration is internally coherent. (The companion manuscript
anchors at a different cell, $N = 35$ with the OL+BDC design, and
is therefore not directly comparable here.)
:::

## Power by analysis specification

```{r fig-power-by-spec, echo = FALSE, fig.cap = "Power versus carryover half-life under the exponential DGP, stratified by analysis specification. The matched specification M2 tracks closely with the nominally inferior M1 and M3 at low half-lives but retains a modest advantage at $t_{1/2} = 1.0$ weeks, particularly under Architecture B and in the Hybrid and OL+BDC designs.", out.width = "90%"}
knitr::include_graphics("../../figures/02-power-by-spec.pdf")
```

::: {.bullets data-latex=""}
\textbf{Specifications diverge progressively with carryover
half-life; M2's advantage over M1 and M3 emerges at
$t_{1/2} = 0.5$ weeks and grows at $t_{1/2} = 1.0$ weeks.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item All three specifications are near-identical at
  $t_{1/2} = 0$; divergence is proportional to carryover
  strength.
\item M1 and M3 are effectively indistinguishable across the
  exponential-DGP grid.
\item M2's advantage is consistent across Hybrid and OL+BDC
  design strata, not an artifact of the reference cell.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
The three analysis specifications yield near-identical power at
$t_{1/2} = 0$, where there is no carryover to model, and diverge
progressively as carryover half-life increases. The matched
specification M2 retains a modest but consistent advantage under
Architecture~B and the Hybrid and OL+BDC designs; M1 and M3 are
effectively indistinguishable across the exponential-DGP grid.
The figure makes clear that the M2 advantage is not an artifact
of the reference cell: it emerges at $t_{1/2} = 0.5$ weeks and
grows through $t_{1/2} = 1.0$ weeks in each design stratum.
:::

## Decay-form mis-specification

```{r fig-heatmap-mismatch, echo = FALSE, fig.cap = "Power at the critical cell ($t_{1/2} = 1.0$ weeks, Architecture B, Hybrid design, $N = 70$, $c_{bm} = 0.45$) across DGP decay forms (rows) and analysis specifications (columns). Under the Weibull heavy-tailed DGP ($k = 0.7$) and the linear DGP, the exposure-weighted (M2) specification loses its advantage because its assumed exponential decay no longer matches truth.", out.width = "70%"}
knitr::include_graphics("../../figures/02-heatmap-matched-vs-mismatched.pdf")
```

::: {.bullets data-latex=""}
\textbf{The heatmap exposes two directions of mis-specification:
wrong analysis specification given the true DGP, and wrong DGP
assumed by the analyst.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item Matched reference: exponential DGP $\times$ M2 analysis,
  power $0.860 \pm 0.016$.
\item Same-row off-diagonal cells: power penalty of M1 or M3
  when the true DGP is exponential.
\item M2-column off-diagonal cells: cost of assuming exponential
  when the true DGP is linear or Weibull.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
The heatmap cell at the intersection of the exponential DGP row
and the M2 analysis column is the matched reference
(power $0.860 \pm 0.016$). Cells in the same row but other
columns quantify the power penalty of analyzing under the wrong
specification when the DGP is exponential; cells in the M2 column
but other rows quantify the cost of assuming exponential decay
in the analysis when the true DGP is linear or Weibull. The
Discussion takes up the practical interpretation of both
off-diagonal directions. This is taken up further in §5.2.
:::

## Type I error control and cluster-robust recalibration

```{r fig-type1, echo = FALSE, fig.cap = "Empirical Type I error rates under the null ($c_{bm} = 0$), pooled across DGP decay forms and carryover half-lives, stratified by design and analysis specification. Boxplots cover the distribution across cells at $N = 70$; the dashed reference line marks the nominal 0.05 level.", out.width = "95%"}
knitr::include_graphics("../../figures/02-type1-boxplots.pdf")
```

::: {.bullets data-latex=""}
\textbf{Type I error is conservative, not inflated: mean 0.039
across 540 null cells, the working-covariance conservatism the
companion study documents rather than ideal control.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item Mean empirical Type I error 0.039, below nominal 0.05: the
  conservative side of the documented mis-calibration.
\item Cell-level maximum 0.084 is within the 540-cell
  multiplicity range; no specification or architecture inflates
  the rate.
\item The conservatism is removed by a cluster-robust standard
  error (block S6, below) and leaves the comparison intact.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
Mean empirical Type I error across all 540 null cells is 0.039,
below the nominal 0.05 level. The cell-level maximum is 0.084,
which falls within the binomial Monte Carlo 95% interval
$[0.033, 0.073]$ for nominal 0.05 at $n_{sim} = 500$ once the
multiplicity across 540 cells is acknowledged (the expected
maximum among 540 independent binomial draws of size 500 at
$p = 0.05$ exceeds 0.07). The shortfall below nominal is not
sampling slack but the conservatism of the model-based
interaction test that the companion calibration study
[@pmsimstats_calibration2026] traces to the corCAR1 working
covariance: its standard error is over-estimated by six to ten
percent and the realized rejection rate falls correspondingly.
The conservatism is uniform across specifications and
architectures, which is why it leaves the comparison intact;
the cluster-robust recalibration reported below (block S6) confirms
that a CR2 cluster-robust standard error restores the rejection rate
to nominal without disturbing the specification ranking. Table \@ref(tab:tab-type1-cr2) makes the
per-specification conservatism and its cluster-robust correction
explicit at the null cell.
:::

```{r tab-type1-cr2, echo = FALSE}
suppressPackageStartupMessages(library(dplyr))
t1 <- readRDS(here::here(
  'analysis/scripts/carryover-sensitivity/output/diag-null-type1-cr2.rds'))
tab <- t1$summary |>
  transmute(
    Specification    = spec_display(spec),
    `$\\kappa$ model` = sprintf('%.2f', mod_se_ratio),
    `Type I model`    = sprintf('%.3f', type1_mod),
    `$\\kappa$ CR2`    = sprintf('%.2f', cr2_se_ratio),
    `Type I CR2`      = sprintf('%.3f', type1_cr2))
knitr::kable(tab, booktabs = TRUE, escape = FALSE,
  caption = 'Conservatism of the model-based interaction test and its
    cluster-robust correction, at the null cell ($c_{bm} = 0$,
    exponential DGP, $t_{1/2} = 1.0$ weeks, Hybrid design, $N = 70$,
    Architecture B, 3{,}000 replicates). $\\kappa$ is the ratio of the
    mean standard error to the empirical standard deviation of the
    interaction estimate; $\\kappa > 1$ indicates a conservative test.
    Type I is the empirical rejection rate of
    $H_0: \\beta_{bm:X} = 0$ at $\\alpha = 0.05$, with Monte Carlo
    standard error about 0.004. All three specifications are
    conservative under the model-based test, most so the
    exposure-weighted M2 ($\\kappa = 1.10$, Type I $= 0.029$); the CR2
    bias-reduced cluster-robust standard error restores $\\kappa$ to
    near one and Type I to nominal (0.045 to 0.049) for every
    specification.')
```

::: {.bullets data-latex=""}
\textbf{Cluster-robust recalibration recovers the power the conservative model-based test forfeits and preserves the specification ranking in the information-rich cells, while its precision cost dominates under heavy dropout.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item The companion calibration study shows the model-based interaction test is conservative; this block re-fits all three specifications under both the model-based and a CR2 cluster-robust standard error at the reference cell and four stress cells.
\item In the information-rich cells CR2 restores calibration ($\kappa \to 1$) and recovers two to five points of power while preserving the M2 $>$ M1, M3 ranking.
\item CR2 spends precision (Satterthwaite df falls from $\approx 480$ to $\approx 23$ to $\approx 5$); under heavy dropout the conservatism vanishes, CR2 turns mildly anti-conservative, and recalibration is moot because power has collapsed.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
The companion calibration study [@pmsimstats_calibration2026] establishes
that the model-based Wald test of the interaction in the lme model with a
corCAR1 residual is conservative: at the reference cell its standard error
is over-estimated by six to ten percent, its realized Type I error sits
near 0.03, and the conservatism is asymptotic, removed by a CR2
bias-reduced cluster-robust standard error. Two consequences bear on the
present study, which evaluates that same model-based test. The mean null
rejection rate reported above, 0.039, is not merely good control but this
conservatism surfacing across the grid; and the absolute power figures
throughout are correspondingly mild under-statements. The pre-registered block S6 re-fits
all three specifications under both the model-based test and the CR2
cluster-robust test, clustered on participant with Satterthwaite degrees
of freedom, at the reference cell and four stress cells (Table
\@ref(tab:tab-s6)), to ask whether recalibration disturbs the
specification ranking and what it costs.
:::

```{r tab-s6, echo = FALSE}
suppressPackageStartupMessages({library(dplyr)})
s6 <- readRDS(here::here(
  'analysis/scripts/carryover-sensitivity/output/diag-s6-cr2.rds'))
ord <- c('Reference (N=70)', 'Small N (N=35)',
         'High autocorr (rho=0.9)',
         'Half-life mis-spec (assume 0.25)', '30% MCAR dropout')
lab <- c('Reference, $N=70$', 'Small $N=35$',
         'High autocorr, $\\rho=0.9$', 'Half-life mis-spec',
         '30\\% MCAR dropout')
tab <- s6$summary |>
  filter(spec == 'A2') |>
  left_join(s6$gaps, by = 'cell') |>
  mutate(o = match(cell, ord)) |>
  arrange(o) |>
  transmute(
    Cell = lab[o],
    `$\\kappa$ model` = sprintf('%.2f', mod_se_ratio),
    `$\\kappa$ CR2`   = sprintf('%.2f', cr2_se_ratio),
    `Power model`     = sprintf('%.3f', power_mod),
    `Power CR2`       = sprintf('%.3f', power_cr2),
    `M2$-$M1 model`   = sprintf('%+.3f', gap_mod),
    `M2$-$M1 CR2`     = sprintf('%+.3f', gap_cr2),
    `CR2 df`          = sprintf('%.0f', cr2_df))
knitr::kable(tab, booktabs = TRUE, escape = FALSE,
  caption = 'Cluster-robust recalibration of the exposure-weighted
    specification M2 at the reference cell and four stress cells
    (500 replicates, $c_{bm} = 0.45$, Architecture B). $\\kappa$ is the
    ratio of the mean standard error to the empirical standard
    deviation of the interaction estimate; a value above one is a
    conservative test. The M2$-$M1 columns give the power gap between
    the two specifications under each inference, a check that
    recalibration preserves the ranking. CR2 df is the mean
    Satterthwaite degrees of freedom for the cluster-robust test; the
    model-based test carries roughly 480. Computed with the
    \\texttt{clubSandwich} package; see the companion calibration study
    for the mechanism. The S6 cells are an independent 500-replicate
    run, so the model-based M2 power at the reference cell (0.829)
    differs from the headline-grid value (0.860) by within one Monte
    Carlo standard error ($\\approx 0.016$); the two should not be
    read as conflicting.') |>
  kableExtra::kable_styling(latex_options = "scale_down")
```

```{r fig-type1-cr2, echo = FALSE, fig.cap = "Empirical Type I error at the null ($c_{bm} = 0$) under the model-based and CR2 cluster-robust inference, across the five S6 cells and the three specifications. The model-based test (open circles) is conservative, sitting near 0.03; the CR2 test (filled) is restored to the nominal 0.05 (dashed line), within the binomial Monte Carlo band (gray). 500 null replicates per cell.", out.width = "95%"}
knitr::include_graphics("../../figures/02-type1-cr2.pdf")
```

::: {.orig data-latex=""}
The effect of recalibration on test size is shown directly in
Figure \@ref(fig:fig-type1-cr2). Under the model-based test the null
rejection rate sits near 0.03 across all five cells and all three
specifications (mean 0.035), the same conservatism documented above;
the CR2 cluster-robust test restores it to a mean of 0.046, within
the binomial Monte Carlo interval around the nominal 0.05. This
matters for the study's central question. A test that is not
correctly sized cannot be preferred on power alone, because its
power is not comparable to a correctly-sized competitor's; the CR2
restoration is therefore the precondition that licenses the power
comparison among M1, M2, and M3. Under CR2 all three hold nominal
$\alpha$, so the M2 advantage documented above is a comparison among
correctly-sized tests, not an artifact of differential
conservatism.
:::

::: {.orig data-latex=""}
At the reference cell and the information-rich stress cells the result
matches the companion study. The model-based standard-error ratio
$\kappa$ runs from 1.07 to 1.26 (Table \@ref(tab:tab-s6)), the documented
conservatism; the CR2 ratio is restored to between 0.93 and 1.01; and the
correctly-sized test recovers two to five percentage points of power. The
M2 advantage is preserved throughout: the M2-minus-M1 gap is held or
slightly widened under recalibration, from $+0.080$ to $+0.092$ at the
reference cell, $+0.066$ to $+0.086$ at $N = 35$, and $+0.052$ to
$+0.032$ at $\rho = 0.9$, where every specification approaches the power
ceiling and the gaps compress. Mis-specifying the analyst half-life
leaves M1 and M3 untouched, as they do not use it, and costs M2 alone a
few points of power, recovered in part by recalibration. The ranking that
is this study's contribution is therefore invariant to the conservatism.
:::

::: {.orig data-latex=""}
Recalibration is neither free nor uniform. The CR2 Satterthwaite degrees
of freedom fall from the model-based value of roughly 480 to about 23 at
$N = 70$ and 12 at $N = 35$, the price of a working-covariance-agnostic
standard error. Under heavy dropout the trade reverses. At 30 percent
missingness the model-based test is no longer conservative, its
standard-error ratio near 0.98; the CR2 estimator turns mildly
anti-conservative, ratio near 0.85, on the few degrees of freedom that
remain (about five); and its power falls slightly below the model-based
test's. All three specifications collapse to about 0.13 power with a
negligible M2-minus-M1 gap of 0.01. There the recalibration is moot: the
information lost to dropout, not the calibration of the standard error,
governs the analysis, in keeping with this study's broader finding that
preventing dropout outweighs the choice of specification. The practical
recommendation follows. In the information-rich regime under
Architecture~B, the recommended analysis is the exposure-weighted
specification M2 reported with a CR2 cluster-robust standard error, which
is correctly sized and recovers the power the conventional test forfeits
to conservatism. We emphasize the scope of this claim: the companion
calibration study validated the cluster-robust recalibration and its
ranking-invariance only at the Architecture~B, Hybrid, $N = 70$ cells, so
the conclusion that recalibration preserves the ranking and that the
reported power figures are mildly conservative lower bounds applies to
those cells, not to the crossover design or to Architecture~A, where M2
does not lead in the first place.
:::


## Sensitivity analyses

::: {.bullets data-latex=""}
\textbf{Five Tier 2 marginal sensitivity blocks probe single-axis
perturbations around the Tier 1 reference configuration.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item Reference configuration: Architecture B, Hybrid, $N = 70$,
  $c_{bm} = 0.45$, exponential DGP, $t_{1/2} = 1.0$ weeks.
\item Each block varies one axis; all others are held at
  reference values.
\item Blocks are not pooled with Tier 1 or with each other;
  design rationale in \texttt{simulation-grid-plan.md}.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
Before going into the individual blocks, we note the general
design. Five Tier 2 marginal sensitivity blocks probe the
robustness of the Tier 1 ranking to single-axis perturbations
around the reference configuration (Architecture B, Hybrid
design, $N = 70$, $c_{bm} = 0.45$, exponential DGP,
$t_{1/2} = 1.0$ weeks). Each block varies one axis and holds the
others at their reference values. Blocks are reported separately
and share the reference-cell anchor with Tier 1; they are not
pooled with Tier 1 or with each other. The design rationale is
documented in
`analysis/scripts/carryover-sensitivity/simulation-grid-plan.md`.
:::

### S1: within-factor autocorrelation

```{r fig-sens-s1, echo = FALSE, fig.cap = "Power by analysis specification across AR(1) autocorrelation $\\rho \\in \\{0.5, 0.7, 0.8, 0.9\\}$. Hendrickson (2020) uses $\\rho = 0.8$; docs/02 uses $\\rho = 0.7$ after positive-definiteness reductions.", out.width = "70%"}
knitr::include_graphics("../../figures/02-sens-S1.pdf")
```

::: {.bullets data-latex=""}
\textbf{Autocorrelation lifts all three specifications
proportionally; the M2 advantage of approximately 0.06 is
preserved at every $\rho$ level.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item Power increases monotonically with $\rho$ for all
  specifications.
\item M2 advantage: $\approx 0.06$ absolute throughout
  $\rho \in [0.5, 0.9]$.
\item Autocorrelation and carryover act additively on the
  specification ranking; no multiplicative interaction.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
All three specifications gain power monotonically with $\rho$.
At $\rho = 0.5$, M2 power is $0.798 \pm 0.018$ versus M1
$0.738 \pm 0.020$ and M3 $0.738 \pm 0.020$; at $\rho = 0.9$,
M2 reaches $0.956 \pm 0.009$ versus M1 $0.906 \pm 0.013$ and
M3 $0.910 \pm 0.013$ (point estimate $\pm$ MCSE,
$n_{sim} = 500$). Contrary to the prior expectation that M2
would be the most $\rho$-sensitive specification, all three
exhibit comparable relative gains across the explored range.
The M2 advantage over M1 and M3, approximately 0.06 in absolute
power throughout, is preserved at every AR(1) level. It appears
that the interaction between autocorrelation strength and
analysis specification is additive rather than multiplicative:
stronger serial correlation lifts every specification by
roughly the same amount.
:::

### S2: analyst-truth half-life mismatch

```{r fig-sens-s2, echo = FALSE, fig.cap = "Power of M1, M2, and M3 as the analyst's assumed half-life departs from the true DGP half-life. M1 and M3 are invariant to the assumed half-life by construction; M2's power degrades with mis-specification.", out.width = "90%"}
knitr::include_graphics("../../figures/02-sens-S2.pdf")
```

::: {.bullets data-latex=""}
\textbf{M2 retains its advantage over M1 and M3 even when the
analyst's assumed half-life departs from the true value by a
factor of four.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item M2 power ranges from 0.766 to 0.964 across eight S2 cells;
  M1 and M3 average 0.840 each (invariant by construction).
\item M2's curvature is gentle enough that a two-fold
  mis-specification does not erase its advantage.
\item M2 is a defensible default when the pharmacokinetic
  half-life is known only approximately.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
M2 is markedly more robust to analyst-truth half-life mismatch
than the prior expectation predicted. Across the eight S2 cells
(DGP $t_{1/2} \in \{0.5, 1.0\}$ weeks crossed with analyst-assumed
$t_{1/2} \in \{0.25, 0.5, 1.0, 2.0\}$ weeks), M2 averages 0.868,
ranging from 0.766 (DGP 1.0 weeks analyzed at 0.25 weeks) to
0.964 (DGP 0.5 weeks analyzed at 0.25 weeks); per-cell MCSE
ranges from 0.008 to 0.019. M1 and M3 are invariant to the
analyst-assumed half-life by construction, as they do not consume
the parameter, and their power averages 0.840 each across the same
eight cells. The exposure-weighted predictor's curvature is
sufficiently gentle that mis-specification of the assumed
half-life by a factor of two does not erode M2's advantage over
M1 and M3 at either DGP half-life. The result suggests that the
exposure-weighted (M2) specification is a defensible default even when the
true pharmacokinetic half-life is only approximately known at the
design stage.
:::

### S3: dropout

```{r fig-sens-s3, echo = FALSE, fig.cap = "Power at dropout rates $\\in \\{0, 0.1, 0.2, 0.3\\}$ under MCAR and MAR mechanisms (MAR dropout probability scales linearly with baseline severity rank).", out.width = "90%"}
knitr::include_graphics("../../figures/02-sens-S3.pdf")
```

::: {.bullets data-latex=""}
\textbf{M2's advantage narrows under heavy dropout because its
signal source -- the exposure-weighted off-drug contrast -- is
eroded by dropout.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item M2$-$M1 gap: $\approx 0.10$ at dropout 0, declining to
  $\approx 0.03$ at a missing-completely-at-random (MCAR) rate
  of 0.30.
\item Under missing-at-random (MAR) dropout the gap narrows
  further at rate 0.30
  ($\approx 0.004$).
\item M2 advantage is preserved at low-to-moderate dropout
  (rate $\leq 0.20$).
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
All three specifications lose power monotonically with dropout
rate. The MCAR baseline ($N = 70$, dropout = 0) gives M1
$0.764 \pm 0.019$, M2 $0.866 \pm 0.015$, M3 $0.764 \pm 0.019$;
at MCAR rate 0.30 these collapse to M1 $0.122 \pm 0.015$, M2
$0.148 \pm 0.016$, M3 $0.124 \pm 0.015$. M2's advantage narrows
under heavy dropout: the absolute M2$-$M1 gap is approximately
0.10 at zero dropout, 0.07 at rate 0.10, 0.04 at rate 0.20, and
0.03 at rate 0.30 under MCAR. Under MAR with the same marginal
rate the gap narrows further at the highest rate (M2$-$M1
$\approx 0.004$ at rate 0.30 MAR). The pattern is consistent
with the exposure-weighted (M2) specification's signal source residing in
the exposure-weighted contrast contributed by off-drug
observations: when those observations are preferentially
dropped, M2's advantage contracts toward M1's. At
low-to-moderate dropout (rate $\leq 0.20$) the M2 advantage is
preserved.
:::

### S4: biomarker-moderation effect-size curve

```{r fig-sens-s4, echo = FALSE, fig.cap = "Power across the biomarker-moderation effect-size range $c_{bm} \\in \\{0, 0.10, 0.20, 0.30, 0.45, 0.60\\}$ under Architecture B. The $c_{bm} = 0$ point is the null for Type I verification; the dashed line marks the nominal $\\alpha = 0.05$.", out.width = "70%"}
knitr::include_graphics("../../figures/02-sens-S4.pdf")
```

::: {.bullets data-latex=""}
\textbf{M2's advantage is largest at intermediate effect sizes
and shrinks toward ceiling as $c_{bm}$ increases.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item M2$-$M1 advantage peaks at $+0.102$ absolute at
  $c_{bm} = 0.45$, declining to $+0.042$ at $c_{bm} = 0.60$.
\item Type I error at $c_{bm} = 0$ is conservative for all
  specifications (consistent with AR(1) reference distribution).
\item The M2 advantage is most practically consequential when
  targeting moderate biomarker-treatment interactions.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
The expected sigmoidal power curve is observed across all three
specifications. At $c_{bm} = 0$, empirical Type I error is below
the nominal level at $0.026 \pm 0.007$ (M1), $0.028 \pm 0.007$
(M2), and $0.030 \pm 0.008$ (M3); the conservative behavior is
consistent with the AR(1) reference distribution at moderate
sample sizes. Power rises through the mid-range and saturates at
$c_{bm} \geq 0.45$: at $c_{bm} = 0.6$, M1 reaches
$0.916 \pm 0.012$, M2 $0.958 \pm 0.009$, and M3
$0.916 \pm 0.012$. The M2 advantage over M1 is largest at
intermediate effect sizes, $+0.082$ absolute at $c_{bm} = 0.30$
($0.444$ versus $0.362$) and $+0.102$ at $c_{bm} = 0.45$
($0.820$ versus $0.718$), shrinking to $+0.042$ at $c_{bm} = 0.60$
where all specifications approach ceiling. The M2 advantage is
therefore most practically consequential in designs targeting
moderate biomarker-treatment interactions.
:::

### S5: autocorrelation by carryover (exploratory)

```{r fig-sens-s5, echo = FALSE, fig.cap = "Power across carryover half-lives at $\\rho \\in \\{0.5, 0.8\\}$, by analysis specification. This block tests whether the $\\rho$-sensitivity of the M1/M2/M3 ranking depends on carryover strength.", out.width = "90%"}
knitr::include_graphics("../../figures/02-sens-S5.pdf")
```

::: {.bullets data-latex=""}
\textbf{The M2 advantage of $\approx 0.096$ is constant across
the $\rho \times t_{1/2}$ space, confirming additive rather than
interactive effects.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item M2$-$M1 = 0.096 at both $\rho = 0.5$ and $\rho = 0.8$
  when $t_{1/2} = 1.0$ weeks.
\item Autocorrelation and carryover act additively; no
  qualitative interaction with specification ranking.
\item Block S5 functions as methodological reassurance that
  the Tier 1 ranking is preserved across the $\rho$-by-$t_{1/2}$
  grid.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
The M2 $>$ M1 $\approx$ M3 ranking is invariant to the
$\rho \times t_{1/2}$ interaction. At $\rho = 0.5$,
$t_{1/2} = 1.0$ weeks, M2 $-$ M1 = 0.096 (M2 $0.778 \pm 0.019$,
M1 $0.682 \pm 0.021$); at $\rho = 0.8$, $t_{1/2} = 1.0$ weeks,
M2 $-$ M1 = 0.096 (M2 $0.910 \pm 0.013$, M1 $0.814 \pm 0.017$).
The absolute M2 advantage is constant across the
$\rho \times t_{1/2}$ space, indicating that autocorrelation and
carryover strength act additively on the specification ranking
rather than interactively. Block S5 therefore functions as a
methodological reassurance: the Tier 1 ranking is preserved at
every level of $\rho$ and $t_{1/2}$ in the explored grid.
:::

## High-precision rerun confirmation

::: {.bullets data-latex=""}
\textbf{The high-precision rerun confirms the Tier 1 M2 $>$ M1
ranking at MCSE 0.020 with paired McNemar $p = 6.6 \times
10^{-17}$ at the highest-leverage cell.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item 24-cell rerun (600 reps/cell, OL+BDC); 100\% convergence;
  MCSE $\approx 0.020$.
\item At $N = 70$, $t_{1/2} = 1.0$ weeks: M2$-$M1 gap $+13.7$
  percentage points, McNemar $p = 6.6 \times 10^{-17}$.
\item M3 recovery of the M2$-$M1 gap is uneven: 74\% at
  $N = 35$, $t_{1/2} = 1.0$ weeks; 48\%--57\% at larger $N$.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
We endeavored to provide an independent, higher-resolution
confirmation of the Tier 1 ranking. A high-precision rerun at
600 replicates per cell covered a 24-cell expansion (analysis
specification $\in \{$M1, M2, M3$\} \times c_{bm} \in
\{0, 0.45\} \times N \in \{35, 70\} \times
t_{1/2}^{\text{DGP}} \in \{0.5, 1.0\}$ weeks, OL+BDC design),
executed using 8-core `parallel::mclapply`. Wall-clock time was
436 seconds; convergence was 100\% across all 14{,}400 fits.
Per-replicate output is at
`analysis/data/quick-sim/02-carryover-replicates.rds`; the
Morris ADEMP cell-level summary is at `02-carryover-summary.txt`.
Cell MCSE on power is approximately 0.020 in the middle of the
range. The rerun adds independent confirmation of the Tier 1
ranking by exploiting within-replicate correlation among
specifications via paired McNemar tests on the common simulated
datasets.

Type I error across the twelve null cells of the rerun fell in
$[0.003, 0.053]$, so all three specifications hold nominal alpha
at the explored $N$ and DGP half-life combinations, in agreement
with the Tier 1 result in the preceding section.

The M2-versus-M1 gap is statistically clear at three of the four
$(N, t_{1/2})$ alternative cells under $c_{bm} = 0.45$. The
binding cell is at $t_{1/2}^{\text{DGP}} = 1.0$ weeks: at
$N = 35$, $\Delta_{\text{M2-M1}} = +5.7$ percentage points
($0.400 \to 0.457$), McNemar $p = 1.2 \times 10^{-4}$; at
$N = 70$, the gap widens to $+13.7$ percentage points
($0.737 \to 0.873$), McNemar $p = 6.6 \times 10^{-17}$. The
shorter-half-life cells show $\Delta_{\text{M2-M1}} \approx +2$
to $+3$ percentage points with McNemar $p \in [0.004, 0.015]$,
a small but statistically resolvable advantage; it is reduced
because at $t_{1/2}^{\text{DGP}} = 0.5$ weeks the carryover
decays quickly enough that the binary on-drug M1 specification
already captures most of the structure that M2 is designed to
recover.

M3's recovery of the M2-versus-M1 gap is uneven across the
rerun grid. Only the small-$N$ short-half-life cell ($N = 35$,
$t_{1/2}^{\text{DGP}} = 0.5$) reaches and slightly overshoots
the manuscript's headline 70\% recovery target (recovery 119\%;
M3 nominally above M2 by 0.5 percentage points). At $N = 35$,
$t_{1/2}^{\text{DGP}} = 1.0$ weeks, recovery is 74\%; the
larger-$N$ cells fall to 57\% and 48\%. Paired McNemar on M2
versus M3 finds M2 strictly dominant only at the high-leverage
corner ($N = 70$, $t_{1/2}^{\text{DGP}} = 1.0$ weeks,
$\Delta = +7.2$ percentage points, $p = 1.2 \times 10^{-7}$);
in the other three cells the paired difference lies between
$-0.5$ and $+1.5$ percentage points with McNemar $p \geq 0.18$,
and the two specifications are operationally tied at this
resolution. Resolving the M2-versus-M3 gap at the intermediate
cells with statistical clarity would require approximately
2{,}000 replicates per cell at the current MCSE.

The substantive implication, carried into the Discussion, is
that the summary statement 'M2 dominates M1 and M3 captures
roughly 70\% of the gap' is correct in spirit at the
prazosin-calibrated reference cell ($N = 35$,
$t_{1/2}^{\text{DGP}} = 1.0$ weeks), but the M3-as-default
heuristic is supported only at small $N$ and short half-life.
The production-grade recommendation in the Discussion reports
the recovery percentages across the full
$(N, t_{1/2}^{\text{DGP}})$ grid.
:::

# Discussion

## The principal finding

::: {.bullets data-latex=""}
\textbf{M2 leads M1 and M3 by approximately 10 percentage
points under Architecture B in the Hybrid and OL+BDC designs, but
the ranking is conditional: it reverses under the crossover design
and under Architecture A.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item Most favorable cell: M2 = 0.860, M1 = 0.766, M3 = 0.770;
  $\approx 5$ MCSEs of separation at $n_{sim} = 500$.
\item Ranking M2 $>$ M1 $\approx$ M3 holds only under
  Architecture B in the non-crossover designs; M2 is the highest
  in only 19\% of the 360 non-null cells.
\item Under the crossover (CO) design M2 is strongly dominated
  (M2 = 0.488 vs M1 = 0.830 at the matched cell), consistent with
  the companion finding that the crossover AR(1) structure
  neutralises the carryover-correlation channel M2 exploits.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
The central result of this factorial study requires careful
statement, because the advantage of the exposure-weighted
specification M2 is real but conditional. Under Architecture B in
the Hybrid and OL+BDC designs, M2 produced the highest power for
detecting the biomarker-treatment interaction; the advantage
over the binary-treatment specification M1 and the lagged-nuisance
specification M3 was approximately 10 percentage points at the
most favorable configuration (Architecture B, Hybrid design,
$N = 70$, $c_{bm} = 0.45$, $t_{1/2} = 1.0$ weeks; M2 = 0.860,
M1 = 0.766, M3 = 0.770), corresponding to approximately five Monte
Carlo standard errors of separation at $n_{sim} = 500$. This
advantage does not generalize across the grid. M2 was the
highest-power specification in only 68 of the 360 non-null cells
(19 percent); under the crossover (CO) design it was substantially
dominated (for example M2 = 0.488 versus M1 = 0.830 and M3 = 0.834
at the matched CO cell, $t_{1/2} = 1.0$ weeks), and under
Architecture A it was marginally dominated in every design. The
crossover behavior has a clear mechanism, established in the
companion manuscript: the crossover design's within-subject AR(1)
structure dominates the carryover-mediated correlation channel
that Architecture B encodes, leaving nothing for the
exposure-weighted predictor to recover, so M2 collapses to or
below the binary specification. The ranking M2 $>$ M1 $\approx$ M3
therefore holds only under Architecture B in the non-crossover
designs; where it holds, it is preserved across sample sizes and
all carryover half-lives in the Tier 1 grid.
:::

## Robustness to decay-form variation

::: {.bullets data-latex=""}
\textbf{M2 retains its advantage across all five DGP decay forms,
even when the analyst assumes exponential and the true form is
linear or Weibull.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item M2 does not fall below M1 or M3 under any evaluated
  mis-specification condition.
\item The linear DGP produces the largest M2 penalty (sharp
  cutoff poorly approximated by an exponential tail), yet the
  advantage is preserved.
\item M2 is more robust to decay-form mis-specification than
  anticipated.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
A natural concern about any exposure-weighted (M2) specification is that
its advantage over a naive alternative may evaporate when the
assumed form departs from truth. We have examined this directly.
The heatmap in Figure~\@ref(fig:fig-heatmap-mismatch) crosses DGP
decay form against analysis specification at the reference cell
($t_{1/2} = 1.0$ week, Architecture B, Hybrid design, $N = 70$,
$c_{bm} = 0.45$). M2 retains its advantage across all five DGP
decay forms (linear, exponential, Weibull at
$k \in \{0.7, 1.0, 1.5\}$) when the analyst correctly specifies
the form. But what happens when the analyst assumes exponential
while the true DGP is not? The off-diagonal cells show that M2
degrades modestly under mis-specification but does not fall below
M1 or M3 in any evaluated condition. The linear DGP produces the
largest M2 penalty, because the sharp cutoff at $2 t_{1/2}$ is
poorly approximated by an exponential tail, yet even there the
advantage is preserved. On balance, M2 is more robust to
decay-form mis-specification than one might have anticipated.
:::

## Robustness to half-life mis-specification

::: {.bullets data-latex=""}
\textbf{Block S2 contradicts the Jones-Kenward expectation: M3
does not overtake M2 at any evaluated half-life mismatch, because
M2's curvature is too gentle to be costly under mis-specification.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item M2 power span across a four-fold mis-specification:
  $\approx 0.066$; M2 remains above M1 and M3 throughout.
\item Four-fold undershoot (assumed 0.25 wks, true 1.0 wks):
  M2 = 0.766 vs. M1/M3 $\approx 0.74$.
\item M3 closes the gap as half-life information degrades, but
  practical magnitudes are too small for M3 to overtake M2.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
Block S2 provides the strongest practical reassurance in the
study. Jones and Kenward [-@jones2014crossover] developed the
crossover-trial framework underlying the M3 specification with
the expectation that a non-parametric carryover indicator should
outperform a parametric model under mis-specification, because
the lagged indicator does not commit to a functional form. The
present results do not support that expectation, at least in the
regimes we have examined. When the analyst's assumed half-life
departs from the true half-life by a factor of two or four
(assumed $t_{1/2} \in \{0.25, 0.5, 1.0, 2.0\}$ at true
$t_{1/2} = 1.0$), M2 power ranges from 0.766 to 0.832, a span
of roughly 0.066 across the full range and roughly 0.024 across
the factor-of-two window (numerical values are given in
Figure~\@ref(fig:fig-sens-s2)). M1 and M3, by construction not
consuming the assumed half-life, are invariant to this
mis-specification. M2 retains its advantage over M1 and M3 at
every assumed half-life examined: even the four-fold undershoot
to an assumed half-life of 0.25 weeks leaves M2 at 0.766, still
above M1 and M3 at roughly 0.74. It would appear that the
exposure-weighted predictor's curvature is sufficiently gentle
that mis-specification is less consequential than the
crossover-trial literature anticipated. M3 does close the gap
as half-life information degrades, consistent in spirit with
the Jones-Kenward expectation, but the practical magnitudes are
too small for M3 to overtake M2 at any evaluated operating
point. The result suggests that M2 is a defensible default even
when the pharmacokinetic half-life is known only approximately at
the design stage.
:::

## Autocorrelation sensitivity

::: {.bullets data-latex=""}
\textbf{Contrary to expectation, M2's `corCAR1` structure does
not extract disproportionately more information at higher
$\rho$; the advantage is approximately constant across
$\rho \in [0.5, 0.9]$.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item M2 advantage: $\approx 6$ pp at $\rho = 0.5$, $\approx 5$
  pp at $\rho = 0.9$.
\item All three specifications gain power proportionally with
  $\rho$; relative gains are comparable.
\item Autocorrelation acts additively, not interactively, on
  the specification ranking.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
Block S1 asks whether the M2 advantage is concentrated at
particular AR(1) strengths or is approximately constant across
the explored range. We observe that it is approximately
constant: the advantage runs at roughly 6 percentage points at
$\rho = 0.5$ and narrows to roughly 5 points at $\rho = 0.9$, a
range too narrow to be practically meaningful. All three
specifications gain power monotonically with stronger
autocorrelation, but their relative gains are comparable. This
is contrary to the prior expectation that M2's continuous-time
`corCAR1` structure would extract disproportionately more
information at higher $\rho$. It appears, rather, that
autocorrelation and carryover strength act additively on the
specification ranking: stronger serial correlation lifts every
specification by roughly the same amount, leaving the
M2 $>$ M1 $\approx$ M3 ordering undisturbed.
:::

## Dropout and the limits of specification refinement

::: {.bullets data-latex=""}
\textbf{Dropout is far more consequential than the carryover
specification choice; at 30\% dropout all three specifications
are near floor and effectively indistinguishable.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item At 30\% MCAR, all specifications collapse to power
  0.12--0.25; M2 provides no protection against the power loss
  itself.
\item MAR at 10\% is approximately equivalent to MCAR at 12\%,
  because MAR preferentially drops high-severity participants.
\item Dropout prevention is more consequential than
  specification refinement; trialists should prioritise
  accordingly.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
Block S3 is, in some ways, the most sobering result in the
study. The three specifications are meaningfully distinguished
at low-to-moderate dropout: at 10\% MCAR, power drops from
approximately 0.86 to approximately 0.56 across all
specifications, and M2 retains a modest absolute advantage
throughout. At 30\% dropout, however, all specifications are
near floor (0.12 to 0.25) and are effectively
indistinguishable. Unfortunately, M2 provides no protection
against the power loss itself; it merely organizes the remaining
information more efficiently.

Under MAR dropout biased by baseline severity, the pattern is
qualitatively similar but shifted slightly: MAR dropout at 10\%
is comparable to MCAR dropout at approximately 12\%, which is
expected because MAR preferentially drops high-severity
participants who contribute more to the interaction signal under
Architecture B.

The practical implication is direct. Dropout prevention, whether
through shorter trial durations, participant engagement
strategies, or adaptive scheduling, is far more consequential
for the biomarker-interaction test than the choice of
analysis-model carryover specification. Trialists who invest in
refining the carryover specification while tolerating high
dropout rates have, in our view, the priorities somewhat
reversed.
:::

## Architecture dependence

::: {.bullets data-latex=""}
\textbf{Architecture B amplifies the M2 benefit because carryover
erodes the biomarker-response correlation directly, and M2
recovers more of that attenuated signal.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item Architecture B is more sensitive to carryover under all
  three decay forms and all Weibull shapes examined.
\item M2's advantage is present only under Architecture B; under
  Architecture A it is marginally dominated, consistent with the
  correlation-erosion mechanism being absent there.
\item The M2 $>$ M1 $\approx$ M3 ranking holds under
  Architecture B (non-crossover designs) but not under
  Architecture A.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
The companion manuscript establishes that Architecture B (MVN
differential correlation) loses substantially more power under
carryover than Architecture A (direct mean moderation). The
present results confirm that this qualitative gap is robust to
decay-form variation: Architecture B is more sensitive to
carryover under all three decay forms and at all five Weibull
shape values examined. The M2 advantage, however, is confined to
Architecture B: under Architecture A it is marginally negative in
every design (for example, at the Hybrid cell, $t_{1/2} = 1.0$
weeks, M2 = 0.960 versus M1 = 0.972). This is consistent with the
mechanism. Under Architecture B, carryover erodes the
biomarker-response correlation directly, and the continuous
exposure-weighted predictor recovers more of that attenuated
signal than either the binary on-drug indicator or the
lagged-nuisance term can; under Architecture A the treatment
effect remains proportional at every timepoint, so there is no
attenuated correlation for the exposure-weighted predictor to
recover and the simpler binary indicator is marginally preferable.
The M2 $>$ M1 $\approx$ M3 ranking therefore holds under
Architecture B in the non-crossover designs but not under
Architecture A, a dependence that trialists must weigh against the
assumed data-generating mechanism.
:::

## The latent-subtype question

::: {.bullets data-latex=""}
\textbf{The present simulation does not evaluate a finite-mixture
DGP; whether M2 dominance extends to discrete-subtype data is an
open question.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item A mixture DGP would produce discrete responder/non-responder
  profiles rather than a graded covariance structure.
\item Under a mixture DGP, M3's form-free carryover indicator
  might gain an advantage over M2.
\item The present results cannot address this; a mixture DGP
  evaluation is identified as a direction for subsequent work.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
A reviewer of the companion manuscript broached the question of
whether Architecture B's covariance-based interaction is
adequately modeled by the MVN differential-correlation
parameterization, or whether an explicit finite-mixture DGP
would be more faithful to the latent-subtype hypothesis. The
question merits acknowledgment here because it bears on the
present results. A true mixture DGP would produce
biomarker-response associations that are qualitatively different
from those generated by a continuous correlation: discrete
responder and non-responder subgroups rather than a graded
covariance structure. Under a mixture DGP, the lagged-treatment
specification M3 might gain an advantage over M2, because M3
does not assume a specific decay form and may be more tolerant
of the non-smooth response profiles that mixture data can
produce. The present simulation does not evaluate a mixture DGP
and cannot address this question directly. We have not been able
to determine, as yet, whether the M2 dominance we observe extends
to genuinely discrete-subtype data-generating processes; this is
a direction that merits investigation in subsequent work.
:::

## Comparison with prior simulation studies

::: {.bullets data-latex=""}
\textbf{The present factorial replicates Hendrickson's baseline
and extends it by contrasting three analysis specifications, with
the M2-versus-M1 advantage being a new contribution.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item Hendrickson (2020): power 0.82 reproduced here at
  $0.860 \pm 0.016$; M1/M3 comparators are the new element.
\item Jones-Kenward expectation (M3 dominates under
  mis-specification) is not supported at the prazosin-calibrated
  parameters.
\item Sturdevant-Lumley (2021) address carryover testing after
  treatment cessation, a related substantive problem; the two-tier
  reporting structure here follows Morris et al. (2019).
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
Four prior simulation programs provide direct quantitative
benchmarks for the present results. We discuss each in turn,
focusing on cells where the prior parameter ranges overlap with
ours and noting where the design targets differ.

Hendrickson [-@hendrickson2020] introduced the
biomarker-treatment interaction question for aggregated
N-of-1 trials and reported power for the exposure-weighted (M2)
analysis under exponential carryover at $t_{1/2} = 1$ week,
$N = 70$, $c_{bm} = 0.45$, Hybrid design. The reported power
was 0.82 at 1{,}000 replicates under a single architecture.
The present 500-replicate factorial reproduces this result at
the same cell to $0.860 \pm 0.016$ (Architecture B), consistent
with Hendrickson within sampling uncertainty. The principal
methodological extension here is the addition of M1 and M3 as
comparators; Hendrickson did not contrast analysis-side
specifications, so the M2-versus-M1 advantage of approximately
10 percentage points is a new contribution rather than a
replication.

Jones and Kenward [-@jones2014crossover] developed the
crossover-trial framework underlying the M3 specification with
the expectation, noted above, that the non-parametric lagged-on
indicator should outperform a parametric carryover model under
decay-form mis-specification. The present S2 sensitivity block
directly tests this expectation and finds it is not supported in
the regimes we have examined: although M2's power varies by
roughly 0.066 across a four-fold half-life mis-specification,
M2 retains its advantage at every assumed half-life, so M3 does
not become preferred at any evaluated mismatch level. The
Jones-Kenward expectation is preserved in spirit, in that M3
closes the gap as half-life information degrades, but the
practical magnitudes are too small for M3 to overtake M2 in the
explored window. The disagreement is informative: it appears
that the exposure-weighted predictor's curvature is gentle enough
that mis-specification is less costly than the crossover-trial
literature anticipates, at least at the prazosin-calibrated
parameters.

Senn [-@senn2016mastering] established the
variance-decomposition foundation for within-subject designs,
predicting that within-subject contrasts are disproportionately
advantageous when between-patient variance
$\sigma^2_{\text{between}}$ is large relative to within-patient
variance. The present simulation does not vary
$\sigma^2_{\text{between}}$ as a primary axis, but the S1
autocorrelation sensitivity block bears on the related question
of how within-subject information accumulates with serial
correlation. Our finding that M2's advantage is approximately
constant across $\rho \in [0.5, 0.9]$ is qualitatively
consistent with Senn's prediction that within-subject methods
extract roughly proportional information across correlation
regimes; the absence of a strong $\rho$ interaction with the
analysis specification is the new result here.

Sturdevant and Lumley [-@sturdevant2021carryover] address the
related but distinct problem of testing for carryover effects
after treatment cessation using a mixed-effects model, in the
context of threshold-crossing diagnosis under open-label
treatment. Their work bears on the substantive carryover-testing
question rather than on study design, but it reinforces the point
that carryover is a first-order inferential concern in
within-subject designs and not merely a nuisance. The two-tier
structure of the present study, a principal factorial covering the
main contrast supplemented by marginal sensitivity blocks anchored
at a reference cell, follows the general simulation-reporting
guidance of Morris et al. [-@morris2019simulation, Section 7]
rather than any single prior study, and our adoption of it is
documented in the `simulation-grid-plan.md` design document. The
two-tier structure proved useful in practice: the Tier 1 factorial
establishes the conditional M2 ranking, while the Tier 2 blocks
probe its robustness to single-axis perturbations.

The companion manuscripts at
`analysis/report/01-dgp-mean-moderation-vs-mvn/` and
`analysis/report/05-nof1-design-sensitivity/` provide
complementary benchmarks: the first establishes the
architecture-dependence of carryover-induced power loss
confirmed here in the architecture-dependence subsection, and
the second reports the effect of carryover on the Hybrid N-of-1
design's saturation profile confirmed in §4.4.
:::

## Limitations

::: {.bullets data-latex=""}
\textbf{Five limitations bound the generalizability of the
findings: single prazosin parameterization, incomplete analysis
strategy space, OFAT sensitivity blocks, omitted OL design, and
monotone dropout assumption.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item Absolute power values are specific to the prazosin-PTSD
  parameterization; rankings are expected to be more portable
  but are not verified across alternative trajectory shapes.
\item Three specifications do not exhaust the carryover-aware
  model space; weighted analysis, random slopes, and observation
  exclusion are not evaluated.
\item Sensitivity blocks are OFAT; joint perturbations (e.g.,
  high $\rho$ combined with high dropout) are not fully explored.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
Several limitations should be noted when interpreting these
results.

First, the simulation draws its response-trajectory parameters
from a single configuration calibrated to the Hendrickson (2020)
prazosin-PTSD framework (Gompertz maxima, rates, and dispersions).
The absolute power values are specific to this parameterization.
The ranking of analysis specifications is expected to be more
portable, but we have not been able to verify this across
alternative response profiles. Trialists working in therapeutic
areas with substantially different trajectory shapes should
regard the power estimates as illustrative rather than
definitive.

Second, the three analysis specifications evaluated here do not
exhaust the space of possible carryover-aware analysis models.
Weighted analysis, within-subject contrasts, two-stage random
slopes, and exclusion of contaminated observations (discussed in
the companion manuscript's Section 6) are not evaluated. We have
endeavored to select the three specifications that represent the
most natural practical alternatives for the N-of-1 context, but
a more comprehensive comparison would cross the present grid with
these additional strategies, and others may merit consideration.

Third, the sensitivity blocks are one-factor-at-a-time
perturbations around a single reference configuration.
Unfortunately, joint perturbations, for example high $\rho$
combined with high dropout, may produce interactions that the
marginal blocks do not capture. The exploratory block S5
($\rho \times$ carryover) provides one such joint check and
does not reveal a qualitative interaction, which is reassuring,
but the space of possible joint combinations is large and our
exploration of it is necessarily incomplete.

Fourth, the Tier 1 grid omits the OL design because
OL lacks the off-drug contrast required by the M1 and M3
specifications. For trialists considering an OL design, the
analysis-model question is moot: only M2 (or the `bm:t`
trajectory-based test from the companion manuscript) is
applicable.

Fifth, dropout is modeled as monotone: once a participant
misses a timepoint, all subsequent timepoints are missing. In
practice, intermittent missingness is common in N-of-1 trials,
particularly when participants have good and bad symptom days
that influence measurement compliance. Intermittent missingness
would presumably be less destructive than monotone dropout
because it preserves some late-trial observations that
contribute to the treatment contrast. We note this as a
realistic departure from the modeled mechanism that future
work should address.

The headline Tier 1 ranking is further supported by the
high-precision rerun reported at the end of the Results section,
which resolves the M2-versus-M1 gap to MCSE 0.020 with paired
McNemar $p < 10^{-3}$ at the headline cells.
:::

# Conclusions

1. **M2 (exposure-weighted) is the recommended default only
   under Architecture B in the multi-period designs.** The
   continuous exposure-weighted predictor `Dbc` with an assumed
   exponential decay produces the highest power for detecting the
   biomarker-treatment interaction under carryover, but this
   advantage is confined to Architecture B (MVN differential
   correlation) combined with the Hybrid or OL+BDC design. In
   that region M2 attains the strictly highest power in 82.5% and
   87.5% of carryover-affected, non-null cells respectively, with
   an advantage over the binary-treatment (M1) and lagged-nuisance
   (M3) alternatives of approximately 10 percentage points at the
   reference configuration; the lead there is robust to DGP
   decay-form variation, within-factor autocorrelation, and
   analyst-truth half-life mismatch. Outside this region M2 holds
   no advantage: under the crossover design (Architecture B), and
   under Architecture A in every design, M2 never attains the
   strictly highest power and M1 is at least as good (see
   Conclusion 5). Across the full 540-cell factorial M2 is
   therefore the strict winner in only 17.8% of cells (28.3%
   restricting to carryover-affected, non-null cells). Where M2 is
   used, report it with a CR2 cluster-robust standard error, which
   restores nominal size; the ranking is invariant to this
   recalibration (Block S6).

2. **Half-life mis-specification does not displace M2.** M2
   power varies by roughly 0.066 across a four-fold
   mis-specification of the analyst's assumed half-life (Block
   S2), but M2 retains its advantage over M1 and M3 at every
   assumed half-life examined. Trialists who have only an
   approximate pharmacokinetic estimate can adopt M2 with
   confidence that a mis-specified half-life does not cost M2
   its ranking.

3. **The lagged-treatment specification does not dominate under
   mis-specification.** Contrary to the expectation from the
   crossover-trial literature, M3 does not become preferable when
   the analyst's decay-form assumption is wrong. M3's
   model-free character provides robustness to assumed half-life
   but not enough efficiency gain to overtake M2 at any evaluated
   operating point.

4. **Dropout is more consequential than the carryover
   specification.** Power degrades sharply with dropout rate
   under all specifications. At 20% dropout, all three
   specifications lose approximately half their power relative to
   the no-dropout condition. Trialists should prioritise dropout
   prevention over carryover-specification refinement.

5. **The DGP architecture determines whether M2 has any
   advantage at all.** Under Architecture B (MVN differential
   correlation), carryover-induced correlation erosion makes the
   continuous predictor's information advantage consequential and
   M2 leads. Under Architecture A (direct mean moderation), the
   treatment effect stays proportional at every timepoint, there
   is no correlation erosion to recover, and M2 is marginally
   dominated by the binary specification in every design. The
   recommendation for M2 is therefore conditional on the assumed
   architecture, not merely modulated in magnitude by it.

6. **Reporting recommendations.** Simulation studies evaluating
   biomarker-moderated treatment effects under carryover should
   report: (a) the analysis-model carryover specification used,
   (b) the assumed decay form and half-life, (c) sensitivity of
   power estimates to at least one alternative specification,
   and (d) the anticipated dropout rate, which is the dominant
   source of power loss in the evaluated grid.

7. **Type I error is conservative, and recalibrable.** All three
   specifications hold below the nominal 0.05 across the full
   Tier 1 grid and the S4 effect-size curve, with no inflation
   under any DGP decay form, architecture, or carryover level.
   The shortfall is the working-covariance conservatism the
   companion calibration study documents; a CR2 cluster-robust
   standard error removes it and recovers a few points of power
   without disturbing the ranking (Block S6).

# Conflict of interest

The authors declare no conflicts of interest.

# Funding

This work received no specific funding.

# Data availability statement

::: {.bullets data-latex=""}
\textbf{All simulation outputs and drivers are versioned in the
pmsimstats-ng repository; exact paths are given in the
Reproducibility section.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item Per-replicate Monte Carlo outputs and cell-level summaries:
  \texttt{analysis/data/}.
\item Simulation drivers:
  \texttt{analysis/scripts/carryover-sensitivity/}.
\item Exact file paths listed in the Reproducibility section
  below.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
The data and code supporting the findings of this study are
available in the pmsimstats-ng project repository.
Per-replicate Monte Carlo outputs and ADEMP-compliant cell-level
summaries are versioned under `analysis/data/`; simulation
drivers are at `analysis/scripts/`. Exact paths for this
manuscript are listed in the Reproducibility section below.
:::

# Reproducibility

::: {.bullets data-latex=""}
\textbf{The manuscript is fully reproducible from pinned package
versions in \texttt{renv.lock}, deterministic seeds, and
pre-computed checkpoints; no simulation code re-executes during
knit.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item Render environment: R 4.6.0, zzcollab Docker container;
  the production simulation ran under R 4.5.3 (see §3.6); packages
  pinned in \texttt{renv.lock}.
\item Principal outputs: \texttt{01-factorial.rds} and
  \texttt{04-sensitivity.rds}; cell-level summaries at
  \texttt{analysis/data/02-grid-summary.rds}.
\item Manuscript Rmd loads pre-computed checkpoints only; no
  simulation code runs during render.
\item The three specifications are recorded as \texttt{A1},
  \texttt{A2}, \texttt{A3} in the archived simulation output and are
  reported here as M1, M2, M3; the mapping is applied at display
  time and the stored values are unchanged.
\end{itemize}
:::

::: {.rgt data-latex=""}
rgt to complete.
:::

::: {.orig data-latex=""}
This manuscript was rendered with R 4.6.0 inside the project's
zzcollab Docker container (the production simulation itself ran
under R 4.5.3, as recorded in §3.6; image hash recorded in
`Dockerfile`;
package set captured in `renv.lock`). Principal packages used by
the simulation pipeline are `nlme` (continuous-time linear-mixed
analysis with `corCAR1` residual correlation), `parallel`
(8-core `mclapply` for parameter-grid sweeps), `data.table` (the
`original-extended` simulation collection), `furrr` and `purrr`
(the `tidyverse` simulation collection), and `rmarkdown` plus
`bookdown` for the manuscript build.

**Random-number generator.** Each simulation driver sets
`set.seed(20260507)` at the top of the replicate loop and uses
`parallel::mc.set.seed` inside `mclapply`. Per-replicate seeds
are derived from the master seed as \texttt{master + rep\_idx}.

**Data and code availability.** Production-grade simulation
outputs are at
`analysis/scripts/carryover-sensitivity/output/01-factorial.rds`
(principal factorial) and `04-sensitivity.rds` (sensitivity
blocks S1 through S5). Cell-level summaries are at
`analysis/data/02-grid-summary.rds` and
`analysis/data/02-sensitivity-summary.rds`. Driver scripts are
`01-run-factorial.R`, `04-run-sensitivity-blocks.R`, and
`02-summarize-grid.R` under
`analysis/scripts/carryover-sensitivity/`. The simulation grid
plan and design rationale are documented at
`analysis/scripts/carryover-sensitivity/simulation-grid-plan.md`,
which serves as the Morris ADEMP pre-registration. High-precision
prototype rerun outputs are at
`analysis/data/quick-sim/02-carryover-replicates.rds`
(per-replicate) and `02-carryover-summary.txt` (Morris ADEMP
cell-level summary). The manuscript source, paper-specific
bibliography (`references.bib`), and SIM CSL file are at
`analysis/report/02-carryover-sensitivity/`.
:::

# References

::: {#refs}
:::

```{r footer, echo=FALSE, results='asis'}
src <- sub("^/Users/zenn/(Library/CloudStorage/)?Dropbox/prj/", "~/prj/",
           knitr::current_input(dir = TRUE))
cat(sprintf(
  "\n\\vfill\n\n---\n\n*Rendered on %s.*  \n*Source: `%s`*\n",
  format(Sys.time(), "%Y-%m-%d at %H:%M %Z"),
  src
))
```
