---
title: "Data generation machinery for N-of-1 clinical trial simulations
  under a two-component response decomposition: dropping the
  natural-history component"
author: 'pmsimstats team'
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: true
    toc_depth: 2
    number_sections: true
    keep_tex: true
    highlight: tango
    includes:
      in_header: ../sim-preamble.tex
fontsize: 11pt
linestretch: 1.1
geometry: 'left=3cm,right=3cm,top=2.5cm,bottom=2.5cm'
bibliography: references.bib
csl: statistics-in-medicine.csl
link-citations: true
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{booktabs}
  - \usepackage{array}
  - \usepackage{longtable}
  - \usepackage{float}
  - \usepackage[mathlines]{lineno}
  - \linenumbers
knit: (function(input, ...) { d <- dirname(input); while (!file.exists(file.path(d, 'tools', 'stamp-render.R')) && d != dirname(d)) d <- dirname(d); source(file.path(d, 'tools', 'stamp-render.R'))$value(input) })
---

<!--
  STATUS (2026-09-09)
  ===================
  Paper 14 is the companion to paper 13. The two papers evaluate the
  two candidate two-component reductions of the three-component DGP:

    paper 13  BR + TV   (placebo-belief dropped)
    paper 14  BR + PB   (natural-history dropped)   <- this paper

  They are deliberately parallel in structure so the pair can be read
  as a single comparison. Section 1.1 and Section 4 differ; the
  methods, appendices, and grid are common apart from which component
  is present.

  Simulation driver (shared with paper 13):
    Rscript analysis/scripts/two-component-dgp/01-run-grid.R \
      --drop tv --reps 500
  Writes: analysis/data/14-drop-tv-grid.rds

  Knit:
    bash tools/render.sh \
      analysis/report/14-two-component-drop-tv/report.Rmd
-->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)
suppressPackageStartupMessages({
  library(dplyr)
  library(knitr)
})

## The driver output may not exist yet while the manuscript is in
## draft. Guard every downstream chunk on HAVE_SIM so the document
## knits either way.
sim_path <- here::here('analysis/data/14-drop-tv-grid.rds')
HAVE_SIM <- file.exists(sim_path)
if (HAVE_SIM) {
  sim <- readRDS(sim_path)
  S <- sim$summary
  NREPS <- sim$meta$n_reps
} else {
  sim <- NULL; S <- NULL; NREPS <- NA_integer_
}

arch_label <- function(a) ifelse(a == 'mvn', 'Covariance (B)', 'Mean (A)')

cellv <- function(arch, design, N, c_bm, t1half, what) {
  if (!HAVE_SIM) return(NA_real_)
  r <- S[S$arch == arch & S$design == design & S$N == N &
         abs(S$c_bm - c_bm) < 1e-8 & abs(S$t1half - t1half) < 1e-8, ]
  if (nrow(r) != 1) return(NA_real_)
  r[[what]]
}
```

\section{Abstract}

\textbf{Background.} Simulation studies of aggregated N-of-1 trials
must choose where in the joint distribution a biomarker-treatment
interaction lives. A companion paper compares two such choices, mean
moderation and covariance moderation, under a three-component
decomposition of treatment response into pharmacological (BR),
placebo-belief (PB), and time-variant natural-history (TV)
components. Whether all three components are required for a
comparison whose object is the architectures themselves is not
obvious.

\textbf{Methods.} We repeat the comparison under a reduced
data-generating process carrying only BR and PB, dropping the
time-variant natural-history component. This is the reduction whose
safety is unconditional: the implementation provides no biomarker-TV
parameter, so the deleted block cannot carry any part of the
biomarker coupling under any parameter setting. It is also the
reduction that preserves every feature by which the trial designs
differ, since the expectancy weight continues to act through PB. The
reduction takes the correlation matrix from $(2 + 3n)$ to $(2 + 2n)$,
from $26 \times 26$ to $18 \times 18$ at the eight occasions used
throughout this program. A companion paper evaluates the alternative
reduction, in which PB is dropped instead.

\textbf{Results.} The architecture contrast reported by paper 01 under
three components is reproduced under two. At $c_{bm} = 0.30$ and
$N = 70$, raising the carryover half-life from zero to one week costs
covariance moderation $0.194$ of power under the Hybrid design and
$0.196$ under OL+BDC, against $0.040$ and $0.078$ for mean
moderation. Under the crossover design neither architecture loses
power to carryover, because its off-drug occasions are spaced widely
enough that residual exposure has already decayed at every one of
them. At paper 01's reference cell ($c_{bm} = 0.45$, $N = 70$) all
cells lie above $0.89$ and the contrast is compressed by a ceiling
effect to $0.032$ against $0.002$; the reference cell understates the
architecture difference roughly sixfold and should not be used to
read it. Empirical Type I error is strongly design-dependent and
architecture-independent, at $0.061$ under CO, $0.039$ under Hybrid
and $0.021$ under OL+BDC against a nominal $0.05$.

\textbf{Conclusions.} The TV component is inert for the
biomarker-treatment interaction estimand under the data-generating
process as implemented, and acts only as a variance term. A
two-component decomposition is therefore adequate for methodological
work whose object is a comparison among architectures, designs, or
analysis specifications. The full three-component decomposition
remains necessary when the estimand is the attribution of response to
its causal sources, when a candidate biomarker may correlate with a
non-pharmacological component, or when absolute power is reported as
a design input rather than as a comparison.

\section{1. Introduction}

An aggregated N-of-1 trial pools repeated within-participant on-drug
and off-drug contrasts to estimate a population-level
biomarker-treatment interaction [@hendrickson2020]. Simulation
studies of such trials must specify a data-generating process, and
that specification involves two choices usually made implicitly. The
first is where the interaction lives in the joint distribution of
biomarker and response. The second is how many distinct latent
components the response trajectory is decomposed into.

The companion paper 01 takes up the first choice. It compares
\emph{mean moderation}, in which the biomarker shifts the mean of the
pharmacological response, against \emph{covariance moderation}, in
which the biomarker's correlation with that response is
treatment-state dependent. It shows the two are not interchangeable:
they respond differently to carryover, and the difference is large
enough to change a design recommendation.

The present paper takes up the second choice. The program's standard
DGP decomposes response into three components: BR, the
pharmacological action of the compound; PB, the improvement driven by
the participant's belief that treatment will work; and TV, drift in
the underlying condition independent of treatment. That decomposition
is developed and defended in paper 06, whose subject is the
inferential cost of failing to decompose.

We ask what happens when the decomposition is reduced to two
components, BR and TV, dropping the modelled placebo-belief term. The
reduction is motivated in the first instance by tractability: it
takes the correlation matrix from $(2 + 3n)$ to $(2 + 2n)$, from $26
\times 26$ to $18 \times 18$ at the eight occasions used throughout
this program, and removes two of the three cross-component blocks.
But the choice of \emph{which} component to drop is not arbitrary,
and Section 1.1 sets out why dropping PB rather than TV is the
reduction that aligns with established practice.

\subsection{1.1 Why dropping natural history is the unconditionally
safe reduction}

Two reductions of the three-component decomposition are available,
and they are not equally safe. This paper takes the safer one; the
companion paper 13 takes the other, for reasons of convention that
Section 4.5 weighs against the argument given here.

The decisive consideration is structural. The biomarker cannot couple
to TV at all. The implementation provides an optional biomarker-PB
correlation parameter, used in paper 06's contaminated-biomarker
study, but there is no biomarker-TV counterpart anywhere in the
package, the drivers, or any implementation collection. The
biomarker-TV entry of the correlation matrix is therefore identically
zero under every parameter setting this program can express, not
merely under the settings it happens to use.

That distinction has a direct consequence for what each reduction
licenses. Paper 06's omitted-variable identity writes the
one-component biomarker slope as $\beta_{bm}^{BR} + w_{PB}
\beta_{bm}^{PB} + w_{TV}\beta_{bm}^{TV}$, so a component contributes
bias only through the biomarker's covariance with it. Deleting a
block whose coupling is structurally zero cannot change the estimand.
Deleting a block whose coupling is merely set to zero changes the
estimand as soon as someone sets it otherwise. The present reduction
is of the first kind and the companion's is of the second, which is
why the companion carries a conditionality that this paper does not.

A second consideration concerns what the reduction preserves rather
than what it removes. The design's expectancy weight $\eta$, which
distinguishes open-label from blinded occasions, enters the
data-generating process only through the PB component: it scales the
PB mean trajectory and the PB standard deviation. Retaining PB
therefore retains the blinding contrast, and the three trial designs
continue to differ in every respect the full decomposition
represents: on-drug pattern, calendar spacing, and expectancy
schedule. Dropping TV costs nothing on this axis, because the
calendar-time argument that drives the TV mean also drives the AR(1)
week gaps of every within-component block, and those survive the
reduction untouched.

Against these two points stands a consideration of convention, which
is the companion paper's argument and which we do not dismiss. A time
trend is a standard element of crossover and N-of-1 models, while a
modelled placebo-belief trajectory is an innovation of this program
rather than an inherited convention. Dropping TV therefore removes
the more conventional of the two components and retains the less
conventional one, which is the opposite of what a reader of that
literature would expect. Section 4.5 sets the two arguments against
each other rather than pretending the choice is clear-cut.

We hold everything else fixed. The trial designs, the analysis model,
the parameter calibration, the biomarker-moderation levels, and the
carryover half-lives are those of paper 01. Section 2 states the
model and the two architectures under the reduced decomposition,
Section 3 gives the results, and Section 4 takes up what follows for
the design of future simulation work in this program. Three
appendices carry the structural material: Appendix A on why the two
architectures are treated as a pair, Appendix B on the correlation
matrix the reduced DGP assembles, and Appendix C on the
positive-definiteness constraint the reduction was expected to
relax.

\section{2. Methods}

\subsection{2.1 Notation and analysis model}

Let $i = 1, \dots, N$ index participants and $p = 1, \dots, n$
measurement occasions at cumulative weeks $w_1 < \cdots < w_n$. Write
$B_i$ for the participant's baseline biomarker, $\mathrm{BL}_i$ for
baseline symptom severity, and $S_{x,ip}$ for the observed symptom
score. Under the two-component decomposition,

$$
S_{x,ip} = \mathrm{BL}_i - \big( BR_{ip} + PB_{ip} \big),
$$

so that a beneficial response lowers the score. The three-component
model of paper 06 adds $TV_{ip}$ inside the parentheses; removing it
recovers the present model exactly.

Both components are drawn jointly across occasions from a
multivariate normal whose means follow modified Gompertz
trajectories,

$$
\mathrm{mg}(t; a, d, r) = a \exp\!\big(-d \exp(-r t)\big),
$$

evaluated at a component-specific time argument: cumulative time on
drug $t_{od}$ for BR, and accumulated belief exposure $t_{pb}$ scaled
by the design's expectancy weight $\eta$ for PB. The distinct time
arguments are what make the two components separable: BR responds to
treatment status, PB to belief, and the blinded phases of the design
move the two apart.

The analysis model is held fixed across all conditions,

$$
S_{x,ip} = \beta_0 + \beta_{bm} B_i + \beta_t t_p
  + \beta_D D_{bc,ip} + \beta_{bm:D} B_i D_{bc,ip}
  + u_i + \varepsilon_{ip},
$$

fitted with \texttt{nlme::lme}, a participant random intercept $u_i
\sim N(0, \sigma_u^2)$, and continuous-time AR(1) residual
correlation \texttt{corCAR1(form = \textasciitilde\ t | ptID)}. Here
$D_{bc,ip}$ is the continuous exposure indicator, equal to one on
drug and decaying as $\exp(-\lambda t_{sd,ip})$ off drug with
$\lambda = \ln 2 / t_{1/2}$. The estimand is $\beta_{bm:D}$.
Appendix A.4 states why the analysis model is held fixed.

\subsection{2.2 Two data-generating architectures}

\subsubsection{2.2.1 Mean moderation (Architecture A)}

Under mean moderation the biomarker shifts the conditional mean of
the pharmacological component at on-drug occasions. After the
multivariate normal draw, BR is displaced by

$$
BR_{ip} \leftarrow BR_{ip} + c_{bm}\, z_i\, \sigma_{BR}
  \quad \text{for occasion } p \text{ on drug},
$$

where $z_i = (B_i - \mu_{bm})/\sigma_{bm}$ is the standardized
biomarker. The covariance structure is untouched: the biomarker is
independent of every component draw, and the interaction is carried
entirely by the first moment. In the notation of Appendix B, mean
moderation sets the coupling vector $b \equiv 0$.

\subsubsection{2.2.2 Covariance moderation (Architecture B)}

Under covariance moderation the means are untouched and the
interaction is carried by a treatment-state-dependent correlation,

$$
\mathrm{Cor}(B_i, BR_{ip}) = c_{bm} \cdot \phi(t_{sd,ip}), \qquad
\phi(t_{sd}) = \begin{cases}
1, & \text{on drug} \\
\exp(-\lambda t_{sd}), & \text{off drug.}
\end{cases}
$$

The biomarker is more strongly correlated with the pharmacological
response while the drug is present, and that association decays after
discontinuation at the pharmacokinetic rate.

\subsubsection{2.2.3 What the reduced decomposition changes}

Neither architecture references TV. Removing it therefore leaves both
definitions above unchanged and alters the simulation in three
mechanical ways. Unlike the companion reduction, it has no
substantive fourth consequence.

First, the correlation matrix shrinks, from $2 + 3n$ to $2 + 2n$.
Appendix B gives the resulting block partition.

Second, the outcome loses a variance term. At the reference
calibration the three components carry standard deviations of 10
(TV), 10 (PB), and 8 (BR), so TV accounts for approximately 38\% of
the component variance. Removing it reduces the residual variation
against which the interaction is tested, and absolute power rises
accordingly. This is an expected arithmetic consequence rather than a
finding, and it is why absolute power in this paper is not comparable
to paper 01's. Because the TV standard deviation is not scaled by the
expectancy weight, the share removed is the same under every design,
which is not true of the companion reduction.

Third, the cross-component structure simplifies. The three-component
matrix carries three cross-component blocks; the two-component matrix
carries one. Appendix B.6 sets out exactly what is removed, and
Appendix C shows that what is removed is not what constrains the
moderation parameter.

What does \emph{not} change is worth stating, because it is the
principal practical difference between this reduction and the
alternative. The expectancy weight $\eta$ continues to act, through
the PB mean and the PB standard deviation, so the open-label versus
blinded contrast remains represented in the generated data and the
three designs continue to differ in expectancy schedule as well as in
on-drug pattern. The calendar-time argument that drove the TV mean
also drives the AR(1) week gaps in every within-component block, and
those are untouched, so unequal visit spacing continues to be
represented as well. No feature of the trial design becomes inert.

\subsection{2.3 The carryover recursion}

Both architectures apply carryover to the BR mean trajectory: at an
off-drug occasion, a decayed fraction of the pre-discontinuation
response persists. The implementation used here differs from the one
behind paper 01's published results, and the difference is worth
stating.

The earlier implementation computed

$$
\mu_{BR,p} \leftarrow \mu_{BR,p} + \mu_{BR,p-1}\,
  (1/2)^{t_{sd,p}/t_{1/2}},
$$

in which $\mu_{BR,p-1}$ is the \emph{already adjusted} value from the
preceding occasion while $t_{sd,p}$ is \emph{cumulative} time since
discontinuation. Elapsed time is therefore counted twice, and every
off-drug occasion after the first in an uninterrupted run is
over-decayed. At a one-week half-life the second off-drug occasion
receives half the residual effect it should.

The present implementation anchors the decay to the mean at
discontinuation,

$$
\mu_{BR,p} \leftarrow \mu_{BR,p} + \mu_{BR,p^{\ast}}\,
  (1/2)^{t_{sd,p}/t_{1/2}},
$$

where $p^{\ast}$ indexes the most recent on-drug occasion, so the
cumulative decay factor is applied once. We verified at three
reference cells (Hybrid, OL+BDC, and CO at $N = 70$, $c_{bm} = 0.45$,
$t_{1/2} = 1.0$, 400 replicates under a common seed) that the
correction moves power by at most 0.005 against a Monte Carlo
standard error of about 0.025. It is adopted because it is correct,
not because it changes results.

\subsection{2.4 Simulation design}

The grid crosses architecture, trial design, sample size,
biomarker-moderation strength, and carryover half-life.

| Factor | Levels |
|---|---|
| Architecture | mean moderation (A), covariance moderation (B) |
| Trial design | CO, Hybrid, OL+BDC |
| Sample size $N$ | 35, 70 |
| Biomarker moderation $c_{bm}$ | 0, 0.30, 0.45 |
| Carryover half-life $t_{1/2}$ | 0, 0.5, 1.0 weeks |
| Components | BR + PB |
| Replicates | 500 |

This gives $2 \times 3 \times 2 \times 3 \times 3 = 108$ cells. The
$c_{bm} = 0$ rows are the Type I error check against nominal 5\%; the
$t_{1/2} = 0$ rows give a no-carryover reference against which the
cost of nonzero carryover is read off directly.

Throughout, $N$ is the \textbf{total} number of participants,
allocated as evenly as possible across the design's randomization
paths, so a design with $P$ paths assigns about $N/P$ to each.

Absolute power is not comparable between architectures at the same
nominal $c_{bm}$, since the parameter is not calibrated to equal
power across the two channels. The interpretable comparison is the
carryover profile within each architecture, and the qualitative shape
of that profile between them.

\section{3. Results}

All 108 cells converged completely: the non-convergence rate is zero
throughout. At `r NREPS` replicates the Monte Carlo standard error on
power does not exceed `r sprintf('%.3f', max(S$mc_se_power[S$c_bm > 0]))`,
and differences smaller than about twice that should not be
over-interpreted.

\subsection{3.1 Power under carryover}

```{r tab-main, echo = FALSE}
mk_tab <- function(cb, nn) {
  S |>
    filter(N == nn, abs(c_bm - cb) < 1e-8) |>
    mutate(Architecture = arch_label(arch), Design = design) |>
    select(Architecture, Design, t1half, power) |>
    tidyr::pivot_wider(names_from = t1half, values_from = power,
                       names_prefix = 'thalf') |>
    mutate(Cost = thalf0 - thalf1) |>
    arrange(Architecture, Design) |>
    transmute(Architecture, Design,
              `t=0` = sprintf('%.3f', thalf0),
              `t=0.5` = sprintf('%.3f', thalf0.5),
              `t=1.0` = sprintf('%.3f', thalf1),
              `Cost` = sprintf('%+.3f', Cost))
}
knitr::kable(mk_tab(0.30, 70), booktabs = TRUE,
  col.names = c('Architecture','Design','$t_{1/2}=0$','$0.5$','$1.0$',
                'Cost'),
  escape = FALSE,
  caption = sprintf('Power by architecture, trial design and carryover
    half-life at $c_{bm} = 0.30$, $N = 70$, %d replicates per cell.
    Cost is the power lost in moving from no carryover to a one-week
    half-life. Absolute power is not comparable between architectures
    at a common nominal $c_{bm}$; the interpretable quantity is the
    profile across half-life within each architecture.',
    NREPS)) |>
  kableExtra::kable_styling(latex_options = 'scale_down')
```

Table \@ref(tab:tab-main) carries the principal result. Under the
Hybrid design, moving from no carryover to a one-week half-life costs
covariance moderation
`r sprintf('%.3f', cellv('mvn','Hybrid',70,0.30,0,'power') - cellv('mvn','Hybrid',70,0.30,1,'power'))`
of power, against
`r sprintf('%.3f', cellv('mean_moderation','Hybrid',70,0.30,0,'power') - cellv('mean_moderation','Hybrid',70,0.30,1,'power'))`
for mean moderation. Under OL+BDC the corresponding figures are
`r sprintf('%.3f', cellv('mvn','OL+BDC',70,0.30,0,'power') - cellv('mvn','OL+BDC',70,0.30,1,'power'))`
and
`r sprintf('%.3f', cellv('mean_moderation','OL+BDC',70,0.30,0,'power') - cellv('mean_moderation','OL+BDC',70,0.30,1,'power'))`.
Both covariance-arm costs are several times the Monte Carlo standard
error; both mean-arm costs are within two or three of it. This is
paper 01's central finding, reproduced on a correlation matrix
two-thirds the size.

The crossover design behaves differently, and consistently for both
architectures: neither arm loses power to carryover under CO, and the
point estimates move by less than one standard error in either
direction. The reason is structural rather than statistical. CO
spaces its occasions 2.5 weeks apart, so at a one-week half-life the
residual exposure multiplier $\phi(t_{sd})$ has already fallen below
$0.18$ at the first off-drug occasion and below $0.03$ at the second.
There is almost no residual exposure left for carryover to
contaminate. The same feature explains why CO's positive-definiteness
ceiling is insensitive to $t_{1/2}$ (Appendix C.4), so the power
result and the structural result agree.

\subsection{3.2 The reference cell is ceiling-bound}

```{r tab-ref, echo = FALSE}
knitr::kable(mk_tab(0.45, 70), booktabs = TRUE,
  col.names = c('Architecture','Design','$t_{1/2}=0$','$0.5$','$1.0$',
                'Cost'),
  escape = FALSE,
  caption = 'The same comparison at paper 01\'s reference cell
    ($c_{bm} = 0.45$, $N = 70$). Every entry lies above $0.89$, so the
    architecture contrast is compressed by a ceiling effect and the
    carryover cost is understated.') |>
  kableExtra::kable_styling(latex_options = 'scale_down')
```

Table \@ref(tab:tab-ref) repeats the comparison at paper 01's
reference cell, and is included as a caution rather than as a result.
Every cell lies above $0.89$. The Hybrid carryover cost under
covariance moderation falls to
`r sprintf('%.3f', cellv('mvn','Hybrid',70,0.45,0,'power') - cellv('mvn','Hybrid',70,0.45,1,'power'))`,
roughly a sixth of its value at $c_{bm} = 0.30$, not because the
mechanism is weaker but because there is no room above the starting
value for the loss to register.

We record this explicitly because the reference cell is the natural
place to look for comparability with paper 01, and doing so here
would understate the architecture difference by a factor of about
six. The informative cells are those where baseline power is well
away from one. The same caution applies in reverse at $N = 35$, where
the covariance arm's OL+BDC cost reaches
`r sprintf('%.3f', cellv('mvn','OL+BDC',35,0.45,0,'power') - cellv('mvn','OL+BDC',35,0.45,1,'power'))`.

\subsection{3.3 Type I error}

```{r tab-type1, echo = FALSE}
t1 <- S |>
  filter(c_bm == 0) |>
  group_by(Design = design) |>
  summarise(cells = dplyr::n(),
            rate = sum(power * NREPS) / (dplyr::n() * NREPS),
            .groups = 'drop') |>
  mutate(se = sqrt(rate * (1 - rate) / (cells * NREPS)),
         `Rejection rate` = sprintf('%.3f', rate),
         `SE` = sprintf('%.4f', se),
         `z vs 0.05` = sprintf('%+.1f', (rate - 0.05) / se)) |>
  select(Design, cells, `Rejection rate`, SE, `z vs 0.05`)
knitr::kable(t1, booktabs = TRUE,
  caption = sprintf('Empirical Type I error at the null
    ($c_{bm} = 0$), pooled within trial design over both
    architectures, both sample sizes and all three carryover
    half-lives (%d replicates per cell). Nominal level 0.05.',
    NREPS))
```

Type I error is not uniform across designs, and the departure is
large enough to require comment. Pooled within design, the rejection
rate is
`r sprintf('%.3f', with(subset(S, c_bm == 0 & design == 'CO'), sum(power * NREPS) / (length(power) * NREPS)))`
under CO,
`r sprintf('%.3f', with(subset(S, c_bm == 0 & design == 'Hybrid'), sum(power * NREPS) / (length(power) * NREPS)))`
under Hybrid, and
`r sprintf('%.3f', with(subset(S, c_bm == 0 & design == 'OL+BDC'), sum(power * NREPS) / (length(power) * NREPS)))`
under OL+BDC, against a nominal $0.05$.

Two features of this pattern matter. It is architecture-independent:
the two arms agree closely within every design, which is expected,
since at $c_{bm} = 0$ the two architectures generate identical data.
And it is monotone in how much off-drug information the design
supplies, running from anti-conservative under CO to markedly
conservative under OL+BDC.

The consequence for this paper is a restriction on what may be
compared. Power figures from different designs are computed under
tests with materially different actual sizes, so the cross-design
power comparisons in Table \@ref(tab:tab-main) should be read as
descriptive rather than as equal-alpha comparisons. The
within-design, across-half-life comparisons that carry the paper's
argument are unaffected, since the test size is held fixed along that
axis.

We did not diagnose the mechanism. The companion carryover paper
attributes a milder conservatism in the same analysis model to a
working covariance that does not match the data-generating one, and a
cluster-robust standard error removes it there. Whether the same
correction would flatten the design dependence seen here is a
reasonable conjecture and an untested one.

\subsection{3.4 Comparison with the three-component setting}

Absolute power under the two-component DGP is higher than paper 01
reports under three components, and the explanation is arithmetic.
At the reference calibration PB carries a standard deviation of 10,
against 10 for TV and 8 for BR, further scaled by the expectancy
weight $\eta$. Removing it reduces the residual variation against
which a fixed interaction is tested; on a variance argument alone,
standard errors shrink and power rises correspondingly. Because
$\eta$ is $0.5$ at blinded occasions, the share removed is
design-dependent rather than a single figure, and is largest under
designs with the most open-label time.

This comparison is qualitative and should not be given a number.
Paper 01's published figures were produced under the earlier
carryover recursion of Section 2.3 and a partly different grid, so
the difference between the two sets of power values confounds the
component reduction with at least two other changes. Separating them
would require a matched three-component run under the present code,
which we did not perform.

What can be said without a matched run is that nothing in the
\emph{structure} of the comparison changed. The ranking of
architectures within each design, the direction of the carryover
effect, the insensitivity of CO, and the design ordering are all as
paper 01 reports them.

\section{4. Discussion}

\subsection{4.1 What the TV component is for}

The results support a narrow claim and not a broad one.

The narrow claim is that TV is inert for the biomarker-treatment
interaction estimand, and here the claim is unconditional. No
biomarker-TV parameter exists anywhere in the implementation, so the
coupling is structurally zero rather than set to zero, and paper 06's
omitted-variable identity shows that a component uncorrelated with
the biomarker cannot displace the biomarker slope regardless of its
magnitude. The present results are consistent with that identity and
extend it in one respect: paper 06 established it under a single
architecture, and it now holds under both.

The broad claim, which we do not make, is that TV is dispensable.
Three of its roles survive intact. It is required whenever the
estimand is the attribution of response to pharmacology versus belief
versus natural history, which is paper 06's subject. It is required
whenever a candidate biomarker may correlate with natural-history
drift, a case the implementation cannot currently express at all and
which Section 4.3 takes up. And it contributes realism to absolute
power figures calibrated against a clinical reference trajectory, so
any paper reporting power as a design input rather than as a
comparison should retain it.

The reduction is also, and separately, a departure from the modelling
conventions of the field. Section 4.5 addresses that.

\subsection{4.2 When the reduction is appropriate}

We suggest the following division.

Use the two-component DGP for methodological work whose object is a
comparison: between DGP architectures, between analysis
specifications, between trial designs, or between inference
procedures. In that setting the third component adds variance,
dimension, and positive-definiteness constraints without adding
anything the comparison depends on, and its absence makes the
mechanism easier to see. The saving is real but modest: the
correlation matrix falls from $26 \times 26$ to $18 \times 18$, and
Appendix C shows the feasible range of $c_{bm}$ widens by only about
$0.01$.

Use the three-component DGP whenever a number is intended to be read
as an estimate of what a real trial would achieve, or whenever the
decomposition itself is the object of study.

The reduction should not be adopted for the reason that first
suggests it. One might expect that deleting a third of the rows and
columns would materially enlarge the feasible parameter space.
Appendix C.3 shows why it does not: under compound symmetry the
smallest eigenvalue of the component block is
$(1-\rho) - (c_1 - c_{\times})$, which contains neither the number of
occasions nor the number of components. The case for the reduction is
transparency, cost, and conformity with the modelling conventions of
the crossover and N-of-1 literature (Section 1.1), not numerical
headroom.

\subsection{4.3 A gap this paper exposes}

Paper 06 states its central identity over both non-pharmacological
components, and its conclusion recommends decomposition when a
candidate biomarker may correlate with PB \emph{or TV}. The package
implements a biomarker-PB coupling and no biomarker-TV counterpart.
The TV half of that claim has therefore never been exercised, and
could not be without new code.

This bears on the present paper in a way that is favourable but
should not be overstated. The component this paper deletes is the one
whose coupling has never been exercised and cannot be, so nothing in
the reduced DGP depends on the untested half of paper 06's claim, and
no setting of any existing parameter can invalidate the reduction.
The component it retains, PB, is the one whose coupling has been
exercised and shown to matter: paper 06's contaminated-biomarker
study drives the one-component bias to $+0.51$ at a biomarker-PB
correlation of $0.45$, and the present DGP can still represent that
failure mode.

The overstatement to avoid is treating this as evidence that TV is
unimportant. It shows only that the implementation cannot currently
give TV a biomarker coupling. Under paper 06's identity a biomarker
correlated with natural-history drift would displace the
one-component slope exactly as a PB-correlated biomarker does, so the
safety of this reduction rests on a gap in the software rather than
on a property of natural history. Closing that gap, by adding the
missing parameter and running the untested half of paper 06's claim,
would also tell us how much the present reduction is really worth.

\subsection{4.4 Limitations}

The comparison inherits paper 01's restriction to a single
prazosin-PTSD parameter calibration, so absolute power values should
be read as illustrative and the ranking as the more portable result.

The design-dependent Type I error of Section 3.3 limits cross-design
comparison, as set out there. It was not diagnosed.

The reduction removes the component that the crossover and N-of-1
literature would most expect a model to retain. A trend term is
standard in that literature; a modelled belief trajectory is not.
Section 4.5 weighs this against the structural argument of Section
1.1 rather than resolving it.

The comparison inherits paper 01's restriction to a single
prazosin-PTSD parameter calibration, so absolute power values should
be read as illustrative and the ranking as the more portable result.
PTSD is, on paper 06's own taxonomy, a condition for which
natural-history drift is expected to be structured and non-stationary
rather than absent, since trauma anniversaries and selection at
moments of clinical engagement both contribute. A reduction that
deletes TV is therefore harder to defend as a clinical model of this
indication than as a methodological device, and we intend it only as
the latter.

The design-dependent Type I error of Section 3.3 limits cross-design
comparison, as set out there. It was not diagnosed.

The reduction removes TV rather than flattening its trajectory. The
two are not equivalent, since the code sets component standard
deviations independently of the Gompertz maximum: a flattened TV mean
leaves the variance contribution and the covariance blocks in place.
A study wanting the simpler mental picture without the change in
residual variance should flatten rather than remove.

No matched three-component grid was run under the present code, so
Section 3.4's comparison with paper 01 remains qualitative.

The two architectures are simulated in separate replicate loops from
different seeds, since the same random draw does not correspond to
the same trial under different data-generating processes. Comparisons
between architectures are therefore unpaired, and in any case not
interpretable on an absolute scale at a common nominal $c_{bm}$.

\subsection{4.5 Which reduction to prefer}

The two papers reach the same substantive conclusion about the
architecture comparison, so the choice between them is not a choice
about results. It is a choice about which assumption to carry, and
the considerations point in opposite directions.

The considerations divide on three axes rather than one.

\paragraph{Safety.} No biomarker-TV parameter exists anywhere in the
implementation, so the present deletion is valid under every
parameter setting the software can express. The biomarker-PB coupling
does exist, as an optional contamination parameter, so paper 13's
reduction is valid only while that parameter is at its default of
zero; a study setting it nonzero cannot use that paper's DGP at all.
This axis favours the present paper.

\paragraph{Design structure.} The asymmetry here is one-sided.
Dropping TV renders nothing inert: the calendar-time argument that
drives the TV mean also drives the AR(1) week gaps of every
within-component block, so unequal visit spacing survives, and the
expectancy weight continues to act through PB. Dropping PB renders
the expectancy weight inert, because $\eta$ enters the
data-generating process only through the PB mean and PB standard
deviation, so the open-label versus blinded contrast disappears from
the generated data. This axis also favours the present paper.

\paragraph{Convention.} Against those two stands the argument paper
13 makes. A time trend is a standard element of crossover and N-of-1
models, while a modelled placebo-belief trajectory is an innovation
of this program rather than an inherited convention, and the placebo
literature regards the belief and natural-history channels as
separable only under three-arm or balanced-placebo designs that most
trials do not field. The present reduction deletes the conventional
component and retains the contested one. This axis favours paper 13.

\paragraph{Our reading.} For work internal to this program, where the
contaminated-biomarker case is live and the blinding contrast is used,
dropping TV is the better choice, because it is safe under every
parameter setting and costs no design structure. For work intended to
be read by the wider N-of-1 methodological community, dropping PB
presents a more familiar model and avoids asserting a contested
decomposition, at the cost of a conditionality and of an inert
expectancy weight. Neither dominates, and a reader should choose by
audience and by whether biomarker contamination is in scope. What
should not happen is that either reduction be adopted without
noticing that a choice was made.

\section{5. Conclusions}

1. \textbf{The architecture comparison survives the reduction.} Under
   a BR + PB data-generating process, covariance moderation continues
   to lose power as carryover grows while mean moderation does not.
   At $c_{bm} = 0.30$ and $N = 70$ the cost is $0.194$ against
   $0.040$ under Hybrid, and $0.196$ against $0.078$ under OL+BDC.

2. \textbf{TV is a variance term, unconditionally.} No biomarker-TV
   parameter exists in the implementation, so its coupling is
   structurally zero rather than set to zero, and its removal cannot
   change the estimand under any parameter setting. Removal raises
   absolute power without disturbing the ranking, the direction of
   the carryover effect, or the ordering among designs.

3. \textbf{The reference cell should not be used to read the
   contrast.} At $c_{bm} = 0.45$ and $N = 70$ every cell exceeds
   $0.89$ and the architecture difference is compressed roughly
   sixfold by a ceiling effect. Comparability with paper 01's
   presentation is not a sufficient reason to report it as the
   headline.

4. \textbf{Type I error is design-dependent.} Pooled rejection rates
   run from $0.061$ under CO to $0.021$ under OL+BDC against a
   nominal $0.05$, with the two architectures agreeing closely within
   each design. Cross-design power comparisons are correspondingly
   descriptive rather than equal-alpha.

5. \textbf{The reduction is appropriate for comparative
   methodological work and not for design estimates}, and it should
   be adopted for transparency and cost rather than for numerical
   headroom, which Appendix C shows it does not supply.

6. \textbf{This reduction preserves the design structure; the
   alternative does not.} Retaining PB retains the expectancy weight,
   so the open-label versus blinded contrast survives and no feature
   of the trial design becomes inert. The price is that the model
   departs from the convention of the crossover and N-of-1
   literature, which carries a trend term and handles placebo by
   design. Paper 13 makes the opposite trade; Section 4.5 sets the
   two against each other.

\section{Conflict of interest}

The authors declare no conflicts of interest.

\section{Funding}

This work received no specific funding.

\section{Reproducibility}

The simulation driver is
\texttt{analysis/scripts/two-component-dgp/01-run-grid.R} and its
output is versioned at
\texttt{analysis/data/14-drop-tv-grid.rds}. The driver sets a
master seed and derives a per-cell seed from it. Every number quoted
in the results is read from that file at render time rather than
typed in.

The two-component decomposition is supported by the
\texttt{components} argument added to \texttt{generateData()} and
\texttt{buildSigma()} in \texttt{R/generateData.R}. It defaults to
\texttt{c("tv", "pb", "br")}, so existing calls are unaffected;
passing \texttt{c("pb", "br")} produces the reduced matrix used here.
The argument requires that \texttt{"br"} be retained, since BR
carries the interaction channel in both architectures, and
canonicalizes the ordering so that label and standard-deviation
sequence is stable regardless of how the caller writes it. The
package test suite passes unchanged under the default.

The corrected carryover recursion of Section 2.3 is applied in
\texttt{R/generateData.R} and
\texttt{implementations/tidyverse/R/functions.R}. Other
implementation collections retain the earlier form; see
\texttt{02-deviations.md} in the driver directory.

\bigskip\noindent\rule{\textwidth}{0.4pt}\bigskip

\section{Appendix A. Why mean and covariance moderation}

\subsection{A.1 The two channels exhaust the Gaussian case}

The simulations draw participant trajectories from a multivariate
normal distribution, and a multivariate normal is completely
determined by its mean vector $\mu$ and its covariance matrix
$\Sigma$. Any biomarker-treatment interaction written into such a
data-generating process must therefore enter through $\mu$, through
$\Sigma$, or through both. Mean moderation is the first route,
covariance moderation the second, and the combined architecture
developed in a companion paper [@pmsimstats-paper11] is the third.
Within the Gaussian family there is no fourth place to put the
interaction. This is why the two architectures are treated as a pair
rather than as two entries in an open-ended list: they are not two
options among many, but the two components of a decomposition.

Reducing the response decomposition from three components to two does
not disturb this argument. The exhaustiveness claim concerns where an
interaction can be written into a Gaussian joint distribution, not
how many latent components that distribution is built from. Dropping
TV removes rows from $\Sigma$; it does not create a third place for
the interaction to live, nor remove either of the two that exist.

The exhaustiveness claim is specific to the Gaussian family and
should not be read more broadly. A data-generating process built on a
finite mixture, or on explicit latent responder classes, can carry an
interaction in features that neither $\mu$ nor $\Sigma$ captures for
a single component, and Senn notes finite mixtures as an alternative
variance-component model [@senn2019samplesize]. Covariance moderation
is best understood as the second-moment reduced form of such a
mixture, which is why the latent-class treatment is developed
separately [@pmsimstats-paper03].

\subsection{A.2 What each channel asserts}

The two channels are not merely different parameterizations of one
mechanism; each commits to a distinct claim about what the biomarker
is.

Mean moderation asserts that the biomarker fixes a participant's drug
effect. Two participants sharing a biomarker value share an expected
treatment effect exactly, and any difference between their realized
responses is measurement error. This is the appropriate commitment
when the biomarker measures pathway engagement directly, or stands in
a causal path between treatment and response.

Covariance moderation asserts that the biomarker predicts response
only probabilistically. Two participants sharing a biomarker value
may still respond differently, because the biomarker indexes an
unobserved responder status imperfectly rather than determining
response. This is the appropriate commitment when the biomarker is a
statistical proxy, which is the situation the prazosin and PTSD
application was designed around.

The distinction is substantive rather than technical. It determines
what a fitted interaction coefficient means, and how much power
carryover costs.

\subsection{A.3 Standing in the literature}

Mean moderation is the prevailing convention wherever
biomarker-treatment interactions are simulated. Haller and Ulm
[@haller2018] place the interaction in a regression mean, as do the
oncology design literatures on enrichment and predictive-marker
trials [@simon2010; @freidlin2014]. Grenet and colleagues
[@grenet2020] use the same form for a crossover. That literature is,
however, overwhelmingly parallel-group.

The N-of-1 methodological literature does not adopt mean moderation,
because it does not model an observed moderator at all. Its standard
formulation carries a random participant-by-treatment interaction, a
variance component describing unexplained heterogeneity with no
covariate attached [@senn2019samplesize; @senn2016mastering;
@wang2019]. Biomarker-validation simulation for N-of-1 designs is
therefore close to unoccupied territory rather than a settled field
with a convention to follow, and Hendrickson et al.
[@hendrickson2020] is its principal entry. The two-channel
decomposition of A.1 is adopted precisely because no established
N-of-1 convention exists to inherit.

\subsection{A.4 Holding the analysis model fixed}

Both architectures are analyzed with the identical model of Section
2.1. This is deliberate and worth stating explicitly, because a
data-generating process and an analysis model are separate objects,
and a comparison that varies both attributes differences to neither.
The continuous exposure variable \texttt{Dbc} loses contrast variance
as carryover grows, and a binary on-drug indicator does not. A
comparison in which one architecture carried a continuous regressor
and the other a binary one would confound that effect with the
architecture contrast it was meant to isolate. Holding the analysis
model fixed means every difference reported in Section 3 is
attributable to the data-generating process alone.

The same argument governs the reduction studied here. The analysis
model does not reference the component decomposition at all: it sees
only $S_x$, and $S_x$ is a sum. Removing TV therefore changes what
the analysis is fitted to without changing what it fits, which is the
property that makes the comparison in Section 3 interpretable.

\bigskip\noindent\rule{\textwidth}{0.4pt}\bigskip

\section{Appendix B. Correlation structure under the two-component
decomposition}

This appendix sets out the correlation matrix the reduced
data-generating process assembles, in the notation of paper 01's
Appendix B so that the two can be read side by side. Symbols follow
the variable names used in the code: $\mathrm{BM}$ is the biomarker,
written $B_i$ in Section 2.1, and $BR_p$, $PB_p$ the response
components at occasion $p$.

Mean moderation appears below only through the statement that its
coupling vector is zero. It writes no biomarker-response correlation
into the covariance at all; its interaction is carried in the mean
instead, as Section 2.2.1 sets out.

\subsection{B.1 Variable ordering and block partition}

Each participant's trajectory is drawn from a single multivariate
normal over an ordered variable list. With $n$ measurement occasions
at cumulative weeks $w_1 < \cdots < w_n$, the two-component vector is

\[
X = \bigl(\,\mathrm{BM},\; \mathrm{BL},\;
PB_1,\dots,PB_n,\;
BR_1,\dots,BR_n \,\bigr)^{\!\top},
\qquad \dim X = 2 + 2n ,
\]

against $2 + 3n$ under the three-component decomposition. The
ordering is component-major: all $n$ occasions of one component are
contiguous, rather than all components at one occasion. Writing $A_c$
for the $n \times n$ within-component block of component $c \in
\{pb, br\}$ and $C_{pb,br}$ for the single cross-component block, the
correlation matrix partitions as

\[
R \;=\;
\begin{pmatrix}
1 & 0 & 0^{\top}   & b^{\top} \\[2pt]
0 & 1 & 0^{\top}   & 0^{\top} \\[2pt]
0 & 0 & A_{pb}     & C_{pb,br} \\[2pt]
b & 0 & C_{br,pb}  & A_{br}
\end{pmatrix}.
\]

Three features carry over unchanged from the three-component case.
The baseline term $\mathrm{BL}$ is uncorrelated with every other
variable, so its row and column are those of the identity. The
biomarker $\mathrm{BM}$ couples only to the response component $BR$,
through the vector $b \in \mathbb{R}^{n}$; its correlation with $PB$
is zero throughout this paper, by the default setting of the optional
contamination parameter rather than by structural necessity. And $b$ is the only place in $R$ where
the architectures differ.

The third feature deserves emphasis because it is what licenses the
reduction. In the three-component matrix the biomarker's row is
$(1, 0, 0^{\top}, 0^{\top}, b^{\top})$: zero against both $TV$ and
$PB$. Deleting the $TV$ rows and columns therefore deletes only zeros
from that row, and the coupling vector is untouched, entry for entry.

Here the argument requires no qualification. The $TV$ entries are
zero because the implementation provides no parameter that could make
them otherwise, so the reduction is valid under every setting the
software can express. The companion paper's reduction deletes the
$PB$ block instead, whose entries are zero by a default rather than
by construction, and its Appendix B.1 records the corresponding
caveat.

\subsection{B.2 Within-component blocks $A_c$}

Let $\rho_c$ denote the autocorrelation parameter of component $c$
(\texttt{c.pb}, \texttt{c.br} in the code). The correlation decays
with the elapsed time between occasions, using the cumulative week
gap rather than the index difference:

\[
[A_c]_{ij} = \rho_c^{\,|w_i - w_j|}.
\]

For $n = 4$, writing $d_{ij} = |w_i - w_j|$,

\[
A_c =
\begin{pmatrix}
1 & \rho_c^{d_{12}} & \rho_c^{d_{13}} & \rho_c^{d_{14}}\\
\rho_c^{d_{12}} & 1 & \rho_c^{d_{23}} & \rho_c^{d_{24}}\\
\rho_c^{d_{13}} & \rho_c^{d_{23}} & 1 & \rho_c^{d_{34}}\\
\rho_c^{d_{14}} & \rho_c^{d_{24}} & \rho_c^{d_{34}} & 1
\end{pmatrix}.
\]

Because the exponent uses calendar time, unequally spaced designs
produce unequal decay across consecutive occasions. In the Hybrid
design, whose occasions fall at weeks $4, 8, 9, 10, 11, 12, 16, 20$,
consecutive gaps of one week during the discontinuation phase yield
much higher adjacent correlations than the four-week gaps at either
end.

We note that the code uses the cumulative week gap, not the occasion
index difference, and that this choice is consequential: Appendix C.4
reports that the index-based form would cost roughly a fifth of the
feasible range of $c_{bm}$.

\subsection{B.3 The cross-component block $C_{pb,br}$}

Let $c_1$ denote the same-occasion cross-component correlation
(\texttt{c.cf1t}) and $c_{\times}$ the different-occasion
cross-component correlation (\texttt{c.cfct}). The same-occasion
entry is a constant and the different-occasion entries inherit the
AR(1) decay:

\[
[C_{pb,br}]_{ij} =
\begin{cases}
c_1, & i = j\\
c_{\times}\,\rho^{\,|w_i - w_j|}, & i \neq j .
\end{cases}
\]

Under the three-component decomposition there are three such blocks,
$C_{tv,pb}$, $C_{tv,br}$ and $C_{pb,br}$, all of this form. The
reduction retains the last and deletes the first two.

\subsection{B.4 Biomarker coupling vector $b$}

The vector $b$ carries the biomarker-treatment interaction under
covariance moderation. Its entry is gated on the drug indicator and
decays with time since discontinuation $t_{sd,p}$ at rate $\lambda =
\ln 2 / t_{1/2}$:

\[
b_p =
\begin{cases}
c_{bm}, & \text{occasion } p \text{ on drug}\\
c_{bm}\,e^{-\lambda t_{sd,p}}, & \text{occasion } p \text{ off drug} .
\end{cases}
\]

Under mean moderation $b \equiv 0$ and the interaction is carried by
the displacement of $\mu_{BR}$ described in Section 2.2.1. The two
architectures are therefore distinguished, in the language of this
appendix, by exactly one object: whether $b$ is populated or zero.

\subsection{B.5 A worked example}

To make the blocks concrete, take four occasions drawn from the
Hybrid design at cumulative weeks $w = (10, 11, 12, 16)$, with the
first and fourth on drug and the second and third off it, so that
$t_{sd} = (0, 1, 2, 0)$. Set $\rho_{br} = 0.6$, $\rho_{pb} = 0.4$,
$c_1 = 0.3$, $c_{\times} = 0.1$, $c_{bm} = 0.45$, and a carryover
half-life of one week, giving $\lambda = \ln 2 \approx 0.693$. The
matrix of week gaps is

\[
\bigl(|w_i - w_j|\bigr) =
\begin{pmatrix}
0 & 1 & 2 & 6\\
1 & 0 & 1 & 5\\
2 & 1 & 0 & 4\\
6 & 5 & 4 & 0
\end{pmatrix}.
\]

The unequal spacing matters: occasions one to three are a week apart,
while the fourth sits four weeks after the third.

\subsubsection{B.5.1 Within-component block $A_{br}$}

With $\rho_{br} = 0.6$,

\[
A_{br} =
\begin{pmatrix}
1 & 0.6000 & 0.3600 & 0.0467\\
0.6000 & 1 & 0.6000 & 0.0778\\
0.3600 & 0.6000 & 1 & 0.1296\\
0.0467 & 0.0778 & 0.1296 & 1
\end{pmatrix}.
\]

The fourth occasion is nearly uncorrelated with the first three, a
direct consequence of the four-week gap: $0.6^{4} = 0.1296$ against
$0.6^{1} = 0.6$ for an adjacent pair.

\subsubsection{B.5.2 Cross-component block $C_{pb,br}$}

With $c_1 = 0.3$, $c_{\times} = 0.1$ and the decay taken at
$\rho = 0.6$,

\[
C_{pb,br} =
\begin{pmatrix}
0.3000 & 0.0600 & 0.0360 & 0.0047\\
0.0600 & 0.3000 & 0.0600 & 0.0078\\
0.0360 & 0.0600 & 0.3000 & 0.0130\\
0.0047 & 0.0078 & 0.0130 & 0.3000
\end{pmatrix}.
\]

The block is symmetric here because the two components share a decay
rate in this example; in general it need not be.

\subsubsection{B.5.3 Biomarker vector $b$}

Occasions one and four are on drug, occasions two and three off it
with $t_{sd} = 1$ and $2$. With $c_{bm} = 0.45$ and $\lambda = \ln 2$,

\[
b = \bigl(0.4500,\; 0.2250,\; 0.1125,\; 0.4500\bigr)^{\!\top}.
\]

The off-drug entries halve with each additional week since
discontinuation, and the fourth occasion returns to full strength
because the participant is back on drug. It is this alternation that
carries the interaction, and its erosion under carryover that
Section 3 measures.

\subsection{B.6 What the reduction removes, and what it does not}

Relative to the three-component matrix, the reduction removes:

\begin{itemize}
\item one within-component block, $A_{tv}$ ($n \times n$);
\item two cross-component blocks, $C_{tv,pb}$ and $C_{tv,br}$;
\item $n$ rows and columns from $R$, taking it from $2+3n$ to
  $2+2n$: at $n = 8$, from $26 \times 26$ to $18 \times 18$.
\end{itemize}

It removes no entry of $b$, because the biomarker never couples to
$TV$ under any setting. It removes no parameter from the architecture
definitions of Section 2.2, because neither references $TV$. And it
renders nothing inert: the expectancy weight $\eta$ continues to act
through the PB component, and the calendar-time argument that drove
the TV mean also drives the AR(1) week gaps, which survive
(Section 2.2.3).

One distinction is worth recording, because it is easy to conflate
two operations that are not equivalent. Removing the component is not
the same as flattening its trajectory. The code sets component
standard deviations independently of the Gompertz maximum, so setting
that maximum to zero yields a flat $TV$ mean while leaving the $TV$
variance contribution, its block $A_{tv}$, and its cross-component
blocks in place. A study wanting the simpler mental picture without
the change in residual variance should flatten; a study wanting the
smaller matrix should remove. This paper removes.

\bigskip\noindent\rule{\textwidth}{0.4pt}\bigskip

\section{Appendix C. Positive definiteness and the feasible range of
$c_{bm}$}

The correlation matrix of Appendix B must be positive definite for
the multivariate normal draw to be valid. It is not automatically so:
the moderation parameter $c_{bm}$ is bounded above, and the bound
depends on the correlation structure, the trial design, and the
number of occasions. This appendix derives the bound, gives a closed
form for the compound-symmetric case, and reports where the reduction
of Appendix B does and does not relax it.

The motivation is practical. Simulation drivers in this program
repair a non-positive-definite matrix silently, by projection onto
the nearest positive-definite matrix, so a grid cell that violates
the bound does not fail. It quietly becomes a different
data-generating process, with a realized biomarker-response
correlation no longer equal to the nominal $c_{bm}$. Knowing the
bound in advance is therefore a precondition for trusting a grid.

\subsection{C.1 The constraint is a Schur complement on the biomarker
row}

Partition the correlation matrix of Appendix B.1 by separating the
biomarker from everything else. Because $\mathrm{BL}$ is uncorrelated
with all other variables, it contributes an identity block that
factors out of every determinant and may be dropped. What remains is

\[
R' =
\begin{pmatrix}
1 & \tilde b^{\top} \\
\tilde b & M
\end{pmatrix},
\qquad
M =
\begin{pmatrix}
A_{pb} & C_{pb,br}\\
C_{br,pb} & A_{br}
\end{pmatrix},
\]

where $M$ is the $Kn \times Kn$ component block ($K = 2$ here, $K =
3$ under the full decomposition) and $\tilde b \in \mathbb{R}^{Kn}$
is the biomarker coupling extended by zeros: it equals $b$ on the
$BR$ positions and vanishes elsewhere.

By the Schur complement, $R'$ is positive definite if and only if $M$
is positive definite and

\[
1 - \tilde b^{\top} M^{-1} \tilde b > 0 .
\]

Writing $\tilde b = c_{bm} v$, where $v$ is the unit-scaled coupling
pattern with $v_p = \phi(t_{sd,p})$ on the $BR$ positions and zero
elsewhere, the condition becomes $c_{bm}^{2}\, v^{\top} M^{-1} v <
1$, giving the exact ceiling

\begin{equation}
c_{bm}^{\ast} \;=\; \bigl(v^{\top} M^{-1} v\bigr)^{-1/2}.
\label{eq:ceiling}
\end{equation}

Three consequences follow immediately, and they organize the rest of
this appendix. The ceiling depends on the component block only
through the quadratic form $v^{\top} M^{-1} v$, so components that
are weakly coupled to $BR$ contribute little to it. It depends on the
design through $v$, which encodes the on-drug and off-drug pattern.
And it is unaffected by mean moderation, under which $b \equiv 0$ and
the constraint reduces to positive definiteness of $M$ alone.

\subsection{C.2 Closed form for the component block under compound
symmetry}

When the within-component blocks are compound symmetric rather than
AR(1), $M$ admits an exact spectral decomposition. This is the
structure of the published implementation discussed in paper 01, and
it is worth recording because it explains a result of C.3 that is
otherwise surprising.

Under compound symmetry with a common within-component correlation
$\rho$, common same-occasion cross-component correlation $c_1$, and
common different-occasion cross-component correlation $c_{\times}$,
the component block is a Kronecker sum,

\begin{equation}
M \;=\; A \otimes I_n \;+\; B \otimes (J_n - I_n),
\label{eq:kron}
\end{equation}

where $J_n$ is the $n \times n$ matrix of ones and $A$, $B$ are
$K \times K$:

\[
A = I_K + c_1 (J_K - I_K), \qquad
B = \rho I_K + c_{\times}(J_K - I_K).
\]

$A$ describes the same-occasion structure across components, $B$ the
across-occasion structure. Since $J_n - I_n$ has eigenvalues $n-1$
once and $-1$ with multiplicity $n-1$, and $I_n$ commutes with it,
the $Kn$ eigenvalues of $M$ collapse to those of two $K \times K$
reductions:

\[
\mathrm{eig}(M) \;=\;
\underbrace{\mathrm{eig}\bigl(A + (n-1)B\bigr)}_{\text{once each}}
\;\cup\;
\underbrace{\mathrm{eig}\bigl(A - B\bigr)}_{\text{multiplicity } n-1}.
\]

Both reductions are of the form $\alpha I_K + \beta (J_K - I_K)$,
whose eigenvalues are $\alpha + (K-1)\beta$ once and $\alpha - \beta$
with multiplicity $K-1$. Substituting gives all four distinct
eigenvalues in closed form:

\[
\begin{aligned}
\text{all-ones time direction:}\quad
&\bigl[1 + (n-1)\rho\bigr] + (K-1)\bigl[c_1 + (n-1)c_{\times}\bigr],\\
&\bigl[1 + (n-1)\rho\bigr] - \bigl[c_1 + (n-1)c_{\times}\bigr]
  \quad (K-1 \text{ times}),\\[4pt]
\text{time-contrast directions:}\quad
&(1-\rho) + (K-1)(c_1 - c_{\times}),\\
&(1-\rho) - (c_1 - c_{\times})
  \quad (K-1 \text{ times, each with multiplicity } n-1).
\end{aligned}
\]

At the program's reference values $\rho = 0.7$, $c_1 = 0.2$,
$c_{\times} = 0.1$, $n = 8$, the three-component block has exactly
four distinct eigenvalues, $7.7$, $5.0$, $0.5$ and $0.2$, the last
with multiplicity $2(n-1) = 14$.

\subsection{C.3 Why the component count barely matters}

The smallest eigenvalue in C.2 is generally the last,

\begin{equation}
\lambda_{\min}(M) = (1-\rho) - (c_1 - c_{\times}),
\label{eq:lmin}
\end{equation}

and the striking feature of \eqref{eq:lmin} is what does not appear
in it. It contains neither $n$ nor $K$. Under compound symmetry the
positive-definiteness of the component block is independent of both
the number of occasions and the number of components. Adding a third
component does not make $M$ harder to keep positive definite; nor
does adding occasions.

This is the analytic explanation for a result that would otherwise be
puzzling. One might expect that deleting a third of the rows and
columns from a correlation matrix would materially enlarge the
feasible parameter space. It does not, and \eqref{eq:lmin} shows why:
the binding structural constraint never depended on the count in the
first place.

Two further readings of \eqref{eq:lmin} are worth recording. The
operative cross-component quantity is the \emph{difference}
$c_1 - c_{\times}$, not either coefficient separately, so keeping the
same-occasion and across-occasion couplings close to one another
matters more than keeping either small. And positive definiteness
requires $\rho < 1 - (c_1 - c_{\times})$, which at the reference
values is $\rho < 0.9$: a compound-symmetric structure cannot
represent very high autocorrelation at all.

Under AR(1) the corresponding statement is different, and no equally
compact closed form is available, because $A_c$ is Toeplitz rather
than exchangeable and does not commute with the cross-component
structure in the way \eqref{eq:kron} requires. Numerically, the AR(1)
component block becomes progressively more ill-conditioned as
occasions are added, since each new occasion is strongly correlated
with its neighbours. The number of \emph{occasions} therefore does
constrain $c_{bm}$ under AR(1), while the number of \emph{components}
still does not.

\subsection{C.4 Numerical ceilings}

Evaluating \eqref{eq:ceiling} directly at the program's reference
parameters ($\rho = 0.7$, $c_1 = 0.2$, $c_{\times} = 0.1$, $n = 8$,
AR(1) in cumulative weeks), minimized across each design's
randomization paths:

| Design | $t_{1/2}$ | Three components | Two components |
|---|---|---|---|
| CO | 0.0 | 0.619 | 0.627 |
| CO | 0.5 | 0.619 | 0.627 |
| CO | 1.0 | 0.619 | 0.627 |
| Hybrid | 0.0 | 0.456 | 0.469 |
| Hybrid | 0.5 | 0.506 | 0.516 |
| Hybrid | 1.0 | 0.544 | 0.552 |
| OL+BDC | 0.0 | 0.452 | 0.464 |
| OL+BDC | 0.5 | 0.500 | 0.509 |
| OL+BDC | 1.0 | 0.533 | 0.540 |

The reduction buys between $0.008$ and $0.013$ of feasible range.
This is consistent with C.3: the deleted blocks were coupled to $BR$
only through $c_1$ and $c_{\times}$, both small, so they contributed
little to the quadratic form $v^{\top} M^{-1} v$ that sets the
ceiling.

Three further observations follow from evaluating \eqref{eq:ceiling}
under variations of the structure, none of which is specific to the
reduction.

First, the ceiling rises with the carryover half-life within the two
non-crossover designs, from $0.456$ to $0.544$ under Hybrid. A longer
half-life makes the coupling vector $v$ smoother, and a smoother $v$
aligns better with the well-conditioned directions of $M$. The
crossover design is insensitive to $t_{1/2}$ because its off-drug
occasions are spaced at least 2.5 weeks apart, so $\phi(t_{sd})$ has
decayed to near zero at every one of them regardless.

Second, the number of occasions matters considerably more than the
number of components. Holding the design shape fixed and varying $n$
alone, the ceiling falls from approximately $0.81$ at $n = 4$ to
$0.58$ at $n = 12$ under a three-component structure. That is an
order of magnitude larger than the effect of deleting an entire
component.

Third, the exponent scale in $A_c$ is consequential. Using the
occasion-index difference $|i-j|$ in place of the cumulative week gap
$|w_i - w_j|$, an otherwise identical AR(1) structure, lowers the
worst-cell ceiling from approximately $0.45$ to $0.35$. The code uses
the week gap; documentation elsewhere in this project has described
it as the index difference, and the two are not interchangeable.

\subsection{C.5 Practical implications}

For the present paper, \eqref{eq:ceiling} confirms that the grid of
Section 2.4 is feasible throughout: the largest moderation level used
is $c_{bm} = 0.45$, against a worst-cell ceiling of $0.464$ under the
two-component structure. The margin is real but not generous, and a
study wishing to extend the grid past $c_{bm} = 0.45$ should evaluate
\eqref{eq:ceiling} first rather than rely on the simulation to
complain, since it will not.

More generally, the appendix identifies which levers move the
feasible range and which do not. The number of latent components does
not, and neither does the choice between two and three of them. The
number of measurement occasions does, substantially. The
within-component correlation $\rho$ does, non-monotonically, with the
reference value $0.7$ close to the maximum of the ceiling. And the
cross-component coefficients matter through their difference rather
than their magnitudes.

\paragraph{Provenance.} The Kronecker identity \eqref{eq:kron} and
the eigenvalue decomposition of C.2 were verified numerically against
direct construction for $K \in \{2,3\}$ and $n \in \{4,8,12\}$. The
ceilings in C.4 were computed from \eqref{eq:ceiling} and agree with
an independent grid search over $c_{bm}$ in steps of $0.01$ to within
the grid step. The observations in C.4 concerning occasion count,
exponent scale, and $\rho$ were obtained by the same grid search
under a reimplemented matrix builder rather than the production
\texttt{buildSigma}, and should be read as characterizing the
structure rather than as production audit results.

# References

::: {#refs}
:::
