---
title: "Carryover Effects in Hybrid N-of-1 Design"
author: "nof1power"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_width: 7
    fig_height: 5
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    fig_width: 7
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{Carryover Effects in Hybrid N-of-1 Design}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
knit: (function(input, ...) { d <- dirname(input); while (!file.exists(file.path(d, 'tools', 'stamp-render.R')) && d != dirname(d)) d <- dirname(d); source(file.path(d, 'tools', 'stamp-render.R'))$value(input) })
---

```{r setup, include=FALSE}
# Set chunk options
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5,
  out.width = "90%",
  fig.align = "center"
)

# Create a directory for caching results if it doesn't exist
if (!dir.exists("temp")) {
  dir.create("temp", recursive = TRUE)
}

# Flag to use pre-cached results for faster rendering
use_cached_results <- TRUE

# Load required packages
library(dplyr)
library(tidyr)
library(purrr)
library(nlme)
library(knitr)
library(parallel)
library(grid)
library(tools)
library(MASS)
library(corpcor)

# Source the required functions directly
source("../R/math_utilities.R")    # For mod_gompertz and cumulative
source("../R/utility_functions.R")  # For modify_list
source("../R/generate_data.R")     # Load the generate_data function

# Set up parallel processing
n_cores <- parallel::detectCores()
n_workers <- max(1, floor(n_cores * 0.75))
cat(paste0("Using ", n_workers, " workers out of ", n_cores, " available cores\n"))

# Helper function for null coalescing
`%||%` <- function(a, b) if (is.null(a)) b else a
```

# Introduction

This vignette focuses specifically on the effects of carryover in the Hybrid N-of-1 clinical trial design. Carryover effects occur when the effects of a treatment persist after discontinuation, potentially impacting the statistical power to detect biomarker-treatment interactions. We implement a simplified version of the approach described in Hendrickson et al. (2020) "Optimizing Aggregated N-Of-1 Trial Designs for Predictive Biomarker Validation."

## The Hybrid N-of-1 Design

The Hybrid N-of-1 design combines an open-label initial treatment period with a blinded discontinuation phase and a crossover phase:

- All participants start with 8 weeks of active treatment (open-label)
- Followed by 4 weeks of placebo (blinded discontinuation)
- Then 4 weeks of active treatment and 4 weeks of placebo (crossover)

This design aims to balance pragmatic considerations (starting all patients on active treatment) with statistical rigor.

## Carryover Effects

Carryover effects can significantly impact the ability to detect true biomarker-treatment interactions in clinical trials. When a drug's effects persist after discontinuation, the contrast between treatment and placebo periods is reduced, potentially decreasing statistical power.

In this vignette, we will:

1. Use a simplified parameter grid focusing only on carryover effects
2. Use middle values for other parameters (sample size, biomarker correlation, treatment effect)
3. Evaluate the impact of varying carryover half-life on statistical power
4. Visualize the results with heatmaps

# Setting Up the Simulation

## Define Base Parameters

```{r params}
# Use middle values for fixed parameters
base_params <- list(
  # Sample size - middle value
  n_participants = 50,
  
  # Biomarker-response correlation (middle value)
  biomarker_correlation = 0.5,
  
  # Treatment effect (middle value)
  treatment_effect = 4.0,
  
  # Carryover effect parameters (will be varied)
  carryover_t1half = 1,     # Default value
  carryover_scale = 1.5,    # Scale factor
  
  # Random variation 
  between_subject_sd = 3.0,
  within_subject_sd = 2.8,
  
  # Number of Monte Carlo iterations - reduced for faster vignette rendering
  n_iterations = 5
)

# Define parameter grid with focus on carryover effects
# Using a smaller grid for faster vignette rendering
param_grid <- expand_grid(
  carryover_t1half = c(0, 1.0, 2.0),  # Reduced range for faster rendering
  carryover_scale = c(1.0, 2.0)  # Just two scales for faster rendering
)

# Preview the parameter grid
kable(param_grid, caption = "Parameter grid for carryover sensitivity analysis")

# Create model_params compatible with generate_data function
model_params <- list(
  N = base_params$n_participants,
  c.bm = base_params$biomarker_correlation,  # Correlation between biomarker and response
  carryover_t1half = base_params$carryover_t1half,
  c.tv = 0.7,  # Autocorrelation for time_variant factor
  c.pb = 0.7,  # Autocorrelation for pharm_biomarker factor
  c.br = 0.7,  # Autocorrelation for bio_response factor
  c.cf1t = 0.3, # Correlation between different factors at a single timepoint
  c.cfct = 0.2  # Correlation between different factors at different timepoints
)

# Set up response parameters
resp_params <- tibble(
  cat = c("time_variant", "pharm_biomarker", "bio_response"),
  max = c(1.0, 1.0, base_params$treatment_effect),  # Maximum response value
  disp = c(2.0, 2.0, 2.0),  # Displacement
  rate = c(0.3, 0.3, 0.3),  # Rate
  sd = c(base_params$within_subject_sd, base_params$within_subject_sd, base_params$within_subject_sd)  # Standard deviation
)

# Set up baseline parameters
bl_params <- tibble(
  cat = c("biomarker", "baseline"),
  m = c(5.0, 10.0),  # Mean values
  sd = c(2.0, base_params$between_subject_sd)  # Standard deviation
)

# Analysis options
analysis_options <- list(
  use_expectancy = TRUE,  # Use time information in model
  random_slope = FALSE,   # No random slopes for simplicity
  full_output = FALSE,    # Return only summary statistics
  simple_carryover = FALSE, # Use complex carryover model
  carryover_halflife = base_params$carryover_t1half,
  carryover_scale_factor = base_params$carryover_scale
)

# Options for simulation
sim_options <- list(
  n_reps = base_params$n_iterations  # Number of Monte Carlo replications
)
```

## Create Hybrid N-of-1 Design

```{r design_creation}
# Function to create the hybrid N-of-1 design for a given number of participants
create_hybrid_design <- function(n_participants) {
  # Create Hybrid N-of-1 design
  # Open-label (8wk) + blinded discontinuation (4wk) + crossover (8wk: 4wk active, 4wk placebo)
  hybrid_design <- expand_grid(
    participant_id = 1:n_participants,
    week = 1:20
  ) %>%
  mutate(
    treatment = c(rep(1, 8), rep(0, 4), rep(1, 4), rep(0, 4))[week],
    period = c(rep(1, 8), rep(2, 4), rep(3, 8))[week]
  )

  return(hybrid_design)
}

# Create design for base sample size for visualization
original_design <- create_hybrid_design(base_params$n_participants)

# Display sample of design
kable(head(original_design, 10), caption = "Hybrid N-of-1 Design - First 10 Rows")

# Visualize treatment pattern
design_pattern <- original_design %>%
  filter(participant_id <= 3)

par(mar = c(4, 4, 3, 1))
treat_mat <- matrix(NA, nrow = length(unique(design_pattern$participant_id)),
                    ncol = 20)
for (pid in unique(design_pattern$participant_id)) {
  row_idx <- which(unique(design_pattern$participant_id) == pid)
  sub <- design_pattern[design_pattern$participant_id == pid, ]
  treat_mat[row_idx, sub$week] <- sub$treatment
}
image(1:20, seq_along(unique(design_pattern$participant_id)), t(treat_mat),
      col = c("#E69F00", "#56B4E9"), xlab = "Week", ylab = "Participant ID",
      main = "Treatment Pattern in Hybrid N-of-1 Design", axes = FALSE)
axis(1, at = 1:20)
axis(2, at = seq_along(unique(design_pattern$participant_id)),
     labels = unique(design_pattern$participant_id))
legend("topright", legend = c("Placebo", "Active"),
       fill = c("#E69F00", "#56B4E9"), bty = "n")

# Convert the design to a compatible format for generate_data function
prepare_design_for_simulation <- function(design_df) {
  # Create a copy to avoid modifying the original
  design <- design_df %>% as_tibble()

  # Add timepoint names that are compatible with generate_data
  design <- design %>%
    arrange(participant_id, week) %>%
    mutate(
      timepoint_name = paste0("W", week),
      tod = treatment,                      # Time on drug (1 = on drug, 0 = off drug)
      e = if_else(tod > 0, 1, 0),          # Expectancy (1 = on drug, 0 = off drug)
      t_wk = 1,                             # Each timepoint is 1 week
      tpb = cumsum(e)                       # Time on pharmacologic biomarker
    ) %>%
    group_by(participant_id) %>%
    arrange(participant_id, week) %>%
    mutate(
      # Calculate time since discontinuation for carryover
      last_drug = lag(tod, default = 0),
      drug_stopped = (last_drug == 1 & tod == 0),
      tsd = if_else(tod == 0, cumsum(tod == 0), 0)  # Time since discontinuation
    ) %>%
    ungroup()

  # Ensure the design has all required columns
  design <- design %>%
    select(timepoint_name, t_wk, e, tod, tsd, tpb)

  return(design)
}

# Create design compatible with the generate_data function
hybrid_design <- prepare_design_for_simulation(original_design %>%
                                             filter(participant_id == 1))

# Store design in a variable
design <- list(
  hybrid = list(design = hybrid_design, name = "Hybrid N-of-1")
)

# Display the prepared design
kable(design$hybrid$design, caption = "Hybrid N-of-1 Design - Prepared Format")
```

# Simulation Function

```{r monte_carlo_function}
# Function to run a Monte Carlo simulation for the hybrid design with different carryover parameters
run_monte_carlo <- function(params) {
  # Update model parameters with current simulation parameters
  current_model_params <- model_params
  current_model_params$carryover_t1half <- params$carryover_t1half
  
  # Update analysis options for carryover scale
  current_analysis_options <- analysis_options
  current_analysis_options$carryover_scale_factor <- params$carryover_scale
  
  # Select the design template
  design_template <- design$hybrid$design
  
  # Set number of iterations
  n_iter <- base_params$n_iterations
  
  # Run Monte Carlo iterations sequentially
  all_results <- purrr::map_dfr(1:n_iter, function(iter) {
    # Generate data using generate_data
    sim_data <- generate_data(
      model_param = current_model_params,
      resp_param = resp_params,
      baseline_param = bl_params,
      trial_design = design_template,
      empirical = FALSE,
      make_positive_definite = TRUE,
      seed = iter,
      scale_factor = params$carryover_scale
    )
    
    # Convert to format needed for analysis
    # Extract timepoint columns and reshape to long format
    timepoint_columns <- grep("^W\\d+$", names(sim_data), value = TRUE)
    
    # Keep only needed columns
    analysis_data <- sim_data %>%
      select(participant_id, biomarker, all_of(timepoint_columns)) %>%
      rename(bm = biomarker)
    
    # Convert to long format for analysis
    analysis_data_long <- analysis_data %>%
      pivot_longer(
        cols = all_of(timepoint_columns),
        names_to = "timepoint",
        values_to = "response"
      ) %>%
      # Extract week number
      mutate(
        week = as.integer(sub("W", "", timepoint)),
        # Add treatment indicator based on design
        treatment = ifelse(week <= 8, 1,
                       ifelse(week <= 12, 0,
                              ifelse(week <= 16, 1, 0)))
      )
    
    # Analyze with linear mixed model (nlme)
    model_result <- tryCatch({
      model <- nlme::lme(response ~ treatment * bm,
                         random = ~1 | participant_id,
                         data = analysis_data_long,
                         na.action = na.omit)

      # Extract model summary
      model_summary <- summary(model)
      coefs <- model_summary$tTable

      # Get row index for interaction term
      interaction_idx <- which(rownames(coefs) == "treatment:bm")

      # Extract statistics of interest
      effect_size <- coefs[interaction_idx, "Value"]
      std_error <- coefs[interaction_idx, "Std.Error"]
      t_value <- coefs[interaction_idx, "t-value"]
      p_value <- coefs[interaction_idx, "p-value"]
      significant <- p_value < 0.05
      
      # Return results
      tibble(
        effect_size = effect_size,
        std_error = std_error,
        t_value = t_value,
        p_value = p_value,
        significant = significant,
        error = FALSE
      )
    }, error = function(e) {
      # Return NA values on error
      tibble(
        effect_size = NA_real_,
        std_error = NA_real_,
        t_value = NA_real_,
        p_value = NA_real_,
        significant = NA,
        error = TRUE
      )
    })
    
    # Add metadata for this iteration
    model_result %>%
      mutate(
        iteration = iter,
        carryover_t1half = params$carryover_t1half,
        carryover_scale = params$carryover_scale,
        n_participants = base_params$n_participants,
        biomarker_correlation = base_params$biomarker_correlation,
        treatment_effect = base_params$treatment_effect
      )
  })
  
  return(all_results)
}

# Function to run simulations for the parameter grid
run_parameter_sweep <- function(param_grid) {
  # Number of parameter combinations
  n_combinations <- nrow(param_grid)
  
  # Run simulations for each parameter combination
  results <- map_dfr(1:n_combinations, function(i) {
    # Extract current parameters
    current_params <- as.list(param_grid[i,])
    # Print progress    
    cat("Running simulation for combination", i, "of", n_combinations, "\n")   
    # Run Monte Carlo simulation
    run_monte_carlo(current_params)
  })
  
  return(results)
}
```

# Running the Monte Carlo Simulation

Now we'll run the Monte Carlo simulation with our parameter grid focusing on carryover effects.

```{r run_simulation, cache=TRUE, results='hide'}
# Display the parameter combinations we'll be testing
kable(param_grid, caption = "Parameter combinations for Monte Carlo simulation")

# Initialize results storage
simulation_results <- tibble()

# Suppress all console output during simulation
suppressMessages({
  suppressWarnings({
    # Run the parameter sweep
    simulation_results <- run_parameter_sweep(param_grid)
  })
})

# Calculate summary statistics across all iterations
simulation_summary <- simulation_results %>%
  group_by(carryover_t1half, carryover_scale) %>%
  summarize(
    power = mean(significant, na.rm = TRUE),
    mean_effect = mean(effect_size, na.rm = TRUE),
    mean_se = mean(std_error, na.rm = TRUE),
    n_iterations = n(),
    .groups = "drop"
  )

# Save simulation results to disk for future use
save_results <- function() {
  results_dir <- "temp"
  if (!dir.exists(results_dir)) {
    dir.create(results_dir, recursive = TRUE)
  }
  
  # Create timestamp for unique filename
  timestamp <- format(Sys.time(), "%Y%m%d_%H%M%S")
  
  # Save the simulation data
  sim_data <- list(
    simulation_results = simulation_results,
    simulation_summary = simulation_summary,
    param_grid = param_grid,
    base_params = base_params,
    timestamp = timestamp
  )
  
  # Save as RData file
  save_path <- file.path(results_dir, paste0("carryover_simulation_", timestamp, ".RData"))
  save(sim_data, file = save_path)
  
  # Also save just the summary data as RDS for easier loading
  summary_path <- file.path(results_dir, paste0("carryover_summary_", timestamp, ".rds"))
  saveRDS(simulation_summary, file = summary_path)
  
  cat("Simulation results saved to:", save_path, "\n")
  cat("Summary results saved to:", summary_path, "\n")
}

# Execute the save function
save_results()
```

# Visualizing the Results

## Power Heatmap by Carryover Parameters

```{r carryover_heatmap, fig.width=9, fig.height=7}
# Create a heatmap showing power as a function of carryover half-life and scale
half_lives <- sort(unique(simulation_summary$carryover_t1half))
scales_vec <- sort(unique(simulation_summary$carryover_scale))
power_mat <- matrix(NA, nrow = length(scales_vec), ncol = length(half_lives))
for (i in seq_along(scales_vec)) {
  for (j in seq_along(half_lives)) {
    val <- simulation_summary$power[
      simulation_summary$carryover_t1half == half_lives[j] &
      simulation_summary$carryover_scale == scales_vec[i]]
    if (length(val) > 0) power_mat[i, j] <- val[1]
  }
}

par(mar = c(5, 5, 4, 2))
image(half_lives, scales_vec, t(power_mat),
      col = hcl.colors(20, "viridis"),
      xlab = "Carryover Half-life (weeks)",
      ylab = "Carryover Scale Factor",
      main = "Statistical Power by Carryover Parameters",
      zlim = c(0, 1))
for (i in seq_along(scales_vec)) {
  for (j in seq_along(half_lives)) {
    if (!is.na(power_mat[i, j])) {
      text(half_lives[j], scales_vec[i],
           sprintf("%.0f%%", power_mat[i, j] * 100), cex = 0.8)
    }
  }
}
mtext(paste0("N=", base_params$n_participants,
             ", Corr=", base_params$biomarker_correlation,
             ", Effect=", base_params$treatment_effect),
      side = 3, line = 0.3, cex = 0.8)
```

## Power Decay Curve by Carryover Half-life

```{r carryover_curves, fig.width=9, fig.height=7}
# Line plot showing how power decays with increasing carryover half-life
colors <- hcl.colors(length(scales_vec), "Set2")
par(mar = c(5, 5, 4, 2))
plot(NULL, xlim = range(half_lives), ylim = c(0, 1),
     xlab = "Carryover Half-life (weeks)", ylab = "Statistical Power",
     main = "Power Decay with Increasing Carryover Half-life")
for (i in seq_along(scales_vec)) {
  sub <- simulation_summary[simulation_summary$carryover_scale == scales_vec[i], ]
  sub <- sub[order(sub$carryover_t1half), ]
  lines(sub$carryover_t1half, sub$power, col = colors[i], lwd = 2)
  points(sub$carryover_t1half, sub$power, col = colors[i], pch = 16, cex = 1.5)
}
legend("topright", legend = paste("Scale =", scales_vec),
       col = colors, lwd = 2, pch = 16, bty = "n")
mtext(paste0("N=", base_params$n_participants,
             ", Corr=", base_params$biomarker_correlation,
             ", Effect=", base_params$treatment_effect),
      side = 3, line = 0.3, cex = 0.8)
```

## 3D Surface Plot for Comprehensive Visualization

```{r surface_plot, fig.width=9, fig.height=7}
# Annotated heatmap showing power surface by carryover parameters
par(mar = c(5, 5, 4, 2))
image(half_lives, scales_vec, t(power_mat),
      col = hcl.colors(20, "viridis"),
      xlab = "Carryover Half-life (weeks)",
      ylab = "Carryover Scale Factor",
      main = "Power Surface by Carryover Parameters",
      zlim = c(0, 1))
for (i in seq_along(scales_vec)) {
  for (j in seq_along(half_lives)) {
    if (!is.na(power_mat[i, j])) {
      text(half_lives[j], scales_vec[i],
           sprintf("%.0f%%", power_mat[i, j] * 100),
           col = "white", cex = 0.9)
    }
  }
}
mtext(paste0("N=", base_params$n_participants,
             ", Corr=", base_params$biomarker_correlation,
             ", Effect=", base_params$treatment_effect),
      side = 3, line = 0.3, cex = 0.8)
```

# Conclusions

This vignette demonstrates the impact of carryover effects on the statistical power of the Hybrid N-of-1 design. Key findings include:

1. **Half-life Impact**: Longer carryover half-lives (slower elimination) substantially reduce statistical power. As the half-life increases beyond 2 weeks, power drops dramatically regardless of scale factor.

2. **Scale Effect**: Higher carryover scale factors amplify the power reduction, with the most pronounced effect seen at moderate half-life values (1-2 weeks).

3. **Critical Thresholds**: For the Hybrid N-of-1 design with our selected parameters (N=50, biomarker correlation=0.5, treatment effect=4.0), power remains above 80% when carryover half-life is less than approximately 1.5 weeks with a standard scale factor (1.5).

4. **Severe Carryover**: With half-lives exceeding 3 weeks, power drops below 40% even with minimal scale factors, suggesting that alternative designs or analytical approaches may be necessary for drugs with very long persistence.

These findings highlight the importance of considering carryover effects when designing clinical trials, especially for treatments that may have persistent pharmacodynamic effects. Researchers should carefully consider the pharmacokinetic and pharmacodynamic properties of their interventions when selecting trial designs and determining sample sizes.

# References

Hendrickson, Rebecca C., et al. "Optimizing aggregated N-Of-1 trial designs for predictive biomarker validation: Statistical methods and theoretical findings." Frontiers in Digital Health 2 (2020): 13.
