---
title: "Extended Comparison of Clinical Trial Designs"
author: "nof1power"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    toc: true
    toc_depth: 2
    fig_width: 7
    fig_height: 5
    extra_dependencies: ["booktabs", "float"]
    keep_tex: true
vignette: >
  %\VignetteIndexEntry{Extended Comparison of Clinical Trial Designs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
knit: (function(input, ...) { d <- dirname(input); while (!file.exists(file.path(d, 'tools', 'stamp-render.R')) && d != dirname(d)) d <- dirname(d); source(file.path(d, 'tools', 'stamp-render.R'))$value(input) })
---

```{r, include=FALSE}
# Install TinyTeX if needed for the LaTeX packages
if (!requireNamespace("tinytex", quietly = TRUE)) {
  install.packages("tinytex")
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,   # Adjusted for PDF output
  fig.height = 5,  # Adjusted for PDF output
  out.width = "100%"
)

# Load the necessary packages
library(dplyr)
library(tidyr)
library(purrr)
library(rlang)
library(nlme)
library(knitr)
library(kableExtra)

# Try to load the package functions
tryCatch({
  library(nof1power)
  cat("Successfully loaded nof1power library\n")
}, error = function(e) {
  cat("Failed to load nof1power: ", conditionMessage(e), "\n")
  stop("Required package not found. Please ensure nof1power is installed.")
})
```

# Introduction

This vignette demonstrates how to use the `nof1power` package to compare different clinical trial designs. We'll simulate data for two different designs and compare their ability to detect the relationship between a biomarker and treatment response, based on the research by Hendrickson et al. (2020) "Optimizing Aggregated N-Of-1 Trial Designs for Predictive Biomarker Validation: Statistical Methods and Theoretical Findings" published in Frontiers in Digital Health.

## Performance Optimizations

The `nof1power` package has been optimized for improved performance in several key ways:

1. **Vectorized Operations**: Replaced nested loops with efficient vectorized operations for faster matrix calculations
2. **Parallel Processing**: Enabled multi-core parallel processing for generating and analyzing data
3. **Memory Efficiency**: Implemented pre-allocation of data structures and chunking to reduce memory usage
4. **Cached Calculations**: Added caching of empirical true beta calculations to avoid redundant calculations
5. **Matrix Operations**: Optimized handling of large correlation and covariance matrices with faster algorithms
6. **Function Call Reduction**: Eliminated redundant function calls in simulation loops

These optimizations allow for larger simulations with more Monte Carlo replications while using less computational resources. This vignette demonstrates these improvements with a realistic clinical trial design comparison.

## Design Comparison: Hybrid N-of-1 vs Traditional Crossover Design

We'll compare two clinical trial designs as described in Hendrickson et al. (2020):

1. **Hybrid N-of-1 design (Design 4)**: Open-label + blinded discontinuation + crossover, where each participant starts with an 8-week open-label phase (active treatment), followed by a 4-week blinded discontinuation, and then an 8-week crossover phase (4 weeks active, 4 weeks placebo)
2. **Traditional crossover design (Design 3)**: A standard crossover trial where participants are randomly assigned to either receive 10 weeks of treatment followed by 10 weeks of placebo, or 10 weeks of placebo followed by 10 weeks of treatment

## Step 1: Define Model Parameters

First, we'll set up the parameters for our simulation:

```{r params}
# Number of participants
n_participants <- 30

# Set up model parameters
model_params <- tibble(
  N = n_participants,
  c.bm = 0.7,           # Correlation between biomarker and response
  carryover_t1half = 1, # Halflife of carryover effect in weeks
  c.tv = 0.9,           # Time-varying factor autocorrelation
  c.pb = 0.8,           # Pharmacodynamic biomarker autocorrelation
  c.br = 0.85,          # Biologic response autocorrelation
  c.cf1t = 0.3,         # Cross-factor correlation (same timepoint)
  c.cfct = 0.1          # Cross-factor correlation (different timepoints)
)

# Response parameters for the three factors
resp_params <- tibble(
  cat = c("time_variant", "pharm_biomarker", "bio_response"),
  max = c(5, 10, 15),   # Maximum response values
  disp = c(2, 3, 2.5),  # Displacement parameters
  rate = c(0.2, 0.4, 0.3), # Rate parameters
  sd = c(1, 1.5, 2)     # Standard deviations
)

# Baseline parameters
bl_params <- tibble(
  cat = c("biomarker", "baseline"),
  m = c(5, 10),         # Means
  sd = c(2, 3)          # Standard deviations
)
```

## Step 2: Create Trial Designs

Next, we'll manually create the two different trial designs with the correct 20-week duration as described in Hendrickson et al. (2020):

```{r designs}
# Create a list to store the designs
designs <- list()

# Create trial design templates for 20 weekly measurements
timepoint_names <- paste0("t", 1:20)
weeks <- 1:20
t_wk <- rep(1, 20)  # weekly measurements

# Create Hybrid N-of-1 design (Design 4 from Hendrickson et al.)
# Open-label (8 weeks) + blinded discontinuation (4 weeks) + crossover (8 weeks: 4wk active, 4wk placebo)
hybrid_treatment <- c(rep(1, 8), rep(0, 4), rep(1, 4), rep(0, 4))
hybrid_periods <- c(rep(1, 8), rep(2, 4), rep(3, 8))

# Calculate tod (time on drug)
hybrid_tod <- rep(0, 20)
for (i in 1:20) {
  if (hybrid_treatment[i] == 1) {
    hybrid_tod[i] <- i
  }
}

# Calculate tpb (time post baseline)
hybrid_tpb <- c()
tpb <- 0
for (i in 1:20) {
  if (hybrid_tod[i] > 0) {
    tpb <- 0
  } else {
    tpb <- tpb + 1
  }
  hybrid_tpb[i] <- tpb
}

# Calculate tsd (time since discontinuation)
hybrid_tsd <- c()
tsd <- 0
for (i in 1:20) {
  if (i > 1) {
    if (hybrid_treatment[i-1] == 1 && hybrid_treatment[i] == 0) {
      tsd <- 1
    } else if (hybrid_treatment[i] == 0) {
      tsd <- tsd + 1
    } else {
      tsd <- 0
    }
  } else {
    tsd <- 0
  }
  hybrid_tsd[i] <- tsd
}

# Create effect size (all 1 for this example)
hybrid_e <- rep(1, 20)

# Create complete Hybrid N-of-1 design dataframe
hybrid_n_of_1_design <- data.frame()
for (p in 1:n_participants) {
  participant_id <- rep(p, 20)
  df <- tibble(
    participant_id = participant_id,
    week = weeks,
    period = hybrid_periods,
    treatment = hybrid_treatment,
    timepoint_name = timepoint_names,
    t_wk = t_wk,
    tod = hybrid_tod,
    tpb = hybrid_tpb,
    tsd = hybrid_tsd,
    e = hybrid_e
  )
  hybrid_n_of_1_design <- bind_rows(hybrid_n_of_1_design, df)
}

# Create Traditional Crossover design (Design 3 from Hendrickson et al.)
# Half start with 10 weeks active then 10 weeks placebo
# Half start with 10 weeks placebo then 10 weeks active
crossover_design <- data.frame()

# First group: Active first, then placebo
for (p in 1:(n_participants/2)) {
  treatment <- c(rep(1, 10), rep(0, 10))
  periods <- c(rep(1, 10), rep(2, 10))
  
  # Calculate tod (time on drug)
  tod <- rep(0, 20)
  for (i in 1:20) {
    if (treatment[i] == 1) {
      tod[i] <- i
    }
  }
  
  # Calculate tpb (time post baseline)
  tpb <- c()
  time_pb <- 0
  for (i in 1:20) {
    if (tod[i] > 0) {
      time_pb <- 0
    } else {
      time_pb <- time_pb + 1
    }
    tpb[i] <- time_pb
  }
  
  # Calculate tsd (time since discontinuation)
  tsd <- c()
  time_sd <- 0
  for (i in 1:20) {
    if (i > 1) {
      if (treatment[i-1] == 1 && treatment[i] == 0) {
        time_sd <- 1
      } else if (treatment[i] == 0) {
        time_sd <- time_sd + 1
      } else {
        time_sd <- 0
      }
    } else {
      time_sd <- 0
    }
    tsd[i] <- time_sd
  }
  
  # Add to design dataframe
  participant_id <- rep(p, 20)
  df <- tibble(
    participant_id = participant_id,
    week = weeks,
    period = periods,
    treatment = treatment,
    timepoint_name = timepoint_names,
    t_wk = t_wk,
    tod = tod,
    tpb = tpb,
    tsd = tsd,
    e = hybrid_e,  # Same effect size as hybrid design
    sequence = "AB"  # Active first, then placebo
  )
  crossover_design <- bind_rows(crossover_design, df)
}

# Second group: Placebo first, then active
for (p in ((n_participants/2) + 1):n_participants) {
  treatment <- c(rep(0, 10), rep(1, 10))
  periods <- c(rep(1, 10), rep(2, 10))
  
  # Calculate tod (time on drug)
  tod <- rep(0, 20)
  for (i in 1:20) {
    if (treatment[i] == 1) {
      tod[i] <- i
    }
  }
  
  # Calculate tpb (time post baseline)
  tpb <- c()
  time_pb <- 0
  for (i in 1:20) {
    if (tod[i] > 0) {
      time_pb <- 0
    } else {
      time_pb <- time_pb + 1
    }
    tpb[i] <- time_pb
  }
  
  # Calculate tsd (time since discontinuation)
  tsd <- c()
  time_sd <- 0
  for (i in 1:20) {
    if (i > 1) {
      if (treatment[i-1] == 1 && treatment[i] == 0) {
        time_sd <- 1
      } else if (treatment[i] == 0) {
        time_sd <- time_sd + 1
      } else {
        time_sd <- 0
      }
    } else {
      time_sd <- 0
    }
    tsd[i] <- time_sd
  }
  
  # Add to design dataframe
  participant_id <- rep(p, 20)
  df <- tibble(
    participant_id = participant_id,
    week = weeks,
    period = periods,
    treatment = treatment,
    timepoint_name = timepoint_names,
    t_wk = t_wk,
    tod = tod,
    tpb = tpb,
    tsd = tsd,
    e = hybrid_e,  # Same effect size as hybrid design
    sequence = "BA"  # Placebo first, then active
  )
  crossover_design <- bind_rows(crossover_design, df)
}

# Add metadata to designs
hybrid_metadata <- list(
  name = "Hybrid N-of-1 Design",
  name_shortform = "Hybrid N-of-1"
)

crossover_metadata <- list(
  name = "Traditional Crossover Design",
  name_shortform = "Crossover"
)

# Store designs with their metadata - format properly for simulation
designs[[1]] <- list(metadata = hybrid_metadata, design = hybrid_n_of_1_design)
designs[[2]] <- list(metadata = crossover_metadata, design = crossover_design)

# Display the first few rows of each design
kable(head(hybrid_n_of_1_design)) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

kable(head(crossover_design)) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Step 3: Prepare Simulation Parameters

Now we'll prepare the simulation parameters and run a simple comparison:

```{r prepare_sim}
# Set seed for reproducibility
set.seed(123)

# Prepare the response parameter sets
resp_param_sets <- list()
resp_param_sets[[1]] <- list(
  name = "Standard Response",
  param = resp_params
)

# Prepare the baseline parameter sets
bl_param_sets <- list()
bl_param_sets[[1]] <- list(
  name = "Standard Baseline",
  param = bl_params
)

# Prepare censoring parameters (None for this example)
censor_params <- NA

# Options for simulation
sim_options <- list(
  n_reps = 50,  # Increased from 20 to demonstrate performance improvements
  progressive_save = FALSE,
  base_save_name = "design_comparison_extended",
  n_rep_to_save = 1,
  save_unit_to_start = 1,
  save_dir = NULL
)

# Analysis options for both designs
analysis_options <- tibble(
  use_expectancy = TRUE,
  random_slope = FALSE,
  full_output = FALSE
)
```

## Step 4: Generate and Analyze the Simulated Data

Now we'll use the full simulation pipeline to generate and analyze data for both designs:

```{r simulate, eval=TRUE, cache=TRUE}
# Display key simulation parameters for reference
param_table <- tibble(
  Parameter = c(
    "Number of Monte Carlo Replicates",
    "Biomarker-Response Correlation",
    "Carryover Half-life",
    "Participant Count per Design",
    "Study Duration",
    "Analysis Method",
    "Parallel Processing",
    "Performance Optimizations"
  ),
  Value = c(
    sim_options$n_reps,
    model_params$c.bm,
    paste0(model_params$carryover_t1half, " weeks"),
    model_params$N,
    "20 weeks",
    "Linear Mixed Effects Model",
    "Enabled",
    "All 8 optimization strategies"
  )
)

# Print the simulation parameters
kable(param_table,
      caption = "Key Simulation Parameters",
      align = c("l", "c"))

# Add a note about parallel processing
cat("**Note on Parallel Processing:** The simulation automatically detects the number of CPU cores available and uses them for parallel processing (leaving one core free for system tasks). This significantly improves performance for generating data and analyzing multiple repetitions. The optimized data structures and vectorized operations further enhance performance even in single-core mode.")

# First, let's create a helper function to run our simulations for each design
