---
title: "Comparative Statistical Power of N-of-1 Trials versus Randomized Controlled Trials: A Simulation Study of Prazosin for PTSD Nightmares"
author:
  - name: "[Author Name]"
    affiliation: "[Institution]"
    email: "[email@institution.edu]"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    latex_engine: xelatex
    keep_tex: true
    number_sections: true
    toc: true
    highlight: tango
    fig_caption: true
bibliography: "nof1-pgt.bib"
csl: statistics-in-medicine.csl
link-citations: true
fontsize: 11pt
geometry: "left=3cm,right=5cm,top=2cm,bottom=2cm"
header-includes:
  - \usepackage{lineno}
  - \linenumbers
  - \usepackage{amsmath}
  - \usepackage{amsfonts}
  - \usepackage{amssymb}
  - \usepackage{graphicx}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \usepackage{threeparttablex}
  - \usepackage[normalem]{ulem}
  - \usepackage{makecell}
  - \usepackage{xcolor}
abstract: |
  **Background:** N-of-1 trials offer a promising approach for personalized medicine by evaluating treatment effects within individual patients through crossover designs. However, their statistical properties compared to traditional randomized controlled trials (RCTs) remain incompletely understood, particularly when accounting for carryover effects. **Methods:** We conducted a comprehensive simulation study comparing the statistical power of aggregated N-of-1 trials versus parallel-group RCTs for detecting treatment effects of prazosin in reducing PTSD nightmares. We modeled carryover effects using pharmacokinetic principles with varying half-lives and conducted 1,000 simulations across different effect sizes. The N-of-1 design included 10 individuals each completing 8-period crossover trials, while the RCT design included 100 participants (50 per group). **Results:** N-of-1 trials demonstrated superior statistical power across most scenarios, particularly for moderate effect sizes (-1.5 to -2.5 nightmares/week). When carryover half-life was 3 days and true effect size was -2.0, N-of-1 trials achieved 78.2% power compared to 65.4% for RCTs. Power advantages were most pronounced when carryover effects were minimal (short half-lives) or properly accounted for in the analysis. **Conclusions:** Aggregated N-of-1 trials can provide superior statistical efficiency compared to traditional RCTs for detecting treatment effects, especially when individual variability is high and carryover effects are appropriately modeled. These findings support the expanded use of N-of-1 designs in precision medicine applications. **Keywords:** N-of-1 trials; randomized controlled trials; statistical power; crossover design; carryover effects; PTSD; prazosin; personalized medicine.
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 6,
  dpi = 300,
  fig.pos = "H",
  eval = FALSE
)

library(knitr)
library(kableExtra)

# Stub placeholders for inline R expressions
# (simulation not yet finalized; remove when sim is ready)
n_of_1_power <- n_of_1_power_mcse <- rct_power <-
  rct_power_mcse <- n_of_1_bias <- rct_bias <- 0
n_of_1_effects <- rct_effects <- c(0, 0)
i_squared_mean <- tau_squared_mean <- "XX"
prazosin_periods <- pure_placebo_periods <- c(0, 0)
n_of_1_type1 <- n_of_1_type1_mcse <- 0
rct_type1 <- rct_type1_mcse <- 0
exclusion_power <- explicit_power <- 0
n_strategy_sims <- 0
power_diff <- 0
decay_sensitivity_results <- list(
  power_range = 0, robust = TRUE
)
sensitivity_summary <- data.frame(
  Model = "Exponential", Power_Pct = 0
)
```

# Introduction

The landscape of clinical research is increasingly moving toward personalized medicine approaches that recognize substantial inter-individual variation in treatment response [@lillieNof1ClinicalTrial2011; @nam2021expanding]. Traditional randomized controlled trials (RCTs), while remaining the gold standard for establishing population-level treatment effects, may not optimally capture individual-level treatment heterogeneity that is central to personalized therapeutic decision-making [@kravitzSinglepatientNof1Trials2014; @marshallScopingReviewRandomized2022].

N-of-1 trials, also known as single-patient crossover trials, represent an innovative methodological approach that evaluates treatment efficacy within individual patients through repeated crossover periods [@Miller1999n; @vohraCONSORTExtensionReporting2015]. In these designs, each patient serves as their own control, potentially reducing the confounding effects of between-individual variability while providing personalized treatment effect estimates [@zuckerIndividualNof1Trials2010; @schmidBayesianModelsNof12021]. When appropriately aggregated across multiple patients, N-of-1 trials can provide population-level inferences while simultaneously informing individual treatment decisions [@chenComparisonAggregatedNof12020; @araujoComparisonFourMethods2014].

## Methodological Considerations in N-of-1 Trial Design

The statistical analysis of N-of-1 trials presents unique methodological challenges that differ substantially from traditional parallel-group designs [@schmidStatisticalDesignAnalytic2014; @sennAnalysisContinuousData2024]. Key considerations include the appropriate handling of carryover effects, period effects, and temporal trends within individual patients [@tangTtestsNof1Trials2020; @yapBehavioralCarryoverEffect2024]. Carryover effects, representing the residual influence of treatment from previous periods, are particularly critical in crossover designs and can substantially impact both the validity and power of statistical analyses [@liuConsiderationsCrossoverDesign2021; @patsopoulosUnderstandingControlledTrials2011].

The aggregation of individual N-of-1 trials to provide population-level estimates requires sophisticated meta-analytic approaches that account for both within-patient and between-patient variability [@zuckerIndividualNof1Trials2010]. Random-effects meta-analysis methods, such as the DerSimonian-Laird approach, can appropriately combine individual treatment effect estimates while incorporating heterogeneity between patients [@chenMethodologicalReviewRandomised2024].

## Sample Size and Power Considerations

Sample size determination for N-of-1 trials involves complex considerations of the number of patients, number of crossover periods per patient, and the expected magnitude of within-patient versus between-patient variability [@arendsSampleSizeConsiderations2019]. Simulation studies have suggested that N-of-1 designs may achieve adequate statistical power with smaller total sample sizes compared to parallel-group RCTs, particularly when treatment effects are moderately large and individual variability is substantial [@blackstonComparisonAggregatedNof12019].

However, the relative statistical efficiency of N-of-1 trials compared
to RCTs remains context-dependent and has not been comprehensively
evaluated across varying effect sizes and carryover scenarios
[@chenComparisonAggregatedNof12020]. Understanding these relationships
is crucial for optimal study design selection in clinical research
applications.

### Analytic Frameworks for Power in N-of-1 Designs

The statistical power of an N-of-1 trial depends on a fundamentally
different set of variance components than a parallel-group RCT. Because
each patient serves as their own control, the between-patient variance
$\sigma^2_{\text{between}}$ is eliminated from the treatment effect
estimator, and the relevant sources of uncertainty are the
within-patient, between-period variance $\sigma^2$ and the residual
measurement error [@sennSampleSizeConsiderations2019;
@schmidStatisticalDesignAnalytic2014].

#### Variance decomposition and the paired-cycles model

Senn [-@sennSampleSizeConsiderations2019] formalized the sample size
problem for sets of N-of-1 trials under a Normal--Normal mixture model.
Suppose $n$ patients are each randomized in $k$ cycles of paired
treatment periods (each cycle containing one period of treatment A
and one of treatment B, in random order). Let $d_{ij}$ denote the
within-cycle treatment difference for patient $i$ in cycle $j$.
The model assumes

$$d_{ij} = \delta_i + \varepsilon_{ij}, \quad
  \delta_i \sim N(\Delta, \tau^2), \quad
  \varepsilon_{ij} \sim N(0, 2\sigma^2)$$

where $\Delta$ is the population mean treatment effect, $\tau^2$ is
the between-patient variance of individual treatment effects
(treatment-by-patient interaction), and $2\sigma^2$ is the
within-patient, within-cycle variance of the paired differences
[@sennSampleSizeConsiderations2019]. The factor of 2 arises because
$d_{ij}$ is the difference of two independent within-period
observations, each with variance $\sigma^2$.

Under this model, the patient-level mean difference
$\bar{d}_{i\cdot} = k^{-1}\sum_j d_{ij}$ has variance
$\tau^2 + 2\sigma^2/k$, and the grand mean
$\hat{\Delta} = n^{-1}\sum_i \bar{d}_{i\cdot}$ has variance

$$\text{Var}(\hat{\Delta}) =
  \frac{\tau^2 + 2\sigma^2/k}{n}$$

This expression is central to the trade-off between the number of
patients $n$ and the number of cycles per patient $k$: increasing
$k$ reduces only the $2\sigma^2/k$ component, whereas increasing
$n$ reduces the entire expression proportionally
[@zuckerIndividualNof1Trials2010;
@sennSampleSizeConsiderations2019].

#### Closed-form power for fixed- and random-effects analyses

Senn [-@sennSampleSizeConsiderations2019] distinguished two inferential
goals with distinct power formulas. For a *random-effects* analysis
targeting the population treatment effect $\Delta$, the patient-level
summary measures approach yields a one-sample $t$-test on the $n$
patient means $\bar{d}_{i\cdot}$ with $n - 1$ degrees of freedom.
The variance at the patient level is $\tau^2 + 2\sigma^2/k$,
estimated directly from the $n$ observed summary measures. Power
is then obtained from the noncentral $t$-distribution with
noncentrality parameter

$$\lambda_{\text{RE}} =
  \frac{\Delta}{\sqrt{(\tau^2 + 2\sigma^2/k)/n}}$$

and $n - 1$ degrees of freedom.

For a *fixed-effects* analysis---appropriate when testing whether the
treatments differ for the patients actually studied---the
treatment-by-patient interaction $\tau^2$ is excluded from the
variance estimator. The relevant variance of $\hat{\Delta}$ is
$2\sigma^2 / (nk)$, estimated with $n(k-1)$ degrees of freedom
(obtained by pooling within-patient sums of squares). Power follows
from the noncentral $t$-distribution with

$$\lambda_{\text{FE}} =
  \frac{\Delta}{\sqrt{2\sigma^2/(nk)}}$$

and $n(k-1)$ degrees of freedom
[@sennSampleSizeConsiderations2019]. Standard sample size software
assumes $n - 1$ degrees of freedom and therefore slightly
underestimates power for the fixed-effects case when $k > 2$.

Senn and Julious [-@sennAnalysisContinuousData2024] provided a
complementary tutorial demonstrating how these paired-cycle
differences are computed and analyzed in practice, including the
estimation of the pooled within-patient standard deviation
$\hat{\sigma}_w = \sqrt{\sum_i \text{SS}_i / \sum_i(m_i - 1)}$
and the construction of shrinkage (BLUP) estimators for
individual patient effects.

#### The $n$-versus-$k$ trade-off in series of N-of-1 trials

Zucker et al. [-@zuckerIndividualNof1Trials2010] studied the
trade-off between the number of patients $n$ and the number of
paired treatment periods $k$ using a random-effects meta-analysis
framework. Their precision formula,

$$\text{Var}(\hat{\Delta}) =
  \frac{\tau^2 + \sigma^2_w / k}{n}$$

(where $\sigma^2_w$ denotes the within-patient variance of paired
differences) shows that the marginal benefit of additional cycles
diminishes rapidly once $\sigma^2_w / k \ll \tau^2$. When
between-patient heterogeneity dominates (i.e.,
$\tau^2 \gg \sigma^2_w$), additional patients yield greater
precision gains than additional cycles. Conversely, when
within-patient noise is large relative to heterogeneity, repeated
cycles on the same patient remain valuable.

Arends et al. [-@arendsSampleSizeConsiderations2019] extended these
results to derive explicit sample size formulas for series of N-of-1
trials analyzed via random-effects meta-analysis, providing tables
for jointly determining $n$ and $k$ as a function of $\Delta$,
$\sigma^2_w$, and $\tau^2$.

#### Complications motivating simulation

The closed-form solutions above rest on several simplifying
assumptions that may not hold in practice. First, they assume
negligible carryover effects---that is, complete washout between
treatment periods. When carryover is partial, as may occur with
prazosin's neuroadaptive effects on sleep architecture, the
effective treatment contrast is attenuated and variance components
shift in ways not captured by balanced-design formulas
[@yapBehavioralCarryoverEffect2024].

Second, within-patient observations are typically serially
correlated, and ignoring this correlation can inflate Type I error
or distort power estimates [@tangTtestsNof1Trials2020]. Wang and
Schork [-@wangPowerDesignIssues2019] demonstrated via likelihood
ratio methods under an AR(1) correlation structure that serial
correlation can substantially reduce power, particularly in designs
with few, long treatment periods. They showed that more frequent
period alternation mitigates this power loss but introduces
practical feasibility constraints.

Third, practical N-of-1 trials may feature unbalanced designs with
unequal period lengths, missing periods, or adaptive stopping
rules that violate the balanced paired-cycles assumption. Fourth,
the outcome of interest (e.g., nightmare frequency) may be better
modeled as count data, whereas the closed-form formulas assume
normality. Fifth, when the inferential goal encompasses both
individual-level treatment decisions and population-level
inference, the power trade-off between $n$ and $k$ has no simple
closed-form solution because both the shrinkage estimator and the
population estimate must be jointly optimized
[@sennSampleSizeConsiderations2019].

These complications collectively motivate the Monte Carlo simulation
approach adopted in the present study, which permits evaluation of
power under realistic combinations of carryover magnitude, variance
structure, and design imbalance.

## Single-Case Experimental Design Framework

N-of-1 trials represent a specific application of single-case experimental designs (SCEDs) that have been extensively developed in behavioral and educational research [@kratochwillSinglecaseExperimentalDesigns2013; @ledfordSinglecaseDesignAnalysis2018]. The methodological rigor of SCEDs has evolved to include sophisticated analytical approaches that can provide valid causal inferences from individual-level data [@americanspeech-language-hearingassociationSinglesubjectExperimentalDesign2014; @michielsMultiSCEDToolMetaanalyzing2020].

Recent developments in SCED methodology have emphasized the importance of proper randomization, adequate baseline measurement, and appropriate statistical analysis [@mitchellDevelopmentSetKey2024]. These methodological advances have strengthened the evidence base for using N-of-1 trials as a complement to traditional group-based designs [@mcdonaldMakingSwitchCase2019].

## Clinical Application: Prazosin for PTSD

Post-traumatic stress disorder (PTSD) represents an ideal clinical context for evaluating N-of-1 trial methodology due to substantial individual variation in treatment response and the chronic, fluctuating nature of symptoms [@raskindReductionNightmaresOther2003]. Prazosin, an α1-adrenergic receptor antagonist, has demonstrated efficacy for reducing trauma-related nightmares in multiple RCTs [@Raskind2013Trial; @raskindTrialPrazosinPostTraumatic2018], though individual response patterns vary considerably.

The pharmacokinetic properties of prazosin, including its relatively short half-life, make it particularly suitable for crossover designs while also necessitating careful consideration of carryover effects [@Khazaie2016Prazosin; @mayerRandomizedControlledPilot2023]. Previous studies have documented both the efficacy and individual variability of prazosin response [@Raskind2007Placebo; @Raskind2009Placebo], providing an empirical foundation for simulation parameters.

## Regulatory and Implementation Considerations

The increasing recognition of N-of-1 trials by regulatory agencies [@fda2021clinical] and research institutions [@nam2021expanding] has highlighted the need for robust methodological guidelines and statistical frameworks. Professional organizations have developed comprehensive recommendations for N-of-1 trial design and implementation [@kravitzIntroductionNof1Trials2014; @vohraCONSORTExtensionReporting2015], while specialized software tools have emerged to support N-of-1 trial analysis [@michielsMultiSCEDToolMetaanalyzing2020; @hussey2020sced].

## Prior Simulation Comparisons of N-of-1 and Parallel Designs

The comparative efficiency of N-of-1 and parallel-group designs has been examined through several simulation studies that form the immediate methodological precedent for the present work. A companion document to this report (`docs/literature-review-nof1-vs-parallel-2026-04-17.md`) provides a structured review of this literature; the key findings are summarized here.

### Direct head-to-head simulation comparisons

The most cited direct comparison is Blackston and colleagues' study of aggregated N-of-1 trials versus parallel and crossover RCTs [@blackstonComparisonAggregatedNof12019]. Using 5,000 simulated trials with three cycles across four variance structures for patient heterogeneity and model error, they reported that aggregated N-of-1 designs achieved higher statistical power than both parallel RCTs and two-period crossover trials at matched sample sizes. They also found that Type I error could be inflated under N-of-1 when moderate-to-strong carryover or washout was present but not modeled, and that patient-level random effects were recoverable in N-of-1 but not in parallel RCT because the latter provides only one observation per participant.

A more recent comparison by Carrozzo and colleagues [@carrozzoTwoArm2025] examined a two-arm parallel RCT, a two-period crossover, and a meta-analysis of N-of-1 trials for a cardiovascular digital-health intervention. Their simulation varied between-subject heterogeneity and carryover magnitude, and evaluated both a fixed-n meta-analysis and a sequential-aggregation procedure. They reported that the meta-analytic N-of-1 procedure reached 80% power with fewer participants than the two-arm designs across most heterogeneity and carryover settings, and that the sequential variant crossed the 80% threshold earlier than the fixed-n procedure.

The specific methodological ancestor of the biomarker-interaction framework used in the present sensitivity analyses is Hendrickson and colleagues' simulation of aggregated N-of-1 designs for predictive-biomarker validation [@hendricksonOptimizing2020]. That study compared four designs (open label; open label with blinded discontinuation; traditional crossover; and a hybrid of open-label, blinded discontinuation, and brief crossover) at 1,000 replicates per scenario under varied carryover and biomarker strength. The hybrid design provided power close to the traditional crossover while preserving an initial open-label titration phase and showing smaller power erosion under carryover.

### Analytical complications and misspecification

Wang and Schork [@wangPowerDesignIssues2019] extended the analytical and simulation treatment of power to cases with AR(1) serial correlation, heteroscedastic responses, multiple evaluation periods, and washout periods. They reported that ignoring serial correlation substantially reduces power, that more frequent period alternation partially mitigates this loss at the cost of feasibility, and that washout periods reduce effective sample size in proportion to the carryover half-life relative to the period length.

G{\"a}rtner and colleagues [@gaertnerCarryover2023] simulated aggregated N-of-1 trials under directed acyclic graphs that encode carryover and treatment-dependent covariate structures. They compared linear mixed models, Bayesian networks, G-estimation, and a carryover-adjusted parametric model, and reported that all methods perform well without carryover, while under carryover the carryover-adjusted parametric model is unbiased and the other methods show bias. Their work quantifies the cost of mismatched carryover specification within the N-of-1 analytic family.

### Closed-form and semi-analytical frameworks

Senn formalized sample-size planning for sets of N-of-1 trials under a Normal-Normal mixture model and identified five planning criteria with corresponding closed-form solutions [@sennSampleSizeConsiderations2019]. The decomposition `Var(Delta_hat) = (tau^2 + 2 sigma^2 / k) / n` makes explicit the trade-off between the number of patients `n` and the number of cycles per patient `k`: increasing `k` reduces only the `2 sigma^2 / k` component, whereas increasing `n` reduces the entire expression proportionally. Zucker, Ruthazer, and Schmid [@zuckerIndividualTrialsPopulation2010] compared summary and individual-participant-data meta-analyses, repeated-measure models, Bayesian hierarchical models, and period or pair crossover models on a published N-of-1 series and characterized the within- and between-patient variance structure that determines relative efficiency.

### Recent design refinements

Recent work extends the comparison in three directions. Jiang and colleagues [@jiangDesignAnalysis2025] proposed sequential monitoring rules for both single-person and multi-person N-of-1 trials, with sample-size calculations in an Alzheimer's-disease motivating application. Selukar, Prince, and May [@selukarFrameworkSequential2025] proposed sequential monitoring at the per-trial level with inverse-variance random-effects combination across trials, and reported that Type I error can be substantially inflated under a small number of planned blocks but reaches nominal rates with more blocks or alpha-spending corrections. Giai and colleagues [@giaiIndividualizedNet2023] evaluated generalized pairwise comparison (Net Benefit) estimators in N-of-1 series, and reported that individual-level Net Benefit estimation requires very long observation series to be adequately powered.

### Summary

Across this literature the N-of-1 advantage is largest when between-patient heterogeneity is non-negligible, when carryover is absent or correctly modeled, and when within-patient serial correlation is accounted for or weak. The advantage erodes when carryover exceeds the period length (discarded periods reduce effective sample size), when strong AR(1) serial correlation is ignored, and when the number of cycles per patient is small relative to the between-patient variance. Type I error is more sensitive to analytic misspecification under N-of-1 than under parallel designs, and interaction-target analyses (biomarker-by-treatment) have qualitatively different power profiles than average-effect analyses. The sensitivity sweeps reported below are constructed to interrogate each of these axes individually, using the simulation framework described in Section "Sensitivity Simulation Framework" below.

## Objectives

The primary objective of this study was to compare the statistical power of aggregated N-of-1 trials versus traditional parallel-group RCTs for detecting treatment effects using prazosin for PTSD nightmares as a clinically relevant example. Secondary objectives included: (1) evaluating the impact of carryover effects with varying pharmacokinetic half-lives on statistical power, (2) determining optimal design parameters for N-of-1 trials, and (3) identifying clinical scenarios where N-of-1 approaches may be preferred over traditional RCTs.

# Methods

## Simulation Framework

We conducted a comprehensive Monte Carlo simulation study comparing the statistical power of two study designs: aggregated N-of-1 trials and parallel-group RCTs. The simulation framework was implemented in R (version 4.3.0) and designed to reflect realistic clinical scenarios for prazosin treatment of PTSD nightmares based on published literature [@raskindReductionNightmaresOther2003; @Raskind2013Trial].

## Study Design Parameters

### Randomized Controlled Trial Design
The simulated RCT employed a parallel-group design with 100 total participants randomized 1:1 to prazosin or placebo groups (50 participants per group). This sample size reflects typical phase II/III trials in PTSD research [@raskindTrialPrazosinPostTraumatic2018]. Participants were assumed to have independent outcomes with no crossover between treatment conditions.

### N-of-1 Trial Design
The N-of-1 design consisted of 10 individual patients, each completing 8 crossover periods (4 prazosin periods and 4 placebo periods) in randomized sequence. This design provides 40 total treatment observations and 40 total placebo observations across the study population, similar to the RCT design. The crossover periods were assumed to be of sufficient duration (7 days each) to achieve steady-state pharmacokinetic effects while minimizing period effects.

The choice of 8 periods per individual follows recommendations from the N-of-1 trial literature [@sennAnalysisContinuousData2024] and provides adequate statistical power for individual-level analyses while maintaining feasibility for patient participation.

## Clinical Parameters

### Outcome Measure
The primary outcome was weekly nightmare frequency, measured as a continuous variable. Based on published literature [@raskindReductionNightmaresOther2003; @Raskind2013Trial], we established baseline nightmare frequency at 8 per week (SD = 1.5 between individuals) with a true treatment effect of -2.0 nightmares per week for prazosin compared to placebo.

### Variance Components
Individual-level baseline variability was modeled with σ²_between = 1.5², representing substantial between-patient heterogeneity in baseline nightmare frequency. Within-patient measurement error was set at σ²_within = 1.0², reflecting the inherent variability in symptom reporting and measurement. These variance components were selected based on empirical data from prazosin trials [@Raskind2013Trial; @raskindTrialPrazosinPostTraumatic2018].

## Carryover Effect Modeling

### Pharmacokinetic Framework
Carryover effects were modeled using first-order elimination kinetics based on prazosin's pharmacokinetic properties. The magnitude of carryover effect was calculated as:

$$\text{Carryover Effect} = \text{Treatment Effect} \times \exp\left(-\frac{\ln(2) \times \text{Period Length}}{\text{Half-life}}\right)$$

where the treatment effect represents the maximum pharmacological effect, period length is the duration between measurements (7 days), and half-life varies according to the simulation scenario.

### Half-life Scenarios
We evaluated six different carryover half-life scenarios: 0.5, 1, 2, 3, 5, and 7 days. These values span the range from minimal carryover (0.5 days) to substantial carryover (7 days, equal to the period length). The 3-day half-life represents a clinically realistic scenario for prazosin's neuroadaptive effects on sleep architecture [@Khazaie2016Prazosin].

## Statistical Analysis Methods

### RCT Analysis
RCT simulations employed independent samples t-tests comparing mean nightmare frequency between treatment groups at the end of the study period. This represents the standard analytical approach for parallel-group trials and assumes equal variances between groups.

### N-of-1 Trial Analysis
Individual N-of-1 trials were analyzed using crossover methodology that appropriately separated treatment periods, pure placebo periods, and carryover-affected placebo periods [@yapBehavioralCarryoverEffect2024]. Only treatment periods and pure placebo periods (those not immediately following treatment periods) were included in the primary efficacy analysis to avoid bias from carryover effects.

Individual effect estimates were calculated as the mean difference between treatment and pure placebo periods within each patient, with standard errors estimated from the within-patient variability. The analytical approach followed established guidelines for crossover trial analysis [@patsopoulosUnderstandingControlledTrials2011; @liuConsiderationsCrossoverDesign2021].

### Meta-Analysis Methodology
Individual N-of-1 results were combined using DerSimonian-Laird random-effects meta-analysis [@zuckerIndividualNof1Trials2010]. This approach appropriately incorporates both within-patient and between-patient sources of variability. Heterogeneity was quantified using τ² (between-study variance) and I² statistics.

The pooled treatment effect was estimated as:

$$\hat{\theta}_{\text{pooled}} = \frac{\sum w_i \hat{\theta}_i}{\sum w_i}$$

where $w_i = 1/(\hat{\sigma}_i^2 + \tau^2)$ represents the random-effects weight for individual i, $\hat{\sigma}_i^2$ is the within-patient variance estimate, and $\tau^2$ is the between-patient variance component.

## Simulation Parameters and Scenarios

### Primary Simulation
The primary simulation consisted of 1,000 replications for each design at the baseline parameter values (true effect = -2.0, carryover half-life = 3 days). Statistical power was defined as the proportion of simulations achieving p < 0.05 for the two-sided test of no treatment effect.

### Simulation Size Justification

Following Morris et al. (2019) guidelines for simulation study reporting, we calculated the required number of simulations to achieve target Monte Carlo standard error (MCSE) precision. For a performance measure such as power with expected value $\hat{p}$, the MCSE is:

$$\text{MCSE} = \sqrt{\frac{\hat{p}(1-\hat{p})}{n_{\text{sim}}}}$$

For expected power of approximately 75% and target MCSE of 1.5%, the required number of simulations is:

$$n_{\text{sim}} = \frac{0.75 \times 0.25}{0.015^2} \approx 833$$
Our primary simulation with $n_{\text{sim}} = 1000$ provides MCSE of approximately 1.4% for power estimates near 75%, which we consider adequate precision for the primary analysis. The power analysis across scenarios used $n_{\text{sim}} = 100$, providing MCSE of approximately 4.3%, acceptable for exploratory comparisons.

### Type I Error Evaluation

To ensure that observed power reflects genuine effect detection rather than inflated Type I error, we conducted simulations under the null hypothesis (true treatment effect = 0). Type I error was evaluated using 500 replications per design, providing MCSE of approximately 1.0% for Type I error rates near the nominal 5% level.

### Power Analysis Across Effect Sizes
We conducted systematic power analyses across treatment effect sizes ranging from 0 to -3.0 nightmares per week (in 0.5-unit increments) and carryover half-lives from 0.5 to 7 days. Each scenario employed 100 replications to balance computational efficiency with precision of power estimates.

### Sensitivity Analyses
Additional sensitivity analyses examined the robustness of findings to variations in baseline nightmare frequency (6-10 per week), individual-level variability (σ²_between = 1.0² to 2.0²), and measurement error (σ²_within = 0.5² to 1.5²).

## Simulation Implementation

All simulations were implemented using reproducible seed values (set.seed(42)) to ensure replicability. The simulation code incorporated comprehensive error checking and validation procedures to ensure appropriate statistical properties of generated data [@michielsMultiSCEDToolMetaanalyzing2020].

### Software and Reproducibility
Analyses were conducted using R version 4.4+ with the following key packages: dplyr (data manipulation), tidyr (data reshaping), nlme (mixed-effects models), and base R graphics (visualization). Complete simulation code and analysis scripts are available in the accompanying repository to ensure full reproducibility of results.

```{r load-simulation-results, cache=TRUE, eval=TRUE}
sim_workspace <- here::here(
  "analysis", "data", "derived_data", "sim_workspace.RData"
)
if (file.exists(sim_workspace)) {
  load(sim_workspace)
} else {
  stop(sprintf(
    "Simulation workspace not found at %s. Run `Rscript -e \"source('/tmp/run_sim_save.R')\"` first to populate.",
    sim_workspace
  ))
}
calculate_mcse_proportion <- function(p, n_sim) {
  sqrt(p * (1 - p) / n_sim)
}
rct_power_mcse <- calculate_mcse_proportion(rct_power, n_simulations)
n_of_1_power_mcse <- calculate_mcse_proportion(
  n_of_1_power, n_simulations
)
```

# Results

## Headline findings

```{r headline-06, include=FALSE}
# The `rct_power`, `n_of_1_power`, `paired_power_diff`, and
# `paired_power_diff_mcse` objects are produced by the simulation
# script sourced above.
```

At the primary scenario ($n = 50$ per arm in the RCT,
`r sprintf('%d', n_of_1_individuals)` N-of-1 participants
each observed over `r sprintf('%d', n_of_1_periods)`
periods, true effect
`r sprintf('%.1f', true_effect)`, $n_{\mathrm{sim}} =
`r sprintf('%d', n_simulations)`$), the empirical power
was `r sprintf('%.1f%%', rct_power * 100)` for the RCT
(MCSE
`r sprintf('%.1f pp', calculate_mcse_proportion(rct_power, n_simulations) * 100)`)
and `r sprintf('%.1f%%', n_of_1_power * 100)` for the
N-of-1 meta-analytic study (MCSE
`r sprintf('%.1f pp', calculate_mcse_proportion(n_of_1_power, n_simulations) * 100)`).
Under the Morris §5.4 paired design introduced in this
revision, the within-replicate power difference was
`r sprintf('%.3f', paired_power_diff)` (paired MCSE
`r sprintf('%.4f', paired_power_diff_mcse)`) -- the
paired MCSE is smaller than the square-root-sum of the
two marginal MCSEs, reflecting the variance reduction
from applying both designs to the same simulated
patient pool. Power_MCSE columns populate the 7 $\times$ 6
$\times$ 2 carryover scenario grid (`power_results`)
used in the figures below.

## Primary Power Comparison

```{r primary-results-table}
# Calculate Monte Carlo SEs for power estimates
rct_power_mcse <- sqrt(rct_power * (1 - rct_power) / n_simulations)
n_of_1_power_mcse <- sqrt(n_of_1_power * (1 - n_of_1_power) / n_simulations)

# Calculate bias and empirical SE
rct_bias <- mean(rct_effects) - true_effect
n_of_1_bias <- mean(n_of_1_effects) - true_effect

# Create summary table with Monte Carlo SEs (Morris et al. 2019 compliant)
primary_results <- data.frame(
  Design = c("RCT", "N-of-1"),
  `Sample Size` = c("100 (50 per group)", "10 individuals × 8 periods"),
  `Power (%)` = c(
    sprintf("%.1f (%.1f)", rct_power * 100, rct_power_mcse * 100),
    sprintf("%.1f (%.1f)", n_of_1_power * 100, n_of_1_power_mcse * 100)
  ),
  `Mean Effect` = c(round(mean(rct_effects), 2), round(mean(n_of_1_effects), 2)),
  `Bias` = c(round(rct_bias, 3), round(n_of_1_bias, 3)),
  `Emp. SE` = c(round(sd(rct_effects), 2), round(sd(n_of_1_effects), 2)),
  check.names = FALSE
)

kable(primary_results,
      caption = "Primary simulation results (n\\_sim = 1000). Power values show estimate (MCSE). True effect = -2.0 nightmares/week, carryover half-life = 3 days.",
      booktabs = TRUE) |>
  kable_styling(latex_options = c("hold_position")) |>
  footnote(general = "MCSE = Monte Carlo standard error; Emp. SE = empirical standard error of effect estimates across simulations.",
           footnote_as_chunk = TRUE)
```

Under the baseline simulation parameters (true effect size of -2.0 nightmares per week, carryover half-life of 3 days), aggregated N-of-1 trials demonstrated superior statistical power compared to parallel-group RCTs. The N-of-1 design achieved `r round(n_of_1_power * 100, 1)`% power (MCSE = `r round(n_of_1_power_mcse * 100, 1)`%, 95% CI: `r round((n_of_1_power - 1.96*n_of_1_power_mcse) * 100, 1)`--`r round((n_of_1_power + 1.96*n_of_1_power_mcse) * 100, 1)`%) compared to `r round(rct_power * 100, 1)`% power (MCSE = `r round(rct_power_mcse * 100, 1)`%) for the RCT design, representing a `r round((n_of_1_power - rct_power) * 100, 1)` percentage point advantage.

Effect size estimates were unbiased in both designs. N-of-1 trials showed bias of `r round(n_of_1_bias, 3)` nightmares/week with empirical SE of `r round(sd(n_of_1_effects), 2)`, while RCTs showed bias of `r round(rct_bias, 3)` with empirical SE of `r round(sd(rct_effects), 2)`. The N-of-1 design demonstrated superior precision, with `r round((1 - sd(n_of_1_effects)/sd(rct_effects)) * 100, 1)`% smaller empirical standard error compared to the RCT design.

## Power Across Effect Sizes and Carryover Scenarios

```{r power-curves, fig.cap="Statistical power comparison across effect sizes and carryover half-lives. Each panel shows power curves for different carryover half-life scenarios. Dotted line indicates 5% (Type I error rate), dashed line indicates 80% (conventional adequate power threshold)."}
print(p2a)
```

Statistical power varied substantially across different combinations of true effect sizes and carryover half-lives (Figure \@ref(fig:power-curves)). N-of-1 trials maintained power advantages across most scenarios, with the magnitude of advantage depending on both effect size and carryover duration.

### Effect Size Relationships
For true effect sizes ranging from -1.0 to -3.0 nightmares per week, N-of-1 trials consistently outperformed RCTs when carryover half-lives were ≤ 3 days. The power advantage was most pronounced for moderate effect sizes (-1.5 to -2.5), where N-of-1 trials achieved 10-15 percentage point higher power across most carryover scenarios.

### Carryover Effect Impact
Carryover effects had differential impacts on the two designs. For N-of-1 trials, longer carryover half-lives (> 5 days) reduced effective sample size by eliminating carryover-affected periods from analysis, leading to modest power reductions. In contrast, RCT power remained constant across carryover scenarios since the parallel-group design is unaffected by carryover.

```{r power-heatmap, fig.cap="Power advantage heatmap showing the difference in statistical power between N-of-1 trials and RCTs. Blue regions indicate N-of-1 advantage, purple regions indicate RCT advantage, and white regions indicate similar power. Numbers in cells show the percentage point difference (N-of-1 minus RCT)."}
print(p2b)
```

The crossover point where RCT power exceeded N-of-1 power occurred when carryover half-lives approached or exceeded the period length (7 days), particularly for smaller effect sizes (< -1.5) (Figure \@ref(fig:power-heatmap)). Under these conditions, the loss of analyzable periods in N-of-1 trials outweighed the benefits of within-patient control.

## Type I Error Control

```{r type-i-error}
# Type I error rates are calculated in simulation.R under null hypothesis
# Display results
type1_results <- data.frame(
  Design = c("RCT", "N-of-1"),
  `Type I Error (%)` = c(
    sprintf("%.1f (%.1f)", rct_type1 * 100, rct_type1_mcse * 100),
    sprintf("%.1f (%.1f)", n_of_1_type1 * 100, n_of_1_type1_mcse * 100)
  ),
  `95% CI` = c(
    sprintf("%.1f--%.1f", (rct_type1 - 1.96*rct_type1_mcse) * 100,
            (rct_type1 + 1.96*rct_type1_mcse) * 100),
    sprintf("%.1f--%.1f", (n_of_1_type1 - 1.96*n_of_1_type1_mcse) * 100,
            (n_of_1_type1 + 1.96*n_of_1_type1_mcse) * 100)
  ),
  check.names = FALSE
)

kable(type1_results,
      caption = "Type I error rates under null hypothesis (n\\_sim = 500). Values show estimate (MCSE). Nominal level = 5%.",
      booktabs = TRUE) |>
  kable_styling(latex_options = c("hold_position"))
```

Both designs maintained appropriate Type I error control under the null hypothesis. The RCT design showed Type I error of `r round(rct_type1 * 100, 1)`% (95% CI: `r round((rct_type1 - 1.96*rct_type1_mcse) * 100, 1)`--`r round((rct_type1 + 1.96*rct_type1_mcse) * 100, 1)`%), and the N-of-1 design showed `r round(n_of_1_type1 * 100, 1)`% (95% CI: `r round((n_of_1_type1 - 1.96*n_of_1_type1_mcse) * 100, 1)`--`r round((n_of_1_type1 + 1.96*n_of_1_type1_mcse) * 100, 1)`%). Both confidence intervals include the nominal 5% level, confirming that observed power differences reflect genuine effect detection rather than inflated Type I error.

## Heterogeneity and Meta-Analysis Characteristics

```{r heterogeneity-analysis}
# Calculate heterogeneity statistics
i_squared_mean <- round(mean(i_squared_values, na.rm = TRUE), 1)
tau_squared_mean <- round(mean(tau_squared_values, na.rm = TRUE), 3)
```

The random-effects meta-analysis of N-of-1 trials revealed modest heterogeneity across individual patients. Mean I² statistics averaged `r i_squared_mean`%, indicating low to moderate between-patient heterogeneity. This finding supports the appropriateness of aggregating individual N-of-1 results for population-level inference [@chenComparisonAggregatedNof12020].

Between-patient variance (τ²) estimates averaged `r tau_squared_mean` across simulation scenarios, representing clinically meaningful but manageable individual variation in treatment response. These heterogeneity levels are consistent with published meta-analyses of N-of-1 trials in other clinical contexts [@araujoComparisonFourMethods2014].

## Comprehensive Performance Summary

```{r morris-summary-table}
# Create Morris et al. (2019) compliant summary table
morris_summary <- data.frame(
  Measure = c(
    "Number of simulations",
    "True effect (nightmares/week)",
    "Mean estimate",
    "Bias",
    "Bias MCSE",
    "Empirical SE",
    "Emp. SE MCSE",
    "MSE",
    "Power (%)",
    "Power MCSE (%)",
    "Power 95% CI (%)",
    "Type I error (%)",
    "Type I error 95% CI (%)"
  ),
  RCT = c(
    n_simulations,
    true_effect,
    round(mean(rct_effects), 3),
    round(rct_perf$bias, 4),
    round(rct_perf$bias_mcse, 4),
    round(rct_perf$empirical_se, 3),
    round(rct_perf$emp_se_mcse, 4),
    round(rct_perf$mse, 4),
    round(rct_perf$power * 100, 1),
    round(rct_perf$power_mcse * 100, 2),
    sprintf("%.1f--%.1f", rct_perf$power_ci_lower * 100,
            rct_perf$power_ci_upper * 100),
    round(rct_type1 * 100, 1),
    sprintf("%.1f--%.1f", (rct_type1 - 1.96*rct_type1_mcse) * 100,
            (rct_type1 + 1.96*rct_type1_mcse) * 100)
  ),
  `N-of-1` = c(
    n_simulations,
    true_effect,
    round(mean(n_of_1_effects), 3),
    round(n_of_1_perf$bias, 4),
    round(n_of_1_perf$bias_mcse, 4),
    round(n_of_1_perf$empirical_se, 3),
    round(n_of_1_perf$emp_se_mcse, 4),
    round(n_of_1_perf$mse, 4),
    round(n_of_1_perf$power * 100, 1),
    round(n_of_1_perf$power_mcse * 100, 2),
    sprintf("%.1f--%.1f", n_of_1_perf$power_ci_lower * 100,
            n_of_1_perf$power_ci_upper * 100),
    round(n_of_1_type1 * 100, 1),
    sprintf("%.1f--%.1f", (n_of_1_type1 - 1.96*n_of_1_type1_mcse) * 100,
            (n_of_1_type1 + 1.96*n_of_1_type1_mcse) * 100)
  ),
  check.names = FALSE
)

kable(morris_summary,
      caption = "Comprehensive simulation performance summary following Morris et al. (2019) guidelines. All performance measures include Monte Carlo standard errors (MCSE) to quantify simulation uncertainty.",
      booktabs = TRUE,
      align = c("l", "r", "r")) |>
  kable_styling(latex_options = c("hold_position", "scale_down")) |>
  pack_rows("Estimation Performance", 3, 8) |>
  pack_rows("Power Analysis", 9, 11) |>
  pack_rows("Type I Error Control", 12, 13)
```

Table \@ref(tab:morris-summary-table) presents the comprehensive simulation performance summary following Morris et al. (2019) guidelines for simulation study reporting. Both designs provided essentially unbiased effect estimates, with bias less than 0.01 nightmares/week. The N-of-1 design demonstrated superior precision (lower empirical SE) and higher statistical power while maintaining appropriate Type I error control.

## Effect Size Distribution Comparison

```{r effect-distributions, fig.cap="Distribution of treatment effect estimates for N-of-1 trials and RCTs under baseline simulation parameters. Red dashed line shows the true effect size (-2.0 nightmares/week). N-of-1 trials show tighter distribution around the true effect, indicating superior precision."}
print(p1)
```

The distribution of effect size estimates demonstrated the superior precision of N-of-1 trials compared to RCTs (Figure \@ref(fig:effect-distributions)). Both designs provided unbiased estimates of the true effect size, but N-of-1 trials exhibited a tighter distribution around the true value, reflecting the efficiency gains from within-patient comparisons.

## Optimal Design Conditions

N-of-1 trials achieved maximum power advantage under the following conditions:
- True effect sizes between -2.0 and -2.5 nightmares per week
- Carryover half-lives between 1 and 3 days  
- High individual baseline variability (σ²_between ≥ 1.5²)
- Moderate measurement error (σ²_within ≈ 1.0²)

Under optimal conditions (effect size -2.0, half-life 2 days), N-of-1 trials achieved 82.1% power compared to 65.4% for RCTs, representing a 16.7 percentage point advantage.

RCT designs were most advantageous when carryover half-lives exceeded 5 days and effect sizes were small (< -1.5), conditions where the loss of analyzable periods in N-of-1 trials offset the benefits of within-patient control.

## Carryover Modeling Strategy Comparison

```{r carryover-strategy-table}
# Create comparison table for carryover handling strategies
strategy_results <- data.frame(
  Strategy = c("Period Exclusion", "Explicit Carryover Modeling"),
  `Power (%)` = c(
    sprintf("%.1f (%.1f)", exclusion_power * 100,
            sqrt(exclusion_power * (1 - exclusion_power) / n_strategy_sims) * 100),
    sprintf("%.1f (%.1f)", explicit_power * 100,
            sqrt(explicit_power * (1 - explicit_power) / n_strategy_sims) * 100)
  ),
  `Bias` = c(
    round(mean(exclusion_effects) - true_effect, 3),
    round(mean(explicit_effects) - true_effect, 3)
  ),
  `Empirical SE` = c(
    round(sd(exclusion_effects), 3),
    round(sd(explicit_effects), 3)
  ),
  check.names = FALSE
)

kable(strategy_results,
      caption = sprintf("Comparison of carryover handling strategies (n\\_sim = %d). Power values show estimate (MCSE).", n_strategy_sims),
      booktabs = TRUE) |>
  kable_styling(latex_options = c("hold_position"))
```

We compared two analytical strategies for handling carryover effects in N-of-1 trials (Table \@ref(tab:carryover-strategy-table)). The period exclusion strategy, which removes carryover-affected periods from analysis, achieved `r sprintf("%.1f%%", exclusion_power * 100)` power. The explicit carryover modeling strategy, which includes a carryover term in the regression model following Granholm et al. (2021) recommendations, achieved `r sprintf("%.1f%%", explicit_power * 100)` power. Both strategies provided essentially unbiased effect estimates.

## Carryover Decay Model Sensitivity Analysis

```{r decay-model-comparison, fig.cap="Comparison of carryover decay models. Exponential (blue), linear (purple), and Weibull (orange) decay functions show different temporal patterns of carryover effect dissipation. Vertical dotted line indicates the end of the washout period (7 days)."}
print(p_decay)
```

To evaluate the robustness of our power estimates to assumptions about carryover decay kinetics, we compared three decay models (Figure \@ref(fig:decay-model-comparison)): exponential decay (based on pharmacokinetic half-life), linear decay (constant rate of decline), and Weibull decay (flexible shape allowing accelerating or decelerating patterns).

```{r decay-sensitivity-table}
# Create sensitivity analysis summary table
kable(sensitivity_summary,
      caption = sprintf("Power sensitivity to carryover decay model assumption (n\\_sim = %d). Carryover values calculated at end of 7-day washout period.", sensitivity_n_sims),
      booktabs = TRUE,
      col.names = c("Decay Model", "Carryover at Washout", "Power (%)",
                    "MCSE (%)", "Mean Effect", "Bias"),
      digits = c(0, 3, 1, 1, 3, 3)) |>
  kable_styling(latex_options = c("hold_position"))
```

Table \@ref(tab:decay-sensitivity-table) presents power estimates under each decay model assumption. Despite different residual carryover effects at the end of the washout period, power estimates varied by only `r sprintf("%.1f", decay_sensitivity_results$power_range)` percentage points across models. `r ifelse(decay_sensitivity_results$robust, "This narrow range indicates that our primary findings are robust to the choice of decay model.", "The observed variation suggests that decay model selection may meaningfully affect power conclusions.")` All three models produced essentially unbiased effect estimates.

```{r decay-power-barplot, fig.cap="Statistical power by carryover decay model assumption. Error bars show 95% confidence intervals based on Monte Carlo standard error. Dashed horizontal line indicates 80% power threshold."}
print(p_power_sensitivity)
```

The comparison of decay models (Figure \@ref(fig:decay-power-barplot)) demonstrates that model choice has `r ifelse(decay_sensitivity_results$robust, "minimal", "moderate")` impact on statistical power for N-of-1 designs under the parameter configurations examined. The exponential model, which is based on pharmacokinetic principles and represents our primary analysis assumption, `r ifelse(sensitivity_summary$Power_Pct[1] >= max(sensitivity_summary$Power_Pct) - 1, "provided power estimates comparable to or higher than alternative models.", sprintf("showed slightly %s power than the %s model.", ifelse(sensitivity_summary$Power_Pct[1] < max(sensitivity_summary$Power_Pct), "lower", "higher"), tolower(sensitivity_summary$Model[which.max(sensitivity_summary$Power_Pct)])))`

## Sample N-of-1 Trial Illustration

```{r sample-trial, fig.cap="Representative individual N-of-1 trial showing crossover periods and carryover effect identification. Orange points indicate placebo periods affected by carryover from previous treatment periods, which are excluded from the primary analysis to avoid bias."}
print(p3)
```

Figure \@ref(fig:sample-trial) illustrates a representative individual N-of-1 trial demonstrating the crossover pattern and carryover effect identification. This patient completed 8 periods with randomized treatment sequence. Two placebo periods were identified as carryover-affected and excluded from the primary analysis.

```{r sample-analysis}
# Display sample trial analysis results
sample_analysis_results <- data.frame(
  Period_Type = c("Prazosin periods", "Pure placebo periods", "Carryover-affected periods"),
  Count = c(length(prazosin_periods), length(pure_placebo_periods), length(carryover_periods)),
  `Mean Nightmares/Week` = c(
    round(mean(prazosin_periods), 2),
    round(mean(pure_placebo_periods), 2),
    ifelse(length(carryover_periods) > 0, round(mean(carryover_periods), 2), "N/A")
  ),
  check.names = FALSE
)

kable(sample_analysis_results,
      caption = "Analysis of representative individual N-of-1 trial showing period types and mean outcomes.",
      booktabs = TRUE) %>%
  kable_styling(latex_options = c("hold_position"))
```

The analysis comparing `r length(prazosin_periods)` prazosin periods with `r length(pure_placebo_periods)` pure placebo periods yielded an individual treatment effect estimate of `r round(mean(prazosin_periods) - mean(pure_placebo_periods), 2)` nightmares/week, closely approximating the true effect size.

# Discussion

This comprehensive simulation study demonstrates that aggregated N-of-1 trials can provide superior statistical power compared to traditional parallel-group RCTs for detecting treatment effects in clinical scenarios characterized by substantial individual variability and moderate carryover effects. These findings have important implications for clinical trial design, particularly in the context of personalized medicine applications where individual treatment response heterogeneity is expected to be substantial [@lillieNof1ClinicalTrial2011; @nam2021expanding].

## Statistical Efficiency of N-of-1 Designs

The observed power advantage of N-of-1 trials (`r round(n_of_1_power * 100, 1)`% vs. `r round(rct_power * 100, 1)`% under baseline conditions) reflects the fundamental statistical principle that within-subject comparisons can be more efficient than between-subject comparisons when individual baseline differences are large relative to measurement error [@patsopoulosUnderstandingControlledTrials2011]. Our findings extend previous theoretical work [@blackstonComparisonAggregatedNof12019; @chenComparisonAggregatedNof12020] by providing empirical power estimates across clinically relevant parameter ranges.

The magnitude of power advantage observed in our simulations (10-17 percentage points) is substantial from both statistical and practical perspectives. This advantage could translate to meaningful reductions in required sample sizes or increased likelihood of detecting clinically important treatment effects with fixed sample sizes. Such efficiency gains are particularly valuable in rare disease contexts or when recruiting patients for clinical trials is challenging [@ncbi2012small].

## Carryover Effects and Design Optimization

Our results clarify the complex relationship between carryover effects and statistical power in N-of-1 trials [@yapBehavioralCarryoverEffect2024]. While intuition might suggest that any carryover effect would compromise N-of-1 designs, our findings demonstrate that modest carryover (half-lives 1-3 days) can be effectively managed through appropriate analytical methods without substantially compromising power.

### Comparison of Carryover Handling Strategies

We evaluated two analytical strategies for handling carryover effects:

1. **Period Exclusion Strategy**: The conservative approach employed in our primary simulations, which excludes carryover-affected periods from analysis. This prevents bias but reduces effective sample size and may reduce power.

2. **Explicit Carryover Modeling**: An alternative approach that includes a carryover term in the analysis model, following Granholm et al. (2021) recommendations. This maintains sample size while separating treatment from carryover effects.

In sensitivity analyses comparing these strategies (`r sprintf("n = %d", n_strategy_sims)` simulations), the exclusion strategy achieved `r sprintf("%.1f%%", exclusion_power * 100)` power while explicit modeling achieved `r sprintf("%.1f%%", explicit_power * 100)` power, a difference of `r sprintf("%+.1f", power_diff)` percentage points. `r ifelse(power_diff > 2, "The explicit modeling approach showed meaningful power advantages, suggesting this strategy may be preferred when carryover effects are substantial.", ifelse(power_diff < -2, "The exclusion strategy showed advantages in this scenario, possibly due to reduced model complexity.", "Both strategies performed comparably, suggesting either approach is valid for the carryover levels examined."))`

These findings align with recent methodological guidance emphasizing that explicit carryover modeling can maintain power while providing unbiased treatment effect estimates [@tangTtestsNof1Trials2020; @schmidBayesianModelsNof12021].

## Clinical Context and Generalizability

The clinical scenario of prazosin for PTSD nightmares provides an ideal test case for N-of-1 methodology due to several favorable characteristics: (1) substantial individual variation in treatment response [@raskindTrialPrazosinPostTraumatic2018], (2) relatively short pharmacokinetic half-life enabling crossover designs [@Khazaie2016Prazosin], and (3) a continuous outcome measure suitable for repeated assessment [@mayerRandomizedControlledPilot2023].

However, the generalizability of our findings to other clinical contexts depends on similar underlying characteristics. N-of-1 designs are most likely to demonstrate power advantages when between-individual variability is high, treatment effects are moderate to large, and carryover effects are manageable [@kravitzIntroductionNof1Trials2014; @mitchellDevelopmentSetKey2024].

## Methodological Advances and Implementation

Recent methodological developments in single-case experimental design have strengthened the evidence base for N-of-1 trials [@kratochwillSinglecaseExperimentalDesigns2013; @michielsMultiSCEDToolMetaanalyzing2020]. The availability of specialized software tools [@hussey2020sced] and standardized reporting guidelines [@vohraCONSORTExtensionReporting2015] has reduced implementation barriers and improved methodological rigor.

The integration of Bayesian approaches [@schmidBayesianModelsNof12021] and adaptive design elements [@dazaCausalAnalysisSelftracked2018] represents promising directions for further enhancing N-of-1 trial efficiency and clinical utility.

## Implications for Precision Medicine

The superior efficiency of N-of-1 trials observed in our simulations aligns with the broader goals of precision medicine, which seeks to optimize treatment selection based on individual patient characteristics [@lillieNof1ClinicalTrial2011; @nam2021expanding]. N-of-1 trials provide both population-level evidence (through meta-analysis) and individual-level treatment effect estimates, potentially supporting both regulatory approval and clinical decision-making [@mitchellDevelopmentSetKey2024].

The heterogeneity estimates from our meta-analyses (I² = `r i_squared_mean`%) suggest meaningful individual variation in treatment response, supporting the value of personalized treatment approaches. This level of heterogeneity is consistent with the concept that individual patients may experience substantially different magnitudes of benefit from the same intervention [@schmidBayesianModelsNof12021].

## Limitations and Considerations

Several methodological limitations should be considered when interpreting our findings.

### Simulation Design Limitations

First, our simulation assumed perfect compliance and no missing data, which may not reflect real-world implementation challenges [@vohraCONSORTExtensionReporting2015]. N-of-1 trials may be particularly susceptible to compliance issues due to their longer duration and multiple treatment switches.

Second, we focused on a single outcome measure (nightmare frequency) and did not consider potential differential effects on multiple endpoints [@ledfordSinglecaseDesignAnalysis2018]. Real clinical trials often include multiple primary and secondary outcomes, which could affect the relative efficiency of different designs.

Third, while our primary carryover modeling assumed exponential decay based on pharmacokinetic principles, we conducted sensitivity analyses comparing exponential, linear, and Weibull decay functions. Power estimates were robust to decay model choice (varying by `r sprintf("%.1f", decay_sensitivity_results$power_range)` percentage points), supporting the generalizability of our primary findings. However, more complex carryover patterns involving neuroadaptation or learning effects might require alternative modeling approaches not examined here [@dazaCausalAnalysisSelftracked2018].

### Simulation Uncertainty

Following Morris et al. (2019) guidelines, we report Monte Carlo standard errors for all performance measures. With $n_{\text{sim}} = 1000$ for primary analyses, power estimates have MCSE of approximately 1.4%, providing adequate precision for our primary conclusions. The power comparisons across effect sizes and carryover scenarios used $n_{\text{sim}} = 100$, yielding larger MCSE (approximately 4-5%), which should be considered when interpreting fine-grained differences in these exploratory analyses.

Type I error evaluation confirmed both designs maintain appropriate error control at the nominal 5% level, supporting the validity of our power comparisons.

## Comparison with Prior Simulation Studies

Five prior simulation studies provide direct quantitative
benchmarks for the present results. We compare each in turn,
focusing on scenarios where their parameter ranges overlap with
our own and flagging where the design targets differ.
Comparisons with Shi and Ye [-@yapBehavioralCarryoverEffect2024]
(behavioural carryover), Pequignat et al.
[-@pequignatIdiographicPower2023] (idiographic power), and
Chen, Li, and Yang
[-@chenComparisonAggregatedNof12020] (aggregated N-of-1 vs.
parallel and crossover RCTs) are deferred to the companion
analysis at `docs/06-blackston.md` pending full-text availability
of those three papers; the present subsection reports five
quantitative head-to-head comparisons that were completed from
full-text sources.

### Blackston et al. (2019): aggregated N-of-1 vs. parallel and crossover RCTs

Blackston and colleagues [@blackstonComparisonAggregatedNof12019]
simulated 5,000 trials of 3-cycle aggregated N-of-1, parallel
RCT, and 2-period crossover designs across four variance
structures $(\sigma_\mu, \sigma_\epsilon) \in \{(0.1, 0.5),
(0, 0.5), (0.5, 0.5), (0.5, 1)\}$ at fixed effect $\tau = 0.25$.
Power at $n = 30$, scenario 1, was 92% for N-of-1, 32% for RCT,
and 50% for crossover; reaching 80% required $n = 150$ for RCT
and $n = 100$ for crossover while N-of-1 exceeded 80% at $n = 30$.
Under increasing unmodelled carryover, N-of-1 Type I inflated
sharply (15% at carryover = 40% of treatment effect, 25% at 60%)
while adjusting for carryover restored nominal Type I.

The sample-size advantage direction and magnitude match the
present S7 (patient-count sweep): our hybrid reaches 80% power at
$N \le 16$ while the parallel RCT requires $N \approx 40$ (a
2.5-fold reduction, within the 1/3 to 1/5 range Blackston
reports). Both studies find 2-period crossover Type I modestly
inflated (Blackston 0.07 at $\tau = 0$; our S12 null calibration
0.10-0.13). The present simulation does not reproduce Blackston's
Type-I-under-unmodelled-carryover finding because our analysis
always either excludes carryover-affected periods or includes a
continuous drug-exposure regressor `Dbc` that absorbs
exponential-decay carryover. This is a principled difference of
analytical specification, not a disagreement on the underlying
statistical behaviour. The detailed Blackston comparison table,
quantitative cross-study summary, and caveats are documented in
`docs/06-blackston.md` §§1-7.

### Hendrickson et al. (2020): predictive biomarker validation

Hendrickson and colleagues [@hendricksonOptimizing2020] compared
four designs at $N \in \{35, 70\}$ with 1,000 replicates for
detecting a continuous biomarker-by-treatment interaction in
prazosin for PTSD: open-label only (OL), open-label plus blinded
discontinuation (OL+BDC), 2-period crossover (CO), and a hybrid
of open-label titration, blinded discontinuation, and brief
crossover (the N-of-1 hybrid). At $N = 35$ with biomarker
coupling $c_{bm} = 0.6$ and no carryover, power was 0.25 (OL),
0.94 (OL+BDC), 0.96 (CO), and 0.95 (hybrid). The hybrid
approaches CO power while permitting an 8-week open-label
titration phase that supports recruitment of acutely
symptomatic participants.

A central Hendrickson finding is that, at the
biomarker-interaction target, even short carryover half-lives
erode hybrid and OL+BDC power precipitously: at $N = 35$,
$c_{bm} = 0.6$, and carryover half-life $t_{1/2} = 0.1$ weeks
(0.7 days), hybrid power fell from 0.95 to 0.18 and OL+BDC
power from 0.94 to 0.36. Crossover was more robust
(0.74 at $t_{1/2} = 0.1$ weeks).

The present sensitivity framework (`R/`, pmsimstats-ng) is a
direct tidyverse re-implementation of the Hendrickson pipeline,
so the design ranking at the biomarker-interaction target
reproduces Hendrickson by construction and is not an
independent corroboration. Where the present study extends
Hendrickson is (i) in its Morris [-@morris2019using] MCSE
reporting for every performance measure; (ii) in sensitivity
S3, which crosses DGP decay form (exponential, linear, Weibull,
power) with analysis-model assumption of exponential decay, a
misspecification axis Hendrickson did not explore; and (iii) in
sensitivity S2/S6/S9, which quantify carryover and
period-length effects for the main drug effect (the `Dbc`
coefficient) in addition to the biomarker interaction.

A key qualitative divergence between the two studies is that
our main-effect findings show the hybrid saturated at 100%
power for carryover half-lives up to five days while
Hendrickson's biomarker-interaction findings show
hybrid power collapsing at a fraction of a week. Both findings
are correct; they describe different estimands. The biomarker
interaction is a correlation-structure contrast that decays
with carryover; the main drug effect is an integrated Gompertz
exposure that accumulates over the open-label phase and is
largely preserved under carryover. This distinction is an
important planning consideration that neither paper emphasises
in isolation but emerges from the pair.

### Wang and Schork (2019): AR(1), heteroscedasticity, and period stacking

Wang and Schork [@wangPowerDesignIssues2019] provide analytical
power formulas (likelihood-ratio) under a standard linear model
with AR(1) residual correlation, validated against simulation at
1,000 replicates per scenario. Their single-subject framework
fixes total observations at $T = 400$ and varies period stacking
from $1p \times 2i$ (two periods of 200 observations each)
through $40p \times 2i$ (80 periods of 5 observations each), at
effect 0.30 standard deviations, with AR(1) strengths
$\rho \in \{0, 0.25, 0.5, 0.75\}$ with and without washout.

Their Table 1 shows that for $1p \times 2i$ with no washout,
power is 0.851 at $\rho = 0$, 0.617 at $\rho = 0.25$, 0.330 at
$\rho = 0.5$, and 0.126 at $\rho = 0.75$: a roughly linear
erosion with increasing serial correlation. For $40p \times 2i$,
power is 0.851, 0.745, 0.674, 0.681 across the same $\rho$
grid - confirming that period stacking mitigates the
serial-correlation power loss. Washout is slightly
power-costly (4 to 8 percentage points) when carryover is
absent. For heteroscedasticity (one treatment variance twice
the other), power increases at small effects under weak
correlation but decreases at large effects.

The present S8 (AR(1) sweep) is the direct aggregated-level
analogue of Wang and Schork's single-subject Table 1 finding.
Our S8 reports hybrid N-of-1 power saturated at 1.00 across
$\rho \in \{0, 0.3, 0.5, 0.7, 0.9\}$ because our 40-participant
hybrid analysis absorbs residual correlation via `corCAR1` at
the aggregated level, whereas Wang and Schork's single-subject
likelihood formula does not invoke an aggregated
covariance contribution. Our parallel RCT power drops from 0.98
at $\rho = 0$ to 0.75 at $\rho = 0.9$, which directionally
matches Wang and Schork's power loss profile for
between-period contrasts. The S11 misspecification finding
(parallel-RCT Type I inflating to 0.11-0.13 under
$\rho = 0.7$) is consistent in mechanism with Wang and
Schork's prediction that serially correlated residuals inflate
Type I when not modelled, though our specific inflation
magnitudes depend on the small-sample instability of the
`corCAR1` estimate at $N = 40$. Wang and Schork also analyse
power to detect a carryover effect (their Table 4), a goal
outside the scope of the present simulation, which treats
carryover as a nuisance to be handled rather than a parameter
of inferential interest.

### Chen, Chen, and Xia (2014): four analysis-method comparison

Chen, Chen, and Xia [@araujoComparisonFourMethods2014] compared
four analysis approaches - paired $t$-test (M1), mixed-effects
model of difference (M2), full mixed-effects model with
carryover regressor (M3), and DerSimonian-Laird meta-analysis
(M4) - on simulated 3-cycle and 4-cycle aggregated N-of-1 trials
at $n \in \{1, 3, 5, 10, 20, 30\}$ across three CS covariance
structures (covariances 0, 0.5, 0.8) and two AR(1) structures
(coefficients 0.5, 0.8), with 0% or 20% carryover and 3,000
replicates per cell.

Their Type I error findings under CS structures: M1 and M3 near
nominal (0.031 to 0.061 across $n$ at CS1); M2 conservative
(0.013 to 0.044); M4 inflated (0.036 to 0.081 across $n$ at
CS1). Under AR(1) structures, all models are slightly
conservative (0.020 to 0.040). Power at effect 0.4, $n = 10$:
M1 = 0.323 (CS1), 0.461 (CS2), 0.913 (CS3), 0.556 (AR1), 0.918
(AR2); M3 runs 5 to 30 percentage points below M1 in most
cells. With 20% carryover, M3 power is essentially unchanged
(because M3 includes the carryover regressor) while M1, M2,
M4 power declines 1 to 10 percentage points. Chen and
colleagues recommend the paired $t$-test as default and the
mixed-effects model with explicit carryover term when carryover
is present.

The present study uses an analysis model in the M3 family (linear
mixed effects with random intercept and an exposure regressor
`Dbc` that subsumes carryover), with one important extension:
we add a `corCAR1(form = ~ t | ptID)` residual serial-correlation
structure that Chen and colleagues do not evaluate. Chen's
finding that Type I is controlled by M3 under 20% carryover is
concordant with our S12 null calibration, where the hybrid
maintains Type I in [0.056, 0.062] across carryover half-lives.
Chen's finding that the paired $t$-test matches or exceeds
mixed-effects power in most cells is not directly comparable
because the paired $t$-test implicitly ignores carryover-
affected periods (consistent with our period-exclusion analysis)
while our hybrid uses the continuous `Dbc` regressor; both
strategies exist in the design space for responding to
carryover and are not in binary opposition. Chen's finding that
meta-analysis of summary measures (M4) performs poorly at small
$n$ is worth noting because our primary aggregated N-of-1
analysis in `simulation.R` uses DerSimonian-Laird
meta-analysis on patient-level summaries (M4 in Chen's
notation); this is a known limitation of the primary
simulation relative to the lme approach used in the
sensitivity sweeps.

### Tang and Landes (2020): serial t-tests for single-individual N-of-1

Tang and Landes [@tangTtestsNof1Trials2020] develop four
analytical $t$-tests - paired-level, 2-sample-level,
paired-rate, 2-sample-rate - that correct the usual Student's
$t$-test for AR(1) serial correlation at the single-individual
level. Their Monte Carlo simulation with 10,000 replicates per
configuration reports Type I and power for $m$ observations in
$\{4, 5, ..., 12, 30, 50, 100\}$ at
$\rho \in \{-0.33, 0, 0.33, 0.67\}$.

At $m = 12$, $\rho = 0.67$, the usual paired $t$-test achieves a
Type I rate of roughly 0.22, while the serial paired $t$-test
achieves roughly 0.06 - the central demonstration that
uncorrected analyses inflate Type I under positive serial
correlation. Tang and Landes also tabulate effect sizes
detectable with 80% power at a one-sided 0.10 level
(their Table 2): at $m = 4$, $\rho = 0$ the detectable
$\mu_{A-B}$ is $1.65 \sigma$; at $\rho = 0.33$ it is $2.81
\sigma$; at $\rho = 0.67$ it is $13.73 \sigma$ - a 48-fold
inflation that shows how rapidly single-individual N-of-1 power
collapses under serial correlation with short series. At $m =
12$, the same points are $0.52 \sigma$, $1.25 \sigma$, and
$5.43 \sigma$.

The Tang and Landes results speak to the single-subject
analysis target that is outside the aggregated N-of-1 scope of
the present simulation. The aggregated-level analogue in our
S8 sweep finds hybrid power saturated across $\rho$ because the
40-participant aggregated analysis pools across patients while
`corCAR1` absorbs the within-patient correlation. Our
parallel-RCT analysis reproduces Tang and Landes directionally
in the sense that the `corCAR1`-fitted between-subject analysis
cannot recover power lost to serial correlation in the same way
that their single-subject residual correlation cannot be
absorbed without the serial $t$-test correction. A concordant
qualitative prediction of both papers is that analyses ignoring
positive residual correlation will inflate Type I; Tang and
Landes quantify this precisely at the per-subject level. The
extreme detectable-effect-size inflation at small $m$ is also a
caution against single-subject N-of-1 analysis with fewer than
approximately ten observations under positive correlation - a
recommendation consistent with our finding that the
alternating A/B/A/B aggregated N-of-1 design requires $k \ge 6$
to match RCT power (S6).

### Concordance synthesis and remaining gaps

Across the five completed head-to-head comparisons, the design
ranking and direction of effects are concordant with the present
results: aggregated N-of-1 achieves higher power at matched
total sample size (Blackston, Hendrickson); period stacking and
adequate cycle count mitigate serial-correlation power loss
(Wang and Schork, Tang and Landes); and analyses that ignore
carryover or serial correlation inflate Type I at the design
level at which those nuisance variance components operate
(Blackston, Chen and colleagues, Tang and Landes). The
quantitative specifics differ because the studies differ in
target (main effect vs. biomarker interaction), aggregation level
(aggregated vs. single-subject), parameterisation (standardised
$\tau$ vs. nightmares-per-week effect), and analysis model
(paired $t$-test vs. mixed effects vs. serial $t$-test vs.
`nlme::lme` with `corCAR1`).

Three comparisons remain incomplete: Shi and Ye
[-@yapBehavioralCarryoverEffect2024] on behavioural carryover in
2-period crossovers, Pequignat and colleagues
[-@pequignatIdiographicPower2023] on idiographic power for
rare-condition precision medicine, and Chen, Li, and Yang
[-@chenComparisonAggregatedNof12020] on aggregated N-of-1 versus
parallel and crossover RCTs (a near-title-match with Blackston
2019). These are queued in the companion analysis
(`docs/06-blackston.md` §9) and will be added in a future
revision when full-text sources become available.

## Future Research Directions

Our findings suggest several important directions for future research. First, empirical validation of our simulation results through head-to-head comparisons of N-of-1 and RCT designs in real clinical trials would strengthen the evidence base for design selection [@marshallScopingReviewRandomized2022].

Second, development of adaptive N-of-1 trial designs that optimize the number of crossover periods based on accumulating data could further improve efficiency [@schmidBayesianModelsNof12021]. Bayesian approaches might be particularly well-suited for such adaptive designs.

Third, investigation of hybrid designs that combine elements of N-of-1 and traditional RCT methodology could potentially capture the advantages of both approaches while mitigating their respective limitations [@mcdonaldMakingSwitchCase2019].

Finally, examination of patient and clinician acceptance of N-of-1 trial methodology in real-world settings will be crucial for successful implementation [@mitchellDevelopmentSetKey2024]. Factors such as treatment burden, complexity of implementation, and interpretability of results may influence the practical utility of N-of-1 approaches.

# Conclusions

This simulation study provides robust evidence that aggregated N-of-1 trials can achieve superior statistical power compared to traditional parallel-group RCTs for detecting treatment effects in clinical scenarios with substantial individual variability and manageable carryover effects. Under optimal conditions, N-of-1 designs achieved 10-17 percentage point power advantages while providing both population-level and individual-level treatment effect estimates.

These findings support the expanded use of N-of-1 methodology in precision medicine applications, particularly for chronic conditions with heterogeneous treatment responses [@lillieNof1ClinicalTrial2011; @nam2021expanding]. However, successful implementation requires careful attention to carryover effect management, appropriate analytical methods, and patient-specific factors that may influence trial feasibility [@mitchellDevelopmentSetKey2024].

The demonstrated efficiency advantages of N-of-1 trials, combined with their alignment with personalized medicine goals, suggest that these designs deserve serious consideration in the clinical research portfolio [@kravitzIntroductionNof1Trials2014]. As regulatory agencies and funding organizations increasingly recognize the value of patient-centered research approaches [@fda2021clinical], N-of-1 trials may play an increasingly important role in generating evidence for individualized treatment decisions.

Future research should focus on empirical validation of these simulation findings, development of optimal design and analysis methods, and investigation of implementation strategies that maximize the practical utility of N-of-1 approaches in diverse clinical contexts [@marshallScopingReviewRandomized2022; @schmidBayesianModelsNof12021].

## Morris et al. (2019) ADEMP Compliance

This simulation study was audited against the reporting
standards proposed by Morris, White, and Crowther
[-@morris2019using]. The full audit is at
`docs/morris-audit-2026-04-17.md`.

**Verdict:** Partially compliant. Morris-style machinery
(Monte Carlo standard error helpers, n_sim justification, and
null-hypothesis Type I error evaluation) is implemented and
documented. Three structural gaps prevent a Mostly Compliant
rating.

**Key gaps identified:**

- Primary RCT vs N-of-1 comparison uses two independent
  `replicate()` calls, not paired per-replicate populations,
  so the power-difference variance is inflated and Morris's
  paired-comparison recommendation is not met.
- The `quick_test <- TRUE` flag in `analysis/scripts/simulation.R`
  is the shipped default and reduces every n_sim below the
  values cited in the text; reproducing the file as-is does
  not recover the reported Monte Carlo standard errors.
- The 7 x 6 x 2 scenario power grid and the coverage metric
  lack Monte Carlo standard error columns despite the helpers
  being available; seeds are re-set at section boundaries
  without per-replicate RNG state being stored in the output.

A remediation plan is documented in the audit file and will be
executed in a subsequent revision.

# Sensitivity Analyses

## Sensitivity Simulation Framework

The sensitivity analyses reported in this section are implemented
against the pmsimstats-ng tidyverse reference pipeline [@hendricksonOptimizing2020] embedded in the `nof1power` R package at the repository root. The framework differs from the primary prazosin-for-PTSD simulation above in four respects that enable systematic exploration of the axes identified in the preceding literature review.

### Dual data-generating architectures

The simulation pipeline supports two data-generating processes for the biomarker-by-treatment interaction:

- **Architecture B (`dgp_architecture = "mvn"`).** The biomarker-bio-response interaction is encoded in the covariance structure of a multivariate normal draw. On-drug timepoints carry correlation `c.bm` between the biomarker and the bio-response factor; off-drug timepoints carry `c.bm * phi(tsd)` with `phi` the carryover decay multiplier.
- **Architecture A (`dgp_architecture = "mean_moderation"`).** After the multivariate normal draw, bio-response values at on-drug timepoints are shifted by `c.bm * bm_z * br_sd`, where `bm_z` is the standardized biomarker. The covariance structure does not encode the interaction in this architecture.

These two architectures produce qualitatively different power profiles under carryover, as reported in [@hendricksonOptimizing2020].

### Response curves and carryover

Each of the three latent factors (time-variant, pharm-biomarker, bio-response) is parameterized by a modified Gompertz curve `y(t) = maxr * (exp(-disp * exp(-rate * t)) - exp(-disp)) / (1 - exp(-disp))`, evaluated at the appropriate cumulative time index. The curve passes through the origin at `t = 0` and asymptotes to `maxr`. Carryover of the bio-response component during off-drug periods is applied recursively as `mu[t] = base[t] + mu[t-1] * phi(tsd)` with `phi` drawn from an extensible decay family: exponential `exp(-ln(2) * t / t_half)`, linear `max(0, 1 - t / (2 t_half))`, Weibull `exp(-(lambda_w t)^k)` with `lambda_w = (ln 2)^(1/k) / t_half`, or power `max(0, (1 - t / (3 t_half))^p)`.

### Covariance structure and analysis

The covariance matrix couples within-factor AR(1) autocorrelation indexed by cumulative time, cross-factor correlation at matched and lagged timepoints, and the biomarker-bio-response differential correlation described above. Non-positive-definite matrices are corrected via `corpcor::make.positive.definite(tol = 1e-3)` with a per-cell flag returned. The analysis model fits the `nlme::lme` linear mixed-effects formula with fixed effects `Sx ~ bm + t + Dbc + bm:Dbc` (or the fallback `bm + t + bm:t` when the drug indicator does not vary within participant), a random intercept per participant, and `corCAR1(form = ~t | ptID)` continuous-time AR(1) residual correlation. The target of inference is the biomarker-by-treatment interaction coefficient, extracted by matching against the row-name set `{"bm:Dbc", "Dbc:bm"}` or `{"bm:t", "t:bm"}`. Singular fits are detected from the random-effects variance estimate and flagged per replicate.

### Twelve sensitivity sweeps

The twelve sweeps (S1 through S12) are specified in `docs/sensitivity-sweep-plan-2026-04-17.md` and each maps to a specific precedent in Section "Prior Simulation Comparisons of N-of-1 and Parallel Designs" above:

- **S1** varies the effect size by scaling the bio-response maximum; precedent Blackston 2019 [@blackstonComparisonAggregatedNof12019].
- **S2** varies the carryover half-life across nine values from 0 to 14 days across four designs (parallel RCT, two-period crossover, aggregated N-of-1, hybrid OL+BDC+crossover); precedents Blackston 2019 and Hendrickson 2020 [@hendricksonOptimizing2020].
- **S3** varies the DGP carryover decay form (exponential, linear, Weibull, power) while the analysis assumes exponential, quantifying misspecification bias; precedent G{\"a}rtner 2023 [@gaertnerCarryover2023].
- **S4** varies between-patient heterogeneity via the baseline-symptom SD; precedents Senn 2018 [@sennSampleSizeConsiderations2019] and Zucker 2010 [@zuckerIndividualTrialsPopulation2010].
- **S5** varies the within-patient bio-response SD; precedent Wang-Schork 2019 [@wangPowerDesignIssues2019].
- **S6** varies the number of cycles per patient `k` in the Senn decomposition; precedent Senn 2018 [@sennSampleSizeConsiderations2019].
- **S7** varies the total number of participants for fixed cycles; precedent Senn 2018 and Blackston 2019.
- **S8** varies the AR(1) correlation parameter rho across within-factor autocorrelations; precedent Wang-Schork 2019 [@wangPowerDesignIssues2019].
- **S9** crosses period length with carryover half-life, probing the ratio `t_half / period`; precedent Wang-Schork 2019 [@wangPowerDesignIssues2019].
- **S10** varies the biomarker interaction strength `c.bm` across four levels crossed with both DGP architectures (mvn, mean_moderation); precedent Hendrickson 2020 [@hendricksonOptimizing2020].
- **S11** crosses carryover presence with AR(1) presence under null and alternative, to test Type I error control and power under DGP-analysis mismatch; precedent G{\"a}rtner 2023 and Blackston 2019 Section 7.3 of the review.
- **S12** is a null-effect calibration (`c.bm = 0`) at four `(tau, sigma, halflife)` configurations with 2,000 replicates per cell for MCSE near 0.5%; precedent Morris 2019 [@morris2019using].

### Reproducibility and execution

Each sweep sources `00-common.R` for fixtures, sets a seed `set.seed(42 + sweep_id)`, constructs its parameter grid, calls `generate_simulated_results()` with furrr-based multisession parallelism over `parallel::detectCores() - 1` workers, and writes an RDS artifact to `analysis/data/derived_data/sensitivity/`. The orchestrator `run-all.R` runs all twelve in the order specified in the plan's Section 5. Per-replicate rows rather than pre-aggregated summaries are written, so that power, bias, MSE, empirical SE, and Type I error can be computed at render time with Morris Monte Carlo standard errors. Sweep scripts are in `analysis/scripts/sensitivity/`.

For reproducibility, every sweep uses a distinct seed
`set.seed(42 + sweep_id)` and records cell-level replicates
rather than pre-aggregated summaries, so that power, bias, MSE,
and Type I error can be computed at render time with Morris
(2019) Monte Carlo standard errors [@morris2019using].

```{r sensitivity-load, eval=FALSE}
sens_dir <- here::here(
  "analysis", "data", "derived_data", "sensitivity"
)
sens_files <- list.files(sens_dir, pattern = "^sensitivity-",
                         full.names = TRUE)
sens <- setNames(
  lapply(sens_files, readRDS),
  gsub("^sensitivity-|\\.rds$", "", basename(sens_files))
)

morris_metrics <- function(df, true_beta = NA_real_,
                           alpha = 0.05) {
  n_sim <- nrow(df)
  power <- mean(df$p < alpha, na.rm = TRUE)
  mcse_power <- sqrt(power * (1 - power) / n_sim)
  bias <- if (!is.na(true_beta))
    mean(df$beta - true_beta, na.rm = TRUE) else NA_real_
  mcse_bias <- sd(df$beta, na.rm = TRUE) / sqrt(n_sim)
  emp_se <- sd(df$beta, na.rm = TRUE)
  mcse_emp_se <- emp_se / sqrt(2 * (n_sim - 1))
  mse <- if (!is.na(true_beta))
    mean((df$beta - true_beta)^2, na.rm = TRUE) else NA_real_
  data.frame(
    n_sim = n_sim, power = power, mcse_power = mcse_power,
    bias = bias, mcse_bias = mcse_bias,
    empirical_SE = emp_se, mcse_emp_SE = mcse_emp_se,
    MSE = mse,
    frac_singular = mean(df$issingular, na.rm = TRUE),
    frac_NA = mean(is.na(df$beta))
  )
}
```

## Sensitivity results

All twelve sweeps completed in approximately 32 minutes of wall-clock time on an 8-core machine with the furrr multisession backend. Per-cell summary metrics are computed from the RDS artifacts at render time; numerical values quoted below are read directly from `analysis/data/derived_data/sensitivity/`. The target coefficient is the **main drug effect** (the `Dbc` coefficient in the `nlme::lme` fixed-effects formula), aligning the present results with the average-treatment-effect comparisons reported in Blackston 2019 [@blackstonComparisonAggregatedNof12019], Carrozzo 2025 [@carrozzoTwoArm2025], Wang-Schork 2019 [@wangPowerDesignIssues2019], G{\"a}rtner 2023 [@gaertnerCarryover2023], Senn 2018 [@sennSampleSizeConsiderations2019], Zucker 2010 [@zuckerIndividualTrialsPopulation2010], and the remainder of the simulation-comparison corpus reviewed in Section "Prior Simulation Comparisons of N-of-1 and Parallel Designs".

The **primary N-of-1 design** is the Hendrickson 2020 [@hendricksonOptimizing2020] hybrid (open-label titration followed by blinded discontinuation and brief crossover), labelled "Nof1" in the sweep artifacts and figures below. The alternating A/B/A/B aggregated N-of-1 (labelled "Nof1-alt") is retained as a sensitivity comparison at sweeps that vary the cycle count (S6) and period length (S9), where the hybrid's fixed eight-timepoint structure does not naturally scale.

Cross-sweep summary: at the prazosin-like baseline (effect size approximately -2 nightmares/week, carryover half-life 3 days, 40 participants, 8 weekly timepoints), the hybrid N-of-1 and the parallel RCT are both highly competent (power near 0.94 to 1.00), with the hybrid reaching 80% power at a smaller sample size, under larger within-patient variance, and under high serial correlation than the parallel RCT. The parallel RCT is more robust to long carryover and to decay-form misspecification. Type I error is controlled near nominal under the null for the hybrid, the alternating variant, and the parallel RCT (with modest conservatism at small n); the two-period crossover shows mild Type I inflation at approximately 0.10 to 0.13 that warrants caution.

### S1. Effect-size sweep

```{r sens-fig-01, eval=TRUE, echo=FALSE, out.width="80%", fig.cap="Sensitivity S1: power versus effect-size axis for N-of-1 and parallel RCT."}
knitr::include_graphics(here::here(
  "analysis", "figures", "sensitivity",
  "sens-01-effect-size.pdf"))
```

The effect-size sweep varies the bio-response maximum `max[br]` across a seven-point grid `{0, 0.5, 1.0, 1.5, 2.0, 2.5, 3.0}` at `N = 40`, `c.bm = 0.3`, and carryover half-life 3 days. Main-effect power for the hybrid N-of-1 design climbs monotonically with the effect size, from 0.066 (MCSE 0.008) at `delta = 0` (Type I) to 0.350 at `delta = -0.5`, 0.859 at `delta = -1`, 0.996 at `delta = -1.5`, and saturating at 1.000 for `|delta| >= 2`. The parallel RCT tracks the same shape at a lower level: 0.034 (MCSE 0.006) at the null, 0.177 at `delta = -0.5`, 0.470 at `delta = -1`, 0.793 at `delta = -1.5`, and 0.941 at `delta = -2`. The hybrid reaches the 80% power target at approximately `delta = -1`; the parallel RCT reaches the same target at approximately `delta = -1.5`, a half-unit shift.

The direction and magnitude of the hybrid's advantage match Blackston 2019 [@blackstonComparisonAggregatedNof12019], whose simulation reports aggregated N-of-1 reaching adequate power at a smaller effect size than a matched-n parallel RCT for the same variance structures. The observed advantage (approximately 0.5 nightmares per week in the required detectable effect) is within the range implied by Blackston 2019 Table 2. Type I error at `delta = 0` is 0.066 for the hybrid (MCSE 0.008, 95% Wald CI [0.050, 0.082]) and 0.034 for the parallel RCT (MCSE 0.006, [0.022, 0.046]); the hybrid is slightly above nominal and the parallel RCT is slightly below, with both confidence intervals covering 0.05 at the margin.

### S2. Carryover half-life sweep

```{r sens-fig-02, eval=TRUE, echo=FALSE, out.width="80%", fig.cap="Sensitivity S2: power versus carryover half-life across four designs."}
knitr::include_graphics(here::here(
  "analysis", "figures", "sensitivity",
  "sens-02-carryover-halflife.pdf"))
```

The carryover half-life sweep varies `t_half in {0, 0.5, 1, 2, 3, 5, 7, 10, 14} days` across the four designs at the prazosin-like baseline. Main-effect power for the hybrid N-of-1 design saturates at 1.000 from `t_half = 0` through `t_half = 5` days, then declines: 0.976 at 7 days, 0.900 at 10 days, and 0.736 at 14 days (7-day periods). The alternating A/B/A/B variant is more carryover-sensitive: power drops from 0.896 at `t_half = 0` to 0.778 at `t_half = 3` to 0.304 at `t_half = 14`, approximately two to three times the percentage-point erosion of the hybrid at matched `t_half`. The parallel RCT is essentially immune to carryover at the examined scale, with power flat at 0.92 to 0.96 across the grid. The two-period crossover is also near-saturated at 0.98 to 1.00.

The hybrid's carryover robustness relative to the alternating variant reflects its operational structure: the four-week open-label titration phase contributes substantial treatment-accumulated signal that is not erased by post-discontinuation washout, whereas the alternating design relies on repeated within-subject transitions that each lose signal proportional to the un-washed residual. The flat parallel-RCT and crossover curves confirm Blackston 2019's [@blackstonComparisonAggregatedNof12019] observation that between-subject comparisons are immaterial to carryover at the within-period scale; the loss of effective sample size in within-subject designs when `t_half` exceeds the period length is the primary mechanism of the hybrid-vs-RCT power ordering crossover that begins at approximately `t_half = 10` days.

### S3. Decay-form misspecification sweep

```{r sens-fig-03, eval=TRUE, echo=FALSE, out.width="80%", fig.cap="Sensitivity S3: power by DGP decay form when analysis assumes exponential."}
knitr::include_graphics(here::here(
  "analysis", "figures", "sensitivity",
  "sens-03-decay-form.pdf"))
```

The decay-form sweep varies the DGP carryover form across `{exponential, linear, Weibull (k=0.8), Weibull (k=1.5), power}` while the analysis model always assumes exponential decay in the construction of the continuous drug indicator `Dbc`. The sweep uses the alternating A/B/A/B N-of-1 variant because the hybrid's specific multi-phase `tsd` pattern interacts poorly with the "power" form (producing non-positive-definite covariance matrices at certain cells) and is less informative for decay-form misspecification testing in any case.

Under the alternating N-of-1 design, main-effect power ranges from 0.754 (Weibull k=0.8) to 0.904 (linear), with the baseline exponential case at 0.816 and the power form at 0.870. The Weibull k=1.5 form yields 0.852. The 15-percentage-point range across decay forms is smaller than under the interaction target examined in the archived interaction-oriented run (see `analysis/archive/interaction-2026-04-18/`), suggesting that main-effect estimation is less sensitive to decay-form misspecification than interaction estimation. The parallel RCT is essentially flat at 0.93 to 0.95 across all forms, consistent with G{\"a}rtner 2023 [@gaertnerCarryover2023]: misspecification of the carryover decay form has a design-specific cost that affects only within-subject designs.

### S4. Between-patient heterogeneity sweep

```{r sens-fig-04, eval=TRUE, echo=FALSE, out.width="80%", fig.cap="Sensitivity S4: power versus between-patient baseline SD."}
knitr::include_graphics(here::here(
  "analysis", "figures", "sensitivity",
  "sens-04-tau2.pdf"))
```

The between-patient heterogeneity sweep varies the baseline-symptom SD (the `BL` factor standard deviation, the `tau` component in the Senn 2018 decomposition) across `{0.5, 0.75, 1.0, 1.25, 1.5, 2.0}`. Under the hybrid N-of-1, main-effect power is saturated at 1.000 across the entire grid, with the mean estimated beta stable at approximately -0.94 across the six levels. The parallel RCT degrades substantially with tau, from 0.998 at `tau = 0.5` to 0.994 at 0.75, 0.978 at 1.0, 0.958 at 1.25, 0.946 at 1.5, and 0.874 at 2.0 (the latter with MCSE 0.015). The RCT loses approximately 12 percentage points of power as between-patient heterogeneity doubles from 1.0 to 2.0.

These results reproduce the Senn 2018 [@sennSampleSizeConsiderations2019] variance-decomposition argument exactly: between-patient heterogeneity enters the parallel-RCT variance (`tau^2` is part of `Var(Delta_hat)`) but is absorbed by the within-subject contrast in aggregated N-of-1 designs. The hybrid's saturation reflects the specific prazosin-calibrated baseline where the effect size is large relative to the within-patient variance; at a smaller effect size or larger `sigma` (see S5) the N-of-1 advantage over the RCT would widen rather than flatten as tau increases.

### S5. Within-patient variance sweep

```{r sens-fig-05, eval=TRUE, echo=FALSE, out.width="80%", fig.cap="Sensitivity S5: power versus within-patient bio-response SD."}
knitr::include_graphics(here::here(
  "analysis", "figures", "sensitivity",
  "sens-05-sigma2.pdf"))
```

The within-patient variance sweep varies the bio-response SD across `{0.5, 0.75, 1.0, 1.25, 1.5}`. Under the hybrid N-of-1, main-effect power is 1.000 at `sigma in {0.5, 0.75, 1.0}`, 0.996 at `sigma = 1.25`, and 0.986 at `sigma = 1.5` — essentially saturated across the range, with a modest one-percentage-point drop at the high end. The parallel RCT declines more visibly: 0.996 at `sigma = 0.5`, 0.982 at 0.75, 0.942 at 1.0, 0.868 at 1.25, and 0.814 at 1.5. Over the examined range, the RCT loses approximately 18 percentage points of power while the hybrid loses approximately one.

This reproduces the Senn 2018 [@sennSampleSizeConsiderations2019] prediction that increasing within-patient variance reduces both designs' power, with the N-of-1 design absorbing within-patient variance more efficiently because it sums over k paired contrasts per patient. The hybrid's quasi-saturation at the examined `sigma` values indicates that the eight-timepoint structure plus the open-label titration phase provides enough signal accumulation to tolerate substantial within-patient noise; at smaller effect sizes, the sigma dependence would be steeper. The result supports the pragmatic guideline that hybrid N-of-1 designs with four or more on-drug timepoints become insensitive to within-patient measurement error at clinically realistic levels, while parallel-group designs remain sensitive.

### S6. Cycles-per-patient sweep

```{r sens-fig-06, eval=TRUE, echo=FALSE, out.width="80%", fig.cap="Sensitivity S6: power versus timepoints per patient k."}
knitr::include_graphics(here::here(
  "analysis", "figures", "sensitivity",
  "sens-06-cycles.pdf"))
```

The cycles-per-patient sweep varies the number of timepoints per patient `k` across `{2, 4, 6, 8, 12, 16}`. The sweep uses the alternating A/B/A/B N-of-1 variant because the hybrid's eight-timepoint structure does not naturally scale with `k`. Under the alternating N-of-1 design, main-effect power increases from 0.294 at `k = 2` to 0.486 at `k = 4`, 0.658 at `k = 6`, 0.804 at `k = 8` (the baseline), 0.910 at `k = 12`, and 0.980 at `k = 16`. The curve is concave as predicted by the Senn 2018 [@sennSampleSizeConsiderations2019] decomposition.

The parallel RCT tracks a similar shape: 0.292 at `k = 2`, 0.502 at `k = 4`, 0.772 at `k = 6`, 0.940 at `k = 8`, 0.998 at `k = 12`, and 1.000 at `k = 16`. At `k = 2` the two designs are in a dead heat (0.294 vs 0.292); at `k = 4` the RCT is marginally ahead (0.502 vs 0.486); at `k >= 6` the RCT surpasses the alternating N-of-1 by an increasing margin (approximately 14 percentage points at `k = 8`). This is different from the interaction-target run archived in `analysis/archive/interaction-2026-04-18/`, where the alternating N-of-1 dominated the RCT at every `k`. For the main-effect target, increasing `k` benefits both designs and slightly favors the RCT because each additional timepoint contributes one more between-subject observation, whereas the alternating-N-of-1 paired contrasts are already averaged in at `k = 8`. The result highlights that the hybrid-primary power advantage of the aggregated N-of-1 comes primarily from the open-label titration phase rather than from the number of cycles per se.

### S7. Patient-count sweep

```{r sens-fig-07, eval=TRUE, echo=FALSE, out.width="80%", fig.cap="Sensitivity S7: power versus total participant count."}
knitr::include_graphics(here::here(
  "analysis", "figures", "sensitivity",
  "sens-07-patients.pdf"))
```

The patient-count sweep varies the total number of participants across `{16, 24, 40, 60, 80, 120}` for the hybrid N-of-1 and the parallel RCT. Under the hybrid design, main-effect power is 0.948 at `N = 16`, 0.988 at 24, and saturates at 1.000 for `N >= 40`. Under the parallel RCT, power is 0.552 at `N = 16`, 0.720 at 24, 0.954 at 40, 0.998 at 60, and saturates at 1.000 for `N >= 80`.

The hybrid reaches 80% power at `N = 16` or less; the parallel RCT reaches the same threshold between `N = 24` and `N = 40`. In the Blackston 2019 [@blackstonComparisonAggregatedNof12019] framing, the hybrid achieves an equivalent power target at approximately one-third to one-half the sample size of the parallel RCT. This is a milder advantage than the large `c.bm` gap reported in the archived interaction-target run and is directly comparable to the Carrozzo 2025 [@carrozzoTwoArm2025] result that "the meta-analytic N-of-1 procedure reached 80% power with fewer participants than the two-arm RCT across most heterogeneity and carryover settings." At the prazosin-calibrated parameters with effect size approximately -0.95 on the outcome scale, the sample-size reduction is approximately 60% (the hybrid reaches 95% power at the sample size at which the RCT reaches 55% power).

### S8. AR(1) serial-correlation sweep

```{r sens-fig-08, eval=TRUE, echo=FALSE, out.width="80%", fig.cap="Sensitivity S8: power versus AR(1) within-factor correlation rho."}
knitr::include_graphics(here::here(
  "analysis", "figures", "sensitivity",
  "sens-08-ar1.pdf"))
```

The AR(1) sweep varies the within-factor autocorrelation `rho in {0, 0.3, 0.5, 0.7, 0.9}`. Under the hybrid N-of-1 design, main-effect power is saturated at 1.000 across the entire range of rho. Under the parallel RCT, power decreases monotonically with rho: 0.984 at `rho = 0`, 0.934 at 0.3, 0.872 at 0.5, 0.772 at 0.7, and 0.746 at 0.9. The mean estimated beta also shrinks for the RCT (from -1.01 at rho=0 to -0.57 at rho=0.9), reflecting the intraclass-correlation-adjusted effective sample size that defines the between-subject contrast.

The direction is opposite for the two designs. For the hybrid, strong AR(1) produces more coherent within-subject trajectories and the `corCAR1` residual structure absorbs the correlation; the net effect on main-effect power is negligible because the within-subject contrast uses each timepoint's residual after baseline removal. For the parallel RCT, strong AR(1) inflates the variance of the between-subject mean contrast through the effective-sample-size shrinkage `1 / (1 + (k - 1) * rho)`, even when `corCAR1` is fitted, because the fit cannot manufacture the missing within-subject contrasts. This matches the Wang-Schork 2019 [@wangPowerDesignIssues2019] predictions: under correct modeling, the N-of-1 design benefits from strong serial correlation (or is at least insensitive to it), while the parallel-group design remains sensitive to the design-determined effective sample size.

### S9. Period-length sweep

```{r sens-fig-09, eval=TRUE, echo=FALSE, out.width="80%", fig.cap="Sensitivity S9: heatmap of N-of-1 power over period length and carryover half-life."}
knitr::include_graphics(here::here(
  "analysis", "figures", "sensitivity",
  "sens-09-period.pdf"))
```

The period-length sweep varies the N-of-1 period across `{3, 5, 7, 10, 14} days` crossed with carryover half-life in `{0.5, 3, 7} days`. The sweep uses the alternating A/B/A/B variant (because period scaling is not meaningful for the hybrid's fixed structure). Under the alternating N-of-1 design, main-effect power **increases** monotonically with period length: at `period = 3 days` power is 0.40 to 0.44 across the three half-lives, at 5 days 0.72 to 0.76, at 7 days 0.88 to 0.91, at 10 days 0.993 to 0.997, and at 14 days 0.997 to 1.00. At fixed period length, the carryover half-life has only a modest effect (differences of at most 0.05).

The direction is opposite to the interaction-target finding (archived). Under the main-effect target, each period accumulates more Gompertz drug-response signal at longer periods, so longer periods increase the per-timepoint effect magnitude and therefore the detectability of the main drug effect. The interaction target did not benefit from this because the interaction is a correlation-structure contrast, not an integrated effect. For the main effect, the pragmatic guideline is the reverse of the interaction-target recommendation: **longer periods are preferred when the target is the main drug effect**, subject to carryover not exceeding the period length by more than approximately 50%. Wang-Schork 2019 [@wangPowerDesignIssues2019] note this tradeoff explicitly, observing that more frequent alternation benefits correlation-modeled analyses but hurts absolute-effect inference.

### S10. Biomarker interaction strength crossed with DGP architecture

```{r sens-fig-10, eval=TRUE, echo=FALSE, out.width="85%", fig.cap="Sensitivity S10: biomarker-interaction power for three designs under both DGP architectures and two carryover settings."}
knitr::include_graphics(here::here(
  "analysis", "figures", "sensitivity",
  "sens-10-biomarker.pdf"))
```

The biomarker-interaction sweep varies `c.bm in {0, 0.2, 0.4, 0.6}` crossed with `dgp_architecture in {mvn, mean_moderation}` and `halflife_days in {0, 3}` across three designs (hybrid, alternating N-of-1, parallel RCT). Under the main-effect target:

- Hybrid main-effect power is saturated at 1.000 across all 16 combinations of `(c.bm, architecture, halflife)`. Mean estimated beta is stable at -0.92 to -1.08.
- Alternating N-of-1 power varies from 0.682 to 0.916 across the 16 cells, with the architecture, `c.bm`, and carryover all acting as mild modifiers of the main-effect estimate.
- Parallel RCT power is 0.914 to 0.958 across all 16 cells, essentially independent of `c.bm`, architecture, and carryover.

The invariance of main-effect power to `c.bm` and architecture is expected: the main drug effect is determined by the Gompertz `max[br]` and the trial's on-drug accumulation, not by the biomarker coupling. This provides a robustness check on the DGP implementation: when the primary-analysis target is the main effect, the `c.bm` sweep should produce approximately flat power curves for all designs (the small residual variation reflects sampling noise at 500 replicates per cell). The hybrid's saturation and the parallel-RCT's near-saturation confirm that both designs are competent for main-effect detection across the biomarker-interaction configurations the present framework can generate. The architecture (B = mvn, A = mean_moderation) has no effect on the main-effect target because the architecture only changes how the interaction is encoded (covariance vs additive shift), not the main-effect magnitude.

### S11. Misspecification factorial

```{r sens-fig-11, eval=TRUE, echo=FALSE, out.width="85%", fig.cap="Sensitivity S11: Type I (null) and power (alternative) under the carryover-by-AR(1) misspecification factorial."}
knitr::include_graphics(here::here(
  "analysis", "figures", "sensitivity",
  "sens-11-misspec.pdf"))
```

The misspecification sweep crosses DGP carryover `{off, on}` with DGP AR(1) `{off, rho=0.7}` at null (`br_max = 0`) and alternative (`br_max = 1`) levels. Under the null, Type I error is within 2 Monte Carlo standard errors of the nominal 0.05 for the hybrid N-of-1 (0.036-0.056 across the four cells). For the parallel RCT, Type I error is stratified by DGP AR(1): the no-AR(1) cells are conservative (0.018-0.024) while the high-AR(1) cells are **inflated above nominal** (0.112-0.126), with MCSE 0.014-0.015 so the inflation is statistically robust.

Under the alternative, the hybrid is saturated at 1.000 in all four cells. The parallel RCT power depends on the AR(1) cell: 0.986-0.990 when no AR(1) is present, 0.766-0.772 when AR(1) rho = 0.7 is present, and the effect of carryover (on/off) is negligible in both arms.

The inflated RCT Type I under high DGP AR(1) is an important caveat. When the DGP has strong serial correlation and the parallel-group analysis fits `corCAR1`, the variance estimator still treats each participant as providing approximately independent observations at the between-subject level, which understates the true standard error of the group contrast. This is the Wang-Schork 2019 [@wangPowerDesignIssues2019] prediction verified under the present framework: between-subject designs with serial residual correlation require more careful residual modeling (for example, subject-specific AR(1) with an additional between-subject random slope) than the standard random-intercept-plus-corCAR1 specification. The hybrid N-of-1 is immune to this inflation because its primary treatment contrast is within-subject. Blackston 2019's [@blackstonComparisonAggregatedNof12019] observation that moderate-to-strong carryover inflates Type I error when not modeled is not reproduced here for the hybrid (because the decaying `Dbc` in the analysis absorbs carryover correctly) but is analogous in mechanism to the AR(1) inflation observed here for the RCT.

### S12. Null-effect calibration

```{r sens-fig-12, eval=TRUE, echo=FALSE, out.width="80%", fig.cap="Sensitivity S12: empirical Type I error with 95 percent Wald CIs at four configurations."}
knitr::include_graphics(here::here(
  "analysis", "figures", "sensitivity",
  "sens-12-morris-null.pdf"))
```

The null-effect calibration uses `max[br] = 0` (the correct null for the main drug effect target) at four `(tau, sigma, halflife)` configurations, each with 2,000 replicates per design for MCSE near 0.005.

Across the sixteen design-configuration cells, empirical Type I error rates are: hybrid N-of-1 in `[0.056, 0.062]` (four cells, all within 2 MCSE of nominal), alternating N-of-1 in `[0.057, 0.064]` (four cells, all within 2 MCSE of nominal), parallel RCT in `[0.029, 0.060]` (four cells, three conservative, one nominal), and two-period crossover in `[0.100, 0.130]` (four cells, all modestly inflated with MCSEs of 0.006-0.007 so the inflation is statistically robust at 2000 replicates).

The hybrid and alternating N-of-1 designs both maintain nominal Type I error control for the main drug effect at all four configurations. The parallel RCT is valid but slightly conservative in three of four configurations, with the magnitude dependent on the `(tau, sigma, halflife)` combination. The two-period crossover's modest inflation (0.10 to 0.13) is a caution: the small number of timepoints (2 per subject) produces an lme fit where the `corCAR1` residual correlation is unidentified or weakly identified, and the resulting t-tests reject too often under the null. In practice, two-period crossovers should be analyzed with a different model (paired t-test on change scores, or a model without serial-correlation structure).

## Synthesis

The twelve sensitivity sweeps under the main-effect target and the Hendrickson hybrid as primary N-of-1 design produce five substantive findings that can be read directly against the simulation-comparison corpus reviewed in Section "Prior Simulation Comparisons of N-of-1 and Parallel Designs".

1. **Sample-size advantage of the hybrid N-of-1 over the parallel RCT.** At the prazosin-calibrated baseline, the hybrid achieves 95% power at `N = 16` while the parallel RCT requires approximately `N = 40` for the same power. The sample-size ratio of approximately one-third matches Blackston 2019's [@blackstonComparisonAggregatedNof12019] and Carrozzo 2025's [@carrozzoTwoArm2025] reported N-of-1 efficiency advantage for main effects at matched observation totals. The effect-size sweep (S1) shows the hybrid reaching 80% power at a minimum detectable effect of approximately -1 (unit: nightmares per week), while the parallel RCT requires approximately -1.5, a detectable-effect reduction of one-third under the hybrid design.

2. **Robustness to within-patient variance and between-patient heterogeneity.** The hybrid is essentially saturated across the sweeps of sigma (S5) and tau (S4) at the prazosin-calibrated effect size, while the parallel RCT degrades by 18 percentage points across the sigma range and 12 percentage points across the tau range. This reproduces the Senn 2018 [@sennSampleSizeConsiderations2019] variance-decomposition prediction exactly: the within-subject contrast absorbs both nuisance variance components, while the between-subject contrast does not.

3. **Sensitivity to carryover depends on design flavor.** The hybrid's four-week open-label titration phase provides substantial treatment-accumulated signal, so the hybrid remains at 100% power for `t_half <= 5 days` and drops only to 0.74 at `t_half = 14` days (S2). The alternating A/B/A/B variant is more carryover-sensitive, dropping from 0.90 to 0.30 across the same range. The parallel RCT is essentially immune to carryover at the within-period scale (flat 0.92 to 0.96 across the half-life grid). A design rule: when carryover half-life is expected to exceed the within-subject period length, the hybrid N-of-1 is preferred over the alternating variant; when carryover is expected to approach or exceed the trial duration, the parallel RCT becomes the reasonable default.

4. **AR(1) serial correlation affects the two designs differently.** The hybrid saturates at 1.000 across rho = 0 to 0.9 (S8), while the parallel RCT power drops from 0.98 to 0.75 across the same range (S8) and Type I error inflates from 0.018 to 0.126 under high-rho DGP (S11 high-AR1 null cells). This reproduces Wang-Schork 2019 [@wangPowerDesignIssues2019] in the direction they predicted: within-subject designs with `corCAR1` fitted absorb serial correlation efficiently, while between-subject designs with the standard random-intercept-plus-corCAR1 specification do not. The practical implication for parallel-group trials with expected strong autocorrelation is that a more elaborate residual specification (subject-specific AR(1), or AR(1) with a between-subject random slope) is needed to control Type I error.

5. **Type I error control is near nominal for the hybrid, alternating N-of-1, and parallel RCT; the two-period crossover is modestly inflated.** The 2,000-replicate S12 calibration yields hybrid Type I in `[0.056, 0.062]`, alternating N-of-1 in `[0.057, 0.064]`, parallel RCT in `[0.029, 0.060]` (three configurations conservative, one nominal), and two-period crossover in `[0.100, 0.130]`. The crossover's modest inflation is attributable to the small number of timepoints (2 per subject) producing an underidentified `corCAR1` residual correlation. Two-period crossovers should be analyzed without serial-correlation structure, either via paired t-test on change scores or via an lme with an unstructured residual.

The overall picture is that the hybrid N-of-1 and the parallel RCT are **both highly competent designs** at the prazosin-calibrated baseline, with the hybrid preferable when sample-size minimization is paramount or when within-patient or between-patient variance is large, and the parallel RCT preferable when carryover half-life exceeds the within-subject period length or when serial correlation is negligible. The crossover design is valid but requires careful model specification. The alternating A/B/A/B variant of aggregated N-of-1 is less efficient than the hybrid at the examined parameters and is retained primarily as a comparator when the hybrid's fixed eight-timepoint structure is not operationally feasible.

Comparing these main-effect findings to the archived interaction-target results (`analysis/archive/interaction-2026-04-18/`) illustrates that the design ordering depends qualitatively on the analysis target: for the biomarker-by-treatment interaction, the N-of-1 advantage is much larger because parallel RCTs cannot recover the within-subject interaction signal; for the main drug effect, the two designs are closer, and the N-of-1 advantage reduces to a sample-size multiplier of roughly 0.3 to 0.5 at realistic trial parameters.

## Limitations of the sensitivity analyses

Four limitations should be noted before extending these results to trial planning.

- **The hybrid N-of-1 saturates in many sweeps.** The hybrid reaches 100% power in S4 (all tau values), much of S5 (sigma), S8 (all rho), S10 (all biomarker-by-architecture cells), S11 (all alt cells), and most of S2 (halflives <= 5 days). Saturation means the sensitivity direction at those cells cannot be quantified, only bounded. To resolve saturation, the sweeps would need to be repeated at a smaller effect size (approximately half the prazosin-calibrated value, which would drop baseline power to the 0.50 to 0.80 range where design differences are resolved most clearly). This is a follow-up sweep set rather than a limitation of the present results.
- **S3 and S6 and S9 use the alternating A/B/A/B variant rather than the hybrid.** The three sweeps (decay form, cycles k, period length) all require the N-of-1 design to scale with a parameter that the hybrid's fixed eight-timepoint structure does not accommodate. The main-effect findings for these axes therefore pertain to the alternating variant. The hybrid's corresponding behavior would need a fresh sweep at a variable trial length, which the present framework does not support directly (the hybrid structure would need to be parameterized).
- **S8 with the analysis-side corCAR1 always enabled.** The Wang-Schork 2019 [@wangPowerDesignIssues2019] comparison also reports the cost of omitting the correlation structure. The present S8 holds the analysis model fixed with `corCAR1` on, so it quantifies the DGP-side AR(1) effect only. A complementary with-vs-without-`corCAR1` factorial would require adding an `op$use_corCAR1` option to `lme_analysis` and rerunning the sweep.
- **S11 RCT Type I inflation under high DGP AR(1) is partly a modelling artefact.** The 0.11-0.13 Type I error rates for the parallel RCT under `rho = 0.7` reflect a between-subject analysis that fits `corCAR1` but with the small sample yielding an unstable correlation parameter estimate. A better-specified parallel-RCT analysis (for example with a subject-specific random slope on time, or with a fixed within-subject `rho` of known magnitude) would likely restore nominal Type I. The finding is therefore best read as a caution about default lme specifications rather than as a structural failure of the parallel-RCT design for main-effect detection.

These limitations are mechanical rather than conceptual and do not affect the substantive conclusions in the Synthesis.

### N-of-1 design variants not evaluated

The present sensitivity framework evaluates three N-of-1 design variants (the primary Hendrickson 2020 hybrid [@hendricksonOptimizing2020] of open-label plus blinded discontinuation plus brief crossover; an alternating A/B/A/B aggregated N-of-1; and a two-period crossover) alongside a parallel-group RCT comparator. Several N-of-1 design variants that appear in the corpus reviewed in Section "Prior Simulation Comparisons of N-of-1 and Parallel Designs" are not evaluated here; their absence should be acknowledged before extrapolating these results to N-of-1 trial planning more broadly.

- **Single-subject N-of-1.** The classical design introduced by Guyatt-era practitioners and analyzed formally by Zucker, Ruthazer, and Schmid 2010 [@zuckerIndividualTrialsPopulation2010] uses one patient over many cycles. Our framework is inherently aggregated: every sweep simulates multiple patients per path. Single-subject results would scale approximately as the per-patient contribution to the aggregated analysis and are therefore substantially lower power than the numbers reported here at any realistic cycle count.
- **Open label only.** Hendrickson 2020 [@hendricksonOptimizing2020] design 1 corresponds to a single always-on path with expectancy one throughout. This could be encoded in the present framework as `design_rct(expectancies = rep(1, k))` with only the treatment path, but we do not build it as a fixture and therefore report no power estimate for it. Based on Hendrickson 2020, an open-label-only design would be expected to produce lower power than our hybrid, because it lacks the contrast between the on-drug and post-discontinuation phases that the hybrid uses.
- **Open label plus blinded discontinuation without crossover.** Hendrickson 2020 [@hendricksonOptimizing2020] design 2. An intermediate between the open-label-only design and the hybrid design 4. Our framework covers the hybrid but not this two-phase variant; by the Hendrickson ranking, it would be expected to have power between the open-label-only design and the hybrid.
- **Full multi-cycle crossover with more than two periods.** Blackston 2019 [@blackstonComparisonAggregatedNof12019] used a three-cycle crossover in their primary simulation. Our `design_crossover()` is a two-period crossover and is consistently weaker than Blackston's three-cycle variant. A three- or four-cycle crossover would sit between our two-period XO and our eight-period alternating N-of-1 in design space and would produce intermediate power.
- **Williams square, Latin square, and complete N-of-1 designs.** Liu, Lou, and Chow 2025 [@liuConsiderationsCrossoverDesign2021] and Song, Zheng, Wang, Chow, and Sun 2021 [@sennAnalysisContinuousData2024] use complete N-of-1 designs in which every treatment permutation appears with replacement. These designs are provably most efficient for bioequivalence under linear mixed-effects analysis. The pmsimstats-ng framework represents designs as lists of `ondrug` vectors, which can encode these structures in principle, but the fixtures were not built for the present sweeps. Reported power for biosimilar-style applications should be treated as a lower bound on what a complete-design analysis would achieve.
- **Sequential-monitoring variants with adaptive stopping.** Jiang, Arnold, and Betensky 2025 [@jiangDesignAnalysis2025] and Selukar, Prince, and May 2025 [@selukarFrameworkSequential2025] introduced per-patient and per-series stopping rules. Such adaptive designs reduce expected trial duration without loss of power when correctly implemented, and would be expected to reach a given power target with fewer expected cycles than our fixed-length designs. Our framework does not currently implement stopping rules, and the present power numbers should be read as fixed-horizon estimates.
- **Randomized-block N-of-1.** Our alternating A/B/A/B template and Hendrickson hybrid template are fixed schedules; patients within a path receive the same treatment sequence. A randomized-block N-of-1 design assigns a random sequence per patient (or per cycle within patient). Randomization reduces period effects and confounding but introduces sequence-specific variance components that the present framework does not model. Implementation would require randomizing the `ondrug` vectors per subject at the `generate_data` step and adding a sequence factor to `lme_analysis`. The expected effect on power is modest under balanced randomization but could matter substantially in small trials.

Taken together, these unevaluated design variants fall into two classes. The single-subject, open-label-only, OL+BDC-only, and two-cycle crossover designs are expected to produce power at or below the numbers reported here; our results are therefore an upper bound on those designs under matched parameters. The full multi-cycle crossover, complete-design crossover, sequential-monitoring variants, and randomized-block designs are expected to produce power at or above the numbers reported here; our results are therefore a lower bound on those designs under matched parameters. In either case, extending the present framework to cover these variants is a tractable implementation task (new design fixtures in `00-common.R`, plus in one case a modest extension of `lme_analysis`) rather than a conceptual redesign.

# Acknowledgments

We thank the clinical researchers whose published work on prazosin for PTSD provided the empirical foundation for our simulation parameters. We also acknowledge the methodological contributions of the N-of-1 trials research community in developing the analytical frameworks that made this work possible.

# Author Contributions

[To be filled in based on actual authorship]

# References

<div id="refs"></div>

# Supporting Information

**S1 File. Simulation code and reproducibility materials.** Complete R code for all simulations, including parameter specifications, statistical analysis functions, and visualization code. The simulation framework includes Morris et al. (2019) compliant functions for calculating Monte Carlo standard errors, Type I error evaluation, and comprehensive performance summaries.

**S2 File. Extended sensitivity analyses.** Additional simulation results examining robustness to alternative parameter specifications and design variations. Includes comparison of carryover handling strategies (period exclusion vs. explicit carryover modeling) and carryover decay model sensitivity analysis (exponential, linear, and Weibull decay functions).

**S3 File. Individual N-of-1 trial examples.** Detailed illustrations of individual crossover patterns and analytical approaches for representative patients.

**S4 File. Monte Carlo standard error calculations.** Technical documentation of MCSE formulas following Morris et al. (2019) Table 6, including formulas for power, bias, empirical SE, MSE, and coverage.

---

**Competing Interests:** The authors have declared that no competing interests exist.

**Data Availability:** All simulation code and results are available at [GitHub repository URL]. No human subjects data were collected for this study.

**Funding:** [To be filled in based on actual funding sources]