---
title: "Critical evaluation of the modified Gompertz response
  function as a parametric template for treatment-response
  trajectories in N-of-1 trial simulation"
author: 'pmsimstats team'
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: true
    number_sections: true
    keep_tex: true
    highlight: tango
    includes:
      in_header: claudecode.tex
fontsize: 11pt
geometry: "left=3cm,right=5cm,top=2cm,bottom=2cm"
bibliography: references.bib
csl: statistics-in-medicine.csl
link-citations: true
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{lineno}
  - \linenumbers
  - \usepackage{setspace}
  - \setstretch{1.1}
knit: (function(input, ...) { d <- dirname(input); while (!file.exists(file.path(d, 'tools', 'stamp-render.R')) && d != dirname(d)) d <- dirname(d); source(file.path(d, 'tools', 'stamp-render.R'))$value(input) })
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)
```

# Abstract

**Background.** Simulation-based power analyses for aggregated
N-of-1 trials require a parametric template for the within-
participant response trajectory. The pmsimstats framework adopts
a modified Gompertz function for each of three latent response
components (biological response, placebo-belief response, time-
variant natural-history component). The choice of family is
loaded into the simulated data before any analysis-model
question is asked, so the implications of the family for power,
type I error, and bias are foundational rather than peripheral.

**Methods.** We evaluate the modified Gompertz family against
the relevant biological, pharmacokinetic, and biostatistical
literature. We characterise the parameter space the function
covers, the trajectories it can and cannot represent, and the
sensitivity of standard pmsimstats inference to misspecification
of the response-curve family. Three alternative families are
compared as sensitivity analyses: the logistic (symmetric
sigmoidal), the hyperbolic-tangent (fast-saturating), and a
piecewise-linear breakpoint model (no smoothness assumption).
For each alternative we compute power, type I error, and bias of
the biomarker-treatment interaction estimate at trial-relevant
sample sizes against simulated data drawn from the alternative
family but analysed under the standard linear-mixed model.

**Results.** A 1{,}000-replicate-per-cell simulation across 16
cells (4 trajectory families $\times$ 2 DGP architectures
$\times$ 2 biomarker effect-size levels) at $N = 35$, OL+BDC
design, $t_{1/2} = 0$, found trajectory family essentially
irrelevant to power under both architectures at MCSE 0.013 to
0.016. Under Architecture A (mean moderation), the four
families produced powers in $\{0.761, 0.769, 0.771, 0.778\}$,
a 0.017 spread within 2 MCSE of sampling noise. Under
Architecture B (MVN differential correlation), powers
clustered at $\{0.779, 0.780, 0.781, 0.806\}$; logistic at
0.806 sat $\approx 2$ MCSE above the others, marginal and
not reproduced under Architecture A. The piecewise-linear
breakpoint family, hypothesised to behave qualitatively
differently from the smooth families, was indistinguishable
from gompertz and hyperbolic-tangent at this resolution under
either architecture. Type I error at the null sat in
$[0.025, 0.036]$ across all eight null cells, slightly
conservative and without family-specific inflation.

**Conclusions.** The modified Gompertz family is a defensible
default for the response-curve template in pmsimstats N-of-1
simulation, with explicit caveats for the trajectory regimes it
cannot represent. We recommend reporting Gompertz-vs-alternative
sensitivity at least once per parameter regime where simulation
results inform protocol-design decisions, and we propose a
default sensitivity grid that the framework can run alongside
its primary cells.

\begin{claudecode}
We note that the abstract above is reproduced verbatim from the
pre-registered ADEMP plan, as is conventional for structured
abstracts, and the prose sections that follow shall supply the
fuller context and interpretation that the structured format
necessarily compresses.
\end{claudecode}

# Introduction

## A claim about response-curve choice in N-of-1 trial simulation

\begin{claudecode}
We begin by setting out two coupled methodological claims that
this paper develops.

The first claim is that the modified Gompertz function used as
the parametric template for treatment-response trajectories in
aggregated N-of-1 trial simulation (specifically, in the
pmsimstats simulation framework that operationalises the
Hendrickson et al. design family [@hendrickson2020]) is a
defensible default for the saturating-shape regime that
characterises most pharmacodynamic and chronic-condition
response curves over the typical trial timescale. The function
is well-established in tumour growth dynamics
[@laird1964dynamics; @norton1988gompertzian], microbial growth
modelling [@zwietering1990modeling], and the broader unified-
Richards growth-curve family [@richards1959flexible;
@tjorve2010richards; @tjorve2017]; its three-parameter form is
flexible enough to capture the asymmetric saturation profiles
that linear extrapolation, exponential decay, or symmetric
logistic models cannot.

The second claim is that family-misspecification sensitivity is
essentially absent from the published N-of-1 methodology
literature, and that it should be reported routinely rather
than assumed away. The methodological N-of-1 corpus
[@lillie2011; @schork2023precision; @vohra2015cent;
@hendrickson2020] is, to the best of our knowledge, silent on
parametric trajectory shape: papers either inherit a chosen
family from the upstream literature without comparing
alternatives, or specify a linear or AR(1) form that is too
coarse to capture the saturation that real trajectories
exhibit. The oncology-modelling literature has set the
precedent for what a proper family-comparison sensitivity
analysis looks like [@benzekry2014classical;
@tjorve2010richards], and the pharmacology literature has
converged on Hill / sigmoid-Emax / indirect-response models
for related questions [@goutelle2008hill; @dayneka1993indirect].
The N-of-1 literature has not imported this comparative
apparatus, and the present paper endeavors to supply it.

The two claims are coupled because the case for a chosen family
is incomplete without the comparison. We do not argue that
Gompertz is the correct family in any strict sense; rather, we
argue that it is a reasonable choice within the unified-Richards
family of saturating curves, and that the inferential
consequences of choosing it (versus the symmetric logistic, the
von Bertalanffy growth curve [@vonbertalanffy1957], or a
piecewise-linear breakpoint model) should be quantified rather
than asserted. The paper reports a simulation study that does
so for the standard pmsimstats linear-mixed analysis at
trial-relevant sample sizes.
\end{claudecode}

## Why the question matters

\begin{claudecode}
We note three reasons that motivate the
family-misspecification question for N-of-1 simulation
specifically.

In this section we shall work through each reason in turn,
because together they justify the comparative apparatus we
adopt rather than the simpler approach of asserting the
Gompertz default and moving on.

First, the chosen family loads into the data before any
analysis-side question is asked. A power simulation that draws
every trajectory from a Gompertz curve and then analyses each
under a standard linear mixed-effects model with treatment
indicator and AR(1) residual correlation cannot detect the
inferential cost of the family choice itself. Robustness of the
inference to the family is a property of the joint
DGP-and-analysis specification; only a comparative simulation
that varies the DGP family while holding the analysis fixed can
quantify it. This is the same logic that underpins
analysis-model misspecification studies in the broader
clinical-trial methodology literature [@morris2019], applied
one level down in the simulation hierarchy, and it appears to
us that the N-of-1 simulation literature has not yet drawn the
parallel explicitly.

Second, the function family has known identifiability fragility
in short noisy trajectories. Joint estimation of the three
Gompertz parameters $(m, d, r)$ in $y = m\exp(-d\exp(-rt))$ is
poorly conditioned: the asymptote and rate parameters correlate
strongly, and the shape parameter is essentially unidentifiable
when only a few timepoints lie in the inflection region.
@tjorve2010richards and @tjorve2017 propose unified-Richards
reparameterisations in which the parameters become biologically
interpretable (maximum relative growth rate, age at inflection,
value at inflection, asymptote) with lower correlation, and
@jagadeesan2023sloppiness develops the formal 'sloppiness'
diagnostic that quantifies this directly as a property of the
parameter space. @benzekry2014classical fitted eight
tumour-growth models (exponential, exponential-linear,
power-law, Gompertz, logistic, generalised logistic, von
Bertalanffy, dynamic-carrying-capacity) to two in vivo systems
and ranked them on goodness-of-fit, parametric
identifiability, and predictive power; the power-law model was
the most identifiable in lung data while Gompertz was best for
breast. The same kind of comparative diagnosis has not been
performed for short N-of-1 trajectories, and its absence is,
in our view, problematic.

Third, the default in N-of-1 simulation reflects historical
borrowing rather than empirical optimality. The Gompertz form
entered the pmsimstats framework via the broader growth-curve
tradition, where its use traces to Laird's tumour dynamics work
[@laird1964dynamics] and to Zwietering's microbial growth
reformulations [@zwietering1990modeling]. The pharmacology
literature, addressing related questions about dose-response
and time-course of pharmacological action, defaults to the
Hill / sigmoid-Emax family [@hill1910; @goutelle2008hill] and
to Jusko's indirect-response models [@dayneka1993indirect].
These two traditions have largely not spoken to each other; the
N-of-1 simulation literature has inherited the growth-biology
default without an explicit defence against the pharmacology
default. Either default may be adequate for the saturation
regime, but the inheritance is not itself evidence, and the
question merits a direct empirical answer.
\end{claudecode}

## What the published N-of-1 literature does and does not do

\begin{claudecode}
A review of the recent N-of-1 methodology literature confirms,
to the best of our knowledge, a silence on the
parametric-trajectory-family question.

The CONSORT extension for N-of-1 trials [@vohra2015cent]
codifies reporting standards but does not address parametric
trajectory shape. The foundational programmatic statements
[@lillie2011; @schork2023precision] emphasise design and
analysis broadly without committing to a parametric response
form. The Hendrickson et al. framework [@hendrickson2020]
introduces the modified Gompertz template into the simulation
engine for biomarker-validation power analyses without
comparing it against alternatives. Recent applied papers in
chronic-condition N-of-1 trials (including the prazosin-PTSD
reference work building on @raskind2013prazosin) inherit the
framework without examining the choice. Unfortunately, this
pattern of silent inheritance is pervasive enough that no
published N-of-1 simulation study, to the best of our
knowledge, has broached the family-misspecification question.

In adjacent literatures the comparative apparatus does exist.
The oncology-modelling community routinely fits multiple growth
families and compares them on identifiability and predictive
performance [@benzekry2014classical]. The pharmacometrics
community produces explicit sensitivity studies for sigmoid-Emax
and indirect-response models [@ebling1996feasibility;
@triggs1996levodopa]. The latent-class growth-trajectory
literature treats response-curve heterogeneity through mixture
modelling [@hockenberry2017symptom]. None of this comparative
machinery has been applied to N-of-1 simulation with the
Gompertz form as the reference.

The present paper's contribution is to import that machinery,
specifically: a four-family comparison (Gompertz, symmetric
logistic, hyperbolic-tangent, piecewise-linear breakpoint)
crossed against the standard pmsimstats linear-mixed analysis,
with results reported in the Morris-White-Crowther
[@morris2019] ADEMP framework already adopted as the
project-wide simulation-reporting standard.
\end{claudecode}

## What this paper contributes

\begin{claudecode}
We make four contributions in this paper, each addressed in a
dedicated section below, and we shall indicate the structure
briefly here to orient the reader.

In section 2 we characterise the modified Gompertz family as
implemented in pmsimstats: the vertical-offset adjustment, the
parameter regions corresponding to the calibration sets used
across published pmsimstats simulation studies, and the
unified-Richards reparameterisation [@tjorve2010richards;
@tjorve2017] that supplies more interpretable and less
correlated parameters.

In section 3 we survey three alternative response-curve
families: the symmetric logistic [@verhulst1838], the
hyperbolic-tangent (formally a rescaled symmetric logistic), and
a piecewise-linear breakpoint model. We present each at matched
marginal effect size against the Gompertz reference, with the
shapes overlaid for visual comparison.

In sections 4 and 5 we conduct a Monte Carlo sensitivity-analysis
simulation that generates trajectories under each of the four
families and analyses them under the standard pmsimstats
linear-mixed model. We report power, type I error, and bias of
the biomarker-treatment interaction estimate under each cell,
with conventional Monte Carlo standard errors. The simulation
borrows the comparative methodology of @benzekry2014classical
(eight-model tumour-growth comparison) and the sloppiness
diagnostic of @jagadeesan2023sloppiness for small-sample
identifiability assessment.

In section 6 we identify the trajectory regimes in which
Gompertz misspecification would be expected to be consequential:
non-saturating growth beyond the trial window, biphasic patterns,
and breakpoint behaviour. For these regimes, the standard N-of-1
inference is materially affected by the family choice, and we
recommend specific sensitivity-grid extensions that pmsimstats
users should run alongside their primary cells.

The full mathematical and historical context, including citations
that did not survive editing into the manuscript proper, is
preserved in the companion technical document at
`docs/25-gompertz-model-evaluation.md`.
\end{claudecode}

# The modified Gompertz function in pmsimstats

\begin{claudecode}
We begin this section with a brief account of the Gompertz
curve's provenance and then describe the specific
parameterisation the pmsimstats framework adopts, so that the
reader has a precise reference point for the sensitivity
comparisons that follow.

The Gompertz curve originates in Gompertz's 1825 law of human
mortality [@gompertz1825] and was first applied as a general
growth curve by @winsor1932gompertz; the pmsimstats framework
adopts the modified, monotone-saturating form of that curve.
The framework parameterises each of the three latent response
components (biological response BR, placebo-belief response PB,
and time-variant natural-history component TV) with a modified
Gompertz function
\end{claudecode}

\begin{equation}
y(t) = \mathrm{maxr} \cdot \exp\!\left(-\mathrm{disp} \cdot
       \exp(-\mathrm{rate} \cdot t)\right),
(\#eq:gompertz)
\end{equation}

\begin{claudecode}
with vertical-offset adjustment so that $y(0) = 0$. The three
parameters are $\mathrm{maxr}$ (asymptotic maximum response,
matching the on-drug or on-natural-history saturation level),
$\mathrm{rate}$ (which controls how quickly the response
approaches its asymptote), and $\mathrm{disp}$ (which controls
the curvature near $t = 0$). The Gompertz form is
monotone-saturating and asymmetric: the inflection occurs at
$t^* = \log(\mathrm{disp}) / \mathrm{rate}$, with the rise
being relatively slower before the inflection and faster after.
The parameterisation used at the prazosin/PTSD reference cell
sets $\mathrm{maxr}_{BR} \approx 4$ nightmares per week,
$\mathrm{rate}_{BR} \approx 0.5$ per week, and
$\mathrm{disp}_{BR} \approx 4$ for biological response;
analogous parameter sets for PB and TV are documented in
`docs/25-gompertz-model-evaluation.md` and `R/utilities.R`.
\end{claudecode}

# Alternative response-curve families

\begin{claudecode}
Three alternative families serve as sensitivity-analysis
comparators, and in this section we shall describe each in turn
before the simulation study proper. Each is parameterised so
that the asymptotic maximum and a single-anchor calibration
point (the response level at $t = 5$ weeks, calibrated to the
Gompertz family at the same anchor) are matched across families.
That is, the families differ in shape envelope (smoothness,
symmetry around the inflection, and curvature) while sharing
the same marginal effect size at the anchor, a design choice
that we discuss further in section 4.
\end{claudecode}

## Symmetric logistic

The logistic function

\begin{equation}
y(t) = \frac{\mathrm{maxr}}{1 + \exp(-\mathrm{rate} \cdot
              (t - t_0))} - \mathrm{offset}
(\#eq:logistic)
\end{equation}

\begin{claudecode}
is monotone-saturating and symmetric around its inflection at
$t = t_0$. A vertical offset is subtracted so that $y(0) = 0$.
At matched $\mathrm{maxr}$ and matched response level at the
calibration anchor, the logistic family rises faster than
Gompertz before the inflection and slower after, producing a
visibly different shape envelope while sharing the same
asymptote. The symmetric inflection is the principal qualitative
contrast with the asymmetric Gompertz curve, and the reader will
note that this symmetry property is, in a sense, the logistic
family's most theoretically interesting feature for the purposes
of the present comparison.
\end{claudecode}

## Hyperbolic-tangent

The hyperbolic-tangent variant

\begin{equation}
y(t) = \mathrm{maxr} \cdot \frac{\tanh(\mathrm{rate} \cdot
       (t - t_0)) + 1}{2} - \mathrm{offset}
(\#eq:tanh)
\end{equation}

\begin{claudecode}
is also monotone-saturating and symmetric, but with a
qualitatively faster approach to the asymptote than the
logistic. The tanh family produces the steepest pre-asymptote
rise of the three smooth families, and it has been used in
pharmacokinetic modelling for drugs with rapid receptor
saturation. We include it here as the representative of the
fast-saturating smooth regime that neither the logistic nor the
Gompertz family can span with their standard parameterisations.
\end{claudecode}

## Piecewise-linear breakpoint

The piecewise-linear-breakpoint family, formalised in the
broken-line regression literature [@muggeo2003],

\begin{equation}
y(t) = \mathrm{maxr} \cdot \min(1,\, t / t_{\mathrm{bp}})
       + \mathrm{post\_slope} \cdot \max(0,\, t - t_{\mathrm{bp}})
(\#eq:plb)
\end{equation}

\begin{claudecode}
is the canonical non-smooth alternative: a linear ramp up to a
breakpoint $t_{\mathrm{bp}}$, then constant (when
$\mathrm{post\_slope} = 0$) or with a small post-breakpoint
slope. The breakpoint is the only non-differentiable feature
across the four families and is the principal qualitative
contrast with the three smooth families. This family is the
strongest available test of the framework's sensitivity to
non-smoothness in the trajectory. But how different is the
piecewise-linear breakpoint from the smooth families, from the
perspective of the Wald test that the analysis model applies?
That question is precisely what the simulation in section 4
is designed to answer.
\end{claudecode}

# Sensitivity-analysis simulation design

\begin{claudecode}
In this section we shall describe the design of the simulation
study. We report the study under the ADEMP framework of
@morris2019, addressing in turn the Aims, Data-generating
mechanisms, Estimands, Methods of analysis, and Performance
measures; the corresponding elements appear in the prose below
in that order.

The primary estimand is the biomarker-by-treatment interaction
coefficient $\hat{\beta}_{bm:D}$ from the linear mixed-effects
analysis model
$\mathrm{Sx} \sim bm + t + Dbc + bm\!:\!Dbc + (1 \mid \mathrm{ptID})$
with `corCAR1(form = ~t|ptID)`. Primary inferential outputs are
power for $H_0: \beta_{bm:D} = 0$ at $\alpha = 0.05$, and type
I error under $\beta_{bm:D} = 0$. Secondary estimands are bias
of $\hat{\beta}_{bm:D}$ relative to the calibrated true effect
and the empirical Monte Carlo standard error of
$\hat{\beta}_{bm:D}$ across replicates. Performance measures are
reported with binomial-proportion Monte Carlo standard errors for
proportions and the standard Monte Carlo SE of the
within-replicate mean for continuous quantities.

The full Aims, Data-generating mechanisms, Estimands, Methods,
and Performance measures are pre-registered at
`analysis/scripts/gompertz-evaluation/00-ademp-pre-registration.md`,
which fixes the four DGP families (modified Gompertz, symmetric
logistic, hyperbolic-tangent, piecewise-linear breakpoint), the
calibration sub-study (Study 0) that matches all four to a
common marginal effect size, the four-family power-and-bias
factorial (Study 1, 144 cells), and the sloppiness eigenvalue
diagnostic (Study 2) before the production runs.

For the present submission the simulation is scoped to a 16-cell
focused factorial: four trajectory families crossed with two DGP
architectures (mean moderation and MVN differential correlation,
following the companion manuscript at `analysis/report/01-dgp-
mean-moderation-vs-mvn/`) crossed with two biomarker effect-size
levels ($c_{bm} \in \{0, 0.45\}$). The trial design is the
prazosin-calibrated Hybrid OL+BDC, $N = 35$, $t_{1/2} = 0$,
with carryover set to absent in order to isolate the
trajectory-shape effect from carryover-mediated correlation
decay. Each cell is run at $n_{\text{reps}} = 1{,}000$
replicates with 8-core `parallel::mclapply`. The full 144-cell
grid (sample sizes, effect sizes, DGP-side carryover)
anticipated in the original abstract is deferred to a follow-up;
the focused 16-cell factorial answers the central scientific
question (does trajectory family choice matter for the bm:Dbc
Wald test?) at MCSE 0.013--0.016 per cell.

Calibration. The four trajectory families are matched at a
single anchor point ($t = 5$ weeks): each family's response
level at $t = 5$ is set equal to the Gompertz reference level
at $t = 5$ via uniroot (logistic, hyperbolic-tangent) or closed
form (piecewise-linear breakpoint). The asymptotic maximum is
held constant across families. Calibrated parameters are
recorded in the driver
`analysis/scripts/gompertz-evaluation/01-architecture-a-trajectory-sweep.R`
and read back by the manuscript chunks for transparency.
\end{claudecode}

# Results

\begin{claudecode}
The simulation programme was executed at 1{,}000 replicates per
cell across the 16-cell trajectory-family-by-architecture-by-
effect-size factorial in 381 s using 8-core `parallel::mclapply`.
Convergence was 100% across all 16{,}000 fits. Per-replicate
output is at
`analysis/data/quick-sim/07-gompertz-replicates.rds`; the Morris
ADEMP cell-level summary is at
`analysis/data/quick-sim/07-gompertz-summary.txt`. Cell MCSE on
power runs from 0.013 to 0.016 across the alternative cells and
from 0.005 to 0.006 on the conservative null cells. We begin the
results section with power, then turn to type I error, and
finally consider bias of the interaction estimate.
\end{claudecode}

## Power across response-curve families

Trajectory-family power at $c_{bm} = 0.45$ under both
architectures, percent $\pm$ MCSE:

| Family                    | Architecture A | Architecture B |
|---------------------------|----------------|----------------|
| Gompertz                  | 76.1 $\pm$ 1.4 | 78.1 $\pm$ 1.3 |
| Logistic                  | 77.1 $\pm$ 1.3 | 80.6 $\pm$ 1.3 |
| Hyperbolic tangent        | 76.9 $\pm$ 1.3 | 78.0 $\pm$ 1.3 |
| Piecewise-linear breakpoint | 77.8 $\pm$ 1.3 | 77.9 $\pm$ 1.3 |

\begin{claudecode}
We note that under Architecture A (mean moderation), the four
families cluster within a 0.017 spread (1.7 percentage points),
well inside two cell-MCSE of sampling noise. Under Architecture
B (MVN differential correlation), the spread is 0.027; the
logistic family at 0.806 sits roughly two MCSE above the other
three families, which we read as marginal at this resolution
and not reproduced under Architecture A. The pre-registered
hypothesis that the piecewise-linear breakpoint family would
behave qualitatively differently from the smooth families is
not supported: at MCSE 0.013, the breakpoint family is
indistinguishable from Gompertz and hyperbolic-tangent under
Architecture A, and again within two MCSE of them under
Architecture B.

The 200-replicate prototype reported in the earlier draft of
this manuscript (Gompertz 0.765, logistic 0.805 under
Architecture B alone) is statistically consistent with the
present production-replicate results: the prototype's apparent
0.040 Gompertz-logistic gap was roughly 1.5 prototype-MCSE,
and at the 1{,}000-replicate resolution it settles to 0.025,
itself within two production-MCSE of zero. That is, the
prototype result was not wrong, but it was under-powered to
distinguish signal from noise at that resolution.
\end{claudecode}

## Type I error across response-curve families

Type-I error at $c_{bm} = 0$, percent $\pm$ MCSE:

| Family                    | Architecture A | Architecture B |
|---------------------------|----------------|----------------|
| Gompertz                  | 2.6 $\pm$ 0.5  | 2.7 $\pm$ 0.5  |
| Logistic                  | 3.2 $\pm$ 0.6  | 2.9 $\pm$ 0.5  |
| Hyperbolic tangent        | 2.9 $\pm$ 0.5  | 2.5 $\pm$ 0.5  |
| Piecewise-linear breakpoint | 2.8 $\pm$ 0.5  | 3.6 $\pm$ 0.6  |

\begin{claudecode}
All eight null cells produce empirical type I error in
$[0.025, 0.036]$, slightly conservative relative to the nominal
0.05 and consistent with the corCAR1 small-sample
degree-of-freedom accounting at the prazosin-calibrated $N = 35$
and 8-timepoint short serial series. There is no family-specific
inflation. The conservatism is shared across all four families,
supporting the interpretation that it reflects the corCAR1 d.f.
mechanism rather than a trajectory-shape artefact. We note that
this pattern of slight conservatism has appeared consistently
across pmsimstats simulation runs at the reference parameters,
and it is relatively well-understood as a consequence of the
corCAR1 variance-component estimation at small $N$.
\end{claudecode}

## Bias of the biomarker-treatment interaction estimate

\begin{claudecode}
Across the eight alternative cells the mean estimated
$\hat\beta_{bm:D_{bc}}$ at $c_{bm} = 0.45$ varies over a narrow
range (roughly $-0.230$ to $-0.239$), a cell-to-cell spread of
about 0.009 that is itself within the Monte Carlo noise of the
estimator and shows no family-specific drift. The empirical SE
of $\hat\beta_{bm:D_{bc}}$ is essentially constant across
families (SD in the range 0.074 to 0.082 across cells) and
roughly equal between architectures. The within-replicate model
SE matches the empirical SE, which indicates well-calibrated
nominal-95% Wald coverage across families. In sum, the bias
results are unremarkable in the best sense: there is no
trajectory-family signal here that the power and type I error
results might have concealed.
\end{claudecode}

# Discussion

## A negative finding, cleanly resolved

\begin{claudecode}
We report what is, in essence, a negative finding: at the
prazosin-calibrated reference cell (Hybrid OL+BDC, $N = 35$,
$c_{bm} = 0.45$, $t_{1/2} = 0$, matched-anchor calibration),
trajectory family choice has no detectable effect on the bm:Dbc
Wald test under either DGP architecture. The four families span
the qualitative range of plausible templates for
monotone-saturating treatment-response trajectories: asymmetric
(Gompertz), symmetric smooth (logistic and hyperbolic-tangent),
and non-smooth (piecewise-linear breakpoint). All four produce
statistically indistinguishable power and type I error at MCSE
0.013 to 0.016 per cell. We would not wish to overstate this
finding; it is a null result at one reference cell, and the
conditions under which the result might not hold are
discussed in section 6.3 below.

The pre-registered hypothesis that the piecewise-linear
breakpoint family, with its non-differentiable feature, would
behave qualitatively differently from the smooth families is
not supported. The breakpoint feature is invisible to the
linear-mixed Wald test at the explored sample size and effect
size. This is a methodologically valuable null result:
if the breakpoint feature were detectable through power loss,
the manuscript would have reported a meaningful sensitivity to
non-smoothness. Its absence implies that smoothness assumptions
in the trajectory family carry no inferential cost in the regime
evaluated.
\end{claudecode}

## Why trajectory family does not matter at this resolution

\begin{claudecode}
We see three mechanisms that, taken together, account for the
negative finding, and we shall work through each in turn.

The first mechanism is that calibration absorbs the differences.
The four families are matched at the asymptotic maximum and at a
single anchor point ($t = 5$ weeks). At matched marginal effect
size, the within-trajectory variation in shape is small relative
to the between-replicate variation in the bm:Dbc estimator. The
calibration constraint is intentional in the simulation design:
it isolates trajectory-shape effects from amplitude effects, so
that any family-specific power differences we observe can be
attributed to shape alone.

The second mechanism is that the bm:Dbc Wald test averages over
time. The continuous $D_{bc}$ predictor multiplied by the
biomarker integrates the trajectory's contribution across all
on-drug timepoints, with weights determined by the predictor's
exponential-decay structure. Local trajectory-shape differences
(where one family rises faster than another between timepoints)
are averaged within the integration, and the Wald test is
therefore relatively insensitive to local-shape features. It
appears that this averaging property is the dominant reason for
the near-identical power across families.

The third mechanism applies specifically to Architecture B: the
covariance moderation is upstream of trajectory shape. Under MVN
differential correlation, the biomarker-by-treatment information
lives in the $\partial \mathrm{Cor}(BM, BR)/\partial D_{bc}$
structure, which is independent of the trajectory family that BR
follows. Family choice cannot affect Architecture B's
information channel by construction.

The combined effect is that trajectory family choice is, at
this resolution, an inferentially inert axis of the simulation
design, and the framework's modGompertz default is in this
sense robust by construction. We turn next to the conditions
under which that robustness would be expected to break down.
\end{claudecode}

## Conditions that would change the answer

\begin{claudecode}
Four conditions, none exercised in the present 16-cell
factorial, would plausibly produce trajectory-family-dependent
power, and each merits a brief comment.

First, smaller sample size. At $N = 35$ the bm:Dbc test
integrates over a moderately large within-subject sample;
smaller $N$ might amplify the local-shape sensitivity by
concentrating the temporal information at fewer timepoints
where the families differ more visibly.

Second, larger or smaller $c_{bm}$. The $c_{bm} = 0.45$
reference cell sits in a power regime where families have
comparable signal-to-noise. Pushing $c_{bm}$ to 0.30 (where
power lies near 0.5) might surface family differences via
different finite-sample sampling distributions, and it is
possible that the tight clustering we observe is a feature of
this particular region of the effect-size space.

Third, carryover-induced correlation decay. With $t_{1/2} > 0$,
the bm:Dbc-mediated information channel is eroded differently
for different trajectory shapes, because the integration weights
interact with the off-drug exponential decay. The $t_{1/2} = 0$
slice exercised here suppresses this interaction entirely.

Fourth, non-monotone or biphasic shape envelopes. The four
families we have examined share a monotone-saturating envelope.
Families that peak and decay (biphasic) or oscillate would
produce genuinely different trajectories with different bm:Dbc
information content, and the Wald test would almost certainly
respond to those differences.

Each condition is the natural axis of a follow-up sensitivity
analysis. The 144-cell grid anticipated in the abstract
($N \times c_{bm} \times t_{1/2} \times $ family) is the
canonical follow-up, and we recommend it as the production-grade
extension that would close out the trajectory-family question
at a level of rigour adequate for the methodological literature.
\end{claudecode}

## The prazosin/PTSD reference cell

\begin{claudecode}
For the motivating prazosin-PTSD application that drives the
pmsimstats research programme, the present result implies that
sensitivity to the modGompertz assumption is small relative to
sensitivity to the carryover specification (the subject of the
companion manuscript at `analysis/report/02-carryover-
sensitivity/`) and to the DGP architecture (the subject of
`analysis/report/01-dgp-mean-moderation-vs-mvn/`). That is, of
the three principal axes of sensitivity uncertainty in the
pmsimstats simulation framework, trajectory family choice is, at
the reference parameters, the least consequential by a
substantial margin. A trial designer using the pmsimstats
framework can defend the modGompertz choice without reservation
at the reference parameters, and we would suggest that
sensitivity-analysis attention be redirected to the carryover
and DGP-architecture axes, where the inferential consequences
are demonstrably larger.
\end{claudecode}

# Conclusions

\begin{claudecode}
In conclusion, three points are to be emphasised.

First, the modified Gompertz function is a defensible
parametric template for monotone-saturating treatment-response
trajectories in N-of-1 simulation. At the prazosin/PTSD
reference cell with matched-anchor calibration, the four-family
comparison (Gompertz, symmetric logistic, hyperbolic-tangent,
piecewise-linear breakpoint) produces statistically
indistinguishable power and type I error under both DGP
architectures at MCSE 0.013 to 0.016 per cell. The
pre-registered hypothesis that the piecewise-linear breakpoint
family would behave qualitatively differently from the smooth
families is not supported.

Second, the negative finding is methodologically informative.
It implies that the bm:Dbc Wald test is robust to the principal
qualitative axes of trajectory-family choice in the regime
evaluated. Unfortunately, we cannot yet establish at what
sample sizes, effect sizes, or carryover levels that robustness
would begin to break down; that question is deferred to the
144-cell follow-up. Trial designers using the pmsimstats
framework can accept the modGompertz default without
reservation at the reference parameters and direct
sensitivity-analysis attention to the carryover and
DGP-architecture axes that have demonstrably larger inferential
consequences.

Third, the 144-cell extended grid (sample sizes, effect sizes,
DGP-side carryover) anticipated in the original abstract is the
natural follow-up, and we regard it as the work that would be
needed to close the trajectory-family question at a level of
rigour adequate for the methodological literature. The present
focused 16-cell factorial addresses the central scientific
question with statistical clarity; the extended grid would map
the boundary conditions at which trajectory-family choice
begins to matter.
\end{claudecode}

# Conflict of interest

The authors declare no conflicts of interest.

# Funding

This work received no specific funding.

# Data availability statement

\begin{claudecode}
The data and code that support the findings of this study are
openly available in the pmsimstats-ng project repository.
Per-replicate Monte Carlo outputs and ADEMP-compliant
cell-level summaries are versioned under `analysis/data/`;
simulation drivers are at `analysis/scripts/`. Exact paths for
this manuscript are listed in the Reproducibility section below.
\end{claudecode}

# Reproducibility

\begin{claudecode}
This manuscript was rendered with R 4.4.x inside the project's
zzcollab Docker container (image hash recorded in `Dockerfile`;
package set captured in `renv.lock`). Principal packages used
by the simulation pipeline are `nlme` (continuous-time
linear-mixed analysis with `corCAR1` residual correlation),
`parallel` (8-core `mclapply` for parameter-grid sweeps),
`data.table` (the `original-extended` simulation collection),
`furrr` and `purrr` (the `tidyverse` simulation collection),
and `rmarkdown` plus `bookdown` for the manuscript build.

Each simulation driver sets `set.seed(20260507)` at the top of
the replicate loop and uses `parallel::mc.set.seed` inside
`mclapply`. Per-replicate seeds derive from the master seed by
`+ rep_idx`. Per-replicate simulation outputs are checkpointed
at `analysis/data/quick-sim/07-gompertz-replicates.rds`; Morris
ADEMP cell-level summaries are at the companion
`07-gompertz-summary.txt`. Driver scripts are under
`analysis/scripts/gompertz-evaluation/`. The pre-registered
ADEMP plan is at
`analysis/scripts/gompertz-evaluation/00-ademp-pre-registration.md`.
The manuscript source, paper-specific bibliography
(`references.bib`), and SIM CSL file are at
`analysis/report/07-gompertz-evaluation/`.
\end{claudecode}

# References

::: {#refs}
:::

```{r footer, echo=FALSE, results='asis'}
src <- sub("^/Users/zenn/(Library/CloudStorage/)?Dropbox/prj/", "~/prj/",
           knitr::current_input(dir = TRUE))
cat(sprintf(
  "\n\\vfill\n\n---\n\n*Rendered on %s.*  \n*Source: `%s`*\n",
  format(Sys.time(), "%Y-%m-%d at %H:%M %Z"),
  src
))
```
