---
title: "Critical evaluation of the modified Gompertz response
  function as a parametric template for treatment-response
  trajectories in N-of-1 trial simulation"
author: 'pmsimstats team'
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: true
    toc_depth: 2
    number_sections: true
    keep_tex: true
    highlight: tango
    includes:
      in_header: claudecode.tex
fontsize: 11pt
linestretch: 1.1
geometry: 'left=3cm,right=3cm,top=2.5cm,bottom=2.5cm'
bibliography: references.bib
csl: statistics-in-medicine.csl
link-citations: true
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{float}
  - \usepackage[mathlines]{lineno}
  - \linenumbers
knit: (function(input, ...) { d <- dirname(input); while (!file.exists(file.path(d, 'tools', 'stamp-render.R')) && d != dirname(d)) d <- dirname(d); source(file.path(d, 'tools', 'stamp-render.R'))$value(input) })
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = 'center'
)
suppressPackageStartupMessages({
  suppressMessages(devtools::load_all(here::here(), quiet = TRUE))
  library(ggplot2)
})

# Trajectory-family calibration, reproduced from the simulation driver
# analysis/scripts/gompertz-evaluation/02-faithful-trajectory-sweep.R so
# that the introductory figure shows the curves the simulation used.
data(extracted_rp)
br_p     <- as.list(extracted_rp[cat == 'br'])
maxr_v   <- br_p$max
br_disp  <- br_p$disp
br_rate  <- br_p$rate
br_sd    <- br_p$sd
t_anchor <- 5
g_anchor <- modgompertz(t_anchor, maxr_v, br_disp, br_rate)

fam_par <- list(
  gompertz = list(p2 = br_disp, p3 = br_rate),
  logistic = list(p2 = 0.5, p3 = uniroot(function(x)
    shape_logistic(t_anchor, maxr_v, 0.5, x) - g_anchor,
    interval = c(-5, 20), tol = 1e-10)$root),
  hyperbolic_tangent = list(p2 = 0.5, p3 = uniroot(function(x)
    shape_hyperbolic_tangent(t_anchor, maxr_v, 0.5, x) - g_anchor,
    interval = c(-5, 20), tol = 1e-10)$root),
  piecewise_linear_breakpoint = list(
    p2 = t_anchor * maxr_v / g_anchor, p3 = 0))

fam_lab <- c(gompertz = 'Gompertz', logistic = 'Logistic',
             hyperbolic_tangent = 'Hyperbolic tangent',
             piecewise_linear_breakpoint = 'Piecewise linear')
fam_cols <- c('#1f78b4', '#e31a1c', '#33a02c', '#6a3d9a')

traj <- function(fam, tt) trajectoryShape(
  fam, tt, maxr_v, fam_par[[fam]]$p2, fam_par[[fam]]$p3)

theme_paper <- theme_bw(base_size = 9) +
  theme(panel.grid.minor = element_blank(),
        strip.background = element_rect(fill = 'grey92', colour = NA),
        legend.position = 'top')
```

# Abstract {-}

**Background.** Simulation-based power analyses for aggregated
N-of-1 trials require a parametric template for the within-
participant response trajectory. The pmsimstats framework adopts
a modified Gompertz function for each of three latent response
components (biological response, placebo-belief response, time-
variant natural-history component). The choice of family is loaded
into the simulated data before any analysis-model question is
asked. That is, its implications for power, Type I error, and bias
are foundational rather than peripheral.

**Methods.** We evaluate the modified Gompertz family against the
relevant biological, pharmacokinetic, and biostatistical
literature. We characterize the parameter space the function
covers, the trajectories it can and cannot represent, and the
sensitivity of standard pmsimstats inference to misspecification of
the response-curve family. We compare three alternative families as
sensitivity analyses: the logistic (symmetric sigmoidal), the
hyperbolic-tangent (fast-saturating), and a piecewise-linear
breakpoint model (no smoothness assumption). For each alternative we
compute power, Type I error, and the sampling distribution of the
biomarker-treatment interaction estimate at trial-relevant sample
sizes, against simulated data drawn from the alternative family but
analyzed under the standard linear-mixed model.

**Results.** We ran a 1{,}000-replicate-per-cell simulation across
16 cells (4 trajectory families $\times$ 2 data-generating process
(DGP) architectures $\times$ 2 biomarker effect-size levels) at
$N = 70$, open-label plus blinded discontinuation (OL+BDC) design,
$t_{1/2} = 0$. We find that the inferential consequence of
trajectory family is architecture-dependent. Under the
covariance-moderation architecture (multivariate normal (MVN)
differential correlation), where the biomarker-treatment
interaction lives in the covariance structure and is therefore
mean-independent, trajectory family is exactly inert. With matched
random-number seeds the four families return identical rejection
decisions, so the production-seed spread $\{0.774, 0.777, 0.791,
0.804\}$ is sampling noise. Under the mean-moderation architecture,
implemented in its faithful multiplicative form in which the
biomarker moderation rides the response trajectory, the four
families separate: powers were $\{0.743, 0.760, 0.778, 0.782\}$ for
Gompertz, hyperbolic-tangent, piecewise-linear, and logistic
respectively, a 0.039 spread. The Gompertz default sits roughly two
Monte Carlo standard errors (MCSE) below the logistic and is the
lowest-power, that is mildly conservative, family. Type I error at
the null sat in $[0.019, 0.038]$ across all eight null cells,
slightly conservative and without family-specific inflation, and
the estimator was unbiased there (mean $\hat\beta_{bm:D}$ within
0.004 of zero in every null cell). At the alternative the mean
interaction estimate was family-invariant ($-0.230$ to $-0.239$
across the eight cells, a spread within Monte Carlo noise) with a
near-constant empirical standard error, so the power separation is
a precision effect rather than a shift in the estimate. We do not
quantify bias at the alternative: the matched-anchor calibration
fixes the marginal effect size across families but yields no
closed-form true $\beta_{bm:D}$ on the outcome scale to difference
against.

**Conclusions.** The modified Gompertz family is a defensible
default for the response-curve template in pmsimstats N-of-1
simulation, with two qualifications. Under a covariance-channel
(MVN) data-generating process, trajectory family cannot affect the
interaction test by construction, so the choice is inferentially
free. Under a mean-moderation data-generating process, trajectory
family does carry a modest inferential cost at the few-percent
level, and the Gompertz default is mildly conservative relative to
the symmetric and breakpoint alternatives. We recommend reporting
Gompertz-versus-alternative sensitivity at least once per parameter
regime where simulation results inform protocol-design decisions,
with particular attention to mean-moderation regimes, and we
propose a default sensitivity grid that the framework can run
alongside its primary cells.

# Introduction

## A claim about response-curve choice in N-of-1 trial simulation

The first claim is that the modified Gompertz function used as
the parametric template for treatment-response trajectories in
aggregated N-of-1 trial simulation (specifically, in the
pmsimstats simulation framework that operationalizes the
Hendrickson et al. design family [@hendrickson2020]) is a
defensible default for the saturating-shape regime that
characterizes most pharmacodynamic and chronic-condition
response curves over the typical trial timescale. The function
is well-established in tumor growth dynamics
[@laird1964dynamics; @norton1988gompertzian], microbial growth
modeling [@zwietering1990modeling], and the broader unified-
Richards growth-curve family [@richards1959flexible;
@tjorve2010richards; @tjorve2017]; its three-parameter form is
flexible enough to capture the asymmetric saturation profiles
that linear extrapolation, exponential decay, or symmetric
logistic models cannot.

The second claim is that family-misspecification sensitivity is
essentially absent from the published N-of-1 methodology
literature, and that it should be reported routinely rather
than assumed away. The methodological N-of-1 corpus
[@lillie2011; @schork2023precision; @vohraCONSORTExtensionReporting2015;
@hendrickson2020] is, to the best of our knowledge, silent on
parametric trajectory shape: papers either inherit a chosen
family from the upstream literature without comparing
alternatives, or specify a linear or AR(1) form that is too
coarse to capture the saturation that real trajectories
exhibit. The oncology-modeling literature has set the
precedent for what a proper family-comparison sensitivity
analysis looks like [@benzekry2014classical;
@tjorve2010richards], and the pharmacology literature has
converged on Hill / sigmoid-Emax / indirect-response models
for related questions [@goutelle2008hill; @dayneka1993indirect].
The N-of-1 literature has not imported this comparative
apparatus, and the present paper endeavors to supply it.

The two claims are coupled because the case for a chosen family
is incomplete without the comparison. We do not argue that
Gompertz is the correct family in any strict sense. Rather, we
argue that it is a reasonable choice within the unified-Richards
family of saturating curves, and that the inferential
consequences of choosing it should be quantified rather than
asserted. The comparators are the symmetric logistic, the von
Bertalanffy growth curve [@vonbertalanffy1957], and a
piecewise-linear breakpoint model. The paper reports a
simulation study that quantifies those consequences for the
standard pmsimstats linear-mixed analysis at trial-relevant
sample sizes.

## What the trajectory template is for

A power simulation for an aggregated N-of-1 trial has to turn a
*design* into a *distribution*. The design is a schedule: a sequence
of visits, and for each visit a statement of whether the participant
is receiving active treatment and what the participant believes about
that assignment. A schedule carries no magnitudes. To emit an outcome
the simulator needs a rule that maps elapsed time, in each of several
distinct senses, onto an expected symptom score. The parametric
trajectory template is that rule. It is the only place in the
data-generating process where the size and the time course of a
treatment response are specified, and everything downstream
(covariance construction, censoring, analysis, power) operates on
what it produces.

The template is applied three times over, once per latent response
component, and the three applications differ in which clock they
read. In the reference implementation (\texttt{generateData}),
the biological response BR is evaluated at \texttt{tod}, the
cumulative time the participant has spent on active drug. The
expectancy or placebo-belief component PB is evaluated at
\texttt{tpb}, the accumulated time under a nonzero expectancy,
and is then scaled by the visit's expectancy weight. The
time-varying natural-history component TV is evaluated at
calendar time since baseline. The observed score is their sum
plus a correlated error. That the three components read three
different clocks is what makes them separable by design rather
than by assumption. A blinded-discontinuation block advances
calendar time and belief time while freezing time on drug, so
the components come apart in a way that no single-clock
specification could represent. Absent a trajectory template
there is no clock to read, and being on
drug degenerates to a binary label whose only expressible effect is a
step change.

The template therefore does four jobs simultaneously. It fixes the
magnitude of the treatment response at each scheduled visit. It fixes
the time course, and hence how much of the on-drug window actually
carries signal rather than sitting at the asymptote. It renders the
three latent components separable by giving each its own time
argument. And it supplies the level from which carryover decays after
withdrawal, since the post-discontinuation value is defined relative
to the on-drug trajectory attained at the moment drug stops.

What the template is not is a quantity that anyone estimates. The
analysis model is
\texttt{Sx \textasciitilde{} bm + t + Dbc + bm:Dbc} with a random
intercept and \texttt{corCAR1} residual correlation. It is linear in
calendar time and linear in the continuous drug indicator, and it
contains no free parameter corresponding to a response shape. The
template is a pure data-generating nuisance structure: chosen once,
before any simulation is run, entering every cell of every parameter
grid, and never confronted with the data it generates. That asymmetry
is what makes the choice both consequential and easy to leave
unexamined. Every power number the framework reports is conditional
on it, and no diagnostic internal to a standard run would reveal a
poor choice.

```{r fig-families, fig.width = 7, fig.height = 3.4, out.width = "100%", fig.cap = "The trajectory-family manipulation, at the calibration used throughout this paper. Panel (a): the biological-response mean curve under each of the four families, against cumulative time on drug, matched at the asymptote and at the week-5 anchor (dashed line). Panel (b): each family's deviation from the Gompertz reference. Dotted vertical lines and plotted points mark the six distinct on-drug durations the OL+BDC design actually samples (weeks 4, 8, 12, 16, 17, 18). The families are separated at the first two on-drug visits and have effectively converged by the third."}
t_grid    <- seq(0, 20, by = 0.05)
visit_tod <- c(4, 8, 12, 16, 17, 18)

mk <- function(tt, mode) do.call(rbind, lapply(names(fam_lab), function(f)
  data.frame(
    tod    = tt,
    family = fam_lab[[f]],
    y      = if (mode == 'level') traj(f, tt)
             else traj(f, tt) - traj('gompertz', tt),
    panel  = if (mode == 'level') '(a) Calibrated BR trajectory'
             else '(b) Deviation from Gompertz')))

d_fam  <- rbind(mk(t_grid, 'level'), mk(t_grid, 'dev'))
d_vis  <- mk(visit_tod, 'dev')
d_fam$family <- factor(d_fam$family, levels = unname(fam_lab))
d_vis$family <- factor(d_vis$family, levels = unname(fam_lab))

dev_early <- max(abs(d_vis$y[d_vis$tod <= 8]))
dev_late  <- max(abs(d_vis$y[d_vis$tod >= 12]))

ggplot(d_fam, aes(tod, y, colour = family)) +
  geom_hline(yintercept = 0, linewidth = 0.3, colour = 'grey70') +
  geom_vline(xintercept = visit_tod, linetype = 'dotted',
             colour = 'grey75', linewidth = 0.3) +
  geom_vline(xintercept = t_anchor, linetype = 'dashed',
             colour = 'grey40', linewidth = 0.4) +
  geom_line(linewidth = 0.6) +
  geom_point(data = d_vis, size = 1.5) +
  facet_wrap(~ panel, scales = 'free_y') +
  scale_colour_manual(values = fam_cols) +
  scale_x_continuous(breaks = c(0, 4, 5, 8, 12, 16, 20)) +
  labs(x = 'Time on drug (weeks)',
       y = 'BR response (CAPS-IV points)', colour = NULL) +
  theme_paper
```

Figure \@ref(fig:fig-families) makes the manipulation concrete at the
calibration this paper uses. Panel (a) shows that the four families
span a genuine range of shape envelopes: the piecewise-linear
breakpoint rises linearly and saturates abruptly at week
`r round(fam_par$piecewise_linear_breakpoint$p2, 1)`, the
hyperbolic-tangent is the most delayed early and then the steepest
through its inflection, and the Gompertz and logistic curves sit
between the two. Panel (a) also shows what the matched-anchor
calibration costs: because all four families are pinned to the same
asymptote and to the same value at week 5, they must cross there, and
the room left for them to differ is bounded. Panel (b) reports the
residual. Across the six on-drug durations the OL+BDC design
actually samples, the largest departure of any family from the
Gompertz reference is
`r sprintf('%.2f', dev_early)` CAPS-IV points at the first two visits,
against a BR residual standard deviation of
`r br_sd`, and falls to
`r sprintf('%.2f', dev_late)` points from week 12 onward. Two features
of the design conspire here: the response has essentially saturated by
the third on-drug visit, and the blinded-discontinuation block adds
visits that are closely spaced in calendar time but nearly identical
in time on drug. In this design, therefore, the trajectory-family
question is almost entirely a question about the first eight weeks on
drug, and any answer we report should be read as conditional on a
visit schedule that samples the discriminating region twice.

## How the Hendrickson framework treats the template, and what it estimates

The framework this paper evaluates [@hendrickson2020] treats the
trajectory template as a fixed calibration input rather than as an
object of inference, and it does so explicitly. The package ships a
single response-parameter set as data, one
$(\mathrm{maxr}, \mathrm{disp}, \mathrm{rate}, \sigma)$ quadruple per
latent component, documented as a deliberately noncommittal, 'tabula
rasa' set chosen so that the three components together could
plausibly reproduce the authors' existing prazosin-PTSD randomized
trial. The biological-response component is assigned
$\mathrm{maxr}_{BR} = `r sprintf('%.2f', maxr_v)`$ CAPS-IV total
severity points, $\mathrm{disp}_{BR} = `r br_disp`$,
and $\mathrm{rate}_{BR} = `r br_rate`$ per week;
the natural-history and expectancy components are given identical
parameters to one another, the authors having declined to assume any
difference between them. Published sensitivity analyses sweep the
*parameters* of that family, the rates and the maxima in particular,
but the family itself is held fixed throughout. The sweeps therefore
explore the interior of the Gompertz family and never test its
boundary, which is the gap the present paper addresses.

The estimand is unambiguous, and it is not the treatment effect. The
framework is built for predictive-biomarker validation, and its
analysis routine returns exactly one coefficient, documented in the
source as the coefficient corresponding to an interaction between
biomarker and outcome. That coefficient is \texttt{bm:Dbc}, the
biomarker-by-drug interaction, or, in the degenerate case where a
design leaves no within-participant variation in drug state, the
\texttt{bm:t} biomarker-by-time slope fitted on ever-on-drug
participants only. The drug main effect \texttt{Dbc} is carried in the
model as an adjustment; it is never extracted, never reported, and
never powered for. Power, in this framework, means power to detect
that the treatment effect *varies with* the biomarker, not power to
detect that the drug works. The treatment main effect is a separate
question, taken up in the companion manuscript at
\texttt{analysis/report/04-treatment-main-effect/}.

That separation is what gives the present question its structure, and
it is worth stating plainly because the framework as published does
not. The trajectory template governs the treatment effect: it sets
the BR mean curve, which is exactly the \texttt{Dbc} channel. The
estimand is an interaction. Whether the template can reach the
estimand at all therefore depends on how the biomarker moderation is
coupled to the trajectory, and that coupling is the architecture
choice, not a property of the template. Under Hendrickson's own
encoding the moderation is placed in the BM-BR block of the
correlation matrix while the template populates only the mean vector,
so template and estimand occupy formally disjoint parts of the
multivariate normal specification. Under a mean-moderation encoding
the moderation multiplies the trajectory and the two are coupled
directly. A reader of the original framework could reasonably take
the Gompertz response curve for an implementation detail; it is
better understood as a choice whose inferential relevance is
contingent on the interaction encoding, which is the finding this
paper reports in the Discussion.

## Why the question matters

First, the chosen family loads into the data before any
analysis-side question is asked. Consider a power simulation
that draws every trajectory from a Gompertz curve and then
analyzes each under a standard linear mixed-effects model with
treatment indicator and AR(1) residual correlation. Such a
simulation cannot detect the inferential cost of the family
choice itself. Robustness of the inference to the family is a
property of the joint DGP-and-analysis specification. Only a
comparative simulation that varies the DGP family while holding
the analysis fixed can quantify it. This is the same logic that
underpins analysis-model misspecification studies in the
broader clinical-trial methodology literature
[@morris2019simulation], applied one level down in the
simulation hierarchy. It appears to us that the N-of-1
simulation literature has not yet drawn the parallel explicitly.

Second, the function family has known identifiability fragility
in short noisy trajectories. Joint estimation of the three
Gompertz parameters $(m, d, r)$ in $y = m\exp(-d\exp(-rt))$ is
poorly conditioned: the asymptote and rate parameters correlate
strongly, and the shape parameter is essentially unidentifiable
when only a few timepoints lie in the inflection region.
@tjorve2010richards and @tjorve2017 propose unified-Richards
reparameterizations in which the parameters become biologically
interpretable (maximum relative growth rate, age at inflection,
value at inflection, asymptote) with lower correlation, and
@jagadeesan2023sloppiness develops the formal 'sloppiness'
diagnostic that quantifies this directly as a property of the
parameter space. @benzekry2014classical fitted eight
tumor-growth models (exponential, exponential-linear,
power-law, Gompertz, logistic, generalized logistic, von
Bertalanffy, dynamic-carrying-capacity) to two in vivo systems
and ranked them on goodness-of-fit, parametric
identifiability, and predictive power; the power-law model was
the most identifiable in lung data while Gompertz was best for
breast. The same kind of comparative diagnosis has not been
performed for short N-of-1 trajectories, and its absence is,
in our view, problematic.

Third, the default in N-of-1 simulation reflects historical
borrowing rather than empirical optimality. The Gompertz form
entered the pmsimstats framework via the broader growth-curve
tradition, where its use traces to Laird's tumor dynamics work
[@laird1964dynamics] and to Zwietering's microbial growth
reformulations [@zwietering1990modeling]. The pharmacology
literature, addressing related questions about dose-response
and time-course of pharmacological action, defaults to the
Hill / sigmoid-Emax family [@hill1910; @goutelle2008hill] and
to Jusko's indirect-response models [@dayneka1993indirect].
These two traditions have largely not spoken to each other; the
N-of-1 simulation literature has inherited the growth-biology
default without an explicit defense against the pharmacology
default. Either default may be adequate for the saturation
regime, but the inheritance is not itself evidence, and the
question merits a direct empirical answer.

## What the published N-of-1 literature does and does not do

A review of the recent N-of-1 methodology literature confirms,
to the best of our knowledge, a silence on the
parametric-trajectory-family question.

The CONSORT extension for N-of-1 trials [@vohraCONSORTExtensionReporting2015]
codifies reporting standards but does not address parametric
trajectory shape. The foundational programmatic statements
[@lillie2011; @schork2023precision] emphasize design and
analysis broadly without committing to a parametric response
form. The Hendrickson et al. framework [@hendrickson2020]
introduces the modified Gompertz template into the simulation
engine for biomarker-validation power analyses without
comparing it against alternatives. Recent applied papers in
chronic-condition N-of-1 trials (including the prazosin-PTSD
reference work building on @raskind2013) inherit the
framework without examining the choice. Unfortunately, this
pattern of silent inheritance is pervasive enough that no
published N-of-1 simulation study, to the best of our
knowledge, has broached the family-misspecification question.

In adjacent literatures the comparative apparatus does exist.
The oncology-modeling community routinely fits multiple growth
families and compares them on identifiability and predictive
performance [@benzekry2014classical]. The pharmacometrics
community produces explicit sensitivity studies for sigmoid-Emax
and indirect-response models [@ebling1996feasibility;
@triggs1996levodopa]. The latent-class growth-trajectory
literature treats response-curve heterogeneity through mixture
modeling [@hockenberry2017symptom]. None of this comparative
machinery has been applied to N-of-1 simulation with the
Gompertz form as the reference.

The present paper's contribution is to import that machinery,
specifically: a four-family comparison (Gompertz, symmetric
logistic, hyperbolic-tangent, piecewise-linear breakpoint)
crossed against the standard pmsimstats linear-mixed analysis,
with results reported in the Morris-White-Crowther
[@morris2019simulation] ADEMP framework already adopted as the
project-wide simulation-reporting standard.

## What this paper contributes

In section 2 we characterize the modified Gompertz family as
implemented in pmsimstats: the vertical-offset adjustment, the
parameter regions corresponding to the calibration sets used
across published pmsimstats simulation studies, and the
unified-Richards reparameterization [@tjorve2010richards;
@tjorve2017] that supplies more interpretable and less
correlated parameters.

In section 3 we survey three alternative response-curve
families: the symmetric logistic [@verhulst1838], the
hyperbolic-tangent (formally a rescaled symmetric logistic), and
a piecewise-linear breakpoint model. We present each at matched
marginal effect size against the Gompertz reference, with the
shapes overlaid for visual comparison.

In sections 4 and 5 we conduct a Monte Carlo sensitivity-analysis
simulation that generates trajectories under each of the four
families and analyzes them under the standard pmsimstats
linear-mixed model. We report power, Type I error, and the mean
and empirical standard error of the biomarker-treatment
interaction estimate under each cell, with conventional MCSEs.
The simulation borrows the comparative methodology of
@benzekry2014classical (eight-model tumor-growth comparison).
The sloppiness diagnostic of
@jagadeesan2023sloppiness motivates the identifiability framing
in this Introduction but is not itself computed in the present
16-cell focused submission; it is Study 2 of the ADEMP
pre-registration (`00-ademp-pre-registration.md`) and remains
future work, not a result reported in Sections 4-5 below.

In section 6 we identify the trajectory regimes in which
Gompertz misspecification would be expected to be consequential:
non-saturating growth beyond the trial window, biphasic patterns,
and breakpoint behavior. For these regimes, the standard N-of-1
inference is materially affected by the family choice, and we
recommend specific sensitivity-grid extensions that pmsimstats
users should run alongside their primary cells.

The full mathematical and historical context, including citations
that did not survive editing into the manuscript proper, is
preserved in the companion technical document at
\texttt{docs/25-gompertz-model-evaluation.md}.

# The modified Gompertz function in pmsimstats

The Gompertz curve originates in Gompertz's 1825 law of human
mortality [@gompertz1825] and was first applied as a general
growth curve by @winsor1932gompertz; the pmsimstats framework
adopts the modified, monotone-saturating form of that curve.
The framework parameterizes each of the three latent response
components (biological response BR, placebo-belief response PB,
and time-variant natural-history component TV) with a modified
Gompertz function

\begin{equation}
y(t) = \mathrm{maxr} \cdot \exp\!\left(-\mathrm{disp} \cdot
       \exp(-\mathrm{rate} \cdot t)\right),
(\#eq:gompertz)
\end{equation}

with vertical-offset adjustment so that $y(0) = 0$. The three
parameters are $\mathrm{maxr}$ (asymptotic maximum response,
matching the on-drug or on-natural-history saturation level),
$\mathrm{rate}$ (which controls how quickly the response
approaches its asymptote), and $\mathrm{disp}$ (which controls
the curvature near $t = 0$). The Gompertz form is
monotone-saturating and asymmetric: the inflection occurs at
$t^* = \log(\mathrm{disp}) / \mathrm{rate}$, with the rise
being relatively slower before the inflection and faster after.
The parameterization used at the prazosin/PTSD reference cell is the
\texttt{extracted\_rp} set shipped with the package, which sets
$\mathrm{maxr}_{BR} = `r sprintf('%.2f', maxr_v)`$ CAPS-IV total
severity points, $\mathrm{disp}_{BR} = `r br_disp`$, and
$\mathrm{rate}_{BR} = `r br_rate`$ per week for the biological
response, with a within-timepoint residual standard deviation of
`r br_sd` points. The expectancy and natural-history components are
assigned identical parameters to one another
($\mathrm{maxr} = `r sprintf('%.2f', as.list(extracted_rp[cat == 'tv'])$max)`$,
$\mathrm{disp} = 5$, $\mathrm{rate} = 0.35$); the full set is
documented in \texttt{docs/25-gompertz-model-evaluation.md} and in the
\texttt{extracted\_rp} data documentation.

# Alternative response-curve families

Three alternative families serve as sensitivity-analysis
comparators, and in this section we shall describe each in turn
before the simulation study proper. Each is parameterized so
that the asymptotic maximum and a single-anchor calibration
point (the response level at $t = 5$ weeks, calibrated to the
Gompertz family at the same anchor) are matched across families.
That is, the families differ in shape envelope (smoothness,
symmetry around the inflection, and curvature) while sharing
the same marginal effect size at the anchor, a design choice
that we discuss further in section 4.

## Symmetric logistic

The logistic function

\begin{equation}
y(t) = \frac{\mathrm{maxr}}{1 + \exp(-\mathrm{rate} \cdot
              (t - t_0))} - \mathrm{offset}
(\#eq:logistic)
\end{equation}

is monotone-saturating and symmetric around its inflection at
$t = t_0$. A vertical offset is subtracted so that $y(0) = 0$.
At matched $\mathrm{maxr}$ and matched response level at the
calibration anchor, the logistic family rises faster than
Gompertz before the inflection and slower after, producing a
visibly different shape envelope while sharing the same
asymptote. The symmetric inflection is the principal qualitative
contrast with the asymmetric Gompertz curve, and the reader will
note that this symmetry property is, in a sense, the logistic
family's most theoretically interesting feature for the purposes
of the present comparison.

## Hyperbolic-tangent

The hyperbolic-tangent variant

\begin{equation}
y(t) = \mathrm{maxr} \cdot \frac{\tanh(\mathrm{rate} \cdot
       (t - t_0)) + 1}{2} - \mathrm{offset}
(\#eq:tanh)
\end{equation}

is also monotone-saturating and symmetric, but with a
qualitatively faster approach to the asymptote than the
logistic. The tanh family produces the steepest pre-asymptote
rise of the three smooth families, and it has been used in
pharmacokinetic modeling for drugs with rapid receptor
saturation. We include it here as the representative of the
fast-saturating smooth regime that neither the logistic nor the
Gompertz family can span with their standard parameterizations.

## Piecewise-linear breakpoint

The piecewise-linear-breakpoint family, formalized in the
broken-line regression literature [@muggeo2003],

\begin{equation}
y(t) = \mathrm{maxr} \cdot \min(1,\, t / t_{\mathrm{bp}})
       + \mathrm{post\_slope} \cdot \max(0,\, t - t_{\mathrm{bp}})
(\#eq:plb)
\end{equation}

is the canonical non-smooth alternative: a linear ramp up to a
breakpoint $t_{\mathrm{bp}}$, then constant (when
$\mathrm{post\_slope} = 0$) or with a small post-breakpoint
slope. The breakpoint is the only non-differentiable feature
across the four families and is the principal qualitative
contrast with the three smooth families. This family is the
strongest available test of the framework's sensitivity to
non-smoothness in the trajectory. But how different is the
piecewise-linear breakpoint from the smooth families, from the
perspective of the Wald test that the analysis model applies?
That question is precisely what the simulation in section 4
is designed to answer.

# Sensitivity-analysis simulation design

The primary estimand is the biomarker-treatment interaction
coefficient $\hat{\beta}_{bm:D}$ from the linear mixed-effects
analysis model
\texttt{Sx \textasciitilde{} bm + t + Dbc + bm:Dbc + (1|ptID)}
with \texttt{corCAR1(form = \textasciitilde{}t|ptID)}.
Primary inferential outputs are
power for $H_0: \beta_{bm:D} = 0$ at $\alpha = 0.05$, and Type
I error under $\beta_{bm:D} = 0$. Secondary performance measures
are the mean and the empirical standard deviation of
$\hat{\beta}_{bm:D}$ across replicates. The ADEMP pre-registration
lists bias of $\hat{\beta}_{bm:D}$ as a secondary measure; it is
reported here only at the null, where the true value is zero by
construction. At the alternative the matched-anchor calibration
equalizes the marginal effect size across families but does not
deliver a closed-form true $\beta_{bm:D}$ on the CAPS-IV scale, so no
reference value is available to difference against and the mean
estimate is reported in place of a bias. Performance measures are
reported with binomial-proportion MCSEs for
proportions and the standard MCSE of the
within-replicate mean for continuous quantities.

The full Aims, Data-generating mechanisms, Estimands, Methods,
and Performance measures are pre-registered at
\texttt{analysis/scripts/gompertz-evaluation/00-ademp-pre-registration.md},
which fixes the four DGP families (modified Gompertz, symmetric
logistic, hyperbolic-tangent, piecewise-linear breakpoint), the
calibration sub-study (Study 0) that matches all four to a
common marginal effect size, the four-family power-and-bias
factorial (Study 1, 144 cells), and the sloppiness eigenvalue
diagnostic (Study 2) before the production runs.

For the present submission the simulation is scoped to a 16-cell
focused factorial: four trajectory families crossed with two DGP
architectures (mean moderation and MVN differential correlation,
following the companion manuscript at
\texttt{analysis/report/01-dgp-mean-moderation-vs-mvn/})
crossed with two biomarker effect-size levels
($c_{bm} \in \{0, 0.45\}$). The trial design is the
prazosin-calibrated Hybrid OL+BDC, $N = 70$, $t_{1/2} = 0$,
with carryover set to absent in order to isolate the
trajectory-shape effect from carryover-mediated correlation
decay. Following the compendium convention, $N$ is the total
across randomization paths: the driver allocates 35 participants
to each of the design's two paths, and the paths are concatenated
before analysis. Each cell is run at $n_{\text{reps}} = 1{,}000$
replicates with 8-core \texttt{parallel::mclapply}. The full
144-cell grid (sample sizes, effect sizes, DGP-side carryover)
anticipated in the original abstract is deferred to a follow-up;
the focused 16-cell factorial answers the central scientific
question (does trajectory family choice matter for the \texttt{bm:Dbc}
Wald test?) at MCSE 0.013--0.014 per alternative cell.

**Notation for the moderation parameter.** Because this factorial
crosses both architectures, $c_{bm}$ is used here as a common label
for the biomarker-moderation strength at a matched numeric value.
Strictly, the parameter is the correlation $c_{bm}$ under
the covariance-moderation architecture and the dimensionless multiplier $\beta_{bm}$ under
the mean-moderation architecture, the two being calibrated to a common scale because
the Architecture-A shift is applied as $\beta_{bm}\,b_i\,\sigma_{BR}$
on the standardized biomarker $b_i$. The companion architecture
manuscript keeps the symbols distinct; we adopt the shared label
only where a statement applies to both architectures at the matched
value, and the architecture is named explicitly wherever a result
is architecture-specific.

Calibration. The four trajectory families are matched at a
single anchor point ($t = 5$ weeks): each family's response
level at $t = 5$ is set equal to the Gompertz reference level
at $t = 5$ via uniroot (logistic, hyperbolic-tangent) or closed
form (piecewise-linear breakpoint). The asymptotic maximum is
held constant across families. Calibrated parameters are
recorded in the driver
\texttt{analysis/scripts/gompertz-evaluation/02-faithful-trajectory-sweep.R}.

Trajectory-family encoding. The family is routed through the
data-generating process rather than imposed post hoc: it sets
the BR mean curve inside the covariance construction via the
\texttt{br\_family} argument to \texttt{buildSigma}. Under
the mean-moderation architecture the biomarker-treatment
interaction is implemented in the multiplicative form of the
companion manuscript at
\texttt{analysis/report/01-dgp-mean-moderation-vs-mvn/},
$Y_{it} = \mu_0 + \beta_1 D_{it}(1 + \beta_{bm} B_i) +
\epsilon_{it}$, so the biomarker-dependent component of the
treatment effect rides the response trajectory $BR(t)$ rather
than entering as a flat shift. We normalize the per-timepoint
moderation so that its mean on-drug magnitude equals
$\sigma_{BR}$, holding the marginal interaction effect size
fixed across families while letting the temporal profile vary by
family shape. Under the covariance-moderation architecture (MVN differential
correlation) the interaction is encoded in the BM-BR covariance
block, which is independent of the BR mean; the family therefore
enters only through the mean curve and, as we verify by a
matched-seed check, cannot move the interaction test.

# Results

The simulation program was executed at 1{,}000 replicates per
cell across the 16-cell trajectory-family-by-architecture-by-
effect-size factorial in 597 s using 8-core
\texttt{parallel::mclapply}.
Convergence was 100\% across all 16{,}000 fits. Per-replicate
output is at
\texttt{analysis/data/quick-sim/07-gompertz-faithful-replicates.rds};
the Morris ADEMP cell-level summary is at
\texttt{analysis/data/quick-sim/07-gompertz-faithful-summary.txt}.
Cell MCSE on power runs from 0.013 to 0.014 across the
alternative cells and from 0.004 to 0.006 on the conservative
null cells. We begin the results section with power, then turn
to Type I error, and finally consider the sampling distribution of
the interaction estimate.

## Power across response-curve families

Trajectory-family power at $c_{bm} = 0.45$, percent $\pm$ MCSE.
The mean-moderation architecture is the faithful
trajectory-scaled mean-moderation form. The covariance-moderation
architecture is the MVN covariance channel:

| Family                    | Mean | Covariance |
|---------------------------|----------------|----------------|
| Gompertz                  | 74.3 $\pm$ 1.4 | 77.4 $\pm$ 1.3 |
| Logistic                  | 78.2 $\pm$ 1.3 | 79.1 $\pm$ 1.3 |
| Hyperbolic tangent        | 76.0 $\pm$ 1.4 | 77.7 $\pm$ 1.3 |
| Piecewise-linear breakpoint | 77.8 $\pm$ 1.3 | 80.4 $\pm$ 1.3 |

The two architectures behave qualitatively differently. Under
the covariance-moderation architecture (MVN differential
correlation), the biomarker-treatment interaction is carried
entirely by the covariance structure, which is independent of
the BR mean curve. Consequently the trajectory family, which
sets that mean curve, cannot alter the interaction test. We
confirm this directly. In a matched-seed check the four families
return bit-for-bit identical rejection decisions, so the
production-seed cell spread ($0.774$ to $0.804$) is sampling
noise rather than a family effect. Under the mean-moderation
architecture, implemented in the faithful multiplicative form in
which the biomarker-dependent component of the treatment effect
rides the response trajectory, the families separate: power
ranges from $0.743$ (Gompertz) through $0.760$
(hyperbolic-tangent) and $0.778$ (piecewise-linear) to $0.782$
(logistic), a $0.039$ spread. The Gompertz-logistic gap of
$0.039$ is approximately two standard errors of the difference
(SE $\approx 0.019$), and the Gompertz default is the
lowest-power, that is the mildly conservative, family of the
four. The pre-registered hypothesis that the piecewise-linear
breakpoint family would behave qualitatively differently from
the smooth families is not supported in the direction
anticipated. The breakpoint family tracks the logistic and the
hyperbolic-tangent, not the Gompertz default.

An earlier draft of this manuscript reported essentially no
family effect under either architecture. That result is now
understood to have been an artifact of the interaction encoding
rather than a property of trajectory family. In that draft the
Architecture-A biomarker moderation was applied as a flat
per-timepoint shift of magnitude $\beta_{bm}\sigma_{BR}$,
independent of the response trajectory. As a result the
trajectory family, which enters only through the BR mean curve,
could not affect the \texttt{bm:Dbc} interaction under the
mean-moderation architecture any more than under the
mean-independent covariance-moderation architecture. We then
implemented the mean-moderation architecture in its faithful
multiplicative form (section 4), in which the biomarker-dependent
component scales with the response trajectory in keeping with
the companion manuscript at
\texttt{analysis/report/01-dgp-mean-moderation-vs-mvn/}. Under
that implementation the mean-moderation families separate at the
few-percent level, while the covariance-channel invariance under
the covariance-moderation architecture is preserved. The latter
is a genuine mean-independence property rather than an artifact.

## Type I error across response-curve families

Type I error at $c_{bm} = 0$, percent $\pm$ MCSE:

| Family                    | Mean | Covariance |
|---------------------------|----------------|----------------|
| Gompertz                  | 1.9 $\pm$ 0.4  | 2.6 $\pm$ 0.5  |
| Logistic                  | 3.8 $\pm$ 0.6  | 2.2 $\pm$ 0.5  |
| Hyperbolic tangent        | 2.7 $\pm$ 0.5  | 2.6 $\pm$ 0.5  |
| Piecewise-linear breakpoint | 2.8 $\pm$ 0.5  | 2.7 $\pm$ 0.5  |

All eight null cells produce empirical Type I error in
$[0.019, 0.038]$, slightly conservative relative to the nominal
0.05 and consistent with the corCAR1 small-sample
degree-of-freedom accounting at the prazosin-calibrated $N = 70$
and 8-timepoint short serial series. At the null the biomarker
moderation term is zero, so the trajectory-scaling that
distinguishes the Architecture-A families under the alternative
is inactive here, and the null cells differ across families only
by sampling noise. There is no family-specific inflation; the
highest cell, logistic under the mean-moderation architecture at 0.038, is within
two MCSE of the others. The conservatism is shared across all
four families, supporting the interpretation that it reflects
the corCAR1 d.f. mechanism rather than a trajectory-shape
artifact, and it has appeared consistently across pmsimstats
simulation runs at the reference parameters.

## Sampling distribution of the biomarker-treatment interaction estimate

At the null the true $\beta_{bm:D}$ is zero by construction, and the
estimator is unbiased there: the mean estimate across the eight null
cells lies in $[-0.004, 0.003]$, within Monte Carlo noise of zero and
without family-specific drift. This is the only bias statement the
present design supports, and it holds for both architectures.

Across the eight alternative cells the mean estimated
$\hat\beta_{bm:D_{bc}}$ at $c_{bm} = 0.45$ varies over a narrow
range (roughly $-0.230$ to $-0.239$), a cell-to-cell spread of
about 0.009 that is itself within the Monte Carlo noise of the
estimator and shows no family-specific drift; the Gompertz and
logistic cells that differ most in power have nearly identical
mean estimates. The empirical SE of $\hat\beta_{bm:D_{bc}}$ is
essentially constant across families (SD in the range 0.073 to
0.083 across cells) and roughly equal between architectures.
That the Architecture-A power separation appears with a
family-invariant mean estimate and near-constant empirical SE
indicates it is a precision
effect operating through the test statistic's sensitivity to the
temporal profile of the signal, not a shift in the estimand or
its sampling dispersion. The matched-anchor calibration holds
the marginal effect size fixed across families by construction,
so the residual family differences are differences in the
temporal shape of the moderation, which is precisely the axis
the design isolates.

# Discussion

## An architecture-dependent finding

The central finding is that the inferential consequence of
trajectory-family choice is architecture-dependent. At the
prazosin-calibrated reference cell (Hybrid OL+BDC, $N = 70$,
$c_{bm} = 0.45$, $t_{1/2} = 0$, matched-anchor calibration),
the four families span the qualitative range of plausible
templates for monotone-saturating treatment-response
trajectories: asymmetric (Gompertz), symmetric smooth (logistic
and hyperbolic-tangent), and non-smooth (piecewise-linear
breakpoint). Under the covariance-moderation architecture, where the interaction is
carried by the BM-BR covariance and is therefore independent of
the BR mean curve, the family is inert by construction. Under
the mean-moderation architecture, where the interaction is a mean effect that, in
its faithful multiplicative form, scales with the response
trajectory, the families separate over a 0.039 power range and
the Gompertz default is the lowest-power, that is the mildly
conservative, family. We would not wish to overstate the
Architecture-A effect; it is modest, on the order of two Monte
Carlo standard errors at this sample size and effect size, and
the conditions under which it would grow or shrink are discussed
in section 6.3. But it is real and direction-consistent, and it
overturns the apparent null reported in the earlier flat-
moderation draft.

The pre-registered hypothesis that the piecewise-linear
breakpoint family, with its non-differentiable feature, would
behave qualitatively differently from the smooth families is not
supported in the direction anticipated. The breakpoint family
carries no distinctive power signature attributable to its
non-smoothness; under the mean-moderation architecture it sits with the logistic
and hyperbolic-tangent families, above the Gompertz default, and
under the covariance-moderation architecture it is inert like the others.

What the operative axis is instead, we cannot say from this design,
and we decline to name one. The obvious candidate, the level the
response has reached at the visits that discriminate the families, does
not order the powers. Figure \@ref(fig:fig-families) shows that the
two on-drug visits carrying essentially all of the family contrast are
weeks 4 and 8. At week 4 the hyperbolic-tangent family is lowest and
the piecewise-linear highest; at week 8 the ordering reverses, with
Gompertz lowest and hyperbolic-tangent highest. Neither ordering
matches the power ranking, in which Gompertz is lowest and the
hyperbolic-tangent second lowest. Nor is the ranking itself firmly
established: with four families compared post hoc and no multiplicity
adjustment, the 0.039 Gompertz-logistic gap is the largest of six
pairwise contrasts and is the only one that separates at even two
standard errors. The safe reading is that the Gompertz default is not
above the others in power and may be slightly below them, and that
identifying the mechanism requires the extended grid of section 6.3
rather than further interpretation of these sixteen cells.

## Why the architectures differ

We see two mechanisms that, taken together, account for the
architecture contrast, and we shall work through each in turn.
The first is the exact mean-independence of the covariance
channel under the covariance-moderation architecture. The second
is the residual profile-sensitivity of the trajectory-scaled
mean channel under the mean-moderation architecture. The
matched-anchor calibration and the time-averaging of the Wald
test damp this second mechanism but do not eliminate it.

The first mechanism damps the Architecture-A effect, because
calibration removes the amplitude differences and leaves only
profile differences. The four families are matched at the
asymptotic maximum and at a single anchor point ($t = 5$ weeks),
and the trajectory-scaled moderation is normalized to a common
mean on-drug magnitude. At matched marginal effect size, the
residual within-trajectory variation in shape is small relative
to the between-replicate variation in the \texttt{bm:Dbc}
estimator. The calibration constraint is intentional in the
simulation design, since it isolates trajectory-shape effects
from amplitude effects. That is why the family-specific power
differences we observe under the mean-moderation architecture
can be attributed to the temporal profile of the moderation
alone, and why that effect is on the order of a few percentage
points rather than larger.

The second mechanism, also damping, is that the \texttt{bm:Dbc}
Wald test averages over time. The continuous $D_{bc}$ predictor
multiplied by the biomarker integrates the trajectory's
contribution across all on-drug timepoints, with weights
determined by the predictor's exponential-decay structure.
Trajectory-shape differences, where one family rises faster than
another between timepoints, are partly averaged within the
integration. The Wald test is therefore relatively, though not
perfectly, insensitive to profile features. Under the
mean-moderation architecture the averaging is incomplete: the
residual profile difference survives into the test statistic and
produces the observed few-percent power spread. Under the
covariance-moderation architecture there is no profile term to
survive, because the interaction does not enter the mean at all.

Mean-independence applies specifically to the
covariance-moderation architecture, where the covariance
moderation is upstream of trajectory shape. Under MVN
differential correlation, the biomarker-treatment information
lives in the $\partial \mathrm{Cor}(BM, BR)/\partial D_{bc}$
structure, which is independent of the trajectory family that BR
follows. The family enters only the BR mean, to which the
interaction test is invariant. This is not a small-effect
statement but an exact one, and we verify it directly. Holding
the random-number seed fixed across families, the four
Architecture-B cells return bit-for-bit identical rejection
decisions. The production-seed spread under the
covariance-moderation architecture is therefore sampling noise,
whereas the Architecture-A spread is a genuine, if modest,
profile effect.

The combined effect is that trajectory-family choice is
inferentially free under the covariance channel and carries a
modest, mildly conservative, cost under the mean channel. The
framework's modified Gompertz default is robust by exact
construction under the covariance-moderation architecture. Under
the mean-moderation architecture it is the lowest-power of the
four families, which makes it a safe rather
than an optimal default: a trial designer who adopts it under a
mean-moderation DGP will, if anything, slightly under-state
achievable power. We turn next to the conditions under which the
Architecture-A effect would be expected to grow.

## Conditions that would change the answer

Four conditions, none exercised in the present 16-cell
factorial, would plausibly amplify the modest Architecture-A
family effect, and each merits a brief comment. Because
the covariance-moderation architecture is mean-independent, the first three conditions
(sample size, effect size, carryover) bear on the mean-moderation architecture
only; the fourth, non-monotone shape, is the only one that could
in principle reach the covariance-moderation architecture, and then only by changing the
covariance construction rather than the mean.

First, smaller sample size. At $N = 70$ the \texttt{bm:Dbc} test
integrates over a moderately large within-subject sample;
smaller $N$ might amplify the local-shape sensitivity by
concentrating the temporal information at fewer timepoints
where the families differ more visibly.

Second, larger or smaller $c_{bm}$. The $c_{bm} = 0.45$
reference cell sits in a power regime where families have
comparable signal-to-noise. Pushing $c_{bm}$ to 0.30 (where
power lies near 0.5) might surface family differences via
different finite-sample sampling distributions, and it is
possible that the tight clustering we observe is a feature of
this particular region of the effect-size space.

Third, carryover-induced correlation decay. With $t_{1/2} > 0$,
the \texttt{bm:Dbc}-mediated information channel is eroded differently
for different trajectory shapes, because the integration weights
interact with the off-drug exponential decay. The $t_{1/2} = 0$
slice exercised here suppresses this interaction entirely.

Fourth, non-monotone or biphasic shape envelopes. The four
families we have examined share a monotone-saturating envelope.
Families that peak and decay (biphasic) or oscillate would
produce genuinely different trajectories with different \texttt{bm:Dbc}
information content, and the Wald test would almost certainly
respond to those differences.

Each condition is the natural axis of a follow-up sensitivity
analysis. The 144-cell grid anticipated in the abstract
($N \times c_{bm} \times t_{1/2} \times $ family) is the
canonical follow-up, and we recommend it as the production-grade
extension that would close out the trajectory-family question
at a level of rigor adequate for the methodological literature.

## The prazosin/PTSD reference cell

For the motivating prazosin-PTSD application that drives the
pmsimstats research program, the present result implies that
sensitivity to the modified Gompertz assumption is small relative to
sensitivity to the carryover specification (the subject of the
companion manuscript at
\texttt{analysis/report/02-carryover-sensitivity/}) and to
the DGP architecture (the subject of
\texttt{analysis/report/01-dgp-mean-moderation-vs-mvn/}). Of the
three principal axes of sensitivity uncertainty in the
pmsimstats simulation framework, trajectory family choice is, at
the reference parameters, the least consequential by a
substantial margin. It is not, however, entirely free: under the
mean-moderation DGP the modified Gompertz default is the lowest-power
family by a few percentage points, so a trial designer who
adopts it is making a mildly conservative choice. Under the
covariance-channel DGP the choice is inferentially free. We
would suggest that sensitivity-analysis attention be directed
chiefly to the carryover and DGP-architecture axes, where the
inferential consequences are larger, while reporting at least
one Gompertz-vs-alternative comparison whenever the operative
DGP is mean moderation.

# Conclusions

First, the modified Gompertz function is a defensible
parametric template for monotone-saturating treatment-response
trajectories in N-of-1 simulation, with an architecture-dependent
qualification. At the prazosin/PTSD reference cell with
matched-anchor calibration, the four-family comparison (Gompertz,
symmetric logistic, hyperbolic-tangent, piecewise-linear
breakpoint) produces exactly equivalent inference under the
covariance-channel architecture, where the interaction is
mean-independent, and a modest 0.039 power spread under the
mean-moderation architecture, where the biomarker moderation
rides the response trajectory. In the latter case the Gompertz
default is the lowest-power of the four families, which makes it
a mildly conservative rather than an optimal default. The
pre-registered hypothesis that the piecewise-linear breakpoint
family would behave qualitatively differently from the smooth
families is not supported in the direction anticipated.

Second, the finding is methodologically informative precisely
because it is architecture-dependent. It implies that the \texttt{bm:Dbc}
Wald test is exactly robust to trajectory-family choice when the
interaction is encoded in the covariance, and only mildly
sensitive to it, at the level of a few percentage points of
power, when the interaction is a trajectory-scaled mean effect.
We cannot yet establish at what sample sizes, effect sizes, or
carryover levels the Architecture-A effect would grow; that
question is deferred to the 144-cell follow-up. Trial designers
using the pmsimstats framework can adopt the modified Gompertz default
with confidence under a covariance-channel DGP, and with the
understanding that it is mildly conservative under a
mean-moderation DGP, while directing sensitivity-analysis
attention to the carryover and DGP-architecture axes that have
larger inferential consequences.

Third, the 144-cell extended grid (sample sizes, effect sizes,
DGP-side carryover) anticipated in the original abstract is the
natural follow-up, and we regard it as the work that would be
needed to close the trajectory-family question at a level of
rigor adequate for the methodological literature. The present
focused 16-cell factorial addresses the central scientific
question with statistical clarity; the extended grid would map
the boundary conditions at which trajectory-family choice
begins to matter.

# Conflict of interest

The authors declare no conflicts of interest.

# Funding

This work received no specific funding.

# Data availability statement

The data and code that support the findings of this study are
openly available in the pmsimstats-ng project repository.
Per-replicate Monte Carlo outputs and ADEMP-compliant
cell-level summaries are versioned under
\texttt{analysis/data/}; simulation drivers are at
\texttt{analysis/scripts/}. Exact paths for this manuscript
are listed in the Reproducibility section below.

# Reproducibility

This manuscript was rendered with R 4.4.x inside the project's
zzcollab Docker container (image hash recorded in
\texttt{Dockerfile}; package set captured in
\texttt{renv.lock}). Principal packages used by the simulation
pipeline are \texttt{nlme} (continuous-time linear-mixed
analysis with \texttt{corCAR1} residual correlation),
\texttt{parallel} (8-core \texttt{mclapply} for
parameter-grid sweeps), \texttt{data.table} (the
\texttt{original-extended} simulation collection),
\texttt{furrr} and \texttt{purrr} (the \texttt{tidyverse}
simulation collection), and \texttt{rmarkdown} plus
\texttt{bookdown} for the manuscript build.

The simulation driver
\texttt{analysis/scripts/gompertz-evaluation/02-faithful-trajectory-sweep.R}
sets \texttt{set.seed(20260617)} and assigns each of the 16 cells
a distinct seed offset, so replicate streams are independent
across cells but reproducible. The trajectory family is routed
through the data-generating process via the \texttt{br\_family}
argument to \texttt{buildSigma}, and the Architecture-A
moderation is trajectory-scaled via the
\texttt{moderation\_scaling = "trajectory"} argument to
\texttt{generateData} (both default to the original behavior, so
all other pmsimstats runs are unchanged). Per-replicate
simulation outputs are checkpointed at
\texttt{analysis/data/quick-sim/07-gompertz-faithful-replicates.rds};
Morris ADEMP cell-level summaries are at the companion
\texttt{07-gompertz-faithful-summary.txt}. The pre-registered
ADEMP plan is at
\texttt{analysis/scripts/gompertz-evaluation/00-ademp-pre-registration.md},
and deviations from it are recorded in
\texttt{02-deviations-log.md}. The manuscript source,
paper-specific bibliography (\texttt{references.bib}), and SIM
CSL file are at
\texttt{analysis/report/07-gompertz-evaluation/}.

# References

::: {#refs}
:::
