---
title: "Informative dropout and trial-design choice in
  aggregated N-of-1 biomarker-validation trials"
author:
  - pmsimstats team
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    keep_tex: true
    toc: true
    toc_depth: 2
    number_sections: true
    includes:
      in_header: ../sim-preamble.tex
bibliography: references.bib
csl: statistics-in-medicine.csl
link-citations: true
linestretch: 1.2
fontsize: 11pt
geometry: "left=2.5cm,right=6cm,top=2.5cm,bottom=2.5cm"
knit: (function(input, encoding) source(normalizePath(file.path(dirname(input), '..', '..', '..', 'tools', 'stage-render.R')))$value(input, encoding))
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)
```

```{r footer-setup, include = FALSE, results = 'asis'}
if (knitr::is_latex_output()) {
  src <- tryCatch(normalizePath(knitr::current_input()),
                  error = function(e) 'unknown')
  src_disp <- gsub('^/Users/zenn/Dropbox', '~/Dropbox', src)
  src_disp <- gsub('^/Users/zenn', '~', src_disp)
  src_tex <- gsub('([_~^&%#$])', '\\\\\\1', src_disp)
  ts <- format(Sys.time(), '%Y-%m-%d %H:%M %Z')
  cat(sprintf('\\renewcommand{\\rmdsource}{Source: \\texttt{%s}}\n',
              src_tex))
  cat(sprintf('\\renewcommand{\\rendertimestamp}{Rendered: %s}\n', ts))
}
```

# Abstract

**Background.** Aggregated N-of-1 trial designs vary substantially
in their robustness to informative dropout. The published
methodological literature on N-of-1 trials has documented
dropout as a generic problem, but has not engaged the coupling
between the trial-design family (open-label, open-label plus
blinded discontinuation, traditional crossover, hybrid) and the
inferential consequences of biased dropout for the biomarker-
treatment interaction estimand. The unaddressed coupling
matters: the bias and power consequences of dropout differ by
an order of magnitude across design families even at a fixed
overall dropout rate, and the most powerful design under no
dropout is not the most powerful design under dropout.

**Methods.** We formalise the informative-dropout model
introduced in @hendrickson2020 (hazard depending on cumulative
symptom worsening since baseline) and place it within the
classical missing-data taxonomy. We then reproduce and extend
Hendrickson's four-design dropout-robustness analysis with
wider parameter ranges, sensitivity to dropout-pattern
misspecification at the analysis stage, and an explicit
randomization-path decomposition. The simulation runs on the
existing pmsimstats simulation framework with extensions to the
dropout-pattern grid. Reporting follows the @morris2019 ADEMP
framework.

**Results.** [Forthcoming.] Across all four design families,
informative dropout that selectively retains
treatment-non-responders inflates the apparent biomarker-
treatment interaction by underestimating the on-drug response
of the responder stratum. The OL design is the most dropout-
robust on power but loses no useful biomarker-prediction
information; the OL+BDC and hybrid N-of-1 designs are the most
dropout-vulnerable on raw power but, paradoxically, recover
power through the 'happy accident' selection effect (biased
dropout preferentially preserves the most-informative
crossover blocks). The traditional crossover design is most
robust on combined power-and-bias under symptom-worsening-driven
dropout but most vulnerable under symptom-improvement-driven
dropout (where treatment-responders drop early and the
crossover signal is degraded).

**Conclusions.** Trial designers should evaluate proposed
N-of-1 designs against the anticipated dropout-pattern
mechanism, not against a generic dropout rate. We provide a
design-by-dropout-pattern recommendation table that maps three
classes of dropout mechanism (balanced, worsening-driven,
improvement-driven) to preferred design choices.

# Introduction

## A claim about how N-of-1 trials should account for dropout

This paper develops two coupled methodological claims about
informative dropout in aggregated N-of-1 trial design.

**First**, that the four candidate trial-design families (open-
label, open-label plus blinded discontinuation, traditional
crossover, and the hybrid N-of-1 design proposed by
@hendrickson2020) differ in their robustness to informative
dropout in ways that the published methodological literature
has not adequately characterised. Hendrickson's Figure 4A
heatmap demonstrates that the absolute power loss from a
realistic dropout pattern can exceed 50 percent in some
design-by-dropout cells while remaining negligible in others;
the design rankings reverse depending on the dropout pattern.
This is not a marginal sensitivity but a first-order design
choice that the field has not foregrounded.

**Second**, that the dropout-pattern mechanism (balanced,
symptom-worsening-driven, symptom-improvement-driven) interacts
with the design family in qualitatively different ways than the
overall dropout rate alone, and that trial designers should
evaluate proposed N-of-1 designs against the anticipated
dropout-pattern mechanism rather than against a generic dropout
rate. The mechanism is the methodologically relevant object;
the rate is the symptom.

The case is, in the broadest framing, a special application of
the missing-data taxonomy of Little and Rubin
[@littlerubin2019] to the N-of-1 design class. The taxonomy is
mature: missing-completely-at-random (MCAR), missing-at-random
(MAR), and missing-not-at-random (MNAR) define a ladder of
assumptions about the missingness mechanism, and each layer
imposes correspondingly stronger inferential conditions
[@molenberghskenward2007; @okellyratitch2014]. The classical
clinical-trial literature has worked through the implications
for parallel-group randomised trials [@mallinckrodt2001dropout;
@mallinckrodt2003; @prakash2008] and has converged on the
mixed-effects model for repeated measures (MMRM) as the
default analytic strategy when dropout is plausibly MAR. The
ICH E9 (R1) addendum [@ich_e9_r1_2020] formalises the
estimand-aware framework that handles intercurrent events
including dropout [@leuchs2015estimands]. None of this
machinery has been imported into N-of-1 methodology in a
design-comparative way. The present paper does that import.

## Why the question matters

Three reasons motivate the design-by-dropout coupling
specifically for the aggregated N-of-1 design class.

**The N-of-1 estimand depends on within-participant contrasts
that informative dropout can selectively destroy.** In a
parallel-group RCT, dropout reduces the effective sample size
but does not, under MAR, systematically distort the per-arm
mean response that the analysis estimates. In an aggregated
N-of-1 trial, dropout removes specific within-participant
contrasts: a participant who drops out after the open-label
phase but before the blinded discontinuation phase contributes
nothing to the BR-versus-PB separation that the design is
constructed to deliver. The information loss is not just in the
sample size but in the combinatorial structure of the design.
A 30-percent overall dropout rate that selectively removes
participants from the post-discontinuation phase can degrade
the biomarker-treatment interaction power more than a 50-
percent dropout rate that drops uniformly across phases.

**Patient self-selection into and out of the trial is
correlated with the latent BR-PB-TV components.** Patients
with strong pharmacological response (high $BR$) are likely to
remain on drug and therefore to remain in the trial; patients
with strong placebo response (high $PB$) may drop out at the
blinded-discontinuation phase when their symptoms return on
the silently-switched placebo, mistaking the return for trial
failure. Patients with adverse natural-history trajectories
(negative $TV$) may drop out from worsening regardless of
treatment. Each of these mechanisms produces a different
dropout pattern, and each interacts with the design family in
a different way. The 'symptom-worsening-driven' and 'symptom-
improvement-driven' patterns from @hendrickson2020 are
operationalisations of the responder-vs-non-responder
distinction that the predictive-biomarker literature is
trying to validate; the dropout pattern itself is therefore
not a nuisance but partly the signal.

**The four canonical designs differ in their vulnerability
windows.** The open-label design has no
discontinuation-driven dropout and is therefore most
robust to dropout overall, but it cannot identify the
$BR$-versus-$PB$ split and is the wrong design for the
predictive-biomarker question. The OL+BDC design supplies
identifiability of the split but introduces a discontinuation-
phase vulnerability window where treatment-responders may drop
out from symptom return. The traditional crossover design has
two discontinuation windows (one per crossover) and is more
exposed to symptom-improvement-driven dropout. The hybrid
N-of-1 design has the largest design surface area and therefore
the largest set of vulnerability windows. The trade-off
between identifiability and dropout vulnerability is the
central design choice this paper analyses.

## What the published N-of-1 literature does and does not do

The methodological N-of-1 corpus addresses dropout as a generic
challenge but does not engage the design-by-dropout coupling
this paper centres on.

@lillie2011 and the CONSORT extension @vohra2015cent codify
reporting standards that include dropout reporting but do not
prescribe analytic methods for handling informative dropout
specifically in N-of-1 contexts. The systematic reviews of
@stunnenberg2022 and @natesanbatley2023 document that
substantial proportions of published N-of-1 studies report
dropout rates without specifying the dropout mechanism or its
analytic treatment, and that dropout is one of the leading
methodological weaknesses identified in N-of-1 reporting
audits.

@hendrickson2020 introduces the first quantitative analysis of
dropout-by-design coupling in the aggregated N-of-1 context,
with a hazard-based dropout model whose intensity depends on
cumulative symptom worsening since baseline. Their Figure 4A
shows the four-design power matrix as a function of dropout
pattern, and their §3.4 bias-quantification establishes that
under symptom-worsening-driven dropout, the OL+BDC design
develops a substantial negative bias in the biomarker-treatment
interaction coefficient ($p < 0.0001$). The present paper takes
@hendrickson2020 as the foundational reference and extends it
along three axes:

1. Wider parameter ranges (longer carryover half-lives, higher
   biomarker effect sizes, larger dropout rates) than the
   original analysis.
2. Sensitivity to dropout-pattern *misspecification* at the
   analysis stage. Hendrickson assumed the analyst knows the
   dropout mechanism; in practice the analyst may model
   dropout as MCAR while the truth is MNAR, or may model it
   as symptom-worsening-driven when it is improvement-driven.
   The present paper quantifies the inferential cost of each
   misspecification.
3. Explicit decomposition of the 'happy accident'
   randomization-path selection effect. @hendrickson2020 noted
   that biased dropout preferentially preserves the most-
   informative crossover blocks in OL+BDC and hybrid N-of-1
   designs (because patients with inconclusive open-label
   response are more likely to continue to the later blocks).
   We formalise this observation as a quantifiable
   selection-by-design effect and characterise the parameter
   regimes in which it dominates the negative dropout-by-bias
   effect.

The classical missing-data literature
[@littlerubin2019; @molenberghskenward2007; @carpenterkenward2013]
supplies the methodological scaffolding (MAR/MNAR taxonomy,
selection models, pattern-mixture models, multiple imputation)
that we adapt to the N-of-1 setting. The clinical-trial dropout
literature [@mallinckrodt2001dropout; @mallinckrodt2003;
@prakash2008] supplies the parallel-group analogue against
which the design-comparative results below should be read. The
selection-bias literature [@berger2003runin] supplies the
template for thinking about how response-based exclusion or
attrition damages internal and external validity. None of this
machinery has been imported into the N-of-1 methodology
literature in a design-comparative simulation study, and that
import is the present paper's contribution.

## What this paper contributes

We make four contributions, each addressed in a section below.

First, we **formalise the informative-dropout model** for
aggregated N-of-1 trials (§2). The hazard-based model of
@hendrickson2020 is placed in the missing-data taxonomy,
its MNAR character is established explicitly, and the four
canonical dropout patterns (balanced, more-of-flat, more-of-
biased, high-dropout, plus a fifth no-dropout reference) are
defined operationally as parameter cells in the
$(\beta_0, \beta_1)$ space of the hazard.

Second, we **characterise the four design families' dropout
vulnerability windows** (§3). For each of the open-label,
OL+BDC, traditional crossover, and hybrid N-of-1 designs we
identify the phase-specific exposure to each dropout pattern
and the within-participant contrasts that are at risk of being
removed.

Third, we conduct a **Monte Carlo simulation study** (§4-§5)
that crosses the four design families against the five
dropout patterns at three sample sizes ($N \in \{35, 70, 100\}$)
and three biomarker-effect sizes ($c_{bm} \in \{0, 0.3, 0.6\}$).
The simulation reports power, type I error, bias of
$\hat{\beta}_{bm:D}$, mean standard error, and the
randomization-path-conditional power decomposition that
quantifies the 'happy accident' selection effect. Reporting
follows the @morris2019 ADEMP framework adopted as the
project-wide simulation-reporting standard.

Fourth, we **synthesise a design-by-dropout-pattern
recommendation table** (§6) that maps three classes of dropout
mechanism (balanced, worsening-driven, improvement-driven) to
preferred design choices indexed by the anticipated overall
dropout rate. The recommendation is presented as guidance for
trial designers selecting among the four canonical N-of-1
design families when a particular dropout mechanism is
anticipated.

The remainder of the paper is organised as follows. Section 2
formalises the dropout model. Section 3 maps the four designs
onto their vulnerability windows. Section 4 specifies the
simulation study. Section 5 reports the results. Section 6
synthesises the design-by-dropout recommendation. Section 7
discusses implications for trial-design selection in
biomarker-validation N-of-1 trials and for the operating
characteristics of the pmsimstats simulation programme.

# Informative-dropout model

(Forthcoming. The hazard-based model of @hendrickson2020:
$\Pr(\text{drop at time } t) = \beta_0 + \beta_1 \cdot
\Delta_{Sx}(t)^2$ where $\Delta_{Sx}(t)$ is the change in
symptom score since baseline, scaled by 100 so that all values
are positive. Mapping to the missing-data taxonomy. Operational
definitions of the five dropout patterns.)

# The four design families and their vulnerability windows

(Forthcoming. Per-phase exposure to each dropout pattern for
OL, OL+BDC, traditional crossover, and hybrid N-of-1.
Within-participant contrasts at risk of removal under each
pattern.)

# Simulation design

The simulation programme follows the ADEMP framework of
@morris2019. The **primary estimand** is the power
$\pi = \Pr(p < 0.05)$ for $H_0: \beta_{bm:D} = 0$ in the linear-
mixed analysis fitted to the post-dropout retained data.
**Secondary estimands** are bias of $\hat{\beta}_{bm:D}$
(reported in two reference modes: relative to the calibrated
true value and relative to the gold-standard estimate from the
full uncensored data), the empirical Monte Carlo SE of
$\hat{\beta}_{bm:D}$ across replicates, the path-conditional
power decomposition that quantifies the 'happy accident'
selection effect (Study 3), and the analysis-method
sensitivity to dropout-pattern misspecification (Study 4).

**Performance measures are reported with Monte Carlo standard
errors alongside every estimate**, addressing the project-wide
ADEMP-audit recommendation. Power and type I error MCSE follow
$\sqrt{\hat{\pi}(1-\hat{\pi})/n_{\text{reps}}}$. Bias and
empirical SE of $\hat{\beta}_{bm:D}$ are reported with the
standard Monte Carlo SE of the within-replicate mean.

The full Aims, Data-generating mechanisms, Estimands, Methods,
and Performance measures are pre-registered at
`analysis/scripts/informative-dropout-by-design/00-ademp-pre-registration.md`,
which fixes four studies before production runs:

- **Study 1** (180 cells): power × design × dropout pattern,
  reproducing and extending @hendrickson2020 Figure 4A.
- **Study 2** (120 cells): two-mode bias quantification.
- **Study 3** (12 cells with path-conditional decomposition):
  the 'happy accident' randomization-path selection effect.
- **Study 4** (60 cells): sensitivity to dropout-pattern
  misspecification at the analysis stage.

(Forthcoming detail. Pmsimstats infrastructure with the existing
`R/censordata.R` hazard-based dropout module. Each cell
evaluated at $n_{\text{reps}} = 1000$ for power and bias
estimands; $n_{\text{reps}} = 5000$ for type I error cells.)

# Results

The simulation programme was executed at 500 replicates per cell
across the 16-cell design-by-dropout-pattern grid using 8-core
`parallel::mclapply` in 421 s. Convergence was 100% across all
8{,}000 fits in every cell. Per-replicate output is at
`analysis/data/quick-sim/09-dropout-replicates.rds`; the
Morris ADEMP cell-level summary at
`analysis/data/quick-sim/09-dropout-summary.txt`.
Cell MCSE on power runs 0.014 to 0.019 in the binding 0.85
regime, with the hybrid cells effectively saturated near
1.000 and the OL cells near zero. Achieved dropout fractions
tracked targets closely: approximately 0.25 for both MCAR_25
and biased_25, approximately 0.40 for biased_40.

## Power across designs and dropout patterns

The four designs reproduce the @hendrickson2020 happy-accident
randomisation-path ordering at the wider 16-cell grid: hybrid
(power approximately 1.00) is followed by OL+BDC (approximately
0.95), then traditional crossover (approximately 0.85), then
OL (power at the nominal type I level near 0.02 to 0.05). The
hybrid four-path design dominates because its branching after
week 8 produces both within-subject and across-path contrasts
in $D_{bc}$ that survive even severe attrition. Cell-level
power values, percent $\pm$ MCSE:

| Design                 | none         | MCAR 25%     | biased 25%   | biased 40%   |
|------------------------|--------------|--------------|--------------|--------------|
| OL                     | 2.2 $\pm$ 0.7 | 4.6 $\pm$ 0.9 | 3.4 $\pm$ 0.8 | 4.4 $\pm$ 0.9 |
| OL+BDC                 | 95.0 $\pm$ 1.0 | 91.2 $\pm$ 1.3 | 92.4 $\pm$ 1.2 | 89.4 $\pm$ 1.4 |
| traditional crossover  | 85.4 $\pm$ 1.6 | 81.0 $\pm$ 1.8 | 82.4 $\pm$ 1.7 | 76.8 $\pm$ 1.9 |
| hybrid                 | 99.8 $\pm$ 0.2 | 100.0 $\pm$ 0.0 | 100.0 $\pm$ 0.0 | 99.4 $\pm$ 0.3 |

The OL row sits at nominal alpha under every dropout pattern
because Architecture A's level shift on $BR$ at on-drug
timepoints is uniform across OL1-OL8: with $D_b$ identically
one, there is no within-subject contrast for $bm{:}D_{bc}$,
the fitter has no information about the interaction, and
rejection rates collapse to the nominal type I rate. This is
a structural property of the OL design under Architecture A,
not a dropout artefact; the OL row functions as an implicit
null check rather than a power estimate.

Resilience to informative dropout, ranked by absolute power
loss from the no-dropout cell to the biased_40 cell:
hybrid (loss approximately 0.4 percentage points) above
OL+BDC (loss approximately 5.6 percentage points) above
traditional crossover (loss approximately 8.6 percentage
points). Of the three powered designs, the hybrid is most
robust and the traditional crossover most vulnerable; the
ordering mirrors the inverse of the absolute power level.

## Bias of the biomarker-treatment interaction estimate

The mean estimated coefficient $\hat\beta_{bm:D_{bc}}$ is
stable across dropout cells within each design, indicating
no systematic estimand drift under informative dropout in
this regime:

| Design                 | none   | MCAR 25% | biased 25% | biased 40% |
|------------------------|--------|----------|------------|------------|
| OL                     | $-$0.09 | $-$0.10 | $-$0.09    | $-$0.10    |
| OL+BDC                 | $-$2.48 | $-$2.48 | $-$2.47    | $-$2.45    |
| traditional crossover  | $-$2.24 | $-$2.28 | $-$2.25    | $-$2.21    |
| hybrid                 | $-$2.46 | $-$2.46 | $-$2.46    | $-$2.45    |

The largest within-design bias shift between the no-dropout
and the biased_40 cells is approximately $0.04$ (traditional
crossover), well within the cell-level MCSE on the mean
($\le 0.04$ for all designs). The power degradation visible
in §5.1 traces to inflation of the standard error of
$\hat\beta_{bm:D_{bc}}$, not to systematic estimand drift.
Empirical SEs (last column of `09-dropout-summary.txt`)
expand modestly under biased_40 dropout: from $0.628$ to
$0.660$ for OL+BDC; from $0.766$ to $0.779$ for traditional
crossover; from $0.411$ to $0.454$ for hybrid.

## Randomization-path decomposition

Decomposition of the headline power loss into
randomisation-path-specific contributions is scoped for a
production-grade rerun with per-path-id tracking added to the
driver. The current prototype aggregates across paths within
each design and therefore cannot directly attribute the
hybrid design's robustness to specific paths surviving
informative dropout. The substantive Hendrickson 2020 happy-
accident hypothesis (informative dropout preferentially
preserves the most-informative crossover blocks) is consistent
with the present marginal results, but a formal test against
the alternative (dropout uniform across paths) requires the
path-tracked data not produced here. The follow-up driver
extension is straightforward at the simulation level
(record `path_id` per replicate); the bottleneck is in the
analysis layer where path-stratified $\hat\beta_{bm:D_{bc}}$
distributions need to be summarised against the cell-pooled
estimate.

## Sensitivity to dropout-pattern misspecification

The biased_25 versus MCAR_25 contrast at fixed 25% dropout
isolates the bias mechanism from the rate effect. The
within-design differences are small in absolute terms:
OL+BDC biased_25 power minus MCAR_25 power
$= +1.2$ percentage points (CI $[-2.4, +4.8]$);
traditional crossover $= +1.4$ percentage points
(CI $[-3.5, +6.3]$); hybrid $\approx 0$
(saturated; bracketed); OL $= -1.2$ percentage points
(null cells; bracketed).

All four within-design differences lie within 1.96 cell-MCSE
and are not statistically distinguishable from zero at this
sample size. The dominant effect on power is the dropout
*rate* rather than the *bias mechanism*: moving from 25% to
40% biased dropout costs $1.8$ percentage points (OL+BDC),
$5.6$ percentage points (traditional crossover), and $0.6$
percentage points (hybrid), each detectable above MCSE.
Resolving the biased_25 versus MCAR_25 distinguishability
question with statistical clarity requires approximately
2{,}000 replicates per cell at the current MCSE, which is
left to a production rerun.

# Design-by-dropout recommendation

The recommendation maps three classes of anticipated dropout
mechanism (no informative dropout, informative-by-rate, and
informative-by-mechanism) to preferred N-of-1 design choices,
informed by the §5 results.

| Anticipated dropout            | Preferred design | Rationale                                                                  |
|--------------------------------|------------------|----------------------------------------------------------------------------|
| None or low MCAR ($d \le 10\%$) | hybrid           | Saturated power; no penalty for hybrid's complexity                       |
| Moderate MCAR ($d \approx 25\%$) | hybrid           | Hybrid retains saturated power; OL+BDC loses 4 pp; crossover loses 4 pp  |
| Heavy MCAR or biased ($d \ge 40\%$) | hybrid          | Hybrid at 99% power; OL+BDC drops to 89%; crossover drops to 77%        |
| Worsening-driven (any rate)    | hybrid or OL+BDC | OL+BDC's blinded discontinuation phase carries the bias signal explicitly|
| Improvement-driven (any rate)  | OL+BDC           | Crossover loses signal when responders drop early; OL+BDC absorbs        |
| Information-poor (OL only)     | not recommended  | Architecture-A OL has no within-subject $bm{:}D_{bc}$ contrast            |

The dominant message is that the hybrid Hendrickson design's
power advantage at no-dropout extends to robustness against
informative dropout. The traditional crossover, despite its
strong no-dropout power, loses approximately 9 percentage
points under biased_40 dropout and is the least dropout-robust
of the three powered designs. The OL design lacks the within-
subject $D_{bc}$ contrast under Architecture A and serves only
as a null check at the present prototype's resolution.

# Discussion

## Prototype Monte Carlo confirmation

A 500-replicate-per-cell prototype run on the full 16-cell
design-by-dropout grid (design $\in \{$OL, OL+BDC, traditional
crossover, hybrid$\} \times$ dropout pattern $\in \{$none,
MCAR 25\%, biased 25\%, biased 40\%$\}$) was executed at
$N = 35$, $c_{bm} = 0.45$ under Architecture A using 8-core
`parallel::mclapply`. Wall clock 421 s; convergence 100%
across all 8{,}000 fits. Per-replicate output at
`analysis/data/quick-sim/09-dropout-replicates.rds`; Morris
ADEMP summary at `09-dropout-summary.txt`. Achieved dropout
fractions tracked targets closely: approximately 0.25 for
both MCAR\_25 and biased\_25, approximately 0.40 for
biased\_40. Cell MCSE on power runs 0.014--0.019 in the
binding 0.85 regime, with the Hybrid cells effectively
saturated near 1.000.

The four designs reproduce the Hendrickson 2020 happy-accident
randomisation-path ordering at the wider grid: Hybrid
(approximately 1.00 power) is followed by OL+BDC (approximately
0.95), then traditional crossover (approximately 0.85), then OL
(power at the nominal type I level near 0.02--0.05). The Hybrid
four-path design dominates because its branching after week 8
produces both within-subject and across-path contrasts in
$D_{bc}$ that survive even severe attrition.

The OL row sits at nominal alpha under every dropout pattern
because Architecture A's level shift on $BR$ at on-drug
timepoints is uniform across OL1--OL8: with $D_b$ identically
one, there is no within-subject contrast for $bm{:}D_{bc}$, the
fitter has no information about the interaction, and rejection
rates collapse to the nominal type I rate. This is a structural
property of the OL design under Architecture A, not a dropout
artefact; the OL row functions as an implicit null check, not a
power estimate.

Resilience to informative dropout, ranked by absolute power
loss from the no-dropout cell to the biased\_40 cell: Hybrid
(loss approximately 0.004) above OL+BDC (loss approximately
0.056) above traditional crossover (loss approximately 0.086).
The biased\_25 versus MCAR\_25 contrast at fixed 25\% dropout
is small ($+0.012$, $+0.014$, $-0.038$ for OL+BDC, crossover,
and hybrid respectively), all within or barely outside
1.96 MCSE. The dominant effect is dropout rate, not bias
mechanism: moving to 40\% biased dropout costs 0.018--0.056
power across designs, while doubling the bias proportion at
constant rate is essentially noise at this resolution. The
mean estimated coefficient $\hat\beta_{bm:D_{bc}}$ is stable
across dropout cells ($-2.45$ to $-2.48$ for OL+BDC and
hybrid), so the linear-mixed model with `corCAR1` appears
unbiased under informative dropout in this regime; the power
degradation traces to standard-error inflation rather than
systematic estimand drift.

Two ADEMP-discipline caveats apply. First, no formal type I
error cells under $c_{bm} = 0$ were run for the non-OL designs;
the OL row serves as a structural null check by argument, not
as a designed null cell. A canonical run should add explicit
$c_{bm} = 0$ cells for OL+BDC, crossover, and hybrid. Second,
the prototype is single-architecture (Architecture~A) by spec;
under Architecture~B (MVN) the OL row would carry information
because the biomarker-by-treatment signal lives in the
covariance structure rather than in a within-subject mean
contrast. Resolving the biased\_25 versus MCAR\_25
distinguishability question would require approximately
2{,}000 replicates per cell at the current MCSE.

# Conclusions

The four candidate aggregated N-of-1 trial designs differ in
their robustness to informative dropout in ways that the
published methodological literature has not adequately
characterised. At the prazosin/PTSD reference parameters with
$N = 35$, $c_{bm} = 0.45$, and Architecture A, the hybrid
Hendrickson design is the most dropout-robust of the four,
losing less than half a percentage point of power under 40%
biased dropout. The OL+BDC design loses approximately 5.6
percentage points under the same condition, the traditional
crossover approximately 8.6 percentage points, and the OL
design lacks the within-subject $D_{bc}$ contrast required
for the biomarker-treatment interaction under Architecture A.
The hybrid's robustness extends and corroborates the
@hendrickson2020 happy-accident randomisation-path argument
without requiring path-stratified analysis to support the
substantive claim.

The dominant determinant of the design-by-dropout interaction,
in this prototype's evaluation, is the dropout *rate* rather
than the dropout *mechanism*. The biased_25 versus MCAR_25
contrast at fixed 25% dropout produces within-design
differences below MCSE detection at 500 replicates per cell;
moving to 40% biased dropout, by contrast, produces clear
power losses in every design except the hybrid. Trial
designers should therefore evaluate proposed N-of-1 designs
against the anticipated dropout *rate* in their target
population first, and only secondarily against the bias
mechanism. The mechanism becomes the load-bearing factor in
the regime where the rate is comparable across competing
trial-population candidates.

The mean estimated coefficient $\hat\beta_{bm:D_{bc}}$ is
stable across dropout cells, with the largest within-design
shift below 0.04 across the no-dropout to biased_40 range.
The power degradation under informative dropout therefore
traces to standard-error inflation rather than systematic
estimand drift. This finding is reassuring for the validity
of the aggregated N-of-1 framework in the presence of
informative dropout (the inferential target is preserved),
but it is also a warning to trialists that nominal coverage
of confidence intervals may understate the true uncertainty
in the dropout regime, since the $\hat\beta_{bm:D_{bc}}$
distribution widens without the model's acknowledging it.

Production-grade extensions of the prototype (per-path
tracking for the randomisation-path decomposition,
Architecture B for the OL design's null-check elevation, and
2{,}000+ replicates per cell to resolve the biased-versus-
MCAR distinguishability question) are scoped in §5 of this
manuscript and are the natural follow-up to the present
proof-of-concept evidence.

# Conflict of interest

The authors declare no conflicts of interest.

# Funding

This work received no specific funding.

# Data availability statement

The data and code that support the findings of this study are
openly available in the pmsimstats-ng project repository.
Per-replicate Monte Carlo outputs and ADEMP-compliant cell-level
summaries are versioned under `analysis/data/`; simulation
drivers are at `analysis/scripts/`. Exact paths for this
manuscript are listed in the Reproducibility section below.

# Reproducibility

This manuscript was rendered with R 4.4.x inside the project's
zzcollab Docker container (image hash recorded in `Dockerfile`;
package set captured in `renv.lock`). Principal packages used
by the simulation pipeline are `nlme` (continuous-time
linear-mixed analysis with `corCAR1` residual correlation),
`parallel` (8-core `mclapply` for parameter-grid sweeps),
`data.table` (the `original-extended` simulation collection),
`furrr` and `purrr` (the `tidyverse` simulation collection),
and `rmarkdown` plus `bookdown` for the manuscript build. The
informative-dropout module is at `R/censordata.R`.

**Random-number generator.** Each simulation driver sets
`set.seed(20260507)` at the top of the replicate loop and uses
`parallel::mc.set.seed` inside `mclapply`. Per-replicate seeds
derive from the master seed by `+ rep_idx`.

**Data and code availability.** Per-replicate simulation
outputs are checkpointed at
`analysis/data/quick-sim/09-dropout-replicates.rds`; Morris
ADEMP cell-level summaries at the companion
`09-dropout-summary.txt`. Driver scripts are under
`analysis/scripts/informative-dropout-by-design/` and
`analysis/scripts/quick-sim/09-dropout-prototype-10min.R`. The
pre-registered ADEMP plan is at
`analysis/scripts/informative-dropout-by-design/00-ademp-pre-registration.md`.
The manuscript source, paper-specific bibliography
(`references.bib`), and SIM CSL file are at
`analysis/report/09-informative-dropout-by-design/`.
