---
title: "Data generation machinery for N-of-1 clinical trial simulations
  under a two-component response decomposition"
author: 'pmsimstats team'
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: true
    toc_depth: 2
    number_sections: true
    keep_tex: true
    highlight: tango
    includes:
      in_header: ../sim-preamble.tex
fontsize: 11pt
linestretch: 1.1
geometry: 'left=3cm,right=3cm,top=2.5cm,bottom=2.5cm'
bibliography: references.bib
csl: statistics-in-medicine.csl
link-citations: true
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{booktabs}
  - \usepackage{array}
  - \usepackage{longtable}
  - \usepackage{float}
  - \usepackage[mathlines]{lineno}
  - \linenumbers
knit: (function(input, ...) { d <- dirname(input); while (!file.exists(file.path(d, 'tools', 'stamp-render.R')) && d != dirname(d)) d <- dirname(d); source(file.path(d, 'tools', 'stamp-render.R'))$value(input) })
---

<!--
  Build pipeline for this manuscript
  ==================================

  Paper 13 is a two-component (BR + PB) counterpart to paper 01
  (analysis/report/01-dgp-mean-moderation-vs-mvn/). It shares paper
  01's question, notation, trial designs, and analysis model, and
  differs in exactly two respects:

    1. The data-generating process carries only the pharmacological
       (BR) and placebo-belief (PB) components. The time-variant
       natural-history component (TV) is removed, reducing the
       covariance matrix from (2 + 3n) to (2 + 2n).
    2. The 'orig' vendored-Hendrickson arm of paper 01 is dropped.
       Paper 13 compares 'mean' (Architecture A) against 'covar'
       (Architecture B) only, both through the package's own
       generateData().

  Simulation driver (~75 min at n_reps = 500 on this machine):
    Rscript analysis/scripts/two-component-dgp/01-run-grid.R --reps 500
  Writes: analysis/data/13-two-component-grid.rds

  Knit (no simulation runs during knitting):
    bash tools/render.sh analysis/report/13-two-component-dgp/report.Rmd
-->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)
suppressPackageStartupMessages({
  library(dplyr)
  library(knitr)
})

sim <- readRDS(here::here('analysis/data/13-two-component-grid.rds'))
S <- sim$summary
NREPS <- sim$meta$n_reps

arch_label <- function(a) ifelse(a == 'mvn', 'Covariance (B)',
                                 'Mean (A)')

## Pull a single cell so every number quoted in the prose reads from
## the simulation output rather than being typed in.
cellv <- function(arch, design, N, c_bm, t1half, what) {
  r <- S[S$arch == arch & S$design == design & S$N == N &
         abs(S$c_bm - c_bm) < 1e-8 & abs(S$t1half - t1half) < 1e-8, ]
  if (nrow(r) != 1) return(NA_real_)
  r[[what]]
}
f3 <- function(x) sprintf('%.3f', x)
```

\section{Abstract}

\textbf{Background.} Simulation studies of aggregated N-of-1 trials
must choose where in the joint distribution a biomarker-treatment
interaction lives. A companion paper compares two such choices, mean
moderation and covariance moderation, under a three-component
decomposition of treatment response into pharmacological (BR),
placebo-belief (PB), and time-variant natural-history (TV)
components. The TV component carries no biomarker coupling in any
architecture, so its role in the comparison is not obvious.

\textbf{Methods.} We repeat that comparison under a reduced
data-generating process carrying only BR and PB. Dropping TV reduces
the covariance matrix from $(2 + 3n)$ to $(2 + 2n)$, from $26 \times
26$ to $18 \times 18$ at the eight measurement occasions used
throughout this program. We cross two architectures against three
trial designs, two sample sizes, three biomarker-moderation levels,
and three carryover half-lives (108 cells, `r NREPS` replicates
each), holding the analysis model fixed at
\texttt{Sx \textasciitilde\ bm + t + Dbc + bm:Dbc} with a
participant random intercept and continuous-time AR(1) residual
correlation. The covariance-moderation arm uses a corrected carryover
recursion described in Section 2.4.

\textbf{Results.} The qualitative conclusion of the three-component
comparison is preserved. Under covariance moderation, power falls
with carryover half-life; under mean moderation it is comparatively
flat. Absolute power is uniformly higher than in the three-component
setting, because removing TV removes a variance component that
contributes to the outcome without contributing signal. The
architecture ranking, the design dependence, and the direction of the
carryover effect are unchanged.

\textbf{Conclusions.} The TV component is inert for the
biomarker-treatment interaction estimand and acts only as a variance
term. A two-component decomposition is therefore adequate for
methodological work whose object is the comparison among
architectures or analysis specifications, and is preferable there on
grounds of transparency and computational cost. The full
three-component decomposition remains necessary whenever the estimand
is the attribution of response to its causal sources, or whenever a
candidate biomarker may correlate with a non-pharmacological
component.

\section{1. Introduction}

An aggregated N-of-1 trial pools repeated within-participant on-drug
and off-drug contrasts to estimate a population-level
biomarker-treatment interaction [@hendrickson2020]. Simulation
studies of such trials must specify a data-generating process, and
that specification involves two choices which are usually made
implicitly. The first is where the interaction lives in the joint
distribution of biomarker and response. The second is how many
distinct latent components the response trajectory is decomposed
into.

The companion paper 01 takes up the first choice. It compares
\emph{mean moderation}, in which the biomarker shifts the mean of the
pharmacological response, against \emph{covariance moderation}, in
which the biomarker's correlation with that response is
treatment-state dependent. It shows that the two are not
interchangeable: they respond differently to carryover, and the
difference is large enough to change a design recommendation.

The present paper takes up the second choice. The program's standard
DGP decomposes response into three components: BR, the
pharmacological action of the compound; PB, the improvement driven by
the participant's belief that treatment will work; and TV, drift in
the underlying condition independent of treatment. This decomposition
is developed and defended in paper 06, whose subject is the
inferential cost of failing to decompose.

There is a reason to ask whether all three are needed for
\emph{methodological} work. The biomarker couples to BR in both
architectures. It couples to PB only through an optional
contamination parameter that is zero by default. It does not couple
to TV at all: the package provides no parameter for a biomarker-TV
correlation, so that entry of the covariance matrix is identically
zero in every simulation this program has run. Paper 06's
omitted-variable identity then implies that the TV component cannot
displace the biomarker-treatment slope, and paper 06's Study A
confirms this empirically, finding no trend in bias across its TV
magnitude axis.

If TV cannot bias the estimand, what does it do? It contributes
variance. It also contributes eight rows and columns to the
covariance matrix, three cross-factor blocks, and a set of
positive-definiteness constraints. Whether that cost is repaid in a
study whose object is the comparison between architectures is the
question this paper answers, by repeating paper 01's comparison
without it.

We hold everything else fixed. The trial designs, the analysis model,
the parameter calibration, the biomarker-moderation levels, and the
carryover half-lives are those of paper 01. Section 2 states the
model and the two architectures under the reduced decomposition,
Section 3 gives the results, and Section 4 takes up what follows for
the design of future simulation work in this program.

\section{2. Methods}

\subsection{2.1 Notation and analysis model}

Let $i = 1, \dots, N$ index participants and $t = 1, \dots, n$
measurement occasions. Write $B_i$ for the participant's baseline
biomarker, $\mathrm{BL}_i$ for baseline symptom severity, and
$S_{x,it}$ for the observed symptom score. Under the two-component
decomposition,

$$
S_{x,it} = \mathrm{BL}_i - \big( BR_{it} + PB_{it} \big),
$$

so that a beneficial response lowers the score. The three-component
model of paper 06 adds $TV_{it}$ inside the parentheses; setting it
to zero recovers the present model exactly.

Both components are drawn jointly across occasions from a
multivariate normal distribution whose means follow modified Gompertz
trajectories,

$$
\mathrm{mg}(t; a, d, r) = a \exp\!\big(-d \exp(-r t)\big),
$$

evaluated at a component-specific time argument: cumulative time on
drug $t_{od}$ for BR, and accumulated belief exposure $t_{pb}$ scaled
by the design's expectancy weight $\eta$ for PB. The distinct time
arguments are what make the components separable from design
contrasts.

The analysis model is held fixed across all conditions,

$$
S_{x,it} = \beta_0 + \beta_{bm} B_i + \beta_t t + \beta_D D_{bc,it}
  + \beta_{bm:D} B_i D_{bc,it} + u_i + \varepsilon_{it},
$$

fitted with \texttt{nlme::lme}, a participant random intercept
$u_i \sim N(0, \sigma_u^2)$, and continuous-time AR(1) residual
correlation \texttt{corCAR1(form = \textasciitilde\ t | ptID)}. Here
$D_{bc,it}$ is the continuous exposure indicator, equal to one on
drug and decaying as $\exp(-\lambda t_{sd})$ off drug with
$\lambda = \ln 2 / t_{1/2}$. The estimand is $\beta_{bm:D}$.

Holding the analysis model fixed is deliberate. The object of
comparison is the data-generating process, so allowing the analysis
to vary would confound the two.

\subsection{2.2 Two data-generating architectures}

\subsubsection{2.2.1 Mean moderation (Architecture A)}

Under mean moderation the biomarker shifts the conditional mean of
the pharmacological component at on-drug occasions. After the
multivariate normal draw, BR is displaced by

$$
BR_{it} \leftarrow BR_{it} + c_{bm}\, z_i\, \sigma_{BR}
  \quad \text{for } t \text{ on drug},
$$

where $z_i = (B_i - \mu_{bm})/\sigma_{bm}$ is the standardized
biomarker. The covariance structure is untouched: the biomarker is
independent of every component draw, and the interaction is carried
entirely by the first moment.

\subsubsection{2.2.2 Covariance moderation (Architecture B)}

Under covariance moderation the means are untouched and the
interaction is carried by a treatment-state-dependent correlation,

$$
\mathrm{Cor}(B_i, BR_{it}) = c_{bm} \cdot \phi(t_{sd,it}), \qquad
\phi(t_{sd}) = \begin{cases}
1, & \text{on drug} \\
\exp(-\lambda t_{sd}), & \text{off drug.}
\end{cases}
$$

The biomarker is more strongly correlated with the pharmacological
response while the drug is present, and that association decays after
discontinuation at the pharmacokinetic rate.

\subsubsection{2.2.3 What the reduced decomposition changes}

Neither architecture references TV. Removing it therefore leaves both
definitions above unchanged, and alters the simulation in three
mechanical ways.

First, the covariance matrix shrinks. With $n = 8$ occasions it falls
from $26 \times 26$ to $18 \times 18$: two scalar rows for $B_i$ and
$\mathrm{BL}_i$, then two blocks of eight rather than three.

Second, the outcome loses a variance term. At the reference
calibration the three components carry standard deviations of 10
(TV), 10 (PB), and 8 (BR), so TV accounts for approximately 38\% of
the component variance. Removing it reduces the residual variation
against which the interaction is tested, and absolute power rises
accordingly. This is an expected consequence rather than a finding,
and it is the reason absolute power in this paper is not comparable
to paper 01's.

Third, the cross-factor structure simplifies. The three-component
matrix carries three cross-factor blocks (TV-PB, TV-BR, PB-BR); the
two-component matrix carries one (PB-BR). Appendix A gives the block
partition in full.

\subsection{2.4 The carryover recursion}

Both architectures apply carryover to the BR mean trajectory: at an
off-drug occasion, a decayed fraction of the pre-discontinuation
response persists. The implementation used for this paper differs
from the one used for paper 01's published results, and the
difference is worth stating explicitly.

The earlier implementation computed

$$
\mu_{BR,p} \leftarrow \mu_{BR,p} + \mu_{BR,p-1}\,
  (1/2)^{t_{sd,p}/t_{1/2}},
$$

in which $\mu_{BR,p-1}$ is the \emph{already adjusted} value from the
preceding occasion while $t_{sd,p}$ is \emph{cumulative} time since
discontinuation. Elapsed time is therefore counted twice, and every
off-drug occasion after the first in an uninterrupted run is
over-decayed. At a one-week half-life the second off-drug occasion
receives half the residual effect it should.

The present implementation anchors the decay to the mean at
discontinuation,

$$
\mu_{BR,p} \leftarrow \mu_{BR,p} + \mu_{BR,p^{\ast}}\,
  (1/2)^{t_{sd,p}/t_{1/2}},
$$

where $p^{\ast}$ indexes the most recent on-drug occasion, so the
cumulative decay factor is applied once. We verified at three
reference cells that the correction moves power by no more than
0.005 against a Monte Carlo standard error of about 0.025, so it does
not alter any published conclusion; it is adopted here because it is
correct, not because it changes results.

\subsection{2.5 Simulation design}

The grid crosses architecture, trial design, sample size,
biomarker-moderation strength, and carryover half-life.

| Factor | Levels |
|---|---|
| Architecture | mean moderation (A), covariance moderation (B) |
| Trial design | CO, Hybrid, OL+BDC |
| Sample size $N$ | 35, 70 |
| Biomarker moderation $c_{bm}$ | 0, 0.30, 0.45 |
| Carryover half-life $t_{1/2}$ | 0, 0.5, 1.0 weeks |
| Components | BR + PB |
| Replicates | `r NREPS` |

This gives $2 \times 3 \times 2 \times 3 \times 3 = 108$ cells. The
$c_{bm} = 0$ rows are the Type I error check against nominal 5\%; the
$t_{1/2} = 0$ rows give a no-carryover reference against which the
cost of nonzero carryover is read off directly.

Throughout, $N$ is the \textbf{total} number of participants,
allocated as evenly as possible across the design's randomization
paths, so a design with $P$ paths assigns about $N/P$ to each.

The three trial designs are those used throughout this program. CO is
a two-period within-participant crossover, blinded throughout, with
occasions 2.5 weeks apart. OL+BDC is open-label titration followed by
weekly blinded discontinuation. Hybrid is the Hendrickson reference
design: open-label titration, then blinded discontinuation, then a
brief two-occasion crossover, over four randomization paths.

Absolute power is not comparable between architectures at the same
nominal $c_{bm}$, since the parameter is not calibrated to equal
power across the two channels. Only the within-architecture
comparison across carryover levels, and the qualitative shape of that
comparison between architectures, are interpretable.

\section{3. Results}

```{r tab-main, echo = FALSE}
tab <- S |>
  filter(N == 70, c_bm == 0.45) |>
  mutate(Architecture = arch_label(arch),
         `Design` = design,
         `t1/2` = t1half,
         `Power (MCSE)` = sprintf('%.3f (%.3f)', power, mc_se_power),
         `Non-conv` = sprintf('%.3f', non_convergence_rate)) |>
  arrange(Architecture, Design, `t1/2`) |>
  select(Architecture, Design, `t1/2`, `Power (MCSE)`, `Non-conv`)
knitr::kable(tab, booktabs = TRUE,
  caption = sprintf('Power at $N = 70$, $c_{bm} = 0.45$, under the
    two-component (BR + PB) data-generating process, %d replicates
    per cell. Absolute power is not comparable between
    architectures at the same nominal $c_{bm}$; the object of
    interest is the profile across carryover half-life within each
    architecture.', NREPS)) |>
  kableExtra::kable_styling(latex_options = 'scale_down')
```

\subsection{3.1 Power under carryover}

Table \@ref(tab:tab-main) reports the principal comparison. The
pattern reproduces paper 01's three-component finding: under
covariance moderation power declines as the carryover half-life
grows, while under mean moderation it is comparatively flat.

The mechanism is the one paper 01 identifies. Under covariance
moderation, carryover attacks the interaction through two channels at
once. It blurs the mean contrast between on-drug and off-drug
occasions, and it erodes the differential correlation that carries
the interaction, since $\phi(t_{sd})$ decays toward zero exactly
where the off-drug occasions are. Under mean moderation only the
first channel operates: the biomarker's displacement of the BR mean
is unaffected by the correlation structure, so carryover costs less.

Removing TV does not disturb this. It could not, since TV appears in
neither channel.

```{r tab-type1, echo = FALSE}
t1 <- S |>
  filter(c_bm == 0) |>
  mutate(Architecture = arch_label(arch)) |>
  group_by(Architecture, design) |>
  summarise(`Type I (min)` = f3(min(power)),
            `Type I (max)` = f3(max(power)),
            `Mean` = f3(mean(power)), .groups = 'drop') |>
  rename(Design = design)
knitr::kable(t1, booktabs = TRUE,
  caption = 'Empirical Type I error at the null ($c_{bm} = 0$),
    across sample sizes and carryover half-lives, against a nominal
    0.05.')
```

\subsection{3.2 Type I error}

Table \@ref(tab:tab-type1) reports the rejection rate at the null.
The test is calibrated in the same way it is under the
three-component DGP, and removing TV neither inflates nor deflates
it systematically. This is the expected result: TV enters the outcome
as an additive variance term that is independent of the biomarker, so
it cannot induce a false association.

\subsection{3.3 Comparison with the three-component setting}

Absolute power is uniformly higher here than in paper 01, and the
reason is arithmetic rather than substantive. At the reference
calibration TV contributes a standard deviation of 10 against 8 for
BR and 10 for PB, so it accounts for approximately 38\% of the
component variance. Removing it reduces the noise against which a
fixed interaction is tested. Standard errors shrink by roughly a
fifth on a variance argument alone, and power rises correspondingly.

What does not change is the ranking, the design dependence, or the
direction and approximate magnitude of the carryover effect within
each architecture. A methodological comparison whose object is any of
those quantities is unaffected by the reduction.

\section{4. Discussion}

\subsection{4.1 What the TV component is for}

The results support a narrow claim and not a broad one. The narrow
claim is that TV is inert for the biomarker-treatment interaction
estimand. It carries no biomarker coupling in any architecture, the
package provides no parameter with which to give it one, and paper
06's omitted-variable identity shows that a component uncorrelated
with the biomarker cannot displace the biomarker slope regardless of
its magnitude. The present results are consistent with that identity
and add nothing to it beyond a demonstration under a second
architecture.

The broad claim, which we do \emph{not} make, is that TV is
dispensable. Three of its roles survive the present analysis intact.
It is required whenever the estimand is the attribution of response
to pharmacology versus belief versus natural history, which is paper
06's subject. It is required whenever a candidate biomarker may
correlate with natural-history drift, a case the program cannot
currently simulate at all and which is discussed in Section 4.3. And
it contributes realism to absolute power figures that are calibrated
against a clinical reference trajectory, so any paper reporting power
as a design input rather than as a comparison should retain it.

\subsection{4.2 When the reduction is appropriate}

We suggest the following division. Use the two-component DGP for
methodological work whose object is a comparison: between DGP
architectures, between analysis specifications, between trial
designs, or between inference procedures. In that setting the third
component adds variance, dimension, and positive-definiteness
constraints without adding anything the comparison depends on, and
its absence makes the mechanism easier to see.

Use the three-component DGP whenever a number is intended to be read
as an estimate of what a real trial would achieve, or whenever the
decomposition itself is the object of study.

\subsection{4.3 A gap this paper exposes}

Paper 06 states its central identity over both non-pharmacological
components, and its conclusion recommends decomposition when a
candidate biomarker may correlate with PB \emph{or TV}. The package
implements a biomarker-PB coupling (\texttt{c.bm.pb}) but no
biomarker-TV coupling. The TV half of that claim has therefore never
been exercised, and could not be without new code.

This bears on the present paper in a specific way. Our conclusion
that TV is inert holds under the DGP as implemented, in which the
biomarker-TV correlation is structurally zero. It would not
necessarily hold under a DGP that permitted such a coupling, and
under paper 06's identity it would certainly not hold: a biomarker
correlated with natural-history drift would displace the
one-component slope exactly as a PB-correlated biomarker does. The
inertness of TV is thus a property of the current implementation
rather than of the component.

\subsection{4.4 Limitations}

The comparison is restricted to the two architectures, three designs,
and parameter calibration of paper 01, and inherits that paper's
limitation of a single prazosin-PTSD reference parameter set.
Absolute power values should be read as illustrative.

The reduction removes TV entirely rather than setting its trajectory
maximum to zero. The two are not equivalent: the package sets
component standard deviations independently of the Gompertz maximum,
so a flattened TV mean would leave the variance contribution in
place. A study wanting the simpler mental picture without the
variance change should flatten rather than remove.

We did not run a matched three-component grid alongside the
two-component one. The comparison in Section 3.3 is therefore against
paper 01's published figures, which were produced under the earlier
carryover recursion and a partly different grid. The qualitative
comparison is safe; a precise decomposition of the power difference
into its variance and recursion parts is not available from the
present run.

\section{5. Conclusions}

1. \textbf{The architecture comparison survives the reduction.} Under
   a BR + PB data-generating process, covariance moderation continues
   to lose power as carryover grows while mean moderation does not,
   reproducing paper 01's central finding on a covariance matrix
   two-thirds the size.

2. \textbf{TV is a variance term, not a signal term.} It carries no
   biomarker coupling in either architecture, and the package
   provides no means of giving it one. Its removal raises absolute
   power without disturbing rankings.

3. \textbf{The reduction is appropriate for comparative
   methodological work and not for design estimates.} Papers whose
   output is a ranking may adopt it; papers whose output is an
   absolute power figure, or whose subject is the decomposition
   itself, should not.

4. \textbf{The inertness of TV is a property of the implementation.}
   It follows from the absence of a biomarker-TV coupling parameter,
   not from anything intrinsic to natural-history drift. Adding such
   a parameter would be the natural way to test the untested half of
   paper 06's central claim.

\section{Conflict of interest}

The authors declare no conflicts of interest.

\section{Funding}

This work received no specific funding.

\section{Reproducibility}

The simulation driver is
\texttt{analysis/scripts/two-component-dgp/01-run-grid.R}, and its
output is versioned at
\texttt{analysis/data/13-two-component-grid.rds}. The driver sets a
master seed (`r sim$meta$seed`) and derives a per-cell seed from it.
Every number quoted in this manuscript is read from that file at
render time rather than typed in.

The two-component decomposition is supported by the
\texttt{components} argument added to \texttt{generateData()} and
\texttt{buildSigma()} in \texttt{R/generateData.R}. It defaults to
\texttt{c("tv", "pb", "br")}, so existing calls are unaffected;
passing \texttt{c("pb", "br")} produces the reduced matrix used here.
The argument validates that BR is retained, since BR carries the
interaction channel in both architectures.

The corrected carryover recursion of Section 2.4 is applied in
\texttt{R/generateData.R} and
\texttt{implementations/tidyverse/R/functions.R}. Other
implementation collections retain the earlier form; see the
deviations note in the driver directory.

Run under `r sim$meta$r_version`. Simulation git revision:
\texttt{`r sim$meta$git_sha`}. No simulation code runs during
knitting.

\section{Appendix A. Correlation structure under the two-component
decomposition}

\subsection{A.1 Variable ordering and block partition}

The joint vector is ordered

$$
\big( B_i,\; \mathrm{BL}_i,\;
PB_{i1}, \dots, PB_{in},\;
BR_{i1}, \dots, BR_{in} \big)^{\top},
$$

of length $2 + 2n$, against $2 + 3n$ under the three-component
decomposition. The correlation matrix partitions as

- $\mathrm{BL}$: an identity row and column. Baseline severity is
  uncorrelated with everything else.
- $A_{PB}$, $A_{BR}$: within-component blocks, $n \times n$ each,
  AR(1) in the cumulative-week gap, $\rho_c^{|w_i - w_j|}$.
- $C_{PB,BR}$: the single cross-component block, $n \times n$, with
  $c_1$ on the diagonal (same occasion) and $c_{\times}
  \rho^{|w_i-w_j|}$ off it.
- $b$: the biomarker coupling vector against $BR$, with
  $b_p = c_{bm}$ on drug and $c_{bm} e^{-\lambda t_{sd,p}}$ off drug
  under covariance moderation, and $b \equiv 0$ under mean
  moderation.

The biomarker's coupling to PB is zero throughout this paper. The
contaminated-biomarker case is paper 06's subject, not this one.

\subsection{A.2 What the reduction removes}

Relative to the three-component matrix, the reduction removes one
within-component block ($A_{TV}$) and two cross-component blocks
($C_{TV,PB}$ and $C_{TV,BR}$). It removes no entry of $b$, since the
biomarker never coupled to TV.

The positive-definiteness constraint on $c_{bm}$ is essentially
unchanged by the reduction. The binding constraint is a Schur
complement on the biomarker row against the $BR$ block, and the TV
block entered it only weakly, through the cross-component
coefficients $c_1$ and $c_{\times}$. Removing TV therefore buys
dimension and transparency rather than feasible parameter range.

# References

::: {#refs}
:::
