---
title: "Sensitivity of carryover-mitigation analysis strategies to
  biomarker data-generating process: a 3 x 3 simulation study"
author:
  - pmsimstats team
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    keep_tex: true
    toc: false
    number_sections: true
bibliography: references.bib
csl: statistics-in-medicine.csl
link-citations: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.path = "../figures/02-",
  cache.path = "_cache/"
)
library(tibble)
library(dplyr)
library(knitr)
```

# Abstract

Carryover effects in aggregated N-of-1 and crossover trials degrade
the detectability of biomarker-treatment interactions. Analysis
strategies that acknowledge carryover---binary-treatment,
exposure-weighted, and lagged-treatment models---differ in how
they parameterise residual drug effect, and each makes implicit
assumptions about the functional form of the decay. We conduct a
simulation study crossing three data-generating carryover decay
forms (linear, exponential, Weibull) against three analysis-model
carryover specifications, under both mean-moderation (Architecture
A) and multivariate-normal differential-correlation (Architecture
B) biomarker data-generating processes. Results quantify the power
penalty of analysis-model mis-specification and identify which
mitigation strategy is most robust across the decay-form grid.

# Introduction

Biomarker-moderated treatment effects in within-subject designs
depend on how residual drug effect is modelled in both the
data-generating process (DGP) and the analysis model. The DGP
architecture comparison in the companion manuscript and in
`docs/02-dgp-mean-moderation-vs-mvn.tex` establishes that
carryover has qualitatively different power consequences under
the two architectures [@hendrickson2020]. That prior work held
the decay form fixed at exponential and evaluated a single
analysis model.

This paper addresses a separate question: given that the trial
designer will specify a carryover decay form (explicitly or
implicitly) and the analyst will choose an analysis-model
carryover specification, how sensitive is the realised power to
mis-specification at each layer? The practical motivation is
that the true pharmacokinetic decay of residual drug effect is
rarely known with precision at the design stage, and the
analyst typically defaults to a convenient parametric form.

We perform a factorial simulation study crossing three DGP decay
forms (linear, exponential, Weibull) with three analysis-model
carryover specifications (binary-treatment, exposure-weighted,
lagged-treatment). We run the full grid under both DGP
architectures to establish whether robustness rankings transfer
across architectures.

# Simulation design

## Data-generating processes

Both DGP architectures from the companion manuscript are
retained:

- **Architecture A (direct mean moderation).** Biomarker $B_i$
  enters the mean structure:
  $Y_{it} = \mu_0 + \beta_1 D_{it}^{*}(1 + \beta_{bm} B_i)
  + \epsilon_{it}$, where $D_{it}^{*}$ is the effective
  exposure at timepoint $t$ under the chosen decay form.
- **Architecture B (MVN differential correlation).** Biomarker
  and biological response are jointly drawn from a multivariate
  normal with treatment-state-dependent correlation
  $\text{Cor}(B_i, BR_{it}) = c_{bm} \cdot \phi(t_{sd})$, where
  $\phi$ is the decay form.

In both architectures the decay function $\phi : \mathbb{R}_+
\to [0, 1]$ maps time since drug discontinuation $t_{sd}$ to a
residual-effect multiplier. We evaluate three forms
parameterised to match a common half-life $t_{1/2}$:

### Linear decay

$$
\phi_{\text{lin}}(t_{sd}) =
  \max\left(0,\, 1 - \frac{t_{sd}}{2 t_{1/2}}\right)
$$

Residual effect vanishes completely at $t_{sd} = 2 t_{1/2}$.
This form is biologically implausible for drugs with
first-order elimination kinetics but represents a defensible
pragmatic assumption when only the "carryover has fully worn
off by time X" information is available.

### Exponential decay

$$
\phi_{\text{exp}}(t_{sd}) =
  \exp\left(-\lambda t_{sd}\right), \qquad
  \lambda = \frac{\ln 2}{t_{1/2}}
$$

The form used throughout the companion manuscript and currently
implemented in `implementations/tidyverse/R/functions.R`.
Corresponds to first-order elimination kinetics and is the
default assumption in classical pharmacokinetics.

### Weibull decay

$$
\phi_{\text{weib}}(t_{sd}) =
  \exp\left(-\left(\lambda_w t_{sd}\right)^{k}\right), \qquad
  \lambda_w = \frac{(\ln 2)^{1/k}}{t_{1/2}}
$$

A shape parameter $k$ controls whether the tail is heavier
($k < 1$, slow elimination) or lighter ($k > 1$, accelerated
washout) than exponential. Exponential is recovered at $k = 1$.
We report results at $k = 0.7$ (slow elimination) and $k = 1.5$
(accelerated washout); the latter captures drugs with
saturable clearance or active-transporter kinetics.

**Implementation note.** The three decay forms are implemented
in `implementations/tidyverse/R/functions.R` via the
`carryover_decay()` helper and a `carryover_form` / `weibull_shape`
pair of `model_param` fields plumbed through
`apply_carryover_to_component()`, `build_correlation_matrix()`,
and `build_sigma_matrix()`. Backward compatibility is preserved
by defaulting to the exponential form. The
`implementations/original-extended/` collection has not been
updated and supports only the exponential form; cross-validation
across implementations is therefore limited to exponential cells.

## Analysis-model carryover specifications

The analysis model is `nlme::lme(Sx ~ bm * X + t, random = ~1
| ptID, correlation = corCAR1(form = ~ t | ptID))` for all
specifications. The three specifications differ in the drug
exposure predictor $X$:

### A1: binary-treatment (carryover ignored)

$X = D_{it}$, the on-drug indicator. This is the naive
specification that disregards any residual drug effect during
washout. It serves as a reference for the damage done by
ignoring carryover.

### A2: exposure-weighted (matched-decay continuous predictor)

$X = \text{Dbc}_{it}$, the continuous exposure-decayed
predictor matching the analyst's assumed decay form and
half-life. This is the current pmsimstats-ng default
specification. When the analyst's assumed form matches the
true DGP form the specification is well-matched; when it does
not, the predictor is mis-specified.

### A3: lagged-treatment (estimated carryover term)

$X = D_{it}$ augmented with a lagged-treatment indicator
$L_{it} = D_{i,t-1}(1 - D_{it})$ that marks off-drug
timepoints following an on-drug timepoint. The carryover
magnitude is estimated rather than assumed and does not
require the analyst to posit a decay form. Follows the
classical crossover-trial approach surveyed in
[@jones2014crossover].

## Factorial grid

The principal comparison is a $3 \times 3$ factorial of DGP
decay form against analysis-model specification, replicated
under each DGP architecture:

```{r grid-table}
grid <- tibble::tribble(
  ~`DGP decay`,   ~`A1 binary`, ~`A2 exposure-weighted`, ~`A3 lagged-treatment`,
  "Linear",       "cell 1",     "cell 2",                "cell 3",
  "Exponential",  "cell 4",     "cell 5 (matched)",      "cell 6",
  "Weibull",      "cell 7",     "cell 8",                "cell 9"
)
knitr::kable(
  grid,
  caption = "Principal $3 \\times 3$ simulation grid. Cell 5 is the
    matched reference: exponential DGP analysed under the current
    pmsimstats-ng default. Off-diagonal cells quantify the cost of
    DGP-analysis mis-specification. The full grid is crossed with
    Architecture $\\in \\{A, B\\}$."
)
```

## Parameter settings

| Parameter | Values |
|---|---|
| Biomarker moderation ($c_{bm}$ / $\beta_{bm}$) | 0.0, 0.30, 0.45 |
| Carryover half-life ($t_{1/2}$) | 0.0, 0.5, 1.0 weeks |
| Weibull shape ($k$) | 0.7, 1.5 |
| Sample size ($N$) | 35, 70 |
| Trial design | CO, N-of-1 (Hybrid), OL+BDC |
| DGP architecture | A, B |
| Replicates per cell | 500 (target); 50 for development |

The $c_{bm} = 0$ condition serves as the type I error check
(target: nominal 5%). The $t_{1/2} = 0$ condition removes
carryover entirely and provides a reference for the best-case
power at each $(N, \text{design}, c_{bm})$ combination.

## Outcome metrics

For each cell we report:

1. **Power** to reject $H_0: \beta_{\text{bm:X}} = 0$ at
   $\alpha = 0.05$.
2. **Type I error** at $c_{bm} = 0$.
3. **Bias** of the interaction estimate: $\hat{\beta} -
   \beta_{\text{true}}$ on the effect-size scale calibrated
   across architectures (see
   `docs/19-mean-moderation-implementation-notes.tex` for the
   calibration derivation).
4. **Monte Carlo SE** on all reported estimates.

## Computing environment

Simulations are orchestrated from
`analysis/02-carryover-sensitivity/` and use
`implementations/tidyverse/` as the primary implementation
(only collection with linear and Weibull decay support). Shared
helpers in `analysis/02-carryover-sensitivity/simulation-core.R`
define design presets (OL, CO, Hybrid, OLBDC), multi-path data
generation, long-form preparation with `Db` / `Dbc` / `L`
columns, and the three analysis-spec fitters.
Reproducibility metadata (R version, `implementations/tidyverse/`
commit SHA, simulation seed, elapsed time) are recorded in each
output `.rds` file.

Parallel execution uses `furrr::future_map` with the `multicore`
backend. Sigma matrices are cached per unique
(design, $t_{1/2}$, decay form) tuple before the replicate
loop begins.

# Results

All results reported below derive from the development-scale
factorial run (50 replicates per cell, 540 cells, computation
time ~9 min on eight cores). Monte Carlo standard errors on
individual power estimates are approximately
$\sqrt{p(1-p)/50} \approx 0.07$ at $p = 0.5$; within-figure
differences smaller than this should be read with caution.
A production-scale run (500 replicates per cell) is scheduled
before manuscript submission.

## Matched-decay baseline

Under the matched cell (exponential DGP, A2 exposure-weighted
analysis, Architecture~B, $N = 70$, $c_{bm} = 0.45$, Hybrid
design), power declines from 0.96 at $t_{1/2} = 0$ to 0.94 at
$t_{1/2} = 0.5$ to 0.82 at $t_{1/2} = 1.0$ weeks. These values
are within Monte Carlo error of the Architecture~B numbers
reported in the companion manuscript for the same parameter
settings (0.82 at $t_{1/2} = 1.0$), confirming that the
simulation pipeline reproduces the baseline result and
validating the calibration.

## Power by analysis specification

```{r fig-power-by-spec, echo = FALSE, fig.cap = "Power versus carryover half-life under the exponential DGP, stratified by analysis specification. The matched specification A2 tracks closely with the nominally inferior A1 and A3 at low half-lives but retains a modest advantage at $t_{1/2} = 1.0$ weeks, particularly under Architecture B and in the Hybrid and OL+BDC designs.", out.width = "90%"}
knitr::include_graphics("../figures/02-power-by-spec.pdf")
```

The three analysis specifications yield near-identical power at
$t_{1/2} = 0$ (no carryover), diverging as carryover increases.
The matched specification A2 retains a modest advantage under
Architecture~B and Hybrid / OL+BDC designs; A1 and A3 are
effectively indistinguishable across the exponential-DGP grid.

## Decay-form mis-specification

```{r fig-heatmap-mismatch, echo = FALSE, fig.cap = "Power at the critical cell ($t_{1/2} = 1.0$ weeks, Architecture B, Hybrid design, $N = 70$, $c_{bm} = 0.45$) across DGP decay forms (rows) and analysis specifications (columns). Under the Weibull heavy-tailed DGP ($k = 0.7$) and the linear DGP, the A2 matched-decay specification loses its advantage because its assumed exponential decay no longer matches truth.", out.width = "70%"}
knitr::include_graphics("../figures/02-heatmap-matched-vs-mismatched.pdf")
```

The heatmap cell at the intersection of exponential DGP and
A2 analysis is the matched reference (power 0.82). Cells in
the same row but other columns quantify the power penalty of
analysing under the wrong spec when the DGP is exponential.
Cells in the A2 column but other rows quantify the cost of
assuming exponential in the analysis when the DGP is actually
linear or Weibull.

## Type I error control

```{r fig-type1, echo = FALSE, fig.cap = "Empirical type-I error rates under the null ($c_{bm} = 0$), pooled across DGP decay forms and carryover half-lives, stratified by design and analysis specification. Boxplots cover the distribution across cells at $N = 70$; the dashed reference line marks the nominal 0.05 level.", out.width = "95%"}
knitr::include_graphics("../figures/02-type1-boxplots.pdf")
```

Mean empirical type-I error across all null cells is 0.040,
below the nominal 0.05 level, with maximum 0.14 (within the
Monte Carlo 95% interval $[0.01, 0.17]$ at 50 replicates). No
systematic inflation is apparent for any analysis specification
or architecture, consistent with well-behaved reference
distributions for the $bm \times \text{treatment}$ interaction
test under all three specifications.

# Discussion

*(Pending. Principal findings:*

- *Which analysis specification is most robust across DGP decay
  forms?*
- *Is the architecture-dependent ranking of Section 6 of the
  companion manuscript (which was derived under a single decay
  form) stable when decay form varies?*
- *Recommended default for a trialist who will specify the
  analysis model prospectively without knowing the true decay
  form.*
- *Relation to the latent-subtype tension raised by reviewer RH
  in the companion manuscript: if the true DGP is closer to an
  explicit mixture model than to any of the three parametric
  decay forms evaluated here, the lagged-treatment
  specification's non-parametric character may be doubly
  attractive.*
- *Limitations: we do not evaluate trial-design-level
  mitigations (longer washout, design modification; see
  `docs/02` Section 6.6) because these are orthogonal to the
  analysis-model question addressed here.)*

# Conclusions

*(Pending. Reporting recommendations for trialists planning a
biomarker-moderated N-of-1 study when carryover is anticipated:
choice of decay-form assumption, choice of analysis-model
specification, and sensitivity analyses to report alongside the
primary power calculation.)*

# References
