---
title: "Sensitivity of carryover-mitigation analysis strategies to
  biomarker data-generating process: a 3 x 3 simulation study"
author:
  - pmsimstats team
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    keep_tex: true
    toc: false
    number_sections: true
bibliography: references.bib
csl: statistics-in-medicine.csl
link-citations: true
---

<!--
  Build pipeline for this manuscript
  ==================================

  This Rmd does NOT run any simulation code. It loads pre-computed
  summaries and figures produced by the scripts below. Re-run the
  relevant steps when the DGP, analysis specs, or parameter grid
  change. All paths are relative to the repo root.

  1. Tier 1 principal factorial (heaviest step, ~85 min on 8 cores):
       Rscript analysis/02-carryover-sensitivity/01-run-factorial.R
     Writes: analysis/02-carryover-sensitivity/output/01-factorial.rds

  2. Tier 2 sensitivity blocks S1..S5 (~4 min on 8 cores; requires
     Tier 1 not to be running concurrently - both request all cores):
       Rscript analysis/02-carryover-sensitivity/04-run-sensitivity-blocks.R
     Writes: analysis/02-carryover-sensitivity/output/04-sensitivity.rds

  3. Summarise both tiers into cell-level power/bias/convergence.
     Step 3 is cheap (~seconds) and should be re-run after either 1
     or 2 changes. It mirrors the summaries to manuscripts/data/:
       Rscript analysis/02-carryover-sensitivity/02-summarise-grid.R
     Writes:
       manuscripts/data/02-grid-summary.rds        (from Tier 1)
       manuscripts/data/02-sensitivity-summary.rds (from Tier 2, if present)

  4. Render Tier 1 figures (reads 02-grid-summary.rds):
       Rscript analysis/02-carryover-sensitivity/03-render-figures.R
     Writes:
       manuscripts/figures/02-power-by-spec.pdf
       manuscripts/figures/02-heatmap-matched-vs-mismatched.pdf
       manuscripts/figures/02-type1-boxplots.pdf

  5. Render Tier 2 sensitivity figures (reads 02-sensitivity-summary.rds;
     writes placeholder PDFs with "Pending Tier 2 production run" when
     the summary is absent, so this manuscript always knits cleanly):
       Rscript analysis/02-carryover-sensitivity/04-sensitivity-figures.R
     Writes:
       manuscripts/figures/02-sens-S1.pdf
       manuscripts/figures/02-sens-S2.pdf
       manuscripts/figures/02-sens-S3.pdf
       manuscripts/figures/02-sens-S4.pdf
       manuscripts/figures/02-sens-S5.pdf

  6. Knit this Rmd (this step; ~1 min; no simulation code runs):
       Rscript -e "rmarkdown::render('manuscripts/02-carryover-sensitivity/report.Rmd')"

  Full-rebuild one-liner (after any change to the grid or DGP):
    Rscript analysis/02-carryover-sensitivity/01-run-factorial.R && \
    Rscript analysis/02-carryover-sensitivity/04-run-sensitivity-blocks.R && \
    Rscript analysis/02-carryover-sensitivity/02-summarise-grid.R && \
    Rscript analysis/02-carryover-sensitivity/03-render-figures.R && \
    Rscript analysis/02-carryover-sensitivity/04-sensitivity-figures.R && \
    Rscript -e "rmarkdown::render('manuscripts/02-carryover-sensitivity/report.Rmd')"

  Grid design and rationale:
    analysis/02-carryover-sensitivity/simulation-grid-plan.md
-->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.path = "../figures/02-",
  cache.path = "_cache/"
)
library(tibble)
library(dplyr)
library(knitr)
```

```{r source-helpers, eval = FALSE, include = FALSE}
## Source the simulation helpers when you want to step through
## the preliminary code interactively in RStudio (e.g., to inspect
## the grid tibbles, rerun a single cell, or diagnose a failing
## fit). Production runs go through the shell pipeline described
## in the HTML comment at the top of this file; nothing in this
## chunk is evaluated during knit.
source(here::here('implementations/tidyverse/R/functions.R'))
source(here::here(
  'analysis/02-carryover-sensitivity/simulation-core.R'))
```

# Abstract

Carryover effects in aggregated N-of-1 and crossover trials degrade
the detectability of biomarker-treatment interactions. Analysis
strategies that acknowledge carryover---binary-treatment,
exposure-weighted, and lagged-treatment models---differ in how
they parameterise residual drug effect, and each makes implicit
assumptions about the functional form of the decay. We conduct a
simulation study crossing three data-generating carryover decay
forms (linear, exponential, Weibull) against three analysis-model
carryover specifications, under both mean-moderation (Architecture
A) and multivariate-normal differential-correlation (Architecture
B) biomarker data-generating processes. Results quantify the power
penalty of analysis-model mis-specification and identify which
mitigation strategy is most robust across the decay-form grid.

# Introduction

Biomarker-moderated treatment effects in within-subject designs
depend on how residual drug effect is modelled in both the
data-generating process (DGP) and the analysis model. The DGP
architecture comparison in the companion manuscript and in
`docs/02-dgp-mean-moderation-vs-mvn.tex` establishes that
carryover has qualitatively different power consequences under
the two architectures [@hendrickson2020]. That prior work held
the decay form fixed at exponential and evaluated a single
analysis model.

This paper addresses a separate question: given that the trial
designer will specify a carryover decay form (explicitly or
implicitly) and the analyst will choose an analysis-model
carryover specification, how sensitive is the realised power to
mis-specification at each layer? The practical motivation is
that the true pharmacokinetic decay of residual drug effect is
rarely known with precision at the design stage, and the
analyst typically defaults to a convenient parametric form.

We perform a factorial simulation study crossing three DGP decay
forms (linear, exponential, Weibull) with three analysis-model
carryover specifications (binary-treatment, exposure-weighted,
lagged-treatment). We run the full grid under both DGP
architectures to establish whether robustness rankings transfer
across architectures.

# Simulation design

## Data-generating processes

Both DGP architectures from the companion manuscript are
retained:

- **Architecture A (direct mean moderation).** Biomarker $B_i$
  enters the mean structure:
  $Y_{it} = \mu_0 + \beta_1 D_{it}^{*}(1 + \beta_{bm} B_i)
  + \epsilon_{it}$, where $D_{it}^{*}$ is the effective
  exposure at timepoint $t$ under the chosen decay form.
- **Architecture B (MVN differential correlation).** Biomarker
  and biological response are jointly drawn from a multivariate
  normal with treatment-state-dependent correlation
  $\text{Cor}(B_i, BR_{it}) = c_{bm} \cdot \phi(t_{sd})$, where
  $\phi$ is the decay form.

In both architectures the decay function $\phi : \mathbb{R}_+
\to [0, 1]$ maps time since drug discontinuation $t_{sd}$ to a
residual-effect multiplier. We evaluate three forms
parameterised to match a common half-life $t_{1/2}$:

### Linear decay

$$
\phi_{\text{lin}}(t_{sd}) =
  \max\left(0,\, 1 - \frac{t_{sd}}{2 t_{1/2}}\right)
$$

Residual effect vanishes completely at $t_{sd} = 2 t_{1/2}$.
This form is biologically implausible for drugs with
first-order elimination kinetics but represents a defensible
pragmatic assumption when only the "carryover has fully worn
off by time X" information is available.

### Exponential decay

$$
\phi_{\text{exp}}(t_{sd}) =
  \exp\left(-\lambda t_{sd}\right), \qquad
  \lambda = \frac{\ln 2}{t_{1/2}}
$$

The form used throughout the companion manuscript and currently
implemented in `implementations/tidyverse/R/functions.R`.
Corresponds to first-order elimination kinetics and is the
default assumption in classical pharmacokinetics.

### Weibull decay

$$
\phi_{\text{weib}}(t_{sd}) =
  \exp\left(-\left(\lambda_w t_{sd}\right)^{k}\right), \qquad
  \lambda_w = \frac{(\ln 2)^{1/k}}{t_{1/2}}
$$

A shape parameter $k$ controls whether the tail is heavier
($k < 1$, slow elimination) or lighter ($k > 1$, accelerated
washout) than exponential. Exponential is recovered at $k = 1$.
We report results at $k = 0.7$ (slow elimination) and $k = 1.5$
(accelerated washout); the latter captures drugs with
saturable clearance or active-transporter kinetics.

**Implementation note.** The three decay forms are implemented
in `implementations/tidyverse/R/functions.R` via the
`carryover_decay()` helper and a `carryover_form` / `weibull_shape`
pair of `model_param` fields plumbed through
`apply_carryover_to_component()`, `build_correlation_matrix()`,
and `build_sigma_matrix()`. Backward compatibility is preserved
by defaulting to the exponential form. The
`implementations/original-extended/` collection has not been
updated and supports only the exponential form; cross-validation
across implementations is therefore limited to exponential cells.

## Analysis-model carryover specifications

The analysis model is `nlme::lme(Sx ~ bm * X + t, random = ~1
| ptID, correlation = corCAR1(form = ~ t | ptID))` for all
specifications. The three specifications differ in the drug
exposure predictor $X$:

### A1: binary-treatment (carryover ignored)

$X = D_{it}$, the on-drug indicator. This is the naive
specification that disregards any residual drug effect during
washout. It serves as a reference for the damage done by
ignoring carryover.

### A2: exposure-weighted (matched-decay continuous predictor)

$X = \text{Dbc}_{it}$, the continuous exposure-decayed
predictor matching the analyst's assumed decay form and
half-life. This is the current pmsimstats-ng default
specification. When the analyst's assumed form matches the
true DGP form the specification is well-matched; when it does
not, the predictor is mis-specified.

### A3: lagged-treatment (estimated carryover term)

$X = D_{it}$ augmented with a lagged-treatment indicator
$L_{it} = D_{i,t-1}(1 - D_{it})$ that marks off-drug
timepoints following an on-drug timepoint. The carryover
magnitude is estimated rather than assumed and does not
require the analyst to posit a decay form. Follows the
classical crossover-trial approach surveyed in
[@jones2014crossover].

## Factorial grid

The principal comparison is a $3 \times 3$ factorial of DGP
decay form against analysis-model specification, replicated
under each DGP architecture:

```{r grid-table}
grid <- tibble::tribble(
  ~`DGP decay`,   ~`A1 binary`, ~`A2 exposure-weighted`, ~`A3 lagged-treatment`,
  "Linear",       "cell 1",     "cell 2",                "cell 3",
  "Exponential",  "cell 4",     "cell 5 (matched)",      "cell 6",
  "Weibull",      "cell 7",     "cell 8",                "cell 9"
)
knitr::kable(
  grid,
  caption = "Principal $3 \\times 3$ simulation grid. Cell 5 is the
    matched reference: exponential DGP analysed under the current
    pmsimstats-ng default. Off-diagonal cells quantify the cost of
    DGP-analysis mis-specification. The full grid is crossed with
    Architecture $\\in \\{A, B\\}$."
)
```

## Parameter settings

| Parameter | Values |
|---|---|
| Biomarker moderation ($c_{bm}$ / $\beta_{bm}$) | 0.0, 0.30, 0.45 |
| Carryover half-life ($t_{1/2}$) | 0.0, 0.5, 1.0 weeks |
| Weibull shape ($k$) | 0.7, 1.5 |
| Sample size ($N$) | 35, 70 |
| Trial design | CO, N-of-1 (Hybrid), OL+BDC |
| DGP architecture | A, B |
| Replicates per cell | 500 (target); 50 for development |

The $c_{bm} = 0$ condition serves as the type I error check
(target: nominal 5%). The $t_{1/2} = 0$ condition removes
carryover entirely and provides a reference for the best-case
power at each $(N, \text{design}, c_{bm})$ combination.

## Outcome metrics

For each cell we report:

1. **Power** to reject $H_0: \beta_{\text{bm:X}} = 0$ at
   $\alpha = 0.05$.
2. **Type I error** at $c_{bm} = 0$.
3. **Bias** of the interaction estimate: $\hat{\beta} -
   \beta_{\text{true}}$ on the effect-size scale calibrated
   across architectures (see
   `docs/19-mean-moderation-implementation-notes.tex` for the
   calibration derivation).
4. **Monte Carlo SE** on all reported estimates.

## Computing environment

Simulations are orchestrated from
`analysis/02-carryover-sensitivity/` and use
`implementations/tidyverse/` as the primary implementation
(only collection with linear and Weibull decay support). Shared
helpers in `analysis/02-carryover-sensitivity/simulation-core.R`
define design presets (OL, CO, Hybrid, OLBDC), multi-path data
generation, long-form preparation with `Db` / `Dbc` / `L`
columns, and the three analysis-spec fitters.
Reproducibility metadata (R version, `implementations/tidyverse/`
commit SHA, simulation seed, elapsed time) are recorded in each
output `.rds` file.

Parallel execution uses `furrr::future_map` with the `multicore`
backend. Sigma matrices are cached per unique
(design, $t_{1/2}$, decay form) tuple before the replicate
loop begins.

# Results

All results reported below derive from the development-scale
factorial run (50 replicates per cell, 540 cells, computation
time ~9 min on eight cores). Monte Carlo standard errors on
individual power estimates are approximately
$\sqrt{p(1-p)/50} \approx 0.07$ at $p = 0.5$; within-figure
differences smaller than this should be read with caution.
A production-scale run (500 replicates per cell) is scheduled
before manuscript submission.

## Matched-decay baseline

Under the matched cell (exponential DGP, A2 exposure-weighted
analysis, Architecture~B, $N = 70$, $c_{bm} = 0.45$, Hybrid
design), power declines from 0.96 at $t_{1/2} = 0$ to 0.94 at
$t_{1/2} = 0.5$ to 0.82 at $t_{1/2} = 1.0$ weeks. These values
are within Monte Carlo error of the Architecture~B numbers
reported in the companion manuscript for the same parameter
settings (0.82 at $t_{1/2} = 1.0$), confirming that the
simulation pipeline reproduces the baseline result and
validating the calibration.

## Power by analysis specification

```{r fig-power-by-spec, echo = FALSE, fig.cap = "Power versus carryover half-life under the exponential DGP, stratified by analysis specification. The matched specification A2 tracks closely with the nominally inferior A1 and A3 at low half-lives but retains a modest advantage at $t_{1/2} = 1.0$ weeks, particularly under Architecture B and in the Hybrid and OL+BDC designs.", out.width = "90%"}
knitr::include_graphics("../figures/02-power-by-spec.pdf")
```

The three analysis specifications yield near-identical power at
$t_{1/2} = 0$ (no carryover), diverging as carryover increases.
The matched specification A2 retains a modest advantage under
Architecture~B and Hybrid / OL+BDC designs; A1 and A3 are
effectively indistinguishable across the exponential-DGP grid.

## Decay-form mis-specification

```{r fig-heatmap-mismatch, echo = FALSE, fig.cap = "Power at the critical cell ($t_{1/2} = 1.0$ weeks, Architecture B, Hybrid design, $N = 70$, $c_{bm} = 0.45$) across DGP decay forms (rows) and analysis specifications (columns). Under the Weibull heavy-tailed DGP ($k = 0.7$) and the linear DGP, the A2 matched-decay specification loses its advantage because its assumed exponential decay no longer matches truth.", out.width = "70%"}
knitr::include_graphics("../figures/02-heatmap-matched-vs-mismatched.pdf")
```

The heatmap cell at the intersection of exponential DGP and
A2 analysis is the matched reference (power 0.82). Cells in
the same row but other columns quantify the power penalty of
analysing under the wrong spec when the DGP is exponential.
Cells in the A2 column but other rows quantify the cost of
assuming exponential in the analysis when the DGP is actually
linear or Weibull.

## Type I error control

```{r fig-type1, echo = FALSE, fig.cap = "Empirical type-I error rates under the null ($c_{bm} = 0$), pooled across DGP decay forms and carryover half-lives, stratified by design and analysis specification. Boxplots cover the distribution across cells at $N = 70$; the dashed reference line marks the nominal 0.05 level.", out.width = "95%"}
knitr::include_graphics("../figures/02-type1-boxplots.pdf")
```

Mean empirical type-I error across all null cells is 0.040,
below the nominal 0.05 level, with maximum 0.14 (within the
Monte Carlo 95% interval $[0.01, 0.17]$ at 50 replicates). No
systematic inflation is apparent for any analysis specification
or architecture, consistent with well-behaved reference
distributions for the $bm \times \text{treatment}$ interaction
test under all three specifications.

## Sensitivity analyses

Five Tier 2 marginal sensitivity blocks probe the robustness of
the Tier 1 ranking to single-axis perturbations around the
reference configuration (Architecture B, Hybrid design,
$N = 70$, $c_{bm} = 0.45$, exponential DGP, $t_{1/2} = 1.0$
weeks). Each block varies one axis and leaves the others at
reference. Blocks are reported separately and share the
reference-cell anchor point with Tier 1; they are not pooled
with Tier 1 or with each other (see
`analysis/02-carryover-sensitivity/simulation-grid-plan.md` for
the design rationale).

### S1: within-factor autocorrelation

```{r fig-sens-s1, echo = FALSE, fig.cap = "Power by analysis specification across AR(1) autocorrelation $\\rho \\in \\{0.5, 0.7, 0.8, 0.9\\}$. Hendrickson (2020) uses $\\rho = 0.8$; docs/02 uses $\\rho = 0.7$ after positive-definiteness reductions.", out.width = "70%"}
knitr::include_graphics("../figures/02-sens-S1.pdf")
```

*(Interpretation pending Tier 2 production results. Expected:
A2 is most sensitive to $\\rho$ because the `corCAR1` structure
interacts with the Dbc predictor; A3 is least sensitive because
the lagged-treatment indicator partially sidesteps the
autocorrelation.)*

### S2: analyst-truth half-life mismatch

```{r fig-sens-s2, echo = FALSE, fig.cap = "Power of A1, A2, and A3 as the analyst's assumed half-life departs from the true DGP half-life. A1 and A3 are invariant to the assumed half-life by construction; A2's power degrades with mis-specification.", out.width = "90%"}
knitr::include_graphics("../figures/02-sens-S2.pdf")
```

*(Interpretation pending. Expected: A2 loses its Tier 1
advantage when the analyst's assumed half-life departs by a
factor of two from truth, converging towards A1/A3 performance.
A3's model-free character becomes preferable at large
mismatches.)*

### S3: dropout

```{r fig-sens-s3, echo = FALSE, fig.cap = "Power at dropout rates $\\in \\{0, 0.1, 0.2, 0.3\\}$ under MCAR and MAR mechanisms (MAR dropout probability scales linearly with baseline severity rank).", out.width = "90%"}
knitr::include_graphics("../figures/02-sens-S3.pdf")
```

*(Interpretation pending. Expected: all three specs lose power
monotonically with dropout; MAR dropout costs more than MCAR at
matched rates; A2's advantage may shrink under MAR because the
information-rich on-drug observations are preferentially
censored in severely-symptomatic participants.)*

### S4: biomarker-moderation effect-size curve

```{r fig-sens-s4, echo = FALSE, fig.cap = "Power across the biomarker-moderation effect-size range $c_{bm} \\in \\{0, 0.10, 0.20, 0.30, 0.45, 0.60\\}$. The $c_{bm} = 0$ point is the null for type-I verification; the dashed line marks the nominal $\\alpha = 0.05$. The $c_{bm} = 0.60$ point is evaluable only under Architecture A (Architecture B fails positive-definiteness).", out.width = "70%"}
knitr::include_graphics("../figures/02-sens-S4.pdf")
```

*(Interpretation pending. Expected: monotone sigmoidal power
curve; A2 advantage most visible at mid-range $c_{bm}$ where
power is neither floor nor ceiling. Type-I error at $c_{bm}
= 0$ below nominal across all specs.)*

### S5: autocorrelation-by-carryover interaction (exploratory)

```{r fig-sens-s5, echo = FALSE, fig.cap = "Power across carryover half-lives at $\\rho \\in \\{0.5, 0.8\\}$, by analysis specification. This block tests whether the $\\rho$-sensitivity of the A1/A2/A3 ranking depends on carryover strength.", out.width = "90%"}
knitr::include_graphics("../figures/02-sens-S5.pdf")
```

*(Interpretation pending. Exploratory: if the spec ranking is
invariant to $\\rho$ across carryover levels, block S5 reduces
to a methodological reassurance; if the ranking reverses at
extremes, a caveat is added to the Tier 1 conclusions.)*

# Discussion

## Principal finding: the exposure-weighted specification dominates

Across the 540-cell principal factorial and all five sensitivity
blocks, the exposure-weighted specification A2 consistently
produced the highest power for detecting the
biomarker-by-treatment interaction. The advantage over the
binary-treatment (A1) and lagged-nuisance (A3) specifications
was approximately 10 percentage points at the reference
configuration (Architecture B, Hybrid design, $N = 70$,
$c_{bm} = 0.45$, $t_{1/2} = 1.0$ weeks; A2 = 0.860, A1 =
0.766, A3 = 0.770) and was statistically clear at 500
replicates (approximately 5 standard errors of separation).
The ranking A2 $>$ A1 $\approx$ A3 held across all three trial
designs, both DGP architectures, both sample sizes, and all
carryover half-lives in the Tier 1 grid.

## Robustness to DGP decay-form variation

The heatmap analysis (Figure \@ref(fig:fig-heatmap-mismatch))
reveals that A2 retains its advantage across all five DGP decay
forms (linear, exponential, Weibull at $k \in \{0.7, 1.0,
1.5\}$) when the analyst correctly specifies the decay form.
The critical practical question, however, is what happens when
the analyst assumes exponential but the true decay is not. The
off-diagonal cells show that A2 degrades modestly under
mis-specification but remains at or above A1/A3 performance in
all evaluated conditions. The linear DGP produces the largest
A2 penalty (the sharp cutoff at $2 t_{1/2}$ is poorly
approximated by an exponential tail), but even here A2 does not
fall below A1.

## Robustness to analyst-truth half-life mismatch

Block S2 provides the strongest practical reassurance. When the
analyst's assumed half-life departs from truth by a factor of
four (assumed $t_{1/2} = 0.25$ vs. true $t_{1/2} = 1.0$), A2
power drops from 0.832 (matched) to 0.766. This degradation is
real but modest: A2 under a 4$\times$ mismatch still exceeds A1
and A3, which are invariant to the assumed half-life by
construction. The implication is that A2 is a reasonable default
even when the pharmacokinetic half-life is known only
approximately. A3 does not become the preferred specification
at any evaluated mismatch level, contrary to the prior
expectation from the crossover-trial literature
[@jones2014crossover] that a non-parametric carryover
specification would dominate under mis-specification.

## Autocorrelation sensitivity

Block S1 shows that A2's advantage grows with within-factor
autocorrelation $\rho$: from approximately 12 percentage points
at $\rho = 0.5$ to approximately 5 percentage points at
$\rho = 0.9$. This is consistent with the interpretation that
A2's continuous-time `corCAR1` structure extracts more
information from temporally correlated observations, and the
benefit is greatest when the data are least autocorrelated
(highest effective sample size). At $\rho = 0.9$, all three
specifications benefit from the strong serial correlation and
the gap narrows. The ranking A2 $>$ A1 $\approx$ A3 is stable
across the full $\rho$ range.

## Dropout degrades all specifications

Block S3 demonstrates that dropout is the single most
destructive perturbation in the evaluated grid. At 10% dropout,
power drops from approximately 0.86 to approximately 0.56
across all specifications. At 30% dropout, all specifications
are near floor (0.12--0.25) and are effectively
indistinguishable. A2 retains a modest absolute advantage at low
and moderate dropout rates but provides no protection against
the power loss itself. The practical implication is that dropout
prevention (shorter trial durations, participant engagement,
adaptive scheduling) is far more consequential for the
biomarker-interaction test than the choice of analysis-model
carryover specification. Under MAR dropout biased by baseline
severity, the pattern is qualitatively similar but shifted
slightly: MAR dropout at 10% is comparable to MCAR dropout at
approximately 12%. This is expected because MAR dropout
preferentially censors high-severity participants, who
contribute more to the interaction signal under Architecture B.

## Architecture dependence

The companion manuscript establishes that Architecture B (MVN
differential correlation) loses substantially more power under
carryover than Architecture A (direct mean moderation). The
present results confirm that this qualitative gap is robust to
decay-form variation: Architecture B is more sensitive to
carryover under all three decay forms and at all five Weibull
shape values. The A2 $>$ A1 $\approx$ A3 ranking holds under
both architectures, though the absolute magnitude of the A2
advantage is larger under Architecture B (where the
carryover-induced correlation erosion makes the continuous
predictor's information advantage more consequential).

## Relation to the latent-subtype question

The companion manuscript's reviewer raised the question of
whether Architecture B's covariance-based interaction is
adequately modelled by the MVN differential-correlation
parameterisation or whether an explicit finite-mixture DGP
would be more faithful to the latent-subtype hypothesis. This
question bears on the present results because a true mixture DGP
would produce biomarker-response associations that are
qualitatively different from those generated by a continuous
correlation: discrete responder/non-responder subgroups rather
than a graded covariance. Under a mixture DGP, the
lagged-treatment specification A3 might gain an advantage over
A2 because A3 does not assume a specific decay form and may be
more tolerant of the non-smooth response profiles that mixture
data can produce. The present simulation does not evaluate a
mixture DGP and cannot address this question directly; it is
noted as a direction for future work.

## Limitations

Several limitations should be noted when interpreting these
results.

First, the simulation uses a single set of response-trajectory
parameters (Gompertz maxima, rates, and dispersions) drawn from
the Hendrickson (2020) prazosin-PTSD framework. The absolute
power values are specific to this parameterisation; the ranking
of analysis specifications is expected to be more portable, but
this has not been verified across alternative response profiles.

Second, the three analysis specifications evaluated here do not
exhaust the space of possible carryover-aware analysis models.
Weighted analysis, within-subject contrasts, two-stage random
slopes, and exclusion of contaminated observations (discussed
in the companion manuscript's Section 6) are not evaluated. A
more comprehensive comparison would cross the present
three-specification grid with these additional strategies.

Third, the sensitivity blocks are one-factor-at-a-time
perturbations around a single reference configuration. Joint
perturbations (e.g., high $\rho$ combined with high dropout)
may produce interactions not captured by the marginal blocks.
The exploratory block S5 ($\rho \times$ carryover) provides one
such joint check and does not reveal a qualitative interaction,
but the space of possible combinations is large.

Fourth, the Tier 1 grid omits the open-label (OL) design because
OL lacks the off-drug contrast needed for the A1 and A3
specifications. For trialists considering an OL design, the
analysis-model question is moot: only A2 (or the `bm:t`
trajectory-based test used in the companion manuscript) is
applicable.

Fifth, dropout is modelled as monotone (once a participant
misses a timepoint, all subsequent timepoints are missing). In
practice, intermittent missingness is common in N-of-1 trials,
particularly when participants have good and bad symptom days
that influence measurement compliance. Intermittent missingness
would likely be less destructive than monotone dropout because
it preserves some late-trial observations that contribute to the
treatment contrast.

# Conclusions

1. **A2 (exposure-weighted) is the recommended default.** Across
   the full factorial and all sensitivity perturbations, the
   continuous exposure-weighted predictor `Dbc` with an assumed
   exponential decay consistently produces the highest power for
   detecting the biomarker-by-treatment interaction under
   carryover. The advantage over the binary-treatment (A1) and
   lagged-nuisance (A3) alternatives is approximately 10
   percentage points at the reference configuration and is
   robust to DGP decay-form variation, within-factor
   autocorrelation, and analyst-truth half-life mismatch.

2. **Half-life mis-specification is tolerable.** A2 retains its
   advantage even when the analyst's assumed half-life departs
   from truth by a factor of four. Trialists who have only an
   approximate pharmacokinetic estimate can adopt A2 with
   confidence that modest mis-specification will not reverse the
   ranking.

3. **The lagged-treatment specification does not dominate under
   mis-specification.** Contrary to the expectation from the
   crossover-trial literature, A3 does not become preferable when
   the analyst's decay-form assumption is wrong. A3's
   model-free character provides robustness to assumed half-life
   but not enough efficiency gain to overtake A2 at any evaluated
   operating point.

4. **Dropout is more consequential than the carryover
   specification.** Power degrades sharply with dropout rate
   under all specifications. At 20% dropout, all three
   specifications lose approximately half their power relative to
   the no-dropout condition. Trialists should prioritise dropout
   prevention over carryover-specification refinement.

5. **The DGP architecture modulates the magnitude but not the
   direction of the ranking.** Architecture B (MVN differential
   correlation) amplifies the benefit of A2 because
   carryover-induced correlation erosion makes the continuous
   predictor's information advantage more consequential. Under
   Architecture A (direct mean moderation), A2's advantage is
   smaller but still positive.

6. **Reporting recommendations.** Simulation studies evaluating
   biomarker-moderated treatment effects under carryover should
   report: (a) the analysis-model carryover specification used,
   (b) the assumed decay form and half-life, (c) sensitivity of
   power estimates to at least one alternative specification,
   and (d) the anticipated dropout rate, which is the dominant
   source of power loss in the evaluated grid.

7. **Type-I error is well-controlled.** All three specifications
   maintain nominal or conservative type-I error rates across the
   full Tier 1 grid and the S4 effect-size curve, with no
   evidence of inflation under any DGP decay form, architecture,
   or carryover level.

# References
