---
title: "Three-component decomposition of treatment response in
  aggregated N-of-1 trials: pharmacological, expectation-driven,
  and natural-history components"
author:
  - pmsimstats team
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    number_sections: true
    keep_tex: true
    highlight: tango
fontsize: 11pt
geometry: "left=3cm,right=5cm,top=2cm,bottom=2cm"
bibliography: references.bib
csl: statistics-in-medicine.csl
link-citations: true
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{lineno}
  - \linenumbers
  - \usepackage{setspace}
  - \setstretch{1.1}
knit: (function(input, encoding) source(normalizePath(file.path(dirname(input), '..', '..', '..', 'tools', 'stage-render.R')))$value(input, encoding))
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)
```

```{r footer-setup, include = FALSE, results = 'asis'}
if (knitr::is_latex_output()) {
  src <- tryCatch(normalizePath(knitr::current_input()),
                  error = function(e) 'unknown')
  src_disp <- gsub('^/Users/zenn/Dropbox', '~/Dropbox', src)
  src_disp <- gsub('^/Users/zenn', '~', src_disp)
  src_tex <- gsub('([_~^&%#$])', '\\\\\\1', src_disp)
  ts <- format(Sys.time(), '%Y-%m-%d %H:%M %Z')
  cat(sprintf('\\renewcommand{\\rmdsource}{Source: \\texttt{%s}}\n',
              src_tex))
  cat(sprintf('\\renewcommand{\\rendertimestamp}{Rendered: %s}\n', ts))
}
```

# Abstract

**Background.** Treatment response in aggregated N-of-1 clinical
trials reflects three causally distinct mechanisms: pharmacological
action of the active compound, the patient's belief that the
treatment will work, and natural-history drift in the underlying
condition. Standard analysis models lump these into a single
'treatment effect' plus residual error, which conflates clinically
distinct quantities and compromises both regulatory inference and
predictive-biomarker validation.

**Methods.** We formalise a three-component decomposition of the
within-participant outcome trajectory into a biological response
component (BR), a placebo-belief response component (PB), and a
time-variant natural-history component (TV), each modelled as a
modified Gompertz function with participant-specific random
effects. We characterise the trial-design contrasts (open-label,
blinded discontinuation, crossover) that supply identifiability of
each component, derive the linear-mixed approximations to the full
nonlinear model that retain the pharmacological-component target
estimand, and quantify the inferential cost of failing to
decompose through worked examples and a Monte Carlo simulation
study calibrated to the prazosin-PTSD reference parameters.

**Results.** The framework is developed and pre-registered for
empirical evaluation; a 100-replicate pilot of the analysis
machinery is reported, and a production simulation programme at
1,000 replicates per cell across $N \in \{35, 70, 100, 150\}$ is
pre-registered for execution.

**Conclusions.** The three-component framework is the principled
basis for predictive-biomarker validation in aggregated N-of-1
trials and clarifies the design contrasts that are mandatory
versus optional. A pragmatic analysis specification consistent
with the framework but tractable at trial-relevant sample sizes is
`Sx ~ bm + t + Dbc + phase + Dbc:phase + bm:Dbc + bm:Dbc:phase`
with random intercept and AR(1) residual correlation.

# Introduction

## A claim about how N-of-1 trials should be analysed

This paper advances an opinionated methodological claim:
treatment response in aggregated N-of-1 clinical trials is the
sum of three causally distinct components and **should routinely
be analysed as such**, rather than estimated as a single lumped
quantity. The three components are pharmacological action ($BR$,
the biological response), placebo-belief response ($PB$, the
patient's belief that they are receiving treatment), and
time-variant natural history ($TV$, the trajectory that would
have occurred without intervention). Each is causally distinct,
each is independently of clinical and regulatory interest, and
each is identifiable from a properly designed N-of-1 trial. We
will argue that the standard practice of fitting a single
mixed-effects model with one treatment indicator, prevalent
across the published N-of-1 literature, conflates these three
components, biases the interpretation of treatment effect, and
makes predictive-biomarker validation unanswerable as a
scientific question.

The framework we advocate is not new in its mathematical
ingredients. Each individual piece (mixed-effects models with
trial-phase indicators, placebo-arm contrasts, regression-to-
mean adjustment, growth-curve parameterisation) has appeared
across decades of clinical-trial methodology. What is novel,
and what we argue should become routine, is the **explicit
joint decomposition**: simultaneous estimation of all three
components from a single N-of-1 trajectory using trial-design
contrasts that uniquely identify each one. The published N-of-1
literature has, with rare exceptions, not adopted this
joint-decomposition framing. We will document the gap, explain
why the gap has persisted (parsimony, identifiability concerns,
sample-size constraints, methodological inertia), and propose a
methodology that is tractable at trial-relevant sample sizes
through a phase-augmented linear-mixed-effects approximation to
the full decomposition.

## Why this matters: three failure modes of the lumped analysis

A one-component analysis ('the treatment effect is what we
observe minus baseline') has three identifiable failure modes
that arise when the population displays meaningful heterogeneity
in any of the unmodelled components. None of these is
hypothetical: each has a documented presence in the
chronic-condition trial literature.

First, the **placebo-attribution failure**. When the population
contains both high and low placebo responders, the lumped
treatment effect overestimates the pharmacological component.
The bias scales with the magnitude and variance of $PB$ in the
study population, and it is invariant under sample-size
increases: a larger one-component trial converges on the same
biased estimate. The most consequential example in modern
clinical pharmacology is antidepressant efficacy: @kirsch2008
analysed FDA-submitted data on four widely-prescribed
antidepressants and showed that, except at the most severe
baseline-severity stratum, mean drug-placebo differences fell
below clinical-significance thresholds, with the apparent
'efficacy' largely explained by placebo response and natural
fluctuation; @locher2015 reproduced this in late-life
depression and explicitly invoked regression to the mean.
@sugarman2014 generalised the finding to anxiety. Patient
populations differ systematically in placebo responsiveness,
and the difference has measurable biological correlates: COMT
genotype predicts placebo response in irritable bowel syndrome
(@wang2022comt), indicating that 'placebo' is not a
homogeneous nuisance but a structured biological phenomenon
that interacts with the patient's genome and treatment history.
A trial whose population happens to be placebo-responsive will
report a larger 'drug effect' than one whose population is not,
even when the underlying pharmacology is identical.

Second, the **regression-to-the-mean failure**. When patients
enrol at moments of clinical engagement that coincide with peak
symptom severity, natural-history regression toward typical
severity contributes several points of apparent improvement in
the early weeks of any trial, without any pharmacological
contribution at all. The systematic reviews of
@hrobjartsson2001, @hrobjartsson2004, and @hrobjartsson2010
across more than two hundred trials and sixty conditions
quantified this directly: in trials with both placebo and
no-treatment arms, much of what is conventionally attributed to
the placebo response was indistinguishable from regression to
the mean and natural-history drift. The lumped analysis
absorbs this regression into the treatment-effect estimate and
the analyst has no formal mechanism to recover the true
pharmacological effect.

Third, the **biomarker-validation failure**. The
precision-medicine question that motivates many recent N-of-1
trials is whether a baseline biomarker predicts treatment
response, formalised in the PATH statement [@kent2020path;
@kent2020pathee]. A biomarker that correlates with placebo
responsiveness or with favourable natural-history trajectory
will appear in a one-component analysis as a predictor of the
lumped 'treatment effect' indistinguishable from a true
pharmacological predictor. The biomarker-stratified prescribing
decision that follows from the analysis is wrong, and the
trial's regulatory utility is degraded. The PATH framework
explicitly defers the *mechanism* of treatment-effect
heterogeneity to domain knowledge [@kent2018bmj;
@rekkas2020scoping]; the BR-PB-TV decomposition supplies the
mechanistic substrate that PATH leaves blank. We argue that
predictive-biomarker validation cannot be conducted rigorously
in a lumped-analysis framework regardless of how large the
trial is.

## The current state of the N-of-1 literature

The aggregated N-of-1 trial design has matured rapidly over the
last fifteen years. The foundational programmatic statement is
@lillie2011; the methodological reviews of @duan2013, the
historical-practice audit of @gabler2011, and the more recent
syntheses of @schork2017nutrition, @schork2018rct,
@schork2023precision, and @porcino2022 trace the trajectory
from individual-patient consultations to the
aggregated-meta-analytic framework that supports
population-level inference. Reporting standards are codified in
the CONSORT extension @vohra2015cent (the headline statement)
and @shamseer2015cent (the explanation-and-elaboration
companion); the reporting-quality audits of @li2016cent and
its successors document slow but progressive improvement in
analytic transparency.

Three observations about this literature motivate the present
paper.

**The dominant analytic models are simple.** The audit work of
@li2016cent documents that the published N-of-1 trials of the
preceding three decades used paired t-tests, simple linear
mixed-effects models with a binary treatment indicator, or
hierarchical Bayesian regressions, with no published trial
employing a structural decomposition of treatment response into
pharmacological, expectancy, and natural-history components.
The most recent methodological advances retain this character:
@liao2023 develops Bayesian distributed-lag models with
autocorrelated errors as the state-of-the-art carryover-aware
analysis, but the model still treats the response as a single
quantity to be estimated. @blackston2019 and @carrozzo2025
compare aggregated N-of-1 trials with parallel and crossover
RCT alternatives by simulation and document carryover-induced
type-I-error inflation as a key failure mode, but again at the
level of the lumped treatment effect. None of these papers
decompose the response.

**The placebo-component separation problem is widely known but
largely unaddressed in N-of-1 specifically.** The blinded
discontinuation phase used in the Hendrickson hybrid design
[@hendrickson2020] is one of the few protocol-level mechanisms
in routine N-of-1 use that supplies identifiability of
$PB$ versus $BR$. Other relevant approaches (placebo run-in
designs, open-label-placebo trials, mediator-analysis
applications) have been developed in parallel but rarely
imported into N-of-1 methodology.

**The closest precedent is Kaptchuk's three-component
decomposition of the placebo response itself.** In a randomised
trial of acupuncture-style placebo for irritable bowel
syndrome, @kaptchuk2008components decomposed the *placebo*
response into observation, ritual, and patient-practitioner
relationship components, demonstrating empirically that the
placebo response is itself a sum of identifiable parts rather
than a single 'expectancy' lump. The follow-up open-label
placebo trials of @kaptchuk2010olp, @lembo2021olp, and
@nurko2022 in adolescent functional pain establish that even
disclosed-placebo intervention can produce clinically
meaningful PB-only response. The framework we propose is
structurally analogous to Kaptchuk's decomposition (a
decomposition into causally distinct components, each with its
own identifiability) but applied at the level of the *full*
treatment response (drug, placebo, natural history) rather than
within the placebo response alone. To our knowledge, no
published paper proposes the corresponding decomposition for
the aggregated N-of-1 setting, even though the methodological
ingredients (mixed-effects modelling, blinded-phase contrasts,
trajectory parameterisation) are all in routine use.

The case we develop in the rest of this paper is therefore that
the gap is real, that the methodological tools to fill it are
mature, and that the failure modes of the lumped analysis
(placebo-attribution, regression-to-mean, biomarker-validation)
are large enough in chronic-condition pharmacology to make the
decomposition a methodological priority rather than an academic
refinement.

## The neurobiological and epidemiological reality of the components

The decomposition is not a statistical convenience. Each
component reflects a real causal mechanism documented in
independent literatures.

The biological response component represents the
pharmacological action of the active compound: the chemical
binding, downstream signalling, and physiological response
that a regulatory authority cares about when it grants
approval. The placebo-belief response reflects neurobiologically
real mechanisms documented across pain, depression, Parkinson's,
and PTSD research [@benedetti2014; @frisaldi2020;
@carlino2014; @carlino2016benedetti], mediated by belief-driven
release of endogenous opioids and dopamine, anticipatory
regulation of the hypothalamic-pituitary-adrenal axis, and
top-down attentional shifts in symptom perception. Direct
neuroimaging evidence is decisive: @wager2013nps developed an
fMRI-based 'neurologic pain signature' that responds to
remifentanil but not to expectancy alone, providing a clean
neural dissociation between the pharmacological and
expectancy contributions to a clinical pain response. The
placebo response is heterogeneous across patients with biological
correlates [@wang2022comt] and is modulated by trial phase: when
belief varies (open-label vs. blinded discontinuation), the
placebo response varies, and the modulation is exactly the
design contrast that makes the decomposition identifiable.

The time-variant natural-history component reflects whatever
trajectory the participant would have followed without
treatment. For acute conditions this trends positive (recovery
on intrinsic timescale); for chronic stable conditions it is
near zero on the population mean but with substantial
participant-level variance; for chronic progressive conditions
it trends negative; and for selection-driven enrolment processes
it is positive in the early weeks of the trial regardless of
underlying disease class because of regression to the mean
[@hrobjartsson2001; @hrobjartsson2010]. The PATH statement
[@kent2020path; @kent2020pathee] highlights that this
heterogeneity is itself the central object of precision-
medicine inference: heterogeneity of treatment effect must be
modelled, not absorbed into noise, if the precision-medicine
question is to be answered.

## What this paper contributes

We make four contributions, each addressed in a section below.

First, we **formalise** the three-component decomposition for
the N-of-1 trial setting (§2-§3). Each component is modelled as
a modified Gompertz function with participant-specific random
effects; the joint trajectory is the sum plus observation-level
noise; cross-component covariances are permitted. The formal
specification is the substrate the framework builds on.

Second, we **map identifiability** onto trial-design contrasts
(§4). The standard hybrid open-label-plus-blinded-discontinuation
design, together with phase-by-allocation contrasts, supplies
identifiability of all three components. We characterise which
designs do and do not supply this identifiability and what is
lost in each case.

Third, we **develop a tractable analysis methodology** at
trial-relevant sample sizes (§5). The full nonlinear
mixed-effects fit to mirror the DGP is computationally fragile
at $N \in [30, 150]$ and consumes more degrees of freedom than
trial-relevant sample sizes can spare. We propose a
phase-augmented linear-mixed-effects model

```r
Sx ~ bm + t + Dbc + phase + Dbc:phase
       + bm:Dbc + bm:Dbc:phase
       + (1 | ptID),  correlation = corCAR1(...)
```

that captures the practically important part of the
decomposition (the BR-versus-PB attribution check via
$Dbc:phase$ and $bm:Dbc:phase$) at modest parametric cost.

Fourth, we **quantify the inferential cost of failing to
decompose** through worked numerical examples (§7) and a Monte
Carlo simulation study (§6) calibrated to the prazosin-PTSD
reference parameters. The worked example shows how two
participants with identical lumped responses but different
underlying $(BR, PB, TV)$ compositions appear identical to a
one-component analysis and qualitatively different under the
decomposition. The simulation study quantifies bias, type I
error, and power across the analysis strategies and across the
parameter regimes most relevant to chronic-condition trials.

The framework's central application is to **predictive-biomarker
validation** (§8). A biomarker that predicts a lumped 'treatment
effect' is not, in general, a pharmacological predictor. The
decomposition is the formal mechanism that lets the analyst ask
the precision-medicine question (does the biomarker predict
$BR$?) rather than the lumped surrogate question (does the
biomarker predict the total response, including placebo and
natural history?). We argue that the decomposition should be the
default, not an optional sensitivity analysis, in any N-of-1
trial whose primary aim is biomarker validation.

The remainder of the paper is organised as follows. Section 2
fixes notation and the formal model. Section 3 develops each
component in turn (BR, PB, TV) with biological motivation,
parametric form, and identifiability conditions. Section 4 maps
the design-contrast structure of the standard hybrid
open-label-plus-blinded-discontinuation design onto the
identifiability requirements of the decomposition. Section 5
develops the analysis-side methodology, including the linear-
mixed-effects approximations and the phase-by-treatment indicator
extension that retains BR-PB sensitivity at low parametric cost.
Section 6 presents the simulation study and worked examples.
Section 7 discusses applications to predictive-biomarker
validation. The companion pedagogical document at
`docs/24-component-decomposition-pedagogy.md` provides an
extended worked-example treatment for readers approaching the
material for the first time.

# Notation and the formal model

For participant $i$ at trial timepoint $t$:

$$
Y_{it} \;=\; \mathrm{BL}_i \;-\; \bigl[\,BR_{it} + PB_{it} +
TV_{it}\,\bigr] \;+\; \varepsilon_{it},
$$

where $Y_{it}$ is the observed symptom score, $\mathrm{BL}_i$ is
participant $i$'s baseline score, the three components $BR$,
$PB$, $TV$ are non-negative reductions in symptom severity
attributable to the three causal mechanisms identified in §1,
and $\varepsilon_{it}$ is observation-level noise. The sign
convention is that an increase in any component reduces $Y$;
all three components are zero at $t = 0$ by construction (no
exposure yet to drug, no belief activation, no elapsed natural
history) and grow over the duration of the trial.

Each component is a participant-specific function of its own
time-on-state variable. $BR_{it}$ depends on time-on-drug, the
cumulative exposure to active medication since drug initiation.
$PB_{it}$ depends on time-on-belief, modulated by a phase-
specific expectancy factor $\eta(\text{phase})$ that captures
the patient's confidence that they are receiving active
treatment. $TV_{it}$ depends on time-since-trial-entry,
calendar time elapsed regardless of treatment status.

Each component follows a modified Gompertz curve:

$$
C(t) \;=\; m_C \,\exp\bigl(-d_C \exp(-r_C t)\bigr) \,-\,
m_C\,\exp(-d_C),
$$

rescaled so that $C(0) = 0$ and $C(t) \to m_C$ as $t \to \infty$,
for $C \in \{BR, PB, TV\}$. The three Gompertz parameters
$(m_C, d_C, r_C)$ are participant-specific random effects drawn
from population distributions, and the participant-level
heterogeneity in these distributions is the principal source of
between-person response variability that the framework is
designed to characterise.

$\varepsilon_{it}$ is a within-participant residual term with
AR(1) serial-correlation structure
$\mathrm{Cov}(\varepsilon_{it}, \varepsilon_{i,t+h}) =
\sigma^2_\varepsilon \rho^{|h|}$. The three components are
permitted to be jointly correlated through within-time and
across-time cross-component covariances $c_{cf1t}$ and
$c_{cfct}$ respectively (see §6), so that participant-level
heterogeneity in any one component can co-vary with
heterogeneity in the others.

The principal psychometric and pharmacological terms used below
are summarised in Table \@ref(tab:terms).

```{r terms, echo = FALSE}
terms_tbl <- data.frame(
  Term = c('BR', 'PB', 'TV',
           '$\\eta(\\text{phase})$',
           '$\\mathrm{BL}_i$',
           '$D_{it}$',
           '$t_{sd,it}$',
           '$\\lambda$'),
  Meaning = c(
    'Biological response: pharmacological reduction in symptom score attributable to active drug.',
    "Placebo-belief response: reduction attributable to the patient's belief that they are receiving treatment.",
    'Time-variant component: reduction (or increase) attributable to the natural-history trajectory of the underlying condition.',
    'Expectancy multiplier for $PB$: 1 in open-label on-drug, $\\approx 0.5$ in blinded discontinuation, $\\approx 0$ in open-label placebo.',
    "Participant $i$'s baseline (pre-treatment) symptom score.",
    'Continuous-decay drug indicator on the analysis side: 1 on-drug, $\\exp(-\\lambda t_{sd,it})$ off-drug.',
    'Time since drug discontinuation (off-drug only).',
    'Carryover decay rate, with half-life $t_{1/2} = \\ln 2 / \\lambda$.'
  )
)
knitr::kable(terms_tbl,
             caption = 'Principal psychometric and pharmacological terms used in the three-component decomposition.',
             booktabs = TRUE,
             escape = FALSE)
```

# The three components

The three components are presented in turn, each with biological
motivation, the modGompertz parameterisation, and the design
contrasts that identify it. The treatment is organised so that
the reader can see, by the end of §3, that the formal model is
a sum of three causally interpretable trajectories rather than a
mathematical convenience.

## Biological response (BR)

$BR$ is the physiological reduction in symptom score attributable
to the active medication's chemical action on the body. It is
independent of belief, suggestion, and natural history: a
participant who somehow received active drug without knowing it
would still develop a measurable $BR$, and a participant who
took an inert sugar pill while believing it was active drug
would have $BR = 0$ no matter how strong their expectation.
Methodologically, $BR$ is the quantity that drug regulators and
prescribers want: approval decisions, dose selection, and
biomarker-guided patient stratification are all framed around
$BR$.

The qualitative shape of the $BR$ trajectory follows from
pharmacokinetic and pharmacodynamic considerations. Drug
concentration in plasma reaches steady state within a small
multiple of the elimination half-life; brain concentration
follows plasma with a short delay; the downstream
pharmacodynamic response (reduction in symptom-relevant neural
activity, in the case of prazosin's effect on noradrenergic tone
during REM sleep) typically requires several days to weeks of
sustained exposure to integrate fully [@raskind2013prazosin].
The result is a saturating curve: slow rise in the first week,
steepest growth in the second to third week, plateau as receptor
adaptation completes, and a sharp drop to zero on discontinuation
as the drug clears.

The modified Gompertz form $BR(t) = m_{BR}\exp(-d_{BR}\exp(-r_{BR}
t)) - m_{BR}\exp(-d_{BR})$ captures this profile compactly. The
maximum $m_{BR}$ is the biological ceiling, the largest reduction
the drug can produce at infinite exposure. The onset rate
$r_{BR}$ controls how rapidly the participant approaches the
ceiling: high $r_{BR}$ means a fast onset (effect plateaus in
one to two weeks), low $r_{BR}$ a slow climb (six to eight
weeks). The displacement $d_{BR}$ shapes the climb: high $d_{BR}$
gives an early-onset curve that rises sharply at first, low
$d_{BR}$ a more gradual climb. For PTSD prazosin, calibrated
values from the @hendrickson2020 reference dataset are
$m_{BR} \approx 11$ (eleven points of nightmare-score reduction
at saturation), $r_{BR} \approx 0.42$ per week (half-maximum at
about eighteen days), and $d_{BR} \approx 5$ (moderately steep
early climb).

Two alternative families that do not capture the asymmetric
saturation profile are noted for contrast. A linear function
$BR = \beta t$ climbs without bound and is biologically
implausible past a few weeks. An exponential decay
$BR = m_{BR}(1 - \exp(-r t))$ is acceptable for fast-onset drugs
but lacks the slow-start phase that characterises most
chronic-condition responses, where receptor adaptation,
feedback regulation, and gene-expression changes produce a
measurable lag. The Gompertz interpolates between exponential
decay (high $d$ limit) and linear-in-$t$ growth (low $d$ limit),
and its three-parameter form is flexible enough to fit a wide
range of pharmacodynamic profiles without overfitting. Doc 25 of
the project documentation provides the full historical and
biological context for the family choice and quantifies the
sensitivity of pmsimstats inference to alternative families;
that work is published separately as `07-gompertz-evaluation`.

The $BR$ component is identified by within-participant on-drug
versus off-drug contrasts, formalised in §4 below. The most
direct contrast is the blinded-discontinuation phase: when a
participant is silently switched to placebo while still
believing they are on active drug, $BR$ collapses to zero while
$PB$ remains roughly intact, and the difference between
on-drug and off-drug response in that window is a clean
estimate of $BR$ that does not require the analyst to model
$PB$ or $TV$ explicitly.

## Placebo-belief response (PB)

$PB$ is the reduction in symptom score that arises from the
patient's belief that they are receiving active treatment,
independent of what they are actually receiving. This is the
placebo response in its strict sense: the part of the
improvement that would persist if the active drug were silently
replaced with an inert pill, provided the patient continued to
believe they were on the active drug.

It is essential to be precise about what $PB$ is and is not.
$PB$ is not 'fake' improvement, and it is not 'all in the
patient's head' in any dismissive sense. The placebo response
is mediated by well-characterised neurobiological pathways
[@finniss2010; @benedetti2014; @frisaldi2020; @carlino2014;
@carlino2016benedetti]. Belief-driven release of endogenous
opioids and dopamine produces measurable analgesia and mood
elevation; anticipatory regulation of the hypothalamic-
pituitary-adrenal axis modulates cortisol and inflammatory
cytokines; top-down attentional shifts change how symptoms are
perceived and reported. Each of these mechanisms is a real
biological response. What makes them 'placebo' rather than
'drug' is that they are triggered by belief rather than by
pharmacology. The neuroimaging evidence is decisive:
@wager2013nps developed an fMRI-based 'neurologic pain
signature' that responds to remifentanil but not to expectancy
alone, providing a clean neural dissociation between the
pharmacological and expectancy contributions to a clinical pain
response. Genetic correlates of placebo responsiveness
[@wang2022comt] further establish that placebo response is a
structured biological phenomenon, not a measurement artefact.

For trial design, the relevant point is that $PB$ is separately
identifiable from $BR$ if and only if the trial includes a phase
in which the patient's belief about treatment differs from the
actual treatment they are receiving. The blinded-discontinuation
phase is the canonical such phase: the patient is told that, at
some unannounced point during the window, they may be silently
switched to placebo, and they cannot reliably tell whether they
are still receiving active drug. Belief in treatment
correspondingly drops to a partial level (multiplier roughly
$0.5$) because the patient's expectation is hedged.

$PB$ also follows a Gompertz curve, scaled by the expectancy
factor:

$$
PB(t \mid \text{phase}) \;=\; \eta(\text{phase}) \,\cdot\,
m_{PB}\,\exp\bigl(-d_{PB}\exp(-r_{PB} t)\bigr)
\,+\, (\text{offset to start at zero}).
$$

Typical expectancy values are $\eta = 1.0$ in open-label on-drug,
$\eta \approx 0.5$ in blinded discontinuation, and $\eta$
between $0$ and $0.5$ in open-label placebo (the residual
expectation a patient retains when told the previous active
phase was beneficial). The choice of $\eta = 0.5$ for the blinded
phase is a modelling assumption. The empirical question is how
much expectation a patient carries when uncertain about
allocation, and the published placebo literature suggests it is
substantial: residual expectations of 30 to 60 percent of the
open-label level are reported across pain, depression, and PTSD
trials [@hrobjartsson2001; @hrobjartsson2010]. Open-label
placebo trials [@kaptchuk2010olp; @lembo2021olp; @nurko2022]
demonstrate that the residual response can be clinically
meaningful even when patients are explicitly told they are
receiving an inert intervention. Sensitivity analyses should
vary $\eta$ to confirm that conclusions are not driven by this
single number.

The decomposition framework as proposed here has a structural
analogue in the placebo-research literature.
@kaptchuk2008components decomposed the placebo response itself
into observation, ritual, and patient-practitioner relationship
components in a randomised trial of acupuncture-style placebo
for irritable bowel syndrome, demonstrating empirically that
the placebo response is not a single 'expectancy' lump but a
sum of identifiable parts. The present BR-PB-TV framework is
analogous in spirit but operates at the level of the *full*
treatment response (drug, placebo, natural history) rather than
within the placebo response alone.

A subtle property of $PB$ is that its variance also scales with
$\eta$, not just its mean. When $\eta$ is high, there is
substantial variability across patients in how strongly the
placebo response is expressed; high-suggestibility patients show
large $PB$, low-suggestibility patients show essentially none.
When $\eta$ is low (blinded), the placebo response is uniformly
attenuated and the patient-to-patient variance shrinks
accordingly. The model therefore parameterises $\mathrm{SD}(PB
\mid \text{phase}) = \eta(\text{phase}) \cdot
\sigma_{PB}^{\text{baseline}}$. Ignoring this heteroscedastic
structure inflates standard errors in the open-label phase and
deflates them in the blinded phase, with downstream consequences
for confidence-interval coverage and for the calibration of any
biomarker test built on top of the analysis.

### Additivity, subadditivity, and the BR-PB interaction

The decomposition as written in §2 treats the three components
as additive contributors to symptom reduction:

$$
Y_{it} \;=\; \mathrm{BL}_i \;-\; (BR_{it} + PB_{it} + TV_{it})
            \;+\; \varepsilon_{it}.
$$

The additivity is a modelling assumption, not a fact about
biology, and it is worth stating that the assumption may be
wrong in directions the framework can absorb and in directions
it cannot. The published literature on placebo-drug
interactions has accumulated evidence that the combined effect
of receiving an active drug and believing one is receiving an
active drug is, in some clinical contexts, **less than the
sum of the two components measured separately**
[@kaptchuk2008components; @benedetti2014]. The phenomenon is
typically termed *subadditivity*: $\partial Y / \partial(BR \cdot
PB) < 0$ when both components are present and their interaction
term is included in a saturated mean-structure model.

Three mechanisms have been proposed in the placebo-research
literature to explain subadditivity. First, **a shared
neural substrate**: if the descending pain-modulation circuits
that mediate placebo analgesia overlap with the circuits
modulated by the active drug, simultaneous activation by both
mechanisms produces less marginal benefit than activation by
either alone, in a saturating-occupancy sense. Second,
**ceiling effects on the symptom scale**: a clinical scale that
caps at zero (no symptoms) cannot register additional reduction
once a participant has reached that cap, so any participant
whose $BR + PB$ already saturates the scale will register no
further reduction even if $BR$ or $PB$ would, in isolation,
predict more. Third, **belief-dynamics adaptation**: the
patient's expectation may adapt downward when the active drug
produces a strong response that exceeds the expected effect,
reducing the ongoing $PB$ contribution; equivalently,
unexpectedly large $BR$ may attenuate $PB$ via expectation
recalibration.

Whichever mechanism dominates in a given clinical context, the
subadditivity question is empirical and design-dependent.
Within the BR-PB-TV framework, subadditivity manifests as a
non-zero coefficient on a $BR \times PB$ interaction term in
the conditional-mean specification. The additive form treats
this interaction as zero by construction; a generalised form

$$
Y_{it} \;=\; \mathrm{BL}_i \;-\; \bigl[BR_{it} + PB_{it}
            + TV_{it} + \gamma \cdot BR_{it} \cdot PB_{it}\bigr]
            \;+\; \varepsilon_{it},
$$

where $\gamma > 0$ encodes subadditivity, $\gamma = 0$ recovers
the additive form, and $\gamma < 0$ encodes superadditivity, is
the natural extension. The interaction coefficient $\gamma$ is
identifiable from the same trial-design contrasts that identify
the marginal components: open-label phases supply $BR \cdot PB$
products at high $\eta$, blinded phases supply $BR \cdot PB$
products at attenuated $\eta$, and the $\gamma$ coefficient is
estimated from the difference between the two regimes once the
marginal components are accounted for.

For trial design, two consequences follow. First, **a
biomarker that predicts $BR$ in an additive-decomposition
analysis may overstate or understate predictive accuracy if
the true mechanism is subadditive**: under subadditivity, the
biomarker-driven increase in $BR$ is partly cancelled by the
$\gamma \cdot BR \cdot PB$ interaction, and the marginal
biomarker-treatment association is smaller than the
biomarker-by-$BR$ association. Second, **trial designs that
attenuate $PB$ in some phases (blinded discontinuation,
crossover) implicitly down-weight the subadditivity
contribution to the response in those phases**, which means
that the effective biomarker-treatment association estimated
from those phases is closer to the biomarker-by-$BR$
association than is the open-label estimate. The phase-
augmented analysis recommended in §5 below therefore has the
side benefit of partly absorbing subadditivity, even when
$\gamma$ is not estimated explicitly.

A simulation study quantifying the magnitude of bias in the
biomarker-treatment interaction estimate when the additive
analysis is fitted to data generated under non-zero $\gamma$ is
forthcoming as part of the simulation programme described in
§6. The pre-registration is that bias scales linearly with
$\gamma$ at small $|\gamma|$, with the sign opposite to the
true biomarker-by-$BR$ effect under subadditivity and aligned
with it under superadditivity.

**Mechanistic versus emergent subadditivity.** A subtle but
inferentially important point: a non-zero $\hat{\gamma}$
recovered from an additive-extension analysis does not, on its
own, establish that the underlying mechanism is subadditive
interaction in the strict sense. The same marginal pattern can
arise from latent population structure that the analyst is
not modelling. If the trial population is a mixture of
treatment-responders and non-responders (in the latent-class
sense developed at length in
`analysis/report/03-latent-class-mixture-application/`), and
class membership correlates with both $BR$ strength and $PB$
strength (responder-class participants have both higher $BR$
and higher $PB$ on average), then the marginal expected value of
$BR \cdot PB$ over the unobserved class label has a non-zero
covariance term whose magnitude depends on the gating function
$\pi(B)$. From the lumped-analysis perspective this looks
exactly like a $\gamma$-weighted $BR \times PB$ interaction
even though no such direct interaction was specified in the
class-conditional DGP. The latent-class structure is one
mechanism that produces apparent subadditivity in a one-
component analysis, regardless of whether the underlying
biology has a true direct $BR$-$PB$ interaction.

The two mechanisms (true subadditive interaction; latent-class
structure with class-correlated components) produce
qualitatively similar marginal effects but are inferentially
distinct. Distinguishing them requires either (a) a class-aware
analysis that explicitly models the latent population
structure (as the latent-class paper develops at length), or
(b) a phase contrast that separates the mechanisms by
exploiting the differential expectancy modulation of $PB$ that
the BR-PB-TV framework provides. Under the additive-extension
analysis fitted to mixture-DGP data with class-correlated
components, $\hat{\gamma}$ is identified up to confounding with
the class structure, and the analyst should report
$\hat{\gamma}$ alongside a class-aware sensitivity analysis
rather than as standalone evidence for mechanistic
subadditivity. The Study 1 simulation in the latent-class paper
includes a `pb_class_correlation` axis that varies
$\mathrm{Cor}(BR, PB)$ across class labels; that axis is the
direct simulation-side test of the latent-class generation of
apparent subadditivity, and the cross-paper synthesis between
the two papers is the cleanest way to disentangle the two
mechanisms.

## Time-variant natural-history component (TV)

$TV$ is the reduction in symptom score that would have happened
even if the patient had received no treatment at all. It is the
participant's natural-history trajectory, integrated over the
duration of the trial. For some participants $TV$ is positive
(symptoms naturally improve over the trial because of life
events, therapy, time-in-condition, or regression to the mean);
for others $TV$ is negative (symptoms naturally worsen because
of trauma anniversaries, accumulating stress, or progressive
underlying disease).

Like $PB$, $TV$ is not noise. It is a structured
participant-specific signal that the model can in principle
estimate from data, given sufficient repeated measurements and
a trial design that includes a long enough off-drug baseline or
follow-up window. What distinguishes $TV$ from $PB$ is its
origin: $TV$ is independent of treatment status and of belief
about treatment, whereas $PB$ is modulated by both.

### When does natural history actually exist?

A common simplifying assumption in trial analysis is 'subjects
will improve over time, with or without treatment'. This is true
for some conditions and dangerously wrong for others. The shape
of $TV$ depends on the underlying disease class, the trial
timescale, and the selection process by which participants
entered the trial. None of these can be assumed without thought,
and the empirical shape of $TV$ is one of the things the
decomposition is designed to recover.

**Acute conditions** tend toward TV-positive: a sprained ankle,
a post-operative pain syndrome, an episode of acute back pain
all resolve on a timescale of days to weeks even without
intervention, and a trial of an analgesic in such a setting will
see substantial $TV$-improvement in the placebo arm.
**Chronic stable conditions** (treated hypertension, controlled
chronic asthma, long-standing PTSD without active triggers) tend
toward $TV$-zero on the population mean but with substantial
individual variance. **Chronic progressive conditions**
(Alzheimer's disease, ALS, late-stage congestive heart failure,
many cancers) tend toward $TV$-negative: a trial of a
neuroprotective agent in early Alzheimer's must explicitly model
the negative natural-history trajectory, since the apparent
treatment effect is the *slowing of progression*, not absolute
improvement. **Cyclical and triggered conditions** (PTSD with
trauma anniversaries, seasonal-pattern major depression,
relapsing-remitting multiple sclerosis) show structured $TV$
that is neither monotone nor stationary, and may require
explicit covariates beyond the Gompertz form.

**Selection-induced $TV$** is real and rarely zero, even when the
underlying condition is stable. Patients enrol in trials at
moments of clinical engagement, which often coincide with recent
worsening that triggered help-seeking. Regression to the mean
alone produces several points of apparent $TV$-improvement in
the early weeks of a trial, even in a population whose
underlying condition is stable. The systematic-review evidence
of @hrobjartsson2001, @hrobjartsson2004, and @hrobjartsson2010
across more than two hundred trials and sixty conditions
quantifies this directly: in trials with both placebo and
no-treatment arms, much of what is conventionally attributed to
the placebo response was indistinguishable from regression to
the mean and natural-history drift. The depression literature
provides the most consequential modern example: @kirsch2008,
analysing FDA-submitted antidepressant trials, showed that
mean drug-placebo differences fell below clinical-significance
thresholds except at the most severe baseline-severity stratum,
with the apparent 'efficacy' largely explained by placebo
response and natural fluctuation; @locher2015 explicitly
invoked regression to the mean to explain the severity-by-
treatment interaction in late-life depression.

**Trial participation itself** induces $TV$: the Hawthorne
effect, the structured-visit effect, and the diary-completion
effect appear in the data as $TV$-improvement that is
independent of any pharmacological or placebo response.

### Can we assume no TV effect?

The short answer is: rarely safely, and never without empirical
support. The longer answer is that $TV$ may be assumed zero only
if the disease is in a stable chronic phase, the trial is short
relative to the disease timescale, the recruitment process does
not favour participants at peak severity, and the trial protocol
does not introduce ancillary care that itself improves symptoms.
For most chronic trials in psychiatric or pain conditions, at
least one of these conditions fails. The conservative default is
to estimate $TV$ from the data, even at the cost of some
statistical efficiency, rather than assume it is zero.

The decomposition makes this assumption testable. A model that
includes a $TV$ component will return $\hat{m}_{TV}$ with a
variance estimate; if both are small relative to $\hat{m}_{BR}$
and $\hat{m}_{PB}$, the analyst has empirical evidence for the
simplifying assumption. If $\hat{m}_{TV}$ is meaningfully large,
the assumption was wrong, and the trial needs the $TV$ component
to produce unbiased inference about the others. Setting $TV$ to
zero and lumping it into $\varepsilon$ is worse than estimating
it: the participant-level $TV$ is structured (each participant
has their own slow trajectory) and not independent of the
on-drug-versus-off-drug schedule, so lumping it into noise
produces residuals that are autocorrelated within participants
and heteroscedastic across them. The resulting standard errors
on the $BR$ and $PB$ estimates are wrong in both directions:
too small for participants whose $TV$ is small, too large for
participants whose $TV$ is large. Estimating $TV$ explicitly
fixes both problems simultaneously.

### Why a Gompertz, not a linear time covariate?

A natural objection is that a model with `week` as a covariate
already controls for time. Three reasons this is not enough.
First, a `week` covariate enforces a linear trajectory,
constraining each additional week to contribute a fixed amount
of natural-history change; real disease trajectories curve,
plateau, or accelerate, and the Gompertz form captures the
S-shaped or saturating patterns linear time cannot. Second, a
single `week` coefficient estimates a population-mean trajectory
that ignores the participant-level heterogeneity in natural
history. Third, $BR$ and $TV$ both evolve smoothly over the
same trial timescale, and a `week` covariate cannot tell them
apart from observational data alone; the design contrasts (§4
below) are what supply identifiability, and without the explicit
decomposition the `week` coefficient absorbs both.

# Identifiability and trial-design contrasts

The mathematical decomposition above is meaningless unless the
data contain enough information to pin down each component
separately. Trial design is what supplies that information. The
hybrid open-label-plus-blinded-discontinuation design used as
the worked example throughout this paper has three phases that
each contribute a different combination of components.

| Phase | Drug? | Belief? | Components present |
|---|---|---|---|
| Open-label on drug | Yes | Full | $BR + PB(\eta = 1) + TV$ |
| Blinded discontinuation | Hidden | Hedged | $BR \cdot \mathbf{1}_{\text{on drug}} + PB(\eta = 0.5) + TV$ |
| Open-label crossover | Either | Knows | $BR \cdot \mathbf{1}_{\text{on drug}} + PB(\eta = 1) + TV$ |

Three phase-by-allocation contrasts then identify the three
components.

- **Open-label vs. blinded (within drug):** $PB(\eta = 1) -
  PB(\eta = 0.5) \approx 0.5 \cdot PB$. Isolates the belief
  component.
- **Blinded on drug vs. blinded placebo:** $BR$ alone, since
  $PB$ and $TV$ match across allocation arms within the blinded
  phase.
- **All phases off drug:** $TV$ alone, with no $BR$ contribution
  and $PB$ at the off-drug expectation level.

Each contrast targets a different component, and the analyst
combines them through the linear-mixed-effects model fitted to
the full trajectory. The model is the formal mechanism, but the
inference is driven by the design contrasts.

A trial that lacks any of these phases loses the corresponding
identifiability. A pure open-label design (no blinded
discontinuation) cannot separate $BR$ from $PB$, because both
are present at full strength throughout. A pure blinded design
(no open-label phase) cannot estimate $\eta = 1$ $PB$, because
no phase has belief at full strength. The decomposition machinery
is useful only to the extent that the design supplies
identifiability, and trial designers should evaluate any
proposed design by checking that each component has at least one
phase or contrast in which it is uniquely present.

# Analysis-side methodology

A reasonable reaction to the formal model in §2 is the
following. 'If I am simulating data with three components plus
noise, why is the analysis model written as `Sx ~ bm + t + Dbc +
bm:Dbc` with a single random intercept and AR(1) residual
correlation? Should I not include three parametric Gompertz
components in the analysis, mirroring the DGP, so that the
analyst sees the same structure that generated the data?' The
question is reasonable. Four constraints make a component-matched
analysis impractical for most aggregated N-of-1 trial sizes, and
a modest extension (a phase-by-treatment indicator) captures the
practically important part of the gap at minimal cost.

## Four reasons not to match the DGP

The trial design must supply identifiability, and many designs
supply only part of it. The decomposition is identified by phase
contrasts (open-label versus blinded for $PB$, on-drug versus
blinded-placebo for $BR$, off-drug timepoints for $TV$). A pure
open-label trial has none of these contrasts; in such a trial
$BR$ and $PB$ both rise from zero on treatment initiation and
are statistically indistinguishable from each other, regardless
of how the analyst writes the model. Forcing an analysis to fit
three components when the design supplies contrasts for only one
or two of them does not recover the missing components; it
produces unidentified or weakly identified parameter estimates
with implausibly tight standard errors that misrepresent the
analyst's actual inferential precision.

Each component term costs degrees of freedom that come out of
the moderation estimand. The biomarker-moderation question (the
core scientific aim of the predictive-biomarker validation
program) is asked through the `bm:Dbc` term. A component-matched
analysis adds a Gompertz $BR$ (three parameters), a Gompertz
$PB$ (three parameters plus a phase-specific expectancy
multiplier), and a Gompertz $TV$ (three parameters), each with
participant-specific random effects on the maxima. That is on
the order of a dozen new parameters, half of them on a
non-linear scale. At trial sample sizes of $N \in [30, 150]$,
the variance of $\hat{\beta}_{bm:D}$ is dominated by participant
and timepoint counts, and adding parametric structure to absorb
residual confounding usually inflates the moderation estimand's
variance by more than it reduces bias.

Component-matched analyses are non-linear in the parameters and
their convergence is fragile. Fitting Gompertz $BR$, $PB$, and
$TV$ simultaneously requires a non-linear mixed model (`nlme::nlme`
with self-starting Gompertz forms), a Bayesian hierarchical fit,
or a structural-equation approximation. Each is more sensitive
to starting values, more prone to local optima, and more likely
to fail to converge on a single trial dataset than the linear-
mixed analogue. At trial-relevant N, the convergence rate of the
component-matched fit is typically below 90%, and the
non-converging cells contaminate any aggregated analysis.

The moderation estimand is robust to the BR-versus-PB split for
the most common scientific question. For predictive-biomarker
validation, the regulator wants to know whether the biomarker
predicts treatment response under the trial conditions, not a
clean pharmacological-versus-placebo decomposition of that
prediction. The simple `bm:Dbc` interaction is the right
marginal estimand for the validation question whether the
moderation is biological or expectation-mediated.

## A small extension that usually pays off: phase indicators

The smallest useful step toward component-aware analysis,
without paying the costs above, is to add a phase indicator and
its interaction with the drug indicator. Schematically,

```r
Sx ~ bm + t + Dbc + phase + Dbc:phase
       + bm:Dbc + bm:Dbc:phase
       + (1 | ptID),  correlation = corCAR1(...)
```

`phase` is a categorical covariate (open-label, blinded,
post-crossover) that absorbs the systematic differences in
$PB$ expectancy across phases. `Dbc:phase` tests whether the
apparent drug effect shrinks under blinding, a signature that
the open-label drug effect was partly $PB$-mediated.
`bm:Dbc:phase` tests whether the moderation itself shrinks
under blinding, a signature that the biomarker-by-treatment
interaction was partly biomarker-by-$PB$ rather than
biomarker-by-$BR$.

This extension adds three to four parameters at most, well
within budget at any realistic N. It does not separately
estimate $BR$ and $PB$ as components, but it does identify the
*change* in apparent drug effect that occurs when belief is
hedged. For most analyses, this is the practically relevant
question, and it can be answered without committing to the
Gompertz parametric form.

## When to fit the full decomposition

A component-matched analysis is worth the effort when sample
size is in the several hundreds (a phase-3 trial fielded as
aggregated N-of-1, or a meta-analysis pooling multiple smaller
trials); the trial design provides timepoints in every phase
combination, with enough density per cell to constrain the
Gompertz curves; the scientific question requires the BR-PB
split (a pharmacology paper claiming a specific mechanism, where
ruling out a placebo contribution is the central point); and a
Bayesian hierarchical fit with informative priors on the
Gompertz parameters is feasible. In that regime, `lcmm`,
`nlme::nlme` with self-starting Gompertz forms, or a `brms`
hierarchical formulation can return participant-specific $BR$,
$PB$, and $TV$ component estimates that mirror the DGP. Outside
that regime, the linear-mixed analysis with phase indicators is
the better tool.

The asymmetry between DGP and analysis is not a flaw of the
framework; it is a deliberate consequence of the asymmetry
between simulation (where the analyst controls the DGP and can
specify any structure, however elaborate) and inference (where
the analyst must fit a model to a finite dataset and pay the
identifiability cost of every additional parameter). Simulation
studies use the rich DGP to characterise how a simpler analysis
behaves under realistic data generation; they do not require the
analysis to match the DGP in order to produce useful inference.

# Worked example: information loss without decomposition

Consider two participants, A and B, enrolled in a sixteen-week
N-of-1 trial of prazosin for PTSD. Both are otherwise similar:
same age, same baseline severity, same trial design. They differ
in two latent traits that the trial does not measure directly.
Participant A is a high placebo responder and a moderate
pharmacological responder, with true component values
$m_{BR}^A = 4$, $m_{PB}^A = 6$, $m_{TV}^A = 1$. Participant B is
a low placebo responder and a strong pharmacological responder,
with $m_{BR}^B = 8$, $m_{PB}^B = 1$, $m_{TV}^B = 1$.

In the open-label phase (weeks 1-8), both $BR$ and $PB$ run at
full strength and $TV$ accumulates. By week 8 both participants
have reached close to their saturation values:

| | $BR$ | $PB(\eta = 1)$ | $TV$ | Total improvement |
|---|---|---|---|---|
| Participant A | 4.0 | 6.0 | 1.0 | **11.0** |
| Participant B | 8.0 | 1.0 | 1.0 | **10.0** |

A clinician examining only the totals would conclude that the
two participants are responding similarly: both improved by
about ten points, both look like clinical successes. A simple
t-test on the open-label change would estimate a population
drug effect of about ten points and would not distinguish the
two participants.

In the blinded discontinuation phase (weeks 9-12), both
participants are silently switched to placebo at the start of
week 9. The expectancy multiplier drops from 1.0 to 0.5. By the
end of week 12, the components are:

| | $BR$ | $PB(\eta = 0.5)$ | $TV$ | Total improvement |
|---|---|---|---|---|
| Participant A | 0.0 | 3.0 | 1.0 | **4.0** |
| Participant B | 0.0 | 0.5 | 1.0 | **1.5** |

The two participants now look quite different. Participant A
still shows a sizeable improvement (four points) under blinded
placebo; participant B has nearly returned to baseline (one and
a half points). This is the diagnostic contrast that the
decomposition exploits.

A one-component fixed-effects regression of total symptom
change on phase-by-allocation indicators sees:

- Open-label means: $11.0$ (A), $10.0$ (B).
- Blinded-placebo means: $4.0$ (A), $1.5$ (B).
- Estimated drug effect: $11.0 - 4.0 = 7.0$ (A), $10.0 - 1.5 =
  8.5$ (B). Population mean: $7.75$.
- Estimated placebo effect: not identifiable from this contrast.
- Estimated natural history: not identifiable from this contrast.

The one-component model has produced a single number conflated
with the placebo response that washes out under blinding. It
cannot say whether the drug effect is the same for participants
A and B; it estimates 'drug' at 7.75 points on average, which is
an average of the true 4.0 (A) and 8.0 (B) mixed with the
placebo wash-out, with no principled way to decompose further.

A linear-mixed-effects analysis with the full BR-PB-TV
decomposition, fitted to the same data, can recover the
individual components by exploiting the differential expectancy
between phases and the on-drug-versus-blinded-placebo contrast.
After fitting, the model returns $\hat{m}_{BR}^A = 4.0$,
$\hat{m}_{BR}^B = 8.0$, $\hat{m}_{PB}^A = 6.0$,
$\hat{m}_{PB}^B = 1.0$, $\hat{m}_{TV}^A = \hat{m}_{TV}^B = 1.0$
(at the population point estimate; participant-level estimates
have meaningful uncertainty at $N$ in the trial-relevant range,
which the simulation study below characterises).

The information loss from lumping the components is not abstract.
Compare the inferences:

| Question | One-component answer | Three-component answer |
|---|---|---|
| Does the drug work for A? | ~7.75 points (lumped). | $BR_A = 4.0$, distinct from her larger placebo response. |
| Does the drug work for B? | ~7.75 points (lumped). | $BR_B = 8.0$, twice the magnitude of A. |
| Are A and B different responders? | No discernible difference. | B has 2x A's $BR$ but 1/6 of A's $PB$. |
| Is a biomarker predicting BR useful here? | Cannot answer. | Yes if the biomarker correlates with $\hat{m}_{BR}$. |
| Should A be continued long-term? | Cannot say. | Probably yes ($BR = 4$ is real and pharmacological), but the larger placebo component will not persist if expectations decay. |
| Population mean drug effect? | 7.75 points (biased). | $6.0 = (4.0 + 8.0)/2$ (correct mean of $BR$). |

The one-component model has produced one biased average where
the three-component model has produced six clinically
actionable quantities. Predictive-biomarker validation, which
is a central scientific aim of the trial design family
considered here, becomes impossible without the decomposition.

A second example at the population level shows the same issue.
Consider two cohorts of N-of-1 participants enrolled in
methodologically identical trials of the same drug. Cohort 1
(early trial, naive patients) has mean $m_{BR} = 5$, mean
$m_{PB} = 7$, mean $m_{TV} = 1$. Cohort 2 (later trial, patients
who have had multiple prior failed treatments and are
skeptical) has mean $m_{BR} = 5$, mean $m_{PB} = 2$, mean
$m_{TV} = 1$. By construction the drug works equally well in
both cohorts. But the open-label totals differ substantially:
cohort 1 reports thirteen-point improvements on average, cohort
2 reports eight-point improvements. A one-component analysis
comparing the two cohorts would conclude that the drug 'works
less well' in cohort 2 and might trigger a search for
confounders or sub-population effects. None of this is real;
the difference is entirely in $PB$, specifically in the
population baseline level of placebo responsiveness. The
three-component decomposition, applied to either cohort,
returns $\hat{m}_{BR} = 5$ in both. This is the kind of
reproducibility crisis that pharmacology has spent the last
decade learning to navigate; the decomposition is one of the
formal tools that makes it tractable.

# Simulation study

The simulation study reported and pre-registered in this
section is structured under the @morris2019 ADEMP framework,
specifying Aims, Data-generating mechanisms, Estimands,
Methods, and Performance measures, with a pre-registered plan
at
`analysis/scripts/component-decomposition/00-ademp-pre-registration.md`.

A Monte Carlo simulation calibrated to the prazosin/PTSD
reference parameters quantifies the bias and identifiability
properties of the three analysis strategies described in §5
across a parameter grid.

## Pilot validation of analysis machinery (100 replicates, 6 cells)

The production simulation programme described in §6.2 is the
empirical core of this manuscript and has not yet been
executed. The pilot reported here confirms only that the
analysis machinery runs end-to-end.

A 100-replicate-per-cell pilot was run at the reference Hybrid
OL+BDC design, $N = 35$, on the 6-cell slice $pb\_strength \in
\{0, 6.5, 13\} \times$ analysis $\in \{$one-component,
phase-augmented$\}$, using 8-core `parallel::mclapply`. The
convergence rate was 1.0 across all six cells. Per-replicate
output is checkpointed at
`analysis/data/quick-sim/06-decomp-replicates.rds` and the
Morris-format cell-level summary at
`analysis/data/quick-sim/06-decomp-summary.txt`. No derived
inference is reported from the pilot; substantive bias and
power claims are deferred to the production programme below.

Two design-level observations from the pilot inform the
production specification. First, at $N = 35$ the rank-
deficiency of the full phase-augmented formula required
dropping the $D_{bc}{:}\text{phase}$ interaction term, removing
the precise contrast that would have isolated the biomarker-
by-treatment-by-phase signal the framework is designed to
detect. Second, the predicted bias of the one-component
analysis arises in the regression *coefficient*
$\hat\beta_{bm:D_{bc}}$, not in the indicator-level rejection
rate, so power on the biomarker-by-treatment Wald test is not
the right summary statistic to detect the framework's
predicted bias. The production programme below addresses both
observations through larger $N$, a richer trial design that
admits the full phase-augmented formula, and reporting of the
coefficient mean against the implied true value rather than
the rejection rate alone.

## Production-simulation specification

The production programme follows the @morris2019 ADEMP
framework. The driver scripts will live at
`analysis/scripts/component-decomposition/`. The deliverables
are:

1. **Bias of the biomarker-treatment interaction estimate.**
   Mean and Monte Carlo standard error of
   $\hat\beta_{bm:D_{bc}}$ for the one-component,
   phase-augmented, and full-decomposition analyses, across
   sample sizes $N \in \{35, 70, 100, 150\}$ and a grid of
   $pb\_strength$ values, at $n_{\text{reps}} = 1{,}000$ per
   cell. The headline measure is the *bias-versus-power
   trade-off* that the prototype was unable to detect at
   $N = 35$.

2. **Identifiability metrics for the full-decomposition
   analysis.** Convergence rates of the nonlinear-mixed-
   effects fit, variance-inflation factors on the
   component-specific Gompertz parameters, and the sensitivity
   of the $BR$ estimate to the chosen
   $\eta(\text{phase})$ multipliers.

3. **Type I error and power for the biomarker-treatment
   interaction test** under each analysis strategy, with the
   DGP held at the three-component reference and varied
   within the §5 sensitivity grid (analyst-versus-truth
   $\eta$ mismatch, $TV$-component magnitude, and dropout).
   Type I cells are run at $n_{\text{reps}} = 5{,}000$ for
   MCSE 0.003.

4. **Participant-level interaction signal** recovered by each
   analysis, as a function of the underlying participant-level
   $(m_{BR}, m_{PB}, m_{TV})$ values, illustrating which
   analysis strategies preserve the predictive-biomarker
   stratification that the lumped-total regression does not.

The simulation will use the existing pmsimstats-ng `tidyverse`
implementation collection. Cell-level summaries will be
checkpointed under `analysis/data/`, with per-replicate
output under `analysis/data/quick-sim/06-decomp-replicates.rds`
(prototype) and `analysis/scripts/component-decomposition/output/`
(production). The ADEMP pre-registration is committed at
`analysis/scripts/component-decomposition/00-ademp-pre-registration.md`
prior to the production runs.

# Application to predictive-biomarker validation

The motivating clinical question for the trial designs in this
project is whether a baseline biomarker predicts treatment
response, and specifically whether the biomarker predicts the
*pharmacological* component of treatment response. This is the
precision-medicine question, formalised in the PATH statement
[@kent2020path], and it depends critically on the BR-PB-TV
decomposition.

A direct regression of total response on baseline biomarker

```
total_response = beta_0 + beta_bm * biomarker + error
```

estimates the biomarker's correlation with whatever the lumped
total contains. If $PB$ varies systematically across the
biomarker distribution, because the biomarker is correlated with
optimism, suggestibility, or treatment-seeking history, then
$\hat{\beta}_{bm}$ contains a placebo-prediction component
inseparable from the pharmacological-prediction component. The
resulting biomarker, applied prospectively, would mis-stratify
patients: those flagged as 'high responders' would include both
true high-$BR$ responders and high-$PB$ responders, and the
latter would not benefit from prescription decisions made on a
pharmacological basis.

The right approach is to regress each component on the
biomarker:

\begin{align*}
BR &= \beta_0^{BR} + \beta_{bm}^{BR} \cdot \mathrm{biomarker} +
\varepsilon^{BR},\\
PB &= \beta_0^{PB} + \beta_{bm}^{PB} \cdot \mathrm{biomarker} +
\varepsilon^{PB},\\
TV &= \beta_0^{TV} + \beta_{bm}^{TV} \cdot \mathrm{biomarker} +
\varepsilon^{TV}.
\end{align*}

A biomarker is a useful pharmacological predictor if and only if
$\hat{\beta}_{bm}^{BR}$ is meaningfully non-zero. Significant
$\hat{\beta}_{bm}^{PB}$ or $\hat{\beta}_{bm}^{TV}$ are diagnostic
findings about the biomarker, that it is partly psychological or
correlates with natural-history characteristics, but they are
not the validation of pharmacological prediction. A biomarker
that correlates only with $PB$ would be valuable for selecting
placebo responders in run-in designs but worthless for
predicting pharmacological efficacy; a biomarker that correlates
only with $TV$ would be a prognostic marker, not a predictive
one.

The decomposition is therefore not optional for the precision-
medicine question. It is the formal mechanism by which the
question is asked at all. The phase-augmented linear-mixed
analysis introduced in §5 retains a useful subset of this
machinery at small N (the $bm:Dbc:phase$ interaction tests
whether the biomarker-by-treatment effect attenuates under
blinding, which is the practical proxy for biomarker-by-PB
contamination); the full decomposition recovers the
component-specific regressions explicitly when N permits.

# Discussion

We note that the three-component decomposition is the formal
expression of the claim that 'treatment response' in an
aggregated N-of-1 trial is not a single quantity but a sum of
three causally distinct contributions, each with its own
pharmacological, psychological, or epidemiological
interpretation. The framework
makes the conditions on trial design explicit (which contrasts
identify which components), supplies an analysis-side
methodology that accommodates trial-relevant sample sizes (the
phase-augmented linear-mixed analysis), and clarifies the target
estimand for biomarker validation (the regression of $BR$ on the
biomarker, not the lumped total).

Three open questions remain. First, the simulation study
described in §6 is forthcoming; the present manuscript reports
the framework rather than its empirical evaluation, and the
quantitative bias-versus-power trade-offs across analysis
strategies are conjectures until that work is run. Second, the
$\eta(\text{phase})$ multipliers used to parameterise the
expectancy modulation of $PB$ are themselves modelling
assumptions, and a sensitivity analysis varying $\eta$ across
plausible values is mandatory before any clinical conclusion is
drawn from the framework's outputs. Third, the framework as
specified treats the three components as monotone Gompertz
trajectories; cyclical or biphasic $TV$ patterns require either
flexible $TV$ specifications or explicit covariates beyond the
Gompertz form, and the boundary between 'flexible-Gompertz' and
'covariate-augmented' specifications is not yet sharply drawn.

The companion pedagogical document at `docs/24-component-
decomposition-pedagogy.md` develops the framework with worked
examples and an extended intuition narrative for readers
approaching the material for the first time. The companion
methodological evaluation of the Gompertz family at
`analysis/report/07-gompertz-evaluation/` quantifies the
sensitivity of the framework's inference to the specific
parametric form chosen for the component trajectories.

## Prototype Monte Carlo confirmation

A 100-replicate-per-cell prototype run on a 6-cell slice
($pb\_strength \in \{0, 6.5, 13\} \times$ analysis $\in
\{$one-component, phase-augmented$\}$) was executed at the
reference Hybrid OL+BDC design, $N = 35$, to probe the
predicted bias of the one-component analysis under a
three-component DGP. Per-replicate output at
`analysis/data/quick-sim/06-decomp-replicates.rds`; Morris
summary at `06-decomp-summary.txt`. With 100 replicates the
MCSE on a power estimate near $0.7$ is approximately $0.046$.

The prototype produced an inverted finding relative to the
framework's prediction. The one-component `bm:Dbc` test
showed power $0.95$, $0.93$, and $0.83$ across $pb\_strength
\in \{0, 6.5, 13\}$, while the phase-augmented analysis showed
$0.65$, $0.65$, and $0.61$: the phase-augmented analysis was
\emph{less} powerful than the one-component analysis across
the entire $pb\_strength$ range. Two factors produce the
inversion. First, at $N = 35$ the rank-deficiency of the full
phase-augmented formula required dropping the `Dbc:phase`
interaction term, which removes the precise contrast that
would have isolated the biomarker-by-treatment-by-phase signal
the framework is designed to detect; without that term the
phase-augmented analysis pays the cost of additional fixed
effects without buying the bias-detection mechanism the
framework promises. Second, the predicted bias of the
one-component analysis arises in the regression
\emph{coefficient}, not in the indicator-level rejection rate.
The mean estimated $\hat\beta_{bm:D_{bc}}$ is approximately
$-2.25$ for both analyses across all three $pb\_strength$
values, so the rejection rate on the biomarker-by-treatment
Wald test is not the right summary statistic to detect the
framework's predicted bias; the bias should be read off the
coefficient mean itself, which the prototype's 100-replicate
resolution does not separate from sampling noise at the
$N = 35$ scale.

A canonical run at 1{,}000 replicates per cell, larger $N$
(70 and 100), and recovery of the `Dbc:phase` contrast (via a
phase-stratified two-stage analysis or a richer trial design
that admits the full formula without rank-deficiency) is
required before the framework's bias-versus-power trade-offs
can be quantified at production rigor. The prototype as
configured is a structural check confirming that the analysis
machinery runs end-to-end on the three-component DGP, not a
substantive test of the manuscript's bias-detection claim. The
§6 simulation programme remains the empirical core of this
manuscript.

# Conclusions

Treatment response in aggregated N-of-1 clinical trials is the
sum of three causally distinct components: pharmacological
action ($BR$), placebo-belief response ($PB$), and time-variant
natural history ($TV$). Decomposing the response into these
components is identifiability-conditional, not optional: the
trial design must include phase contrasts that uniquely identify
each component, and the analyst must use a methodology that
exploits those contrasts. At trial-relevant sample sizes, a
phase-augmented linear-mixed analysis captures the practically
important part of the decomposition without paying the parametric
cost of a full nonlinear-mixed-effects fit. For
predictive-biomarker validation, the decomposition is mandatory:
a biomarker that predicts the lumped total is not, in general,
a pharmacological predictor, and the regulatory and prescribing
decisions that motivate the trial cannot be made without
separating $BR$ from $PB$ and $TV$.

# Conflict of interest

The authors declare no conflicts of interest.

# Funding

This work received no specific funding.

# Data availability statement

The data and code that support the findings of this study are
openly available in the pmsimstats-ng project repository.
Per-replicate Monte Carlo outputs and ADEMP-compliant cell-level
summaries are versioned under `analysis/data/`; simulation
drivers are at `analysis/scripts/`. Exact paths for this
manuscript are listed in the Reproducibility section below.

# Reproducibility

This manuscript was rendered with R 4.4.x inside the project's
zzcollab Docker container (image hash recorded in `Dockerfile`;
package set captured in `renv.lock`). Principal packages used
by the simulation pipeline are `nlme` (continuous-time
linear-mixed analysis with `corCAR1` residual correlation),
`parallel` (8-core `mclapply` for parameter-grid sweeps),
`data.table` (the `original-extended` simulation collection),
`furrr` and `purrr` (the `tidyverse` simulation collection),
and `rmarkdown` plus `bookdown` for the manuscript build.

**Random-number generator.** Each simulation driver sets
`set.seed(20260507)` at the top of the replicate loop and uses
`parallel::mc.set.seed` inside `mclapply`. Per-replicate seeds
derive from the master seed by `+ rep_idx`.

**Data and code availability.** Per-replicate simulation
outputs are checkpointed at
`analysis/data/quick-sim/06-decomp-replicates.rds`; Morris ADEMP
cell-level summaries at the companion `06-decomp-summary.txt`.
Driver scripts are under
`analysis/scripts/component-decomposition/`. The pre-registered
ADEMP plan is at
`analysis/scripts/component-decomposition/00-ademp-pre-registration.md`.
The manuscript source, paper-specific bibliography
(`references.bib`), and SIM CSL file are at
`analysis/report/06-component-decomposition/`.
